### 2020-07-05 paper/policar2020-opentsne.tex
Add initial working version of paper

%% BioMed_Central_Tex_Template_v1.06
%%                                      %
%  bmc_article.tex            ver: 1.06 %
%                                       %

%%IMPORTANT: do not delete the first line of this template
%%It must be present to enable the BMC Submission system to
%%recognise this template!!

%%% additional documentclass options:
%  [doublespacing]
%  [linenumbers]   - put the line numbers on margins

\documentclass[twocolumn]{bmcart}
%\documentclass{bmcart}

%%% Load packages
%\usepackage{amsthm}
%\RequirePackage{natbib}
%\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
\usepackage{hyperref}
\usepackage[utf8]{inputenc} %unicode support
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails

\usepackage{graphicx}
\usepackage{minted}
\usemintedstyle{vs}

\usepackage{amsmath,amssymb}
\usepackage{booktabs}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                             %%
%%  If you wish to display your graphics for   %%
%%  your own use using includegraphic or       %%
%%  includegraphics, then comment out the      %%
%%  following two lines of code.               %%
%%  NB: These line *must* be included when     %%
%%  submitting to BMC.                         %%
%%  All figure files must be submitted as      %%
%%  separate graphics through the BMC          %%
%%  submission process, not included in the    %%
%%  submitted article.                         %%
%%                                             %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\def\includegraphic{}
%\def\includegraphics{}

%%% Put your definitions there:
\startlocaldefs
\endlocaldefs

\begin{document}

%%% Start of article front matter
\begin{frontmatter}

\begin{fmbox}
\dochead{Software}

\title{openTSNE: a modular Python library for t-SNE dimensionality reduction and embedding}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors here                   %%
%%                                          %%
%% Specify information, if available,       %%
%% in the form:                             %%
%%   <key>={<id1>,<id2>}                    %%
%%   <key>=                                 %%
%% Comment or delete the keys which are     %%
%% not used. Repeat \author command as much %%
%% as required.                             %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\author[
   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
   corref={aff1},                       % id of corresponding address, if any
   %noteref={n1},                        % id's of article notes, if any
   email={pavlin.policar@fri.uni-lj.si}   % email address
]{\inits{PGP}\fnm{Pavlin G.} \snm{Poli\v{c}ar}}
\author[
   addressref={aff1,aff2}
]{\inits{MS}\fnm{Martin} \snm{Stra\v{z}ar}}
\author[
   addressref={aff1,aff3}
]{\inits{BZ}\fnm{Bla\v{z}} \snm{Zupan}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors' addresses here        %%
%%                                          %%
%% Repeat \address commands as much as      %%
%% required.                                %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\address[id=aff1]{%
  \orgname{Faculty of Computer and Information Science, University of Ljubljana},
  %\street{Ve\v{c}na Pot 113},
  \postcode{SI 1000}
  \city{Ljubljana},
  \cny{Slovenia}
}
\address[id=aff2]{%
  \orgname{Broad Institute of Harvard and MIT},
  %\street{D\"{u}sternbrooker Weg 20},
  \postcode{MA 02142}
  \city{Cambridge},
  \cny{U.S.A.}
}
\address[id=aff3]{%
  \orgname{Department of Molecular and Human Genetics, Baylor College of Medicine},
  %\street{D\"{u}sternbrooker Weg 20},
  \postcode{TX 77030}
  \city{Houston},
  \cny{U.S.A.}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter short notes here                   %%
%%                                          %%
%% Short notes will be after addresses      %%
%% on first page.                           %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\begin{artnotes}
%\note{Sample of title note}     % note to the article
%\note[id=n1]{Equal contributor} % note, connected to author
%\end{artnotes}

\end{fmbox}% comment this for two column layout

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The Abstract begins here                 %%
%%                                          %%
%% Please refer to the Instructions for     %%
%% authors on http://www.biomedcentral.com  %%
%% and include the section headings         %%
%% accordingly for your article type.       %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{abstractbox}

\begin{abstract}
Point-based visualisations of large, multi-dimensional data from
molecular biology can reveal meaningful clusters. One of the
most popular techniques to construct such visualisations is t-distributed
stochastic neighbor embedding (t-SNE), for which a number of extensions have
recently been proposed to address issues of scalability and the quality of the resulting
visualisations. We introduce openTSNE, a modular Python library that implements
the core t-SNE algorithm and its extensions. The library is orders of magnitude
faster than existing popular implementations, including those from
scikit-learn. Unique to openTSNE is also the mapping of new data to existing embeddings, which
can surprisingly assist in solving batch effects.
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The keywords begin here                  %%
%%                                          %%
%% Put each keyword in separate \kwd{}.     %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{keyword}
\kwd{sample}
\kwd{article}
\kwd{author}
\end{keyword}

% MSC classifications codes, if any
%\begin{keyword}[class=AMS]
%\kwd[Primary ]{}
%\kwd{}
%\kwd[; secondary ]{}
%\end{keyword}

\end{abstractbox}
%
%\end{fmbox}% uncomment this for twcolumn layout

\end{frontmatter}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The Main Body begins here                %%
%%                                          %%
%% Please refer to the instructions for     %%
%% authors on:                              %%
%% http://www.biomedcentral.com/info/authors%%
%% and include the section headings         %%
%% accordingly for your article type.       %%
%%                                          %%
%% See the Results and Discussion section   %%
%% for details on how to create sub-sections%%
%%                                          %%
%% use \cite{...} to cite references        %%
%%  \cite{koon} and                         %%
%%  \cite{oreg,khar,zvai,xjon,schn,pond}    %%
%%  \nocite{smith,marg,hunn,advi,koha,mouse}%%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%% start of article main body
% <put your article body there>

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}

%We implement t-SNE~\cite{maaten2008visualizing}. We include a set of recommendations and tricks from Kobak and Berens~\cite{kobak2019art} as well as incerasing the learning rate as recommended by Belkina \textit{et al.}~\cite{belkina2019automated}. Also support using variable degrees of freedom by Kobak \textit{et al.}~\cite{kobak2019heavy}.
%
%Two approximation schemes, one based on Barnes-Hut~\cite{van2014accelerating} and one on the interpolation-based approach~\cite{linderman2019fast}.
%
%UMAP~\cite{2018arXivUMAP}, whether or not it better preserves global structure or not is in question~\cite{kobak2019umap}.

%% TODO: START FROM FIRST PAPER

The abundance of high-dimensional data sets in molecular biology calls for
techniques for dimensionality reduction, and in particular for methods that can
help in the construction of data visualizations. Popular approaches for
dimensionality reduction include principal component analysis, multidimensional
scaling, and uniform manifold approximation and projections~\cite{2018arXivUMAP}. Among
these, t-distributed stochastic neighbor embedding
(t-SNE)~\cite{maaten2008visualizing} lately received much attention as it can
address high volumes of data and reveal meaningful clustering structure. Most
of the recent reports on single-cell gene expression data start with an
overview of the cell landscape, where t-SNE embeds high-dimensional expression
profiles into a two-dimensional space~\cite{macosko2015highly,cao2019single,tasic2018shared}.
Fig.~\ref{fig:tsne}.a presents an example of one such embedding.

Despite its utility, t-SNE has been criticized for poor scalability
when addressing large data sets, lack of global organization \em t-SNE
focuses on local clusters that are arbitrarily scattered in the
low-dimensional space \em and absence of theoretically-founded implementations to
map new data into existing embeddings~\cite{ding2018interpretable,becht2019dimensionality}. Most of these shortcomings, have
recently been addressed.
\cite{fi_tsne} sped-up the method through an interpolation-based
approximation, reducing the time complexity to be merely linear dependent on the number of samples. \cite{art_of_using_tsne} proposed
several techniques to improve global positioning, including estimating similarities with a mixture of
Gaussian kernels. While no current popular
software library supports mapping of new data into reference embedding,
\cite{parametric_tsne} proposed a related approach
using neural networks.

%% TODO: END FROM FIRST PAPER

\begin{figure*}[htbp]
  \includegraphics[width=\textwidth]{macosko2015}
  \caption{\label{fig:macosko}}
\end{figure*}

\section*{Results}

We introduce openTSNE, a comprehensive Python library that
implements t-SNE and all its recently proposed extensions. The library is
compatible with the Python data science ecosystem (e.g., {\tt numpy}, {\tt
sklearn}, {\tt scanpy}). Its modular design encourages extendibility and
experimentation with various settings and changes in the analysis pipeline.

\begin{figure*}[htbp]
  \includegraphics[width=\textwidth]{tasic2018}
  \caption{\label{fig:tasic}We use openTSNE to create four different visualizations of the Tasic \textit{et al.}~\cite{tasic2018shared} data set containing 21,874 single-cells originating from the mouse neocortex. Cluster annotations and colors are taken from the original publication. Warm colors correspond to excitatory neurons, cool colors correspond to inhibitory neurons, and gray/brown colors correspond to non-neuronal cells. The standard t-SNE \textbf{(a)} emphasizes local structure, while increasing perplexity \textbf{(b)} results in a more meaningful layout of the clusters. We can also combine the two perplexities by using a multiscale kernel affinity model \textbf{(c)} and obtain a good trade-off between global and local structure. Alternatively, we can inspect more fine-grained structure and reveal smaller clusters by using a more heavy-tailed kernel \textbf{(d)}.}
\end{figure*}

The following code is used to generate Fig.~\ref{fig:tasic}
\begin{minted}[fontsize=\footnotesize]{python}
# Perplexity 30 (standard t-SNE)
perp_30 = openTSNE.TSNE(
    perplexity=30, metric="cosine"
).fit(X)

# Perplexity 500 (emphasise global structure)
perp_500 = openTSNE.TSNE(
    perplexity=500, metric="cosine"
).fit(X)

# Multiscale kernel with perplexities 30,500
affinities = openTSNE.affinity.Multiscale(
    X, perplexities=[30, 500], metric="cosine"
)
init = openTSNE.initialization.pca(X)
ms = openTSNE.TSNEEmbedding(init, affinities)
ms = ms.optimize(n_iter=250, exaggeration=12)
ms = ms.optimize(n_iter=250, exaggeration=1)

# Expose finer structure using smaller dof
perp_30 = openTSNE.TSNE(
    perplexity=30, metric="cosine", dof=0.6
).fit(X)
\end{minted}

\begin{figure*}[htbp]
  \includegraphics[width=\textwidth]{cao2019}
  \caption{\label{fig:cao}The data set from Cao \textit{et al.}~\cite{cao2019single} contains expression profiles from 2,058,652 single cells. The data were collected from mice embryos at different developmental time-points at daily intervals after 9.5 to 13.5 days. The exaggeration parameter can be used to highlight the global organization of the clusters and enforce cluster compactness. \textbf{(a)} shows the data is comprised of two main components -- the neural tube and mesenchymal cells -- as well as several other smaller clusters. Additionally, the colors indicate developmental progression with red indicating least-developed cells and blue indicating most developed cells. The overall developmental trajectory is most apparent with higher levels of exaggeration, showing red cells slowly transitioning into blue cells. Progressively easing the exaggeration factor uncovers finer clusters within the larger groups as shown in \textbf{(b)} with exaggeration 2 and subsequently in \textbf{(c)}, where we show the standard t-SNE with no exaggeration. 32,011 putative doublets are excluded from the visualizations.}
\end{figure*}

The following code is used to generate Fig.~\ref{fig:cao}
\begin{minted}[fontsize=\footnotesize]{python}
indices = np.random.permutation(list(range(x.shape[0])))
reverse = np.argsort(indices)
# Split the data into a sample/remainder sets
x_sample, x_rest = x[indices[:25000]], x[indices[25000:]]
y_sample, y_rest = y[indices[:25000]], y[indices[25000:]]
# Create an initial embedding using a subset of the data
sample_embedding = openTSNE.TSNE(
    initialization="spectral",
    perplexity=500,
    metric="cosine",
).fit(x_sample)
# Position the remainder of the points to their nearest
# neighbor in the sample embedding
rest_init = sample_embedding.prepare_partial(x_rest, k=1)
init_full = np.vstack(
    [sample_embedding, rest_init]
)[reverse]
init_full = openTSNE.initialization.rescale(init_full)
# Calculate the affinity model using all the data points
affinities = openTSNE.affinity.PerplexityBasedNN(
    x, perplexity=30, metric="cosine"
)

embedding = openTSNE.TSNEEmbedding(
    init_full, affinities
%)
exag12 = embedding.optimize(n_iter=500, exaggeration=12)
exag4 = exag12.optimize(n_iter=500, exaggeration=4)
exag2 = exag4.optimize(n_iter=500, exaggeration=2)
exag1 = exag2.optimize(n_iter=500, exaggeration=1)
\end{minted}

\begin{figure*}[htbp]
  \includegraphics[width=\textwidth]{transform_hochgerner}
  \caption{\label{fig:transform}openTSNE enables the user to embed new samples into an existing, reference embedding. We construct a t-SNE embedding for the data from Hochgerner \textit{et al.}~\cite{hochgerner2018conserved} containing 24,185 developing, single cells from the mouse hippocampus. The data contains different types of neurons, supporting glia, and other vascular cells (upper left). Cell annotations are taken from the original publications. The colors correspond to Fig.~\ref{fig:tasic}. We then embed new, hippocampal cells collected by Harris \textit{et al.}~\cite{harris2018classes} using the previous embedding as a reference. In their study, Harris \textit{et al.} collected 6,971 single-cells and focused on identifying different types of inhibitory neurons. However, almost half of the collected cells are not neurons and were left uncharacterized. Inspecting where these cells are embedded in the reference embedding (bottom left) reveals that in addition to inhibitory neurons, the data contains a number of supporting glia as well as a small population of endothelial cells. We can verify the accuracy of our approach by inspecting marker genes for the major cell types in the reference (top row) and embedded samples (bottom row).}
\end{figure*}

The following code is used to generate Fig.~\ref{fig:transform}
\begin{minted}[fontsize=\footnotesize]{python}
# Assuming we have prepared a reference embedding and
# assuming we have aligned new_X to the reference X matrix
new_embedding = reference_embedding.transform(new_X)
\end{minted}

We want to mention that this works for mitigating batch effects in our earlier work~\cite{policar2019embedding}.

\begin{figure*}[htbp]
  \includegraphics[width=\textwidth]{benchmarks}
  \caption{\label{fig:benchmarks}We benchmark openTSNE against three popular open-source implementations including scikit-learn~\cite{pedregosa2011scikit}, MulticoreTSNE~\cite{Ulyanov2016}, and FIt-SNE~\cite{linderman2019fast}. Benchmarks were run on a consumer-grade Intel Core i7-7700HQ processor found in laptop computers, and on a server-grade Intel Xeon E5-2650. To generate benchmark data sets of different sizes, we subsampled data from the 10X Genomics 1.3 million mouse brain data set five times, resulting in five different data sets for each size. In total, each implementation was run on 30 different data sets. Notice that openTSNE scales similarly to FIt-SNE, as they both use the same interpolation-based approximation scheme, while scikit-learn and MulticoreTSNE utilize the Barnes-Hut approximation.}
\end{figure*}

\section*{Discussion}
Text for this section \ldots

\section*{Conclusion}
Text for this section \ldots

\section*{Methods}

% approach name in bold, reference, description of the approach and why is
% it useful. focus on theory and math.

\subsection*{t-SNE}
%% TODO: Reword this entirely

Local, non-linear dimensionality reduction by t-SNE is performed as follows.
Given a multi-dimensional data set $\mathbf{X} = \left \{ \mathbf{x}_1,
\mathbf{x}_2, \dots, \mathbf{x}_N \right \} \in \mathbb{R}^D$ where $N$ is the
number of data points in the reference data set, t-SNE aims to find a low
dimensional embedding $\mathbf{Y} = \left \{ \mathbf{y}_1, \mathbf{y}_2, \dots,
\mathbf{y}_N \right \} \in \mathbb{R}^d$ where $d \ll D$, such that if points
$\mathbf{x}_i$ and $\mathbf{x}_j$ are close in the multi-dimensional space,
their corresponding embeddings $\mathbf{y}_i$ and $\mathbf{y}_j$ are also
close. Since t-SNE is primarily used as a visualization tool, $d$ is typically
set to two. The similarity between two data points in t-SNE is defined as:

\begin{equation}
p_{j \mid i} = \frac{\exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{x}_j ) / \sigma_i^2 \right )}
{\sum_{k \neq i } \exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{x}_k ) / \sigma_i^2 \right )}, \quad p_{i \mid i} = 0
\label{eq:gaussian_kernel}
\end{equation}

\noindent where $\mathcal{D}$ is a distance measure. This is then symmetrized to

\begin{equation}
p_{ij} = \frac{p_{j \mid i} + p_{i \mid j}}{2N}.
\label{eq:symmetrize}
\end{equation}

The bandwidth of each Gaussian kernel $\sigma_i$ is selected such that the
perplexity of the distribution matches a user-specified parameter value

\begin{equation}
\text{Perplexity} = 2^{H(P_i)}
\end{equation}

\noindent where $H(P_i)$ is the Shannon entropy of $P_i$,

\begin{equation}
H(P_i) = -\sum_i p_{j \mid i} \log_2 (p_{j \mid i}).
\end{equation}

\noindent Different bandwidths $\sigma_i$ enable t-SNE to adapt to the varying
density of the data in the multi-dimensional space.

The similarity between points $\mathbf{y}_i$ and $\mathbf{y}_j$ in the
embedding space is defined using the $t$-distribution with one degree of
freedom

\begin{equation}
q_{ij} = \frac{\left ( 1 + || \mathbf{y}_i - \mathbf{y}_j ||^2 \right )^{-1}}
{\sum_{k \neq l}\left ( 1 + || \mathbf{y}_k - \mathbf{y}_l ||^2 \right )^{-1}},
\quad q_{ii} = 0.
\label{eq:cauchy_kernel}
\end{equation}

The t-SNE method finds an embedding $\mathbf{Y}$ that minimizes the
Kullback-Leibler (KL) divergence between $\mathbf{P}$ and $\mathbf{Q}$,

\begin{equation}
C = \text{KL}(\mathbf{P} \mid \mid \mathbf{Q}) = \sum_{ij} p_{ij} \log \frac{p_{ij}}{q_{ij}}.
\label{eq:kl_divergence}
\end{equation}

It is then easy to show that the corresponding gradient takes on the form
\begin{equation}
\frac{\partial C}{\partial \mathbf{y}_i} = 4 \sum_{j \neq i} \left ( p_{ij} - q_{ij} \right ) \left ( \mathbf{y}_i - \mathbf{y}_j \right ) \left ( 1 + || \mathbf{y}_i - \mathbf{y}_j || ^2 \right )^{-1}.
\label{eq:tsne_gradient}
\end{equation}

Optimization is performed using batch gradient descent with the delta-bar-delta update rule for increased convergence speed~\cite{jacobs1988increased}. Originally, t-SNE was run for 1000 iterations consisting of two phases. In the first \textit{early exaggeration} phase, the attractive forces between data points is increased by a factor $\rho=12$ so that points can move throughout the embedding uninhibited by repulsive forces. Then the remaining 750 iterations are run with $\rho=1$, which reverts the attractive forces to their original values.

Belkina \textit{et al.} later found that the number of iterations can saftely be decreased to 750 iterations by increasing the learning rate to $\eta=N/12$ instead of the default and somewhat arbitrarily set $\eta=200$~\cite{belkina2019automated}. A welcome side effect of this reduction is also faster runtime.

\subsection*{Efficient Implementation}
A direct evaluation of t-SNE gradients requires $\mathcal{O}(N^2)$ operations, which makes its application impractical to any reasonably-sized data set and beckons for the development of efficient approximation schemes. Van der Maaten observed that the t-SNE gradient may be reformulated in terms of a N-body problem where data points may be interpreted as particles which attract and repel each other~\cite{van2014accelerating}. The gradient may be rewritten as
\begin{equation}
\frac{\partial C}{\partial \mathbf{y}_i} = 4 \left [ \sum_{j \neq i} p_{ij} q_{ij} Z \left ( \mathbf{y}_i - \mathbf{y}_j \right ) -\sum_{j \neq i} q_{ij}^2 Z \left ( \mathbf{y}_i - \mathbf{y}_j \right ) \right ], \label{eq:grad_attr_rep}
\end{equation}
where $Z = \sum_{k \neq l}\left ( 1 + || \mathbf{y}_k - \mathbf{y}_l ||^2 \right )^{-1}$. The two terms may be viewed as the attractive and repulsive forces between particles, respectively. Each term lends itself to efficient approximations, which enable us to greatly reduce the time complexity of t-SNE.

\subsubsection*{Attractive Forces}

Van der Maaten observed that since $p_{ij}$ is calculated in the high-dimensional space using a Gaussian kernel, and that the bandwidth is selected such that only a fixed number of closest neighbors (determined by the perplexity parameter) will contribute to the attractive forces of a given point due to the exponential decay of the Gaussian kernel. Therefore it is sufficient to calculate the attractive forces only over a relatively small number of nearest neighbors instead of all $N$ points. By utilizing tree-based nearest-neighbor search methods, the time complexity is thus reduced to $\mathcal{O}(N \log N)$. Linderman \textit{et al.} further realised that, qualitatively, embeddings are visually virtually indistinguishable when using only \textit{approximate} nearest neighbors, further reducing time complexity to $\mathcal{O}(N)$~\cite{linderman2019fast}.

\subsubsection*{Repulsive Forces}

Examining the second term of Eqn.~(\ref{eq:grad_attr_rep}) we notice that each point indiscriminately exerts a repulsive force on all other points. Van der Maaten proposed an approach based on N-body simulations and used a space-partitioning Barnes-Hut tree approach to approximate the interaction between data points. Briefly, the approach splits the space into quadrants (in the 2D case) and if a given point is far enough away from a quadrant, the points in that quadrant are summarized by a single center of mass. This reduces the time complexity from $\mathcal{O}(N^2)$ to $\mathcal{O}(N \log N)$.

More recently, Linderman \textit{et al.} proposed an alternative approach, FIt-SNE, based on non-uniform convolutions to calculating all pairwise interactions between repelling data points. Briefly, Linderman \textit{et al.} observed that the repulsive forces $\mathbf{R}$ may be rewritten as
\begin{align}
\mathbf{R}_i &= \sum_{j \neq i} q_{ij}^2 Z \left ( \mathbf{y}_i - \mathbf{y}_j \right ) \notag \\
&= \sum_{j \neq i} \frac{\mathbf{y}_i - \mathbf{y}_j}{\left ( 1 + || \mathbf{y}_i - \mathbf{y}_j ||^2 \right )^{2}}
\bigg/
\sum_{k \neq l} \frac{1}{1 + || \mathbf{y}_k - \mathbf{y}_l || ^2} \label{eq:fitsne_rep}
\end{align} 
and computed by evaluating three terms
\begin{align}
\phi_{1, j} &= \sum_{j \neq i} \frac{1}{1 + || \mathbf{y}_j - \mathbf{y}_i ||^2}, \notag \\
\phi_{2, j} &= \sum_{j \neq i} \frac{\mathbf{y}_j}{\left( 1 + || \mathbf{y}_j - \mathbf{y}_i ||^2 \right)^2}, \notag \\
\phi_{3, j} &= \sum_{j \neq i} \frac{1}{\left( 1 + || \mathbf{y}_j - \mathbf{y}_i ||^2 \right)^2}. \label{eq:fitsne_terms}
\end{align}
Then the numerator and denominator of Eqn.~(\ref{eq:fitsne_rep}) can be computed as $\mathbf{y}_i \phi_{1,j} - \phi_{2,j}$ and $Z = \sum_j \phi_{3,j}$, respectively. These interactions are accelerated by interpolating these terms through a grid of equispaced interpolation points. This shifts the computational burden onto the interpolation points and reduces the time complexity to $\mathcal{O}(N)$.

openTSNE provides efficient implementations for all the aforementioned approximations but defaults to using approximate nearest neighbor search and the non-uniform convolutions approach as these enable the user to quickly create informative visualizations for data sets containing up to millions of samples. We would note here that while the default choices introduce some runtime overhead on smaller data sets, this effect is minimal and impacts runtime by only a few seconds compared to the exact nearest-neighbor search Barnes-Hut approximation.

\subsection*{Embedding New Samples}

Unlike other popular dimensionality reduction methods such as principal component analysis or autoencoders, t-SNE is a non-parametric method and does not define an explicit mapping to the embedding space. Therefore embeddings of new data points need to be found through the use of optimization techniques~\cite{policar2019embedding}. When adding new data points to an existing, reference embedding, the reference data points are fixed in place and the new data points are allowed to find their respective positions. The optimization remains the same as in standard t-SNE with only slight modifications to $p_{ij}$ and $q_{ij}$
\begin{align}
p_{j \mid i} &= \frac{\exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{v}_j) /  \sigma_i^2 \right )}{\sum_{i} \exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{v}_j) / \sigma_i^2 \right )}, \\
q_{j \mid i} &= \frac{\left ( 1 + || \mathbf{y}_i - \mathbf{w}_j ||^2 \right )^{-1}}{\sum_{i}\left ( 1 + || \mathbf{y}_i - \mathbf{w}_j ||^2 \right )^{-1}},
\end{align}
\noindent where $\mathbf{V} = \left \{ \mathbf{v}_1, \mathbf{v}_2, \dots,
\mathbf{v}_M \right \} \in \mathbb{R}^D$ where $M$ is the number of samples in
the secondary data set and $\mathbf{W} = \left \{ \mathbf{w}_1, \mathbf{w}_2, \dots,
\mathbf{w}_M \right \} \in \mathbb{R}^d$. Additionally, we omit the
symmetrization step in Eqn.~(\ref{eq:symmetrize}). The gradients of
$\mathbf{w}_j$ with respect to the loss (Eqn.~\ref{eq:kl_divergence}) are:
\begin{equation}
\frac{\partial C}{\partial \mathbf{w}_j} = 2 \sum_i \left ( p_{j \mid i} - q_{j \mid i} \right ) \left ( \mathbf{y}_i - \mathbf{w}_j \right ) \left ( 1 + || \mathbf{y}_i - \mathbf{w}_j || ^2 \right )^{-1}
\label{eq:gradient}
\end{equation}

As in standard t-SNE, a direct calculation of gradients takes $\mathcal{O}(N \cdot M)$ time, but it is straightforward to adapt the Barnes-Hut and FIt-SNE approximation schemes, reducing the time complexity to $\mathcal{O}(M \log N)$ and $\mathcal{O}(\max \{ N, M \})$, respectively. In the FIt-SNE approximation scheme, we additionally exploit the fact that the reference embedding remains fixed throughout the optimization of newly added points, and precompute the interpolation grid. This further reduces the runtime complexity from $\mathcal{O}(\max \{ N, M \})$ to $\mathcal{O}(M)$.

openTSNE is currently the only open-source t-SNE implementation allowing users to add new samples into existing embeddings.

\subsection*{Alternative Perplexity Kernels}

In standard t-SNE, distances are converted to similarities through the use of Gaussian kernels of varying bandwidths. The bandwidths are indirectly determined by the user-provided perplexity parameter, so that a fixed number of nearest data points will have non-zero similarity. One common trick for uncovering the global relations between clusters is to increase perplexity such that more long-range interactions are preserved in the final embedding. However, one unfortunate side effect of increasing perplexity is that smaller clusters get absorbed into larger ones, and less overall clusters are revealed.

Kobak \textit{et al.}~\cite{kobak2019art} suggest that replacing the Gaussian kernel with a mixture of Gaussians may provide better insight into both the local and global structure. Therefore, the affinities in the input space become

\begin{align}
  p_{j\mid i} &\propto \frac{1}{\sigma_{1,i}} \exp \left ( -\mathcal{D}(\mathbf{x}_i, \mathbf{x}_j ) / 2\sigma_{1,i}^2 \right ) \notag \\
  &+ \frac{1}{\sigma_{2,i}} \exp \left ( -\mathcal{D}(\mathbf{x}_i, \mathbf{x}_j ) / 2\sigma_{2,i}^2 \right ).
\end{align}

The bandwidths of each Gaussian $\sigma_{1,i}$ and $\sigma_{2,i}$ is determined by selecting different perplexity values. They show that using a kernel with perplexity 500 captures the long-range interactions which preserve global cluster organization in the final embedding, and combine this with a kernel with perplexity 50, which prevents small, well-defined clusters from being absorbed into larger ones.

\subsection*{Variable Degrees of Freedom}

Standard t-SNE is often used to reveal clustering structure in data at one level of resolution. Different perplexity parameter values can be used to identify global cluster relationships or small, well-isolated groups. Unfortunately, varying perplexity values can be time-consuming as this involves recomputing the KNNG which is often the most expensive part of the t-SNE algorithm. Alternatively,  Kobak \textit{et al.}~\cite{kobak2019heavy} suggest that varying the degree of freedom in the t-distribution used in the embedding space during optimization can be used to explore the clustering structure at different levels of resolution.

Standard t-SNE models similarities between data points in the embedding space using a t-distribution with a single degree of freedom

\begin{equation}
q_{ij} \propto \left ( 1 + || \mathbf{y}_i - \mathbf{y}_j ||^2 / \alpha \right )^{-\alpha} = \frac{1}{\left( 1 + || \mathbf{y}_i - \mathbf{y}_j ||^2 / \alpha \right)^\alpha }.
\end{equation}

In standard t-SNE $\alpha=1$ so this simplifies to Eqn.~(\ref{eq:cauchy_kernel}).

The gradient of the loss function then becomes
\begin{align}
\frac{\partial C}{\partial \mathbf{y}_i} &= 4 \sum_{j \neq i} \left ( p_{ij} - q_{ij} \right ) q_{ij}^{1/\alpha} \left ( \mathbf{y}_i - \mathbf{y}_j \right ),
\end{align}
which can, again, be decomposed into the attractive and repulsive forces as
\begin{align}
\frac{\partial C}{\partial \mathbf{y}_i} &= 4 \sum_{j \neq i} p_{ij} q_{ij}^{1/\alpha} (\mathbf{y}_i - \mathbf{y}_j) \notag \\
&- 4 \sum_{j \neq i} q_{ij}^{\frac{\alpha+1}{\alpha}} / Z (\mathbf{y}_i - \mathbf{y}_j).
\end{align}

Adapting existing approximation schemes to this formulation is straightforward. We provide efficient implementations of both the Barnes-Hut and the FIt-SNE approximation schemes, where we replace the terms from Eqn.~(\ref{eq:fitsne_terms}) with

\begin{align}
\phi_{1,j} &= \sum_{j \neq i} \frac{1}{\left( 1 + || \mathbf{y}_j - \mathbf{y}_i ||^2 / \alpha \right)^{\alpha+1}}, \notag \\
\phi_{2,j} &= \sum_{j \neq i} \frac{\mathbf{y}_j}{\left( 1 + || \mathbf{y}_j - \mathbf{y}_i ||^2 / \alpha \right)^{\alpha+1}}, \notag \\
\phi_{3,j} &= \sum_{j \neq i} \frac{1}{\left( 1 + || \mathbf{y}_j - \mathbf{y}_i ||^2 / \alpha \right)}. \notag
\end{align}

\subsection*{Variable Exaggeration}

Embeddings produced by standard t-SNE often produce clusters separated by thin boundaries and make use of all available space. While this is often desirable, this often obscures the global relationships between clusters as all neighboring clusters appear at the same distance from one another. Other dimensionality reduction methods such as UMAP~\cite{2018arXivUMAP} or ForceAtlas2~\cite{jacomy2014forceatlas2} tend to produce embeddings where clusters appear more compact and the white-space separating the clusters may be interpreted at least partially as a loose measure of distance.

Kobak \textit{et al.}~\cite{todo} showed that the early exaggeration factor $\rho$ in t-SNE may be increased to produce more compact layouts often seen in UMAP or ForceAtlas2. Early exaggeration is a trick introduced by van der Maaten~\cite{maaten2008visualizing} used in the first 250 iterations and increases the attractive forces between data points, enabling neighboring points to move close to one another while ignoring the repulsive forces between data points. In standard t-SNE, the remainder of the iterations are run with $\rho=1$. Kobak \textit{et al.} showed that using $\rho=4$ and $\rho=30$ in the second phase of the optimization result in embeddings visually similar to UMAP embeddings, and ForceAtlas2 embeddings, respectively.

While it is difficult to claim one is better than the other, different settings of $\rho$ may uncover different properties and clustering structures. For example, in single-cell data, standard t-SNE with $\rho=1$ often uncovers distinct groups of cell-types but obscures developmental paths. This problem is exacerbated when dealing with large numbers of data points. Conversely, ForceAtlas2 has often been used to uncover trajectories and transitions between cell types, but larger clusters often absorb small, distinct groups of cells.

\subsection*{Alternative Initialization Schemes}

\subsection*{Ease of Use}

Widespread adoption of novel techniques is almost universally accompanied by easy-to-use and easy-to-install software. While the t-SNE implementation in scikit-learn and Rtsne fits this requirement, it implements the outdated Barnes-Hut approximation scheme which is unsuitable for larger data sets containing up to millions of data points. Other C++ implementations such as MulticoreTSNE and FIt-SNE are more suitable for larger data sets, but do not provide precompiled binaries to their software, requiring users to compile the software themselves. This is problematic from a usability standpoint, as many users often either do not know how to compile C++ source code or use Windows, which does not include a C++ compiler by default.

openTSNE is a Python package which focuses on ease-of-use and extensibility. We provide precompiled binaries for all major Python version on all major platforms, making the installation process as seamless as possible. openTSNE can be installed through the Python Package Index (PyPI) or through conda on the conda-forge channel. openTSNE is designed to be familliar to Python users and loosely follows the scikit-learn API, which is well established in the Python data-science ecosystem. openTSNE implements both the Barnes-Hut and FIt-SNE approximation schemes, allowing it to be used on data sets containing millions of data points. While the Python virtual machine introduces some overhead, the runtime is comparable to that of its pure C++ counterpart -- FIt-SNE. Finally, openTSNE is extensible. Its modular interface enables researchers to quickly experiment with different settings and easily incorporate their own components into the software.

\begin{description}
\item[Alternative initialization schemes] Spectral and PCA, cite dmitry and george paper

\item[Incorporates latest developments] $lr=N/12, n_iter=750$

\end{description}

\begin{table}

\newcommand*\rot{\rotatebox{90}}
\renewcommand{\arraystretch}{1.25}

\begin{tabular}{r c c c c c c}
\setlength\tabcolsep{6pt}
& \rot{scikit-learn} & \rot{MulticoreTSNE} & \rot{FIt-SNE} & \rot{openTSNE} \\
\toprule
PyPI package & \checkmark & \checkmark & & \checkmark \\
conda package & \checkmark & & & \checkmark \\
Precompiled binaries & \checkmark & & & \checkmark \\
\hline
$\mathcal{O}(N \log N)$ & \checkmark & \checkmark & & \checkmark \\
$\mathcal{O}(N)$ & & & \checkmark & \checkmark \\
\hline
Variable degrees of freedom & & & \checkmark & \checkmark \\
Variable exaggeration & & & \checkmark & \checkmark \\
Multiscale Gaussian kernels & & & \checkmark & \checkmark \\
Custom affinity kernels & & & & \checkmark \\
Adding new samples & & & & \checkmark \\
\bottomrule
\end{tabular}
\end{table}

\section*{Availability}
Text for this section \ldots

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Backmatter begins here                   %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{backmatter}

\section*{Competing interests}
  The authors declare that they have no competing interests.

\section*{Author's contributions}
    Text for this section \ldots

\section*{Acknowledgements}
  Text for this section \ldots
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                  The Bibliography                       %%
%%                                                         %%
%%  Bmc_mathpys.bst  will be used to                       %%
%%  create a .BBL file for submission.                     %%
%%  After submission of the .TEX file,                     %%
%%  you will be prompted to submit your .BBL file.         %%
%%                                                         %%
%%                                                         %%
%%  Note that the displayed Bibliography will not          %%
%%  necessarily be rendered by Latex exactly as specified  %%
%%  in the online Instructions for Authors.                %%
%%                                                         %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% if your bibliography is in bibtex format, use those commands:
\bibliographystyle{bmc-mathphys} % Style BST file (bmc-mathphys, vancouver, spbasic).
\bibliography{references}      % Bibliography file (usually '*.bib' )
% for author-year bibliography (bmc-mathphys or spbasic)
% a) write to bib file (bmc-mathphys only)
% @settings{label, options="nameyear"}
% b) uncomment next line
%\nocite{label}

% or include bibliography directly:
% \begin{thebibliography}
% \bibitem{b1}
% \end{thebibliography}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Figures                       %%
%%                               %%
%% NB: this is for captions and  %%
%% Titles. All graphics must be  %%
%% submitted separately and NOT  %%
%% included in the Tex document  %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%
%% Do not use \listoffigures as most will included as separate files

\section*{Figures}
  \begin{figure}[h!]
  \caption{\csentence{Sample figure title.}
      A short description of the figure content
      should go here.}
      \end{figure}

\begin{figure}[h!]
  \caption{\csentence{Sample figure title.}
      Figure legend text.}
      \end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Tables                        %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Use of \listoftables is discouraged.
%%
\section*{Tables}
\begin{table}[h!]
\caption{Sample table title. This is where the description of the table should go.}
      \begin{tabular}{cccc}
        \hline
           & B1  &B2   & B3\\ \hline
        A1 & 0.1 & 0.2 & 0.3\\
        A2 & ... & ..  & .\\
        A3 & ..  & .   & .\\ \hline
      \end{tabular}
\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Additional Files              %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Additional Files}
  \subsection*{Additional file 1 --- Sample additional file title}
    Additional file descriptions text (including details of how to
    view the file, if it is in a non-standard format or the file extension).  This might
    refer to a multi-page table or a figure.

  \subsection*{Additional file 2 --- Sample additional file title}
    Additional file descriptions text.

\end{backmatter}
\end{document}

### 2020-07-06 paper/policar2020-opentsne.tex
Add initial draft

@@
 \startlocaldefs
+\newcommand{\opentsne}{\textsf{openTSNE}}
 \endlocaldefs
@@
 \begin{abstract}
-Point-based visualisations of large, multi-dimensional data from
-molecular biology can reveal meaningful clusters. One of the
-most popular techniques to construct such visualisations is t-distributed
-stochastic neighbor embedding (t-SNE), for which a number of extensions have
-recently been proposed to address issues of scalability and the quality of the resulting
-visualisations. We introduce openTSNE, a modular Python library that implements
-the core t-SNE algorithm and its extensions. The library is orders of magnitude
-faster than existing popular implementations, including those from
-scikit-learn. Unique to openTSNE is also the mapping of new data to existing embeddings, which
-can surprisingly assist in solving batch effects.
+Point-based visualisations of large, multi-dimensional data from molecular biology can reveal meaningful clusters.
+One of the most popular techniques to construct such visualisations is t-distributed stochastic neighbor embedding (t-SNE), for which a number of extensions have recently been proposed to address issues of scalability and the quality of the resulting visualisations.
+We introduce \opentsne, a modular Python library that implements the core t-SNE algorithm and its extensions.
+The library is orders of magnitude faster than existing popular implementations, including those from scikit-learn.
+Unique to \opentsne\ is also the mapping of new data to existing embeddings, which can surprisingly assist in solving batch effects.
 \end{abstract}
@@
 \begin{keyword}
-\kwd{sample}
-\kwd{article}
-\kwd{author}
+\kwd{t-SNE}
+\kwd{embedding}
+\kwd{visualization}
+\kwd{dimensionality reduction}
 \end{keyword}
@@
 help in the construction of data visualizations. Popular approaches for
-dimensionality reduction include principal component analysis, multidimensional
-scaling, and uniform manifold approximation and projections~\cite{2018arXivUMAP}. Among
-these, t-distributed stochastic neighbor embedding
-(t-SNE)~\cite{maaten2008visualizing} lately received much attention as it can
-address high volumes of data and reveal meaningful clustering structure. Most
-of the recent reports on single-cell gene expression data start with an
-overview of the cell landscape, where t-SNE embeds high-dimensional expression
-profiles into a two-dimensional space~\cite{macosko2015highly,cao2019single,tasic2018shared}.
-Fig.~\ref{fig:tsne}.a presents an example of one such embedding.
-
-Despite its utility, t-SNE has been criticized for poor scalability
-when addressing large data sets, lack of global organization \em t-SNE
-focuses on local clusters that are arbitrarily scattered in the
-low-dimensional space \em and absence of theoretically-founded implementations to
-map new data into existing embeddings~\cite{ding2018interpretable,becht2019dimensionality}. Most of these shortcomings, have
-recently been addressed.
-\cite{fi_tsne} sped-up the method through an interpolation-based
-approximation, reducing the time complexity to be merely linear dependent on the number of samples. \cite{art_of_using_tsne} proposed
-several techniques to improve global positioning, including estimating similarities with a mixture of
-Gaussian kernels. While no current popular
-software library supports mapping of new data into reference embedding,
-\cite{parametric_tsne} proposed a related approach
-using neural networks.
-
-%% TODO: END FROM FIRST PAPER
+dimensionality reduction include principal component analysis (PCA), multidimensional
+scaling, t-distributed stochastic
+neighbor embedding (t-SNE)~\cite{maaten2008visualizing}, and uniform manifold approximation and
+projections (UMAP)~\cite{2018arXivUMAP}. Among these, t-SNE lately received much
+attention as it can address high volumes of data and reveal meaningful
+clustering structure. Most of the recent reports on single-cell gene expression
+data start with an overview of the cell landscape, where t-SNE embeds
+high-dimensional expression profiles into a two-dimensional
+space~\cite{macosko2015highly,cao2019single,tasic2018shared}.
+Figs.~\ref{fig:macosko}.a and \ref{fig:macosko}.b are two examples of such
+embeddings.
 
@@
   \includegraphics[width=\textwidth]{macosko2015}
-  \caption{\label{fig:macosko}}
+  \caption{\label{fig:macosko}The figure shows the progression of t-SNE embeddings following theoretical advances, which are all included in \opentsne.
+The data in \textbf{(a)} and \textbf{(b)} represent 44,808 single-cell gene-expression profiles of mouse retinal cells from the study by Macosko \textit{et al.}~\cite{macosko2015highly}.
+Similarly, the data in \textbf{(c)} contains 27,499 expression profiles from mouse retinal cells from Shekhar \textit{et al.}~\cite{shekhar2016comprehensive}.
+\textbf{(a)} We construct a t-SNE embedding following the parameter choices from the original publication from Maaten \& Hinton~\cite{maaten2008visualizing}.
+The visualization shows poor preservation of the global organisation of clusters, resulting from random initialization and an affinity model focused on preserving local neighborhoods.
+\textbf{(b)} shows a modern t-SNE embedding, utilizing the latest theoretical advances and practical recommendations.
+The embedding is constructed using a multi-scale affinity model, preserving both short-range and long-range interactions between data points, and is initialized such that the global layout is meaningful.
+Unlike in \textbf{(a)}, the green and blue clusters representing different sub-types of amacrine and bipolar cells are now localized to the same regions of the space, indicating a higher level of similarity than to other cell types.
+\textbf{(c)} shows how existing t-SNE reference atlases can be used to embed new samples into existing embeddings.
+The positions of new data points correspond to cell types from the reference atlas.}
 \end{figure*}
+
+Despite its utility, t-SNE has often been criticized for poor scalability when
+addressing large data sets, lack of global organization -- t-SNE focuses on local
+clusters that are arbitrarily scattered in the low-dimensional space  -- and
+absence of theoretically-founded implementations to map new data into existing
+embeddings~\cite{ding2018interpretable,becht2019dimensionality}. Most of these
+shortcomings, have recently been addressed. Linderman \textit{et
+al.} sped-up the method through an interpolation-based
+approximation, reducing the time complexity to be merely linear dependent on
+the number of samples~\cite{linderman2019fast}. Kobak \textit{et al.} proposed
+several techniques to improve global positioning, including estimating
+similarities with a mixture of Gaussian kernels~\cite{kobak2019art}. In our previous work, we presented
+a principled approach for embedding new samples into existing visualizations~\cite{policar2019embedding}.
 
@@
 
-We introduce openTSNE, a comprehensive Python library that
-implements t-SNE and all its recently proposed extensions. The library is
-compatible with the Python data science ecosystem (e.g., {\tt numpy}, {\tt
-sklearn}, {\tt scanpy}). Its modular design encourages extendibility and
-experimentation with various settings and changes in the analysis pipeline.
+We introduce \opentsne, a comprehensive Python library that implements t-SNE and all its recently proposed extensions.
+Most notable among these are the implementation of efficient approximation schemes~\cite{van2014accelerating,linderman2019fast} allowing the embedding of millions of data points, the addition of new samples into existing t-SNE embeddings~\cite{policar2019embedding}, better initialization schemes~\cite{kobak2019umap} leading to more globally consistent layouts, variable degrees of freedom~\cite{kobak2019heavy} allowing the inspection of data at different levels of resolution, multiscale similarity kernels~\cite{kobak2019art} which emphasize the preservation of both short-range and long-range interactions, and improved default parameters~\cite{belkina2019automated} which produce better embeddings at lower computational cost.
+The library is compatible with the Python data science ecosystem (e.g., \textsf{numpy}, \textsf{scikit-learn}, \textsf{scanpy}) providing a familiar and intuitive user interface to new users.
+Its modular design encourages extensibility and experimentation with various settings and changes in the analysis pipeline.
+
+\opentsne\ is open-source and has become an integral part of the Python data-science ecosystem.
+The package has received over 600 Github stars, and has been downloaded hundreds of thousands of times, averaging in thousands of weekly downloads.
+The source code is publicly available on Github at \url{https://github.com/pavlin-policar/openTSNE}, and can be installed from \textsf{PyPI} and \textsf{conda-forge} -- the two most widely adopted Python package managers.
+
+Accessibility is one of the core design principles of \opentsne.
+Our aim is to make the latest theoretical advances  in t-SNE as accessible as possible to as many users as possible.
+As such, our API is based on the familiar scikit-learn interface~\cite{sklearn_api}, enabling a smooth transition from other Python data science libraries.
+For example, Fig.~\ref{fig:tasic} depicts four different t-SNE embeddings, each based on recent advances in t-SNE visualizations.
+The code used to generate these embeddings is listed below.
+
+\begin{minted}[
+fontsize=\footnotesize,
+frame=lines,
+framesep=2mm,
+baselinestretch=1.1
+]{python}
+import openTSNE
+from openTSNE.affinity import Multiscale
+# a - standard t-SNE
+tsne1 = openTSNE.TSNE(perplexity=30).fit(X)
+# b - high perplexity  for globa structure
+tsne2 = openTSNE.TSNE(perplexity=500).fit(X)
+# c - multiscale kernel for local/global structure
+aff = Multiscale(X, perplexities=[30, 500])
+tsne3 = openTSNE.TSNE(affinities=aff).fit(X)
+# d - decrease dof for higher resolution
+tsne4 = openTSNE.TSNE(dof=0.6).fit(X)
+\end{minted}
+
+\noindent The code snippet takes as input a \textsf{numpy} array or \textsf{scipy} sparse matrix \texttt{X} and produces four \texttt{TSNEEmbedding} objects.
+These may subsequently be subject to further optimization under different parameter settings, or may be used to embed new samples into the embedding landscape.
+Additional parameters are omitted for clarity, but can be found in accompanying notebooks publicly available at \url{https://github.com/pavlin-policar/opentsne-paper}.
+
+\section*{Discussion}
+
+\subsection*{Uncovering Structure in High-Dimensional Data Sets}
+
+Dimensionality reduction techniques implicitly assume that high-dimensional data lies on a lower-dimensional manifold, which can accurately be captured by a smaller number of dimensions.
+However, there is no evidence to suggest that any data set can accurately be described using only two dimensions, and that any projection onto a two-dimensional plane will inevitably lead to loss of information.
+Therefore, it is beneficial to examine multiple embeddings, each of which provide a different perspective into the topology and other characteristics of the data.
+For instance, in Fig.~\ref{fig:tasic}, we show four embeddings of the same gene-expression data from the mouse brain~\cite{tasic2018shared}.
 
@@
 
-The following code is used to generate Fig.~\ref{fig:tasic}
-\begin{minted}[fontsize=\footnotesize]{python}
-# Perplexity 30 (standard t-SNE)
-perp_30 = openTSNE.TSNE(
-    perplexity=30, metric="cosine"
-).fit(X)
+Fig.~\ref{fig:tasic}.a shows a t-SNE embedding using default parameters, which clearly uncovers numerous clusters.
+Warm colors correspond to excitatory neurons, while cool colors represent inhibitory neurons.
+While clusters of excitatory and inhibitory neurons appear close together, all clusters appear somewhat equidistant from their neighobrs, and the overall relations between clusters are hidden from the user.
+On the other hand, Fig.~\ref{fig:tasic}.b focuses on preserving larger neighborhoods of points, resulting in a much more globally consistent layouts where relations between clusters become more apparent.
+In this embedding, it is evident that there is one large group of excitatory neurons, and two related groups of inhibitory neurons.
+Unfortunately focusing on preserving large neighborhoods often results in smaller clusters being absorbed into larger ones.
+Alternatively, Fig.~\ref{fig:tasic}.c uses multi-scale similarity kernels which aim to preserve both the global organisation of clusters as well as prevent smaller clusters from being obscured by larger ones.
+Fig.~\ref{fig:tasic}.d shows another lever available to users used to uncover clustering structure at different levels of resolution.
+Kobak \textit{et al.}~\cite{kobak2019heavy} found that decreasing the degrees of freedom in the t-distribution an reveal smaller, more compact clusters, otherwise easily missed in other embeddings.
 
-# Perplexity 500 (emphasise global structure)
-perp_500 = openTSNE.TSNE(
-    perplexity=500, metric="cosine"
-).fit(X)
-
-# Multiscale kernel with perplexities 30,500
-affinities = openTSNE.affinity.Multiscale(
-    X, perplexities=[30, 500], metric="cosine"
-)
-init = openTSNE.initialization.pca(X)
-ms = openTSNE.TSNEEmbedding(init, affinities)
-ms = ms.optimize(n_iter=250, exaggeration=12)
-ms = ms.optimize(n_iter=250, exaggeration=1)
-
-# Expose finer structure using smaller dof
-perp_30 = openTSNE.TSNE(
-    perplexity=30, metric="cosine", dof=0.6
-).fit(X)
-\end{minted}
+When faced with a data set containing millions of data points, standard t-SNE embeddings often become unwieldy, with blurred cluster boundaries, large clusters absorbing smaller ones, and relationships between clusters becoming increasingly difficult to interpret.
+Fig.~\ref{fig:cao}.a contains expression profiles of 2,058,652 single cells captured at different time points in mouse development.
+The embedding clearly indicates numerous clusters with some transition between time points, but this is largely difficult to reason about.
+Kobak \& Berens observed that increasing attractive forces between similar data points controlled via the \textit{exaggeration} parameter leads to more compact clusters, which leads to more informative visualizations~\cite{kobak2019art}.
+For instance, \ref{fig:cao}.b increases the doubles the default exaggeration rate and begins to reveal the overall structure of the data.
+Doubling the exaggeration again in \ref{fig:cao}.c, we can now easily observe that the data is comprised of two large clusters, and eight smaller ones.
+Coincidentally, embeddings with this level of exaggeration roughly correspond to embeddings produced by UMAP~\cite{TODO}.
 
@@
 
-The following code is used to generate Fig.~\ref{fig:cao}
-\begin{minted}[fontsize=\footnotesize]{python}
-indices = np.random.permutation(list(range(x.shape[0])))
-reverse = np.argsort(indices)
-# Split the data into a sample/remainder sets
-x_sample, x_rest = x[indices[:25000]], x[indices[25000:]]
-y_sample, y_rest = y[indices[:25000]], y[indices[25000:]]
-# Create an initial embedding using a subset of the data
-sample_embedding = openTSNE.TSNE(
-    initialization="spectral",
-    perplexity=500,
-    metric="cosine",
-).fit(x_sample)
-# Position the remainder of the points to their nearest
-# neighbor in the sample embedding
-rest_init = sample_embedding.prepare_partial(x_rest, k=1)
-init_full = np.vstack(
-    [sample_embedding, rest_init]
-)[reverse]
-init_full = openTSNE.initialization.rescale(init_full)
-# Calculate the affinity model using all the data points
-affinities = openTSNE.affinity.PerplexityBasedNN(
-    x, perplexity=30, metric="cosine"
-)
+Exaggeration can be used to highlight transitions between cell-states in developmental studies.
+Standard t-SNE often produces embeddings with clearly-defined, discrete clusters.
+We have shown that users have access to several parameters to control the granularity of the clustering structure.
+However, such properties are undesired in developmental studies as we assume that the cells follow a continuous path between cell states.
+To this end, other embedding techniques such as UMAP and ForceAtlas2 are often used, as they tend to better capture the continuity between cell states.
+Recently, Kobak \textit{et al.}~\cite{TODO} have shown that embeddings produced by t-SNE with exaggeration values of 4 and $\sim30$ produce embeddings which are visually very similar to UMAP and ForceAtlas2.
+For example, in Fis.~\ref{fig:cao}.a the colors clearly indicate some transitional states, which are made much more apparent in Fig.~\ref{fig:cao}.c.
 
-embedding = openTSNE.TSNEEmbedding(
-    init_full, affinities
-%)
-exag12 = embedding.optimize(n_iter=500, exaggeration=12)
-exag4 = exag12.optimize(n_iter=500, exaggeration=4)
-exag2 = exag4.optimize(n_iter=500, exaggeration=2)
-exag1 = exag2.optimize(n_iter=500, exaggeration=1)
-\end{minted}
+\subsection*{Embedding New Samples}
+
+Unlike other popular dimensionality reduction methods such as principal component analysis or autoencoders, t-SNE is a non-parametric method and does not define an explicit mapping to the embedding space.
+Therefore embeddings of new data points need to be found through the use of optimization techniques~\cite{policar2019embedding}.
+\opentsne\ is currently the only publicly available library allowing users to add new samples to existing embeddings.
+
+Figs.~\ref{fig:macosko}.b and \ref{fig:macosko}.c demonstrate how we can use a previously annotated single-cell data set and embed data from a different single-cell experiment into the reference landscape.
+The reference data taken from Macosko \textit{et al.}~\cite{macosko2015highly} contains gene-expression profiles from mouse retinal cells.
+Taking data from and experiment focusing only on bipolar retinal cells by Shekhar \textit{et al.} and embedding these cells into the reference embedding, we can see that the majority of new cells correctly map to the bipolar cell clusters in the reference.
+
+Embedding single cells into existing reference atlases can also be useful for cell-type classification in cases where cell identities are not yet known.
+For instance, in Fig.~\ref{fig:transform}, we construct a reference embedding using data from Hochgerner \textit{et al.}~\cite{hochgerner2018conserved} containing  gene-expression profiles of cells from the mouse brain.
+The authors provide cell annotations for each cell, and we can verify their accuracy by visualizing well-established gene markers for the major cell types.
+We can then embed similar data from Harris \textit{et al.}, where annotations are provided only for neuronal cells, and quickly identify other non-neuronal cell types, including oligodendrocytes and astrocytes.
+We then use the marker genes to validate that new cells were correctly mapped into the reference landscape.
 
@@
 
-The following code is used to generate Fig.~\ref{fig:transform}
-\begin{minted}[fontsize=\footnotesize]{python}
-# Assuming we have prepared a reference embedding and
-# assuming we have aligned new_X to the reference X matrix
-new_embedding = reference_embedding.transform(new_X)
-\end{minted}
+These two examples demonstrate how \opentsne\ can be used to quickly gain insight into newly-sequenced, single-cell data sets by utilizing already existing, publicly-available, reference atlases.
+This approach is extremely general and is not limited to single-cell gene-expression data, but can be applied to any tabular data.
 
-We want to mention that this works for mitigating batch effects in our earlier work~\cite{policar2019embedding}.
+\subsection*{Versatility}
+
+Versatility is another of  \opentsne s core design principles.
+Kobak \& Berens recently introduced several recommendations and tricks to obtain better and more meaningful t-SNE embeddings~\cite{kobak2019art}.
+These include the use of multi-scale similarity kernels, slowly easing the perplexity parameter, and increasing exaggeration when working with larger data sets.
+\opentsne\ provides an extremely flexible API, where these improvements can be incorporated in a few lines of code.
+Furthermore, \opentsne\ additionally supports custom affinity models, e.g.
+multi-scale Gaussian mixtures or uniform kernels, allowing for quick experimentation.
+This also enables users to construct t-SNE embeddings on non-tabular data, e.g.
+relational data, as the only affinity-model requirement is some notion of similarity between data points.
+Finally, our comprehensive callback system can utilized to monitor and adapt different stages of the optimization phase and has been used to construct visually appealing animations of the t-SNE optimization process.
+
+\subsection*{Speed}
+
+One of t-SNE's common criticisms is poor scalability to large data sets containing up to millions of data points~\cite{becht2019dimensionality}.
+These criticisms are largely based on two factors.
+Until quite recently, most popular implementations of t-SNE were based on the Barnes-Hut approximation scheme developed by van der Maaten in 2014~\cite{van2014accelerating} with asymptotic complexity $\mathcal{O}(N \log N)$.
+More importantly, this misconception was exacerbated further by the fact that the most widely-used implementation of t-SNE came from \textsf{scikit-learn}, which exhibits very poor runtime performance compared to its C++ counterpart -- \textsf{MulticoreTSNE}.
+Even with the Barnes-Hut approximation, the multi-threaded \textsf{MulticoreTSNE} implementation~\cite{Ulyanov2016} is perfectly capable of creating t-SNE embeddings of millions of data points in a matter of hours on widely-accessible, consumer-grade CPUs (Fig.~\ref{fig:benchmarks}).
+Recently, Linderman \textit{et al.} developed a new approximation scheme -- FIt-SNE -- which further reduces the asymptotic time complexity to $\mathcal{O}(N)$, embedding large data sets in a matter of minutes.
+
+Fig.~\ref{fig:benchmarks} shows benchmarks of four Python t-SNE implementations, including \textsf{scikit-learn} v0.23.1, \textsf{MulticoreTSNE} v0.1, \textsf{FIt-SNE} v1.1.0, and \opentsne\ v0.4.3.
+We perform benchmarks on two processors, one representing usage on a personal computer and one on a high-performance computing platform.
+The Intel(R) Core i7 is commonly found in consumer-grade laptop computers, while Intel(R) Xeon(R) processors are often found on high-performance computing machines.
+Benchmarks were run with the original t-SNE parameters for 1000 iterations, as some implementations do not enable their modification.
 
@@
 
-\section*{Discussion}
-Text for this section \ldots
+Fig.~\ref{fig:benchmarks} demonstrates that both \textsf{FIt-SNE} and \opentsne\ exhibit much better runtime performance than their Barnes-Hut counterparts -- \textsf{scikit-learn} and \textsf{MulticoreTSNE}.
+Notice that even though \textsf{FIt-SNE} and \opentsne\ both implement the same approximation scheme, \opentsne\ tends to be slightly slower.
+\opentsne\ is written in Python, which inevitably introduces some runtime overhead compared to its pure C++ counterpart.
+However, Fig.~\ref{fig:benchmarks} demonstrates the runtime of \opentsne\ is comparable to that of \textsf{FIt-SNE} and much faster than its Barnes-Hut competitors.
+Additionally, modern computer processors generally contain more than a single core, allowing us to make use of multi-threading, which drastically reduces the runtime differences between \opentsne\ and \textsf{FIt-SNE}.
+On the server-grade processor, \opentsne\ is actually slightly faster than its pure C++ counterpart.
+We speculate that this is likely due to the higher level API \opentsne\ uses for linear algebra operations, which may make better use of the Intel Math Kernel Library (MKL), which are aggressively optimized on Xeon processors.
+
+Additionally, \opentsne\ provides a flexible API, which allows splitting up the embedding-construction process and caching slow operations, enabling the user to quickly experiment with different parameter settings.
+
+\subsection*{Ease of Use}
+
+Widespread adoption of novel techniques is almost universally accompanied by easy-to-use and easy-to-install software.
+While the t-SNE implementation in \textsf{scikit-learn} fits this requirement, it implements the outdated Barnes-Hut approximation scheme which is unsuitable for large data sets.
+Other C++ implementations such as \textsf{MulticoreTSNE} and \textsf{FIt-SNE} are more suitable to larger data sets, but do not provide precompiled binaries, requiring users to compile the software themselves.
+This is problematic from a usability standpoint, as many users often lack the knowledge to do so.
+This is especially problematic with Windows users, where the operating system does not include a C++ compiler by default and correctly configuring the command line can be an arduous process.
+
+\opentsne\ is a Python package which focuses on accessibility.
+We provide precompiled binaries for all major Python version on all major platforms, making the installation process as seamless as possible.
+\opentsne\ can be installed through the Python Package Index (PyPI) or through conda through the conda-forge channel.
+\opentsne\ is designed to be familiar to Python users and follows the scikit-learn API~\cite{sklearn_api}, which is well established in the Python data-science ecosystem.
+\opentsne\ implements multi-threaded versions of both the Barnes-Hut and FIt-SNE approximation schemes, allowing it to be used on data sets containing millions of data points.
+While the Python virtual machine introduces some overhead, the runtime is comparable to that of its pure C++ counterpart -- \textsf{FIt-SNE}.
+Finally, \opentsne\ is extensible.
+Its modular design enables researchers to quickly experiment with different settings and easily incorporate their own components into the software, and has already been used by various researchers in their own work.
+
+\subsection*{Comparison to UMAP}
+
+Uniform Manifold Approximation and Projection (UMAP)~\cite{2018arXivUMAP} has often been touted as a silver bullet to the problems listed above, exhibiting impressive runtime on large data sets and organizing resulting clusters in a coherent manner.
+However, most performance comparisons to t-SNE were performed against older t-SNE implementations.
+Furthermore, the optimization procedure of UMAP is inherently sequential and does not easily lend itself to parallelization.
+On the other hand, modern t-SNE approximations are highly parallelizable and easily lend themselves to utilizing multiple cores, present on almost every consumer-grade processor.
+The embeddings produced by UMAP, like other force-directed layout algorithms such as t-SNE and ForceAtlas2~\cite{jacomy2014forceatlas2}, are largely dependent on their initialization.
+Many implementations of t-SNE begin with a randomly initialized embedding, leading to a largely scattered and incoherent layout of resulting clusters.
+On the other hand, UMAP embeddings are initialized using Laplacian eigenmaps, a method for the preservation of global relationships between data points.
+Kobak \& Linderman showed in \cite{kobak2019umap} that the supposed superior preservation of global structure is largely the result of this initialization, and that t-SNE, if initialized in similar manner, performed just as well in preserving global coherence.
+Most modern implementation of t-SNE default to such initialization schemes, typically though principal component analysis or spectral embeddings.
+
+We argue here not that either method is superior to the other, but rather, that they are complementary and should be used in tandem. Each may reveal different aspects of the data missed by the other.
 
@@
 
-\subsection*{Ease of Use}
-
-Widespread adoption of novel techniques is almost universally accompanied by easy-to-use and easy-to-install software.
-While the t-SNE implementation in scikit-learn and Rtsne fits this requirement, it implements the outdated Barnes-Hut approximation scheme which is unsuitable for larger data sets containing up to millions of data points.
-Other C++ implementations such as MulticoreTSNE and FIt-SNE are more suitable for larger data sets, but do not provide precompiled binaries to their software, requiring users to compile the software themselves.
-This is problematic from a usability standpoint, as many users often either do not know how to compile C++ source code or use Windows, which does not include a C++ compiler by default.
-
-openTSNE is a Python package which focuses on ease-of-use and extensibility.
-We provide precompiled binaries for all major Python version on all major platforms, making the installation process as seamless as possible.
-openTSNE can be installed through the Python Package Index (PyPI) or through conda on the conda-forge channel.
-openTSNE is designed to be familliar to Python users and loosely follows the scikit-learn API, which is well established in the Python data-science ecosystem.
-openTSNE implements both the Barnes-Hut and FIt-SNE approximation schemes, allowing it to be used on data sets containing millions of data points.
-While the Python virtual machine introduces some overhead, the runtime is comparable to that of its pure C++ counterpart -- FIt-SNE.
-Finally, openTSNE is extensible.
-Its modular interface enables researchers to quickly experiment with different settings and easily incorporate their own components into the software.
-
-\begin{description}
-\item[Alternative initialization schemes] Spectral and PCA, cite dmitry and george paper
-
-\item[Incorporates latest developments] $lr=N/12, n_iter=750$
-
-\end{description}
-
 \begin{table}
@@
 \section*{Availability}
-Text for this section \ldots
+
+openTSNE is distributed under the BSD-3-Clause License and is publicly available as an open-source package at \url{https://github.com/pavlin-policar/openTSNE}.
+openTSNE is also available on PyPI and conda-forge.
+The data sets used and scripts used in this study are included in accompanying notebooks, publicly available at \url{https://github.com/pavlin-policar/opentsne-paper}.
 
@@
 \section*{Acknowledgements}
-  Text for this section \ldots
+We would like to thank Dmitry Kobak for various helpful discussions and best practices when using t-SNE, as well as his contributions to the source code.
+We would also like to thank George Linderman for his help with the FIt-SNE algorithm.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2020-07-06 paper/policar2020-opentsne.tex
Revise manuscript

--- 2020-07-05 paper/policar2020-opentsne.tex
@@
 
-%We implement t-SNE~\cite{maaten2008visualizing}.
-We include a set of recommendations and tricks from Kobak and Berens~\cite{kobak2019art} as well as incerasing the learning rate as recommended by Belkina \textit{et al.}~\cite{belkina2019automated}.
-Also support using variable degrees of freedom by Kobak \textit{et al.}~\cite{kobak2019heavy}.
-%
-%Two approximation schemes, one based on Barnes-Hut~\cite{van2014accelerating} and one on the interpolation-based approach~\cite{linderman2019fast}.
-%
-%UMAP~\cite{2018arXivUMAP}, whether or not it better preserves global structure or not is in question~\cite{kobak2019umap}.
-
-%% TODO: START FROM FIRST PAPER
-
 The abundance of high-dimensional data sets in molecular biology calls for
@@
 help in the construction of data visualizations. Popular approaches for
-dimensionality reduction include principal component analysis, multidimensional
-scaling, and uniform manifold approximation and projections~\cite{2018arXivUMAP}. Among
-these, t-distributed stochastic neighbor embedding
-(t-SNE)~\cite{maaten2008visualizing} lately received much attention as it can
-address high volumes of data and reveal meaningful clustering structure. Most
-of the recent reports on single-cell gene expression data start with an
-overview of the cell landscape, where t-SNE embeds high-dimensional expression
-profiles into a two-dimensional space~\cite{macosko2015highly,cao2019single,tasic2018shared}.
-Fig.~\ref{fig:tsne}.a presents an example of one such embedding.
-
-Despite its utility, t-SNE has been criticized for poor scalability
-when addressing large data sets, lack of global organization \em t-SNE
-focuses on local clusters that are arbitrarily scattered in the
-low-dimensional space \em and absence of theoretically-founded implementations to
-map new data into existing embeddings~\cite{ding2018interpretable,becht2019dimensionality}. Most of these shortcomings, have
-recently been addressed.
-\cite{fi_tsne} sped-up the method through an interpolation-based
-approximation, reducing the time complexity to be merely linear dependent on the number of samples. \cite{art_of_using_tsne} proposed
-several techniques to improve global positioning, including estimating similarities with a mixture of
-Gaussian kernels. While no current popular
-software library supports mapping of new data into reference embedding,
-\cite{parametric_tsne} proposed a related approach
-using neural networks.
-
-%% TODO: END FROM FIRST PAPER
+dimensionality reduction include principal component analysis (PCA), multidimensional
+scaling, t-distributed stochastic
+neighbor embedding (t-SNE)~\cite{maaten2008visualizing}, and uniform manifold approximation and
+projections (UMAP)~\cite{2018arXivUMAP}. Among these, t-SNE lately received much
+attention as it can address high volumes of data and reveal meaningful
+clustering structure. Most of the recent reports on single-cell gene expression
+data start with an overview of the cell landscape, where t-SNE embeds
+high-dimensional expression profiles into a two-dimensional
+space~\cite{macosko2015highly,cao2019single,tasic2018shared}.
+Figs.~\ref{fig:macosko}.a and \ref{fig:macosko}.b are examples of two such
+embeddings.
 
@@
   \includegraphics[width=\textwidth]{macosko2015}
-  \caption{\label{fig:macosko}}
+  \caption{\label{fig:macosko}The figure shows the progression of t-SNE embeddings following theoretical advances, which are all included in \opentsne.
+The data in \textbf{(a)} and \textbf{(b)} represent 44,808 single-cell gene-expression profiles of mouse retinal cells from the study by Macosko \textit{et al.}~\cite{macosko2015highly}.
+Similarly, the data in \textbf{(c)} contains 27,499 expression profiles from mouse retinal cells from Shekhar \textit{et al.}~\cite{shekhar2016comprehensive}.
+\textbf{(a)} We construct a t-SNE embedding following the parameter choices from the original publication from Maaten \& Hinton~\cite{maaten2008visualizing}.
+The visualization shows poor preservation of the global organization of clusters, resulting from random initialization and an affinity model focused on preserving local neighborhoods.
+\textbf{(b)} shows a modern t-SNE embedding, utilizing the latest theoretical advances and practical recommendations.
+The embedding is constructed using a multi-scale affinity model, preserving both short-range and long-range interactions between data points, and is initialized such that the global layout is meaningful.
+Unlike in \textbf{(a)}, the green and blue clusters representing different sub-types of amacrine and bipolar cells are now localized to the same regions of the space, indicating a higher level of similarity than to other cell types.
+\textbf{(c)} shows how existing t-SNE reference atlases can be used to embed new samples into existing embeddings.
+The positions of new data points correspond to cell types from the reference atlas.}
 \end{figure*}
+
+Despite its utility, t-SNE has often been criticized for poor scalability when
+addressing large data sets, lack of global organization -- t-SNE focuses on local
+clusters that are arbitrarily scattered in the low-dimensional space -- and the
+absence of theoretically-founded implementations to map new data into existing
+embeddings~\cite{ding2018interpretable,becht2019dimensionality}. Most of these
+shortcomings have recently been addressed. Linderman \textit{et
+al.} sped-up the method through an interpolation-based
+approximation, reducing the time complexity to be merely linear dependent on
+the number of samples~\cite{linderman2019fast}. Kobak \& Berens proposed
+several techniques to improve global positioning, including estimating
+similarities with a mixture of Gaussian kernels~\cite{kobak2019art}. In our previous work, we presented
+a principled approach for embedding new samples into existing visualizations~\cite{policar2019embedding}.
 
@@
 
-We introduce openTSNE, a comprehensive Python library that
-implements t-SNE and all its recently proposed extensions. The library is
-compatible with the Python data science ecosystem (e.g., {\tt numpy}, {\tt
-sklearn}, {\tt scanpy}). Its modular design encourages extendibility and
-experimentation with various settings and changes in the analysis pipeline.
+We introduce \opentsne, a comprehensive Python library that implements t-SNE and all its recently proposed extensions. Most notable among these are
+\begin{enumerate}
+\item the implementation of efficient approximation schemes~\cite{van2014accelerating,linderman2019fast} allowing the embedding of millions of data points,
+\item the addition of new samples into existing embeddings~\cite{policar2019embedding},
+\item better initialization schemes~\cite{kobak2019umap} leading to more globally consistent layouts,
+\item variable degrees of freedom~\cite{kobak2019heavy} allowing the inspection of data at different levels of resolution,
+\item multiscale similarity kernels~\cite{kobak2019art} which preserves small, well-defined clusters and uncovers global relationships between clusters, and
+\item improved defaults for learning rate and the number of iterations~\cite{belkina2019automated} which produce embeddings at a lower computational cost.
+\end{enumerate}
 
+The library is fully compatible with the Python data-science ecosystem (e.g., \textsf{numpy}, \textsf{scikit-learn}, \textsf{scanpy}) providing a familiar and intuitive user interface to new users.
+Its modular design encourages extensibility and experimentation with various settings and changes in the analysis pipeline.
+
+\opentsne\ is open-source and has become an integral part of the Python data science ecosystem.
+The package has received over 600 Github stars and has been downloaded hundreds of thousands of times, averaging in thousands of weekly downloads.
+The source code is publicly available on Github at \url{https://github.com/pavlin-policar/openTSNE}, and can be installed from \textsf{PyPI} and \textsf{conda-forge} -- the two most widely adopted Python package managers.
+
+Accessibility is one of the core design principles of \opentsne.
+Our aim is to make the latest theoretical advances in t-SNE as accessible as possible to as many users as possible.
+As such, our API is based on the familiar \textsf{scikit-learn} interface~\cite{sklearn_api}, which enables users a smooth transition from existing Python libraries.
+For example, Fig.~\ref{fig:tasic} depicts four different t-SNE visualizations generated by \opentsne, each based on recent advances in the t-SNE algorithm.
+The code used to generate these embeddings is shown below.
+
+\begin{minted}[
+fontsize=\footnotesize,
+frame=lines,
+framesep=2mm,
+baselinestretch=1.1
+]{python}
+import openTSNE
+from openTSNE.affinity import Multiscale
+# a - standard t-SNE
+tsne1 = openTSNE.TSNE(perplexity=30).fit(X)
+# b - high perplexity  for globa structure
+tsne2 = openTSNE.TSNE(perplexity=500).fit(X)
+# c - multiscale kernel for local/global structure
+aff = Multiscale(X, perplexities=[30, 500])
+tsne3 = openTSNE.TSNE(affinities=aff).fit(X)
+# d - decrease dof for higher resolution
+tsne4 = openTSNE.TSNE(dof=0.6).fit(X)
+\end{minted}
+
+\noindent The code snippet takes as input a \textsf{numpy} array or \textsf{scipy} sparse matrix \texttt{X} and produces four \texttt{TSNEEmbedding} objects.
+These may subsequently be subject to further optimization under different parameter settings or may be used to embed new samples into the embedding landscape.
+Additional parameters are omitted for clarity but can be found in accompanying notebooks publicly available at \url{https://github.com/pavlin-policar/opentsne-paper}.
+
+\section*{Discussion}
+
+\subsection*{Uncovering Structure in High-Dimensional Data Sets}
+
+Dimensionality reduction techniques implicitly assume that high-dimensional data lies on a lower-dimensional manifold, which can accurately be captured by a smaller number of dimensions.
+However, there is no evidence to suggest that any data set can accurately be described using only two dimensions and that any projection onto a two-dimensional plane will inevitably lead to loss of information.
+Therefore, it is beneficial to examine multiple embeddings, each of which provides a different perspective into the topology and other characteristics of the data.
+For instance, in Fig.~\ref{fig:tasic}, we show four embeddings of the same gene-expression data from the mouse brain~\cite{tasic2018shared}.
 \begin{figure*}[htbp]
@@
 Warm colors correspond to excitatory neurons, cool colors correspond to inhibitory neurons, and gray/brown colors correspond to non-neuronal cells.
-The standard t-SNE \textbf{(a)} emphasizes local structure, while increasing perplexity \textbf{(b)} results in a more meaningful layout of the clusters.
+The standard t-SNE \textbf{(a)} emphasizes local structure while increasing perplexity \textbf{(b)} results in a more meaningful layout of the clusters.
 We can also combine the two perplexities by using a multiscale kernel affinity model \textbf{(c)} and obtain a good trade-off between global and local structure.
@@
 \end{figure*}
+Fig.~\ref{fig:tasic}.a shows an embedding using default t-SNE parameters, which uncovers numerous clusters.
+Warm colors correspond to excitatory neurons, while cool colors represent inhibitory neurons.
+While different clusters of excitatory and inhibitory neurons appear close to one another, all clusters appear somewhat equidistant from their neighboring clusters, and the overall relations between them are not obvious.
+On the other hand, Fig.~\ref{fig:tasic}.b focuses on preserving larger neighborhoods of points, resulting in a more globally consistent layout where relations between clusters become more apparent.
+In this embedding, it is evident from the increased white-space between groups that there is one large class of excitatory neurons, and two related classes of inhibitory neurons.
+Unfortunately, focusing on preserving large neighborhoods often results in smaller clusters being absorbed into larger ones.
+Alternatively, Fig.~\ref{fig:tasic}.c uses multi-scale similarity kernels which aim to preserve both the global organization of clusters as well as prevent smaller clusters from being absorbed into larger ones.
+On the other hand, Fig.~\ref{fig:tasic}.d shows use the same parameter values as Fig.~\ref{fig:tasic}.a, but at a finer level of resolution.
+This reveals that many of the clusters are composed of other subgroups.
 
-The following code is used to generate Fig.~\ref{fig:tasic}
-\begin{minted}[fontsize=\footnotesize]{python}
-# Perplexity 30 (standard t-SNE)
-perp_30 = openTSNE.TSNE(
-    perplexity=30, metric="cosine"
-).fit(X)
-
-# Perplexity 500 (emphasise global structure)
-perp_500 = openTSNE.TSNE(
-    perplexity=500, metric="cosine"
-).fit(X)
-
-# Multiscale kernel with perplexities 30,500
-affinities = openTSNE.affinity.Multiscale(
-    X, perplexities=[30, 500], metric="cosine"
-)
-init = openTSNE.initialization.pca(X)
-ms = openTSNE.TSNEEmbedding(init, affinities)
-ms = ms.optimize(n_iter=250, exaggeration=12)
-ms = ms.optimize(n_iter=250, exaggeration=1)
-
-# Expose finer structure using smaller dof
-perp_30 = openTSNE.TSNE(
-    perplexity=30, metric="cosine", dof=0.6
-).fit(X)
-\end{minted}
+When faced with a data set containing millions of data points, standard t-SNE embeddings often become unwieldy, with blurred cluster boundaries, large clusters absorbing smaller ones, and relationships between clusters becoming increasingly difficult to interpret.
+Fig.~\ref{fig:cao}.a contains expression profiles of over two million single cells captured at different time points in mouse development.
+The embedding indicates numerous clusters with some transition between time points as indicated by the color-coding, but this is largely difficult to interpret.
+Kobak \& Berens observed that increasing attractive forces between similar data points controlled via the \textit{exaggeration} parameter leads to more compact clusters, and subsequently more informative visualizations~\cite{kobak2019art}.
+For instance, \ref{fig:cao}.b doubles the default exaggeration  which begins to reveal the overall structure of the data.
+Doubling the exaggeration again in \ref{fig:cao}.c, we can now easily observe that the data is comprised of two main groups of cells and eight somewhat smaller clusters.
+The visualization also reveals a small number of very small clusters, possibly corresponding to rare cell types.
 
@@
 
-The following code is used to generate Fig.~\ref{fig:cao}
-\begin{minted}[fontsize=\footnotesize]{python}
-indices = np.random.permutation(list(range(x.shape[0])))
-reverse = np.argsort(indices)
-# Split the data into a sample/remainder sets
-x_sample, x_rest = x[indices[:25000]], x[indices[25000:]]
-y_sample, y_rest = y[indices[:25000]], y[indices[25000:]]
-# Create an initial embedding using a subset of the data
-sample_embedding = openTSNE.TSNE(
-    initialization="spectral",
-    perplexity=500,
-    metric="cosine",
-).fit(x_sample)
-# Position the remainder of the points to their nearest
-# neighbor in the sample embedding
-rest_init = sample_embedding.prepare_partial(x_rest, k=1)
-init_full = np.vstack(
-    [sample_embedding, rest_init]
-)[reverse]
-init_full = openTSNE.initialization.rescale(init_full)
-# Calculate the affinity model using all the data points
-affinities = openTSNE.affinity.PerplexityBasedNN(
-    x, perplexity=30, metric="cosine"
-)
+Exaggeration can be used to highlight transitions between cell-states in developmental studies.
+Standard t-SNE often produces embeddings with clearly-defined, discrete clusters.
+We have shown that we can adjust the level of granularity and resolution of the clusters with several parameters.
+However, discrete clusters are often undesired in developmental studies where we assume cells follow a continuous path between cell states.
+To this end, other embedding techniques such as UMAP and ForceAtlas2~\cite{jacomy2014forceatlas2} are often used, as they tend to better capture the continuity between cell states.
+Recently, Kobak \textit{et al.}~\cite{TODO} have shown that embeddings produced by t-SNE with exaggeration values of 4 and $\sim30$ produce embeddings which are markedly similar to UMAP and ForceAtlas2, respectively.
+For example, in Fig.~\ref{fig:cao}.a, the developmental path between different time points is difficult to observe due to a large number of sprawled out clusters in the embedding.
+On the other hand, the developmental paths are easy to follow between cells in Fig.~\ref{fig:cao}.c, where we have increased the exaggeration factor to $4$.
 
-embedding = openTSNE.TSNEEmbedding(
-    init_full, affinities
-%)
-exag12 = embedding.optimize(n_iter=500, exaggeration=12)
-exag4 = exag12.optimize(n_iter=500, exaggeration=4)
-exag2 = exag4.optimize(n_iter=500, exaggeration=2)
-exag1 = exag2.optimize(n_iter=500, exaggeration=1)
-\end{minted}
+\subsection*{Embedding New Samples}
+
+Unlike other popular dimensionality reduction methods such as principal component analysis or autoencoders, t-SNE is a non-parametric method and does not define an explicit mapping to the embedding space.
+Therefore embeddings of new data points need to be found through the use of optimization techniques~\cite{policar2019embedding}.
+\opentsne\ is currently the only publicly available library allowing users to add new samples to existing embeddings.
+
+Figs.~\ref{fig:macosko}.b and \ref{fig:macosko}.c demonstrate how we can use a previously annotated single-cell data set and embed data from a separate experiment into the reference landscape.
+The reference data taken from Macosko \textit{et al.}~\cite{macosko2015highly} contains gene-expression profiles from mouse retinal cells.
+Embedding samples from a similar experiment focusing only on bipolar retinal cells by Shekhar \textit{et al.}~\cite{shekhar2016comprehensive}, we can observe that the majority of new cells correctly map to the bipolar cell clusters in the reference.
+
+Embedding single cells into existing reference atlases can also be useful for cell-type classification in cases where cell identities are not known in advance.
+For instance, in Fig.~\ref{fig:transform}, we construct a reference embedding using annotated data from Hochgerner \textit{et al.}~\cite{hochgerner2018conserved} containing gene-expression profiles of cells from the mouse brain.
+The authors provide annotations for each cell, and we can verify their accuracy by visualizing well-established gene markers' major cell types.
+We then embed cells from Harris \textit{et al.}~\cite{harris2018classes}, where annotations are provided only for neuronal cells.
+We quickly identify other non-neuronal cell types, including oligodendrocytes and astrocytes.
+We use marker genes to validate that new cells were correctly mapped into the reference landscape.
 
@@
 However, almost half of the collected cells are not neurons and were left uncharacterized.
-Inspecting where these cells are embedded in the reference embedding (bottom left) reveals that in addition to inhibitory neurons, the data contains a number of supporting glia as well as a small population of endothelial cells.
+Inspecting where these cells are embedded in the reference embedding (bottom left) reveals that in addition to inhibitory neurons, the data contains several supporting glial cells as well as a small population of endothelial cells.
 We can verify the accuracy of our approach by inspecting marker genes for the major cell types in the reference (top row) and embedded samples (bottom row).}
@@
 
-The following code is used to generate Fig.~\ref{fig:transform}
-\begin{minted}[fontsize=\footnotesize]{python}
-# Assuming we have prepared a reference embedding and
-# assuming we have aligned new_X to the reference X matrix
-new_embedding = reference_embedding.transform(new_X)
-\end{minted}
+These two examples demonstrate how \opentsne\ can be used to quickly gain insight into newly-sequenced, single-cell data sets by utilizing already existing, publicly-available, reference atlases.
+This approach is extremely general and is not limited to single-cell gene-expression data, but can be applied to any tabular data sets from any domain of science.
 
-We want to mention that this works for mitigating batch effects in our earlier work~\cite{policar2019embedding}.
+\subsection*{Versatility}
+
+Versatility is another of  \opentsne s core design principles.
+Kobak \& Berens recently introduced several recommendations and tricks to obtain better and more meaningful t-SNE visualizations~\cite{kobak2019art}.
+These include the use of multi-scale similarity kernels, perplexity annealing, and increasing exaggeration when working with larger data sets.
+\opentsne\ provides an extremely flexible API, where these improvements can be incorporated in just a few lines of code.
+Furthermore, \opentsne\ supports custom affinity models, e.g.
+multi-scale Gaussian mixtures or uniform kernels, allowing for quick experimentation.
+This also enables users to construct t-SNE embeddings on non-tabular data, e.g.
+relational data, as the only affinity-model requirement is some notion of similarity between data points.
+Finally, our comprehensive callback system can be utilized to monitor and adapt different stages of the optimization phase and has been used to construct visually appealing animations of the t-SNE optimization process.
+
+\subsection*{Speed}
+
+One of t-SNE's common criticisms is poor scalability to large data sets containing up to millions of data points~\cite{becht2019dimensionality}.
+These criticisms are largely based on two factors.
+Until quite recently, most popular implementations of t-SNE were based on the Barnes-Hut approximation scheme developed by van der Maaten in 2014~\cite{van2014accelerating} with asymptotic complexity $\mathcal{O}(N \log N)$.
+More importantly, this criticism was exacerbated further by the fact that the most widely-used implementation of t-SNE came from \textsf{scikit-learn}, which exhibits very poor runtime performance compared to its C++ counterpart -- \textsf{MulticoreTSNE}.
+The multi-threaded \textsf{MulticoreTSNE} implementation~\cite{Ulyanov2016} is perfectly capable of creating t-SNE embeddings of millions of data points in a matter of hours on widely-accessible, consumer-grade processors (Fig.~\ref{fig:benchmarks}).
+However, recently, Linderman \textit{et al.} developed a new approximation scheme -- FIt-SNE -- which further reduces the asymptotic time complexity to $\mathcal{O}(N)$, enabling the embedding of large data sets in a matter of minutes.
+
+Fig.~\ref{fig:benchmarks} shows benchmarks of four Python t-SNE implementations, including those from \textsf{scikit-learn} (v0.23.1), \textsf{MulticoreTSNE} (v0.1), \textsf{FIt-SNE} (v1.1.0), and \opentsne\ (v0.4.3).
+We perform benchmarks on two processors, one representing usage on a personal computer and one on a high-performance computing platform.
+The Intel(R) Core i7 is commonly found in consumer-grade laptop computers, while Intel(R) Xeon(R) processors are found in high-performance computing machines.
+Benchmarks were run with the original t-SNE parameters for 1000 iterations, as some implementations do not allow their modification.
 
@@
   \includegraphics[width=\textwidth]{benchmarks}
-  \caption{\label{fig:benchmarks}We benchmark openTSNE against three popular open-source implementations including scikit-learn~\cite{pedregosa2011scikit}, MulticoreTSNE~\cite{Ulyanov2016}, and FIt-SNE~\cite{linderman2019fast}.
+  \caption{\label{fig:benchmarks}We benchmark openTSNE (v0.4.3) against three popular open-source implementations including scikit-learn~\cite{pedregosa2011scikit} (v0.23.1), MulticoreTSNE~\cite{Ulyanov2016} (v0.1), and FIt-SNE~\cite{linderman2019fast} (v1.1.0).
 Benchmarks were run on a consumer-grade Intel Core i7-7700HQ processor found in laptop computers, and on a server-grade Intel Xeon E5-2650.
@@
 
-\section*{Discussion}
-Text for this section \ldots
+Fig.~\ref{fig:benchmarks} demonstrates that both \textsf{FIt-SNE} and \opentsne\ exhibit better runtime performance than their Barnes-Hut counterparts -- \textsf{scikit-learn} and \textsf{MulticoreTSNE}.
+Notice that even though \textsf{FIt-SNE} and \opentsne\ both implement the same approximation scheme, \opentsne\ tends to be slightly slower.
+\opentsne\ is written in Python, which inevitably incurs some runtime overhead compared to its pure C++ counterpart.
+However, Fig.~\ref{fig:benchmarks} demonstrates the runtime of \opentsne\ is comparable to that of \textsf{FIt-SNE} and much faster than its Barnes-Hut competitors.
+Additionally, modern computer processors generally contain multiple cores, allowing us to make use of multi-threading.
+This further reduces the gap between \opentsne\ and \textsf{FIt-SNE}.
+On the server-grade processor, \opentsne\ is slightly faster than its pure C++ counterpart.
+We speculate that this is could be because \opentsne\ uses \textsf{numpy} for most linear algebra operations, which may make better use of the Intel Math Kernel Library (MKL), which are aggressively optimized on Intel(R) Xeon(R) processors.
+
+Additionally, \opentsne\ provides a flexible API, which allows splitting up the embedding-construction process into several part, caching slow operations.
+This enables users to quickly experiment with different parameter settings and iterate on their final visualizations.
+
+\subsection*{Ease of Use}
+
+Widespread adoption of novel techniques is almost universally accompanied by easy-to-use and easy-to-install software.
+While the t-SNE implementation from \textsf{scikit-learn} fits this requirement, its implementation is unfortunately very slow, making it unsuitable for its application to large data sets.
+Other C++ implementations such as \textsf{MulticoreTSNE} and \textsf{FIt-SNE} exhibit better scaling in larger data sets, but do not provide precompiled binaries, requiring users to compile the software themselves.
+This is problematic from a usability standpoint, as many users often lack the knowledge to do so.
+This is especially problematic with Windows users, where the operating system does not include a C++ compiler by default, and correctly configuring the command line can be an arduous process.
+
+\opentsne\ is a Python package that focuses on accessibility.
+We provide precompiled binaries for all major Python versions on all major platforms, making the installation process as seamless as possible.
+\opentsne\ can be installed through the Python Package Index (PyPI) or through conda on the conda-forge channel.
+\opentsne\ is designed to be familiar to Python users and follows the \textsf{scikit-learn} API~\cite{sklearn_api}, which is well established in the Python data science ecosystem.
+\opentsne\ implements multi-threaded versions of both the Barnes-Hut and FIt-SNE approximation schemes, allowing it to be used on data sets containing millions of data points.
+While the Python virtual machine introduces some performance overhead, the runtime is comparable to that of its pure C++ counterpart -- \textsf{FIt-SNE}.
+Finally, \opentsne\ is extensible.
+Its modular design enables researchers to quickly experiment with different settings and easily incorporate their own components into the software.
 
@@
 
-% approach name in bold, reference, description of the approach and why is
-% it useful. focus on theory and math.
+This section is meant to familiarize the reader with the t-SNE algorithm and its recent extensions.
+This brief review is by no means comprehensive, but is self contained and introduces the mathematics upon which the \opentsne\ library is based.
 
-\subsection*{t-SNE}
-%% TODO: Reword this entirely
+\subsection*{Preliminaries}
+This section is meant to familiarize the reader with the t-SNE algorithm and its recent extensions.
+This brief review is by no means comprehensive, but is self-contained and introduces the mathematics upon which the \opentsne\ library is based.
 
-Local, non-linear dimensionality reduction by t-SNE is performed as follows.
-Given a multi-dimensional data set $\mathbf{X} = \left \{ \mathbf{x}_1,
+\subsection*{Preliminaries}
+
+t-distributed stochastic neighbor embedding (t-SNE) is a non-linear dimensionality reduction method, which aims to find a low-dimensional embedding, where local neighborhoods are preserved.
+More formally, given a multi-dimensional data set $\mathbf{X} = \left \{ \mathbf{x}_1,
 \mathbf{x}_2, \dots, \mathbf{x}_N \right \} \in \mathbb{R}^D$ where $N$ is the
-number of data points in the reference data set, t-SNE aims to find a low
+number of data points in the data set, t-SNE aims to find a low
 dimensional embedding $\mathbf{Y} = \left \{ \mathbf{y}_1, \mathbf{y}_2, \dots,
 \mathbf{y}_N \right \} \in \mathbb{R}^d$ where $d \ll D$, such that if points
-$\mathbf{x}_i$ and $\mathbf{x}_j$ are close in the multi-dimensional space,
+$\mathbf{x}_i$ and $\mathbf{x}_j$ are close in the high-dimensional space,
 their corresponding embeddings $\mathbf{y}_i$ and $\mathbf{y}_j$ are also
 close. Since t-SNE is primarily used as a visualization tool, $d$ is typically
-set to two. The similarity between two data points in t-SNE is defined as:
+set to two. The similarity between two data points in the high-dimensional space is defined as:
 
@@
 
-\noindent where $\mathcal{D}$ is a distance measure. This is then symmetrized to
+\noindent where $\mathcal{D}$ is some distance measure. This is then symmetrized to
 
@@
 \noindent Different bandwidths $\sigma_i$ enable t-SNE to adapt to the varying
-density of the data in the multi-dimensional space.
+density of the data in the multi-dimensional space. Perplexity is sometimes considered the continuous analog to the $k$ nearest neighbors to which distances will be preserved.
 
 The similarity between points $\mathbf{y}_i$ and $\mathbf{y}_j$ in the
-embedding space is defined using the $t$-distribution with one degree of
-freedom
+embedding space is defined using the $t$-distribution with a single degree of
+freedom. This is also known as the Cauchy kernel:
 
@@
 
-The t-SNE method finds an embedding $\mathbf{Y}$ that minimizes the
-Kullback-Leibler (KL) divergence between $\mathbf{P}$ and $\mathbf{Q}$,
+Finally, the Kullback-Leibler (KL) divergence is used as a measure of agreement between distributions $\mathbf{P}$ and $\mathbf{Q}$
 
@@
 
-It is then easy to show that the corresponding gradient takes on the form
+\noindent The t-SNE objective is to find embeddings $\mathbf{Y}$ such that the KL divergence is minimized.
+
+The corresponding gradient takes the form
 \begin{equation}
@@
 
-Optimization is performed using batch gradient descent with the delta-bar-delta update rule for increased convergence speed~\cite{jacobs1988increased}.
+Optimization is performed using batch gradient descent with the delta-bar-delta update rule~\cite{jacobs1988increased}.
 Originally, t-SNE was run for 1000 iterations consisting of two phases.
-In the first \textit{early exaggeration} phase, the attractive forces between data points is increased by a factor $\rho=12$ so that points can move throughout the embedding uninhibited by repulsive forces.
-Then the remaining 750 iterations are run with $\rho=1$, which reverts the attractive forces to their original values.
+In the first \textit{early exaggeration} phase, the attractive forces between data points is increased by a factor $\rho=12$ so that points can more easily move throughout the embedding to find their respective neighbors.
+The remaining 750 iterations are run with $\rho=1$, which reverts the attractive forces to their original values and produces the final embedding.
 
-Belkina \textit{et al.} later found that the number of iterations can saftely be decreased to 750 iterations by increasing the learning rate to $\eta=N/12$ instead of the default and somewhat arbitrarily set $\eta=200$~\cite{belkina2019automated}.
-A welcome side effect of this reduction is also faster runtime.
+Belkina \textit{et al.} later found that faster convergence can be achieved by increasing the learning rate to $\eta=N/12$~\cite{belkina2019automated}.
+The number of iterations can then safely be lowered to 750, leading to faster runtime.
+This convention has been adopted by most modern t-SNE implementations.
 
@@
 
-Van der Maaten observed that since $p_{ij}$ is calculated in the high-dimensional space using a Gaussian kernel, and that the bandwidth is selected such that only a fixed number of closest neighbors (determined by the perplexity parameter) will contribute to the attractive forces of a given point due to the exponential decay of the Gaussian kernel.
+Van der Maaten observed that since $p_{ij}$ is calculated in the high-dimensional space using a Gaussian kernel and that the bandwidth is selected such that only a fixed number of closest neighbors (determined by the perplexity parameter) will contribute to the attractive forces of a given point due to the exponential decay of the Gaussian kernel~\cite{van2014accelerating}.
 Therefore it is sufficient to calculate the attractive forces only over a relatively small number of nearest neighbors instead of all $N$ points.
 By utilizing tree-based nearest-neighbor search methods, the time complexity is thus reduced to $\mathcal{O}(N \log N)$.
-Linderman \textit{et al.} further realised that, qualitatively, embeddings are visually virtually indistinguishable when using only \textit{approximate} nearest neighbors, further reducing time complexity to $\mathcal{O}(N)$~\cite{linderman2019fast}.
+Linderman \textit{et al.} further realized that, qualitatively, embeddings are visually virtually indistinguishable when using only \textit{approximate} nearest neighbors, further reducing time complexity to $\mathcal{O}(N)$~\cite{linderman2019fast}.
 
@@
 Examining the second term of Eqn.~(\ref{eq:grad_attr_rep}) we notice that each point indiscriminately exerts a repulsive force on all other points.
-Van der Maaten proposed an approach based on N-body simulations and used a space-partitioning Barnes-Hut tree approach to approximate the interaction between data points.
+Van der Maaten proposed an approach based on N-body simulations and used a space-partitioning Barnes-Hut tree approach to approximate the interaction between data points~\cite{van2014accelerating}.
 Briefly, the approach splits the space into quadrants (in the 2D case) and if a given point is far enough away from a quadrant, the points in that quadrant are summarized by a single center of mass.
@@
 
-More recently, Linderman \textit{et al.} proposed an alternative approach, FIt-SNE, based on non-uniform convolutions to calculating all pairwise interactions between repelling data points.
+More recently, Linderman \textit{et al.} proposed an alternative approach, FIt-SNE, based on non-uniform convolutions to calculating all pairwise interactions between repelling data points~\cite{linderman2019fast}.
 Briefly, Linderman \textit{et al.} observed that the repulsive forces $\mathbf{R}$ may be rewritten as
@@
 
-openTSNE provides efficient implementations for all the aforementioned approximations but defaults to using approximate nearest neighbor search and the non-uniform convolutions approach as these enable the user to quickly create informative visualizations for data sets containing up to millions of samples.
-We would note here that while the default choices introduce some runtime overhead on smaller data sets, this effect is minimal and impacts runtime by only a few seconds compared to the exact nearest-neighbor search Barnes-Hut approximation.
+\opentsne\ provides efficient implementations for all the aforementioned improvements but defaults to using approximate nearest neighbor search and the non-uniform convolutions approach as these enable the user to quickly create informative visualizations for data sets containing up to millions of samples.
 
@@
 
-Unlike other popular dimensionality reduction methods such as principal component analysis or autoencoders, t-SNE is a non-parametric method and does not define an explicit mapping to the embedding space.
+t-SNE is non-parametric and does not define an explicit mapping from the high-dimensional space to the embedding space.
 Therefore embeddings of new data points need to be found through the use of optimization techniques~\cite{policar2019embedding}.
@@
 
-openTSNE is currently the only open-source t-SNE implementation allowing users to add new samples into existing embeddings.
-
 \subsection*{Alternative Perplexity Kernels}
@@
 In standard t-SNE, distances are converted to similarities through the use of Gaussian kernels of varying bandwidths.
-The bandwidths are indirectly determined by the user-provided perplexity parameter, so that a fixed number of nearest data points will have non-zero similarity.
-One common trick for uncovering the global relations between clusters is to increase perplexity such that more long-range interactions are preserved in the final embedding.
-However, one unfortunate side effect of increasing perplexity is that smaller clusters get absorbed into larger ones, and less overall clusters are revealed.
+The bandwidths are indirectly determined by the user-specified perplexity parameter so that a fixed number of nearest data points will have non-zero similarity.
+One common trick for uncovering the global relations between clusters is to increase perplexity so that more long-range interactions are preserved in the final embedding.
+However, one unfortunate side effect of increasing perplexity is that smaller clusters get absorbed into larger ones.
 
-Kobak \textit{et al.}~\cite{kobak2019art} suggest that replacing the Gaussian kernel with a mixture of Gaussians may provide better insight into both the local and global structure.
-Therefore, the affinities in the input space become
+Kobak \& Berens~\cite{kobak2019art} suggest that replacing the Gaussian kernel with a mixture of Gaussians may provide better insight into both the local and global structure.
+Therefore, the similarities between data points in the input space become
 
@@
 The bandwidths of each Gaussian $\sigma_{1,i}$ and $\sigma_{2,i}$ is determined by selecting different perplexity values.
-They show that using a kernel with perplexity 500 captures the long-range interactions which preserve global cluster organization in the final embedding, and combine this with a kernel with perplexity 50, which prevents small, well-defined clusters from being absorbed into larger ones.
+They show that using a kernel with perplexity 500 captures the long-range interactions which preserve global cluster organization in the final embedding and combine this with a kernel with perplexity 50, which prevents small, well-defined clusters from being absorbed into larger ones.
 
@@
 
-Standard t-SNE is often used to reveal clustering structure in data at one level of resolution.
+Standard t-SNE is often used to reveal the clustering structure in data at one level of resolution.
 Different perplexity parameter values can be used to identify global cluster relationships or small, well-isolated groups.
@@
 
-In standard t-SNE $\alpha=1$ so this simplifies to Eqn.~(\ref{eq:cauchy_kernel}).
+In standard t-SNE $\alpha=1$ so this simplifies to the Cauchy kernel from Eqn.~(\ref{eq:cauchy_kernel}).
 
@@
 In standard t-SNE, the remainder of the iterations are run with $\rho=1$.
-Kobak \textit{et al.} showed that using $\rho=4$ and $\rho=30$ in the second phase of the optimization result in embeddings visually similar to UMAP embeddings, and ForceAtlas2 embeddings, respectively.
+Kobak \textit{et al.} showed that using $\rho=4$ and $\rho=30$ in the second phase of the optimization results in embeddings visually similar to UMAP embeddings, and ForceAtlas2 embeddings, respectively.
 
@@
 
-\subsection*{Ease of Use}
+Similarly to UMAP and ForceAtlas2, t-SNE can also be viewed as a force-directed layout algorithm.
+The first phase of the t-SNE algorithm constructs a $k$-nearest neighbor graph, and the subsequent optimization balances the attractive and repulsive forces between data points.
 
-Widespread adoption of novel techniques is almost universally accompanied by easy-to-use and easy-to-install software.
-While the t-SNE implementation in scikit-learn and Rtsne fits this requirement, it implements the outdated Barnes-Hut approximation scheme which is unsuitable for larger data sets containing up to millions of data points.
-Other C++ implementations such as MulticoreTSNE and FIt-SNE are more suitable for larger data sets, but do not provide precompiled binaries to their software, requiring users to compile the software themselves.
-This is problematic from a usability standpoint, as many users often either do not know how to compile C++ source code or use Windows, which does not include a C++ compiler by default.
+The final positioning of data points using this class of algorithms is largely dependent on the initial positions in the embedding space.
+UMAP is often said to better preserve global structure than t-SNE, but Kobak \& Linderman have recently discovered that this supposed advantage is entirely due to the better default initialization in UMAP than in older implementations of t-SNE~\cite{kobak2019umap}.
+Standard implementations of t-SNE use random initializations, which often scatter resulting clusters throughout the embedding space.
+UMAP injects global organization into the embedding by initializing point positions using Laplacian eigenmaps.
+When initialized in the same manner, both t-SNE and UMAP perform equally well, or equally poorly, dispelling the myth that UMAP is superior in this respect.
 
-openTSNE is a Python package which focuses on ease-of-use and extensibility.
-We provide precompiled binaries for all major Python version on all major platforms, making the installation process as seamless as possible.
-openTSNE can be installed through the Python Package Index (PyPI) or through conda on the conda-forge channel.
-openTSNE is designed to be familliar to Python users and loosely follows the scikit-learn API, which is well established in the Python data-science ecosystem.
-openTSNE implements both the Barnes-Hut and FIt-SNE approximation schemes, allowing it to be used on data sets containing millions of data points.
-While the Python virtual machine introduces some overhead, the runtime is comparable to that of its pure C++ counterpart -- FIt-SNE.
-Finally, openTSNE is extensible.
-Its modular interface enables researchers to quickly experiment with different settings and easily incorporate their own components into the software.
-
-\begin{description}
-\item[Alternative initialization schemes] Spectral and PCA, cite dmitry and george paper
-
-\item[Incorporates latest developments] $lr=N/12, n_iter=750$
-
-\end{description}
+\opentsne\ defaults to using the two leading principal components as initialization, but also provides a spectral approach, similar to that of UMAP.
+The clusters in resulting embeddings are organized in a globally coherent manner, leading to more interpretable visualizations.
 

### 2020-07-06 paper/policar2020-opentsne.tex
Improve titles

@@
 
-\subsection*{Uncovering Structure in High-Dimensional Data Sets}
+\subsection*{Uncovering Structure in High-Dimensional Data}
 
@@
 Therefore, it is beneficial to examine multiple embeddings, each of which provides a different perspective into the topology and other characteristics of the data.
-For instance, in Fig.~\ref{fig:tasic}, we show four embeddings of the same gene-expression data from the mouse brain~\cite{tasic2018shared}.
 \begin{figure*}[htbp]
@@
 \end{figure*}
+For instance, in Fig.~\ref{fig:tasic}, we show four embeddings of the same gene-expression data from the mouse brain~\cite{tasic2018shared}.
 Fig.~\ref{fig:tasic}.a shows an embedding using default t-SNE parameters, which uncovers numerous clusters.
@@
 This section is meant to familiarize the reader with the t-SNE algorithm and its recent extensions.
-This brief review is by no means comprehensive, but is self contained and introduces the mathematics upon which the \opentsne\ library is based.
-
-\subsection*{Preliminaries}
-This section is meant to familiarize the reader with the t-SNE algorithm and its recent extensions.
 This brief review is by no means comprehensive, but is self-contained and introduces the mathematics upon which the \opentsne\ library is based.
@@
 
-\subsection*{Efficient Implementation}
+\subsection*{Efficient Approximation Schemes}
 A direct evaluation of t-SNE gradients requires $\mathcal{O}(N^2)$ operations, which makes its application impractical to any reasonably-sized data set and beckons for the development of efficient approximation schemes.
@@
 
-Standard t-SNE is often used to reveal the clustering structure in data at one level of resolution.
+Standard t-SNE reveals the clustering structure in data at one level of resolution.
 Different perplexity parameter values can be used to identify global cluster relationships or small, well-isolated groups.
@@
 
-\subsection*{Alternative Initialization Schemes}
+\subsection*{Globally Consistent Initialization Schemes}
 
@@
 \begin{table}
+\caption{\label{tab:features}Feature list...}
 
@@
 \begin{tabular}{r c c c c c c}
+\toprule
 \setlength\tabcolsep{6pt}
-& \rot{scikit-learn} & \rot{MulticoreTSNE} & \rot{FIt-SNE} & \rot{openTSNE} \\
+& \rot{\textsf{scikit-learn}} & \rot{\textsf{MulticoreTSNE}} & \rot{\textsf{FIt-SNE}} & \rot{\textsf{openTSNE}} \\
 \toprule
@@
 \hline
-$\mathcal{O}(N \log N)$ & \checkmark & \checkmark & & \checkmark \\
-$\mathcal{O}(N)$ & & & \checkmark & \checkmark \\
+Barnes-Hut ($\mathcal{O}(N \log N)$) & \checkmark & \checkmark & & \checkmark \\
+FIt-SNE ($\mathcal{O}(N)$) & & & \checkmark & \checkmark \\
 \hline
+Multiscale Gaussian kernels & & & \checkmark & \checkmark \\
+Fully-custom affinity kernels & & & & \checkmark \\
 Variable degrees of freedom & & & \checkmark & \checkmark \\
 Variable exaggeration & & & \checkmark & \checkmark \\
-Multiscale Gaussian kernels & & & \checkmark & \checkmark \\
-Custom affinity kernels & & & & \checkmark \\
-Adding new samples & & & & \checkmark \\
+Better initialization & & & \checkmark & \checkmark \\
+Automatic learning rate & & & \checkmark & \checkmark \\
+Embedding new samples & & & & \checkmark \\
 \bottomrule
@@
 
-% or include bibliography directly:
-% \begin{thebibliography}
-% \bibitem{b1}
-% \end{thebibliography}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                               %%
-%% Figures                       %%
-%%                               %%
-%% NB: this is for captions and  %%
-%% Titles. All graphics must be  %%
-%% submitted separately and NOT  %%
-%% included in the Tex document  %%
-%%                               %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%
-%% Do not use \listoffigures as most will included as separate files
-
-\section*{Figures}
-  \begin{figure}[h!]
-  \caption{\csentence{Sample figure title.}
-      A short description of the figure content
-      should go here.}
-      \end{figure}
-
-\begin{figure}[h!]
-  \caption{\csentence{Sample figure title.}
-      Figure legend text.}
-      \end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                               %%
-%% Tables                        %%
-%%                               %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%% Use of \listoftables is discouraged.
-%%
-\section*{Tables}
-\begin{table}[h!]
-\caption{Sample table title. This is where the description of the table should go.}
-      \begin{tabular}{cccc}
-        \hline
-           & B1  &B2   & B3\\ \hline
-        A1 & 0.1 & 0.2 & 0.3\\
-        A2 & ... & ..  & .\\
-        A3 & ..  & .   & .\\ \hline
-      \end{tabular}
-\end{table}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                               %%
-%% Additional Files              %%
-%%                               %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section*{Additional Files}
-  \subsection*{Additional file 1 --- Sample additional file title}
-    Additional file descriptions text (including details of how to
-    view the file, if it is in a non-standard format or the file extension).  This might
-    refer to a multi-page table or a figure.
-
-  \subsection*{Additional file 2 --- Sample additional file title}
-    Additional file descriptions text.
-
 \end{backmatter}

### 2020-07-06 paper/policar2020-opentsne.tex
Reformat tex file with line wrap

@@
 \begin{abstract}
-Point-based visualisations of large, multi-dimensional data from molecular biology can reveal meaningful clusters.
-One of the most popular techniques to construct such visualisations is t-distributed stochastic neighbor embedding (t-SNE), for which a number of extensions have recently been proposed to address issues of scalability and the quality of the resulting visualisations.
-We introduce \opentsne, a modular Python library that implements the core t-SNE algorithm and its extensions.
-The library is orders of magnitude faster than existing popular implementations, including those from scikit-learn.
-Unique to \opentsne\ is also the mapping of new data to existing embeddings, which can surprisingly assist in solving batch effects.
+Point-based visualisations of large, multi-dimensional data from molecular
+	biology can reveal meaningful clusters. One of the most popular
+	techniques to construct such visualisations is t-distributed stochastic
+	neighbor embedding (t-SNE), for which a number of extensions have
+	recently been proposed to address issues of scalability and the quality
+	of the resulting visualisations. We introduce \opentsne, a modular
+	Python library that implements the core t-SNE algorithm and its
+	extensions. The library is orders of magnitude faster than existing
+	popular implementations, including those from scikit-learn. Unique to
+	\opentsne\ is also the mapping of new data to existing embeddings,
+	which can surprisingly assist in solving batch effects.
 \end{abstract}
@@
 help in the construction of data visualizations. Popular approaches for
-dimensionality reduction include principal component analysis (PCA), multidimensional
-scaling, t-distributed stochastic
-neighbor embedding (t-SNE)~\cite{maaten2008visualizing}, and uniform manifold approximation and
-projections (UMAP)~\cite{2018arXivUMAP}. Among these, t-SNE lately received much
-attention as it can address high volumes of data and reveal meaningful
+dimensionality reduction include principal component analysis (PCA),
+multidimensional scaling, t-distributed stochastic neighbor embedding
+(t-SNE)~\cite{maaten2008visualizing}, and uniform manifold approximation and
+projections (UMAP)~\cite{2018arXivUMAP}. Among these, t-SNE lately received
+much attention as it can address high volumes of data and reveal meaningful
 clustering structure. Most of the recent reports on single-cell gene expression
@@
   \includegraphics[width=\textwidth]{macosko2015}
-  \caption{\label{fig:macosko}The figure shows the progression of t-SNE embeddings following theoretical advances, which are all included in \opentsne.
-The data in \textbf{(a)} and \textbf{(b)} represent 44,808 single-cell gene-expression profiles of mouse retinal cells from the study by Macosko \textit{et al.}~\cite{macosko2015highly}.
-Similarly, the data in \textbf{(c)} contains 27,499 expression profiles from mouse retinal cells from Shekhar \textit{et al.}~\cite{shekhar2016comprehensive}.
-\textbf{(a)} We construct a t-SNE embedding following the parameter choices from the original publication from Maaten \& Hinton~\cite{maaten2008visualizing}.
-The visualization shows poor preservation of the global organization of clusters, resulting from random initialization and an affinity model focused on preserving local neighborhoods.
-\textbf{(b)} shows a modern t-SNE embedding, utilizing the latest theoretical advances and practical recommendations.
-The embedding is constructed using a multi-scale affinity model, preserving both short-range and long-range interactions between data points, and is initialized such that the global layout is meaningful.
-Unlike in \textbf{(a)}, the green and blue clusters representing different sub-types of amacrine and bipolar cells are now localized to the same regions of the space, indicating a higher level of similarity than to other cell types.
-\textbf{(c)} shows how existing t-SNE reference atlases can be used to embed new samples into existing embeddings.
-The positions of new data points correspond to cell types from the reference atlas.}
+  \caption{\label{fig:macosko}The figure shows the progression of t-SNE
+	embeddings following theoretical advances, which are all included in
+	\opentsne. The data in \textbf{(a)} and \textbf{(b)} represent 44,808
+	single-cell gene-expression profiles of mouse retinal cells from the
+	study by Macosko \textit{et al.}~\cite{macosko2015highly}. Similarly,
+	the data in \textbf{(c)} contains 27,499 expression profiles from mouse
+	retinal cells from Shekhar \textit{et
+	al.}~\cite{shekhar2016comprehensive}. \textbf{(a)} We construct a t-SNE
+	embedding following the parameter choices from the original publication
+	from Maaten \& Hinton~\cite{maaten2008visualizing}. The visualization
+	shows poor preservation of the global organization of clusters,
+	resulting from random initialization and an affinity model focused on
+	preserving local neighborhoods. \textbf{(b)} shows a modern t-SNE
+	embedding, utilizing the latest theoretical advances and practical
+	recommendations. The embedding is constructed using a multi-scale
+	affinity model, preserving both short-range and long-range interactions
+	between data points, and is initialized such that the global layout is
+	meaningful. Unlike in \textbf{(a)}, the green and blue clusters
+	representing different sub-types of amacrine and bipolar cells are now
+	localized to the same regions of the space, indicating a higher level
+	of similarity than to other cell types. \textbf{(c)} shows how existing
+	t-SNE reference atlases can be used to embed new samples into existing
+	embeddings. The positions of new data points correspond to cell types
+	from the reference atlas.}
 \end{figure*}
@@
 Despite its utility, t-SNE has often been criticized for poor scalability when
-addressing large data sets, lack of global organization -- t-SNE focuses on local
-clusters that are arbitrarily scattered in the low-dimensional space -- and the
-absence of theoretically-founded implementations to map new data into existing
-embeddings~\cite{ding2018interpretable,becht2019dimensionality}. Most of these
-shortcomings have recently been addressed. Linderman \textit{et
-al.} sped-up the method through an interpolation-based
-approximation, reducing the time complexity to be merely linear dependent on
-the number of samples~\cite{linderman2019fast}. Kobak \& Berens proposed
-several techniques to improve global positioning, including estimating
-similarities with a mixture of Gaussian kernels~\cite{kobak2019art}. In our previous work, we presented
-a principled approach for embedding new samples into existing visualizations~\cite{policar2019embedding}.
+addressing large data sets, lack of global organization -- t-SNE focuses on
+local clusters that are arbitrarily scattered in the low-dimensional space --
+and the absence of theoretically-founded implementations to map new data into
+existing embeddings~\cite{ding2018interpretable,becht2019dimensionality}. Most
+of these shortcomings have recently been addressed. Linderman \textit{et al.}
+sped-up the method through an interpolation-based approximation, reducing the
+time complexity to be merely linear dependent on the number of
+samples~\cite{linderman2019fast}. Kobak \& Berens proposed several techniques
+to improve global positioning, including estimating similarities with a mixture
+of Gaussian kernels~\cite{kobak2019art}. In our previous work, we presented a
+principled approach for embedding new samples into existing
+visualizations~\cite{policar2019embedding}.
 
@@
 
-We introduce \opentsne, a comprehensive Python library that implements t-SNE and all its recently proposed extensions. Most notable among these are
-\begin{enumerate}
-\item the implementation of efficient approximation schemes~\cite{van2014accelerating,linderman2019fast} allowing the embedding of millions of data points,
-\item the addition of new samples into existing embeddings~\cite{policar2019embedding},
-\item better initialization schemes~\cite{kobak2019umap} leading to more globally consistent layouts,
-\item variable degrees of freedom~\cite{kobak2019heavy} allowing the inspection of data at different levels of resolution,
-\item multiscale similarity kernels~\cite{kobak2019art} which preserves small, well-defined clusters and uncovers global relationships between clusters, and
-\item improved defaults for learning rate and the number of iterations~\cite{belkina2019automated} which produce embeddings at a lower computational cost.
-\end{enumerate}
+We introduce \opentsne, a comprehensive Python library that implements t-SNE
+and all its recently proposed extensions. Most notable among these are
+\begin{enumerate} \item the implementation of efficient approximation
+			schemes~\cite{van2014accelerating,linderman2019fast}
+			allowing the embedding of millions of data points,
+		\item the addition of new samples into existing
+		embeddings~\cite{policar2019embedding}, \item better
+			initialization schemes~\cite{kobak2019umap} leading to
+			more globally consistent layouts, \item variable
+			degrees of freedom~\cite{kobak2019heavy} allowing the
+		inspection of data at different levels of resolution, \item
+			multiscale similarity kernels~\cite{kobak2019art} which
+			preserves small, well-defined clusters and uncovers
+			global relationships between clusters, and \item
+improved defaults for learning rate and the number of
+iterations~\cite{belkina2019automated} which produce embeddings at a lower
+computational cost.  \end{enumerate}
 
-The library is fully compatible with the Python data-science ecosystem (e.g., \textsf{numpy}, \textsf{scikit-learn}, \textsf{scanpy}) providing a familiar and intuitive user interface to new users.
-Its modular design encourages extensibility and experimentation with various settings and changes in the analysis pipeline.
+The library is fully compatible with the Python data-science ecosystem (e.g.,
+\textsf{numpy}, \textsf{scikit-learn}, \textsf{scanpy}) providing a familiar
+and intuitive user interface to new users. Its modular design encourages
+extensibility and experimentation with various settings and changes in the
+analysis pipeline.
 
-\opentsne\ is open-source and has become an integral part of the Python data science ecosystem.
-The package has received over 600 Github stars and has been downloaded hundreds of thousands of times, averaging in thousands of weekly downloads.
-The source code is publicly available on Github at \url{https://github.com/pavlin-policar/openTSNE}, and can be installed from \textsf{PyPI} and \textsf{conda-forge} -- the two most widely adopted Python package managers.
+\opentsne\ is open-source and has become an integral part of the Python data
+science ecosystem. The package has received over 600 Github stars and has been
+downloaded hundreds of thousands of times, averaging in thousands of weekly
+downloads. The source code is publicly available on Github at
+\url{https://github.com/pavlin-policar/openTSNE}, and can be installed from
+\textsf{PyPI} and \textsf{conda-forge} -- the two most widely adopted Python
+package managers.
 
-Accessibility is one of the core design principles of \opentsne.
-Our aim is to make the latest theoretical advances in t-SNE as accessible as possible to as many users as possible.
-As such, our API is based on the familiar \textsf{scikit-learn} interface~\cite{sklearn_api}, which enables users a smooth transition from existing Python libraries.
-For example, Fig.~\ref{fig:tasic} depicts four different t-SNE visualizations generated by \opentsne, each based on recent advances in the t-SNE algorithm.
-The code used to generate these embeddings is shown below.
+Accessibility is one of the core design principles of \opentsne. Our aim is to
+make the latest theoretical advances in t-SNE as accessible as possible to as
+many users as possible. As such, our API is based on the familiar
+\textsf{scikit-learn} interface~\cite{sklearn_api}, which enables users a
+smooth transition from existing Python libraries. For example,
+Fig.~\ref{fig:tasic} depicts four different t-SNE visualizations generated by
+\opentsne, each based on recent advances in the t-SNE algorithm. The code used
+to generate these embeddings is shown below.
 
@@
 
-\noindent The code snippet takes as input a \textsf{numpy} array or \textsf{scipy} sparse matrix \texttt{X} and produces four \texttt{TSNEEmbedding} objects.
-These may subsequently be subject to further optimization under different parameter settings or may be used to embed new samples into the embedding landscape.
-Additional parameters are omitted for clarity but can be found in accompanying notebooks publicly available at \url{https://github.com/pavlin-policar/opentsne-paper}.
+\noindent The code snippet takes as input a \textsf{numpy} array or
+\textsf{scipy} sparse matrix \texttt{X} and produces four
+\texttt{TSNEEmbedding} objects. These may subsequently be subject to further
+optimization under different parameter settings or may be used to embed new
+samples into the embedding landscape. Additional parameters are omitted for
+clarity but can be found in accompanying notebooks publicly available at
+\url{https://github.com/pavlin-policar/opentsne-paper}.
 
@@
 
-Dimensionality reduction techniques implicitly assume that high-dimensional data lies on a lower-dimensional manifold, which can accurately be captured by a smaller number of dimensions.
-However, there is no evidence to suggest that any data set can accurately be described using only two dimensions and that any projection onto a two-dimensional plane will inevitably lead to loss of information.
-Therefore, it is beneficial to examine multiple embeddings, each of which provides a different perspective into the topology and other characteristics of the data.
-\begin{figure*}[htbp]
+Dimensionality reduction techniques implicitly assume that high-dimensional
+data lies on a lower-dimensional manifold, which can accurately be captured by
+a smaller number of dimensions. However, there is no evidence to suggest that
+any data set can accurately be described using only two dimensions and that any
+projection onto a two-dimensional plane will inevitably lead to loss of
+information. Therefore, it is beneficial to examine multiple embeddings, each
+of which provides a different perspective into the topology and other
+characteristics of the data.  \begin{figure*}[htbp]
 	\includegraphics[width=\textwidth]{tasic2018}
-  \caption{\label{fig:tasic}We use openTSNE to create four different visualizations of the Tasic \textit{et al.}~\cite{tasic2018shared} data set containing 21,874 single-cells originating from the mouse neocortex.
-Cluster annotations and colors are taken from the original publication.
-Warm colors correspond to excitatory neurons, cool colors correspond to inhibitory neurons, and gray/brown colors correspond to non-neuronal cells.
-The standard t-SNE \textbf{(a)} emphasizes local structure while increasing perplexity \textbf{(b)} results in a more meaningful layout of the clusters.
-We can also combine the two perplexities by using a multiscale kernel affinity model \textbf{(c)} and obtain a good trade-off between global and local structure.
-Alternatively, we can inspect more fine-grained structure and reveal smaller clusters by using a more heavy-tailed kernel \textbf{(d)}.}
-\end{figure*}
-For instance, in Fig.~\ref{fig:tasic}, we show four embeddings of the same gene-expression data from the mouse brain~\cite{tasic2018shared}.
-Fig.~\ref{fig:tasic}.a shows an embedding using default t-SNE parameters, which uncovers numerous clusters.
-Warm colors correspond to excitatory neurons, while cool colors represent inhibitory neurons.
-While different clusters of excitatory and inhibitory neurons appear close to one another, all clusters appear somewhat equidistant from their neighboring clusters, and the overall relations between them are not obvious.
-On the other hand, Fig.~\ref{fig:tasic}.b focuses on preserving larger neighborhoods of points, resulting in a more globally consistent layout where relations between clusters become more apparent.
-In this embedding, it is evident from the increased white-space between groups that there is one large class of excitatory neurons, and two related classes of inhibitory neurons.
-Unfortunately, focusing on preserving large neighborhoods often results in smaller clusters being absorbed into larger ones.
-Alternatively, Fig.~\ref{fig:tasic}.c uses multi-scale similarity kernels which aim to preserve both the global organization of clusters as well as prevent smaller clusters from being absorbed into larger ones.
-On the other hand, Fig.~\ref{fig:tasic}.d shows use the same parameter values as Fig.~\ref{fig:tasic}.a, but at a finer level of resolution.
-This reveals that many of the clusters are composed of other subgroups.
+	\caption{\label{fig:tasic}We use openTSNE to create four different
+	visualizations of the Tasic \textit{et al.}~\cite{tasic2018shared} data
+	set containing 21,874 single-cells originating from the mouse
+	neocortex. Cluster annotations and colors are taken from the original
+	publication. Warm colors correspond to excitatory neurons, cool colors
+	correspond to inhibitory neurons, and gray/brown colors correspond to
+	non-neuronal cells. The standard t-SNE \textbf{(a)} emphasizes local
+	structure while increasing perplexity \textbf{(b)} results in a more
+	meaningful layout of the clusters. We can also combine the two
+	perplexities by using a multiscale kernel affinity model \textbf{(c)}
+	and obtain a good trade-off between global and local structure.
+	Alternatively, we can inspect more fine-grained structure and reveal
+	smaller clusters by using a more heavy-tailed kernel \textbf{(d)}.}
+\end{figure*} For instance, in Fig.~\ref{fig:tasic}, we show four embeddings of
+the same gene-expression data from the mouse brain~\cite{tasic2018shared}.
+Fig.~\ref{fig:tasic}.a shows an embedding using default t-SNE parameters, which
+uncovers numerous clusters. Warm colors correspond to excitatory neurons, while
+cool colors represent inhibitory neurons. While different clusters of
+excitatory and inhibitory neurons appear close to one another, all clusters
+appear somewhat equidistant from their neighboring clusters, and the overall
+relations between them are not obvious. On the other hand,
+Fig.~\ref{fig:tasic}.b focuses on preserving larger neighborhoods of points,
+resulting in a more globally consistent layout where relations between clusters
+become more apparent. In this embedding, it is evident from the increased
+white-space between groups that there is one large class of excitatory neurons,
+and two related classes of inhibitory neurons. Unfortunately, focusing on
+preserving large neighborhoods often results in smaller clusters being absorbed
+into larger ones. Alternatively, Fig.~\ref{fig:tasic}.c uses multi-scale
+similarity kernels which aim to preserve both the global organization of
+clusters as well as prevent smaller clusters from being absorbed into larger
+ones. On the other hand, Fig.~\ref{fig:tasic}.d shows use the same parameter
+values as Fig.~\ref{fig:tasic}.a, but at a finer level of resolution. This
+reveals that many of the clusters are composed of other subgroups.
 
-When faced with a data set containing millions of data points, standard t-SNE embeddings often become unwieldy, with blurred cluster boundaries, large clusters absorbing smaller ones, and relationships between clusters becoming increasingly difficult to interpret.
-Fig.~\ref{fig:cao}.a contains expression profiles of over two million single cells captured at different time points in mouse development.
-The embedding indicates numerous clusters with some transition between time points as indicated by the color-coding, but this is largely difficult to interpret.
-Kobak \& Berens observed that increasing attractive forces between similar data points controlled via the \textit{exaggeration} parameter leads to more compact clusters, and subsequently more informative visualizations~\cite{kobak2019art}.
-For instance, \ref{fig:cao}.b doubles the default exaggeration  which begins to reveal the overall structure of the data.
-Doubling the exaggeration again in \ref{fig:cao}.c, we can now easily observe that the data is comprised of two main groups of cells and eight somewhat smaller clusters.
-The visualization also reveals a small number of very small clusters, possibly corresponding to rare cell types.
+When faced with a data set containing millions of data points, standard t-SNE
+embeddings often become unwieldy, with blurred cluster boundaries, large
+clusters absorbing smaller ones, and relationships between clusters becoming
+increasingly difficult to interpret. Fig.~\ref{fig:cao}.a contains expression
+profiles of over two million single cells captured at different time points in
+mouse development. The embedding indicates numerous clusters with some
+transition between time points as indicated by the color-coding, but this is
+largely difficult to interpret. Kobak \& Berens observed that increasing
+attractive forces between similar data points controlled via the
+\textit{exaggeration} parameter leads to more compact clusters, and
+subsequently more informative visualizations~\cite{kobak2019art}. For instance,
+\ref{fig:cao}.b doubles the default exaggeration  which begins to reveal the
+overall structure of the data. Doubling the exaggeration again in
+\ref{fig:cao}.c, we can now easily observe that the data is comprised of two
+main groups of cells and eight somewhat smaller clusters. The visualization
+also reveals a small number of very small clusters, possibly corresponding to
+rare cell types.
 
@@
   \includegraphics[width=\textwidth]{cao2019}
-  \caption{\label{fig:cao}The data set from Cao \textit{et al.}~\cite{cao2019single} contains expression profiles from 2,058,652 single cells.
-The data were collected from mice embryos at different developmental time-points at daily intervals after 9.5 to 13.5 days.
-The exaggeration parameter can be used to highlight the global organization of the clusters and enforce cluster compactness.
-\textbf{(a)} shows the data is comprised of two main components -- the neural tube and mesenchymal cells -- as well as several other smaller clusters.
-Additionally, the colors indicate developmental progression with red indicating least-developed cells and blue indicating most developed cells.
-The overall developmental trajectory is most apparent with higher levels of exaggeration, showing red cells slowly transitioning into blue cells.
-Progressively easing the exaggeration factor uncovers finer clusters within the larger groups as shown in \textbf{(b)} with exaggeration 2 and subsequently in \textbf{(c)}, where we show the standard t-SNE with no exaggeration.
-32,011 putative doublets are excluded from the visualizations.}
+  \caption{\label{fig:cao}The data set from Cao \textit{et
+	al.}~\cite{cao2019single} contains expression profiles from 2,058,652
+	single cells. The data were collected from mice embryos at different
+	developmental time-points at daily intervals after 9.5 to 13.5 days.
+	The exaggeration parameter can be used to highlight the global
+	organization of the clusters and enforce cluster compactness.
+	\textbf{(a)} shows the data is comprised of two main components -- the
+	neural tube and mesenchymal cells -- as well as several other smaller
+	clusters. Additionally, the colors indicate developmental progression
+	with red indicating least-developed cells and blue indicating most
+	developed cells. The overall developmental trajectory is most apparent
+	with higher levels of exaggeration, showing red cells slowly
+	transitioning into blue cells. Progressively easing the exaggeration
+	factor uncovers finer clusters within the larger groups as shown in
+	\textbf{(b)} with exaggeration 2 and subsequently in \textbf{(c)},
+	where we show the standard t-SNE with no exaggeration. 32,011 putative
+	doublets are excluded from the visualizations.}
 \end{figure*}
 
-Exaggeration can be used to highlight transitions between cell-states in developmental studies.
-Standard t-SNE often produces embeddings with clearly-defined, discrete clusters.
-We have shown that we can adjust the level of granularity and resolution of the clusters with several parameters.
-However, discrete clusters are often undesired in developmental studies where we assume cells follow a continuous path between cell states.
-To this end, other embedding techniques such as UMAP and ForceAtlas2~\cite{jacomy2014forceatlas2} are often used, as they tend to better capture the continuity between cell states.
-Recently, Kobak \textit{et al.}~\cite{TODO} have shown that embeddings produced by t-SNE with exaggeration values of 4 and $\sim30$ produce embeddings which are markedly similar to UMAP and ForceAtlas2, respectively.
-For example, in Fig.~\ref{fig:cao}.a, the developmental path between different time points is difficult to observe due to a large number of sprawled out clusters in the embedding.
-On the other hand, the developmental paths are easy to follow between cells in Fig.~\ref{fig:cao}.c, where we have increased the exaggeration factor to $4$.
+Exaggeration can be used to highlight transitions between cell-states in
+developmental studies. Standard t-SNE often produces embeddings with
+clearly-defined, discrete clusters. We have shown that we can adjust the level
+of granularity and resolution of the clusters with several parameters. However,
+discrete clusters are often undesired in developmental studies where we assume
+cells follow a continuous path between cell states. To this end, other
+embedding techniques such as UMAP and ForceAtlas2~\cite{jacomy2014forceatlas2}
+are often used, as they tend to better capture the continuity between cell
+states. Recently, Kobak \textit{et al.}~\cite{TODO} have shown that embeddings
+produced by t-SNE with exaggeration values of 4 and $\sim30$ produce embeddings
+which are markedly similar to UMAP and ForceAtlas2, respectively. For example,
+in Fig.~\ref{fig:cao}.a, the developmental path between different time points
+is difficult to observe due to a large number of sprawled out clusters in the
+embedding. On the other hand, the developmental paths are easy to follow
+between cells in Fig.~\ref{fig:cao}.c, where we have increased the exaggeration
+factor to $4$.
 
@@
 
-Unlike other popular dimensionality reduction methods such as principal component analysis or autoencoders, t-SNE is a non-parametric method and does not define an explicit mapping to the embedding space.
-Therefore embeddings of new data points need to be found through the use of optimization techniques~\cite{policar2019embedding}.
-\opentsne\ is currently the only publicly available library allowing users to add new samples to existing embeddings.
+Unlike other popular dimensionality reduction methods such as principal
+component analysis or autoencoders, t-SNE is a non-parametric method and does
+not define an explicit mapping to the embedding space. Therefore embeddings of
+new data points need to be found through the use of optimization
+techniques~\cite{policar2019embedding}. \opentsne\ is currently the only
+publicly available library allowing users to add new samples to existing
+embeddings.
 
-Figs.~\ref{fig:macosko}.b and \ref{fig:macosko}.c demonstrate how we can use a previously annotated single-cell data set and embed data from a separate experiment into the reference landscape.
-The reference data taken from Macosko \textit{et al.}~\cite{macosko2015highly} contains gene-expression profiles from mouse retinal cells.
-Embedding samples from a similar experiment focusing only on bipolar retinal cells by Shekhar \textit{et al.}~\cite{shekhar2016comprehensive}, we can observe that the majority of new cells correctly map to the bipolar cell clusters in the reference.
+Figs.~\ref{fig:macosko}.b and \ref{fig:macosko}.c demonstrate how we can use a
+previously annotated single-cell data set and embed data from a separate
+experiment into the reference landscape. The reference data taken from Macosko
+\textit{et al.}~\cite{macosko2015highly} contains gene-expression profiles from
+mouse retinal cells. Embedding samples from a similar experiment focusing only
+on bipolar retinal cells by Shekhar \textit{et
+al.}~\cite{shekhar2016comprehensive}, we can observe that the majority of new
+cells correctly map to the bipolar cell clusters in the reference.
 
-Embedding single cells into existing reference atlases can also be useful for cell-type classification in cases where cell identities are not known in advance.
-For instance, in Fig.~\ref{fig:transform}, we construct a reference embedding using annotated data from Hochgerner \textit{et al.}~\cite{hochgerner2018conserved} containing gene-expression profiles of cells from the mouse brain.
-The authors provide annotations for each cell, and we can verify their accuracy by visualizing well-established gene markers' major cell types.
-We then embed cells from Harris \textit{et al.}~\cite{harris2018classes}, where annotations are provided only for neuronal cells.
-We quickly identify other non-neuronal cell types, including oligodendrocytes and astrocytes.
-We use marker genes to validate that new cells were correctly mapped into the reference landscape.
+Embedding single cells into existing reference atlases can also be useful for
+cell-type classification in cases where cell identities are not known in
+advance. For instance, in Fig.~\ref{fig:transform}, we construct a reference
+embedding using annotated data from Hochgerner \textit{et
+al.}~\cite{hochgerner2018conserved} containing gene-expression profiles of
+cells from the mouse brain. The authors provide annotations for each cell, and
+we can verify their accuracy by visualizing well-established gene markers'
+major cell types. We then embed cells from Harris \textit{et
+al.}~\cite{harris2018classes}, where annotations are provided only for neuronal
+cells. We quickly identify other non-neuronal cell types, including
+oligodendrocytes and astrocytes. We use marker genes to validate that new cells
+were correctly mapped into the reference landscape.
 
@@
   \includegraphics[width=\textwidth]{transform_hochgerner}
-  \caption{\label{fig:transform}openTSNE enables the user to embed new samples into an existing, reference embedding.
-We construct a t-SNE embedding for the data from Hochgerner \textit{et al.}~\cite{hochgerner2018conserved} containing 24,185 developing, single cells from the mouse hippocampus.
-The data contains different types of neurons, supporting glia, and other vascular cells (upper left).
-Cell annotations are taken from the original publications.
-The colors correspond to Fig.~\ref{fig:tasic}.
-We then embed new, hippocampal cells collected by Harris \textit{et al.}~\cite{harris2018classes} using the previous embedding as a reference.
-In their study, Harris \textit{et al.} collected 6,971 single-cells and focused on identifying different types of inhibitory neurons.
-However, almost half of the collected cells are not neurons and were left uncharacterized.
-Inspecting where these cells are embedded in the reference embedding (bottom left) reveals that in addition to inhibitory neurons, the data contains several supporting glial cells as well as a small population of endothelial cells.
-We can verify the accuracy of our approach by inspecting marker genes for the major cell types in the reference (top row) and embedded samples (bottom row).}
+  \caption{\label{fig:transform}openTSNE enables the user to embed new samples
+	into an existing, reference embedding. We construct a t-SNE embedding
+	for the data from Hochgerner \textit{et
+	al.}~\cite{hochgerner2018conserved} containing 24,185 developing,
+	single cells from the mouse hippocampus. The data contains different
+	types of neurons, supporting glia, and other vascular cells (upper
+	left). Cell annotations are taken from the original publications. The
+	colors correspond to Fig.~\ref{fig:tasic}. We then embed new,
+	hippocampal cells collected by Harris \textit{et
+	al.}~\cite{harris2018classes} using the previous embedding as a
+	reference. In their study, Harris \textit{et al.} collected 6,971
+	single-cells and focused on identifying different types of inhibitory
+	neurons. However, almost half of the collected cells are not neurons
+	and were left uncharacterized. Inspecting where these cells are
+	embedded in the reference embedding (bottom left) reveals that in
+	addition to inhibitory neurons, the data contains several supporting
+	glial cells as well as a small population of endothelial cells. We can
+	verify the accuracy of our approach by inspecting marker genes for the
+	major cell types in the reference (top row) and embedded samples
+	(bottom row).}
 \end{figure*}
 
-These two examples demonstrate how \opentsne\ can be used to quickly gain insight into newly-sequenced, single-cell data sets by utilizing already existing, publicly-available, reference atlases.
-This approach is extremely general and is not limited to single-cell gene-expression data, but can be applied to any tabular data sets from any domain of science.
+These two examples demonstrate how \opentsne\ can be used to quickly gain
+insight into newly-sequenced, single-cell data sets by utilizing already
+existing, publicly-available, reference atlases. This approach is extremely
+general and is not limited to single-cell gene-expression data, but can be
+applied to any tabular data sets from any domain of science.
 
@@
 
-Versatility is another of  \opentsne s core design principles.
-Kobak \& Berens recently introduced several recommendations and tricks to obtain better and more meaningful t-SNE visualizations~\cite{kobak2019art}.
-These include the use of multi-scale similarity kernels, perplexity annealing, and increasing exaggeration when working with larger data sets.
-\opentsne\ provides an extremely flexible API, where these improvements can be incorporated in just a few lines of code.
-Furthermore, \opentsne\ supports custom affinity models, e.g.
-multi-scale Gaussian mixtures or uniform kernels, allowing for quick experimentation.
-This also enables users to construct t-SNE embeddings on non-tabular data, e.g.
-relational data, as the only affinity-model requirement is some notion of similarity between data points.
-Finally, our comprehensive callback system can be utilized to monitor and adapt different stages of the optimization phase and has been used to construct visually appealing animations of the t-SNE optimization process.
+Versatility is another of  \opentsne s core design principles. Kobak \& Berens
+recently introduced several recommendations and tricks to obtain better and
+more meaningful t-SNE visualizations~\cite{kobak2019art}. These include the use
+of multi-scale similarity kernels, perplexity annealing, and increasing
+exaggeration when working with larger data sets. \opentsne\ provides an
+extremely flexible API, where these improvements can be incorporated in just a
+few lines of code. Furthermore, \opentsne\ supports custom affinity models,
+e.g. multi-scale Gaussian mixtures or uniform kernels, allowing for quick
+experimentation. This also enables users to construct t-SNE embeddings on
+non-tabular data, e.g. relational data, as the only affinity-model requirement
+is some notion of similarity between data points. Finally, our comprehensive
+callback system can be utilized to monitor and adapt different stages of the
+optimization phase and has been used to construct visually appealing animations
+of the t-SNE optimization process.
 
@@
 
-One of t-SNE's common criticisms is poor scalability to large data sets containing up to millions of data points~\cite{becht2019dimensionality}.
-These criticisms are largely based on two factors.
-Until quite recently, most popular implementations of t-SNE were based on the Barnes-Hut approximation scheme developed by van der Maaten in 2014~\cite{van2014accelerating} with asymptotic complexity $\mathcal{O}(N \log N)$.
-More importantly, this criticism was exacerbated further by the fact that the most widely-used implementation of t-SNE came from \textsf{scikit-learn}, which exhibits very poor runtime performance compared to its C++ counterpart -- \textsf{MulticoreTSNE}.
-The multi-threaded \textsf{MulticoreTSNE} implementation~\cite{Ulyanov2016} is perfectly capable of creating t-SNE embeddings of millions of data points in a matter of hours on widely-accessible, consumer-grade processors (Fig.~\ref{fig:benchmarks}).
-However, recently, Linderman \textit{et al.} developed a new approximation scheme -- FIt-SNE -- which further reduces the asymptotic time complexity to $\mathcal{O}(N)$, enabling the embedding of large data sets in a matter of minutes.
+One of t-SNE's common criticisms is poor scalability to large data sets
+containing up to millions of data points~\cite{becht2019dimensionality}. These
+criticisms are largely based on two factors. Until quite recently, most popular
+implementations of t-SNE were based on the Barnes-Hut approximation scheme
+developed by van der Maaten in 2014~\cite{van2014accelerating} with asymptotic
+complexity $\mathcal{O}(N \log N)$. More importantly, this criticism was
+exacerbated further by the fact that the most widely-used implementation of
+t-SNE came from \textsf{scikit-learn}, which exhibits very poor runtime
+performance compared to its C++ counterpart -- \textsf{MulticoreTSNE}. The
+multi-threaded \textsf{MulticoreTSNE} implementation~\cite{Ulyanov2016} is
+perfectly capable of creating t-SNE embeddings of millions of data points in a
+matter of hours on widely-accessible, consumer-grade processors
+(Fig.~\ref{fig:benchmarks}). However, recently, Linderman \textit{et al.}
+developed a new approximation scheme -- FIt-SNE -- which further reduces the
+asymptotic time complexity to $\mathcal{O}(N)$, enabling the embedding of large
+data sets in a matter of minutes.
 
@@
   \includegraphics[width=\textwidth]{benchmarks}
-  \caption{\label{fig:benchmarks}We benchmark openTSNE (v0.4.3) against three popular open-source implementations including scikit-learn~\cite{pedregosa2011scikit} (v0.23.1), MulticoreTSNE~\cite{Ulyanov2016} (v0.1), and FIt-SNE~\cite{linderman2019fast} (v1.1.0).
-Benchmarks were run on a consumer-grade Intel Core i7-7700HQ processor found in laptop computers, and on a server-grade Intel Xeon E5-2650.
-To generate benchmark data sets of different sizes, we subsampled data from the 10X Genomics 1.3 million mouse brain data set five times, resulting in five different data sets for each size.
-In total, each implementation was run on 30 different data sets.
-Notice that openTSNE scales similarly to FIt-SNE, as they both use the same interpolation-based approximation scheme, while scikit-learn and MulticoreTSNE utilize the Barnes-Hut approximation.}
+  \caption{\label{fig:benchmarks}We benchmark openTSNE (v0.4.3) against three
+	popular open-source implementations including
+	scikit-learn~\cite{pedregosa2011scikit} (v0.23.1),
+	MulticoreTSNE~\cite{Ulyanov2016} (v0.1), and
+	FIt-SNE~\cite{linderman2019fast} (v1.1.0). Benchmarks were run on a
+	consumer-grade Intel Core i7-7700HQ processor found in laptop
+	computers, and on a server-grade Intel Xeon E5-2650. To generate
+	benchmark data sets of different sizes, we subsampled data from the 10X
+	Genomics 1.3 million mouse brain data set five times, resulting in five
+	different data sets for each size. In total, each implementation was
+	run on 30 different data sets. Notice that openTSNE scales similarly to
+	FIt-SNE, as they both use the same interpolation-based approximation
+	scheme, while scikit-learn and MulticoreTSNE utilize the Barnes-Hut
+	approximation.}
 \end{figure*}
 
-Fig.~\ref{fig:benchmarks} demonstrates that both \textsf{FIt-SNE} and \opentsne\ exhibit better runtime performance than their Barnes-Hut counterparts -- \textsf{scikit-learn} and \textsf{MulticoreTSNE}.
-Notice that even though \textsf{FIt-SNE} and \opentsne\ both implement the same approximation scheme, \opentsne\ tends to be slightly slower.
-\opentsne\ is written in Python, which inevitably incurs some runtime overhead compared to its pure C++ counterpart.
-However, Fig.~\ref{fig:benchmarks} demonstrates the runtime of \opentsne\ is comparable to that of \textsf{FIt-SNE} and much faster than its Barnes-Hut competitors.
-Additionally, modern computer processors generally contain multiple cores, allowing us to make use of multi-threading.
-This further reduces the gap between \opentsne\ and \textsf{FIt-SNE}.
-On the server-grade processor, \opentsne\ is slightly faster than its pure C++ counterpart.
-We speculate that this is could be because \opentsne\ uses \textsf{numpy} for most linear algebra operations, which may make better use of the Intel Math Kernel Library (MKL), which are aggressively optimized on Intel(R) Xeon(R) processors.
+Fig.~\ref{fig:benchmarks} demonstrates that both \textsf{FIt-SNE} and
+\opentsne\ exhibit better runtime performance than their Barnes-Hut
+counterparts -- \textsf{scikit-learn} and \textsf{MulticoreTSNE}. Notice that
+even though \textsf{FIt-SNE} and \opentsne\ both implement the same
+approximation scheme, \opentsne\ tends to be slightly slower. \opentsne\ is
+written in Python, which inevitably incurs some runtime overhead compared to
+its pure C++ counterpart. However, Fig.~\ref{fig:benchmarks} demonstrates the
+runtime of \opentsne\ is comparable to that of \textsf{FIt-SNE} and much faster
+than its Barnes-Hut competitors. Additionally, modern computer processors
+generally contain multiple cores, allowing us to make use of multi-threading.
+This further reduces the gap between \opentsne\ and \textsf{FIt-SNE}. On the
+server-grade processor, \opentsne\ is slightly faster than its pure C++
+counterpart. We speculate that this is could be because \opentsne\ uses
+\textsf{numpy} for most linear algebra operations, which may make better use of
+the Intel Math Kernel Library (MKL), which are aggressively optimized on
+Intel(R) Xeon(R) processors.
 
-Additionally, \opentsne\ provides a flexible API, which allows splitting up the embedding-construction process into several part, caching slow operations.
-This enables users to quickly experiment with different parameter settings and iterate on their final visualizations.
+Additionally, \opentsne\ provides a flexible API, which allows splitting up the
+embedding-construction process into several part, caching slow operations. This
+enables users to quickly experiment with different parameter settings and
+iterate on their final visualizations.
 
@@
 
-Widespread adoption of novel techniques is almost universally accompanied by easy-to-use and easy-to-install software.
-While the t-SNE implementation from \textsf{scikit-learn} fits this requirement, its implementation is unfortunately very slow, making it unsuitable for its application to large data sets.
-Other C++ implementations such as \textsf{MulticoreTSNE} and \textsf{FIt-SNE} exhibit better scaling in larger data sets, but do not provide precompiled binaries, requiring users to compile the software themselves.
-This is problematic from a usability standpoint, as many users often lack the knowledge to do so.
-This is especially problematic with Windows users, where the operating system does not include a C++ compiler by default, and correctly configuring the command line can be an arduous process.
+Widespread adoption of novel techniques is almost universally accompanied by
+easy-to-use and easy-to-install software. While the t-SNE implementation from
+\textsf{scikit-learn} fits this requirement, its implementation is
+unfortunately very slow, making it unsuitable for its application to large data
+sets. Other C++ implementations such as \textsf{MulticoreTSNE} and
+\textsf{FIt-SNE} exhibit better scaling in larger data sets, but do not provide
+precompiled binaries, requiring users to compile the software themselves. This
+is problematic from a usability standpoint, as many users often lack the
+knowledge to do so. This is especially problematic with Windows users, where
+the operating system does not include a C++ compiler by default, and correctly
+configuring the command line can be an arduous process.
 
-\opentsne\ is a Python package that focuses on accessibility.
-We provide precompiled binaries for all major Python versions on all major platforms, making the installation process as seamless as possible.
-\opentsne\ can be installed through the Python Package Index (PyPI) or through conda on the conda-forge channel.
-\opentsne\ is designed to be familiar to Python users and follows the \textsf{scikit-learn} API~\cite{sklearn_api}, which is well established in the Python data science ecosystem.
-\opentsne\ implements multi-threaded versions of both the Barnes-Hut and FIt-SNE approximation schemes, allowing it to be used on data sets containing millions of data points.
-While the Python virtual machine introduces some performance overhead, the runtime is comparable to that of its pure C++ counterpart -- \textsf{FIt-SNE}.
-Finally, \opentsne\ is extensible.
-Its modular design enables researchers to quickly experiment with different settings and easily incorporate their own components into the software.
+\opentsne\ is a Python package that focuses on accessibility. We provide
+precompiled binaries for all major Python versions on all major platforms,
+making the installation process as seamless as possible. \opentsne\ can be
+installed through the Python Package Index (PyPI) or through conda on the
+conda-forge channel. \opentsne\ is designed to be familiar to Python users and
+follows the \textsf{scikit-learn} API~\cite{sklearn_api}, which is well
+established in the Python data science ecosystem. \opentsne\ implements
+multi-threaded versions of both the Barnes-Hut and FIt-SNE approximation
+schemes, allowing it to be used on data sets containing millions of data
+points. While the Python virtual machine introduces some performance overhead,
+the runtime is comparable to that of its pure C++ counterpart --
+\textsf{FIt-SNE}. Finally, \opentsne\ is extensible. Its modular design enables
+researchers to quickly experiment with different settings and easily
+incorporate their own components into the software.
 
@@
 
-This section is meant to familiarize the reader with the t-SNE algorithm and its recent extensions.
-This brief review is by no means comprehensive, but is self-contained and introduces the mathematics upon which the \opentsne\ library is based.
+This section is meant to familiarize the reader with the t-SNE algorithm and
+its recent extensions. This brief review is by no means comprehensive, but is
+self-contained and introduces the mathematics upon which the \opentsne\ library
+is based.
 
@@
 
-t-distributed stochastic neighbor embedding (t-SNE) is a non-linear dimensionality reduction method, which aims to find a low-dimensional embedding, where local neighborhoods are preserved.
-More formally, given a multi-dimensional data set $\mathbf{X} = \left \{ \mathbf{x}_1,
-\mathbf{x}_2, \dots, \mathbf{x}_N \right \} \in \mathbb{R}^D$ where $N$ is the
-number of data points in the data set, t-SNE aims to find a low
-dimensional embedding $\mathbf{Y} = \left \{ \mathbf{y}_1, \mathbf{y}_2, \dots,
-\mathbf{y}_N \right \} \in \mathbb{R}^d$ where $d \ll D$, such that if points
-$\mathbf{x}_i$ and $\mathbf{x}_j$ are close in the high-dimensional space,
-their corresponding embeddings $\mathbf{y}_i$ and $\mathbf{y}_j$ are also
-close. Since t-SNE is primarily used as a visualization tool, $d$ is typically
-set to two. The similarity between two data points in the high-dimensional space is defined as:
+t-distributed stochastic neighbor embedding (t-SNE) is a non-linear
+dimensionality reduction method, which aims to find a low-dimensional
+embedding, where local neighborhoods are preserved. More formally, given a
+multi-dimensional data set $\mathbf{X} = \left \{ \mathbf{x}_1, \mathbf{x}_2,
+\dots, \mathbf{x}_N \right \} \in \mathbb{R}^D$ where $N$ is the number of data
+points in the data set, t-SNE aims to find a low dimensional embedding
+$\mathbf{Y} = \left \{ \mathbf{y}_1, \mathbf{y}_2, \dots, \mathbf{y}_N \right
+\} \in \mathbb{R}^d$ where $d \ll D$, such that if points $\mathbf{x}_i$ and
+$\mathbf{x}_j$ are close in the high-dimensional space, their corresponding
+embeddings $\mathbf{y}_i$ and $\mathbf{y}_j$ are also close. Since t-SNE is
+primarily used as a visualization tool, $d$ is typically set to two. The
+similarity between two data points in the high-dimensional space is defined as:
 
@@
 
-\noindent where $\mathcal{D}$ is some distance measure. This is then symmetrized to
+\noindent where $\mathcal{D}$ is some distance measure. This is then
+symmetrized to
 
@@
 \noindent Different bandwidths $\sigma_i$ enable t-SNE to adapt to the varying
-density of the data in the multi-dimensional space. Perplexity is sometimes considered the continuous analog to the $k$ nearest neighbors to which distances will be preserved.
+density of the data in the multi-dimensional space. Perplexity is sometimes
+considered the continuous analog to the $k$ nearest neighbors to which
+distances will be preserved.
 
@@
 
-Finally, the Kullback-Leibler (KL) divergence is used as a measure of agreement between distributions $\mathbf{P}$ and $\mathbf{Q}$
+Finally, the Kullback-Leibler (KL) divergence is used as a measure of agreement
+between distributions $\mathbf{P}$ and $\mathbf{Q}$
 
@@
 
-\noindent The t-SNE objective is to find embeddings $\mathbf{Y}$ such that the KL divergence is minimized.
+\noindent The t-SNE objective is to find embeddings $\mathbf{Y}$ such that the
+KL divergence is minimized.
 
@@
 
-Optimization is performed using batch gradient descent with the delta-bar-delta update rule~\cite{jacobs1988increased}.
-Originally, t-SNE was run for 1000 iterations consisting of two phases.
-In the first \textit{early exaggeration} phase, the attractive forces between data points is increased by a factor $\rho=12$ so that points can more easily move throughout the embedding to find their respective neighbors.
-The remaining 750 iterations are run with $\rho=1$, which reverts the attractive forces to their original values and produces the final embedding.
+Optimization is performed using batch gradient descent with the delta-bar-delta
+update rule~\cite{jacobs1988increased}. Originally, t-SNE was run for 1000
+iterations consisting of two phases. In the first \textit{early exaggeration}
+phase, the attractive forces between data points is increased by a factor
+$\rho=12$ so that points can more easily move throughout the embedding to find
+their respective neighbors. The remaining 750 iterations are run with $\rho=1$,
+which reverts the attractive forces to their original values and produces the
+final embedding.
 
-Belkina \textit{et al.} later found that faster convergence can be achieved by increasing the learning rate to $\eta=N/12$~\cite{belkina2019automated}.
-The number of iterations can then safely be lowered to 750, leading to faster runtime.
-This convention has been adopted by most modern t-SNE implementations.
+Belkina \textit{et al.} later found that faster convergence can be achieved by
+increasing the learning rate to $\eta=N/12$~\cite{belkina2019automated}. The
+number of iterations can then safely be lowered to 750, leading to faster
+runtime. This convention has been adopted by most modern t-SNE implementations.
 
-\subsection*{Efficient Approximation Schemes}
-A direct evaluation of t-SNE gradients requires $\mathcal{O}(N^2)$ operations, which makes its application impractical to any reasonably-sized data set and beckons for the development of efficient approximation schemes.
-Van der Maaten observed that the t-SNE gradient may be reformulated in terms of a N-body problem where data points may be interpreted as particles which attract and repel each other~\cite{van2014accelerating}.
-The gradient may be rewritten as
+\subsection*{Efficient Approximation Schemes} A direct evaluation of t-SNE
+gradients requires $\mathcal{O}(N^2)$ operations, which makes its application
+impractical to any reasonably-sized data set and beckons for the development of
+efficient approximation schemes. Van der Maaten observed that the t-SNE
+gradient may be reformulated in terms of a N-body problem where data points may
+be interpreted as particles which attract and repel each
+other~\cite{van2014accelerating}. The gradient may be rewritten as
 \begin{equation}
@@
 \end{equation}
-where $Z = \sum_{k \neq l}\left ( 1 + || \mathbf{y}_k - \mathbf{y}_l ||^2 \right )^{-1}$.
-The two terms may be viewed as the attractive and repulsive forces between particles, respectively.
-Each term lends itself to efficient approximations, which enable us to greatly reduce the time complexity of t-SNE.
+where $Z = \sum_{k \neq l}\left ( 1 + || \mathbf{y}_k - \mathbf{y}_l ||^2
+\right )^{-1}$. The two terms may be viewed as the attractive and repulsive
+forces between particles, respectively. Each term lends itself to efficient
+approximations, which enable us to greatly reduce the time complexity of t-SNE.
 
@@
 
-Van der Maaten observed that since $p_{ij}$ is calculated in the high-dimensional space using a Gaussian kernel and that the bandwidth is selected such that only a fixed number of closest neighbors (determined by the perplexity parameter) will contribute to the attractive forces of a given point due to the exponential decay of the Gaussian kernel~\cite{van2014accelerating}.
-Therefore it is sufficient to calculate the attractive forces only over a relatively small number of nearest neighbors instead of all $N$ points.
-By utilizing tree-based nearest-neighbor search methods, the time complexity is thus reduced to $\mathcal{O}(N \log N)$.
-Linderman \textit{et al.} further realized that, qualitatively, embeddings are visually virtually indistinguishable when using only \textit{approximate} nearest neighbors, further reducing time complexity to $\mathcal{O}(N)$~\cite{linderman2019fast}.
+Van der Maaten observed that since $p_{ij}$ is calculated in the
+high-dimensional space using a Gaussian kernel and that the bandwidth is
+selected such that only a fixed number of closest neighbors (determined by the
+perplexity parameter) will contribute to the attractive forces of a given point
+due to the exponential decay of the Gaussian kernel~\cite{van2014accelerating}.
+Therefore it is sufficient to calculate the attractive forces only over a
+relatively small number of nearest neighbors instead of all $N$ points. By
+utilizing tree-based nearest-neighbor search methods, the time complexity is
+thus reduced to $\mathcal{O}(N \log N)$. Linderman \textit{et al.} further
+realized that, qualitatively, embeddings are visually virtually
+indistinguishable when using only \textit{approximate} nearest neighbors,
+further reducing time complexity to $\mathcal{O}(N)$~\cite{linderman2019fast}.
 
@@
 
-Examining the second term of Eqn.~(\ref{eq:grad_attr_rep}) we notice that each point indiscriminately exerts a repulsive force on all other points.
-Van der Maaten proposed an approach based on N-body simulations and used a space-partitioning Barnes-Hut tree approach to approximate the interaction between data points~\cite{van2014accelerating}.
-Briefly, the approach splits the space into quadrants (in the 2D case) and if a given point is far enough away from a quadrant, the points in that quadrant are summarized by a single center of mass.
-This reduces the time complexity from $\mathcal{O}(N^2)$ to $\mathcal{O}(N \log N)$.
+Examining the second term of Eqn.~(\ref{eq:grad_attr_rep}) we notice that each
+point indiscriminately exerts a repulsive force on all other points. Van der
+Maaten proposed an approach based on N-body simulations and used a
+space-partitioning Barnes-Hut tree approach to approximate the interaction
+between data points~\cite{van2014accelerating}. Briefly, the approach splits
+the space into quadrants (in the 2D case) and if a given point is far enough
+away from a quadrant, the points in that quadrant are summarized by a single
+center of mass. This reduces the time complexity from $\mathcal{O}(N^2)$ to
+$\mathcal{O}(N \log N)$.
 
-More recently, Linderman \textit{et al.} proposed an alternative approach, FIt-SNE, based on non-uniform convolutions to calculating all pairwise interactions between repelling data points~\cite{linderman2019fast}.
-Briefly, Linderman \textit{et al.} observed that the repulsive forces $\mathbf{R}$ may be rewritten as
-\begin{align}
+More recently, Linderman \textit{et al.} proposed an alternative approach,
+FIt-SNE, based on non-uniform convolutions to calculating all pairwise
+interactions between repelling data points~\cite{linderman2019fast}. Briefly,
+Linderman \textit{et al.} observed that the repulsive forces $\mathbf{R}$ may
+be rewritten as \begin{align}
 \mathbf{R}_i &= \sum_{j \neq i} q_{ij}^2 Z \left ( \mathbf{y}_i - \mathbf{y}_j \right ) \notag \\
@@
 \end{align}
-Then the numerator and denominator of Eqn.~(\ref{eq:fitsne_rep}) can be computed as $\mathbf{y}_i \phi_{1,j} - \phi_{2,j}$ and $Z = \sum_j \phi_{3,j}$, respectively.
-These interactions are accelerated by interpolating these terms through a grid of equispaced interpolation points.
-This shifts the computational burden onto the interpolation points and reduces the time complexity to $\mathcal{O}(N)$.
+Then the numerator and denominator of Eqn.~(\ref{eq:fitsne_rep}) can be
+computed as $\mathbf{y}_i \phi_{1,j} - \phi_{2,j}$ and $Z = \sum_j \phi_{3,j}$,
+respectively. These interactions are accelerated by interpolating these terms
+through a grid of equispaced interpolation points. This shifts the
+computational burden onto the interpolation points and reduces the time
+complexity to $\mathcal{O}(N)$.
 
-\opentsne\ provides efficient implementations for all the aforementioned improvements but defaults to using approximate nearest neighbor search and the non-uniform convolutions approach as these enable the user to quickly create informative visualizations for data sets containing up to millions of samples.
+\opentsne\ provides efficient implementations for all the aforementioned
+improvements but defaults to using approximate nearest neighbor search and the
+non-uniform convolutions approach as these enable the user to quickly create
+informative visualizations for data sets containing up to millions of samples.
 
@@
 
-t-SNE is non-parametric and does not define an explicit mapping from the high-dimensional space to the embedding space.
-Therefore embeddings of new data points need to be found through the use of optimization techniques~\cite{policar2019embedding}.
-When adding new data points to an existing, reference embedding, the reference data points are fixed in place and the new data points are allowed to find their respective positions.
-The optimization remains the same as in standard t-SNE with only slight modifications to $p_{ij}$ and $q_{ij}$
+t-SNE is non-parametric and does not define an explicit mapping from the
+high-dimensional space to the embedding space. Therefore embeddings of new data
+points need to be found through the use of optimization
+techniques~\cite{policar2019embedding}. When adding new data points to an
+existing, reference embedding, the reference data points are fixed in place and
+the new data points are allowed to find their respective positions. The
+optimization remains the same as in standard t-SNE with only slight
+modifications to $p_{ij}$ and $q_{ij}$
 \begin{align}
@@
 
-As in standard t-SNE, a direct calculation of gradients takes $\mathcal{O}(N \cdot M)$ time, but it is straightforward to adapt the Barnes-Hut and FIt-SNE approximation schemes, reducing the time complexity to $\mathcal{O}(M \log N)$ and $\mathcal{O}(\max \{ N, M \})$, respectively.
-In the FIt-SNE approximation scheme, we additionally exploit the fact that the reference embedding remains fixed throughout the optimization of newly added points, and precompute the interpolation grid.
-This further reduces the runtime complexity from $\mathcal{O}(\max \{ N, M \})$ to $\mathcal{O}(M)$.
+As in standard t-SNE, a direct calculation of gradients takes $\mathcal{O}(N
+\cdot M)$ time, but it is straightforward to adapt the Barnes-Hut and FIt-SNE
+approximation schemes, reducing the time complexity to $\mathcal{O}(M \log N)$
+and $\mathcal{O}(\max \{ N, M \})$, respectively. In the FIt-SNE approximation
+scheme, we additionally exploit the fact that the reference embedding remains
+fixed throughout the optimization of newly added points, and precompute the
+interpolation grid. This further reduces the runtime complexity from
+$\mathcal{O}(\max \{ N, M \})$ to $\mathcal{O}(M)$.
 
@@
 
-In standard t-SNE, distances are converted to similarities through the use of Gaussian kernels of varying bandwidths.
-The bandwidths are indirectly determined by the user-specified perplexity parameter so that a fixed number of nearest data points will have non-zero similarity.
-One common trick for uncovering the global relations between clusters is to increase perplexity so that more long-range interactions are preserved in the final embedding.
-However, one unfortunate side effect of increasing perplexity is that smaller clusters get absorbed into larger ones.
+In standard t-SNE, distances are converted to similarities through the use of
+Gaussian kernels of varying bandwidths. The bandwidths are indirectly
+determined by the user-specified perplexity parameter so that a fixed number of
+nearest data points will have non-zero similarity. One common trick for
+uncovering the global relations between clusters is to increase perplexity so
+that more long-range interactions are preserved in the final embedding.
+However, one unfortunate side effect of increasing perplexity is that smaller
+clusters get absorbed into larger ones.
 
-Kobak \& Berens~\cite{kobak2019art} suggest that replacing the Gaussian kernel with a mixture of Gaussians may provide better insight into both the local and global structure.
-Therefore, the similarities between data points in the input space become
+Kobak \& Berens~\cite{kobak2019art} suggest that replacing the Gaussian kernel
+with a mixture of Gaussians may provide better insight into both the local and
+global structure. Therefore, the similarities between data points in the input
+space become
 
@@
 
-The bandwidths of each Gaussian $\sigma_{1,i}$ and $\sigma_{2,i}$ is determined by selecting different perplexity values.
-They show that using a kernel with perplexity 500 captures the long-range interactions which preserve global cluster organization in the final embedding and combine this with a kernel with perplexity 50, which prevents small, well-defined clusters from being absorbed into larger ones.
+The bandwidths of each Gaussian $\sigma_{1,i}$ and $\sigma_{2,i}$ is determined
+by selecting different perplexity values. They show that using a kernel with
+perplexity 500 captures the long-range interactions which preserve global
+cluster organization in the final embedding and combine this with a kernel with
+perplexity 50, which prevents small, well-defined clusters from being absorbed
+into larger ones.
 
@@
 
-Standard t-SNE reveals the clustering structure in data at one level of resolution.
-Different perplexity parameter values can be used to identify global cluster relationships or small, well-isolated groups.
-Unfortunately, varying perplexity values can be time-consuming as this involves recomputing the KNNG which is often the most expensive part of the t-SNE algorithm.
-Alternatively, Kobak \textit{et al.}~\cite{kobak2019heavy} suggest that varying the degree of freedom in the t-distribution used in the embedding space during optimization can be used to explore the clustering structure at different levels of resolution.
+Standard t-SNE reveals the clustering structure in data at one level of
+resolution. Different perplexity parameter values can be used to identify
+global cluster relationships or small, well-isolated groups. Unfortunately,
+varying perplexity values can be time-consuming as this involves recomputing
+the KNNG which is often the most expensive part of the t-SNE algorithm.
+Alternatively, Kobak \textit{et al.}~\cite{kobak2019heavy} suggest that varying
+the degree of freedom in the t-distribution used in the embedding space during
+optimization can be used to explore the clustering structure at different
+levels of resolution.
 
-Standard t-SNE models similarities between data points in the embedding space using a t-distribution with a single degree of freedom
+Standard t-SNE models similarities between data points in the embedding space
+using a t-distribution with a single degree of freedom
 
@@
 
-In standard t-SNE $\alpha=1$ so this simplifies to the Cauchy kernel from Eqn.~(\ref{eq:cauchy_kernel}).
+In standard t-SNE $\alpha=1$ so this simplifies to the Cauchy kernel from
+Eqn.~(\ref{eq:cauchy_kernel}).
 
@@
 Adapting existing approximation schemes to this formulation is straightforward.
-We provide efficient implementations of both the Barnes-Hut and the FIt-SNE approximation schemes, where we replace the terms from Eqn.~(\ref{eq:fitsne_terms}) with
+We provide efficient implementations of both the Barnes-Hut and the FIt-SNE
+approximation schemes, where we replace the terms from
+Eqn.~(\ref{eq:fitsne_terms}) with
 
@@
 
-Embeddings produced by standard t-SNE often produce clusters separated by thin boundaries and make use of all available space.
-While this is often desirable, this often obscures the global relationships between clusters as all neighboring clusters appear at the same distance from one another.
-Other dimensionality reduction methods such as UMAP~\cite{2018arXivUMAP} or ForceAtlas2~\cite{jacomy2014forceatlas2} tend to produce embeddings where clusters appear more compact and the white-space separating the clusters may be interpreted at least partially as a loose measure of distance.
+Embeddings produced by standard t-SNE often produce clusters separated by thin
+boundaries and make use of all available space. While this is often desirable,
+this often obscures the global relationships between clusters as all
+neighboring clusters appear at the same distance from one another. Other
+dimensionality reduction methods such as UMAP~\cite{2018arXivUMAP} or
+ForceAtlas2~\cite{jacomy2014forceatlas2} tend to produce embeddings where
+clusters appear more compact and the white-space separating the clusters may be
+interpreted at least partially as a loose measure of distance.
 
-Kobak \textit{et al.}~\cite{todo} showed that the early exaggeration factor $\rho$ in t-SNE may be increased to produce more compact layouts often seen in UMAP or ForceAtlas2.
-Early exaggeration is a trick introduced by van der Maaten~\cite{maaten2008visualizing} used in the first 250 iterations and increases the attractive forces between data points, enabling neighboring points to move close to one another while ignoring the repulsive forces between data points.
-In standard t-SNE, the remainder of the iterations are run with $\rho=1$.
-Kobak \textit{et al.} showed that using $\rho=4$ and $\rho=30$ in the second phase of the optimization results in embeddings visually similar to UMAP embeddings, and ForceAtlas2 embeddings, respectively.
+Kobak \textit{et al.}~\cite{todo} showed that the early exaggeration factor
+$\rho$ in t-SNE may be increased to produce more compact layouts often seen in
+UMAP or ForceAtlas2. Early exaggeration is a trick introduced by van der
+Maaten~\cite{maaten2008visualizing} used in the first 250 iterations and
+increases the attractive forces between data points, enabling neighboring
+points to move close to one another while ignoring the repulsive forces between
+data points. In standard t-SNE, the remainder of the iterations are run with
+$\rho=1$. Kobak \textit{et al.} showed that using $\rho=4$ and $\rho=30$ in the
+second phase of the optimization results in embeddings visually similar to UMAP
+embeddings, and ForceAtlas2 embeddings, respectively.
 
-While it is difficult to claim one is better than the other, different settings of $\rho$ may uncover different properties and clustering structures.
-For example, in single-cell data, standard t-SNE with $\rho=1$ often uncovers distinct groups of cell-types but obscures developmental paths.
-This problem is exacerbated when dealing with large numbers of data points.
-Conversely, ForceAtlas2 has often been used to uncover trajectories and transitions between cell types, but larger clusters often absorb small, distinct groups of cells.
+While it is difficult to claim one is better than the other, different settings
+of $\rho$ may uncover different properties and clustering structures. For
+example, in single-cell data, standard t-SNE with $\rho=1$ often uncovers
+distinct groups of cell-types but obscures developmental paths. This problem is
+exacerbated when dealing with large numbers of data points. Conversely,
+ForceAtlas2 has often been used to uncover trajectories and transitions between
+cell types, but larger clusters often absorb small, distinct groups of cells.
 
@@
 
-Similarly to UMAP and ForceAtlas2, t-SNE can also be viewed as a force-directed layout algorithm.
-The first phase of the t-SNE algorithm constructs a $k$-nearest neighbor graph, and the subsequent optimization balances the attractive and repulsive forces between data points.
+Similarly to UMAP and ForceAtlas2, t-SNE can also be viewed as a force-directed
+layout algorithm. The first phase of the t-SNE algorithm constructs a
+$k$-nearest neighbor graph, and the subsequent optimization balances the
+attractive and repulsive forces between data points.
 
-The final positioning of data points using this class of algorithms is largely dependent on the initial positions in the embedding space.
-UMAP is often said to better preserve global structure than t-SNE, but Kobak \& Linderman have recently discovered that this supposed advantage is entirely due to the better default initialization in UMAP than in older implementations of t-SNE~\cite{kobak2019umap}.
-Standard implementations of t-SNE use random initializations, which often scatter resulting clusters throughout the embedding space.
-UMAP injects global organization into the embedding by initializing point positions using Laplacian eigenmaps.
-When initialized in the same manner, both t-SNE and UMAP perform equally well, or equally poorly, dispelling the myth that UMAP is superior in this respect.
+The final positioning of data points using this class of algorithms is largely
+dependent on the initial positions in the embedding space. UMAP is often said
+to better preserve global structure than t-SNE, but Kobak \& Linderman have
+recently discovered that this supposed advantage is entirely due to the better
+default initialization in UMAP than in older implementations of
+t-SNE~\cite{kobak2019umap}. Standard implementations of t-SNE use random
+initializations, which often scatter resulting clusters throughout the
+embedding space. UMAP injects global organization into the embedding by
+initializing point positions using Laplacian eigenmaps. When initialized in the
+same manner, both t-SNE and UMAP perform equally well, or equally poorly,
+dispelling the myth that UMAP is superior in this respect.
 
-\opentsne\ defaults to using the two leading principal components as initialization, but also provides a spectral approach, similar to that of UMAP.
-The clusters in resulting embeddings are organized in a globally coherent manner, leading to more interpretable visualizations.
+\opentsne\ defaults to using the two leading principal components as
+initialization, but also provides a spectral approach, similar to that of UMAP.
+The clusters in resulting embeddings are organized in a globally coherent
+manner, leading to more interpretable visualizations.
 
@@
 
-openTSNE is distributed under the BSD-3-Clause License and is publicly available as an open-source package at \url{https://github.com/pavlin-policar/openTSNE}.
-openTSNE is also available on PyPI and conda-forge.
-The data sets used and scripts used in this study are included in accompanying notebooks, publicly available at \url{https://github.com/pavlin-policar/opentsne-paper}.
+openTSNE is distributed under the BSD-3-Clause License and is publicly
+available as an open-source package at
+\url{https://github.com/pavlin-policar/openTSNE}. openTSNE is also available on
+PyPI and conda-forge. The data sets used and scripts used in this study are
+included in accompanying notebooks, publicly available at
+\url{https://github.com/pavlin-policar/opentsne-paper}.
 
@@
 \section*{Acknowledgements}
-We would like to thank Dmitry Kobak for various helpful discussions and best practices when using t-SNE, as well as his contributions to the source code.
-We would also like to thank George Linderman for his help with the FIt-SNE algorithm.
+We would like to thank Dmitry Kobak for various helpful discussions and best
+	practices when using t-SNE, as well as his contributions to the source
+	code. We would also like to thank George Linderman for his help with
+	the FIt-SNE algorithm.
 

### 2020-07-06 paper/policar2020-opentsne.tex
Incorporate feature list table

@@
   \caption{\label{fig:benchmarks}We benchmark openTSNE (v0.4.3) against three
-	popular open-source implementations including
+	popular open-source implementations from
 	scikit-learn~\cite{pedregosa2011scikit} (v0.23.1),
@@
 researchers to quickly experiment with different settings and easily
-incorporate their own components into the software.
+incorporate their own components into the software. We provide a full feature
+list and comparison to other popular t-SNE implementations in
+Table~\ref{tab:features}.
+
+\begin{table}
+\caption{\label{tab:features}We compare the feature list of \opentsne to three popular open-source implementations from scikit-learn~(v0.23.1), MulticoreTSNE (v0.1), and FIt-SNE (v1.1.0).
+The first section corresponds to packaging and distribution.
+A properly packaged library is easily accessible to users and can easily be included as a dependency in other software packages.
+The second section lists the two approximation schemes.
+The FIt-SNE approximation scheme is required for t-SNE to scale up to millions of data points.
+The final section provides a list of extensions and improvements to the standard t-SNE algorithm, many of which can produce markedly better visualizations than under default parameter settings.}
+
+\newcommand*\rot{\rotatebox{90}}
+\renewcommand{\arraystretch}{1.25}
+
+\begin{tabular}{r c c c c c c}
+\toprule
+\setlength\tabcolsep{6pt}
+& \rot{\textsf{scikit-learn}} & \rot{\textsf{MulticoreTSNE}} & \rot{\textsf{FIt-SNE}} & \rot{\textsf{openTSNE}} \\
+\toprule
+PyPI package & \checkmark & \checkmark & & \checkmark \\
+conda package & \checkmark & & & \checkmark \\
+Precompiled binaries & \checkmark & & & \checkmark \\
+\hline
+Barnes-Hut ($\mathcal{O}(N \log N)$) & \checkmark & \checkmark & & \checkmark \\
+FIt-SNE ($\mathcal{O}(N)$) & & & \checkmark & \checkmark \\
+\hline
+Multiscale Gaussian kernels & & & \checkmark & \checkmark \\
+Fully-custom affinity kernels & & & & \checkmark \\
+Variable degrees of freedom & & & \checkmark & \checkmark \\
+Variable exaggeration & & & \checkmark & \checkmark \\
+Better initialization & & & \checkmark & \checkmark \\
+Automatic learning rate & & & \checkmark & \checkmark \\
+Embedding new samples & & & & \checkmark \\
+\bottomrule
+\end{tabular}
+\end{table}
 
@@
 
-\begin{table}
-\caption{\label{tab:features}Feature list...}
-
-\newcommand*\rot{\rotatebox{90}}
-\renewcommand{\arraystretch}{1.25}
-
-\begin{tabular}{r c c c c c c}
-\toprule
-\setlength\tabcolsep{6pt}
-& \rot{\textsf{scikit-learn}} & \rot{\textsf{MulticoreTSNE}} & \rot{\textsf{FIt-SNE}} & \rot{\textsf{openTSNE}} \\
-\toprule
-PyPI package & \checkmark & \checkmark & & \checkmark \\
-conda package & \checkmark & & & \checkmark \\
-Precompiled binaries & \checkmark & & & \checkmark \\
-\hline
-Barnes-Hut ($\mathcal{O}(N \log N)$) & \checkmark & \checkmark & & \checkmark \\
-FIt-SNE ($\mathcal{O}(N)$) & & & \checkmark & \checkmark \\
-\hline
-Multiscale Gaussian kernels & & & \checkmark & \checkmark \\
-Fully-custom affinity kernels & & & & \checkmark \\
-Variable degrees of freedom & & & \checkmark & \checkmark \\
-Variable exaggeration & & & \checkmark & \checkmark \\
-Better initialization & & & \checkmark & \checkmark \\
-Automatic learning rate & & & \checkmark & \checkmark \\
-Embedding new samples & & & & \checkmark \\
-\bottomrule
-\end{tabular}
-\end{table}
-
 \section*{Availability}

### 2020-07-07 paper/policar2020-opentsne.tex
Reformat

@@
 and all its recently proposed extensions. Most notable among these are
-\begin{enumerate} \item the implementation of efficient approximation
-			schemes~\cite{van2014accelerating,linderman2019fast}
-			allowing the embedding of millions of data points,
+\begin{enumerate}
+\item the implementation of efficient approximation 
+	schemes~\cite{van2014accelerating,linderman2019fast} allowing the embedding
+	of millions of data points,
 \item the addition of new samples into existing
-		embeddings~\cite{policar2019embedding}, \item better
-			initialization schemes~\cite{kobak2019umap} leading to
-			more globally consistent layouts, \item variable
-			degrees of freedom~\cite{kobak2019heavy} allowing the
-		inspection of data at different levels of resolution, \item
-			multiscale similarity kernels~\cite{kobak2019art} which
-			preserves small, well-defined clusters and uncovers
-			global relationships between clusters, and \item
-improved defaults for learning rate and the number of
+	embeddings~\cite{policar2019embedding},
+\item better initialization schemes~\cite{kobak2019umap} leading to more
+	globally consistent layouts,
+\item variable degrees of freedom~\cite{kobak2019heavy} allowing the inspection
+	of data at different levels of resolution,
+\item multiscale similarity kernels~\cite{kobak2019art} which preserves small,
+	well-defined clusters and uncovers global relationships between clusters,
+	and
+\item improved defaults for learning rate and the number of
 	iterations~\cite{belkina2019automated} which produce embeddings at a lower
-computational cost.  \end{enumerate}
+	computational cost.
+\end{enumerate}
 
@@
 of which provides a different perspective into the topology and other
-characteristics of the data.  \begin{figure*}[htbp]
+characteristics of the data.
+\begin{figure*}[htbp]
 	\includegraphics[width=\textwidth]{tasic2018}
@@
 	smaller clusters by using a more heavy-tailed kernel \textbf{(d)}.}
-\end{figure*} For instance, in Fig.~\ref{fig:tasic}, we show four embeddings of
+\end{figure*}
+or instance, in Fig.~\ref{fig:tasic}, we show four embeddings of
 the same gene-expression data from the mouse brain~\cite{tasic2018shared}.
@@
 
-Fig.~\ref{fig:benchmarks} shows benchmarks of four Python t-SNE implementations, including those from \textsf{scikit-learn} (v0.23.1), \textsf{MulticoreTSNE} (v0.1), \textsf{FIt-SNE} (v1.1.0), and \opentsne\ (v0.4.3).
-We perform benchmarks on two processors, one representing usage on a personal computer and one on a high-performance computing platform.
-The Intel(R) Core i7 is commonly found in consumer-grade laptop computers, while Intel(R) Xeon(R) processors are found in high-performance computing machines.
-Benchmarks were run with the original t-SNE parameters for 1000 iterations, as some implementations do not allow their modification.
+Fig.~\ref{fig:benchmarks} shows benchmarks of four Python t-SNE implementations,
+including those from \textsf{scikit-learn} (v0.23.1), \textsf{MulticoreTSNE}
+(v0.1), \textsf{FIt-SNE} (v1.1.0), and \opentsne\ (v0.4.3). We perform
+benchmarks on two processors, one representing usage on a personal computer and
+one on a high-performance computing platform. The Intel(R) Core i7 is commonly
+found in consumer-grade laptop computers, while Intel(R) Xeon(R) processors are
+found in high-performance computing machines. Benchmarks were run with the
+original t-SNE parameters for 1000 iterations, as some implementations do not
+allow their modification.
 
@@
 \begin{table}
-\caption{\label{tab:features}We compare the feature list of \opentsne to three popular open-source implementations from scikit-learn~(v0.23.1), MulticoreTSNE (v0.1), and FIt-SNE (v1.1.0).
-The first section corresponds to packaging and distribution.
-A properly packaged library is easily accessible to users and can easily be included as a dependency in other software packages.
-The second section lists the two approximation schemes.
-The FIt-SNE approximation scheme is required for t-SNE to scale up to millions of data points.
-The final section provides a list of extensions and improvements to the standard t-SNE algorithm, many of which can produce markedly better visualizations than under default parameter settings.}
+\caption{\label{tab:features}We compare the feature list of \opentsne to three
+popular open-source implementations from scikit-learn~(v0.23.1), MulticoreTSNE
+(v0.1), and FIt-SNE (v1.1.0). The first section corresponds to packaging and 
+distribution. A properly packaged library is easily accessible to users and can
+easily be included as a dependency in other software packages. The second
+section lists the two approximation schemes. The FIt-SNE approximation scheme
+is required for t-SNE to scale up to millions of data points. The final section
+provides a list of extensions and improvements to the standard t-SNE algorithm,
+many of which can produce markedly better visualizations than under default
+parameter settings.}
 

### 2020-07-07 paper/policar2020-opentsne.tex
Minor corrections

@@
 	of data at different levels of resolution,
-\item multiscale similarity kernels~\cite{kobak2019art} which preserves small,
-	well-defined clusters and uncovers global relationships between clusters,
+\item multi-scale similarity kernels~\cite{kobak2019art} which preserve small,
+	well-defined clusters and uncover global relationships between clusters,
 	and
-\item improved defaults for learning rate and the number of
+\item improved defaults for learning rate and number of
 	iterations~\cite{belkina2019automated} which produce embeddings at a lower
@@
 \textsf{numpy}, \textsf{scikit-learn}, \textsf{scanpy}) providing a familiar
-and intuitive user interface to new users. Its modular design encourages
+and intuitive API to new users. Its modular design encourages
 extensibility and experimentation with various settings and changes in the
@@
 Fig.~\ref{fig:tasic} depicts four different t-SNE visualizations generated by
-\opentsne, each based on recent advances in the t-SNE algorithm. The code used
+\opentsne, each based on recent theoretical advances in the t-SNE algorithm. The code used
 to generate these embeddings is shown below.
@@
 any data set can accurately be described using only two dimensions and that any
-projection onto a two-dimensional plane will inevitably lead to loss of
+embedding onto a two-dimensional plane will inevitably lead to loss of
 information. Therefore, it is beneficial to examine multiple embeddings, each
@@
 	\caption{\label{fig:tasic}We use openTSNE to create four different
-	visualizations of the Tasic \textit{et al.}~\cite{tasic2018shared} data
-	set containing 21,874 single-cells originating from the mouse
+    visualizations of the Tasic \textit{et al.}~\cite{tasic2018shared} data,
+    each providing a different perspective into the topology of the data.
+    The data set contains 21,874 single-cells originating from the mouse
 	neocortex. Cluster annotations and colors are taken from the original
@@
 \end{figure*}
-or instance, in Fig.~\ref{fig:tasic}, we show four embeddings of
-the same gene-expression data from the mouse brain~\cite{tasic2018shared}.
-Fig.~\ref{fig:tasic}.a shows an embedding using default t-SNE parameters, which
-uncovers numerous clusters. Warm colors correspond to excitatory neurons, while
+For instance, in Fig.~\ref{fig:tasic}, we show four embeddings of
+the same gene-expression data containing cells from the mouse brain~\cite{tasic2018shared}.
+Fig.~\ref{fig:tasic}.a shows an embedding using default t-SNE parameters.
+Warm colors correspond to excitatory neurons, while
 cool colors represent inhibitory neurons. While different clusters of
 excitatory and inhibitory neurons appear close to one another, all clusters
-appear somewhat equidistant from their neighboring clusters, and the overall
-relations between them are not obvious. On the other hand,
+appear somewhat equidistant from their neighbors, and the overall
+relations between groups are not obvious. On the other hand,
 Fig.~\ref{fig:tasic}.b focuses on preserving larger neighborhoods of points,
@@
 values as Fig.~\ref{fig:tasic}.a, but at a finer level of resolution. This
-reveals that many of the clusters are composed of other subgroups.
-
-When faced with a data set containing millions of data points, standard t-SNE
-embeddings often become unwieldy, with blurred cluster boundaries, large
-clusters absorbing smaller ones, and relationships between clusters becoming
+reveals that many of the clusters are composed of other subgroups and may be
+clustered further into cell sub-types.
+
+When working with data containing millions of data points, standard t-SNE
+embeddings often become unwieldy -- cluster boundaries become blurred, large
+clusters absorb smaller ones, and relationships between clusters become
 increasingly difficult to interpret. Fig.~\ref{fig:cao}.a contains expression
@@
 main groups of cells and eight somewhat smaller clusters. The visualization
-also reveals a small number of very small clusters, possibly corresponding to
+also reveals a number of tiny clusters, possibly corresponding to
 rare cell types.
@@
   \includegraphics[width=\textwidth]{cao2019}
-  \caption{\label{fig:cao}The data set from Cao \textit{et
+  \caption{\label{fig:cao}Increasing the exaggeration parameter produces more
+  	compact clusters, which highlight the global organization of the clusters
+  	and emphasizes continuities between cell states. The data set from Cao \textit{et
 	al.}~\cite{cao2019single} contains expression profiles from 2,058,652
@@
 	developmental time-points at daily intervals after 9.5 to 13.5 days.
-	The exaggeration parameter can be used to highlight the global
-	organization of the clusters and enforce cluster compactness.
 	\textbf{(a)} shows the data is comprised of two main components -- the
@@
 
-Exaggeration can be used to highlight transitions between cell-states in
+Exaggeration can highlight transitions between cell-states in
 developmental studies. Standard t-SNE often produces embeddings with
 clearly-defined, discrete clusters. We have shown that we can adjust the level
-of granularity and resolution of the clusters with several parameters. However,
+of granularity and resolution of the clusters with several parameters in Fig.~\ref{fig:tasic}. However,
 discrete clusters are often undesired in developmental studies where we assume
@@
 between cells in Fig.~\ref{fig:cao}.c, where we have increased the exaggeration
-factor to $4$.
+factor from $1$ to $4$.
 
@@
 cells from the mouse brain. The authors provide annotations for each cell, and
-we can verify their accuracy by visualizing well-established gene markers'
+we can verify their accuracy by visualizing well-established gene markers for the
 major cell types. We then embed cells from Harris \textit{et
@@
 insight into newly-sequenced, single-cell data sets by utilizing already
-existing, publicly-available, reference atlases. This approach is extremely
+existing cell atlases. This approach is extremely
 general and is not limited to single-cell gene-expression data, but can be
@@
 extremely flexible API, where these improvements can be incorporated in just a
-few lines of code. Furthermore, \opentsne\ supports custom affinity models,
-e.g. multi-scale Gaussian mixtures or uniform kernels, allowing for quick
-experimentation. This also enables users to construct t-SNE embeddings on
+few lines of code. Furthermore, \opentsne\ supports fully-custom affinity models,
+enabling users to construct t-SNE embeddings on
 non-tabular data, e.g. relational data, as the only affinity-model requirement
@@
 Fig.~\ref{fig:benchmarks} demonstrates that both \textsf{FIt-SNE} and
-\opentsne\ exhibit better runtime performance than their Barnes-Hut
+\opentsne\ exhibit better scaling than their Barnes-Hut
 counterparts -- \textsf{scikit-learn} and \textsf{MulticoreTSNE}. Notice that
@@
 written in Python, which inevitably incurs some runtime overhead compared to
-its pure C++ counterpart. However, Fig.~\ref{fig:benchmarks} demonstrates the
+its pure C++ counterpart. However, Fig.~\ref{fig:benchmarks} shows the
 runtime of \opentsne\ is comparable to that of \textsf{FIt-SNE} and much faster
@@
 server-grade processor, \opentsne\ is slightly faster than its pure C++
-counterpart. We speculate that this is could be because \opentsne\ uses
+counterpart when utilizing multiple cores. We speculate that this is could be because \opentsne\ uses
 \textsf{numpy} for most linear algebra operations, which may make better use of
-the Intel Math Kernel Library (MKL), which are aggressively optimized on
+the Intel Math Kernel Library (MKL), which is aggressively optimized on
 Intel(R) Xeon(R) processors.
@@
 This section is meant to familiarize the reader with the t-SNE algorithm and
-its recent extensions. This brief review is by no means comprehensive, but is
+its recent extensions. We motivate each extension and introduce standard notation.
+This brief review is by no means comprehensive, but is
 self-contained and introduces the mathematics upon which the \opentsne\ library

### 2020-07-12 paper/policar2020-opentsne.tex
Add references to introduction.

@@
 (t-SNE)~\cite{maaten2008visualizing}, and uniform manifold approximation and
-projections (UMAP)~\cite{2018arXivUMAP}. Among these, t-SNE lately received
-much attention as it can address high volumes of data and reveal meaningful
-clustering structure. Most of the recent reports on single-cell gene expression
-data start with an overview of the cell landscape, where t-SNE embeds
+projections (UMAP)~\cite{2018arXivUMAP}. 
+Among these, t-SNE lately received much attention as it can address high
+volumes of data and reveal meaningful clustering structure. Increased
+resolution and throughput of modern molecular assays has lead to frequent use
+of t-SNE in diverse fields including, but not limited to, single-cell
+transcriptomics~\cite{macosko2015highly,cao2019single,tasic2018shared}, human
+genetics~\cite{hirata2019genetic}, metagenomic
+assembly~\cite{beaulaurier2018metagenomic}, spatial organization of microbial
+communities~\cite{sheth2019spatial} and
+metabolomics~\cite{tkachev2019differences}.
+
+Reports on single cell gene expression data, our running example,
+often start with an overview of the cell landscape, where t-SNE embeds
 high-dimensional expression profiles into a two-dimensional
-space~\cite{macosko2015highly,cao2019single,tasic2018shared}.
-Figs.~\ref{fig:macosko}.a and \ref{fig:macosko}.b are examples of two such
+space. Figs.~\ref{fig:macosko}.a and \ref{fig:macosko}.b are examples of two such
 embeddings.
@@
 characteristics of the data.
+
 \begin{figure*}[htbp]

### 2020-07-12 paper/policar2020-opentsne.tex
Text edits.

--- 2020-07-07 paper/policar2020-opentsne.tex
@@
 The abundance of high-dimensional data sets in molecular biology calls for
-techniques for dimensionality reduction, and in particular for methods that can
-help in the construction of data visualizations. Popular approaches for
-dimensionality reduction include principal component analysis (PCA),
+techniques for dimensionality reduction, in particular for methods that produce informative data visualizations. Popular approaches include 
+principal component analysis (PCA),
 multidimensional scaling, t-distributed stochastic neighbor embedding
 (t-SNE)~\cite{maaten2008visualizing}, and uniform manifold approximation and
-projections (UMAP)~\cite{2018arXivUMAP}. Among these, t-SNE lately received
-much attention as it can address high volumes of data and reveal meaningful
-clustering structure. Most of the recent reports on single-cell gene expression
-data start with an overview of the cell landscape, where t-SNE embeds
+projections (UMAP)~\cite{2018arXivUMAP}. 
+Among these, t-SNE lately received much attention as it can address high
+volumes of data and reveal meaningful clustering structure. Increased
+resolution and throughput of modern molecular assays has lead to frequent use
+of t-SNE in diverse fields including, but not limited to, single-cell
+transcriptomics(scRNA-seq,~\cite{macosko2015highly,cao2019single,tasic2018shared}),
+human genetics~\cite{hirata2019genetic}, metagenomic
+assembly~\cite{beaulaurier2018metagenomic}, spatial organization of microbial
+communities~\cite{sheth2019spatial} and
+metabolomics~\cite{tkachev2019differences}.
+
+Reports on single-cell gene expression data, our running example,
+often start with an overview of the cell landscape, where t-SNE embeds
 high-dimensional expression profiles into a two-dimensional
-space~\cite{macosko2015highly,cao2019single,tasic2018shared}.
-Figs.~\ref{fig:macosko}.a and \ref{fig:macosko}.b are examples of two such
+space. Figs.~\ref{fig:macosko}.a and \ref{fig:macosko}.b are examples of two such
 embeddings.
@@
 	from Maaten \& Hinton~\cite{maaten2008visualizing}. The visualization
-	shows poor preservation of the global organization of clusters,
+	shows no preservation of the global organization of clusters,
 	resulting from random initialization and an affinity model focused on
@@
 
-Despite its utility, t-SNE has often been criticized for poor scalability when
-addressing large data sets, lack of global organization -- t-SNE focuses on
+Despite its utility, t-SNE has often been criticized for its limited scalability, lack of global organization -- t-SNE emphasizes
 local clusters that are arbitrarily scattered in the low-dimensional space --
-and the absence of theoretically-founded implementations to map new data into
+and the absence of theoretically-founded methods to map new data into
 existing embeddings~\cite{ding2018interpretable,becht2019dimensionality}. Most
 of these shortcomings have recently been addressed. Linderman \textit{et al.}
-sped-up the method through an interpolation-based approximation, reducing the
-time complexity to be merely linear dependent on the number of
+sped-up the method through an interpolation-based approximation, achieving linear complexity in the number of
 samples~\cite{linderman2019fast}. Kobak \& Berens proposed several techniques
@@
 \item improved defaults for learning rate and number of
-	iterations~\cite{belkina2019automated} which produce embeddings at a lower
+	iterations~\cite{belkina2019automated} which produce embeddings at a low
 	computational cost.
@@
 
-\opentsne\ is open-source and has become an integral part of the Python data
-science ecosystem. The package has received over 600 Github stars and has been
-downloaded hundreds of thousands of times, averaging in thousands of weekly
-downloads. The source code is publicly available on Github at
+\opentsne\ is open-source and has become adopted by the wider community,
+receeiving over 600 Github stars and averaging thousands of weekly downloads.
+The source code is publicly available on Github at
 \url{https://github.com/pavlin-policar/openTSNE}, and can be installed from
@@
 data lies on a lower-dimensional manifold, which can accurately be captured by
-a smaller number of dimensions. However, there is no evidence to suggest that
-any data set can accurately be described using only two dimensions and that any
-embedding onto a two-dimensional plane will inevitably lead to loss of
-information. Therefore, it is beneficial to examine multiple embeddings, each
-of which provides a different perspective into the topology and other
-characteristics of the data.
+a smaller number of dimensions. However, there is no evidence that
+every data set can accurately be described by using only two dimensions, and any such embedding will therefore inevitably lead to loss of
+information. It is thus beneficial to examine multiple embeddings, each
+of which provides a different perspective into topology and other
+data characteristics.
+
 \begin{figure*}[htbp]
@@
 \end{figure*}
-For instance, in Fig.~\ref{fig:tasic}, we show four embeddings of
-the same gene-expression data containing cells from the mouse brain~\cite{tasic2018shared}.
-Fig.~\ref{fig:tasic}.a shows an embedding using default t-SNE parameters.
-Warm colors correspond to excitatory neurons, while
-cool colors represent inhibitory neurons. While different clusters of
-excitatory and inhibitory neurons appear close to one another, all clusters
-appear somewhat equidistant from their neighbors, and the overall
+
+We illustrate this point by generating four embeddings of the same data on
+single-cell gene expression in mouse brain~\cite{tasic2018shared}.  First,
+Fig.~\ref{fig:tasic}.a shows an embedding using default t-SNE parameters. While
+different clusters of excitatory and inhibitory neurons appear close to one
+another, all clusters appear equidistant from their neighbors, and the overall
 relations between groups are not obvious. On the other hand,
@@
 become more apparent. In this embedding, it is evident from the increased
-white-space between groups that there is one large class of excitatory neurons,
+white space between groups that there is one large class of excitatory neurons,
 and two related classes of inhibitory neurons. Unfortunately, focusing on
@@
 reveals that many of the clusters are composed of other subgroups and may be
-clustered further into cell sub-types.
+clustered further into cell subtypes.
 
@@
 developmental studies. Standard t-SNE often produces embeddings with
-clearly-defined, discrete clusters. We have shown that we can adjust the level
+clearly defined, discrete clusters. We have shown that we can adjust the level
 of granularity and resolution of the clusters with several parameters in Fig.~\ref{fig:tasic}. However,
-discrete clusters are often undesired in developmental studies where we assume
-cells follow a continuous path between cell states. To this end, other
-embedding techniques such as UMAP and ForceAtlas2~\cite{jacomy2014forceatlas2}
-are often used, as they tend to better capture the continuity between cell
+discrete clusters are often undesired in developmental studies where cells are assumed to follow a continuous trajectory between cell states. To this end, other
+embedding techniques such as UMAP and ForceAtlas2~\cite{jacomy2014forceatlas2} tend to better capture the continuity between cell
 states. Recently, Kobak \textit{et al.}~\cite{TODO} have shown that embeddings
@@
 which are markedly similar to UMAP and ForceAtlas2, respectively. For example,
-in Fig.~\ref{fig:cao}.a, the developmental path between different time points
-is difficult to observe due to a large number of sprawled out clusters in the
-embedding. On the other hand, the developmental paths are easy to follow
-between cells in Fig.~\ref{fig:cao}.c, where we have increased the exaggeration
-factor from $1$ to $4$.
+in Fig.~\ref{fig:cao}.a, the developmental trajectory between different time
+points is difficult to observe due to a large number of sprawled out clusters.
+On the other hand, the development becomes easier to follow when the
+exaggeration factor is increased from $1$ to $4$ (Fig.~\ref{fig:cao}.b-c).
 
@@
 Figs.~\ref{fig:macosko}.b and \ref{fig:macosko}.c demonstrate how we can use a
-previously annotated single-cell data set and embed data from a separate
-experiment into the reference landscape. The reference data taken from Macosko
-\textit{et al.}~\cite{macosko2015highly} contains gene-expression profiles from
-mouse retinal cells. Embedding samples from a similar experiment focusing only
-on bipolar retinal cells by Shekhar \textit{et
-al.}~\cite{shekhar2016comprehensive}, we can observe that the majority of new
-cells correctly map to the bipolar cell clusters in the reference.
+previously labeled single-cell data set and embed cells from a separate
+experiment into the reference landscape. The reference data taken from Macosko~\textit{et al.}~\cite{macosko2015highly} contains gene expression profiles from
+mouse retinal cells. By embedding the samples from a similar experiment on bipolar retinal cells by Shekhar~\textit{et al.}~\cite{shekhar2016comprehensive}, 
+we can correctly map to the bipolar cell clusters in the reference.
 
 Embedding single cells into existing reference atlases can also be useful for
-cell-type classification in cases where cell identities are not known in
+cell type classification in cases where cell identities are not known in
 advance. For instance, in Fig.~\ref{fig:transform}, we construct a reference
-embedding using annotated data from Hochgerner \textit{et
+embedding using labeled data from Hochgerner~\textit{et
 al.}~\cite{hochgerner2018conserved} containing gene-expression profiles of
-cells from the mouse brain. The authors provide annotations for each cell, and
+cells from the mouse brain. The authors assign a cell type to each cell, and
 we can verify their accuracy by visualizing well-established gene markers for the
-major cell types. We then embed cells from Harris \textit{et
-al.}~\cite{harris2018classes}, where annotations are provided only for neuronal
+major cell types. We then embed cells from Harris~\textit{et
+al.}~\cite{harris2018classes}, where labels are provided only for neuronal
 cells. We quickly identify other non-neuronal cell types, including
@@
 	types of neurons, supporting glia, and other vascular cells (upper
-	left). Cell annotations are taken from the original publications. The
+	left). Cell labels are taken from the original publications. The
 	colors correspond to Fig.~\ref{fig:tasic}. We then embed new,
@@
 These two examples demonstrate how \opentsne\ can be used to quickly gain
-insight into newly-sequenced, single-cell data sets by utilizing already
-existing cell atlases. This approach is extremely
-general and is not limited to single-cell gene-expression data, but can be
-applied to any tabular data sets from any domain of science.
+insight into newly-sequenced, single-cell data sets by utilizing existing cell
+atlases. This approach is general and not limited to single-cell gene
+expression, but can be used with any tabular data set regardless of field.
 
@@
 of multi-scale similarity kernels, perplexity annealing, and increasing
-exaggeration when working with larger data sets. \opentsne\ provides an
-extremely flexible API, where these improvements can be incorporated in just a
-few lines of code. Furthermore, \opentsne\ supports fully-custom affinity models,
-enabling users to construct t-SNE embeddings on
-non-tabular data, e.g. relational data, as the only affinity-model requirement
-is some notion of similarity between data points. Finally, our comprehensive
-callback system can be utilized to monitor and adapt different stages of the
-optimization phase and has been used to construct visually appealing animations
-of the t-SNE optimization process.
+exaggeration when working with larger data sets. \opentsne\ provides a flexible
+API, where these improvements can be incorporated in just a few lines of code.
+Furthermore, \opentsne\ supports custom affinity models, enabling users to
+construct t-SNE embeddings on non-tabular data, e.g. relational data, as the
+only affinity-model requirement is a notion of similarity between data points,
+i.e. the distance matrix.  Finally, our comprehensive callback system can be
+utilized to monitor and adapt different stages of the optimization phase and
+has been used to construct visually appealing animations of the t-SNE
+optimization process.
 
@@
 
-One of t-SNE's common criticisms is poor scalability to large data sets
+One of t-SNE's common criticisms is limited scalability to large data sets
 containing up to millions of data points~\cite{becht2019dimensionality}. These
@@
 exacerbated further by the fact that the most widely-used implementation of
-t-SNE came from \textsf{scikit-learn}, which exhibits very poor runtime
+t-SNE came from \textsf{scikit-learn}, which exhibits long runtimes
 performance compared to its C++ counterpart -- \textsf{MulticoreTSNE}. The
@@
 
-Fig.~\ref{fig:benchmarks} shows benchmarks of four Python t-SNE implementations,
+Fig.~\ref{fig:benchmarks} benchmarks four Python t-SNE implementations,
 including those from \textsf{scikit-learn} (v0.23.1), \textsf{MulticoreTSNE}
@@
 found in consumer-grade laptop computers, while Intel(R) Xeon(R) processors are
-found in high-performance computing machines. Benchmarks were run with the
-original t-SNE parameters for 1000 iterations, as some implementations do not
+found in high-performance computing machines. Benchmarks were run for 1000 iteration with the original t-SNE parameters, as some implementations do not
 allow their modification.
@@
 written in Python, which inevitably incurs some runtime overhead compared to
-its pure C++ counterpart. However, Fig.~\ref{fig:benchmarks} shows the
+its C++ counterpart. However, Fig.~\ref{fig:benchmarks} shows the
 runtime of \opentsne\ is comparable to that of \textsf{FIt-SNE} and much faster
@@
 is problematic from a usability standpoint, as many users often lack the
-knowledge to do so. This is especially problematic with Windows users, where
-the operating system does not include a C++ compiler by default, and correctly
-configuring the command line can be an arduous process.
+knowledge to do so. This often occurs with Windows operating system, which does
+not include a C++ compiler, making the correct configuration cumbersome.
 
@@
 points. While the Python virtual machine introduces some performance overhead,
-the runtime is comparable to that of its pure C++ counterpart --
+the runtime is comparable to that of its C++ counterpart --
 \textsf{FIt-SNE}. Finally, \opentsne\ is extensible. Its modular design enables

### 2020-09-30 paper/policar2020-opentsne.tex
Edit methods section

@@
-%% BioMed_Central_Tex_Template_v1.06
-%%                                      %
-%  bmc_article.tex            ver: 1.06 %
-%                                       %
-
-%%IMPORTANT: do not delete the first line of this template
-%%It must be present to enable the BMC Submission system to
-%%recognise this template!!
-
-%%% additional documentclass options:
-%  [doublespacing]
-%  [linenumbers]   - put the line numbers on margins
-
 \documentclass[twocolumn]{bmcart}
@@
 \usepackage{booktabs}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                             %%
-%%  If you wish to display your graphics for   %%
-%%  your own use using includegraphic or       %%
-%%  includegraphics, then comment out the      %%
-%%  following two lines of code.               %%
-%%  NB: These line *must* be included when     %%
-%%  submitting to BMC.                         %%
-%%  All figure files must be submitted as      %%
-%%  separate graphics through the BMC          %%
-%%  submission process, not included in the    %%
-%%  submitted article.                         %%
-%%                                             %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\def\includegraphic{}
-%\def\includegraphics{}
-
-%%% Put your definitions there:
 \startlocaldefs
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Enter the authors here                   %%
-%%                                          %%
-%% Specify information, if available,       %%
-%% in the form:                             %%
-%%   <key>={<id1>,<id2>}                    %%
-%%   <key>=                                 %%
-%% Comment or delete the keys which are     %%
-%% not used. Repeat \author command as much %%
-%% as required.                             %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \author[
@@
 ]{\inits{BZ}\fnm{Bla\v{z}} \snm{Zupan}}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Enter the authors' addresses here        %%
-%%                                          %%
-%% Repeat \address commands as much as      %%
-%% required.                                %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Enter short notes here                   %%
-%%                                          %%
-%% Short notes will be after addresses      %%
-%% on first page.                           %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\begin{artnotes}
-%\note{Sample of title note}     % note to the article
-%\note[id=n1]{Equal contributor} % note, connected to author
-%\end{artnotes}
-
 \end{fmbox}% comment this for two column layout
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% The Abstract begins here                 %%
-%%                                          %%
-%% Please refer to the Instructions for     %%
-%% authors on http://www.biomedcentral.com  %%
-%% and include the section headings         %%
-%% accordingly for your article type.       %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \begin{abstractbox}
-
 \begin{abstract}
-Point-based visualisations of large, multi-dimensional data from molecular
+Point-based visualizations of large, multi-dimensional data from molecular
 biology can reveal meaningful clusters. One of the most popular
-	techniques to construct such visualisations is t-distributed stochastic
-	neighbor embedding (t-SNE), for which a number of extensions have
+techniques to construct such visualizations is $t$-distributed stochastic
+neighbor embedding (t-SNE). Several extensions of t-SNE have
 recently been proposed to address issues of scalability and the quality
-	of the resulting visualisations. We introduce \opentsne, a modular
+of the resulting visualizations. We introduce \opentsne, a modular
 Python library that implements the core t-SNE algorithm and its
@@
 popular implementations, including those from scikit-learn. Unique to
-	\opentsne\ is also the mapping of new data to existing embeddings,
+\opentsne\ is also mapping new data to existing embeddings,
 which can surprisingly assist in solving batch effects.
 \end{abstract}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% The keywords begin here                  %%
-%%                                          %%
-%% Put each keyword in separate \kwd{}.     %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \kwd{dimensionality reduction}
+\kwd{Python library}
 \end{keyword}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% The Main Body begins here                %%
-%%                                          %%
-%% Please refer to the instructions for     %%
-%% authors on:                              %%
-%% http://www.biomedcentral.com/info/authors%%
-%% and include the section headings         %%
-%% accordingly for your article type.       %%
-%%                                          %%
-%% See the Results and Discussion section   %%
-%% for details on how to create sub-sections%%
-%%                                          %%
-%% use \cite{...} to cite references        %%
-%%  \cite{koon} and                         %%
-%%  \cite{oreg,khar,zvai,xjon,schn,pond}    %%
-%%  \nocite{smith,marg,hunn,advi,koha,mouse}%%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%% start of article main body
-% <put your article body there>
-
-%%%%%%%%%%%%%%%%
-%% Background %%
-%%
 \section*{Background}
 
-The abundance of high-dimensional data sets in molecular biology calls for
-techniques for dimensionality reduction, in particular for methods that produce informative data visualizations. Popular approaches include 
-principal component analysis (PCA),
-multidimensional scaling, t-distributed stochastic neighbor embedding
-(t-SNE)~\cite{maaten2008visualizing}, and uniform manifold approximation and
-projections (UMAP)~\cite{2018arXivUMAP}. 
-Among these, t-SNE lately received much attention as it can address high
-volumes of data and reveal meaningful clustering structure. Increased
-resolution and throughput of modern molecular assays has lead to frequent use
-of t-SNE in diverse fields including, but not limited to, single-cell
-transcriptomics(scRNA-seq,~\cite{macosko2015highly,cao2019single,tasic2018shared}),
+The abundance of high-dimensional data sets in molecular biology calls for dimensionality reduction techniques, particularly for methods that produce informative data visualizations.
+Popular approaches include
+principal component analysis (PCA), multidimensional scaling, t-distributed stochastic neighbor embedding (t-SNE)~\cite{maaten2008visualizing}, and uniform manifold approximation and projections (UMAP)~\cite{2018arXivUMAP}.
+Among these, t-SNE lately received much attention as it can address high volumes of data and reveal a meaningful clustering structure.
+Increased
+resolution and throughput of modern molecular assays has lead to the frequent use of t-SNE in diverse fields including, but not limited to, single-cell transcriptomics(scRNA-seq,~\cite{macosko2015highly,cao2019single,tasic2018shared}),
 human genetics~\cite{hirata2019genetic}, metagenomic
-assembly~\cite{beaulaurier2018metagenomic}, spatial organization of microbial
-communities~\cite{sheth2019spatial} and
-metabolomics~\cite{tkachev2019differences}.
-
-Reports on single-cell gene expression data, our running example,
-often start with an overview of the cell landscape, where t-SNE embeds
-high-dimensional expression profiles into a two-dimensional
-space. Figs.~\ref{fig:macosko}.a and \ref{fig:macosko}.b are examples of two such
-embeddings.
+assembly~\cite{beaulaurier2018metagenomic}, the spatial organization of microbial communities~\cite{sheth2019spatial} and metabolomics~\cite{tkachev2019differences}.
+Reports on single-cell gene expression data, our running example, often start with an overview of the cell landscape, where t-SNE embeds high-dimensional expression profiles into a two-dimensional
+space. Figs.~\ref{fig:macosko}.a and \ref{fig:macosko}.b show two such embeddings.
 
@@
   \includegraphics[width=\textwidth]{macosko2015}
-  \caption{\label{fig:macosko}The figure shows the progression of t-SNE
-	embeddings following theoretical advances, which are all included in
+\caption{\label{fig:macosko}
+Three t-SNE embeddings demonstrate theoretical advances and corresponding implementations in
 \opentsne. The data in \textbf{(a)} and \textbf{(b)} represent 44,808
-	single-cell gene-expression profiles of mouse retinal cells from the
-	study by Macosko \textit{et al.}~\cite{macosko2015highly}. Similarly,
-	the data in \textbf{(c)} contains 27,499 expression profiles from mouse
+single-cell gene-expression profiles of mouse retinal cells from Macosko \textit{et al.}~\cite{macosko2015highly}. The data in \textbf{(c)} contains 27,499 expression profiles from mouse
 retinal cells from Shekhar \textit{et
@@
 embedding following the parameter choices from the original publication
-	from Maaten \& Hinton~\cite{maaten2008visualizing}. The visualization
+by Maaten \& Hinton~\cite{maaten2008visualizing}. The visualization
 shows no preservation of the global organization of clusters,
 resulting from random initialization and an affinity model focused on
-	preserving local neighborhoods. \textbf{(b)} shows a modern t-SNE
+preserving local neighborhoods. \textbf{(b)} A modern t-SNE
 embedding, utilizing the latest theoretical advances and practical
-	recommendations. The embedding is constructed using a multi-scale
+recommendations constructed using a multi-scale
 affinity model, preserving both short-range and long-range interactions
-	between data points, and is initialized such that the global layout is
+between data points and initialized so that the global layout is
 meaningful. Unlike in \textbf{(a)}, the green and blue clusters
@@
 localized to the same regions of the space, indicating a higher level
-	of similarity than to other cell types. \textbf{(c)} shows how existing
-	t-SNE reference atlases can be used to embed new samples into existing
+of similarity than to other cell types. Embedding in \textbf{(c)} shows how existing
+t-SNE reference atlases can be used to place new samples into existing
 embeddings. The positions of new data points correspond to cell types
-	from the reference atlas.}
+from the reference atlas.
+}
 \end{figure*}
@@
 to improve global positioning, including estimating similarities with a mixture
-of Gaussian kernels~\cite{kobak2019art}. In our previous work, we presented a
-principled approach for embedding new samples into existing
+of Gaussian kernels~\cite{kobak2019art}. In our previous work, we presented a principled approach for embedding new samples into existing
 visualizations~\cite{policar2019embedding}.
@@
 We introduce \opentsne, a comprehensive Python library that implements t-SNE
-and all its recently proposed extensions. Most notable among these are
+and all its recently proposed extensions, including:
 \begin{enumerate}
@@
 	of millions of data points,
-\item the addition of new samples into existing
+\item the addition of new data samples into a fixed existing
 	embeddings~\cite{policar2019embedding},
-\item better initialization schemes~\cite{kobak2019umap} leading to more
+\item improved initialization schemes~\cite{kobak2019umap} leading to more
 	globally consistent layouts,
@@
 \item improved defaults for learning rate and number of
-	iterations~\cite{belkina2019automated} which produce embeddings at a low
+	iterations~\cite{belkina2019automated} which produce embeddings at a lower
 	computational cost.
@@
 
-The library is fully compatible with the Python data-science ecosystem (e.g.,
-\textsf{numpy}, \textsf{scikit-learn}, \textsf{scanpy}) providing a familiar
-and intuitive API to new users. Its modular design encourages
-extensibility and experimentation with various settings and changes in the
-analysis pipeline.
-
-\opentsne\ is open-source and has become adopted by the wider community,
-receeiving over 600 Github stars and averaging thousands of weekly downloads.
-The source code is publicly available on Github at
-\url{https://github.com/pavlin-policar/openTSNE}, and can be installed from
-\textsf{PyPI} and \textsf{conda-forge} -- the two most widely adopted Python
+The \opentsne\ is compatible with the Python data science ecosystem and libraries that include
+\textsf{numpy}, \textsf{scikit-learn}, \textsf{scanpy}), thus providing a familiar and intuitive application program interface to the new users.
+Its modular design encourages extensibility and experimentation with various settings and changes in the analysis pipeline.
+We have released \opentsne\ in an open-source to support adaptations by the wider community.
+At the time of the writing, the library averages several thousands of weekly downloads and has received almost 700 GitHub stars, an appreciation score gained by only $0.3\%$ of public Python repositories.
+The source code is available at
+\url{https://github.com/pavlin-policar/openTSNE} and supports installations from
+\textsf{PyPI} and \textsf{conda-forge}, the two most widely adopted Python
 package managers.
 
-Accessibility is one of the core design principles of \opentsne. Our aim is to
-make the latest theoretical advances in t-SNE as accessible as possible to as
-many users as possible. As such, our API is based on the familiar
-\textsf{scikit-learn} interface~\cite{sklearn_api}, which enables users a
-smooth transition from existing Python libraries. For example,
-Fig.~\ref{fig:tasic} depicts four different t-SNE visualizations generated by
-\opentsne, each based on recent theoretical advances in the t-SNE algorithm. The code used
-to generate these embeddings is shown below.
+Accessibility of the latest theoretical advances in t-SNE is one of the core design principles of \opentsne.
+These advances are accessible through a programming interface of the library, where we closely follow the style of the interface as implemented in
+\textsf{scikit-learn}~\cite{sklearn_api}, a popular library for machine learning.
+The following snippet exemplifies this interface and codes for four visualizations in Fig.~\ref{fig:tasic}, demonstrating the effect of a particular recent theoretical advance in the t-SNE algorithm.
 
@@
 
-\noindent The code snippet takes as input a \textsf{numpy} array or
-\textsf{scipy} sparse matrix \texttt{X} and produces four
-\texttt{TSNEEmbedding} objects. These may subsequently be subject to further
-optimization under different parameter settings or may be used to embed new
-samples into the embedding landscape. Additional parameters are omitted for
-clarity but can be found in accompanying notebooks publicly available at
-\url{https://github.com/pavlin-policar/opentsne-paper}.
+\noindent The code snippet considers a \textsf{numpy} array or
+\textsf{scipy} sparse matrix \texttt{X} and creates four
+\texttt{TSNEEmbedding} objects.
+These objects encode t-SNE embeddings that may subsequently be optimized under different parameter settings or used to embed new samples into the embedding landscape.
+Function calls may include additional parameters not shown in the example code but explained in accompanying notebooks available at \url{https://github.com/pavlin-policar/opentsne-paper}.
 
@@
 
-Dimensionality reduction techniques implicitly assume that high-dimensional
-data lies on a lower-dimensional manifold, which can accurately be captured by
-a smaller number of dimensions. However, there is no evidence that
-every data set can accurately be described by using only two dimensions, and any such embedding will therefore inevitably lead to loss of
-information. It is thus beneficial to examine multiple embeddings, each
-of which provides a different perspective into topology and other
-data characteristics.
+Dimensionality reduction techniques implicitly assume that high-dimensional data lies on a lower-dimensional manifold, which can accurately be captured by a smaller number of dimensions.
+However, there is no evidence that every data set can accurately be described by using only two dimensions, and any such embedding will inevitably lead to a loss of information.
+Thus, it is beneficial to examine multiple embeddings, each of which provides a different perspective on topology and other data characteristics.
 
@@
 
-We illustrate this point by generating four embeddings of the same data on
-single-cell gene expression in mouse brain~\cite{tasic2018shared}.  First,
-Fig.~\ref{fig:tasic}.a shows an embedding using default t-SNE parameters. While
-different clusters of excitatory and inhibitory neurons appear close to one
-another, all clusters appear equidistant from their neighbors, and the overall
-relations between groups are not obvious. On the other hand,
-Fig.~\ref{fig:tasic}.b focuses on preserving larger neighborhoods of points,
-resulting in a more globally consistent layout where relations between clusters
-become more apparent. In this embedding, it is evident from the increased
-white space between groups that there is one large class of excitatory neurons,
-and two related classes of inhibitory neurons. Unfortunately, focusing on
-preserving large neighborhoods often results in smaller clusters being absorbed
-into larger ones. Alternatively, Fig.~\ref{fig:tasic}.c uses multi-scale
-similarity kernels which aim to preserve both the global organization of
-clusters as well as prevent smaller clusters from being absorbed into larger
-ones. On the other hand, Fig.~\ref{fig:tasic}.d shows use the same parameter
-values as Fig.~\ref{fig:tasic}.a, but at a finer level of resolution. This
-reveals that many of the clusters are composed of other subgroups and may be
-clustered further into cell subtypes.
+We illustrate this point by generating four different embeddings of the data on single-cell gene expression in mouse brain~\cite{tasic2018shared}.
+Fig.~\ref{fig:tasic}.a shows an embedding using default t-SNE parameters.
+While different clusters of excitatory and inhibitory neurons appear close to one another, all clusters appear equidistant from their neighbors, and the overall relations between groups are not obvious.
+Embedding in Fig.~\ref{fig:tasic}.b focuses on preserving larger neighborhoods of points, resulting in a more globally consistent layout where relations between clusters become more apparent.
+Here, it is evident from the increased white space between groups that there is one large class of excitatory neurons and two related classes of inhibitory neurons.
+Unfortunately, focusing on preserving large neighborhoods leads to the absorption of smaller clusters by the larger ones.
+Alternatively, Fig.~\ref{fig:tasic}.c uses multi-scale similarity kernels that aim to preserve both the global organization of clusters and prevent the cluster absorption.
+We constructed the embeddings from Fig.~\ref{fig:tasic}.d with the settings used for Fig.~\ref{fig:tasic}.a, but at a finer level of resolution.
+The distinction between these two depictions reveals that many clusters are composed of subgroups representing different cell subtypes.
 
-When working with data containing millions of data points, standard t-SNE
-embeddings often become unwieldy -- cluster boundaries become blurred, large
-clusters absorb smaller ones, and relationships between clusters become
-increasingly difficult to interpret. Fig.~\ref{fig:cao}.a contains expression
-profiles of over two million single cells captured at different time points in
-mouse development. The embedding indicates numerous clusters with some
-transition between time points as indicated by the color-coding, but this is
-largely difficult to interpret. Kobak \& Berens observed that increasing
-attractive forces between similar data points controlled via the
-\textit{exaggeration} parameter leads to more compact clusters, and
-subsequently more informative visualizations~\cite{kobak2019art}. For instance,
-\ref{fig:cao}.b doubles the default exaggeration  which begins to reveal the
-overall structure of the data. Doubling the exaggeration again in
-\ref{fig:cao}.c, we can now easily observe that the data is comprised of two
-main groups of cells and eight somewhat smaller clusters. The visualization
-also reveals a number of tiny clusters, possibly corresponding to
+When dealing with data containing millions of data points, standard t-SNE embeddings often become unwieldy -- cluster boundaries are blurred, large clusters absorb smaller ones, and relationships between clusters become increasingly difficult to interpret.
+We constructed Fig.~\ref{fig:cao}.a from the data containing expression profiles of over two million single cells captured at different time points in mouse development.
+The embedding indicates numerous clusters with transitions between time points, as marked by the color-coding, that are difficult to interpret.
+Kobak \& Berens observed that increasing attractive forces between similar data points controlled via the \textit{exaggeration} parameter leads to more compact clusters, and subsequently, more informative visualizations~\cite{kobak2019art}.
+For instance, \ref{fig:cao}.b doubles the default exaggeration, which uncovers some of the data's overall structure.
+Further doubling the exaggeration in \ref{fig:cao}.c allows us to observe that the data is comprised of two main groups of cells and eight somewhat smaller clusters.
+The visualization also reveals several tiny clusters, possibly corresponding to
 rare cell types.
@@
   \includegraphics[width=\textwidth]{cao2019}
-  \caption{\label{fig:cao}Increasing the exaggeration parameter produces more
-  	compact clusters, which highlight the global organization of the clusters
-  	and emphasizes continuities between cell states. The data set from Cao \textit{et
-	al.}~\cite{cao2019single} contains expression profiles from 2,058,652
-	single cells. The data were collected from mice embryos at different
-	developmental time-points at daily intervals after 9.5 to 13.5 days.
-	\textbf{(a)} shows the data is comprised of two main components -- the
-	neural tube and mesenchymal cells -- as well as several other smaller
-	clusters. Additionally, the colors indicate developmental progression
-	with red indicating least-developed cells and blue indicating most
-	developed cells. The overall developmental trajectory is most apparent
-	with higher levels of exaggeration, showing red cells slowly
-	transitioning into blue cells. Progressively easing the exaggeration
-	factor uncovers finer clusters within the larger groups as shown in
-	\textbf{(b)} with exaggeration 2 and subsequently in \textbf{(c)},
-	where we show the standard t-SNE with no exaggeration. 32,011 putative
-	doublets are excluded from the visualizations.}
+  \caption{\label{fig:cao}
+Increasing the exaggeration parameter leads to compact clusters, highlighting the data's global organization and emphasizing the continuity between cell states.
+The data set from Cao \textit{et al.}~\cite{cao2019single} contains expression profiles from 2,058,652 single cells.
+The data were collected from mice embryos at different development stages at daily intervals after 9.5 to 13.5 days.
+\textbf{(a)} reveals that the data is comprised of two main components -- the neural tube and mesenchymal cells -- as well as several other smaller clusters.
+Additionally, the colors indicate developmental progression with red indicating least-developed cells and blue indicating most developed cells.
+The overall developmental trajectory is most apparent with higher exaggeration levels, showing red cells slowly transitioning into blue cells.
+Progressively easing the exaggeration factor uncovers finer clusters within the larger groups, as shown in
+\textbf{(b)} with exaggeration of two and subsequently in \textbf{(c)}, where we show the standard t-SNE with no exaggeration. 32,011 putative doublets are excluded from the visualizations.
+}
 \end{figure*}
 
-Exaggeration can highlight transitions between cell-states in
-developmental studies. Standard t-SNE often produces embeddings with
-clearly defined, discrete clusters. We have shown that we can adjust the level
-of granularity and resolution of the clusters with several parameters in Fig.~\ref{fig:tasic}. However,
-discrete clusters are often undesired in developmental studies where cells are assumed to follow a continuous trajectory between cell states. To this end, other
-embedding techniques such as UMAP and ForceAtlas2~\cite{jacomy2014forceatlas2} tend to better capture the continuity between cell
-states. Recently, Kobak \textit{et al.}~\cite{TODO} have shown that embeddings
-produced by t-SNE with exaggeration values of 4 and $\sim30$ produce embeddings
-which are markedly similar to UMAP and ForceAtlas2, respectively. For example,
-in Fig.~\ref{fig:cao}.a, the developmental trajectory between different time
-points is difficult to observe due to a large number of sprawled out clusters.
-On the other hand, the development becomes easier to follow when the
-exaggeration factor is increased from $1$ to $4$ (Fig.~\ref{fig:cao}.b-c).
+Exaggeration can highlight transitions between cell states in developmental studies.
+Standard t-SNE often produces embeddings with clearly defined, discrete clusters.
+We can adjust the level of granularity and resolution of the clusters with several parameters in Fig.~\ref{fig:tasic}.
+However, discrete clusters are often undesired in developmental studies where cells' state is changed continuously.
+To this end, other embedding techniques such as UMAP and ForceAtlas2~\cite{jacomy2014forceatlas2} tend to capture the continuity between cell states better.
+Recently, B{\"o}hm \textit{et al.}~\cite{bohm2020unifying} have shown that embeddings produced by t-SNE with exaggeration values of 4 and $\sim30$ construct embeddings which are markedly similar to UMAP and ForceAtlas2, respectively.
+For example, in Fig.~\ref{fig:cao}.a, the developmental trajectory between different time points is difficult to observe due to many sprawled out clusters.
+On the other hand, it is easier to trace the development when we increase the exaggeration factor from $1$ to $4$ (Fig.~\ref{fig:cao}.b-c).
 
@@
 Unlike other popular dimensionality reduction methods such as principal
-component analysis or autoencoders, t-SNE is a non-parametric method and does
-not define an explicit mapping to the embedding space. Therefore embeddings of
-new data points need to be found through the use of optimization
-techniques~\cite{policar2019embedding}. \opentsne\ is currently the only
-publicly available library allowing users to add new samples to existing
-embeddings.
+component analysis or autoencoders, t-SNE is a non-parametric method and does not define an explicit mapping to the embedding space.
+Therefore embeddings of new data points need to be found through optimization~\cite{policar2019embedding}.
+\opentsne\ is currently the only publicly available library allowing users to add new samples to existing embeddings.
 
-Figs.~\ref{fig:macosko}.b and \ref{fig:macosko}.c demonstrate how we can use a
-previously labeled single-cell data set and embed cells from a separate
-experiment into the reference landscape. The reference data taken from Macosko~\textit{et al.}~\cite{macosko2015highly} contains gene expression profiles from
-mouse retinal cells. By embedding the samples from a similar experiment on bipolar retinal cells by Shekhar~\textit{et al.}~\cite{shekhar2016comprehensive}, 
-we can correctly map to the bipolar cell clusters in the reference.
+Figs.~\ref{fig:macosko}.b and \ref{fig:macosko}.c demonstrate how we can use a previously labeled single-cell data set and embed cells from a separate experiment into the reference landscape.
+The reference data from Macosko~\textit{et al.}~\cite{macosko2015highly} contains gene expression profiles from mouse retinal cells.
+By embedding the samples from a similar experiment on bipolar retinal cells by Shekhar~\textit{et al.}~\cite{shekhar2016comprehensive}, we can correctly map the bipolar cell clusters in the reference.
 
-Embedding single cells into existing reference atlases can also be useful for
-cell type classification in cases where cell identities are not known in
-advance. For instance, in Fig.~\ref{fig:transform}, we construct a reference
-embedding using labeled data from Hochgerner~\textit{et
-al.}~\cite{hochgerner2018conserved} containing gene-expression profiles of
-cells from the mouse brain. The authors assign a cell type to each cell, and
-we can verify their accuracy by visualizing well-established gene markers for the
-major cell types. We then embed cells from Harris~\textit{et
-al.}~\cite{harris2018classes}, where labels are provided only for neuronal
-cells. We quickly identify other non-neuronal cell types, including
-oligodendrocytes and astrocytes. We use marker genes to validate that new cells
-were correctly mapped into the reference landscape.
+Embedding single cells into existing reference atlases can also be useful for cell-type classification in cases of unknown cell identities.
+For instance, in Fig.~\ref{fig:transform}, we construct a reference embedding using labeled data from Hochgerner~\textit{et al.}~\cite{hochgerner2018conserved} containing gene-expression profiles of cells from the mouse brain.
+The authors assign a type to each cell.
+We can verify their classification accuracy by visualizing the expression of well-established gene markers for the major cell types.
+We then embed cells from Harris~\textit{et al.}~\cite{harris2018classes} into a constructed atlas.
+In Harris~\textit{et al.}~, labels are provided only for neuronal cells.
+In the resulting mapping, we can quickly identify other non-neuronal cell types, including oligodendrocytes and astrocytes.
+We can further use marker genes to validate that the mapping in the reference landscape was correct.
 
@@
   \includegraphics[width=\textwidth]{transform_hochgerner}
-  \caption{\label{fig:transform}openTSNE enables the user to embed new samples
-	into an existing, reference embedding. We construct a t-SNE embedding
-	for the data from Hochgerner \textit{et
-	al.}~\cite{hochgerner2018conserved} containing 24,185 developing,
-	single cells from the mouse hippocampus. The data contains different
-	types of neurons, supporting glia, and other vascular cells (upper
-	left). Cell labels are taken from the original publications. The
-	colors correspond to Fig.~\ref{fig:tasic}. We then embed new,
-	hippocampal cells collected by Harris \textit{et
-	al.}~\cite{harris2018classes} using the previous embedding as a
-	reference. In their study, Harris \textit{et al.} collected 6,971
-	single-cells and focused on identifying different types of inhibitory
-	neurons. However, almost half of the collected cells are not neurons
-	and were left uncharacterized. Inspecting where these cells are
-	embedded in the reference embedding (bottom left) reveals that in
-	addition to inhibitory neurons, the data contains several supporting
-	glial cells as well as a small population of endothelial cells. We can
-	verify the accuracy of our approach by inspecting marker genes for the
-	major cell types in the reference (top row) and embedded samples
-	(bottom row).}
+  \caption{\label{fig:transform}
+\openTSNE supports embedding new samples into an existing reference t-SNE landscape.
+For the series of visualizations shown in this figure, we first construct a t-SNE embedding for the data from Hochgerner \textit{et al.}~\cite{hochgerner2018conserved} containing 24,185 developing, single cells from the mouse hippocampus.
+The data contains gene expression in different neurons, supporting glia, and other vascular cells (upper left).
+Data points representing cells are colored according to cell-types assigned in the original publication; see the legend from Fig.~\ref{fig:tasic} to map colors to cell-type.
+We then embed new, hippocampal cells collected in a study by Harris \textit{et al.}~\cite{harris2018classes} using the embedding of Hochgerner \textit{et al.} data as a reference.
+In their study, Harris \textit{et al.} collected 6,971 single-cells and focused on identifying different types of inhibitory neurons.
+However, almost half of the collected cells are not neurons and were left uncharacterized.
+Inspecting the embeddings of these cells in the reference embedding (bottom left) reveals that in addition to inhibitory neurons, the data contains several supporting glial cells as well as a small population of endothelial cells.
+We can verify our approach's accuracy by inspecting marker genes for the major cell types in the reference (top row) and embedded samples (bottom row).
+}
 \end{figure*}
 
-These two examples demonstrate how \opentsne\ can be used to quickly gain
-insight into newly-sequenced, single-cell data sets by utilizing existing cell
-atlases. This approach is general and not limited to single-cell gene
-expression, but can be used with any tabular data set regardless of field.
+The examples presented above demonstrate how to use \opentsne\ to quickly gain insight into a newly-sequenced, single-cell data sets by utilizing existing cell atlases.
+The approach is general and not limited to single-cell gene expression, and one can, in principle, apply it to any tabular data set regardless of field.
 
@@
 
-Versatility is another of  \opentsne s core design principles. Kobak \& Berens
-recently introduced several recommendations and tricks to obtain better and
-more meaningful t-SNE visualizations~\cite{kobak2019art}. These include the use
-of multi-scale similarity kernels, perplexity annealing, and increasing
-exaggeration when working with larger data sets. \opentsne\ provides a flexible
-API, where these improvements can be incorporated in just a few lines of code.
-Furthermore, \opentsne\ supports custom affinity models, enabling users to
-construct t-SNE embeddings on non-tabular data, e.g. relational data, as the
-only affinity-model requirement is a notion of similarity between data points,
-i.e. the distance matrix.  Finally, our comprehensive callback system can be
-utilized to monitor and adapt different stages of the optimization phase and
-has been used to construct visually appealing animations of the t-SNE
-optimization process.
+Versatility, the ability to use and combine different optimization approaches to construct the embedding spaces, is another of \opentsne's core design principles.
+Kobak \& Berens recently introduced several recommendations and tricks to obtain better and more meaningful t-SNE visualizations~\cite{kobak2019art}.
+These include multi-scale similarity kernels, perplexity annealing, and increasing exaggeration when working with more massive data sets.
+\opentsne\ provides a flexible program interface to incorporate these improvements in just a few code lines.
+Furthermore, \opentsne\ supports custom affinity models, enabling users to construct t-SNE embeddings on non-tabular relational data: the only requirement imposed by the affinity-model is the availability of similarity between data points, that is, the distance matrix.
+Finally, the \opentsne's comprehensive callback system can be utilized to monitor and adapt different stages of the optimization phase and has been used to construct visually appealing animations of the t-SNE optimization process.
 
@@
 
-One of t-SNE's common criticisms is limited scalability to large data sets
-containing up to millions of data points~\cite{becht2019dimensionality}. These
-criticisms are largely based on two factors. Until quite recently, most popular
-implementations of t-SNE were based on the Barnes-Hut approximation scheme
-developed by van der Maaten in 2014~\cite{van2014accelerating} with asymptotic
-complexity $\mathcal{O}(N \log N)$. More importantly, this criticism was
-exacerbated further by the fact that the most widely-used implementation of
-t-SNE came from \textsf{scikit-learn}, which exhibits long runtimes
-performance compared to its C++ counterpart -- \textsf{MulticoreTSNE}. The
-multi-threaded \textsf{MulticoreTSNE} implementation~\cite{Ulyanov2016} is
-perfectly capable of creating t-SNE embeddings of millions of data points in a
-matter of hours on widely-accessible, consumer-grade processors
-(Fig.~\ref{fig:benchmarks}). However, recently, Linderman \textit{et al.}
-developed a new approximation scheme -- FIt-SNE -- which further reduces the
-asymptotic time complexity to $\mathcal{O}(N)$, enabling the embedding of large
-data sets in a matter of minutes.
+One of the t-SNE's common criticisms is limited scalability to large data sets containing, for instance, millions of data points~\cite{becht2019dimensionality}.
+The culprit for slow response time stems from a particular optimization procedure and its specific implementation in the popular Python library.
+Namely, until quite recently, most popular implementations of t-SNE were based on the Barnes-Hut approximation scheme developed by van der Maaten in 2014~\cite{van2014accelerating} with asymptotic complexity $\mathcal{O}(N \log N)$, where $N$ is the number of data items ({\em e.g.} cells).
+The most widely-used implementation of t-SNE came from \textsf{scikit-learn}, which exhibits long runtimes when compared to its C++ counterpart -- \textsf{MulticoreTSNE}.
+The multi-threaded \textsf{MulticoreTSNE} implementation~\cite{Ulyanov2016} can construct t-SNE embeddings of millions of data points in a matter of hours on widely-accessible, consumer-grade processors (Fig.~\ref{fig:benchmarks}).
+However, recently, Linderman \textit{et al.} developed a new approximation scheme -- FIt-SNE -- which further reduces the asymptotic time complexity to $\mathcal{O}(N)$.
+We have included this scheme in \opentsne\, enabling the embedding of large data sets in a matter of minutes.
 
-Fig.~\ref{fig:benchmarks} benchmarks four Python t-SNE implementations,
-including those from \textsf{scikit-learn} (v0.23.1), \textsf{MulticoreTSNE}
-(v0.1), \textsf{FIt-SNE} (v1.1.0), and \opentsne\ (v0.4.3). We perform
-benchmarks on two processors, one representing usage on a personal computer and
-one on a high-performance computing platform. The Intel(R) Core i7 is commonly
-found in consumer-grade laptop computers, while Intel(R) Xeon(R) processors are
-found in high-performance computing machines. Benchmarks were run for 1000 iteration with the original t-SNE parameters, as some implementations do not
-allow their modification.
+Fig.~\ref{fig:benchmarks} benchmarks four Python t-SNE implementations, including those from \textsf{scikit-learn} (v0.23.1), \textsf{MulticoreTSNE} (v0.1), \textsf{FIt-SNE} (v1.1.0), and our \opentsne\ (v0.4.3).
+We perform benchmarks on two computational platforms, one representing the usage on a personal computer and the other utility of these libraries on a high-performance computing platform.
+The Intel(R) Core i7 is commonly found in consumer-grade laptop computers, while Intel(R) Xeon(R) processors appear in high-performance computing machines.
+Benchmarks were run for $1,000$ iterations with the original t-SNE parameters, as some implementations do not allow their modification.
 
@@
   \includegraphics[width=\textwidth]{benchmarks}
-  \caption{\label{fig:benchmarks}We benchmark openTSNE (v0.4.3) against three
-	popular open-source implementations from
-	scikit-learn~\cite{pedregosa2011scikit} (v0.23.1),
-	MulticoreTSNE~\cite{Ulyanov2016} (v0.1), and
-	FIt-SNE~\cite{linderman2019fast} (v1.1.0). Benchmarks were run on a
-	consumer-grade Intel Core i7-7700HQ processor found in laptop
-	computers, and on a server-grade Intel Xeon E5-2650. To generate
-	benchmark data sets of different sizes, we subsampled data from the 10X
-	Genomics 1.3 million mouse brain data set five times, resulting in five
-	different data sets for each size. In total, each implementation was
-	run on 30 different data sets. Notice that openTSNE scales similarly to
-	FIt-SNE, as they both use the same interpolation-based approximation
-	scheme, while scikit-learn and MulticoreTSNE utilize the Barnes-Hut
-	approximation.}
+  \caption{\label{fig:benchmarks}
+We benchmark openTSNE (v0.4.3) against three popular open-source implementations from scikit-learn~\cite{pedregosa2011scikit} (v0.23.1), MulticoreTSNE~\cite{Ulyanov2016} (v0.1), and FIt-SNE~\cite{linderman2019fast} (v1.1.0).
+Experiments were run on a consumer-grade Intel Core i7-7700HQ processor found in laptop computers, and on a server-grade Intel Xeon E5-2650.
+To generate benchmark data sets of different sizes, we subsampled data from the 10X Genomics 1.3 million mouse brain data set five times, resulting in five different data sets for each size.
+In total, we run each implementation on 30 different data sets.
+Notice that openTSNE scales similarly to FIt-SNE, as they both use the same interpolation-based approximation scheme, while scikit-learn and MulticoreTSNE utilize the Barnes-Hut approximation.
+}
 \end{figure*}
 
-Fig.~\ref{fig:benchmarks} demonstrates that both \textsf{FIt-SNE} and
-\opentsne\ exhibit better scaling than their Barnes-Hut
-counterparts -- \textsf{scikit-learn} and \textsf{MulticoreTSNE}. Notice that
-even though \textsf{FIt-SNE} and \opentsne\ both implement the same
-approximation scheme, \opentsne\ tends to be slightly slower. \opentsne\ is
-written in Python, which inevitably incurs some runtime overhead compared to
-its C++ counterpart. However, Fig.~\ref{fig:benchmarks} shows the
-runtime of \opentsne\ is comparable to that of \textsf{FIt-SNE} and much faster
-than its Barnes-Hut competitors. Additionally, modern computer processors
-generally contain multiple cores, allowing us to make use of multi-threading.
-This further reduces the gap between \opentsne\ and \textsf{FIt-SNE}. On the
-server-grade processor, \opentsne\ is slightly faster than its pure C++
-counterpart when utilizing multiple cores. We speculate that this is could be because \opentsne\ uses
-\textsf{numpy} for most linear algebra operations, which may make better use of
-the Intel Math Kernel Library (MKL), which is aggressively optimized on
-Intel(R) Xeon(R) processors.
+Benchmark results (Fig.~\ref{fig:benchmarks}) confirm that both \textsf{FIt-SNE} and \opentsne\ scale better than their Barnes-Hut counterparts -- \textsf{scikit-learn} and \textsf{MulticoreTSNE}.
+While \opentsne\'s implementation uses Python that incurs some runtime overhead compared to its C++ counterpart \textsf{FIt-SNE}, the two libraries' speeds are surprisingly comparable.
+Modern computer processors contain multiple cores, allowing us to use multi-threading, which further reduces the gap between \opentsne\ and \textsf{FIt-SNE}.
+On the server-grade processor, \opentsne\ is slightly faster than its pure C++ counterpart when utilizing multiple cores.
+\opentsne\ uses \textsf{numpy} for most linear algebra operations, which may make better use of the Intel Math Kernel Library (MKL), which is aggressively optimized on Intel(R) Xeon(R) processors.
 
-Additionally, \opentsne\ provides a flexible API, which allows splitting up the
-embedding-construction process into several part, caching slow operations. This
-enables users to quickly experiment with different parameter settings and
-iterate on their final visualizations.
+\opentsne\ provides a flexible API, allowing splitting up the embedding-construction process into several parts, caching slow operations.
+Running t-SNE optimization in stages enables users to quickly experiment with different parameter settings and iterate on their final visualizations.
 
@@
 
-Widespread adoption of novel techniques is almost universally accompanied by
-easy-to-use and easy-to-install software. While the t-SNE implementation from
-\textsf{scikit-learn} fits this requirement, its implementation is
-unfortunately very slow, making it unsuitable for its application to large data
-sets. Other C++ implementations such as \textsf{MulticoreTSNE} and
-\textsf{FIt-SNE} exhibit better scaling in larger data sets, but do not provide
-precompiled binaries, requiring users to compile the software themselves. This
-is problematic from a usability standpoint, as many users often lack the
-knowledge to do so. This often occurs with Windows operating system, which does
-not include a C++ compiler, making the correct configuration cumbersome.
+Intuitive access and simple installation procedures almost universally correlate with the widespread adoption of novel computational techniques.
+While the t-SNE from \textsf{scikit-learn} fits this requirement, its implementation is prohibitively slow for even moderately-sized data sets that span tens of thousands of data records.
+Other C++ implementations such as \textsf{MulticoreTSNE} and \textsf{FIt-SNE} exhibit better scaling in more massive data sets, but do not provide precompiled binaries and require users to compile the software themselves.
+This problem is critical, for instance, for users of Windows, where the C++ compiler does not come with the system, making the correct configuration of current t-SNE implementations cumbersome.
 
-\opentsne\ is a Python package that focuses on accessibility. We provide
-precompiled binaries for all major Python versions on all major platforms,
-making the installation process as seamless as possible. \opentsne\ can be
-installed through the Python Package Index (PyPI) or through conda on the
-conda-forge channel. \opentsne\ is designed to be familiar to Python users and
-follows the \textsf{scikit-learn} API~\cite{sklearn_api}, which is well
-established in the Python data science ecosystem. \opentsne\ implements
-multi-threaded versions of both the Barnes-Hut and FIt-SNE approximation
-schemes, allowing it to be used on data sets containing millions of data
-points. While the Python virtual machine introduces some performance overhead,
-the runtime is comparable to that of its C++ counterpart --
-\textsf{FIt-SNE}. Finally, \opentsne\ is extensible. Its modular design enables
-researchers to quickly experiment with different settings and easily
-incorporate their own components into the software. We provide a full feature
-list and comparison to other popular t-SNE implementations in
-Table~\ref{tab:features}.
+We have designed \opentsne\ to make it accessible to a broader audience.
+We provide precompiled binaries for all major Python versions on all major platforms, making the installation process as seamless as possible.
+One can install \opentsne\ through the Python Package Index (PyPI) or conda on the conda-forge channel.
+\opentsne\ interface follows that of \textsf{scikit-learn}, which is well established in the Python data science ecosystem.
+\opentsne\ implements multi-threaded versions of both the Barnes-Hut and FIt-SNE approximation schemes, enabling it for data sets containing millions of data points.
+While the Python virtual machine introduces some performance overhead, the runtime is comparable to its C++ counterpart -- \textsf{FIt-SNE}.
+Finally, \opentsne\ is extensible.
+Its modular design enables researchers to quickly experiment with different settings and easily incorporate their components into the software.
+We provide a full feature list and comparison to other popular t-SNE implementations in Table~\ref{tab:features}.
 
 \begin{table}
-\caption{\label{tab:features}We compare the feature list of \opentsne to three
-popular open-source implementations from scikit-learn~(v0.23.1), MulticoreTSNE
-(v0.1), and FIt-SNE (v1.1.0). The first section corresponds to packaging and 
-distribution. A properly packaged library is easily accessible to users and can
-easily be included as a dependency in other software packages. The second
-section lists the two approximation schemes. The FIt-SNE approximation scheme
-is required for t-SNE to scale up to millions of data points. The final section
-provides a list of extensions and improvements to the standard t-SNE algorithm,
-many of which can produce markedly better visualizations than under default
-parameter settings.}
+\caption{\label{tab:features}
+We compare features of \opentsne to three popular open-source implementations from scikit-learn~(v0.23.1), MulticoreTSNE (v0.1), and FIt-SNE (v1.1.0).
+The first section of the comparison addresses packaging and distribution.
+A properly packaged library is easily accessible to users, and developers should easily include it in dependency lists of other software packages.
+The second section of the comparison lists the two existing t-SNE approximation schemes.
+The FIt-SNE approximation scheme is required for t-SNE to scale up to millions of data points.
+The final section provides a list of extensions and improvements to the standard t-SNE algorithm, many of which can produce markedly better visualizations.
+}
 
@@
 primarily used as a visualization tool, $d$ is typically set to two. The
-similarity between two data points in the high-dimensional space is defined as:
+similarity between two data points in the high-dimensional space is defined as
 
@@
 density of the data in the multi-dimensional space. Perplexity is sometimes
-considered the continuous analog to the $k$ nearest neighbors to which
-distances will be preserved.
+interpreted as the continuous analog to the number of nearest neighbors to which
+distances will attempt to be preserved.
 
@@
 embedding space is defined using the $t$-distribution with a single degree of
-freedom. This is also known as the Cauchy kernel:
+freedom (Cauchy kernel)
 
@@
 
-Finally, the Kullback-Leibler (KL) divergence is used as a measure of agreement
+The Kullback-Leibler (KL) divergence is used as a measure of agreement
 between distributions $\mathbf{P}$ and $\mathbf{Q}$
@@
 
-\noindent The t-SNE objective is to find embeddings $\mathbf{Y}$ such that the
+\noindent The objective is to find embeddings $\mathbf{Y}$ such that the
 KL divergence is minimized.
@@
 \begin{equation}
-\frac{\partial C}{\partial \mathbf{y}_i} = 4 \sum_{j \neq i} \left ( p_{ij} - q_{ij} \right ) \left ( \mathbf{y}_i - \mathbf{y}_j \right ) \left ( 1 + || \mathbf{y}_i - \mathbf{y}_j || ^2 \right )^{-1}.
+\frac{\partial C}{\partial \mathbf{y}_i} = 4 \sum_{j \neq i} \left ( p_{ij} - q_{ij} \right ) \left ( \mathbf{y}_i - \mathbf{y}_j \right ) w_{ij},
 \label{eq:tsne_gradient}
 \end{equation}
+where $w_{ij} = \left ( 1 + || \mathbf{y}_i - \mathbf{y}_j || ^2 \right )^{-1}$
+and represents the unnormalized $q_{ij}$.
 
-Optimization is performed using batch gradient descent with the delta-bar-delta
+Optimization is performed with batch gradient descent using the delta-bar-delta
 update rule~\cite{jacobs1988increased}. Originally, t-SNE was run for 1000
-iterations consisting of two phases. In the first \textit{early exaggeration}
-phase, the attractive forces between data points is increased by a factor
-$\rho=12$ so that points can more easily move throughout the embedding to find
+iterations consisting of two phases: in the first \textit{early exaggeration}
+phase, the attractive forces between data points is increased by some factor
+$\rho$, typically set to $12$, so that points in the embedding can more easily move throughout the embedding to find
 their respective neighbors. The remaining 750 iterations are run with $\rho=1$,
@@
 Belkina \textit{et al.} later found that faster convergence can be achieved by
-increasing the learning rate to $\eta=N/12$~\cite{belkina2019automated}. The
-number of iterations can then safely be lowered to 750, leading to faster
-runtime. This convention has been adopted by most modern t-SNE implementations.
+increasing the learning rate to $\eta=N/12$~\cite{belkina2019automated}. As a side-effect, embeddings converge faster, and the
+number of iterations can be lowered to 750, decreasing the overall runtime.
+This convention has been adopted by most modern t-SNE implementations.
 
@@
 efficient approximation schemes. Van der Maaten observed that the t-SNE
-gradient may be reformulated in terms of a N-body problem where data points may
-be interpreted as particles which attract and repel each
-other~\cite{van2014accelerating}. The gradient may be rewritten as
+gradient may be cast as an N-body problem where data points represent
+particles which attract and repel each
+other~\cite{van2014accelerating}. The gradient from Eqn. (\ref{eq:tsne_gradient}) can be rewritten as
 \begin{equation}
@@
 where $Z = \sum_{k \neq l}\left ( 1 + || \mathbf{y}_k - \mathbf{y}_l ||^2
-\right )^{-1}$. The two terms may be viewed as the attractive and repulsive
-forces between particles, respectively. Each term lends itself to efficient
+\right )^{-1}$. This equation can then be viewed as a particule simulation, where 
+the two terms represent the attractive and repulsive
+forces between individual particles. Each term lends itself to efficient
 approximations, which enable us to greatly reduce the time complexity of t-SNE.
@@
 
-Van der Maaten observed that since $p_{ij}$ is calculated in the
-high-dimensional space using a Gaussian kernel and that the bandwidth is
-selected such that only a fixed number of closest neighbors (determined by the
-perplexity parameter) will contribute to the attractive forces of a given point
-due to the exponential decay of the Gaussian kernel~\cite{van2014accelerating}.
-Therefore it is sufficient to calculate the attractive forces only over a
-relatively small number of nearest neighbors instead of all $N$ points. By
+Van der Maaten observed that evaluating the attractive forces between all pairs
+of data points is excessive, and that considering only a handful of nearest
+neighbors at each point is sufficient to obtain a good approximation~\cite{van2014accelerating}.
+Because distances are transformed to similarities using a Gaussian kernel, and
+the bandwidth of each kernel is selected such that only a predefined number of
+neighbors fall within the main probability mass of the bell curve through the perplexity parameter, the remaining
+data points fall into the exponentially decaying tails of each Gaussian distribution.
+These data points are then assigned near-zero probabilities and do not contribute
+to the overall attractive forces of data points.
+Therefore it is sufficient to calculate the attractive forces for only a
+small number of nearest neighbors instead of all $N$ points. By
 utilizing tree-based nearest-neighbor search methods, the time complexity is
 thus reduced to $\mathcal{O}(N \log N)$. Linderman \textit{et al.} further
-realized that, qualitatively, embeddings are visually virtually
+realized that, qualitatively, embeddings are visually
 indistinguishable when using only \textit{approximate} nearest neighbors,
@@
 space-partitioning Barnes-Hut tree approach to approximate the interaction
-between data points~\cite{van2014accelerating}. Briefly, the approach splits
-the space into quadrants (in the 2D case) and if a given point is far enough
-away from a quadrant, the points in that quadrant are summarized by a single
-center of mass. This reduces the time complexity from $\mathcal{O}(N^2)$ to
+between data points~\cite{van2014accelerating}. Briefly, in the 2D case, the approach splits
+the space into quadrants and entire regions may be summarized by simple statistics.
+If a query point is far away from a given quadrant, the repulsive forces exerted
+by all the points in that quadrant onto the query point are summarized by its center of mass. This
+reduces the time complexity from $\mathcal{O}(N^2)$ to
 $\mathcal{O}(N \log N)$.
@@
 More recently, Linderman \textit{et al.} proposed an alternative approach,
-FIt-SNE, based on non-uniform convolutions to calculating all pairwise
+FIt-SNE, based on non-uniform convolutions for calculating all pairwise
 interactions between repelling data points~\cite{linderman2019fast}. Briefly,
-Linderman \textit{et al.} observed that the repulsive forces $\mathbf{R}$ may
+Linderman \textit{et al.} observed that the repulsive forces $\mathbf{R}$ from Eqn. (\ref{eq:grad_attr_rep}) may
 be rewritten as \begin{align}
@@
 computed as $\mathbf{y}_i \phi_{1,j} - \phi_{2,j}$ and $Z = \sum_j \phi_{3,j}$,
-respectively. These interactions are accelerated by interpolating these terms
+respectively. These interactions are calculated by interpolating the terms
 through a grid of equispaced interpolation points. This shifts the
@@
 
-\opentsne\ provides efficient implementations for all the aforementioned
-improvements but defaults to using approximate nearest neighbor search and the
-non-uniform convolutions approach as these enable the user to quickly create
-informative visualizations for data sets containing up to millions of samples.
+% -------- pointless...
+%\opentsne\ provides efficient implementations for all the aforementioned
+%improvements but defaults to using approximate nearest neighbor search and the
+%non-uniform convolutions approach as these enable the user to quickly create
+%informative visualizations for data sets containing up to millions of samples.
 
@@
 techniques~\cite{policar2019embedding}. When adding new data points to an
-existing, reference embedding, the reference data points are fixed in place and
-the new data points are allowed to find their respective positions. The
+existing, reference embedding, the reference data points are fixed in place while
+new data points are allowed to find their respective positions. The
 optimization remains the same as in standard t-SNE with only slight
@@
 \mathbf{v}_M \right \} \in \mathbb{R}^D$ where $M$ is the number of samples in
-the secondary data set and $\mathbf{W} = \left \{ \mathbf{w}_1, \mathbf{w}_2, \dots,
+the new data set and $\mathbf{W} = \left \{ \mathbf{w}_1, \mathbf{w}_2, \dots,
 \mathbf{w}_M \right \} \in \mathbb{R}^d$. Additionally, we omit the
-symmetrization step in Eqn.~(\ref{eq:symmetrize}). The gradients of
+symmetrization step in Eqn.~(\ref{eq:symmetrize}).
+Plugging these
+The gradients of
 $\mathbf{w}_j$ with respect to the loss (Eqn.~\ref{eq:kl_divergence}) are:
 \begin{equation}
-\frac{\partial C}{\partial \mathbf{w}_j} = 2 \sum_i \left ( p_{j \mid i} - q_{j \mid i} \right ) \left ( \mathbf{y}_i - \mathbf{w}_j \right ) \left ( 1 + || \mathbf{y}_i - \mathbf{w}_j || ^2 \right )^{-1}
+\frac{\partial C}{\partial \mathbf{w}_j} = 2 \sum_i \left ( p_{j \mid i} - q_{j \mid i} \right ) \left ( \mathbf{y}_i - \mathbf{w}_j \right ) \left ( 1 + || \mathbf{y}_i - \mathbf{w}_j || ^2 \right )^{-1}.
 \label{eq:gradient}
@@
 
-Kobak \& Berens~\cite{kobak2019art} suggest that replacing the Gaussian kernel
+Kobak \& Berens suggest that replacing the Gaussian kernel
 with a mixture of Gaussians may provide better insight into both the local and
-global structure. Therefore, the similarities between data points in the input
-space become
+global structure~\cite{kobak2019art}. Therefore, the similarities between data points in the input
+space are computed with
 
@@
 
-The bandwidths of each Gaussian $\sigma_{1,i}$ and $\sigma_{2,i}$ is determined
-by selecting different perplexity values. They show that using a kernel with
-perplexity 500 captures the long-range interactions which preserve global
-cluster organization in the final embedding and combine this with a kernel with
-perplexity 50, which prevents small, well-defined clusters from being absorbed
-into larger ones.
+The bandwidth of each Gaussian $\sigma_{1,i}$ and $\sigma_{2,i}$ is determined
+by selecting different perplexity values. Using a kernel with
+perplexity 500 captures long-range interactions which preserve global
+cluster organization. Combining these with a more narrow kernel with
+perplexity 50 prevents small, well-defined clusters from being absorbed
+into larger ones, leading a better overall snapshot into the data structure.
 
@@
 
-Standard t-SNE reveals the clustering structure in data at one level of
+Standard t-SNE reveals the clustering structure at a single level of
 resolution. Different perplexity parameter values can be used to identify
@@
 varying perplexity values can be time-consuming as this involves recomputing
-the KNNG which is often the most expensive part of the t-SNE algorithm.
+the KNN graph which is often the most expensive part of the t-SNE algorithm.
 Alternatively, Kobak \textit{et al.}~\cite{kobak2019heavy} suggest that varying
-the degree of freedom in the t-distribution used in the embedding space during
-optimization can be used to explore the clustering structure at different
+the degree of freedom in the t-distribution can be used to explore the clustering structure at different
 levels of resolution.
@@
 \end{equation}
-
 In standard t-SNE $\alpha=1$ so this simplifies to the Cauchy kernel from
-Eqn.~(\ref{eq:cauchy_kernel}).
-
-The gradient of the loss function then becomes
+Eqn.~(\ref{eq:cauchy_kernel}). The gradient of the loss function then becomes
 \begin{align}
-\frac{\partial C}{\partial \mathbf{y}_i} &= 4 \sum_{j \neq i} \left ( p_{ij} - q_{ij} \right ) q_{ij}^{1/\alpha} \left ( \mathbf{y}_i - \mathbf{y}_j \right ),
+\frac{\partial C}{\partial \mathbf{y}_i} &= 4 \sum_{j \neq i} \left ( p_{ij} - q_{ij} \right ) w_{ij}^{1/\alpha} \left ( \mathbf{y}_i - \mathbf{y}_j \right ),
 \end{align}
-which can, again, be decomposed into the attractive and repulsive forces as
+which can, again, be cast as attractive and repulsive forces between particles
 \begin{align}
-\frac{\partial C}{\partial \mathbf{y}_i} &= 4 \sum_{j \neq i} p_{ij} q_{ij}^{1/\alpha} (\mathbf{y}_i - \mathbf{y}_j) \notag \\
-&- 4 \sum_{j \neq i} q_{ij}^{\frac{\alpha+1}{\alpha}} / Z (\mathbf{y}_i - \mathbf{y}_j).
+\frac{\partial C}{\partial \mathbf{y}_i} &= 4 \sum_{j \neq i} p_{ij} w_{ij}^{1/\alpha} (\mathbf{y}_i - \mathbf{y}_j) \notag \\
+&- 4 \sum_{j \neq i} w_{ij}^{\frac{\alpha+1}{\alpha}} / Z (\mathbf{y}_i - \mathbf{y}_j).
 \end{align}
@@
 Adapting existing approximation schemes to this formulation is straightforward.
-We provide efficient implementations of both the Barnes-Hut and the FIt-SNE
-approximation schemes, where we replace the terms from
-Eqn.~(\ref{eq:fitsne_terms}) with
+\opentsne\ provides efficient implementations of both the Barnes-Hut and the FIt-SNE
+approximation schemes, where we modify the terms from
+Eqn.~(\ref{eq:fitsne_terms}) to
 
@@
 Embeddings produced by standard t-SNE often produce clusters separated by thin
-boundaries and make use of all available space. While this is often desirable,
-this often obscures the global relationships between clusters as all
-neighboring clusters appear at the same distance from one another. Other
+boundaries and make use of all available space. While this is desirable at times,
+it often obscures the global relationships between clusters as all
+neighboring clusters appear about the same distance from one another. Other
 dimensionality reduction methods such as UMAP~\cite{2018arXivUMAP} or
@@
 
-Kobak \textit{et al.}~\cite{todo} showed that the early exaggeration factor
-$\rho$ in t-SNE may be increased to produce more compact layouts often seen in
-UMAP or ForceAtlas2. Early exaggeration is a trick introduced by van der
-Maaten~\cite{maaten2008visualizing} used in the first 250 iterations and
-increases the attractive forces between data points, enabling neighboring
-points to move close to one another while ignoring the repulsive forces between
-data points. In standard t-SNE, the remainder of the iterations are run with
-$\rho=1$. Kobak \textit{et al.} showed that using $\rho=4$ and $\rho=30$ in the
-second phase of the optimization results in embeddings visually similar to UMAP
-embeddings, and ForceAtlas2 embeddings, respectively.
+B{\"o}hm \textit{et al.} showed that the exaggeration factor
+$\rho$ can be used to produce layouts more similar to UMAP and ForceAtlas2~\cite{bohm2020unifying}.
+By incorporating exaggeration into later phases of optimization, t-SNE is able
+introduce more white-space between clusters and better reflect the global
+relations between clusters. B{\"o}hm \textit{et al.} found that using $\rho=4$ and $\rho=30$
+produces embeddings visually similar to UMAP and ForceAtlas2, respectively.
 
@@
 distinct groups of cell-types but obscures developmental paths. This problem is
-exacerbated when dealing with large numbers of data points. Conversely,
+exacerbated when dealing with large numbers of data points. On the other hand,
 ForceAtlas2 has often been used to uncover trajectories and transitions between
 cell types, but larger clusters often absorb small, distinct groups of cells.
+Therefore one may run t-SNE with two different exaggeration settings, one to
+highlight the developmental transitions between cell types and one to identify
+clear populations of cells.
 
@@
 
-Similarly to UMAP and ForceAtlas2, t-SNE can also be viewed as a force-directed
-layout algorithm. The first phase of the t-SNE algorithm constructs a
-$k$-nearest neighbor graph, and the subsequent optimization balances the
-attractive and repulsive forces between data points.
+t-SNE, UMAP, and ForceAtlas2 can all be cast as force-directed layout algorithms,
+which operate on the $k$-nearest neighbor graph. Each method constructs the graph
+and specifies the attractive and repulsive forces in its own way, but ultimately,
+point positions are found by balancing the attractive and repulsive forces between data points.
 
 The final positioning of data points using this class of algorithms is largely
-dependent on the initial positions in the embedding space. UMAP is often said
-to better preserve global structure than t-SNE, but Kobak \& Linderman have
-recently discovered that this supposed advantage is entirely due to the better
-default initialization in UMAP than in older implementations of
-t-SNE~\cite{kobak2019umap}. Standard implementations of t-SNE use random
-initializations, which often scatter resulting clusters throughout the
-embedding space. UMAP injects global organization into the embedding by
-initializing point positions using Laplacian eigenmaps. When initialized in the
-same manner, both t-SNE and UMAP perform equally well, or equally poorly,
-dispelling the myth that UMAP is superior in this respect.
+dependent on the embedding initialization. UMAP found early success as it was
+able to produce visualizations that better captured the global organization of
+clusters. This was due to its initialization scheme, which initialized point
+positions using Laplacian eigenmaps. On the other hand, most implementations of
+t-SNE performed random initialization, which resulted in poor global coherence.
+However, Kobak \& Linderman recently showed that using the same initialization
+for both methods leads to visualizations, which exhibit similarly good global
+coherency. Conversely, when initialized randomly, both methods produce
+visualizations where clusters are arbitrarily positioned in the embedding space.
 
 \opentsne\ defaults to using the two leading principal components as
-initialization, but also provides a spectral approach, similar to that of UMAP.
-The clusters in resulting embeddings are organized in a globally coherent
-manner, leading to more interpretable visualizations.
+initialization, but also provides a spectral approach similar to UMAP.
+This allows clusters in resulting embeddings to be organized in a more globally coherent
+manner, leading to increased visualization interpretability.
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                          %%
-%% Backmatter begins here                   %%
-%%                                          %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \begin{backmatter}
@@
 \section*{Author's contributions}
-    Text for this section \ldots
+P.G.P.
+implemented and maintains \opentsne library.
+M.S.
+helped in development of mathematical foundations and advised on inclusion of the extensions.
+B.Z.
+initiated and supervised the project.
+All three co-authors wrote the manuscript.
 
 \section*{Acknowledgements}
-We would like to thank Dmitry Kobak for various helpful discussions and best
-	practices when using t-SNE, as well as his contributions to the source
-	code. We would also like to thank George Linderman for his help with
-	the FIt-SNE algorithm.
+We want to thank Dmitry Kobak for various helpful discussions and best practices when using t-SNE and his contributions to the source code.
+We would also like to thank George Linderman for his help with the FIt-SNE algorithm.
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                  The Bibliography                       %%
-%%                                                         %%
-%%  Bmc_mathpys.bst  will be used to                       %%
-%%  create a .BBL file for submission.                     %%
-%%  After submission of the .TEX file,                     %%
-%%  you will be prompted to submit your .BBL file.         %%
-%%                                                         %%
-%%                                                         %%
-%%  Note that the displayed Bibliography will not          %%
-%%  necessarily be rendered by Latex exactly as specified  %%
-%%  in the online Instructions for Authors.                %%
-%%                                                         %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% if your bibliography is in bibtex format, use those commands:
 \bibliographystyle{bmc-mathphys} % Style BST file (bmc-mathphys, vancouver, spbasic).
 \bibliography{references}      % Bibliography file (usually '*.bib' )
-% for author-year bibliography (bmc-mathphys or spbasic)
-% a) write to bib file (bmc-mathphys only)
-% @settings{label, options="nameyear"}
-% b) uncomment next line
-%\nocite{label}
 

### 2020-09-30 paper/policar2020-opentsne.tex
Edit methods section

--- 2020-07-12 paper/policar2020-opentsne.tex
@@
 embedding techniques such as UMAP and ForceAtlas2~\cite{jacomy2014forceatlas2} tend to better capture the continuity between cell
-states. Recently, Kobak \textit{et al.}~\cite{TODO} have shown that embeddings
+states. Recently, B{\"o}hm \textit{et al.}~\cite{bohm2020unifying} have shown that embeddings
 produced by t-SNE with exaggeration values of 4 and $\sim30$ produce embeddings
@@
 \begin{equation}
-\frac{\partial C}{\partial \mathbf{y}_i} = 4 \sum_{j \neq i} \left ( p_{ij} - q_{ij} \right ) \left ( \mathbf{y}_i - \mathbf{y}_j \right ) \left ( 1 + || \mathbf{y}_i - \mathbf{y}_j || ^2 \right )^{-1}.
+\frac{\partial C}{\partial \mathbf{y}_i} = 4 \sum_{j \neq i} \left ( p_{ij} - q_{ij} \right ) \left ( \mathbf{y}_i - \mathbf{y}_j \right ) w_{ij},
 \label{eq:tsne_gradient}
 \end{equation}
-
-Optimization is performed using batch gradient descent with the delta-bar-delta
+where $w_{ij} = \left ( 1 + || \mathbf{y}_i - \mathbf{y}_j || ^2 \right )^{-1}$
+and represents the unnormalized $q_{ij}$.
+
+Optimization is performed with batch gradient descent using the delta-bar-delta
 update rule~\cite{jacobs1988increased}. Originally, t-SNE was run for 1000
-iterations consisting of two phases. In the first \textit{early exaggeration}
-phase, the attractive forces between data points is increased by a factor
-$\rho=12$ so that points can more easily move throughout the embedding to find
+iterations consisting of two phases: in the first \textit{early exaggeration}
+phase, the attractive forces between data points is increased by some factor
+$\rho$, typically set to $12$, so that points in the embedding can more easily move throughout the embedding to find
 their respective neighbors. The remaining 750 iterations are run with $\rho=1$,

### 2020-10-01 paper/policar2020-opentsne.tex
More edits to methods section

@@
 iterations consisting of two phases: in the first \textit{early exaggeration}
-phase, the attractive forces between data points is increased by some factor
-$\rho$, typically set to $12$, so that points in the embedding can more easily move throughout the embedding to find
+phase, the attractive forces between data points are increased by some factor
+$\rho$, typically set to $12$, so that points in the embedding can more easily move throughout the space and find
 their respective neighbors. The remaining 750 iterations are run with $\rho=1$,
@@
 where $Z = \sum_{k \neq l}\left ( 1 + || \mathbf{y}_k - \mathbf{y}_l ||^2
-\right )^{-1}$. This equation can then be viewed as a particule simulation, where 
+\right )^{-1}$. This equation can then be viewed as a particle simulation, where 
 the two terms represent the attractive and repulsive
@@
 data points fall into the exponentially decaying tails of each Gaussian distribution.
-These data points are then assigned near-zero probabilities and do not contribute
+These data points are assigned near-zero probabilities and do not contribute
 to the overall attractive forces of data points.
@@
 If a query point is far away from a given quadrant, the repulsive forces exerted
-by all the points in that quadrant onto the query point are summarized by its center of mass. This
+by all the points in that quadrant onto the query point are summarized by a single point. This
 reduces the time complexity from $\mathcal{O}(N^2)$ to
@@
 symmetrization step in Eqn.~(\ref{eq:symmetrize}).
-Plugging these
-The gradients of
-$\mathbf{w}_j$ with respect to the loss (Eqn.~\ref{eq:kl_divergence}) are:
+Plugging these terms into Eqn.~(\ref{eq:kl_divergence}), we obtain the following
+gradient
 \begin{equation}
@@
 
-As in standard t-SNE, a direct calculation of gradients takes $\mathcal{O}(N
+Similarly to standard t-SNE, a direct calculation of gradients takes $\mathcal{O}(N
 \cdot M)$ time, but it is straightforward to adapt the Barnes-Hut and FIt-SNE
@@
 determined by the user-specified perplexity parameter so that a fixed number of
-nearest data points will have non-zero similarity. One common trick for
+nearest data points will be assigned non-zero values. One common trick for
 uncovering the global relations between clusters is to increase perplexity so
@@
 with a mixture of Gaussians may provide better insight into both the local and
-global structure~\cite{kobak2019art}. Therefore, the similarities between data points in the input
-space are computed with
+global structure~\cite{kobak2019art}. For instance, the similarities between data points in the input
+space may instead be computed with
 
@@
 varying perplexity values can be time-consuming as this involves recomputing
-the KNN graph which is often the most expensive part of the t-SNE algorithm.
-Alternatively, Kobak \textit{et al.}~\cite{kobak2019heavy} suggest that varying
+the $k$-nearest neighbor graph which is often the most expensive part of the t-SNE algorithm.
+Alternatively, Kobak \textit{et al.} suggest that varying
 the degree of freedom in the t-distribution can be used to explore the clustering structure at different
-levels of resolution.
+levels of resolution~\cite{kobak2019heavy}.
 
 Standard t-SNE models similarities between data points in the embedding space
-using a t-distribution with a single degree of freedom
+using a t-distribution with a single degree of freedom, but this can be generalized
+to allow for any parameter value
 
@@
 \end{align}
-which can, again, be cast as attractive and repulsive forces between particles
+which can, again, be cast as the interplay between the attractive and repulsive forces between particles
 \begin{align}
@@
 By incorporating exaggeration into later phases of optimization, t-SNE is able
-introduce more white-space between clusters and better reflect the global
+introduce more white-space between clusters which better reflects the global
 relations between clusters. B{\"o}hm \textit{et al.} found that using $\rho=4$ and $\rho=30$
@@
 While it is difficult to claim one is better than the other, different settings
-of $\rho$ may uncover different properties and clustering structures. For
-example, in single-cell data, standard t-SNE with $\rho=1$ often uncovers
-distinct groups of cell-types but obscures developmental paths. This problem is
+of $\rho$ may uncover different properties of the data manifold. For
+example, in single-cell data, standard t-SNE often uncovers
+distinct, well-separated groups of cell-types but obscures transitional paths between cell-states. This problem is
 exacerbated when dealing with large numbers of data points. On the other hand,
-ForceAtlas2 has often been used to uncover trajectories and transitions between
-cell types, but larger clusters often absorb small, distinct groups of cells.
-Therefore one may run t-SNE with two different exaggeration settings, one to
-highlight the developmental transitions between cell types and one to identify
-clear populations of cells.
+ForceAtlas2 has been used successfuly to uncover trajectories and transitions between
+cell types. Unfortunately, when using ForceAtlas2, large clusters often absorb small, distinct groups of cells.
+One might run t-SNE with various parameter settings to avoid switching between
+different algorithms and obtain similar visualizations.
+We might generate an embedding using higher levels of exaggeration to highlight
+the developmental transitions between cell types and
+one with low levels of exaggeration to identify clear populations of cells.
 
@@
 t-SNE, UMAP, and ForceAtlas2 can all be cast as force-directed layout algorithms,
-which operate on the $k$-nearest neighbor graph. Each method constructs the graph
+which operate on the $k$-nearest neighbor graph. Each method constructs the graph differently
 and specifies the attractive and repulsive forces in its own way, but ultimately,
@@
 
-The final positioning of data points using this class of algorithms is largely
+The final positions of data points in this class of algorithms is largely
 dependent on the embedding initialization. UMAP found early success as it was
@@
 
-openTSNE is distributed under the BSD-3-Clause License and is publicly
+\opentsne\ is distributed under the BSD-3-Clause License and is publicly
 available as an open-source package at
-\url{https://github.com/pavlin-policar/openTSNE}. openTSNE is also available on
-PyPI and conda-forge. The data sets used and scripts used in this study are
+\url{https://github.com/pavlin-policar/openTSNE}. \opentsne\ is also available on
+\texttt{PyPI} and \texttt{conda-forge}. The data sets used and scripts used in this study are
 included in accompanying notebooks, publicly available at

### 2020-10-01 paper/policar2020-opentsne.tex
revision of the main text

@@
 To this end, other embedding techniques such as UMAP and ForceAtlas2~\cite{jacomy2014forceatlas2} tend to capture the continuity between cell states better.
-Recently, B{\"o}hm \textit{et al.}~\cite{bohm2020unifying} have shown that embeddings produced by t-SNE with exaggeration values of 4 and $\sim30$ construct embeddings which are markedly similar to UMAP and ForceAtlas2, respectively.
+Recently, Kobak \textit{et al.}~\cite{TODO} have shown that embeddings produced by t-SNE with exaggeration values of 4 and $\sim30$ construct embeddings which are markedly similar to UMAP and ForceAtlas2, respectively.
 For example, in Fig.~\ref{fig:cao}.a, the developmental trajectory between different time points is difficult to observe due to many sprawled out clusters.
@@
 primarily used as a visualization tool, $d$ is typically set to two. The
-similarity between two data points in the high-dimensional space is defined as
+similarity between two data points in the high-dimensional space is defined as:
 
@@
 density of the data in the multi-dimensional space. Perplexity is sometimes
-interpreted as the continuous analog to the number of nearest neighbors to which
-distances will attempt to be preserved.
+considered the continuous analog to the $k$ nearest neighbors to which
+distances will be preserved.
 
@@
 embedding space is defined using the $t$-distribution with a single degree of
-freedom (Cauchy kernel)
+freedom. This is also known as the Cauchy kernel:
 
@@
 
-The Kullback-Leibler (KL) divergence is used as a measure of agreement
+Finally, the Kullback-Leibler (KL) divergence is used as a measure of agreement
 between distributions $\mathbf{P}$ and $\mathbf{Q}$
@@
 
-\noindent The objective is to find embeddings $\mathbf{Y}$ such that the
+\noindent The t-SNE objective is to find embeddings $\mathbf{Y}$ such that the
 KL divergence is minimized.
@@
 \begin{equation}
-\frac{\partial C}{\partial \mathbf{y}_i} = 4 \sum_{j \neq i} \left ( p_{ij} - q_{ij} \right ) \left ( \mathbf{y}_i - \mathbf{y}_j \right ) w_{ij},
+\frac{\partial C}{\partial \mathbf{y}_i} = 4 \sum_{j \neq i} \left ( p_{ij} - q_{ij} \right ) \left ( \mathbf{y}_i - \mathbf{y}_j \right ) \left ( 1 + || \mathbf{y}_i - \mathbf{y}_j || ^2 \right )^{-1}.
 \label{eq:tsne_gradient}
 \end{equation}
-where $w_{ij} = \left ( 1 + || \mathbf{y}_i - \mathbf{y}_j || ^2 \right )^{-1}$
-and represents the unnormalized $q_{ij}$.
 
-Optimization is performed with batch gradient descent using the delta-bar-delta
+Optimization is performed using batch gradient descent with the delta-bar-delta
 update rule~\cite{jacobs1988increased}. Originally, t-SNE was run for 1000
-iterations consisting of two phases: in the first \textit{early exaggeration}
-phase, the attractive forces between data points are increased by some factor
-$\rho$, typically set to $12$, so that points in the embedding can more easily move throughout the space and find
+iterations consisting of two phases. In the first \textit{early exaggeration}
+phase, the attractive forces between data points is increased by a factor
+$\rho=12$ so that points can more easily move throughout the embedding to find
 their respective neighbors. The remaining 750 iterations are run with $\rho=1$,
@@
 Belkina \textit{et al.} later found that faster convergence can be achieved by
-increasing the learning rate to $\eta=N/12$~\cite{belkina2019automated}. As a side-effect, embeddings converge faster, and the
-number of iterations can be lowered to 750, decreasing the overall runtime.
-This convention has been adopted by most modern t-SNE implementations.
+increasing the learning rate to $\eta=N/12$~\cite{belkina2019automated}. The
+number of iterations can then safely be lowered to 750, leading to faster
+runtime. This convention has been adopted by most modern t-SNE implementations.
 
@@
 efficient approximation schemes. Van der Maaten observed that the t-SNE
-gradient may be cast as an N-body problem where data points represent
-particles which attract and repel each
-other~\cite{van2014accelerating}. The gradient from Eqn. (\ref{eq:tsne_gradient}) can be rewritten as
+gradient may be reformulated in terms of a N-body problem where data points may
+be interpreted as particles which attract and repel each
+other~\cite{van2014accelerating}. The gradient may be rewritten as
 \begin{equation}
@@
 where $Z = \sum_{k \neq l}\left ( 1 + || \mathbf{y}_k - \mathbf{y}_l ||^2
-\right )^{-1}$. This equation can then be viewed as a particle simulation, where 
-the two terms represent the attractive and repulsive
-forces between individual particles. Each term lends itself to efficient
+\right )^{-1}$. The two terms may be viewed as the attractive and repulsive
+forces between particles, respectively. Each term lends itself to efficient
 approximations, which enable us to greatly reduce the time complexity of t-SNE.
@@
 
-Van der Maaten observed that evaluating the attractive forces between all pairs
-of data points is excessive, and that considering only a handful of nearest
-neighbors at each point is sufficient to obtain a good approximation~\cite{van2014accelerating}.
-Because distances are transformed to similarities using a Gaussian kernel, and
-the bandwidth of each kernel is selected such that only a predefined number of
-neighbors fall within the main probability mass of the bell curve through the perplexity parameter, the remaining
-data points fall into the exponentially decaying tails of each Gaussian distribution.
-These data points are assigned near-zero probabilities and do not contribute
-to the overall attractive forces of data points.
-Therefore it is sufficient to calculate the attractive forces for only a
-small number of nearest neighbors instead of all $N$ points. By
+Van der Maaten observed that since $p_{ij}$ is calculated in the
+high-dimensional space using a Gaussian kernel and that the bandwidth is
+selected such that only a fixed number of closest neighbors (determined by the
+perplexity parameter) will contribute to the attractive forces of a given point
+due to the exponential decay of the Gaussian kernel~\cite{van2014accelerating}.
+Therefore it is sufficient to calculate the attractive forces only over a
+relatively small number of nearest neighbors instead of all $N$ points. By
 utilizing tree-based nearest-neighbor search methods, the time complexity is
 thus reduced to $\mathcal{O}(N \log N)$. Linderman \textit{et al.} further
-realized that, qualitatively, embeddings are visually
+realized that, qualitatively, embeddings are visually virtually
 indistinguishable when using only \textit{approximate} nearest neighbors,
@@
 space-partitioning Barnes-Hut tree approach to approximate the interaction
-between data points~\cite{van2014accelerating}. Briefly, in the 2D case, the approach splits
-the space into quadrants and entire regions may be summarized by simple statistics.
-If a query point is far away from a given quadrant, the repulsive forces exerted
-by all the points in that quadrant onto the query point are summarized by a single point. This
-reduces the time complexity from $\mathcal{O}(N^2)$ to
+between data points~\cite{van2014accelerating}. Briefly, the approach splits
+the space into quadrants (in the 2D case) and if a given point is far enough
+away from a quadrant, the points in that quadrant are summarized by a single
+center of mass. This reduces the time complexity from $\mathcal{O}(N^2)$ to
 $\mathcal{O}(N \log N)$.
@@
 More recently, Linderman \textit{et al.} proposed an alternative approach,
-FIt-SNE, based on non-uniform convolutions for calculating all pairwise
+FIt-SNE, based on non-uniform convolutions to calculating all pairwise
 interactions between repelling data points~\cite{linderman2019fast}. Briefly,
-Linderman \textit{et al.} observed that the repulsive forces $\mathbf{R}$ from Eqn. (\ref{eq:grad_attr_rep}) may
+Linderman \textit{et al.} observed that the repulsive forces $\mathbf{R}$ may
 be rewritten as \begin{align}
@@
 computed as $\mathbf{y}_i \phi_{1,j} - \phi_{2,j}$ and $Z = \sum_j \phi_{3,j}$,
-respectively. These interactions are calculated by interpolating the terms
+respectively. These interactions are accelerated by interpolating these terms
 through a grid of equispaced interpolation points. This shifts the
@@
 
-% -------- pointless...
-%\opentsne\ provides efficient implementations for all the aforementioned
-%improvements but defaults to using approximate nearest neighbor search and the
-%non-uniform convolutions approach as these enable the user to quickly create
-%informative visualizations for data sets containing up to millions of samples.
+\opentsne\ provides efficient implementations for all the aforementioned
+improvements but defaults to using approximate nearest neighbor search and the
+non-uniform convolutions approach as these enable the user to quickly create
+informative visualizations for data sets containing up to millions of samples.
 
@@
 techniques~\cite{policar2019embedding}. When adding new data points to an
-existing, reference embedding, the reference data points are fixed in place while
-new data points are allowed to find their respective positions. The
+existing, reference embedding, the reference data points are fixed in place and
+the new data points are allowed to find their respective positions. The
 optimization remains the same as in standard t-SNE with only slight
@@
 \mathbf{v}_M \right \} \in \mathbb{R}^D$ where $M$ is the number of samples in
-the new data set and $\mathbf{W} = \left \{ \mathbf{w}_1, \mathbf{w}_2, \dots,
+the secondary data set and $\mathbf{W} = \left \{ \mathbf{w}_1, \mathbf{w}_2, \dots,
 \mathbf{w}_M \right \} \in \mathbb{R}^d$. Additionally, we omit the
-symmetrization step in Eqn.~(\ref{eq:symmetrize}).
-Plugging these terms into Eqn.~(\ref{eq:kl_divergence}), we obtain the following
-gradient
+symmetrization step in Eqn.~(\ref{eq:symmetrize}). The gradients of
+$\mathbf{w}_j$ with respect to the loss (Eqn.~\ref{eq:kl_divergence}) are:
 \begin{equation}
-\frac{\partial C}{\partial \mathbf{w}_j} = 2 \sum_i \left ( p_{j \mid i} - q_{j \mid i} \right ) \left ( \mathbf{y}_i - \mathbf{w}_j \right ) \left ( 1 + || \mathbf{y}_i - \mathbf{w}_j || ^2 \right )^{-1}.
+\frac{\partial C}{\partial \mathbf{w}_j} = 2 \sum_i \left ( p_{j \mid i} - q_{j \mid i} \right ) \left ( \mathbf{y}_i - \mathbf{w}_j \right ) \left ( 1 + || \mathbf{y}_i - \mathbf{w}_j || ^2 \right )^{-1}
 \label{eq:gradient}
@@
 
-Similarly to standard t-SNE, a direct calculation of gradients takes $\mathcal{O}(N
+As in standard t-SNE, a direct calculation of gradients takes $\mathcal{O}(N
 \cdot M)$ time, but it is straightforward to adapt the Barnes-Hut and FIt-SNE
@@
 determined by the user-specified perplexity parameter so that a fixed number of
-nearest data points will be assigned non-zero values. One common trick for
+nearest data points will have non-zero similarity. One common trick for
 uncovering the global relations between clusters is to increase perplexity so
@@
 
-Kobak \& Berens suggest that replacing the Gaussian kernel
+Kobak \& Berens~\cite{kobak2019art} suggest that replacing the Gaussian kernel
 with a mixture of Gaussians may provide better insight into both the local and
-global structure~\cite{kobak2019art}. For instance, the similarities between data points in the input
-space may instead be computed with
+global structure. Therefore, the similarities between data points in the input
+space become
 
@@
 
-The bandwidth of each Gaussian $\sigma_{1,i}$ and $\sigma_{2,i}$ is determined
-by selecting different perplexity values. Using a kernel with
-perplexity 500 captures long-range interactions which preserve global
-cluster organization. Combining these with a more narrow kernel with
-perplexity 50 prevents small, well-defined clusters from being absorbed
-into larger ones, leading a better overall snapshot into the data structure.
+The bandwidths of each Gaussian $\sigma_{1,i}$ and $\sigma_{2,i}$ is determined
+by selecting different perplexity values. They show that using a kernel with
+perplexity 500 captures the long-range interactions which preserve global
+cluster organization in the final embedding and combine this with a kernel with
+perplexity 50, which prevents small, well-defined clusters from being absorbed
+into larger ones.
 
@@
 
-Standard t-SNE reveals the clustering structure at a single level of
+Standard t-SNE reveals the clustering structure in data at one level of
 resolution. Different perplexity parameter values can be used to identify
@@
 varying perplexity values can be time-consuming as this involves recomputing
-the $k$-nearest neighbor graph which is often the most expensive part of the t-SNE algorithm.
-Alternatively, Kobak \textit{et al.} suggest that varying
-the degree of freedom in the t-distribution can be used to explore the clustering structure at different
-levels of resolution~\cite{kobak2019heavy}.
+the KNNG which is often the most expensive part of the t-SNE algorithm.
+Alternatively, Kobak \textit{et al.}~\cite{kobak2019heavy} suggest that varying
+the degree of freedom in the t-distribution used in the embedding space during
+optimization can be used to explore the clustering structure at different
+levels of resolution.
 
 Standard t-SNE models similarities between data points in the embedding space
-using a t-distribution with a single degree of freedom, but this can be generalized
-to allow for any parameter value
+using a t-distribution with a single degree of freedom
 
@@
 \end{equation}
+
 In standard t-SNE $\alpha=1$ so this simplifies to the Cauchy kernel from
-Eqn.~(\ref{eq:cauchy_kernel}). The gradient of the loss function then becomes
+Eqn.~(\ref{eq:cauchy_kernel}).
+
+The gradient of the loss function then becomes
 \begin{align}
-\frac{\partial C}{\partial \mathbf{y}_i} &= 4 \sum_{j \neq i} \left ( p_{ij} - q_{ij} \right ) w_{ij}^{1/\alpha} \left ( \mathbf{y}_i - \mathbf{y}_j \right ),
+\frac{\partial C}{\partial \mathbf{y}_i} &= 4 \sum_{j \neq i} \left ( p_{ij} - q_{ij} \right ) q_{ij}^{1/\alpha} \left ( \mathbf{y}_i - \mathbf{y}_j \right ),
 \end{align}
-which can, again, be cast as the interplay between the attractive and repulsive forces between particles
+which can, again, be decomposed into the attractive and repulsive forces as
 \begin{align}
-\frac{\partial C}{\partial \mathbf{y}_i} &= 4 \sum_{j \neq i} p_{ij} w_{ij}^{1/\alpha} (\mathbf{y}_i - \mathbf{y}_j) \notag \\
-&- 4 \sum_{j \neq i} w_{ij}^{\frac{\alpha+1}{\alpha}} / Z (\mathbf{y}_i - \mathbf{y}_j).
+\frac{\partial C}{\partial \mathbf{y}_i} &= 4 \sum_{j \neq i} p_{ij} q_{ij}^{1/\alpha} (\mathbf{y}_i - \mathbf{y}_j) \notag \\
+&- 4 \sum_{j \neq i} q_{ij}^{\frac{\alpha+1}{\alpha}} / Z (\mathbf{y}_i - \mathbf{y}_j).
 \end{align}
@@
 Adapting existing approximation schemes to this formulation is straightforward.
-\opentsne\ provides efficient implementations of both the Barnes-Hut and the FIt-SNE
-approximation schemes, where we modify the terms from
-Eqn.~(\ref{eq:fitsne_terms}) to
+We provide efficient implementations of both the Barnes-Hut and the FIt-SNE
+approximation schemes, where we replace the terms from
+Eqn.~(\ref{eq:fitsne_terms}) with
 
@@
 Embeddings produced by standard t-SNE often produce clusters separated by thin
-boundaries and make use of all available space. While this is desirable at times,
-it often obscures the global relationships between clusters as all
-neighboring clusters appear about the same distance from one another. Other
+boundaries and make use of all available space. While this is often desirable,
+this often obscures the global relationships between clusters as all
+neighboring clusters appear at the same distance from one another. Other
 dimensionality reduction methods such as UMAP~\cite{2018arXivUMAP} or
@@
 
-B{\"o}hm \textit{et al.} showed that the exaggeration factor
-$\rho$ can be used to produce layouts more similar to UMAP and ForceAtlas2~\cite{bohm2020unifying}.
-By incorporating exaggeration into later phases of optimization, t-SNE is able
-introduce more white-space between clusters which better reflects the global
-relations between clusters. B{\"o}hm \textit{et al.} found that using $\rho=4$ and $\rho=30$
-produces embeddings visually similar to UMAP and ForceAtlas2, respectively.
+Kobak \textit{et al.}~\cite{todo} showed that the early exaggeration factor
+$\rho$ in t-SNE may be increased to produce more compact layouts often seen in
+UMAP or ForceAtlas2. Early exaggeration is a trick introduced by van der
+Maaten~\cite{maaten2008visualizing} used in the first 250 iterations and
+increases the attractive forces between data points, enabling neighboring
+points to move close to one another while ignoring the repulsive forces between
+data points. In standard t-SNE, the remainder of the iterations are run with
+$\rho=1$. Kobak \textit{et al.} showed that using $\rho=4$ and $\rho=30$ in the
+second phase of the optimization results in embeddings visually similar to UMAP
+embeddings, and ForceAtlas2 embeddings, respectively.
 
 While it is difficult to claim one is better than the other, different settings
-of $\rho$ may uncover different properties of the data manifold. For
-example, in single-cell data, standard t-SNE often uncovers
-distinct, well-separated groups of cell-types but obscures transitional paths between cell-states. This problem is
-exacerbated when dealing with large numbers of data points. On the other hand,
-ForceAtlas2 has been used successfuly to uncover trajectories and transitions between
-cell types. Unfortunately, when using ForceAtlas2, large clusters often absorb small, distinct groups of cells.
-One might run t-SNE with various parameter settings to avoid switching between
-different algorithms and obtain similar visualizations.
-We might generate an embedding using higher levels of exaggeration to highlight
-the developmental transitions between cell types and
-one with low levels of exaggeration to identify clear populations of cells.
+of $\rho$ may uncover different properties and clustering structures. For
+example, in single-cell data, standard t-SNE with $\rho=1$ often uncovers
+distinct groups of cell-types but obscures developmental paths. This problem is
+exacerbated when dealing with large numbers of data points. Conversely,
+ForceAtlas2 has often been used to uncover trajectories and transitions between
+cell types, but larger clusters often absorb small, distinct groups of cells.
 
@@
 
-t-SNE, UMAP, and ForceAtlas2 can all be cast as force-directed layout algorithms,
-which operate on the $k$-nearest neighbor graph. Each method constructs the graph differently
-and specifies the attractive and repulsive forces in its own way, but ultimately,
-point positions are found by balancing the attractive and repulsive forces between data points.
+Similarly to UMAP and ForceAtlas2, t-SNE can also be viewed as a force-directed
+layout algorithm. The first phase of the t-SNE algorithm constructs a
+$k$-nearest neighbor graph, and the subsequent optimization balances the
+attractive and repulsive forces between data points.
 
-The final positions of data points in this class of algorithms is largely
-dependent on the embedding initialization. UMAP found early success as it was
-able to produce visualizations that better captured the global organization of
-clusters. This was due to its initialization scheme, which initialized point
-positions using Laplacian eigenmaps. On the other hand, most implementations of
-t-SNE performed random initialization, which resulted in poor global coherence.
-However, Kobak \& Linderman recently showed that using the same initialization
-for both methods leads to visualizations, which exhibit similarly good global
-coherency. Conversely, when initialized randomly, both methods produce
-visualizations where clusters are arbitrarily positioned in the embedding space.
+The final positioning of data points using this class of algorithms is largely
+dependent on the initial positions in the embedding space. UMAP is often said
+to better preserve global structure than t-SNE, but Kobak \& Linderman have
+recently discovered that this supposed advantage is entirely due to the better
+default initialization in UMAP than in older implementations of
+t-SNE~\cite{kobak2019umap}. Standard implementations of t-SNE use random
+initializations, which often scatter resulting clusters throughout the
+embedding space. UMAP injects global organization into the embedding by
+initializing point positions using Laplacian eigenmaps. When initialized in the
+same manner, both t-SNE and UMAP perform equally well, or equally poorly,
+dispelling the myth that UMAP is superior in this respect.
 
 \opentsne\ defaults to using the two leading principal components as
-initialization, but also provides a spectral approach similar to UMAP.
-This allows clusters in resulting embeddings to be organized in a more globally coherent
-manner, leading to increased visualization interpretability.
+initialization, but also provides a spectral approach, similar to that of UMAP.
+The clusters in resulting embeddings are organized in a globally coherent
+manner, leading to more interpretable visualizations.
 
@@
 
-\opentsne\ is distributed under the BSD-3-Clause License and is publicly
+openTSNE is distributed under the BSD-3-Clause License and is publicly
 available as an open-source package at
-\url{https://github.com/pavlin-policar/openTSNE}. \opentsne\ is also available on
-\texttt{PyPI} and \texttt{conda-forge}. The data sets used and scripts used in this study are
+\url{https://github.com/pavlin-policar/openTSNE}. openTSNE is also available on
+PyPI and conda-forge. The data sets used and scripts used in this study are
 included in accompanying notebooks, publicly available at

### 2020-10-01 paper/policar2020-opentsne.tex
Re-add header and fixup opentsne macros

--- 2020-10-01 paper/policar2020-opentsne.tex
@@
+%% BioMed_Central_Tex_Template_v1.06
+%%                                      %
+%  bmc_article.tex            ver: 1.06 %
+%                                       %
+
+%%IMPORTANT: do not delete the first line of this template
+%%It must be present to enable the BMC Submission system to
+%%recognise this template!!
+
+%%% additional documentclass options:
+%  [doublespacing]
+%  [linenumbers]   - put the line numbers on margins
+
 \documentclass[twocolumn]{bmcart}
@@
 	\includegraphics[width=\textwidth]{tasic2018}
-	\caption{\label{fig:tasic}We use openTSNE to create four different
+	\caption{\label{fig:tasic}We use \opentsne\ to create four different
     visualizations of the Tasic \textit{et al.}~\cite{tasic2018shared} data,
@@
   \caption{\label{fig:transform}
-\openTSNE supports embedding new samples into an existing reference t-SNE landscape.
+\opentsne\ supports embedding new samples into an existing reference t-SNE landscape.
 For the series of visualizations shown in this figure, we first construct a t-SNE embedding for the data from Hochgerner \textit{et al.}~\cite{hochgerner2018conserved} containing 24,185 developing, single cells from the mouse hippocampus.

### 2020-10-01 paper/policar2020-opentsne.tex

@@
 
-The abundance of high-dimensional data sets in molecular biology calls for dimensionality reduction techniques, particularly for methods that produce informative data visualizations.
-Popular approaches include
-principal component analysis (PCA), multidimensional scaling, t-distributed stochastic neighbor embedding (t-SNE)~\cite{maaten2008visualizing}, and uniform manifold approximation and projections (UMAP)~\cite{2018arXivUMAP}.
-Among these, t-SNE lately received much attention as it can address high volumes of data and reveal a meaningful clustering structure.
-Increased
-resolution and throughput of modern molecular assays has lead to the frequent use of t-SNE in diverse fields including, but not limited to, single-cell transcriptomics(scRNA-seq,~\cite{macosko2015highly,cao2019single,tasic2018shared}),
+The abundance of high-dimensional data sets in molecular biology calls for
+dimensionality reduction techniques, particularly for methods that produce
+informative data visualizations. Popular approaches include principal component
+analysis (PCA), multidimensional scaling, t-distributed stochastic neighbor
+embedding (t-SNE)~\cite{maaten2008visualizing}, and uniform manifold
+approximation and projections (UMAP)~\cite{2018arXivUMAP}. Among these, t-SNE
+lately received much attention as it can address high volumes of data and reveal
+a meaningful clustering structure. Increased resolution and throughput of modern
+molecular assays has lead to the frequent use of t-SNE in diverse fields
+including, but not limited to, single-cell
+transcriptomics(scRNA-seq,~\cite{macosko2015highly,cao2019single,tasic2018shared}),
 human genetics~\cite{hirata2019genetic}, metagenomic
-assembly~\cite{beaulaurier2018metagenomic}, the spatial organization of microbial communities~\cite{sheth2019spatial} and metabolomics~\cite{tkachev2019differences}.
-Reports on single-cell gene expression data, our running example, often start with an overview of the cell landscape, where t-SNE embeds high-dimensional expression profiles into a two-dimensional
-space. Figs.~\ref{fig:macosko}.a and \ref{fig:macosko}.b show two such embeddings.
+assembly~\cite{beaulaurier2018metagenomic}, the spatial organization of
+microbial communities~\cite{sheth2019spatial} and
+metabolomics~\cite{tkachev2019differences}. Reports on single-cell gene
+expression data, our running example, often start with an overview of the cell
+landscape, where t-SNE embeds high-dimensional expression profiles into a
+two-dimensional space. Figs.~\ref{fig:macosko}.a and \ref{fig:macosko}.b show
+two such embeddings.
 
@@
 \opentsne. The data in \textbf{(a)} and \textbf{(b)} represent 44,808
-single-cell gene-expression profiles of mouse retinal cells from Macosko \textit{et al.}~\cite{macosko2015highly}. The data in \textbf{(c)} contains 27,499 expression profiles from mouse
+single-cell gene-expression profiles of mouse retinal cells from Macosko
+\textit{et al.}~\cite{macosko2015highly}. The data in \textbf{(c)}
+contains 27,499 expression profiles from mouse
 retinal cells from Shekhar \textit{et
@@
 
-Despite its utility, t-SNE has often been criticized for its limited scalability, lack of global organization -- t-SNE emphasizes
-local clusters that are arbitrarily scattered in the low-dimensional space --
-and the absence of theoretically-founded methods to map new data into
-existing embeddings~\cite{ding2018interpretable,becht2019dimensionality}. Most
-of these shortcomings have recently been addressed. Linderman \textit{et al.}
-sped-up the method through an interpolation-based approximation, achieving linear complexity in the number of
-samples~\cite{linderman2019fast}. Kobak \& Berens proposed several techniques
-to improve global positioning, including estimating similarities with a mixture
-of Gaussian kernels~\cite{kobak2019art}. In our previous work, we presented a principled approach for embedding new samples into existing
-visualizations~\cite{policar2019embedding}.
+Despite its utility, t-SNE has often been criticized for its limited
+scalability, lack of global organization -- t-SNE emphasizes local clusters that
+are arbitrarily scattered in the low-dimensional space -- and the absence of
+theoretically-founded methods to map new data into existing
+embeddings~\cite{ding2018interpretable,becht2019dimensionality}. Most of these
+shortcomings have recently been addressed. Linderman \textit{et al.} sped-up the
+method through an interpolation-based approximation, achieving linear complexity
+in the number of samples~\cite{linderman2019fast}. Kobak \& Berens proposed
+several techniques to improve global positioning, including estimating
+similarities with a mixture of Gaussian kernels~\cite{kobak2019art}. In our
+previous work, we presented a principled approach for embedding new samples into
+existing visualizations~\cite{policar2019embedding}.
 
@@
 
-The \opentsne\ is compatible with the Python data science ecosystem and libraries that include
-\textsf{numpy}, \textsf{scikit-learn}, \textsf{scanpy}), thus providing a familiar and intuitive application program interface to the new users.
-Its modular design encourages extensibility and experimentation with various settings and changes in the analysis pipeline.
-We have released \opentsne\ in an open-source to support adaptations by the wider community.
-At the time of the writing, the library averages several thousands of weekly downloads and has received almost 700 GitHub stars, an appreciation score gained by only $0.3\%$ of public Python repositories.
-The source code is available at
+The \opentsne\ is compatible with the Python data science ecosystem and
+libraries that include \textsf{numpy}, \textsf{scikit-learn}, \textsf{scanpy}),
+thus providing a familiar and intuitive application program interface to the new
+users. Its modular design encourages extensibility and experimentation with
+various settings and changes in the analysis pipeline.  We have released
+\opentsne\ in an open-source to support adaptations by the wider community. At
+the time of the writing, the library averages several thousands of weekly
+downloads and has received almost 700 GitHub stars, an appreciation score gained
+by only $0.3\%$ of public Python repositories. The source code is available at
 \url{https://github.com/pavlin-policar/openTSNE} and supports installations from
@@
 
-Accessibility of the latest theoretical advances in t-SNE is one of the core design principles of \opentsne.
-These advances are accessible through a programming interface of the library, where we closely follow the style of the interface as implemented in
-\textsf{scikit-learn}~\cite{sklearn_api}, a popular library for machine learning.
-The following snippet exemplifies this interface and codes for four visualizations in Fig.~\ref{fig:tasic}, demonstrating the effect of a particular recent theoretical advance in the t-SNE algorithm.
+Accessibility of the latest theoretical advances in t-SNE is one of the core
+design principles of \opentsne. These advances are accessible through a
+programming interface of the library, where we closely follow the style of the
+interface as implemented in \textsf{scikit-learn}~\cite{sklearn_api}, a popular
+library for machine learning. The following snippet exemplifies this interface
+and codes for four visualizations in Fig.~\ref{fig:tasic}, demonstrating the
+effect of a particular recent theoretical advance in the t-SNE algorithm.
 
@@
 
-\noindent The code snippet considers a \textsf{numpy} array or
-\textsf{scipy} sparse matrix \texttt{X} and creates four
-\texttt{TSNEEmbedding} objects.
-These objects encode t-SNE embeddings that may subsequently be optimized under different parameter settings or used to embed new samples into the embedding landscape.
-Function calls may include additional parameters not shown in the example code but explained in accompanying notebooks available at \url{https://github.com/pavlin-policar/opentsne-paper}.
+\noindent The code snippet considers a \textsf{numpy} array or \textsf{scipy}
+sparse matrix \texttt{X} and creates four \texttt{TSNEEmbedding} objects. These
+objects encode t-SNE embeddings that may subsequently be optimized under
+different parameter settings or used to embed new samples into the embedding
+landscape. Function calls may include additional parameters not shown in the
+example code but explained in accompanying notebooks available at
+\url{https://github.com/pavlin-policar/opentsne-paper}.
 
@@
 
-Dimensionality reduction techniques implicitly assume that high-dimensional data lies on a lower-dimensional manifold, which can accurately be captured by a smaller number of dimensions.
-However, there is no evidence that every data set can accurately be described by using only two dimensions, and any such embedding will inevitably lead to a loss of information.
-Thus, it is beneficial to examine multiple embeddings, each of which provides a different perspective on topology and other data characteristics.
+Dimensionality reduction techniques implicitly assume that high-dimensional data
+lies on a lower-dimensional manifold, which can accurately be captured by a
+smaller number of dimensions. However, there is no evidence that every data set
+can accurately be described by using only two dimensions, and any such embedding
+will inevitably lead to a loss of information. Thus, it is beneficial to examine
+multiple embeddings, each of which provides a different perspective on topology
+and other data characteristics.
 
@@
 
-We illustrate this point by generating four different embeddings of the data on single-cell gene expression in mouse brain~\cite{tasic2018shared}.
-Fig.~\ref{fig:tasic}.a shows an embedding using default t-SNE parameters.
-While different clusters of excitatory and inhibitory neurons appear close to one another, all clusters appear equidistant from their neighbors, and the overall relations between groups are not obvious.
-Embedding in Fig.~\ref{fig:tasic}.b focuses on preserving larger neighborhoods of points, resulting in a more globally consistent layout where relations between clusters become more apparent.
-Here, it is evident from the increased white space between groups that there is one large class of excitatory neurons and two related classes of inhibitory neurons.
-Unfortunately, focusing on preserving large neighborhoods leads to the absorption of smaller clusters by the larger ones.
-Alternatively, Fig.~\ref{fig:tasic}.c uses multi-scale similarity kernels that aim to preserve both the global organization of clusters and prevent the cluster absorption.
-We constructed the embeddings from Fig.~\ref{fig:tasic}.d with the settings used for Fig.~\ref{fig:tasic}.a, but at a finer level of resolution.
-The distinction between these two depictions reveals that many clusters are composed of subgroups representing different cell subtypes.
+We illustrate this point by generating four different embeddings of the data on
+single-cell gene expression in mouse brain~\cite{tasic2018shared}.
+Fig.~\ref{fig:tasic}.a shows an embedding using default t-SNE parameters. While
+different clusters of excitatory and inhibitory neurons appear close to one
+another, all clusters appear equidistant from their neighbors, and the overall
+relations between groups are not obvious. Embedding in Fig.~\ref{fig:tasic}.b
+focuses on preserving larger neighborhoods of points, resulting in a more
+globally consistent layout where relations between clusters become more
+apparent. Here, it is evident from the increased white space between groups that
+there is one large class of excitatory neurons and two related classes of
+inhibitory neurons. Unfortunately, focusing on preserving large neighborhoods
+leads to the absorption of smaller clusters by the larger ones. Alternatively,
+Fig.~\ref{fig:tasic}.c uses multi-scale similarity kernels that aim to preserve
+both the global organization of clusters and prevent the cluster absorption. We
+constructed the embeddings from Fig.~\ref{fig:tasic}.d with the settings used
+for Fig.~\ref{fig:tasic}.a, but at a finer level of resolution. The distinction
+between these two depictions reveals that many clusters are composed of
+subgroups representing different cell subtypes.
 
-When dealing with data containing millions of data points, standard t-SNE embeddings often become unwieldy -- cluster boundaries are blurred, large clusters absorb smaller ones, and relationships between clusters become increasingly difficult to interpret.
-We constructed Fig.~\ref{fig:cao}.a from the data containing expression profiles of over two million single cells captured at different time points in mouse development.
-The embedding indicates numerous clusters with transitions between time points, as marked by the color-coding, that are difficult to interpret.
-Kobak \& Berens observed that increasing attractive forces between similar data points controlled via the \textit{exaggeration} parameter leads to more compact clusters, and subsequently, more informative visualizations~\cite{kobak2019art}.
-For instance, \ref{fig:cao}.b doubles the default exaggeration, which uncovers some of the data's overall structure.
-Further doubling the exaggeration in \ref{fig:cao}.c allows us to observe that the data is comprised of two main groups of cells and eight somewhat smaller clusters.
-The visualization also reveals several tiny clusters, possibly corresponding to
+When dealing with data containing millions of data points, standard t-SNE
+embeddings often become unwieldy -- cluster boundaries are blurred, large
+clusters absorb smaller ones, and relationships between clusters become
+increasingly difficult to interpret. We constructed Fig.~\ref{fig:cao}.a from
+the data containing expression profiles of over two million single cells
+captured at different time points in mouse development. The embedding indicates
+numerous clusters with transitions between time points, as marked by the
+color-coding, that are difficult to interpret. Kobak \& Berens observed that
+increasing attractive forces between similar data points controlled via the
+\textit{exaggeration} parameter leads to more compact clusters, and
+subsequently, more informative visualizations~\cite{kobak2019art}. For instance,
+\ref{fig:cao}.b doubles the default exaggeration, which uncovers some of the
+data's overall structure. Further doubling the exaggeration in \ref{fig:cao}.c
+allows us to observe that the data is comprised of two main groups of cells and
+eight somewhat smaller clusters. The visualization also reveals several tiny
+clusters, possibly corresponding to
 rare cell types.
@@
 Unlike other popular dimensionality reduction methods such as principal
-component analysis or autoencoders, t-SNE is a non-parametric method and does not define an explicit mapping to the embedding space.
-Therefore embeddings of new data points need to be found through optimization~\cite{policar2019embedding}.
-\opentsne\ is currently the only publicly available library allowing users to add new samples to existing embeddings.
+component analysis or autoencoders, t-SNE is a non-parametric method and does
+not define an explicit mapping to the embedding space. Therefore embeddings of
+new data points need to be found through
+optimization~\cite{policar2019embedding}. \opentsne\ is currently the only
+publicly available library allowing users to add new samples to existing
+embeddings.
 
-Figs.~\ref{fig:macosko}.b and \ref{fig:macosko}.c demonstrate how we can use a previously labeled single-cell data set and embed cells from a separate experiment into the reference landscape.
-The reference data from Macosko~\textit{et al.}~\cite{macosko2015highly} contains gene expression profiles from mouse retinal cells.
-By embedding the samples from a similar experiment on bipolar retinal cells by Shekhar~\textit{et al.}~\cite{shekhar2016comprehensive}, we can correctly map the bipolar cell clusters in the reference.
+Figs.~\ref{fig:macosko}.b and \ref{fig:macosko}.c demonstrate how we can use a
+previously labeled single-cell data set and embed cells from a separate
+experiment into the reference landscape. The reference data from
+Macosko~\textit{et al.}~\cite{macosko2015highly} contains gene expression
+profiles from mouse retinal cells. By embedding the samples from a similar
+experiment on bipolar retinal cells by Shekhar~\textit{et
+al.}~\cite{shekhar2016comprehensive}, we can correctly map the bipolar cell
+clusters in the reference.
 
-Embedding single cells into existing reference atlases can also be useful for cell-type classification in cases of unknown cell identities.
-For instance, in Fig.~\ref{fig:transform}, we construct a reference embedding using labeled data from Hochgerner~\textit{et al.}~\cite{hochgerner2018conserved} containing gene-expression profiles of cells from the mouse brain.
-The authors assign a type to each cell.
-We can verify their classification accuracy by visualizing the expression of well-established gene markers for the major cell types.
-We then embed cells from Harris~\textit{et al.}~\cite{harris2018classes} into a constructed atlas.
-In Harris~\textit{et al.}~, labels are provided only for neuronal cells.
-In the resulting mapping, we can quickly identify other non-neuronal cell types, including oligodendrocytes and astrocytes.
-We can further use marker genes to validate that the mapping in the reference landscape was correct.
+Embedding single cells into existing reference atlases can also be useful for
+cell-type classification in cases of unknown cell identities. For instance, in
+Fig.~\ref{fig:transform}, we construct a reference embedding using labeled data
+from Hochgerner~\textit{et al.}~\cite{hochgerner2018conserved} containing
+gene-expression profiles of cells from the mouse brain. The authors assign a
+type to each cell. We can verify their classification accuracy by visualizing
+the expression of well-established gene markers for the major cell types. We
+then embed cells from Harris~\textit{et al.}~\cite{harris2018classes} into a
+constructed atlas. In Harris~\textit{et al.}~, labels are provided only for
+neuronal cells. In the resulting mapping, we can quickly identify other
+non-neuronal cell types, including oligodendrocytes and astrocytes. We can
+further use marker genes to validate that the mapping in the reference landscape
+was correct.
 
@@
   \caption{\label{fig:transform}
-\opentsne\ supports embedding new samples into an existing reference t-SNE landscape.
-For the series of visualizations shown in this figure, we first construct a t-SNE embedding for the data from Hochgerner \textit{et al.}~\cite{hochgerner2018conserved} containing 24,185 developing, single cells from the mouse hippocampus.
-The data contains gene expression in different neurons, supporting glia, and other vascular cells (upper left).
-Data points representing cells are colored according to cell-types assigned in the original publication; see the legend from Fig.~\ref{fig:tasic} to map colors to cell-type.
-We then embed new, hippocampal cells collected in a study by Harris \textit{et al.}~\cite{harris2018classes} using the embedding of Hochgerner \textit{et al.} data as a reference.
-In their study, Harris \textit{et al.} collected 6,971 single-cells and focused on identifying different types of inhibitory neurons.
-However, almost half of the collected cells are not neurons and were left uncharacterized.
-Inspecting the embeddings of these cells in the reference embedding (bottom left) reveals that in addition to inhibitory neurons, the data contains several supporting glial cells as well as a small population of endothelial cells.
-We can verify our approach's accuracy by inspecting marker genes for the major cell types in the reference (top row) and embedded samples (bottom row).
+\opentsne\ supports embedding new samples into an existing reference t-SNE
+	landscape. For the series of visualizations shown in this figure, we
+	first construct a t-SNE embedding for the data from Hochgerner
+	\textit{et al.}~\cite{hochgerner2018conserved} containing 24,185
+	developing, single cells from the mouse hippocampus. The data contains
+	gene expression in different neurons, supporting glia, and other
+	vascular cells (upper left). Data points representing cells are colored
+	according to cell-types assigned in the original publication; see the
+	legend from Fig.~\ref{fig:tasic} to map colors to cell-type. We then
+	embed new, hippocampal cells collected in a study by Harris \textit{et
+	al.}~\cite{harris2018classes} using the embedding of Hochgerner
+	\textit{et al.} data as a reference. In their study, Harris \textit{et
+	al.} collected 6,971 single-cells and focused on identifying different
+	types of inhibitory neurons. However, almost half of the collected cells
+	are not neurons and were left uncharacterized. Inspecting the embeddings
+	of these cells in the reference embedding (bottom left) reveals that in
+	addition to inhibitory neurons, the data contains several supporting
+	glial cells as well as a small population of endothelial cells. We can
+	verify our approach's accuracy by inspecting marker genes for the major
+	cell types in the reference (top row) and embedded samples (bottom row).
 }
@@
 
-The examples presented above demonstrate how to use \opentsne\ to quickly gain insight into a newly-sequenced, single-cell data sets by utilizing existing cell atlases.
-The approach is general and not limited to single-cell gene expression, and one can, in principle, apply it to any tabular data set regardless of field.
+The examples presented above demonstrate how to use \opentsne\ to quickly gain
+insight into a newly-sequenced, single-cell data sets by utilizing existing cell
+atlases. The approach is general and not limited to single-cell gene expression,
+and one can, in principle, apply it to any tabular data set regardless of field.
 
@@
 
-Versatility, the ability to use and combine different optimization approaches to construct the embedding spaces, is another of \opentsne's core design principles.
-Kobak \& Berens recently introduced several recommendations and tricks to obtain better and more meaningful t-SNE visualizations~\cite{kobak2019art}.
-These include multi-scale similarity kernels, perplexity annealing, and increasing exaggeration when working with more massive data sets.
-\opentsne\ provides a flexible program interface to incorporate these improvements in just a few code lines.
-Furthermore, \opentsne\ supports custom affinity models, enabling users to construct t-SNE embeddings on non-tabular relational data: the only requirement imposed by the affinity-model is the availability of similarity between data points, that is, the distance matrix.
-Finally, the \opentsne's comprehensive callback system can be utilized to monitor and adapt different stages of the optimization phase and has been used to construct visually appealing animations of the t-SNE optimization process.
+Versatility, the ability to use and combine different optimization approaches to
+construct the embedding spaces, is another of \opentsne's core design
+principles. Kobak \& Berens recently introduced several recommendations and
+tricks to obtain better and more meaningful t-SNE
+visualizations~\cite{kobak2019art}. These include multi-scale similarity
+kernels, perplexity annealing, and increasing exaggeration when working with
+more massive data sets. \opentsne\ provides a flexible program interface to
+incorporate these improvements in just a few code lines. Furthermore, \opentsne\
+supports custom affinity models, enabling users to construct t-SNE embeddings on
+non-tabular relational data: the only requirement imposed by the affinity-model
+is the availability of similarity between data points, that is, the distance
+matrix.  Finally, the \opentsne's comprehensive callback system can be utilized
+to monitor and adapt different stages of the optimization phase and has been
+used to construct visually appealing animations of the t-SNE optimization
+process.
 
@@
 
-One of the t-SNE's common criticisms is limited scalability to large data sets containing, for instance, millions of data points~\cite{becht2019dimensionality}.
-The culprit for slow response time stems from a particular optimization procedure and its specific implementation in the popular Python library.
-Namely, until quite recently, most popular implementations of t-SNE were based on the Barnes-Hut approximation scheme developed by van der Maaten in 2014~\cite{van2014accelerating} with asymptotic complexity $\mathcal{O}(N \log N)$, where $N$ is the number of data items ({\em e.g.} cells).
-The most widely-used implementation of t-SNE came from \textsf{scikit-learn}, which exhibits long runtimes when compared to its C++ counterpart -- \textsf{MulticoreTSNE}.
-The multi-threaded \textsf{MulticoreTSNE} implementation~\cite{Ulyanov2016} can construct t-SNE embeddings of millions of data points in a matter of hours on widely-accessible, consumer-grade processors (Fig.~\ref{fig:benchmarks}).
-However, recently, Linderman \textit{et al.} developed a new approximation scheme -- FIt-SNE -- which further reduces the asymptotic time complexity to $\mathcal{O}(N)$.
-We have included this scheme in \opentsne\, enabling the embedding of large data sets in a matter of minutes.
+One of the t-SNE's common criticisms is limited scalability to large data sets
+containing, for instance, millions of data
+points~\cite{becht2019dimensionality}. The culprit for slow response time stems
+from a particular optimization procedure and its specific implementation in the
+popular Python library. Namely, until quite recently, most popular
+implementations of t-SNE were based on the Barnes-Hut approximation scheme
+developed by van der Maaten in 2014~\cite{van2014accelerating} with asymptotic
+complexity $\mathcal{O}(N \log N)$, where $N$ is the number of data items ({\em
+e.g.} cells). The most widely-used implementation of t-SNE came from
+\textsf{scikit-learn}, which exhibits long runtimes when compared to its C++
+counterpart -- \textsf{MulticoreTSNE}. The multi-threaded \textsf{MulticoreTSNE}
+implementation~\cite{Ulyanov2016} can construct t-SNE embeddings of millions of
+data points in a matter of hours on widely-accessible, consumer-grade processors
+(Fig.~\ref{fig:benchmarks}). However, recently, Linderman \textit{et al.}
+developed a new approximation scheme -- FIt-SNE -- which further reduces the
+asymptotic time complexity to $\mathcal{O}(N)$. We have included this scheme in
+\opentsne\, enabling the embedding of large data sets in a matter of minutes.
 
-Fig.~\ref{fig:benchmarks} benchmarks four Python t-SNE implementations, including those from \textsf{scikit-learn} (v0.23.1), \textsf{MulticoreTSNE} (v0.1), \textsf{FIt-SNE} (v1.1.0), and our \opentsne\ (v0.4.3).
-We perform benchmarks on two computational platforms, one representing the usage on a personal computer and the other utility of these libraries on a high-performance computing platform.
-The Intel(R) Core i7 is commonly found in consumer-grade laptop computers, while Intel(R) Xeon(R) processors appear in high-performance computing machines.
-Benchmarks were run for $1,000$ iterations with the original t-SNE parameters, as some implementations do not allow their modification.
+Fig.~\ref{fig:benchmarks} benchmarks four Python t-SNE implementations,
+including those from \textsf{scikit-learn} (v0.23.1), \textsf{MulticoreTSNE}
+(v0.1), \textsf{FIt-SNE} (v1.1.0), and our \opentsne\ (v0.4.3). We perform
+benchmarks on two computational platforms, one representing the usage on a
+personal computer and the other utility of these libraries on a high-performance
+computing platform. The Intel(R) Core i7 is commonly found in consumer-grade
+laptop computers, while Intel(R) Xeon(R) processors appear in high-performance
+computing machines. Benchmarks were run for $1,000$ iterations with the original
+t-SNE parameters, as some implementations do not allow their modification.
 
@@
   \caption{\label{fig:benchmarks}
-We benchmark openTSNE (v0.4.3) against three popular open-source implementations from scikit-learn~\cite{pedregosa2011scikit} (v0.23.1), MulticoreTSNE~\cite{Ulyanov2016} (v0.1), and FIt-SNE~\cite{linderman2019fast} (v1.1.0).
-Experiments were run on a consumer-grade Intel Core i7-7700HQ processor found in laptop computers, and on a server-grade Intel Xeon E5-2650.
-To generate benchmark data sets of different sizes, we subsampled data from the 10X Genomics 1.3 million mouse brain data set five times, resulting in five different data sets for each size.
-In total, we run each implementation on 30 different data sets.
-Notice that openTSNE scales similarly to FIt-SNE, as they both use the same interpolation-based approximation scheme, while scikit-learn and MulticoreTSNE utilize the Barnes-Hut approximation.
+We benchmark openTSNE (v0.4.3) against three popular open-source implementations
+	from scikit-learn~\cite{pedregosa2011scikit} (v0.23.1),
+	MulticoreTSNE~\cite{Ulyanov2016} (v0.1), and
+	FIt-SNE~\cite{linderman2019fast} (v1.1.0). Experiments were run on a
+	consumer-grade Intel Core i7-7700HQ processor found in laptop computers,
+	and on a server-grade Intel Xeon E5-2650. To generate benchmark data
+	sets of different sizes, we subsampled data from the 10X Genomics 1.3
+	million mouse brain data set five times, resulting in five different
+	data sets for each size. In total, we run each implementation on 30
+	different data sets. Notice that openTSNE scales similarly to FIt-SNE,
+	as they both use the same interpolation-based approximation scheme,
+	while scikit-learn and MulticoreTSNE utilize the Barnes-Hut
+	approximation.
 }
@@
 
-Benchmark results (Fig.~\ref{fig:benchmarks}) confirm that both \textsf{FIt-SNE} and \opentsne\ scale better than their Barnes-Hut counterparts -- \textsf{scikit-learn} and \textsf{MulticoreTSNE}.
-While \opentsne\'s implementation uses Python that incurs some runtime overhead compared to its C++ counterpart \textsf{FIt-SNE}, the two libraries' speeds are surprisingly comparable.
-Modern computer processors contain multiple cores, allowing us to use multi-threading, which further reduces the gap between \opentsne\ and \textsf{FIt-SNE}.
-On the server-grade processor, \opentsne\ is slightly faster than its pure C++ counterpart when utilizing multiple cores.
-\opentsne\ uses \textsf{numpy} for most linear algebra operations, which may make better use of the Intel Math Kernel Library (MKL), which is aggressively optimized on Intel(R) Xeon(R) processors.
+Benchmark results (Fig.~\ref{fig:benchmarks}) confirm that both \textsf{FIt-SNE}
+and \opentsne\ scale better than their Barnes-Hut counterparts --
+\textsf{scikit-learn} and \textsf{MulticoreTSNE}. While \opentsne\'s
+implementation uses Python that incurs some runtime overhead compared to its C++
+counterpart \textsf{FIt-SNE}, the two libraries' speeds are surprisingly
+comparable. Modern computer processors contain multiple cores, allowing us to
+use multi-threading, which further reduces the gap between \opentsne\ and
+\textsf{FIt-SNE}. On the server-grade processor, \opentsne\ is slightly faster
+than its pure C++ counterpart when utilizing multiple cores. \opentsne\ uses
+\textsf{numpy} for most linear algebra operations, which may make better use of
+the Intel Math Kernel Library (MKL), which is aggressively optimized on Intel(R)
+Xeon(R) processors.
 
-\opentsne\ provides a flexible API, allowing splitting up the embedding-construction process into several parts, caching slow operations.
-Running t-SNE optimization in stages enables users to quickly experiment with different parameter settings and iterate on their final visualizations.
+\opentsne\ provides a flexible API, allowing splitting up the
+embedding-construction process into several parts, caching slow operations.
+Running t-SNE optimization in stages enables users to quickly experiment with
+different parameter settings and iterate on their final visualizations.
 
@@
 
-Intuitive access and simple installation procedures almost universally correlate with the widespread adoption of novel computational techniques.
-While the t-SNE from \textsf{scikit-learn} fits this requirement, its implementation is prohibitively slow for even moderately-sized data sets that span tens of thousands of data records.
-Other C++ implementations such as \textsf{MulticoreTSNE} and \textsf{FIt-SNE} exhibit better scaling in more massive data sets, but do not provide precompiled binaries and require users to compile the software themselves.
-This problem is critical, for instance, for users of Windows, where the C++ compiler does not come with the system, making the correct configuration of current t-SNE implementations cumbersome.
+Intuitive access and simple installation procedures almost universally correlate
+with the widespread adoption of novel computational techniques. While the t-SNE
+from \textsf{scikit-learn} fits this requirement, its implementation is
+prohibitively slow for even moderately-sized data sets that span tens of
+thousands of data records. Other C++ implementations such as
+\textsf{MulticoreTSNE} and \textsf{FIt-SNE} exhibit better scaling in more
+massive data sets, but do not provide precompiled binaries and require users to
+compile the software themselves. This problem is critical, for instance, for
+users of Windows, where the C++ compiler does not come with the system, making
+the correct configuration of current t-SNE implementations cumbersome.
 
-We have designed \opentsne\ to make it accessible to a broader audience.
-We provide precompiled binaries for all major Python versions on all major platforms, making the installation process as seamless as possible.
-One can install \opentsne\ through the Python Package Index (PyPI) or conda on the conda-forge channel.
-\opentsne\ interface follows that of \textsf{scikit-learn}, which is well established in the Python data science ecosystem.
-\opentsne\ implements multi-threaded versions of both the Barnes-Hut and FIt-SNE approximation schemes, enabling it for data sets containing millions of data points.
-While the Python virtual machine introduces some performance overhead, the runtime is comparable to its C++ counterpart -- \textsf{FIt-SNE}.
-Finally, \opentsne\ is extensible.
-Its modular design enables researchers to quickly experiment with different settings and easily incorporate their components into the software.
-We provide a full feature list and comparison to other popular t-SNE implementations in Table~\ref{tab:features}.
+We have designed \opentsne\ to make it accessible to a broader audience. We
+provide precompiled binaries for all major Python versions on all major
+platforms, making the installation process as seamless as possible. One can
+install \opentsne\ through the Python Package Index (PyPI) or conda on the
+conda-forge channel. \opentsne\ interface follows that of \textsf{scikit-learn},
+which is well established in the Python data science ecosystem. \opentsne\
+implements multi-threaded versions of both the Barnes-Hut and FIt-SNE
+approximation schemes, enabling it for data sets containing millions of data
+points. While the Python virtual machine introduces some performance overhead,
+the runtime is comparable to its C++ counterpart -- \textsf{FIt-SNE}. Finally,
+\opentsne\ is extensible. Its modular design enables researchers to quickly
+experiment with different settings and easily incorporate their components into
+the software. We provide a full feature list and comparison to other popular
+t-SNE implementations in Table~\ref{tab:features}.
 
@@
 \caption{\label{tab:features}
-We compare features of \opentsne to three popular open-source implementations from scikit-learn~(v0.23.1), MulticoreTSNE (v0.1), and FIt-SNE (v1.1.0).
-The first section of the comparison addresses packaging and distribution.
-A properly packaged library is easily accessible to users, and developers should easily include it in dependency lists of other software packages.
-The second section of the comparison lists the two existing t-SNE approximation schemes.
-The FIt-SNE approximation scheme is required for t-SNE to scale up to millions of data points.
-The final section provides a list of extensions and improvements to the standard t-SNE algorithm, many of which can produce markedly better visualizations.
+We compare features of \opentsne to three popular open-source implementations
+	from scikit-learn~(v0.23.1), MulticoreTSNE (v0.1), and FIt-SNE (v1.1.0).
+	The first section of the comparison addresses packaging and
+	distribution. A properly packaged library is easily accessible to users,
+	and developers should easily include it in dependency lists of other
+	software packages. The second section of the comparison lists the two
+	existing t-SNE approximation schemes. The FIt-SNE approximation scheme
+	is required for t-SNE to scale up to millions of data points. The final
+	section provides a list of extensions and improvements to the standard
+	t-SNE algorithm, many of which can produce markedly better
+	visualizations.
 }
@@
 phase, the attractive forces between data points are increased by some factor
-$\rho$, typically set to $12$, so that points in the embedding can more easily move throughout the space and find
-their respective neighbors. The remaining 750 iterations are run with $\rho=1$,
-which reverts the attractive forces to their original values and produces the
-final embedding.
+$\rho$, typically set to $12$, so that points in the embedding can more easily
+move throughout the space and find their respective neighbors. The remaining 750
+iterations are run with $\rho=1$, which reverts the attractive forces to their
+original values and produces the final embedding.
 
 Belkina \textit{et al.} later found that faster convergence can be achieved by
-increasing the learning rate to $\eta=N/12$~\cite{belkina2019automated}. As a side-effect, embeddings converge faster, and the
-number of iterations can be lowered to 750, decreasing the overall runtime.
-This convention has been adopted by most modern t-SNE implementations.
+increasing the learning rate to $\eta=N/12$~\cite{belkina2019automated}. As a
+side-effect, embeddings converge faster, and the number of iterations can be
+lowered to 750, decreasing the overall runtime.  This convention has been
+adopted by most modern t-SNE implementations.
 
@@
 impractical to any reasonably-sized data set and beckons for the development of
-efficient approximation schemes. Van der Maaten observed that the t-SNE
-gradient may be cast as an N-body problem where data points represent
-particles which attract and repel each
-other~\cite{van2014accelerating}. The gradient from Eqn. (\ref{eq:tsne_gradient}) can be rewritten as
+efficient approximation schemes. Van der Maaten observed that the t-SNE gradient
+may be cast as an N-body problem where data points represent particles which
+attract and repel each other~\cite{van2014accelerating}. The gradient from Eqn.
+(\ref{eq:tsne_gradient}) can be rewritten as
 \begin{equation}
@@
 \end{equation}
-where $Z = \sum_{k \neq l}\left ( 1 + || \mathbf{y}_k - \mathbf{y}_l ||^2
-\right )^{-1}$. This equation can then be viewed as a particle simulation, where 
-the two terms represent the attractive and repulsive
-forces between individual particles. Each term lends itself to efficient
-approximations, which enable us to greatly reduce the time complexity of t-SNE.
+where $Z = \sum_{k \neq l}\left ( 1 + || \mathbf{y}_k - \mathbf{y}_l ||^2 \right
+)^{-1}$. This equation can then be viewed as a particle simulation, where the
+two terms represent the attractive and repulsive forces between individual
+particles. Each term lends itself to efficient approximations, which enable us
+to greatly reduce the time complexity of t-SNE.
 
@@
 of data points is excessive, and that considering only a handful of nearest
-neighbors at each point is sufficient to obtain a good approximation~\cite{van2014accelerating}.
-Because distances are transformed to similarities using a Gaussian kernel, and
-the bandwidth of each kernel is selected such that only a predefined number of
-neighbors fall within the main probability mass of the bell curve through the perplexity parameter, the remaining
-data points fall into the exponentially decaying tails of each Gaussian distribution.
-These data points are assigned near-zero probabilities and do not contribute
-to the overall attractive forces of data points.
-Therefore it is sufficient to calculate the attractive forces for only a
-small number of nearest neighbors instead of all $N$ points. By
-utilizing tree-based nearest-neighbor search methods, the time complexity is
-thus reduced to $\mathcal{O}(N \log N)$. Linderman \textit{et al.} further
-realized that, qualitatively, embeddings are visually
-indistinguishable when using only \textit{approximate} nearest neighbors,
-further reducing time complexity to $\mathcal{O}(N)$~\cite{linderman2019fast}.
+neighbors at each point is sufficient to obtain a good
+approximation~\cite{van2014accelerating}.  Because distances are transformed to
+similarities using a Gaussian kernel, and the bandwidth of each kernel is
+selected such that only a predefined number of neighbors fall within the main
+probability mass of the bell curve through the perplexity parameter, the
+remaining data points fall into the exponentially decaying tails of each
+Gaussian distribution.  These data points are assigned near-zero probabilities
+and do not contribute to the overall attractive forces of data points.
+Therefore it is sufficient to calculate the attractive forces for only a small
+number of nearest neighbors instead of all $N$ points. By utilizing tree-based
+nearest-neighbor search methods, the time complexity is thus reduced to
+$\mathcal{O}(N \log N)$. Linderman \textit{et al.} further realized that,
+qualitatively, embeddings are visually indistinguishable when using only
+\textit{approximate} nearest neighbors, further reducing time complexity to
+$\mathcal{O}(N)$~\cite{linderman2019fast}.
 
@@
 space-partitioning Barnes-Hut tree approach to approximate the interaction
-between data points~\cite{van2014accelerating}. Briefly, in the 2D case, the approach splits
-the space into quadrants and entire regions may be summarized by simple statistics.
-If a query point is far away from a given quadrant, the repulsive forces exerted
-by all the points in that quadrant onto the query point are summarized by a single point. This
-reduces the time complexity from $\mathcal{O}(N^2)$ to
-$\mathcal{O}(N \log N)$.
+between data points~\cite{van2014accelerating}. Briefly, in the 2D case, the
+approach splits the space into quadrants and entire regions may be summarized by
+simple statistics.  If a query point is far away from a given quadrant, the
+repulsive forces exerted by all the points in that quadrant onto the query point
+are summarized by a single point. This reduces the time complexity from
+$\mathcal{O}(N^2)$ to $\mathcal{O}(N \log N)$.
 
@@
 interactions between repelling data points~\cite{linderman2019fast}. Briefly,
-Linderman \textit{et al.} observed that the repulsive forces $\mathbf{R}$ from Eqn. (\ref{eq:grad_attr_rep}) may
-be rewritten as \begin{align}
+Linderman \textit{et al.} observed that the repulsive forces $\mathbf{R}$ from
+Eqn. (\ref{eq:grad_attr_rep}) may be rewritten as \begin{align}
 \mathbf{R}_i &= \sum_{j \neq i} q_{ij}^2 Z \left ( \mathbf{y}_i - \mathbf{y}_j \right ) \notag \\
@@
 techniques~\cite{policar2019embedding}. When adding new data points to an
-existing, reference embedding, the reference data points are fixed in place while
-new data points are allowed to find their respective positions. The
+existing, reference embedding, the reference data points are fixed in place
+while new data points are allowed to find their respective positions. The
 optimization remains the same as in standard t-SNE with only slight
@@
 \mathbf{w}_M \right \} \in \mathbb{R}^d$. Additionally, we omit the
-symmetrization step in Eqn.~(\ref{eq:symmetrize}).
-Plugging these terms into Eqn.~(\ref{eq:kl_divergence}), we obtain the following
-gradient
+symmetrization step in Eqn.~(\ref{eq:symmetrize}). Plugging these terms into
+Eqn.~(\ref{eq:kl_divergence}), we obtain the following gradient
 \begin{equation}
@@
 
-Similarly to standard t-SNE, a direct calculation of gradients takes $\mathcal{O}(N
-\cdot M)$ time, but it is straightforward to adapt the Barnes-Hut and FIt-SNE
-approximation schemes, reducing the time complexity to $\mathcal{O}(M \log N)$
-and $\mathcal{O}(\max \{ N, M \})$, respectively. In the FIt-SNE approximation
-scheme, we additionally exploit the fact that the reference embedding remains
-fixed throughout the optimization of newly added points, and precompute the
-interpolation grid. This further reduces the runtime complexity from
-$\mathcal{O}(\max \{ N, M \})$ to $\mathcal{O}(M)$.
+Similarly to standard t-SNE, a direct calculation of gradients takes
+$\mathcal{O}(N \cdot M)$ time, but it is straightforward to adapt the Barnes-Hut
+and FIt-SNE approximation schemes, reducing the time complexity to
+$\mathcal{O}(M \log N)$ and $\mathcal{O}(\max \{ N, M \})$, respectively. In the
+FIt-SNE approximation scheme, we additionally exploit the fact that the
+reference embedding remains fixed throughout the optimization of newly added
+points, and precompute the interpolation grid. This further reduces the runtime
+complexity from $\mathcal{O}(\max \{ N, M \})$ to $\mathcal{O}(M)$.
 
@@
 In standard t-SNE, distances are converted to similarities through the use of
-Gaussian kernels of varying bandwidths. The bandwidths are indirectly
-determined by the user-specified perplexity parameter so that a fixed number of
-nearest data points will be assigned non-zero values. One common trick for
-uncovering the global relations between clusters is to increase perplexity so
-that more long-range interactions are preserved in the final embedding.
-However, one unfortunate side effect of increasing perplexity is that smaller
-clusters get absorbed into larger ones.
+Gaussian kernels of varying bandwidths. The bandwidths are indirectly determined
+by the user-specified perplexity parameter so that a fixed number of nearest
+data points will be assigned non-zero values. One common trick for uncovering
+the global relations between clusters is to increase perplexity so that more
+long-range interactions are preserved in the final embedding.  However, one
+unfortunate side effect of increasing perplexity is that smaller clusters get
+absorbed into larger ones.
 
-Kobak \& Berens suggest that replacing the Gaussian kernel
-with a mixture of Gaussians may provide better insight into both the local and
-global structure~\cite{kobak2019art}. For instance, the similarities between data points in the input
-space may instead be computed with
+Kobak \& Berens suggest that replacing the Gaussian kernel with a mixture of
+Gaussians may provide better insight into both the local and global
+structure~\cite{kobak2019art}. For instance, the similarities between data
+points in the input space may instead be computed with
 
@@
 The bandwidth of each Gaussian $\sigma_{1,i}$ and $\sigma_{2,i}$ is determined
-by selecting different perplexity values. Using a kernel with
-perplexity 500 captures long-range interactions which preserve global
-cluster organization. Combining these with a more narrow kernel with
-perplexity 50 prevents small, well-defined clusters from being absorbed
-into larger ones, leading a better overall snapshot into the data structure.
+by selecting different perplexity values. Using a kernel with perplexity 500
+captures long-range interactions which preserve global cluster organization.
+Combining these with a more narrow kernel with perplexity 50 prevents small,
+well-defined clusters from being absorbed into larger ones, leading a better
+overall snapshot into the data structure.
 
@@
 
-Standard t-SNE reveals the clustering structure at a single level of
-resolution. Different perplexity parameter values can be used to identify
-global cluster relationships or small, well-isolated groups. Unfortunately,
-varying perplexity values can be time-consuming as this involves recomputing
-the $k$-nearest neighbor graph which is often the most expensive part of the t-SNE algorithm.
-Alternatively, Kobak \textit{et al.} suggest that varying
-the degree of freedom in the t-distribution can be used to explore the clustering structure at different
-levels of resolution~\cite{kobak2019heavy}.
+Standard t-SNE reveals the clustering structure at a single level of resolution.
+Different perplexity parameter values can be used to identify global cluster
+relationships or small, well-isolated groups. Unfortunately, varying perplexity
+values can be time-consuming as this involves recomputing the $k$-nearest
+neighbor graph which is often the most expensive part of the t-SNE algorithm.
+Alternatively, Kobak \textit{et al.} suggest that varying the degree of freedom
+in the t-distribution can be used to explore the clustering structure at
+different levels of resolution~\cite{kobak2019heavy}.
 
 Standard t-SNE models similarities between data points in the embedding space
-using a t-distribution with a single degree of freedom, but this can be generalized
-to allow for any parameter value
+using a t-distribution with a single degree of freedom, but this can be
+generalized to allow for any parameter value
 
@@
 Adapting existing approximation schemes to this formulation is straightforward.
-\opentsne\ provides efficient implementations of both the Barnes-Hut and the FIt-SNE
-approximation schemes, where we modify the terms from
+\opentsne\ provides efficient implementations of both the Barnes-Hut and the
+FIt-SNE approximation schemes, where we modify the terms from
 Eqn.~(\ref{eq:fitsne_terms}) to
@@
 Embeddings produced by standard t-SNE often produce clusters separated by thin
-boundaries and make use of all available space. While this is desirable at times,
-it often obscures the global relationships between clusters as all
+boundaries and make use of all available space. While this is desirable at
+times, it often obscures the global relationships between clusters as all
 neighboring clusters appear about the same distance from one another. Other
@@
 
-B{\"o}hm \textit{et al.} showed that the exaggeration factor
-$\rho$ can be used to produce layouts more similar to UMAP and ForceAtlas2~\cite{bohm2020unifying}.
+B{\"o}hm \textit{et al.} showed that the exaggeration factor $\rho$ can be used
+to produce layouts more similar to UMAP and ForceAtlas2~\cite{bohm2020unifying}.
 By incorporating exaggeration into later phases of optimization, t-SNE is able
 introduce more white-space between clusters which better reflects the global
-relations between clusters. B{\"o}hm \textit{et al.} found that using $\rho=4$ and $\rho=30$
-produces embeddings visually similar to UMAP and ForceAtlas2, respectively.
+relations between clusters. B{\"o}hm \textit{et al.} found that using $\rho=4$
+and $\rho=30$ produces embeddings visually similar to UMAP and ForceAtlas2,
+respectively.
 
 While it is difficult to claim one is better than the other, different settings
-of $\rho$ may uncover different properties of the data manifold. For
-example, in single-cell data, standard t-SNE often uncovers
-distinct, well-separated groups of cell-types but obscures transitional paths between cell-states. This problem is
-exacerbated when dealing with large numbers of data points. On the other hand,
-ForceAtlas2 has been used successfuly to uncover trajectories and transitions between
-cell types. Unfortunately, when using ForceAtlas2, large clusters often absorb small, distinct groups of cells.
-One might run t-SNE with various parameter settings to avoid switching between
-different algorithms and obtain similar visualizations.
-We might generate an embedding using higher levels of exaggeration to highlight
-the developmental transitions between cell types and
-one with low levels of exaggeration to identify clear populations of cells.
+of $\rho$ may uncover different properties of the data manifold. For example, in
+single-cell data, standard t-SNE often uncovers distinct, well-separated groups
+of cell-types but obscures transitional paths between cell-states. This problem
+is exacerbated when dealing with large numbers of data points. On the other
+hand, ForceAtlas2 has been used successfuly to uncover trajectories and
+transitions between cell types. Unfortunately, when using ForceAtlas2, large
+clusters often absorb small, distinct groups of cells.  One might run t-SNE with
+various parameter settings to avoid switching between different algorithms and
+obtain similar visualizations.  We might generate an embedding using higher
+levels of exaggeration to highlight the developmental transitions between cell
+types and one with low levels of exaggeration to identify clear populations of
+cells.
 
@@
 
-t-SNE, UMAP, and ForceAtlas2 can all be cast as force-directed layout algorithms,
-which operate on the $k$-nearest neighbor graph. Each method constructs the graph differently
-and specifies the attractive and repulsive forces in its own way, but ultimately,
-point positions are found by balancing the attractive and repulsive forces between data points.
+t-SNE, UMAP, and ForceAtlas2 can all be cast as force-directed layout
+algorithms, which operate on the $k$-nearest neighbor graph. Each method
+constructs the graph differently and specifies the attractive and repulsive
+forces in its own way, but ultimately, point positions are found by balancing
+the attractive and repulsive forces between data points.
 
@@
 \opentsne\ defaults to using the two leading principal components as
-initialization, but also provides a spectral approach similar to UMAP.
-This allows clusters in resulting embeddings to be organized in a more globally coherent
-manner, leading to increased visualization interpretability.
+initialization, but also provides a spectral approach similar to UMAP.  This
+allows clusters in resulting embeddings to be organized in a more globally
+coherent manner, leading to increased visualization interpretability.
 
@@
 available as an open-source package at
-\url{https://github.com/pavlin-policar/openTSNE}. \opentsne\ is also available on
-\texttt{PyPI} and \texttt{conda-forge}. The data sets used and scripts used in this study are
-included in accompanying notebooks, publicly available at
+\url{https://github.com/pavlin-policar/openTSNE}. \opentsne\ is also available
+on \texttt{PyPI} and \texttt{conda-forge}. The data sets used and scripts used
+in this study are included in accompanying notebooks, publicly available at
 \url{https://github.com/pavlin-policar/opentsne-paper}.
@@
 \section*{Author's contributions}
-P.G.P.
-implemented and maintains \opentsne library.
-M.S.
-helped in development of mathematical foundations and advised on inclusion of the extensions.
-B.Z.
-initiated and supervised the project.
-All three co-authors wrote the manuscript.
+P.G.P. implemented and maintains \opentsne library. M.S. helped in development
+of mathematical foundations and advised on inclusion of the extensions.
+B.Z. initiated and supervised the project. All three co-authors wrote
+the manuscript.
 
 \section*{Acknowledgements}
-We want to thank Dmitry Kobak for various helpful discussions and best practices when using t-SNE and his contributions to the source code.
-We would also like to thank George Linderman for his help with the FIt-SNE algorithm.
+We want to thank Dmitry Kobak for various helpful discussions and best practices
+when using t-SNE and his contributions to the source code. We would also
+like to thank George Linderman for his help with the FIt-SNE algorithm.
 

### 2020-10-01 paper/policar2020-opentsne.tex
Rewrap tex file

@@
-%% BioMed_Central_Tex_Template_v1.06
-%%                                      %
-%  bmc_article.tex            ver: 1.06 %
-%                                       %
-
-%%IMPORTANT: do not delete the first line of this template
-%%It must be present to enable the BMC Submission system to
-%%recognise this template!!
-
-%%% additional documentclass options:
-%  [doublespacing]
-%  [linenumbers]   - put the line numbers on margins
-
-\documentclass[twocolumn]{bmcart}
-%\documentclass{bmcart}
+\documentclass[twocolumn]{article}
 
@@
 \usepackage{booktabs}
-\startlocaldefs
+
 \newcommand{\opentsne}{\textsf{openTSNE}}
-\endlocaldefs
 
@@
 
-%%% Start of article front matter
-\begin{frontmatter}
-
-\begin{fmbox}
-\dochead{Software}
-
 \title{openTSNE: a modular Python library for t-SNE dimensionality reduction and embedding}
 
-\author[
-   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
-   corref={aff1},                       % id of corresponding address, if any
-   %noteref={n1},                        % id's of article notes, if any
-   email={pavlin.policar@fri.uni-lj.si}   % email address
-]{\inits{PGP}\fnm{Pavlin G.} \snm{Poli\v{c}ar}}
-\author[
-   addressref={aff1,aff2}
-]{\inits{MS}\fnm{Martin} \snm{Stra\v{z}ar}}
-\author[
-   addressref={aff1,aff3}
-]{\inits{BZ}\fnm{Bla\v{z}} \snm{Zupan}}
-
-\address[id=aff1]{%
-  \orgname{Faculty of Computer and Information Science, University of Ljubljana},
-  %\street{Ve\v{c}na Pot 113},
-  \postcode{SI 1000}
-  \city{Ljubljana},
-  \cny{Slovenia}
-}
-\address[id=aff2]{%
-  \orgname{Broad Institute of Harvard and MIT},
-  %\street{D\"{u}sternbrooker Weg 20},
-  \postcode{MA 02142}
-  \city{Cambridge},
-  \cny{U.S.A.}
-}
-\address[id=aff3]{%
-  \orgname{Department of Molecular and Human Genetics, Baylor College of Medicine},
-  %\street{D\"{u}sternbrooker Weg 20},
-  \postcode{TX 77030}
-  \city{Houston},
-  \cny{U.S.A.}
+\author{
+  Pavlin G. Poli\v{c}ar\textsuperscript{1}\\
+  \texttt{pavlin.policar@fri.uni-lj.si}\\
+  \and
+  Martin Stra\v{zar}\textsuperscript{2}\\
+  \and
+  Bla\v{z} Zupan\textsuperscript{1,3}\\
 }
 
-\end{fmbox}% comment this for two column layout
+\date{
+  \textsuperscript{1}Faculty of Computer and Information Science, University of Ljubljana,\\
+  Ve\v{c}na pot 113, SI-1000 Ljubljana, Slovenia\\ \vspace{2mm}
+  \textsuperscript{2}Broad Institute of Harvard and MIT,\\
+  D\"{u}sternbrooker Weg 20, MA-02142 Cambridge, U.S.A.\\ \vspace{2mm}
+  \textsuperscript{3}Department of Molecular and Human Genetics, Baylor College of Medicine,\\
+  1 Baylor Plaza, TX-77030 Houston, U.S.A.
+}
 
-\begin{abstractbox}
+\maketitle
+
 \begin{abstract}
@@
 popular implementations, including those from scikit-learn. Unique to
-\opentsne\ is also mapping new data to existing embeddings,
+\opentsne\ is also the ability to map new data to existing embeddings,
 which can surprisingly assist in solving batch effects.
@@
 
-\begin{keyword}
-\kwd{t-SNE}
-\kwd{embedding}
-\kwd{visualization}
-\kwd{dimensionality reduction}
-\kwd{Python library}
-\end{keyword}
+\providecommand{\keywords}[1]{\vspace{2mm}\textbf{Keywords ---} #1}
+\begin{keywords}
+t-SNE, embedding, visualization, dimensionality reduction, Python library
+\end{keywords}
 
-% MSC classifications codes, if any
-%\begin{keyword}[class=AMS]
-%\kwd[Primary ]{}
-%\kwd{}
-%\kwd[; secondary ]{}
-%\end{keyword}
-
-\end{abstractbox}
-%
-%\end{fmbox}% uncomment this for twcolumn layout
-
-\end{frontmatter}
-
-\section*{Background}
+\section*{Introduction}
 
@@
 including, but not limited to, single-cell
-transcriptomics(scRNA-seq,~\cite{macosko2015highly,cao2019single,tasic2018shared}),
+transcriptomics (scRNA-seq,~\cite{macosko2015highly,cao2019single,tasic2018shared}),
 human genetics~\cite{hirata2019genetic}, metagenomic
@@
 \textit{et al.}~\cite{macosko2015highly}. The data in \textbf{(c)}
-contains 27,499 expression profiles from mouse
+additionally contains 27,499 expression profiles from mouse
 retinal cells from Shekhar \textit{et
@@
 localized to the same regions of the space, indicating a higher level
-of similarity than to other cell types. Embedding in \textbf{(c)} shows how existing
+of similarity than to other cell types. The embedding in \textbf{(c)} shows how existing
 t-SNE reference atlases can be used to place new samples into existing
@@
 
-The \opentsne\ is compatible with the Python data science ecosystem and
+\opentsne\ is compatible with the Python data science ecosystem and
 libraries that include \textsf{numpy}, \textsf{scikit-learn}, \textsf{scanpy}),
@@
 \opentsne\ in an open-source to support adaptations by the wider community. At
-the time of the writing, the library averages several thousands of weekly
+the time of the writing, the library averages thousands of weekly
 downloads and has received almost 700 GitHub stars, an appreciation score gained
@@
 	perplexities by using a multiscale kernel affinity model \textbf{(c)}
-	and obtain a good trade-off between global and local structure.
+	and obtain a trade-off between global and local structure.
 	Alternatively, we can inspect more fine-grained structure and reveal
@@
 another, all clusters appear equidistant from their neighbors, and the overall
-relations between groups are not obvious. Embedding in Fig.~\ref{fig:tasic}.b
+relations between groups are not obvious. The embedding in Fig.~\ref{fig:tasic}.b
 focuses on preserving larger neighborhoods of points, resulting in a more
@@
 inhibitory neurons. Unfortunately, focusing on preserving large neighborhoods
-leads to the absorption of smaller clusters by the larger ones. Alternatively,
+leads to the absorption of smaller clusters into larger ones. Alternatively,
 Fig.~\ref{fig:tasic}.c uses multi-scale similarity kernels that aim to preserve
@@
   \caption{\label{fig:cao}
-Increasing the exaggeration parameter leads to compact clusters, highlighting the data's global organization and emphasizing the continuity between cell states.
-The data set from Cao \textit{et al.}~\cite{cao2019single} contains expression profiles from 2,058,652 single cells.
-The data were collected from mice embryos at different development stages at daily intervals after 9.5 to 13.5 days.
-\textbf{(a)} reveals that the data is comprised of two main components -- the neural tube and mesenchymal cells -- as well as several other smaller clusters.
-Additionally, the colors indicate developmental progression with red indicating least-developed cells and blue indicating most developed cells.
-The overall developmental trajectory is most apparent with higher exaggeration levels, showing red cells slowly transitioning into blue cells.
-Progressively easing the exaggeration factor uncovers finer clusters within the larger groups, as shown in
-\textbf{(b)} with exaggeration of two and subsequently in \textbf{(c)}, where we show the standard t-SNE with no exaggeration. 32,011 putative doublets are excluded from the visualizations.
+Increasing the exaggeration parameter leads to compact clusters, highlighting
+	the data's global organization and emphasizing the continuity between
+	cell states. The data set from Cao \textit{et al.}~\cite{cao2019single}
+	contains expression profiles from 2,058,652 single cells. The data were
+	collected from mice embryos at different developmental stages at daily
+	intervals after 9.5 to 13.5 days. \textbf{(c)} reveals that the data is
+	comprised of two main components -- the neural tube and mesenchymal
+	cells -- as well as several other smaller clusters. Additionally, the
+	colors indicate developmental progression with red indicating
+	least-developed cells and blue indicating most developed cells. The
+	overall developmental trajectory is most apparent with higher
+	exaggeration levels, showing red cells slowly transitioning into blue
+	cells. Progressively easing the exaggeration factor uncovers finer
+	clusters within the larger groups, as shown in \textbf{(b)} with
+	exaggeration of two and subsequently in \textbf{(a)}, where we show the
+	standard t-SNE with no exaggeration. 32,011 putative doublets are
+	excluded from the visualizations.
 }
@@
 
-Exaggeration can highlight transitions between cell states in developmental studies.
-Standard t-SNE often produces embeddings with clearly defined, discrete clusters.
-We can adjust the level of granularity and resolution of the clusters with several parameters in Fig.~\ref{fig:tasic}.
-However, discrete clusters are often undesired in developmental studies where cells' state is changed continuously.
-To this end, other embedding techniques such as UMAP and ForceAtlas2~\cite{jacomy2014forceatlas2} tend to capture the continuity between cell states better.
-Recently, B{\"o}hm \textit{et al.}~\cite{bohm2020unifying} have shown that embeddings produced by t-SNE with exaggeration values of 4 and $\sim30$ construct embeddings which are markedly similar to UMAP and ForceAtlas2, respectively.
-For example, in Fig.~\ref{fig:cao}.a, the developmental trajectory between different time points is difficult to observe due to many sprawled out clusters.
-On the other hand, it is easier to trace the development when we increase the exaggeration factor from $1$ to $4$ (Fig.~\ref{fig:cao}.b-c).
+Exaggeration can highlight transitions between cell states in developmental
+studies. Standard t-SNE often produces embeddings with clearly defined, discrete
+clusters. We can adjust the level of granularity and resolution of the clusters
+with several parameters in Fig.~\ref{fig:tasic}. However, discrete clusters are
+often undesired in developmental studies where cells' state is changed
+continuously. To this end, other embedding techniques such as UMAP and
+ForceAtlas2~\cite{jacomy2014forceatlas2} tend to better capture the continuity
+between cell states. Recently, B{\"o}hm \textit{et al.}~\cite{bohm2020unifying}
+have shown that embeddings produced by t-SNE with exaggeration values of 4 and
+$\sim30$ construct embeddings which are markedly similar to UMAP and
+ForceAtlas2, respectively. For example, in Fig.~\ref{fig:cao}.a, the
+developmental trajectory between different time points is difficult to observe
+due to many sprawled out clusters. On the other hand, it is easier to trace the
+development when we increase the exaggeration factor from $1$ to $4$
+(Fig.~\ref{fig:cao}.b-c).
 
@@
 more massive data sets. \opentsne\ provides a flexible program interface to
-incorporate these improvements in just a few code lines. Furthermore, \opentsne\
+incorporate these improvements in just a few lines of code. Furthermore, \opentsne\
 supports custom affinity models, enabling users to construct t-SNE embeddings on
 non-tabular relational data: the only requirement imposed by the affinity-model
-is the availability of similarity between data points, that is, the distance
-matrix.  Finally, the \opentsne's comprehensive callback system can be utilized
+is the availability of similarities between data points.
+Finally, \opentsne's comprehensive callback system can be utilized
 to monitor and adapt different stages of the optimization phase and has been
@@
 points~\cite{becht2019dimensionality}. The culprit for slow response time stems
-from a particular optimization procedure and its specific implementation in the
-popular Python library. Namely, until quite recently, most popular
+from a particular optimization procedure and its specific implementation in
+popular Python libraries. Until quite recently, most popular
 implementations of t-SNE were based on the Barnes-Hut approximation scheme
@@
   \caption{\label{fig:benchmarks}
-We benchmark openTSNE (v0.4.3) against three popular open-source implementations
-	from scikit-learn~\cite{pedregosa2011scikit} (v0.23.1),
-	MulticoreTSNE~\cite{Ulyanov2016} (v0.1), and
-	FIt-SNE~\cite{linderman2019fast} (v1.1.0). Experiments were run on a
+We benchmark \opentsne\ (v0.4.3) against three popular open-source implementations
+	from \textsf{scikit-learn}~\cite{pedregosa2011scikit} (v0.23.1),
+	\textsf{MulticoreTSNE}~\cite{Ulyanov2016} (v0.1), and
+	\textsf{FIt-SNE}~\cite{linderman2019fast} (v1.1.0). Experiments were run on a
 	consumer-grade Intel Core i7-7700HQ processor found in laptop computers,
@@
 	data sets for each size. In total, we run each implementation on 30
-	different data sets. Notice that openTSNE scales similarly to FIt-SNE,
+	different data sets. Notice that \opentsne\ scales similarly to \textsf{FIt-SNE},
 	as they both use the same interpolation-based approximation scheme,
-	while scikit-learn and MulticoreTSNE utilize the Barnes-Hut
+	while \textsf{scikit-learn} and \textsf{MulticoreTSNE} utilize the Barnes-Hut
 	approximation.
@@
 and \opentsne\ scale better than their Barnes-Hut counterparts --
-\textsf{scikit-learn} and \textsf{MulticoreTSNE}. While \opentsne\'s
+\textsf{scikit-learn} and \textsf{MulticoreTSNE}. While \opentsne's
 implementation uses Python that incurs some runtime overhead compared to its C++
@@
 platforms, making the installation process as seamless as possible. One can
-install \opentsne\ through the Python Package Index (PyPI) or conda on the
-conda-forge channel. \opentsne\ interface follows that of \textsf{scikit-learn},
+install \opentsne\ through the Python Package Index (\textsf{PyPI}) or \textsf{conda} from the
+\textsf{conda-forge} channel. \opentsne's interface is inspired by \textsf{scikit-learn},
 which is well established in the Python data science ecosystem. \opentsne\
 implements multi-threaded versions of both the Barnes-Hut and FIt-SNE
-approximation schemes, enabling it for data sets containing millions of data
-points. While the Python virtual machine introduces some performance overhead,
+approximation schemes, enabling it to be applied to data sets containing millions of data
+points. While the Python virtual machine inevitably introduces some performance overhead,
 the runtime is comparable to its C++ counterpart -- \textsf{FIt-SNE}. Finally,
 \opentsne\ is extensible. Its modular design enables researchers to quickly
-experiment with different settings and easily incorporate their components into
+experiment with different parameter settings and easily incorporate their own components into
 the software. We provide a full feature list and comparison to other popular
@@
 \caption{\label{tab:features}
-We compare features of \opentsne to three popular open-source implementations
-	from scikit-learn~(v0.23.1), MulticoreTSNE (v0.1), and FIt-SNE (v1.1.0).
-	The first section of the comparison addresses packaging and
-	distribution. A properly packaged library is easily accessible to users,
-	and developers should easily include it in dependency lists of other
-	software packages. The second section of the comparison lists the two
-	existing t-SNE approximation schemes. The FIt-SNE approximation scheme
-	is required for t-SNE to scale up to millions of data points. The final
-	section provides a list of extensions and improvements to the standard
-	t-SNE algorithm, many of which can produce markedly better
-	visualizations.
+We compare features of \opentsne\ to three popular open-source implementations
+	from \textsf{scikit-learn}~(v0.23.1), \textsf{MulticoreTSNE} (v0.1), and
+	\textsf{FIt-SNE} (v1.1.0).  The first section of the comparison
+	addresses packaging and distribution. A properly packaged library is
+	easily accessible to users, and developers should easily include it in
+	dependency lists of other software packages. The second section of the
+	comparison lists the two existing t-SNE approximation schemes. The
+	FIt-SNE approximation scheme is required for t-SNE to scale up to
+	millions of data points. The final section provides a list of extensions
+	and improvements to the standard t-SNE algorithm, many of which can
+	produce markedly better visualizations.
 }
@@
 \toprule
-PyPI package & \checkmark & \checkmark & & \checkmark \\
-conda package & \checkmark & & & \checkmark \\
+\textsf{PyPI} package & \checkmark & \checkmark & & \checkmark \\
+\textsf{conda} package & \checkmark & & & \checkmark \\
 Precompiled binaries & \checkmark & & & \checkmark \\
@@
 
+\fontsize{8}{10}\selectfont
 \section*{Methods}
 
-This section is meant to familiarize the reader with the t-SNE algorithm and
-its recent extensions. We motivate each extension and introduce standard notation.
-This brief review is by no means comprehensive, but is
-self-contained and introduces the mathematics upon which the \opentsne\ library
-is based.
+This section is meant to familiarize the reader with the t-SNE algorithm and its
+recent extensions. We motivate each extension and introduce standard notation.
+This brief review is by no means comprehensive, but is self-contained and
+introduces the mathematics upon which the \opentsne\ library is based.
 
@@
 \url{https://github.com/pavlin-policar/openTSNE}. \opentsne\ is also available
-on \texttt{PyPI} and \texttt{conda-forge}. The data sets used and scripts used
+on \textsf{PyPI} and \textsf{conda-forge}. The data sets used and scripts used
 in this study are included in accompanying notebooks, publicly available at
 \url{https://github.com/pavlin-policar/opentsne-paper}.
-
-\begin{backmatter}
 
@@
 
-\end{backmatter}
 \end{document}

### 2020-10-01 paper/policar2020-opentsne.tex
Edits to main text

@@
   \caption{\label{fig:macosko}
-Three t-SNE embeddings demonstrate theoretical advances and corresponding implementations in
-\opentsne. The data in \textbf{(a)} and \textbf{(b)} represent 44,808
+  We use \opentsne\ to generate three t-SNE embeddings and demonstrate recent
+  theoretical advances. The data in \textbf{(a)} and \textbf{(b)} represent 44,808
   single-cell gene-expression profiles of mouse retinal cells from Macosko
@@
   additionally contains 27,499 expression profiles from mouse
-retinal cells from Shekhar \textit{et
-al.}~\cite{shekhar2016comprehensive}. \textbf{(a)} We construct a t-SNE
+  retinal cells from Shekhar \textit{et al.}~\cite{shekhar2016comprehensive}.
+  \textbf{(a)} We construct a t-SNE
   embedding following the parameter choices from the original publication
@@
   between data points and initialized so that the global layout is
-meaningful. Unlike in \textbf{(a)}, the green and blue clusters
+  as meaningful as possible. Unlike in \textbf{(a)}, the green and blue clusters
   representing different sub-types of amacrine and bipolar cells are now
@@
 	correspond to inhibitory neurons, and gray/brown colors correspond to
-	non-neuronal cells. The standard t-SNE \textbf{(a)} emphasizes local
+	non-neuronal cells. Standard t-SNE \textbf{(a)} emphasizes local
 	structure while increasing perplexity \textbf{(b)} results in a more
@@
   Increasing the exaggeration parameter leads to compact clusters, highlighting
-	the data's global organization and emphasizing the continuity between
-	cell states. The data set from Cao \textit{et al.}~\cite{cao2019single}
+	the data's global organization and emphasizing the continuous nature of
+	cell state transitions. The data set from Cao \textit{et al.}~\cite{cao2019single}
 	contains expression profiles from 2,058,652 single cells. The data were
@@
 	comprised of two main components -- the neural tube and mesenchymal
-	cells -- as well as several other smaller clusters. Additionally, the
+	cells -- as well as several other smaller clusters. The
 	colors indicate developmental progression with red indicating

### 2020-10-01 paper/policar2020-opentsne-BMC.tex
More edits to main text

--- 2020-10-01 paper/policar2020-opentsne.tex
@@
 platforms, making the installation process as seamless as possible. One can
-install \opentsne\ through the Python Package Index (PyPI) or conda on the
-conda-forge channel. \opentsne\ interface follows that of \textsf{scikit-learn},
+install \opentsne\ through the Python Package Index (\textsf{PyPI}) or \textsf{conda} on the
+\textsf{conda-forge} channel. \opentsne\ interface follows that of \textsf{scikit-learn},
 which is well established in the Python data science ecosystem. \opentsne\
@@
 \url{https://github.com/pavlin-policar/openTSNE}. \opentsne\ is also available
-on \texttt{PyPI} and \texttt{conda-forge}. The data sets used and scripts used
+on \textsf{PyPI} and \textsf{conda-forge}. The data sets used and scripts used
 in this study are included in accompanying notebooks, publicly available at

### 2020-10-01 paper/policar2020-opentsne.tex
wide format

@@
-\documentclass[twocolumn]{article}
+% \documentclass[twocolumn]{article}
+\documentclass[letter]{article}
+\usepackage[a4paper, total={6in, 8in}]{geometry}
 
@@
 Point-based visualizations of large, multi-dimensional data from molecular
-biology can reveal meaningful clusters. One of the most popular
+biology are often used to reveal meaningful clusters. One of the most popular
 techniques to construct such visualizations is $t$-distributed stochastic
@@
 of the resulting visualizations. We introduce \opentsne, a modular
-Python library that implements the core t-SNE algorithm and its
-extensions. The library is orders of magnitude faster than existing
-popular implementations, including those from scikit-learn. Unique to
-\opentsne\ is also the ability to map new data to existing embeddings,
+Python library that implements the core t-SNE algorithm and its many
+extensions.
+The library is fast and enables users to embed data sets containing millions of
+data points in a matter of minutes on consumer-grade laptop computers.
+Unique to \opentsne\ is also the ability to map new data to existing embeddings,
 which can surprisingly assist in solving batch effects.
@@
 The abundance of high-dimensional data sets in molecular biology calls for
-dimensionality reduction techniques, particularly for methods that produce
+dimensionality reduction techniques that are able to produce
 informative data visualizations. Popular approaches include principal component
@@
 embedding (t-SNE)~\cite{maaten2008visualizing}, and uniform manifold
-approximation and projections (UMAP)~\cite{2018arXivUMAP}. Among these, t-SNE
-lately received much attention as it can address high volumes of data and reveal
-a meaningful clustering structure. Increased resolution and throughput of modern
+approximation and projections (UMAP)~\cite{2018arXivUMAP}. Among these, t-SNE has
+received much attention as it can address high volumes of data and reveal
+meaningful clustering structure. The increased resolution and throughput of modern
 molecular assays has lead to the frequent use of t-SNE in diverse fields
@@
 expression data, our running example, often start with an overview of the cell
-landscape, where t-SNE embeds high-dimensional expression profiles into a
+landscape, where t-SNE is used to embed high-dimensional expression profiles into a
 two-dimensional space. Figs.~\ref{fig:macosko}.a and \ref{fig:macosko}.b show
@@
 Despite its utility, t-SNE has often been criticized for its limited
-scalability, lack of global organization -- t-SNE emphasizes local clusters that
-are arbitrarily scattered in the low-dimensional space -- and the absence of
+scalability, lack of global organization -- t-SNE identifies well-defined clusters that
+may be arbitrarily scattered throguhout the low-dimensional space -- and the absence of
 theoretically-founded methods to map new data into existing
 embeddings~\cite{ding2018interpretable,becht2019dimensionality}. Most of these
-shortcomings have recently been addressed. Linderman \textit{et al.} sped-up the
-method through an interpolation-based approximation, achieving linear complexity
+shortcomings have recently been addressed. Linderman \textit{et al.} developed
+an interpolation-based approximation scheme which massively improved the
+scalability of t-SNE, achieving linear time complexity
 in the number of samples~\cite{linderman2019fast}. Kobak \& Berens proposed
-several techniques to improve global positioning, including estimating
+several techniques to improve global cluster coherence, including estimating
 similarities with a mixture of Gaussian kernels~\cite{kobak2019art}. In our
@@
 \opentsne\ is compatible with the Python data science ecosystem and
-libraries that include \textsf{numpy}, \textsf{scikit-learn}, \textsf{scanpy}),
-thus providing a familiar and intuitive application program interface to the new
+libraries including \textsf{numpy}, \textsf{scikit-learn}, \textsf{scanpy}),
+providing a familiar and intuitive application program interface to new
 users. Its modular design encourages extensibility and experimentation with
-various settings and changes in the analysis pipeline.  We have released
-\opentsne\ in an open-source to support adaptations by the wider community. At
+various settings and changes to the analysis pipeline. We have released
+\opentsne\ to the open-source community to make recent theoretical advancements
+more accessible to a wider audience. At
 the time of the writing, the library averages thousands of weekly
@@
 
-Accessibility of the latest theoretical advances in t-SNE is one of the core
-design principles of \opentsne. These advances are accessible through a
-programming interface of the library, where we closely follow the style of the
-interface as implemented in \textsf{scikit-learn}~\cite{sklearn_api}, a popular
+Accessibility of the latest theoretical advancments in t-SNE is one of the core
+design principles of \opentsne. This range of advancments is easily accessible through an intuitive
+programming interface, where we closely follow the style
+\textsf{scikit-learn}~\cite{sklearn_api}, a popular and widely used
 library for machine learning. The following snippet exemplifies this interface
-and codes for four visualizations in Fig.~\ref{fig:tasic}, demonstrating the
-effect of a particular recent theoretical advance in the t-SNE algorithm.
+and was used to generate the four visualizations in Fig.~\ref{fig:tasic}, demonstrating the
+effect of a particular recent theoretical advancments in the t-SNE algorithm.
 
@@
 tsne1 = openTSNE.TSNE(perplexity=30).fit(X)
-# b - high perplexity  for globa structure
+# b - high perplexity for global structure
 tsne2 = openTSNE.TSNE(perplexity=500).fit(X)
@@
 sparse matrix \texttt{X} and creates four \texttt{TSNEEmbedding} objects. These
-objects encode t-SNE embeddings that may subsequently be optimized under
+objects represent t-SNE embeddings that may subsequently be optimized under
 different parameter settings or used to embed new samples into the embedding
 landscape. Function calls may include additional parameters not shown in the
-example code but explained in accompanying notebooks available at
+example code but are explained in the accompanying notebooks available at
 \url{https://github.com/pavlin-policar/opentsne-paper}.
@@
 lies on a lower-dimensional manifold, which can accurately be captured by a
-smaller number of dimensions. However, there is no evidence that every data set
-can accurately be described by using only two dimensions, and any such embedding
+small number of dimensions. However, there is no evidence that every data set
+can accurately be described using only two dimensions, and any such embedding
 will inevitably lead to a loss of information. Thus, it is beneficial to examine
@@
 	\includegraphics[width=\textwidth]{tasic2018}
-	\caption{\label{fig:tasic}We use \opentsne\ to create four different
+	\caption{\label{fig:tasic}
+  We use \opentsne\ to create four different
 	visualizations of the Tasic \textit{et al.}~\cite{tasic2018shared} data,
@@
 Fig.~\ref{fig:tasic}.c uses multi-scale similarity kernels that aim to preserve
-both the global organization of clusters and prevent the cluster absorption. We
-constructed the embeddings from Fig.~\ref{fig:tasic}.d with the settings used
-for Fig.~\ref{fig:tasic}.a, but at a finer level of resolution. The distinction
-between these two depictions reveals that many clusters are composed of
-subgroups representing different cell subtypes.
+both the global organization of clusters and prevent smaller cluster absorption. We
+constructed the embedding from Fig.~\ref{fig:tasic}.d with the settings used
+for Fig.~\ref{fig:tasic}.a, but at a finer level of resolution.
+The figure demonstrates that some clusters are composed of numerous, smaller
+subgroups representing different cell subpopulations which are not visible under
+standard parameter settings.
 
@@
 eight somewhat smaller clusters. The visualization also reveals several tiny
-clusters, possibly corresponding to
-rare cell types.
+clusters, possibly corresponding to rare cell types.
 
@@
 with several parameters in Fig.~\ref{fig:tasic}. However, discrete clusters are
-often undesired in developmental studies where cells' state is changed
-continuously. To this end, other embedding techniques such as UMAP and
-ForceAtlas2~\cite{jacomy2014forceatlas2} tend to better capture the continuity
+often undesired in developmental studies where cells' state is assumed to follow
+a continuous transition path. To this end, other embedding techniques such as UMAP and
+ForceAtlas2~\cite{jacomy2014forceatlas2} are used to better capture the continuity
 between cell states. Recently, B{\"o}hm \textit{et al.}~\cite{bohm2020unifying}
-have shown that embeddings produced by t-SNE with exaggeration values of 4 and
+showed that embeddings produced by t-SNE with exaggeration values of 4 and
 $\sim30$ construct embeddings which are markedly similar to UMAP and
@@
 due to many sprawled out clusters. On the other hand, it is easier to trace the
-development when we increase the exaggeration factor from $1$ to $4$
-(Fig.~\ref{fig:cao}.b-c).
+development when we increase the exaggeration factor from $1$ to $2$ to $4$ in
+Figs.~\ref{fig:cao}.b-c.
 
@@
 
-Unlike other popular dimensionality reduction methods such as principal
+Unlike other popular dimensionality reduction techniques such as principal
 component analysis or autoencoders, t-SNE is a non-parametric method and does
@@
 publicly available library allowing users to add new samples to existing
-embeddings.
+embeddings in a principled manner.
 
@@
 al.}~\cite{shekhar2016comprehensive}, we can correctly map the bipolar cell
-clusters in the reference.
+clusters onto the reference embedding.
 
@@
 the expression of well-established gene markers for the major cell types. We
-then embed cells from Harris~\textit{et al.}~\cite{harris2018classes} into a
-constructed atlas. In Harris~\textit{et al.}~, labels are provided only for
+then embed cells from Harris~\textit{et al.}~\cite{harris2018classes} into the
+constructed cell atlas. In Harris~\textit{et al.}~, labels are provided only for
 neuronal cells. In the resulting mapping, we can quickly identify other
@@
 further use marker genes to validate that the mapping in the reference landscape
-was correct.
+is correct.
 
@@
 The examples presented above demonstrate how to use \opentsne\ to quickly gain
-insight into a newly-sequenced, single-cell data sets by utilizing existing cell
+insight into newly-sequenced, single-cell data sets by utilizing existing cell
 atlases. The approach is general and not limited to single-cell gene expression,
@@
 Versatility, the ability to use and combine different optimization approaches to
-construct the embedding spaces, is another of \opentsne's core design
-principles. Kobak \& Berens recently introduced several recommendations and
+construct different embedding spaces, is another of \opentsne's core design
+principles. Kobak \& Berens recently provided several recommendations and
 tricks to obtain better and more meaningful t-SNE
@@
 kernels, perplexity annealing, and increasing exaggeration when working with
-more massive data sets. \opentsne\ provides a flexible program interface to
+massive data sets. \opentsne\ provides a flexible program interface to
 incorporate these improvements in just a few lines of code. Furthermore, \opentsne\
@@
 non-tabular relational data: the only requirement imposed by the affinity-model
-is the availability of similarities between data points.
+is some notion of similarity between data points.
 Finally, \opentsne's comprehensive callback system can be utilized
@@
 developed by van der Maaten in 2014~\cite{van2014accelerating} with asymptotic
-complexity $\mathcal{O}(N \log N)$, where $N$ is the number of data items ({\em
+time complexity $\mathcal{O}(N \log N)$, where $N$ is the number of data items ({\em
 e.g.} cells). The most widely-used implementation of t-SNE came from
@@
 developed a new approximation scheme -- FIt-SNE -- which further reduces the
-asymptotic time complexity to $\mathcal{O}(N)$. We have included this scheme in
+asymptotic time complexity to $\mathcal{O}(N)$. We include this approach in
 \opentsne\, enabling the embedding of large data sets in a matter of minutes.
 
-Fig.~\ref{fig:benchmarks} benchmarks four Python t-SNE implementations,
+Fig.~\ref{fig:benchmarks} benchmarks four popular Python t-SNE implementations,
 including those from \textsf{scikit-learn} (v0.23.1), \textsf{MulticoreTSNE}
@@
 benchmarks on two computational platforms, one representing the usage on a
-personal computer and the other utility of these libraries on a high-performance
+personal computer and the other the utility of these libraries on a high-performance
 computing platform. The Intel(R) Core i7 is commonly found in consumer-grade
@@
 Intuitive access and simple installation procedures almost universally correlate
-with the widespread adoption of novel computational techniques. While the t-SNE
-from \textsf{scikit-learn} fits this requirement, its implementation is
+with the widespread adoption of novel computational techniques. While the t-SNE implementation
+from \textsf{scikit-learn} fits this requirement, the implementation proves
 prohibitively slow for even moderately-sized data sets that span tens of
@@
 compile the software themselves. This problem is critical, for instance, for
-users of Windows, where the C++ compiler does not come with the system, making
+users of the Windows operating system, where the C++ compiler does not come with the system, making
 the correct configuration of current t-SNE implementations cumbersome.
 
-We have designed \opentsne\ to make it accessible to a broader audience. We
+We designed \opentsne\ to be accessible to a broader audience. We
 provide precompiled binaries for all major Python versions on all major
@@
 install \opentsne\ through the Python Package Index (\textsf{PyPI}) or \textsf{conda} from the
-\textsf{conda-forge} channel. \opentsne's interface is inspired by \textsf{scikit-learn},
+\textsf{conda-forge} channel, the two most widely adopted Python package managers.
+\opentsne's interface is inspired by \textsf{scikit-learn},
 which is well established in the Python data science ecosystem. \opentsne\
@@
 \opentsne\ is extensible. Its modular design enables researchers to quickly
-experiment with different parameter settings and easily incorporate their own components into
+experiment with different parameter settings and easily incorporate custom components into
 the software. We provide a full feature list and comparison to other popular
@@
 \caption{\label{tab:features}
-  We compare features of \opentsne\ to three popular open-source implementations
+  Features of \opentsne\ compared to other three popular open-source implementations
 	from \textsf{scikit-learn}~(v0.23.1), \textsf{MulticoreTSNE} (v0.1), and
@@
 }
-
+\begin{center}\small
 \newcommand*\rot{\rotatebox{90}}
@@
 \end{tabular}
+\end{center}
 \end{table}

### 2020-12-15 paper/policar2020-opentsne.tex
Change tasic figure to 2x2 instead of 4x1
conclusion

@@
 \newcommand{\opentsne}{\textsf{openTSNE}}
+
+\renewcommand{\baselinestretch}{1.2}
 
@@
 \begin{figure*}[htbp]
-	\includegraphics[width=\textwidth]{tasic2018}
+  \center
+  \includegraphics[width=0.75\textwidth]{tasic2018}
 	\caption{\label{fig:tasic}
@@
 \section*{Conclusion}
-Text for this section \ldots
 
+Data visualization for efficient exploration and effective communication is integral to scientific progress~\cite{wong2012points}.
+It is of no surprise, then, that a technique called t-SNE that can embed multi-dimensional data within two-dimensional maps has gained such popularity.
+The t-SNE visualizations appear in numerous recent publications, including those in this journal, and are, in particular, instrumental for the advancement in specific fields of biology, such as single-cell transcriptomics.
+Yet, with growing data volume and the need for data integration from various experiments, the original t-SNE exhibits lack of scalability and susceptibility to batch effects.
+We here introduce \opentsne\, an open-source Python t-SNE library that addresses these concerns and includes recent methodological advancements.
+\opentsne\ can scale to millions of data records and fit local cluster exploration and global structure discovery.
+It is also the only existing t-SNE library that enables transfer learning in a principled manner, where new data can be embedded into existing atlases.
+
+\section*{Availability}
+
+\opentsne\ is distributed under the BSD-3-Clause License and is publicly
+available as an open-source package at
+\url{https://github.com/pavlin-policar/openTSNE}. \opentsne\ is also available
+on \textsf{PyPI} and \textsf{conda-forge}. The data sets used and scripts used
+in this study are included in accompanying notebooks, publicly available at
+\url{https://github.com/pavlin-policar/opentsne-paper}.
+
+\section*{Competing interests}
+The authors declare that they have no competing interests.
+
+\section*{Author's contributions}
+P.G.P.
+implemented and maintains \opentsne\ library.
+M.S.
+helped in the development of mathematical foundations and advised on the inclusion of the extensions.
+B.Z.
+initiated and supervised the project.
+All three co-authors wrote the manuscript.
+
+\section*{Acknowledgements}
+We want to thank Dmitry Kobak for various helpful discussions and best practices
+when using t-SNE and his contributions to the source code. We would also
+like to thank George Linderman for his help with the FIt-SNE algorithm.
+
+\bibliographystyle{bmc-mathphys} % Style BST file (bmc-mathphys, vancouver, spbasic).
+\bibliography{references}      % Bibliography file (usually '*.bib' )
+
+\eject
 \fontsize{8}{10}\selectfont
-\section*{Methods}
+\section*{Online Methods}
 
@@
 
-\section*{Availability}
-
-\opentsne\ is distributed under the BSD-3-Clause License and is publicly
-available as an open-source package at
-\url{https://github.com/pavlin-policar/openTSNE}. \opentsne\ is also available
-on \textsf{PyPI} and \textsf{conda-forge}. The data sets used and scripts used
-in this study are included in accompanying notebooks, publicly available at
-\url{https://github.com/pavlin-policar/opentsne-paper}.
-
-\section*{Competing interests}
-The authors declare that they have no competing interests.
-
-\section*{Author's contributions}
-P.G.P. implemented and maintains \opentsne library. M.S. helped in development
-of mathematical foundations and advised on inclusion of the extensions.
-B.Z. initiated and supervised the project. All three co-authors wrote
-the manuscript.
-
-\section*{Acknowledgements}
-We want to thank Dmitry Kobak for various helpful discussions and best practices
-when using t-SNE and his contributions to the source code. We would also
-like to thank George Linderman for his help with the FIt-SNE algorithm.
-
-\bibliographystyle{bmc-mathphys} % Style BST file (bmc-mathphys, vancouver, spbasic).
-\bibliography{references}      % Bibliography file (usually '*.bib' )
-
 \end{document}

### 2021-03-16 paper/policar2020-opentsne.tex
pass through grammarly

@@
 
-This section is meant to familiarize the reader with the t-SNE algorithm and its
-recent extensions. We motivate each extension and introduce standard notation.
-This brief review is by no means comprehensive, but is self-contained and
-introduces the mathematics upon which the \opentsne\ library is based.
+We here introduce notation, review the core t-SNE algorithm, and motivate and provide an overview of its recent extensions. The section presents the mathematics implemented in the \opentsne\ library.
 
@@
 
-t-distributed stochastic neighbor embedding (t-SNE) is a non-linear
-dimensionality reduction method, which aims to find a low-dimensional
-embedding, where local neighborhoods are preserved. More formally, given a
-multi-dimensional data set $\mathbf{X} = \left \{ \mathbf{x}_1, \mathbf{x}_2,
-\dots, \mathbf{x}_N \right \} \in \mathbb{R}^D$ where $N$ is the number of data
-points in the data set, t-SNE aims to find a low dimensional embedding
-$\mathbf{Y} = \left \{ \mathbf{y}_1, \mathbf{y}_2, \dots, \mathbf{y}_N \right
-\} \in \mathbb{R}^d$ where $d \ll D$, such that if points $\mathbf{x}_i$ and
-$\mathbf{x}_j$ are close in the high-dimensional space, their corresponding
-embeddings $\mathbf{y}_i$ and $\mathbf{y}_j$ are also close. Since t-SNE is
-primarily used as a visualization tool, $d$ is typically set to two. The
-similarity between two data points in the high-dimensional space is defined as
+A t-distributed stochastic neighbor embedding (t-SNE) is a non-linear dimensionality reduction method that finds a low-dimensional embedding with preserved neighborhoods.
+More formally, given a multi-dimensional data set $\mathbf{X} = \left \{ \mathbf{x}_1, \mathbf{x}_2, \dots, \mathbf{x}_N \right \} \in \mathbb{R}^D$ where $N$ is the number of data points in the data set, t-SNE aims to find a low dimensional embedding $\mathbf{Y} = \left \{ \mathbf{y}_1, \mathbf{y}_2, \dots, \mathbf{y}_N \right\} \in \mathbb{R}^d$ where $d \ll D$, such that if points $\mathbf{x}_i$ and $\mathbf{x}_j$ are close in the high-dimensional space, their corresponding embeddings $\mathbf{y}_i$ and $\mathbf{y}_j$ are also close.
+Since t-SNE is primarily used as a visualization tool, $d$ is typically set to two.
+The similarity between two data points in the high-dimensional space is defined as
 
@@
 
-The bandwidth of each Gaussian kernel $\sigma_i$ is selected such that the
-perplexity of the distribution matches a user-specified parameter value
+The bandwidth of each Gaussian kernel $\sigma_i$ is selected such that the perplexity of the distribution matches a user-specified parameter value
 
@@
 
-\noindent Different bandwidths $\sigma_i$ enable t-SNE to adapt to the varying
-density of the data in the multi-dimensional space. Perplexity is sometimes
-interpreted as the continuous analog to the number of nearest neighbors to which
-distances will attempt to be preserved.
+\noindent Different bandwidths $\sigma_i$ enable t-SNE to adapt to the varying density of the data in the multi-dimensional space.
+We can view the perplexity as the continuous analog to the number of nearest neighbors to which the distances will be preserved.
 
-The similarity between points $\mathbf{y}_i$ and $\mathbf{y}_j$ in the
-embedding space is defined using the $t$-distribution with a single degree of
-freedom (Cauchy kernel)
+The similarity between points $\mathbf{y}_i$ and $\mathbf{y}_j$ in the embedding space is defined using the $t$-distribution with a single degree of freedom (Cauchy kernel)
 
@@
 
-\noindent The objective is to find embeddings $\mathbf{Y}$ such that the
-KL divergence is minimized.
-
-The corresponding gradient takes the form
+\noindent The objective is to find embeddings $\mathbf{Y}$ that minimize the KL divergence. The corresponding gradient takes the form
 \begin{equation}
@@
 
-Optimization is performed with batch gradient descent using the delta-bar-delta
-update rule~\cite{jacobs1988increased}. Originally, t-SNE was run for 1000
-iterations consisting of two phases: in the first \textit{early exaggeration}
-phase, the attractive forces between data points are increased by some factor
-$\rho$, typically set to $12$, so that points in the embedding can more easily
-move throughout the space and find their respective neighbors. The remaining 750
-iterations are run with $\rho=1$, which reverts the attractive forces to their
+Optimization is performed with batch gradient descent using the delta-bar-delta update rule~\cite{jacobs1988increased}.
+Originally, t-SNE was run for 1000 iterations consisting of two phases: in the first \textit{early exaggeration} phase, the attractive forces between data points are increased by some factor $\rho$, typically set to $12$, so that points in the embedding can more easily move throughout the space and find their respective neighbors.
+The remaining 750 iterations are run with $\rho=1$, which reverts the attractive forces to their
 original values and produces the final embedding.
 
-Belkina \textit{et al.} later found that faster convergence can be achieved by
-increasing the learning rate to $\eta=N/12$~\cite{belkina2019automated}. As a
-side-effect, embeddings converge faster, and the number of iterations can be
-lowered to 750, decreasing the overall runtime.  This convention has been
-adopted by most modern t-SNE implementations.
+Belkina \textit{et al.} later found that faster convergence can be achieved by increasing the learning rate to $\eta=N/12$~\cite{belkina2019automated}.
+As a side-effect, embeddings converge faster, and the number of iterations can be
+lowered to 750, decreasing the overall runtime.  Most modern t-SNE implementations have adopted this convention.
 
-\subsection*{Efficient Approximation Schemes} A direct evaluation of t-SNE
-gradients requires $\mathcal{O}(N^2)$ operations, which makes its application
-impractical to any reasonably-sized data set and beckons for the development of
-efficient approximation schemes. Van der Maaten observed that the t-SNE gradient
-may be cast as an N-body problem where data points represent particles which
-attract and repel each other~\cite{van2014accelerating}. The gradient from Eqn.
+\subsection*{Efficient Approximation Schemes} A direct evaluation of t-SNE gradients requires $\mathcal{O}(N^2)$ operations, which makes its application impractical to any reasonably-sized data set and beckons for the development of efficient approximation schemes.
+Van der Maaten observed that the t-SNE gradient might be cast as an N-body problem where data points represent particles that attract and repel each other~\cite{van2014accelerating}.
+The gradient from Eqn.
 (\ref{eq:tsne_gradient}) can be rewritten as
@@
 where $Z = \sum_{k \neq l}\left ( 1 + || \mathbf{y}_k - \mathbf{y}_l ||^2 \right
-)^{-1}$. This equation can then be viewed as a particle simulation, where the
+)^{-1}$. We can view this equation as a particle simulation, where the
 two terms represent the attractive and repulsive forces between individual
-particles. Each term lends itself to efficient approximations, which enable us
-to greatly reduce the time complexity of t-SNE.
+particles. Each term lends itself to efficient approximations, enabling us
+to reduce the time complexity of t-SNE greatly.
 
@@
 of data points is excessive, and that considering only a handful of nearest
-neighbors at each point is sufficient to obtain a good
-approximation~\cite{van2014accelerating}.  Because distances are transformed to
+neighbors at each point are sufficient to obtain a good
+approximation~\cite{van2014accelerating}.  Because t-SNE transforms the distances to
 similarities using a Gaussian kernel, and the bandwidth of each kernel is
@@
 
-Examining the second term of Eqn.~(\ref{eq:grad_attr_rep}) we notice that each
+Examining the second term of Eqn.~(\ref{eq:grad_attr_rep}), we notice that each
 point indiscriminately exerts a repulsive force on all other points. Van der
@@
 between data points~\cite{van2014accelerating}. Briefly, in the 2D case, the
-approach splits the space into quadrants and entire regions may be summarized by
-simple statistics.  If a query point is far away from a given quadrant, the
+approach splits the space into quadrants, and simple statistics may summarize entire regions.  If a query point is far away from a given quadrant, the
 repulsive forces exerted by all the points in that quadrant onto the query point
@@
 reference embedding remains fixed throughout the optimization of newly added
-points, and precompute the interpolation grid. This further reduces the runtime
+points and precompute the interpolation grid. This further reduces the runtime
 complexity from $\mathcal{O}(\max \{ N, M \})$ to $\mathcal{O}(M)$.
@@
 Combining these with a more narrow kernel with perplexity 50 prevents small,
-well-defined clusters from being absorbed into larger ones, leading a better
-overall snapshot into the data structure.
+well-defined clusters from being absorbed into larger ones, leading to an overall better
+insight into the data structure.
 
@@
 Standard t-SNE reveals the clustering structure at a single level of resolution.
-Different perplexity parameter values can be used to identify global cluster
+We can use different perplexity parameter values to identify global cluster
 relationships or small, well-isolated groups. Unfortunately, varying perplexity
-values can be time-consuming as this involves recomputing the $k$-nearest
-neighbor graph which is often the most expensive part of the t-SNE algorithm.
+values can be time-consuming as this involves recomputing the $k$-nearest neighbor graph, which is often the most expensive part of the t-SNE algorithm.
 Alternatively, Kobak \textit{et al.} suggest that varying the degree of freedom
@@
 
-Embeddings produced by standard t-SNE often produce clusters separated by thin
-boundaries and make use of all available space. While this is desirable at
-times, it often obscures the global relationships between clusters as all
-neighboring clusters appear about the same distance from one another. Other
-dimensionality reduction methods such as UMAP~\cite{2018arXivUMAP} or
-ForceAtlas2~\cite{jacomy2014forceatlas2} tend to produce embeddings where
-clusters appear more compact and the white-space separating the clusters may be
-interpreted at least partially as a loose measure of distance.
+Embeddings produced by standard t-SNE often have clusters separated by
+thin boundaries and use all available space. While this is desirable,
+it often obscures the global relationships between clusters as all
+neighboring clusters appear about the same distance from one
+another. Other dimensionality reduction methods such as
+UMAP~\cite{2018arXivUMAP} or ForceAtlas2~\cite{jacomy2014forceatlas2}
+tend to produce embeddings where clusters appear more compact and the
+white-space separating the clusters may be interpreted at least
+partially as a loose measure of distance.
 
-B{\"o}hm \textit{et al.} showed that the exaggeration factor $\rho$ can be used
-to produce layouts more similar to UMAP and ForceAtlas2~\cite{bohm2020unifying}.
-By incorporating exaggeration into later phases of optimization, t-SNE is able
-introduce more white-space between clusters which better reflects the global
-relations between clusters. B{\"o}hm \textit{et al.} found that using $\rho=4$
-and $\rho=30$ produces embeddings visually similar to UMAP and ForceAtlas2,
-respectively.
+B{\"o}hm \textit{et al.} showed that the exaggeration factor $\rho$
+could be used to produce layouts more similar to UMAP and
+ForceAtlas2~\cite{bohm2020unifying}. By incorporating exaggeration
+into later phases of optimization, t-SNE introduces more white-space
+between clusters which better reflects the global relations between
+clusters. B{\"o}hm \textit{et al.} found that using $\rho=4$ and
+$\rho=30$ produces embeddings visually similar to UMAP and
+ForceAtlas2, respectively.
 
-While it is difficult to claim one is better than the other, different settings
-of $\rho$ may uncover different properties of the data manifold. For example, in
-single-cell data, standard t-SNE often uncovers distinct, well-separated groups
-of cell-types but obscures transitional paths between cell-states. This problem
-is exacerbated when dealing with large numbers of data points. On the other
-hand, ForceAtlas2 has been used successfuly to uncover trajectories and
-transitions between cell types. Unfortunately, when using ForceAtlas2, large
-clusters often absorb small, distinct groups of cells.  One might run t-SNE with
-various parameter settings to avoid switching between different algorithms and
-obtain similar visualizations.  We might generate an embedding using higher
-levels of exaggeration to highlight the developmental transitions between cell
-types and one with low levels of exaggeration to identify clear populations of
-cells.
+While it is difficult to claim one is better than the other, different
+$\rho$ may uncover different properties of the data manifold. For
+example, standard t-SNE often exposes distinct, well-separated
+cell-types in single-cell data but obscures transitional paths between
+cell-states. This problem is exacerbated when dealing with large
+numbers of data points. On the other hand, ForceAtlas2 has been used
+successfully to uncover trajectories and transitions between cell
+types. Unfortunately, when using ForceAtlas2, large clusters often
+absorb small, distinct groups of cells. One might run t-SNE with
+various parameter settings to avoid switching between different
+algorithms and obtain similar visualizations. We might generate an
+embedding using higher levels of exaggeration to highlight the
+developmental transitions between cell types and one with low levels
+of exaggeration to identify clear populations of cells.
 
@@
 t-SNE, UMAP, and ForceAtlas2 can all be cast as force-directed layout
-algorithms, which operate on the $k$-nearest neighbor graph. Each method
-constructs the graph differently and specifies the attractive and repulsive
-forces in its own way, but ultimately, point positions are found by balancing
-the attractive and repulsive forces between data points.
+algorithms, which operate on the $k$-nearest neighbor graph. Each
+method constructs the graph differently and specifies the attractive
+and repulsive forces in its way. Still, ultimately, point positions
+are found by balancing the attractive and repulsive forces between
+data points.
 
-The final positions of data points in this class of algorithms is largely
-dependent on the embedding initialization. UMAP found early success as it was
-able to produce visualizations that better captured the global organization of
-clusters. This was due to its initialization scheme, which initialized point
-positions using Laplacian eigenmaps. On the other hand, most implementations of
-t-SNE performed random initialization, which resulted in poor global coherence.
-However, Kobak \& Linderman recently showed that using the same initialization
-for both methods leads to visualizations, which exhibit similarly good global
-coherency. Conversely, when initialized randomly, both methods produce
-visualizations where clusters are arbitrarily positioned in the embedding space.
+The final positions of data points in this class of algorithms are
+mainly dependent on the embedding initialization. UMAP found early
+success as it produced visualizations that better captured the global
+organization of clusters. This was due to its initialization scheme,
+which initialized point positions using Laplacian eigenmaps. On the
+other hand, most implementations of t-SNE performed random
+initialization, which resulted in poor global coherence. However,
+Kobak \& Linderman recently showed that using the same initialization
+for both methods leads to visualizations, which exhibit similarly good
+global coherency. Conversely, when initialized randomly, both methods
+produce visualizations where clusters are arbitrarily positioned in
+the embedding space.
 
 \opentsne\ defaults to using the two leading principal components as
-initialization, but also provides a spectral approach similar to UMAP.  This
-allows clusters in resulting embeddings to be organized in a more globally
-coherent manner, leading to increased visualization interpretability.
+initialization, and besides, provides a spectral approach similar to
+UMAP. This allows clusters in resulting embeddings to be organized in
+a more globally coherent manner, leading to increased visualization
+interpretability.
 

### 2021-03-16 paper/cover.tex
cover letter

\documentclass[11pt]{article}
\setlength{\parindent}{0em}
\setlength{\parskip}{1em}
\renewcommand{\baselinestretch}{1.1}
\pagestyle{empty}
\usepackage[margin=1in]{geometry}

\begin{document}
\begin{flushright}
Ljubljana, March 17th, 2021
\end{flushright}

\begin{flushleft}
{\bf Dear Editor,}

We would like to submit our manuscript on a modular Python toolbox for t-SNE dimensionality reduction and embedding.

Science, according to Wikipedia, is ``a systematic enterprise that builds and organizes knowledge''. One of the earliest ways means to information organization is a map. Today, an important subfield of data science strives to map multidimensional data onto a two-dimensional plane to expose the structure, relations, provide predictions of functions, and uncover temporal trends. A prominent example of such a technique is t-SNE. For instance, t-SNE can embed expression-profiled single cells into a two-dimensional map, exposing same-typed clusters or charting cell development. Over 1,400 articles from Nature journals alone used t-SNE for data visualization, and Nature Methods has published many papers, including those with t-SNE depiction of single-cell data maps.

In the paper, we report on openTSNE, a Python-based open-source library for t-SNE visualization. openTSNE runs orders of magnitudes faster than comparable Python-based implementations (e.g., scikit-learn) and can handle data sets containing millions of data points. t-SNE has been recently criticized for poor scalability and susceptibility to batch effects. Our proposed implementation addresses all these concerns and includes the lately-proposed methodological advancements, which we review in the paper's online methods section.

Additionally, openTSNE is currently the only t-SNE library that can place new data points into a constructed embedding. In the paper, we show that such embedding of new data effectively mitigates batch effects.

The t-SNE implementation we are submitting a report on is open, available through GitHub, and fosters extensibility and experimentation. The library has already attracted substantial interest within the Python community: it gained 758 GitHub stars, on par with prominent data science packages in Python, such as scanpy for single-cell analysis (854 stars).

We are submitting an original manuscript that has not been considered for publication before. Lin Tang would be an excellent choice for Editor for this manuscript because of his broad knowledge in the field and work in single-cell analytics. For reviewers, we kindly propose Dmitry Kobak (U Tuebingen), an author of recently proposed t-SNE tricks that we implemented in the library, Barbara di Camillo (U Padova), single-cell data scientists, or Cagatay Turkay (U Warwick), a specialist in explainable data visualizations.

Yours faithfully, \\
Pavlin G. Poličar, Martin Stražar, Blaž Zupan
\end{flushleft}
\end{document}

### 2021-03-17 paper/cover.tex
corrected typos

@@
 Science, according to Wikipedia, is ``a systematic enterprise that builds and organizes knowledge''.
-One of the earliest ways means to information organization is a map.
+One of the earliest means of information organization is a map.
 Today, an important subfield of data science strives to map multidimensional data onto a two-dimensional plane to expose the structure, relations, provide predictions of functions, and uncover temporal trends.
 A prominent example of such a technique is t-SNE.
-For instance, t-SNE can embed expression-profiled single cells into a two-dimensional map, exposing same-typed clusters or charting cell development.
-Over 1,400 articles from Nature journals alone used t-SNE for data visualization, and Nature Methods has published many papers, including those with t-SNE depiction of single-cell data maps.
+For instance, t-SNE can embed expression-profiled single cells into a two-dimensional map, exposing regions of same-typed cells or charting cell development.
+Over 1,400 articles from Nature journals alone recently used t-SNE for data visualization.
+Nature Methods has published a number of papers that use t-SNE, including those with a t-SNE depiction of single-cell data maps.
+t-SNE visualizations have appeared on the magazines' covers and have been embraced by the entire data science community.
 
@@
 t-SNE has been recently criticized for poor scalability and susceptibility to batch effects.
-Our proposed implementation addresses all these concerns and includes the lately-proposed methodological advancements, which we review in the paper's online methods section.
+Our proposed implementation addresses these concerns and includes the lately-proposed methodological advancements, which we review in the paper's online methods section.
 
 Additionally, openTSNE is currently the only t-SNE library that can place new data points into a constructed embedding.
-In the paper, we show that such embedding of new data effectively mitigates batch effects.
+The paper shows that such embedding of new data effectively mitigates batch effects and is an approach for visual prediction.
 
-The t-SNE implementation we are submitting a report on is open, available through GitHub, and fosters extensibility and experimentation.
-The library has already attracted substantial interest within the Python community: it gained 758 GitHub stars, on par with prominent data science packages in Python, such as scanpy for single-cell analysis (854 stars).
+Our t-SNE implementation is open, available on GitHub, and fosters extensibility and experimentation.
+The library has already attracted significant interest within the Python community; it gained 758 GitHub stars, on par with prominent data science packages in Python, such as scanpy for single-cell analysis (854 stars).
 

### 2021-03-17 paper/cover.tex
small corrections

@@
 
-In the paper, we report on openTSNE, a Python-based open-source library for t-SNE visualization.
+In the paper, we report on openTSNE, a Python-based open-source library for t-SNE.
 openTSNE runs orders of magnitudes faster than comparable Python-based implementations (e.g., scikit-learn) and can handle data sets containing millions of data points.
@@
 We are submitting an original manuscript that has not been considered for publication before.
-Lin Tang would be an excellent choice for Editor for this manuscript because of his broad knowledge in the field and work in single-cell analytics.
-For reviewers, we kindly propose Dmitry Kobak (U Tuebingen), an author of recently proposed t-SNE tricks that we implemented in the library, Barbara di Camillo (U Padova), single-cell data scientists, or Cagatay Turkay (U Warwick), a specialist in explainable data visualizations.
+{\bf Lin Tang} would be an excellent choice for an editor for this manuscript because of her broad knowledge in the field and work in single-cell analytics.
+For reviewers, we kindly propose {\bf Dmitry Kobak} (U Tuebingen), an author of recently proposed t-SNE tricks that we implemented in the library, {\bf Barbara di Camillo} (U Padova), single-cell data scientists, {\bf Marinka Žitnik} (Harvard), a data integration expert, or {\bf Cagatay Turkay} (U Warwick), a specialist in explainable data visualizations.
 

### 2021-03-17 paper/policar2020-opentsne.tex
emails
grant info

@@
   Martin Stra\v{zar}\textsuperscript{2}\\
+  \texttt{martin.strazar@gmail.com}\\
   \and
   Bla\v{z} Zupan\textsuperscript{1,3}\\
+  \texttt{blaz.zupan@fri.uni-lj.si}\\
 }
@@
 \section*{Competing interests}
-The authors declare that they have no competing interests.
+The authors declare no competing interests.
 
@@
 \section*{Acknowledgements}
+This work was supported by Slovenian Research Agency (P2-0209).
 We want to thank Dmitry Kobak for various helpful discussions and best practices

### 2021-03-17 paper/policar2020-opentsne.tex
minor fixes

@@
 techniques to construct such visualizations is $t$-distributed stochastic
-neighbor embedding (t-SNE). Several extensions of t-SNE have
+neighbor embedding (t-SNE). Several extensions of t\nobreakdash-SNE have
 recently been proposed to address issues of scalability and the quality
 of the resulting visualizations. We introduce \opentsne, a modular
-Python library that implements the core t-SNE algorithm and its many
+Python library that implements the core t\nobreakdash-SNE algorithm and its many
 extensions.
@@
 scalability, lack of global organization -- t-SNE identifies well-defined clusters that
-may be arbitrarily scattered throguhout the low-dimensional space -- and the absence of
+may be arbitrarily scattered throughout the low-dimensional space -- and the absence of
 theoretically-founded methods to map new data into existing
@@
 the time of the writing, the library averages thousands of weekly
-downloads and has received almost 700 GitHub stars, an appreciation score gained
+downloads and has received over 750 GitHub stars, an appreciation score gained
 by only $0.3\%$ of public Python repositories. The source code is available at
@@
 
-Accessibility of the latest theoretical advancments in t-SNE is one of the core
-design principles of \opentsne. This range of advancments is easily accessible through an intuitive
+Accessibility of the latest theoretical advancements in t-SNE is one of the core
+design principles of \opentsne. This range of advancements is easily accessible through an intuitive
 programming interface, where we closely follow the style
@@
 and was used to generate the four visualizations in Fig.~\ref{fig:tasic}, demonstrating the
-effect of a particular recent theoretical advancments in the t-SNE algorithm.
+effect of a particular recent theoretical advancements in the t-SNE algorithm.
 
@@
 component analysis or autoencoders, t-SNE is a non-parametric method and does
-not define an explicit mapping to the embedding space. Therefore embeddings of
+not define an explicit mapping to the embedding space. Therefore, embeddings of
 new data points need to be found through

### 2021-03-17 paper/policar2020-opentsne.tex
Embed bibliography into tex file

@@
 
-\bibliographystyle{bmc-mathphys} % Style BST file (bmc-mathphys, vancouver, spbasic).
-\bibliography{references}      % Bibliography file (usually '*.bib' )
+%% Bibliography
+\begin{thebibliography}{28}
+% BibTex style file: bmc-mathphys.bst (version 2.1), 2014-07-24
+\ifx \bisbn   \undefined \def \bisbn  #1{ISBN #1}\fi
+\ifx \binits  \undefined \def \binits#1{#1}\fi
+\ifx \bauthor  \undefined \def \bauthor#1{#1}\fi
+\ifx \batitle  \undefined \def \batitle#1{#1}\fi
+\ifx \bjtitle  \undefined \def \bjtitle#1{#1}\fi
+\ifx \bvolume  \undefined \def \bvolume#1{\textbf{#1}}\fi
+\ifx \byear  \undefined \def \byear#1{#1}\fi
+\ifx \bissue  \undefined \def \bissue#1{#1}\fi
+\ifx \bfpage  \undefined \def \bfpage#1{#1}\fi
+\ifx \blpage  \undefined \def \blpage #1{#1}\fi
+\ifx \burl  \undefined \def \burl#1{\textsf{#1}}\fi
+\ifx \doiurl  \undefined \def \doiurl#1{\textsf{#1}}\fi
+\ifx \betal  \undefined \def \betal{\textit{et al.}}\fi
+\ifx \binstitute  \undefined \def \binstitute#1{#1}\fi
+\ifx \binstitutionaled  \undefined \def \binstitutionaled#1{#1}\fi
+\ifx \bctitle  \undefined \def \bctitle#1{#1}\fi
+\ifx \beditor  \undefined \def \beditor#1{#1}\fi
+\ifx \bpublisher  \undefined \def \bpublisher#1{#1}\fi
+\ifx \bbtitle  \undefined \def \bbtitle#1{#1}\fi
+\ifx \bedition  \undefined \def \bedition#1{#1}\fi
+\ifx \bseriesno  \undefined \def \bseriesno#1{#1}\fi
+\ifx \blocation  \undefined \def \blocation#1{#1}\fi
+\ifx \bsertitle  \undefined \def \bsertitle#1{#1}\fi
+\ifx \bsnm \undefined \def \bsnm#1{#1}\fi
+\ifx \bsuffix \undefined \def \bsuffix#1{#1}\fi
+\ifx \bparticle \undefined \def \bparticle#1{#1}\fi
+\ifx \barticle \undefined \def \barticle#1{#1}\fi
+\ifx \bconfdate \undefined \def \bconfdate #1{#1}\fi
+\ifx \botherref \undefined \def \botherref #1{#1}\fi
+\ifx \url \undefined \def \url#1{\textsf{#1}}\fi
+\ifx \bchapter \undefined \def \bchapter#1{#1}\fi
+\ifx \bbook \undefined \def \bbook#1{#1}\fi
+\ifx \bcomment \undefined \def \bcomment#1{#1}\fi
+\ifx \oauthor \undefined \def \oauthor#1{#1}\fi
+\ifx \citeauthoryear \undefined \def \citeauthoryear#1{#1}\fi
+\ifx \endbibitem  \undefined \def \endbibitem {}\fi
+\ifx \bconflocation  \undefined \def \bconflocation#1{#1}\fi
+\ifx \arxivurl  \undefined \def \arxivurl#1{\textsf{#1}}\fi
+\csname PreBibitemsHook\endcsname
+
+%%% 1
+\bibitem{maaten2008visualizing}
+\begin{barticle}
+\bauthor{\bsnm{Van Der~Maaten}, \binits{L.}},
+\bauthor{\bsnm{Hinton}, \binits{G.}}:
+\batitle{{Visualizing data using t-SNE}}.
+\bjtitle{Journal of Machine Learning Research}
+\bvolume{9}(\bissue{Nov}),
+\bfpage{2579}--\blpage{2605}
+(\byear{2008})
+\end{barticle}
+\endbibitem
+
+%%% 2
+\bibitem{2018arXivUMAP}
+\begin{botherref}
+\oauthor{\bsnm{{McInnes}}, \binits{L.}},
+\oauthor{\bsnm{{Healy}}, \binits{J.}},
+\oauthor{\bsnm{{Melville}}, \binits{J.}}:
+{UMAP: Uniform Manifold Approximation and Projection for Dimension Reduction}.
+ArXiv e-prints
+(2018).
+\arxivurl{1802.03426}
+\end{botherref}
+\endbibitem
+
+%%% 3
+\bibitem{macosko2015highly}
+\begin{barticle}
+\bauthor{\bsnm{Macosko}, \binits{E.Z.}},
+\bauthor{\bsnm{Basu}, \binits{A.}},
+\bauthor{\bsnm{Satija}, \binits{R.}},
+\bauthor{\bsnm{Nemesh}, \binits{J.}},
+\bauthor{\bsnm{Shekhar}, \binits{K.}},
+\bauthor{\bsnm{Goldman}, \binits{M.}},
+\bauthor{\bsnm{Tirosh}, \binits{I.}},
+\bauthor{\bsnm{Bialas}, \binits{A.R.}},
+\bauthor{\bsnm{Kamitaki}, \binits{N.}},
+\bauthor{\bsnm{Martersteck}, \binits{E.M.}}, \betal:
+\batitle{Highly parallel genome-wide expression profiling of individual cells
+  using nanoliter droplets}.
+\bjtitle{Cell}
+\bvolume{161}(\bissue{5}),
+\bfpage{1202}--\blpage{1214}
+(\byear{2015})
+\end{barticle}
+\endbibitem
+
+%%% 4
+\bibitem{cao2019single}
+\begin{barticle}
+\bauthor{\bsnm{Cao}, \binits{J.}},
+\bauthor{\bsnm{Spielmann}, \binits{M.}},
+\bauthor{\bsnm{Qiu}, \binits{X.}},
+\bauthor{\bsnm{Huang}, \binits{X.}},
+\bauthor{\bsnm{Ibrahim}, \binits{D.M.}},
+\bauthor{\bsnm{Hill}, \binits{A.J.}},
+\bauthor{\bsnm{Zhang}, \binits{F.}},
+\bauthor{\bsnm{Mundlos}, \binits{S.}},
+\bauthor{\bsnm{Christiansen}, \binits{L.}},
+\bauthor{\bsnm{Steemers}, \binits{F.J.}}, \betal:
+\batitle{The single-cell transcriptional landscape of mammalian organogenesis}.
+\bjtitle{Nature}
+\bvolume{566}(\bissue{7745}),
+\bfpage{496}--\blpage{502}
+(\byear{2019})
+\end{barticle}
+\endbibitem
+
+%%% 5
+\bibitem{tasic2018shared}
+\begin{barticle}
+\bauthor{\bsnm{Tasic}, \binits{B.}},
+\bauthor{\bsnm{Yao}, \binits{Z.}},
+\bauthor{\bsnm{Graybuck}, \binits{L.T.}},
+\bauthor{\bsnm{Smith}, \binits{K.A.}},
+\bauthor{\bsnm{Nguyen}, \binits{T.N.}},
+\bauthor{\bsnm{Bertagnolli}, \binits{D.}},
+\bauthor{\bsnm{Goldy}, \binits{J.}},
+\bauthor{\bsnm{Garren}, \binits{E.}},
+\bauthor{\bsnm{Economo}, \binits{M.N.}},
+\bauthor{\bsnm{Viswanathan}, \binits{S.}}, \betal:
+\batitle{Shared and distinct transcriptomic cell types across neocortical
+  areas}.
+\bjtitle{Nature}
+\bvolume{563}(\bissue{7729}),
+\bfpage{72}--\blpage{78}
+(\byear{2018})
+\end{barticle}
+\endbibitem
+
+%%% 6
+\bibitem{hirata2019genetic}
+\begin{barticle}
+\bauthor{\bsnm{Hirata}, \binits{J.}},
+\bauthor{\bsnm{Hosomichi}, \binits{K.}},
+\bauthor{\bsnm{Sakaue}, \binits{S.}},
+\bauthor{\bsnm{Kanai}, \binits{M.}},
+\bauthor{\bsnm{Nakaoka}, \binits{H.}},
+\bauthor{\bsnm{Ishigaki}, \binits{K.}},
+\bauthor{\bsnm{Suzuki}, \binits{K.}},
+\bauthor{\bsnm{Akiyama}, \binits{M.}},
+\bauthor{\bsnm{Kishikawa}, \binits{T.}},
+\bauthor{\bsnm{Ogawa}, \binits{K.}}, \betal:
+\batitle{{Genetic and phenotypic landscape of the major histocompatibilty
+  complex region in the Japanese population}}.
+\bjtitle{Nature Genetics}
+\bvolume{51}(\bissue{3}),
+\bfpage{470}--\blpage{480}
+(\byear{2019})
+\end{barticle}
+\endbibitem
+
+%%% 7
+\bibitem{beaulaurier2018metagenomic}
+\begin{barticle}
+\bauthor{\bsnm{Beaulaurier}, \binits{J.}},
+\bauthor{\bsnm{Zhu}, \binits{S.}},
+\bauthor{\bsnm{Deikus}, \binits{G.}},
+\bauthor{\bsnm{Mogno}, \binits{I.}},
+\bauthor{\bsnm{Zhang}, \binits{X.-S.}},
+\bauthor{\bsnm{Davis-Richardson}, \binits{A.}},
+\bauthor{\bsnm{Canepa}, \binits{R.}},
+\bauthor{\bsnm{Triplett}, \binits{E.W.}},
+\bauthor{\bsnm{Faith}, \binits{J.J.}},
+\bauthor{\bsnm{Sebra}, \binits{R.}}, \betal:
+\batitle{{Metagenomic binning and association of plasmids with bacterial host
+  genomes using DNA methylation}}.
+\bjtitle{Nature Biotechnology}
+\bvolume{36}(\bissue{1}),
+\bfpage{61}
+(\byear{2018})
+\end{barticle}
+\endbibitem
+
+%%% 8
+\bibitem{sheth2019spatial}
+\begin{barticle}
+\bauthor{\bsnm{Sheth}, \binits{R.U.}},
+\bauthor{\bsnm{Li}, \binits{M.}},
+\bauthor{\bsnm{Jiang}, \binits{W.}},
+\bauthor{\bsnm{Sims}, \binits{P.A.}},
+\bauthor{\bsnm{Leong}, \binits{K.W.}},
+\bauthor{\bsnm{Wang}, \binits{H.H.}}:
+\batitle{Spatial metagenomic characterization of microbial biogeography in the
+  gut}.
+\bjtitle{Nature Biotechnology}
+\bvolume{37}(\bissue{8}),
+\bfpage{877}--\blpage{883}
+(\byear{2019})
+\end{barticle}
+\endbibitem
+
+%%% 9
+\bibitem{tkachev2019differences}
+\begin{barticle}
+\bauthor{\bsnm{Tkachev}, \binits{A.}},
+\bauthor{\bsnm{Stepanova}, \binits{V.}},
+\bauthor{\bsnm{Zhang}, \binits{L.}},
+\bauthor{\bsnm{Khrameeva}, \binits{E.}},
+\bauthor{\bsnm{Zubkov}, \binits{D.}},
+\bauthor{\bsnm{Giavalisco}, \binits{P.}},
+\bauthor{\bsnm{Khaitovich}, \binits{P.}}:
+\batitle{Differences in lipidome and metabolome organization of prefrontal
+  cortex among human populations}.
+\bjtitle{Scientific Reports}
+\bvolume{9}(\bissue{1}),
+\bfpage{1}--\blpage{10}
+(\byear{2019})
+\end{barticle}
+\endbibitem
+
+%%% 10
+\bibitem{shekhar2016comprehensive}
+\begin{barticle}
+\bauthor{\bsnm{Shekhar}, \binits{K.}},
+\bauthor{\bsnm{Lapan}, \binits{S.W.}},
+\bauthor{\bsnm{Whitney}, \binits{I.E.}},
+\bauthor{\bsnm{Tran}, \binits{N.M.}},
+\bauthor{\bsnm{Macosko}, \binits{E.Z.}},
+\bauthor{\bsnm{Kowalczyk}, \binits{M.}},
+\bauthor{\bsnm{Adiconis}, \binits{X.}},
+\bauthor{\bsnm{Levin}, \binits{J.Z.}},
+\bauthor{\bsnm{Nemesh}, \binits{J.}},
+\bauthor{\bsnm{Goldman}, \binits{M.}}, \betal:
+\batitle{Comprehensive classification of retinal bipolar neurons by single-cell
+  transcriptomics}.
+\bjtitle{Cell}
+\bvolume{166}(\bissue{5}),
+\bfpage{1308}--\blpage{1323}
+(\byear{2016})
+\end{barticle}
+\endbibitem
+
+%%% 11
+\bibitem{ding2018interpretable}
+\begin{barticle}
+\bauthor{\bsnm{Ding}, \binits{J.}},
+\bauthor{\bsnm{Condon}, \binits{A.}},
+\bauthor{\bsnm{Shah}, \binits{S.P.}}:
+\batitle{Interpretable dimensionality reduction of single cell transcriptome
+  data with deep generative models}.
+\bjtitle{Nature Communications}
+\bvolume{9}(\bissue{1}),
+\bfpage{2002}
+(\byear{2018})
+\end{barticle}
+\endbibitem
+
+%%% 12
+\bibitem{becht2019dimensionality}
+\begin{barticle}
+\bauthor{\bsnm{Becht}, \binits{E.}},
+\bauthor{\bsnm{McInnes}, \binits{L.}},
+\bauthor{\bsnm{Healy}, \binits{J.}},
+\bauthor{\bsnm{Dutertre}, \binits{C.-A.}},
+\bauthor{\bsnm{Kwok}, \binits{I.W.}},
+\bauthor{\bsnm{Ng}, \binits{L.G.}},
+\bauthor{\bsnm{Ginhoux}, \binits{F.}},
+\bauthor{\bsnm{Newell}, \binits{E.W.}}:
+\batitle{{Dimensionality reduction for visualizing single-cell data using
+  UMAP}}.
+\bjtitle{Nature Biotechnology}
+\bvolume{37}(\bissue{1}),
+\bfpage{38}
+(\byear{2019})
+\end{barticle}
+\endbibitem
+
+%%% 13
+\bibitem{linderman2019fast}
+\begin{barticle}
+\bauthor{\bsnm{Linderman}, \binits{G.C.}},
+\bauthor{\bsnm{Rachh}, \binits{M.}},
+\bauthor{\bsnm{Hoskins}, \binits{J.G.}},
+\bauthor{\bsnm{Steinerberger}, \binits{S.}},
+\bauthor{\bsnm{Kluger}, \binits{Y.}}:
+\batitle{{Fast interpolation-based t-SNE for improved visualization of
+  single-cell RNA-seq data}}.
+\bjtitle{Nature Methods}
+\bvolume{16}(\bissue{3}),
+\bfpage{243}--\blpage{245}
+(\byear{2019})
+\end{barticle}
+\endbibitem
+
+%%% 14
+\bibitem{kobak2019art}
+\begin{barticle}
+\bauthor{\bsnm{Kobak}, \binits{D.}},
+\bauthor{\bsnm{Berens}, \binits{P.}}:
+\batitle{{The art of using t-SNE for single-cell transcriptomics}}.
+\bjtitle{Nature Communications}
+\bvolume{10}(\bissue{1}),
+\bfpage{1}--\blpage{14}
+(\byear{2019})
+\end{barticle}
+\endbibitem
+
+%%% 15
+\bibitem{policar2019embedding}
+\begin{bchapter}
+\bauthor{\bsnm{Poli{\v{c}}ar}, \binits{P.G.}},
+\bauthor{\bsnm{Stra{\v{z}}ar}, \binits{M.}},
+\bauthor{\bsnm{Zupan}, \binits{B.}}:
+\bctitle{{Embedding to reference t-SNE space addresses batch effects in
+  single-cell classification}}.
+In: \bbtitle{International Conference on Discovery Science},
+pp. \bfpage{246}--\blpage{260}
+(\byear{2019}).
+\bcomment{Springer}
+\end{bchapter}
+\endbibitem
+
+%%% 16
+\bibitem{van2014accelerating}
+\begin{barticle}
+\bauthor{\bsnm{Van Der~Maaten}, \binits{L.}}:
+\batitle{{Accelerating t-SNE using tree-based algorithms}}.
+\bjtitle{The Journal of Machine Learning Research}
+\bvolume{15}(\bissue{1}),
+\bfpage{3221}--\blpage{3245}
+(\byear{2014})
+\end{barticle}
+\endbibitem
+
+%%% 17
+\bibitem{kobak2019umap}
+\begin{botherref}
+\oauthor{\bsnm{Kobak}, \binits{D.}},
+\oauthor{\bsnm{Linderman}, \binits{G.C.}}:
+{UMAP does not preserve global structure any better than t-SNE when using the
+  same initialization}.
+bioRxiv
+(2019)
+\end{botherref}
+\endbibitem
+
+%%% 18
+\bibitem{kobak2019heavy}
+\begin{bchapter}
+\bauthor{\bsnm{Kobak}, \binits{D.}},
+\bauthor{\bsnm{Linderman}, \binits{G.}},
+\bauthor{\bsnm{Steinerberger}, \binits{S.}},
+\bauthor{\bsnm{Kluger}, \binits{Y.}},
+\bauthor{\bsnm{Berens}, \binits{P.}}:
+\bctitle{{Heavy-tailed kernels reveal a finer cluster structure in t-SNE
+  visualisations}}.
+In: \bbtitle{Joint European Conference on Machine Learning and Knowledge
+  Discovery in Databases},
+pp. \bfpage{124}--\blpage{139}
+(\byear{2019}).
+\bcomment{Springer}
+\end{bchapter}
+\endbibitem
+
+%%% 19
+\bibitem{belkina2019automated}
+\begin{barticle}
+\bauthor{\bsnm{Belkina}, \binits{A.C.}},
+\bauthor{\bsnm{Ciccolella}, \binits{C.O.}},
+\bauthor{\bsnm{Anno}, \binits{R.}},
+\bauthor{\bsnm{Halpert}, \binits{R.}},
+\bauthor{\bsnm{Spidlen}, \binits{J.}},
+\bauthor{\bsnm{Snyder-Cappione}, \binits{J.E.}}:
+\batitle{{Automated optimized parameters for T-distributed stochastic neighbor
+  embedding improve visualization and analysis of large datasets}}.
+\bjtitle{Nature Communications}
+\bvolume{10}(\bissue{1}),
+\bfpage{1}--\blpage{12}
+(\byear{2019})
+\end{barticle}
+\endbibitem
+
+%%% 20
+\bibitem{sklearn_api}
+\begin{bchapter}
+\bauthor{\bsnm{Buitinck}, \binits{L.}},
+\bauthor{\bsnm{Louppe}, \binits{G.}},
+\bauthor{\bsnm{Blondel}, \binits{M.}},
+\bauthor{\bsnm{Pedregosa}, \binits{F.}},
+\bauthor{\bsnm{Mueller}, \binits{A.}},
+\bauthor{\bsnm{Grisel}, \binits{O.}},
+\bauthor{\bsnm{Niculae}, \binits{V.}},
+\bauthor{\bsnm{Prettenhofer}, \binits{P.}},
+\bauthor{\bsnm{Gramfort}, \binits{A.}},
+\bauthor{\bsnm{Grobler}, \binits{J.}},
+\bauthor{\bsnm{Layton}, \binits{R.}},
+\bauthor{\bsnm{VanderPlas}, \binits{J.}},
+\bauthor{\bsnm{Joly}, \binits{A.}},
+\bauthor{\bsnm{Holt}, \binits{B.}},
+\bauthor{\bsnm{Varoquaux}, \binits{G.}}:
+\bctitle{{API} design for machine learning software: experiences from the
+  scikit-learn project}.
+In: \bbtitle{ECML PKDD Workshop: Languages for Data Mining and Machine
+  Learning},
+pp. \bfpage{108}--\blpage{122}
+(\byear{2013})
+\end{bchapter}
+\endbibitem
+
+%%% 21
+\bibitem{jacomy2014forceatlas2}
+\begin{barticle}
+\bauthor{\bsnm{Jacomy}, \binits{M.}},
+\bauthor{\bsnm{Venturini}, \binits{T.}},
+\bauthor{\bsnm{Heymann}, \binits{S.}},
+\bauthor{\bsnm{Bastian}, \binits{M.}}:
+\batitle{{ForceAtlas2, a continuous graph layout algorithm for handy network
+  visualization designed for the Gephi software}}.
+\bjtitle{PloS One}
+\bvolume{9}(\bissue{6}),
+\bfpage{98679}
+(\byear{2014})
+\end{barticle}
+\endbibitem
+
+%%% 22
+\bibitem{bohm2020unifying}
+\begin{botherref}
+\oauthor{\bsnm{B{\"o}hm}, \binits{J.N.}},
+\oauthor{\bsnm{Berens}, \binits{P.}},
+\oauthor{\bsnm{Kobak}, \binits{D.}}:
+{A Unifying Perspective on Neighbor Embeddings along the Attraction-Repulsion
+  Spectrum}.
+arXiv preprint arXiv:2007.08902
+(2020)
+\end{botherref}
+\endbibitem
+
+%%% 23
+\bibitem{hochgerner2018conserved}
+\begin{barticle}
+\bauthor{\bsnm{Hochgerner}, \binits{H.}},
+\bauthor{\bsnm{Zeisel}, \binits{A.}},
+\bauthor{\bsnm{L{\"o}nnerberg}, \binits{P.}},
+\bauthor{\bsnm{Linnarsson}, \binits{S.}}:
+\batitle{{Conserved properties of dentate gyrus neurogenesis across postnatal
+  development revealed by single-cell RNA sequencing}}.
+\bjtitle{Nature Neuroscience}
+\bvolume{21}(\bissue{2}),
+\bfpage{290}--\blpage{299}
+(\byear{2018})
+\end{barticle}
+\endbibitem
+
+%%% 24
+\bibitem{harris2018classes}
+\begin{barticle}
+\bauthor{\bsnm{Harris}, \binits{K.D.}},
+\bauthor{\bsnm{Hochgerner}, \binits{H.}},
+\bauthor{\bsnm{Skene}, \binits{N.G.}},
+\bauthor{\bsnm{Magno}, \binits{L.}},
+\bauthor{\bsnm{Katona}, \binits{L.}},
+\bauthor{\bsnm{Gonzales}, \binits{C.B.}},
+\bauthor{\bsnm{Somogyi}, \binits{P.}},
+\bauthor{\bsnm{Kessaris}, \binits{N.}},
+\bauthor{\bsnm{Linnarsson}, \binits{S.}},
+\bauthor{\bsnm{Hjerling-Leffler}, \binits{J.}}:
+\batitle{{Classes and continua of hippocampal CA1 inhibitory neurons revealed
+  by single-cell transcriptomics}}.
+\bjtitle{PLoS Biology}
+\bvolume{16}(\bissue{6}),
+\bfpage{2006387}
+(\byear{2018})
+\end{barticle}
+\endbibitem
+
+%%% 25
+\bibitem{Ulyanov2016}
+\begin{botherref}
+\oauthor{\bsnm{Ulyanov}, \binits{D.}}:
+{Multicore-TSNE}.
+GitHub
+(2016)
+\end{botherref}
+\endbibitem
+
+%%% 26
+\bibitem{pedregosa2011scikit}
+\begin{barticle}
+\bauthor{\bsnm{Pedregosa}, \binits{F.}},
+\bauthor{\bsnm{Varoquaux}, \binits{G.}},
+\bauthor{\bsnm{Gramfort}, \binits{A.}},
+\bauthor{\bsnm{Michel}, \binits{V.}},
+\bauthor{\bsnm{Thirion}, \binits{B.}},
+\bauthor{\bsnm{Grisel}, \binits{O.}},
+\bauthor{\bsnm{Blondel}, \binits{M.}},
+\bauthor{\bsnm{Prettenhofer}, \binits{P.}},
+\bauthor{\bsnm{Weiss}, \binits{R.}},
+\bauthor{\bsnm{Dubourg}, \binits{V.}}, \betal:
+\batitle{{Scikit-learn: Machine learning in Python}}.
+\bjtitle{The Journal of Machine Learning Research}
+\bvolume{12},
+\bfpage{2825}--\blpage{2830}
+(\byear{2011})
+\end{barticle}
+\endbibitem
+
+%%% 27
+\bibitem{wong2012points}
+\begin{botherref}
+\oauthor{\bsnm{Wong}, \binits{B.}}:
+Points of view: visualizing biological data
+(2012)
+\end{botherref}
+\endbibitem
+
+%%% 28
+\bibitem{jacobs1988increased}
+\begin{barticle}
+\bauthor{\bsnm{Jacobs}, \binits{R.A.}}:
+\batitle{Increased rates of convergence through learning rate adaptation}.
+\bjtitle{Neural Networks}
+\bvolume{1}(\bissue{4}),
+\bfpage{295}--\blpage{307}
+(\byear{1988})
+\end{barticle}
+\endbibitem
+
+\end{thebibliography}
 

### 2021-03-25 paper/policar2020-opentsne.tex
Apply BMC template

--- 2021-03-17 paper/policar2020-opentsne.tex
@@
-% \documentclass[twocolumn]{article}
-\documentclass[letter]{article}
-\usepackage[a4paper, total={6in, 8in}]{geometry}
+%% BioMed_Central_Tex_Template_v1.06
+%%                                      %
+%  bmc_article.tex            ver: 1.06 %
+%                                       %
 
-%%% Load packages
-%\usepackage{amsthm}
-%\RequirePackage{natbib}
-%\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
-\usepackage{hyperref}
-\usepackage[utf8]{inputenc} %unicode support
-%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
-%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
+%%IMPORTANT: do not delete the first line of this template
+%%It must be present to enable the BMC Submission system to
+%%recognise this template!!
 
-\usepackage{graphicx}
+%%% additional documentclass options:
+%  [doublespacing]
+%  [linenumbers]   - put the line numbers on margins
+
+\documentclass[twocolumn]{bmcart}
+
+\usepackage{amsthm,amsmath}
+\usepackage[utf8]{inputenc}
+\usepackage{url}
+
 \usepackage{minted}
@@
 \usepackage{amsmath,amssymb}
+\usepackage{graphicx}
 \usepackage{booktabs}
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%                                             %%
+%%  If you wish to display your graphics for   %%
+%%  your own use using includegraphic or       %%
+%%  includegraphics, then comment out the      %%
+%%  following two lines of code.               %%
+%%  NB: These line *must* be included when     %%
+%%  submitting to BMC.                         %%
+%%  All figure files must be submitted as      %%
+%%  separate graphics through the BMC          %%
+%%  submission process, not included in the    %%
+%%  submitted article.                         %%
+%%                                             %%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%\def\includegraphic{}
+%\def\includegraphics{}
+
+%%% Put your definitions there:
+\startlocaldefs
 \newcommand{\opentsne}{\textsf{openTSNE}}
-
-\renewcommand{\baselinestretch}{1.2}
+\endlocaldefs
 
@@
 
+\begin{frontmatter}
+\begin{fmbox}
+\dochead{Software}
+
 \title{openTSNE: a modular Python library for t-SNE dimensionality reduction and embedding}
 
-\author{
-  Pavlin G. Poli\v{c}ar\textsuperscript{1}\\
-  \texttt{pavlin.policar@fri.uni-lj.si}\\
-  \and
-  Martin Stra\v{zar}\textsuperscript{2}\\
-  \texttt{martin.strazar@gmail.com}\\
-  \and
-  Bla\v{z} Zupan\textsuperscript{1,3}\\
-  \texttt{blaz.zupan@fri.uni-lj.si}\\
+\author[
+  addressref={aff1},
+  corref={aff1},
+  email={pavlin.policar@fri.uni-lj.si}
+]{\inits{P.G.}\fnm{Pavlin G.} \snm{Poli\v{c}ar}}
+\author[
+  addressref={aff2},
+  email={martin.strazar@gmail.com}
+]{\inits{M.S.}\fnm{Martin} \snm{Stra\v{zar}}}
+\author[
+  addressref={aff1,aff3},
+  email={martin.strazar@gmail.com}
+]{\inits{B.Z.}\fnm{Bla\v{z}} \snm{Zupan}}
+
+\address[id=aff1]{
+  \orgdiv{Faculty of Computer and Information Science},
+  \orgname{University of Ljubljana},
+  \street{Ve\v{c}na pot 113},
+  \city{Ljubljana},
+  \cny{Slovenia}
+}
+\address[id=aff2]{%
+  \orgname{Broad Institute of Harvard and MIT},
+  \street{D\"{u}sternbrooker Weg 20},
+  %\postcode{}
+  \city{Cambridge},
+  \cny{U.S.A.}
+}
+\address[id=aff3]{%
+  \orgdiv{Department of Molecular and Human Genetics},
+  \orgname{Baylor College of Medicine},
+  \street{1 Baylor Plaza},
+  %\postcode{}
+  \city{Houston},
+  \cny{U.S.A.}
 }
 
-\date{
-  \textsuperscript{1}Faculty of Computer and Information Science, University of Ljubljana,\\
-  Ve\v{c}na pot 113, SI-1000 Ljubljana, Slovenia\\ \vspace{2mm}
-  \textsuperscript{2}Broad Institute of Harvard and MIT,\\
-  D\"{u}sternbrooker Weg 20, MA-02142 Cambridge, U.S.A.\\ \vspace{2mm}
-  \textsuperscript{3}Department of Molecular and Human Genetics, Baylor College of Medicine,\\
-  1 Baylor Plaza, TX-77030 Houston, U.S.A.
-}
+%\end{fmbox}% comment this for two column layout
 
-\maketitle
-
+\begin{abstractbox}
 \begin{abstract}
@@
 
-\providecommand{\keywords}[1]{\vspace{2mm}\textbf{Keywords ---} #1}
-\begin{keywords}
-t-SNE, embedding, visualization, dimensionality reduction, Python library
-\end{keywords}
+\begin{keyword}
+\kwd{t-SNE}
+\kwd{embedding}
+\kwd{visualization}
+\kwd{dimensionality reduction}
+\kwd{Python library}
+\end{keyword}
 
-\section*{Introduction}
+\end{abstractbox}
+
+\end{fmbox}% uncomment this for two column layout
+
+\end{frontmatter}
+
+\section*{Background}
 
@@
 
-\section*{Conclusion}
+\section*{Conclusions}
 
@@
 
-\section*{Availability}
-
-\opentsne\ is distributed under the BSD-3-Clause License and is publicly
-available as an open-source package at
-\url{https://github.com/pavlin-policar/openTSNE}. \opentsne\ is also available
-on \textsf{PyPI} and \textsf{conda-forge}. The data sets used and scripts used
-in this study are included in accompanying notebooks, publicly available at
-\url{https://github.com/pavlin-policar/opentsne-paper}.
-
-\section*{Competing interests}
-The authors declare no competing interests.
-
-\section*{Author's contributions}
-P.G.P.
-implemented and maintains \opentsne\ library.
-M.S.
-helped in the development of mathematical foundations and advised on the inclusion of the extensions.
-B.Z.
-initiated and supervised the project.
-All three co-authors wrote the manuscript.
-
-\section*{Acknowledgements}
-This work was supported by Slovenian Research Agency (P2-0209).
-We want to thank Dmitry Kobak for various helpful discussions and best practices
-when using t-SNE and his contributions to the source code. We would also
-like to thank George Linderman for his help with the FIt-SNE algorithm.
-
-\bibliographystyle{bmc-mathphys} % Style BST file (bmc-mathphys, vancouver, spbasic).
-\bibliography{references}      % Bibliography file (usually '*.bib' )
-
-\eject
-\fontsize{8}{10}\selectfont
-\section*{Online Methods}
+%\fontsize{8}{10}\selectfont
+\section*{Methods}
 
@@
 
+\begin{backmatter}
+
+\section*{Acknowledgements}
+We want to thank Dmitry Kobak for various helpful discussions and best practices
+when using t-SNE and his contributions to the source code. We would also
+like to thank George Linderman for his help with the FIt-SNE algorithm.
+
+\section*{Funding}
+This work was supported by Slovenian Research Agency (P2-0209).
+
+\section*{Availability of data and materials}
+\opentsne\ is distributed under the BSD-3-Clause License and is publicly
+available as an open-source package at
+\url{https://github.com/pavlin-policar/openTSNE}. \opentsne\ is also available
+on \textsf{PyPI} and \textsf{conda-forge}. The data sets used and scripts used
+in this study are included in accompanying notebooks, publicly available at
+\url{https://github.com/pavlin-policar/opentsne-paper}.
+
+\section*{Competing interests}
+The authors declare that they have no competing interests.
+
+\section*{Authors' contributions}
+P.G.P.
+implemented and maintains \opentsne\ library.
+M.S.
+helped in the development of mathematical foundations and advised on the inclusion of the extensions.
+B.Z.
+initiated and supervised the project.
+All three co-authors wrote the manuscript.
+
+\bibliographystyle{bmc-mathphys} % Style BST file (bmc-mathphys, vancouver, spbasic).
+\bibliography{references}
+
+\end{backmatter}
+
 \end{document}

### 2021-04-25 paper/main.tex
Add JSS paper

\documentclass[article]{jss}

%% -- LaTeX packages and custom commands ---------------------------------------

%% recommended packages
\usepackage{thumbpdf,lmodern}

%% another package (only for this demo article)
\usepackage{framed}

%% new custom commands
\newcommand{\class}[1]{`\code{#1}'}
\newcommand{\fct}[1]{\code{#1()}}

% Custom imports, and defns
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{amssymb}

\newcommand{\opentsne}{\pkg{openTSNE}}

%% -- Article metainformation (author, title, ...) -----------------------------

%% - \author{} with primary affiliation
%% - \Plainauthor{} without affiliations
%% - Separate authors by \And or \AND (in \author) or by comma (in \Plainauthor).
%% - \AND starts a new line, \And does not.
\author{
  Pavlin G. Poli\v{c}ar\\
  Faculty of Computer and Information Science\\
  University of Ljubljana\\
  Ljubljana, Slovenia
  \AND
  Martin Stra\v{z}ar\\
  Broad Institute of MIT and Harvard\\
  Cambridge, MA, USA
  \AND
  Bla\v{z} Zupan\\
  Faculty of Computer and Information Science\\
  University of Ljubljana\\
  Ljubljana, Slovenia
}
\Plainauthor{P. G. Poli\v{c}ar, M. Stra\v{z}ar, B. Zupan}

%% - \title{} in title case
%% - \Plaintitle{} without LaTeX markup (if any)
%% - \Shorttitle{} with LaTeX markup (if any), used as running title
\title{openTSNE: a modular \proglang{Python} library for t-SNE dimensionality reduction and embedding}
\Plaintitle{openTSNE: a modular Python library for t-SNE dimensionality reduction and embedding}
\Shorttitle{openTSNE: Extensible, parallel implementations of t-SNE}

%% - \Abstract{} almost as usual
\Abstract{
One of the most popular techniques for constructing visualizations of large,
high-dimensional data sets is $t$-distributed stochastic neighbor embedding
(t-SNE). Recently, several extensions have been proposed to address
issues of scalability and the quality of the resulting visualizations. We
introduce \opentsne, a modular Python library that implements the core t-SNE
algorithm and its many extensions. The library is fast and scales up to data
sets containing millions of data points. \opentsne\ is distributed under the
BSD-3-Clause License and its source code is publicly available at
\url{https://github.com/pavlin-policar/openTSNE}.
}

%% - \Keywords{} with LaTeX markup, at least one required
%% - \Plainkeywords{} without LaTeX markup (if necessary)
%% - Should be comma-separated and in sentence case.
\Keywords{t-SNE, embedding, visualization, dimensionality reduction, \proglang{Python}}
\Plainkeywords{t-SNE, embedding, visualization, dimensionality reduction}

%% - \Address{} of at least one author
%% - May contain multiple affiliations for each author
%%   (in extra lines, separated by \emph{and}\\).
%% - May contain multiple authors for the same affiliation
%%   (in the same first line, separated by comma).
\Address{
  Pavlin G. Poli\v{c}ar\\
  Faculty of Computer and Information Science, University of Ljubljana\\
  Ve\v{c}na pot 113, Ljubljana, Slovenia\\
  E-mail: \email{pavlin.policar@fri.uni-lj.si}
}

\begin{document}

%% -- Introduction -------------------------------------------------------------

%% - In principle "as usual".
%% - But should typically have some discussion of both _software_ and _methods_.
%% - Use \proglang{}, \pkg{}, and \code{} markup throughout the manuscript.
%% - If such markup is in (sub)section titles, a plain text version has to be
%%   added as well.
%% - All software mentioned should be properly \cite-d.
%% - All abbreviations should be introduced.
%% - Unless the expansions of abbreviations are proper names (like "Journal
%%   of Statistical Software" above) they should be in sentence case (like
%%   "generalized linear models" below).

\section[Introduction]{Introduction} \label{sec:intro}

The ever-growing volumes of high-dimensional data sets in machine learning calls
for efficient dimensionality reduction techniques and their implementations,
which can produce informative data visualizations. Popular approaches include
principal component analysis, multidimensional scaling, t-distributed stochastic
neighbor embedding (t-SNE)~\citep{maaten2008visualizing}, and uniform manifold
approximation and projections (UMAP)~\citep{2018arXivUMAP}. Among these, t-SNE
has received much attention as it can address high volumes of data and reveal
the underlying data structure. For instance, t-SNE is widely used in the
bioinformatics community in areas such as single-cell
transcriptomics~\citep{macosko2015highly,cao2019single,tasic2018shared}, human
genetics~\citep{hirata2019genetic}, metagenomic
assembly~\citep{beaulaurier2018metagenomic}, the spatial organization of
microbial communities~\citep{sheth2019spatial}, and
metabolomics~\citep{tkachev2019differences}. Reports on single-cell gene
expression data, our running example, often start with an overview of the cell
landscape, where t-SNE is used to embed high-dimensional expression profiles
into a two-dimensional space. Figs.~\ref{fig:macosko}.a and \ref{fig:macosko}.b
show two such embeddings.

\begin{figure*}[ht]
  \includegraphics[width=\textwidth]{macosko2015}
  \caption{\label{fig:macosko}
  We use \opentsne\ to generate three t-SNE embeddings and demonstrate recent
  theoretical advances. The data in \textbf{(a)} and \textbf{(b)} represent
  44,808 single-cell gene-expression profiles of mouse retinal cells from 
  \citet{macosko2015highly}. The data in \textbf{(c)} additionally contains
  27,499 expression profiles from mouse retinal cells from
  \citet{shekhar2016comprehensive}. \textbf{(a)} We construct a t-SNE embedding
  following the parameter choices from the original publication by
  \citet{maaten2008visualizing}. The visualization shows no preservation of the
  global organization of clusters, resulting from random initialization and an
  affinity model focused on preserving local neighborhoods. \textbf{(b)} A
  modern t-SNE embedding, utilizing the latest theoretical advances and
  practical recommendations constructed using a multi-scale affinity model,
  preserving both short-range and long-range interactions between data points
  and initialized so that the global layout is as meaningful as possible. Unlike
  in \textbf{(a)}, the green and blue clusters representing different sub-types
  of amacrine and bipolar cells are now localized to the same regions of the
  space, indicating a higher level of similarity than to other cell types. The
  embedding in \textbf{(c)} shows how existing t-SNE reference atlases can be
  used to place new samples into existing embeddings. The positions of new data
  points correspond to cell types from the reference atlas.
}
\end{figure*}

Despite its utility, t-SNE has often been criticized for its limited
scalability, lack of global organization -- t-SNE identifies well-defined
clusters that may be arbitrarily scattered throughout the low-dimensional space
-- and the absence of theoretically-founded methods to map new data into
existing embeddings~\citep{ding2018interpretable,becht2019dimensionality}. Most
of these shortcomings have recently been addressed. \citet{linderman2019fast}
developed FIt-SNE, an efficient approximation scheme which massively improves
the scalability of t-SNE, achieving linear time complexity in the number of
samples. \citet{kobak2019art} proposed several techniques to improve global
cluster coherence, including estimating similarities with a mixture of Gaussian
kernels. In our previous work, we introduced a principled approach for embedding
new samples into existing visualizations~\citep{policar2019embedding}.

However, easy access to efficient implementations almost universally correlates
with the widespread adoption of novel computational techniques. Despite the many
theoretical advancements, most popular t-SNE libraries have been slow to
incorporate them into their implementations. In this paper, we present \opentsne,
our open-source Python implementation of t-SNE and its many recently proposed
extensions. \opentsne\ is easy to install and includes precompiled binaries
available through popular Python package managers. It provides a familiar API,
making it suitable as a drop-in replacement for existing t-SNE implementations.

%% -- Manuscript ---------------------------------------------------------------

%% - In principle "as usual" again.
%% - When using equations (e.g., {equation}, {eqnarray}, {align}, etc.
%%   avoid empty lines before and after the equation (which would signal a new
%%   paragraph.
%% - When describing longer chunks of code that are _not_ meant for execution
%%   (e.g., a function synopsis or list of arguments), the environment {Code}
%%   is recommended. Alternatively, a plain {verbatim} can also be used.
%%   (For executed code see the next section.)

\section{Methods} \label{sec:methods}

We first introduce the relevant notation and briefly review the core t-SNE
algorithm in Sec.~\ref{sec:meth.tsne}. We then address the criticisms of
scalability (Sec.~\ref{sec:meth.approx}), the ability to add new samples to
existing embeddings (Sec.~\ref{sec:meth.transform}), and improvements that
improve the global consistency of resulting embeddings
(Sec.~\ref{sec:meth.global}).

\subsection{t-distributed stochastic neighbor embedding} \label{sec:meth.tsne}

t-distributed stochastic neighbor embedding (t-SNE) is a non-linear
dimensionality reduction method that to aims to find a low-dimensional embedding
where neighborhoods are preserved. More formally, given a multi-dimensional data set
$\mathbf{X} = \left \{ \mathbf{x}_1, \mathbf{x}_2, \dots, \mathbf{x}_N \right \}
\in \mathbb{R}^D$ where $N$ is the number of data points in the data set, t-SNE
aims to find a low dimensional embedding $\mathbf{Y} = \left \{ \mathbf{y}_1,
\mathbf{y}_2, \dots, \mathbf{y}_N \right\} \in \mathbb{R}^d$ where $d \ll D$,
such that if points $\mathbf{x}_i$ and $\mathbf{x}_j$ are close in the
high-dimensional space, their corresponding embeddings $\mathbf{y}_i$ and
$\mathbf{y}_j$ are also close. Since t-SNE is primarily used as a visualization
tool, $d$ is typically set to two. The similarity between two data points in the
high-dimensional space is defined using the Gaussian kernel
\begin{equation}
p_{j \mid i} = \frac{\exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{x}_j ) / \sigma_i^2 \right )}
{\sum_{k \neq i } \exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{x}_k ) / \sigma_i^2 \right )}, \quad p_{i \mid i} = 0 \notag
\label{eq:gaussian_kernel}
\end{equation}
where $\mathcal{D}$ is some distance measure. The values $p_{j \mid i}$
are then symmetrized to
\begin{equation}
p_{ij} = \frac{p_{j \mid i} + p_{i \mid j}}{2N}.
\label{eq:symmetrize}
\end{equation}

The bandwidth $\sigma_i$ of each Gaussian kernel is selected such that the perplexity of the distribution matches a user-specified parameter value
\begin{equation}
\text{Perplexity} = 2^{H(P_i)} \notag
\end{equation}
where $H(P_i)$ is the Shannon entropy of $P_i$,
\begin{equation}
H(P_i) = -\sum_i p_{j \mid i} \log_2 (p_{j \mid i}). \notag
\end{equation}
This enables t-SNE to adapt to the varying density of the data in the
multi-dimensional space. The perplexity can be interpreted as as the continuous
analogue to the number of nearest neighbors to which the distances will be
preserved.

The similarity between points $\mathbf{y}_i$ and $\mathbf{y}_j$ in the embedding
space is defined using the $t$-distribution with a single degree of freedom
(Cauchy kernel)
\begin{equation}
q_{ij} = \frac{\left ( 1 + || \mathbf{y}_i - \mathbf{y}_j ||^2 \right )^{-1}}
{\sum_{k \neq l}\left ( 1 + || \mathbf{y}_k - \mathbf{y}_l ||^2 \right )^{-1}},
\quad q_{ii} = 0.
\label{eq:cauchy_kernel}
\end{equation}

We use the Kullback-Leibler (KL) divergence to measure the agreement
between the two distributions $\mathbf{P}$ and $\mathbf{Q}$
\begin{equation}
C = \text{KL}(\mathbf{P} \mid \mid \mathbf{Q}) = \sum_{ij} p_{ij} \log \frac{p_{ij}}{q_{ij}}.
\label{eq:kl_divergence}
\end{equation}
The objective is to find embeddings $\mathbf{Y}$ that minimize the KL divergence. The corresponding gradient takes the form
\begin{equation}
\frac{\partial C}{\partial \mathbf{y}_i} = 4 \sum_{j \neq i} \left ( p_{ij} - q_{ij} \right ) \left ( \mathbf{y}_i - \mathbf{y}_j \right ) w_{ij},
\label{eq:tsne_gradient}
\end{equation}
where $w_{ij} = \left ( 1 + || \mathbf{y}_i - \mathbf{y}_j || ^2 \right )^{-1}$
and represents the unnormalized $q_{ij}$.

Optimization is performed using batch gradient descent using the delta-bar-delta
update rule~\citep{jacobs1988increased}. The t-SNE optimization proceedure
consists of two phases: in the first \textit{early exaggeration} phase, the
attractive forces between data points are increased by some factor $\rho$,
typically set to $12$, so that points in the embedding can more easily move
throughout the space and settle near their respective neighbors. In the second
phase of the optimization, the attractive forces are then reverted to their
original values with $\rho=1$.

\citet{belkina2019automated} later found that convergence can be sped up by
increasing the learning rate from the standard $\eta=200$ to $\eta=N/12$. As a
side-effect, embeddings converge faster, and the number of iterations can be
lowered from the typical 1000 to 750, decreasing the overall runtime.  Modern
t-SNE implementations, including \pkg{FIt-SNE} and \opentsne, have adopted this
convention.

\subsection{Efficient Approximation Schemes} \label{sec:meth.approx}

A direct evaluation of t-SNE gradients requires $\mathcal{O}(N^2)$ operations,
which makes its application impractical to any reasonably-sized data set and
highlights the need for the development of efficient approximation schemes.
\citet{van2014accelerating} observed that the t-SNE gradient can be cast as an
N-body problem where data points represent particles that attract and repel each
other. The gradient from Eqn.~(\ref{eq:tsne_gradient}) can be rewritten as
\begin{equation}
\frac{\partial C}{\partial \mathbf{y}_i} = 4 \Bigg [
\underbrace{\sum_{j \neq i} p_{ij} q_{ij} Z \left ( \mathbf{y}_i - \mathbf{y}_j \right )}_{\text{attractive forces}}  -
\underbrace{\sum_{j \neq i} q_{ij}^2 Z \left ( \mathbf{y}_i - \mathbf{y}_j \right )}_{\text{repulsive forces}}
\Bigg ], \label{eq:grad_attr_rep} \notag
\end{equation}
where $Z = \sum_{k \neq l}\left ( 1 + || \mathbf{y}_k - \mathbf{y}_l ||^2 \right
)^{-1}$. We can view this equation as a particle simulation, where the
two terms represent the attractive and repulsive forces between individual
particles. Each term lends itself to efficient approximations, enabling us to
reduce the time complexity of t-SNE to $\mathcal{O}(N)$. This allows modern
t-SNE implementations to be leveraged for the visulization of data sets
containing up to millions of data points.

\subsubsection*{Attractive Forces}
\citet{van2014accelerating} observed that evaluating the attractive forces
between all pairs of data points is excessive, and that considering only a
handful of nearest neighbors at each point are sufficient to obtain a good
approximation. Therefore, instead of calculating all pairwise interactions, it
is sufficient to only find and evaluate the attraction to a certain number of
$k$ nearest neighbors. Using tree-based nearest-neighbor search methods,
\citet{van2014accelerating} reduced the time complexity to $\mathcal{O}(N \log
N)$. \citet{linderman2019fast} further realized that, qualitatively, embeddings
are visually indistinguishable when using only \textit{approximate} nearest
neighbors, further reducing time complexity to $\mathcal{O}(N)$.

\subsubsection*{Repulsive Forces}
The repulsive term can similarily be approximated, motivated by methods from
particle simulations. \citet{van2014accelerating} proposed an approach based on
N-body simulations and used a space-partitioning Barnes-Hut tree approach to
approximate the interaction between data points. This reduces the time
complexity from $\mathcal{O}(N^2)$ to $\mathcal{O}(N \log N)$. More recently,
\citet{linderman2019fast} proposed an alternative approach, FIt-SNE, based on
non-uniform convolutions which further reduces the time complexity to
$\mathcal{O}(N)$.

\subsection{Embedding New Samples} \label{sec:meth.transform}

t-SNE is non-parametric and does not define an explicit mapping from the
high-dimensional space to the embedding space. Therefore embeddings of new data
points need to be found through the use of optimization
techniques~\citep{policar2019embedding}. When adding new data points to an
existing, reference embedding, the reference data points are fixed in place
while new data points are allowed to find their respective positions. The
optimization remains the same as in standard t-SNE with only slight
modifications to $p_{ij}$ and $q_{ij}$
\begin{align}
p_{j \mid i} = \frac{\exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{v}_j) /  \sigma_i^2 \right )}{\sum_{i} \exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{v}_j) / \sigma_i^2 \right )}, \qquad
q_{j \mid i} = \frac{\left ( 1 + || \mathbf{y}_i - \mathbf{w}_j ||^2 \right )^{-1}}{\sum_{i}\left ( 1 + || \mathbf{y}_i - \mathbf{w}_j ||^2 \right )^{-1}}, \notag
\end{align}
\noindent where $\mathbf{V} = \left \{ \mathbf{v}_1, \mathbf{v}_2, \dots,
\mathbf{v}_M \right \} \in \mathbb{R}^D$ where $M$ is the number of samples in
the new data set and $\mathbf{W} = \left \{ \mathbf{w}_1, \mathbf{w}_2, \dots,
\mathbf{w}_M \right \} \in \mathbb{R}^d$. Additionally, we omit the
symmetrization step in Eqn.~(\ref{eq:symmetrize}). Plugging these terms into
Eqn.~(\ref{eq:kl_divergence}), we obtain the following gradient
\begin{equation}
\frac{\partial C}{\partial \mathbf{w}_j} = 2 \sum_i \left ( p_{j \mid i} - q_{j \mid i} \right ) \left ( \mathbf{y}_i - \mathbf{w}_j \right ) \left ( 1 + || \mathbf{y}_i - \mathbf{w}_j || ^2 \right )^{-1}. \notag
\label{eq:gradient}
\end{equation}

Similarly to standard t-SNE, a direct calculation of gradients takes
$\mathcal{O}(N \cdot M)$ time, but it is straightforward to adapt the Barnes-Hut
and FIt-SNE approximation schemes, reducing the time complexity to
$\mathcal{O}(M \log N)$ and $\mathcal{O}(M)$, respectively. Special care must be
taken to tune the learning rate during optimization as the existing values
$\eta=200$ and $\eta=N/12$ result in unstable optimization.

\subsection{Improving the Global Structure of Embeddings} \label{sec:meth.global}

One of t-SNE's criticisms is that it fails to preserve long-range distances but
instead focuses on capturing the local structure of the
data manifold~\citep{becht2019dimensionality}. Recently, several approaches have been
proposed to improve the global organization of the resulting embeddings.

\citet{kobak2019umap} showed that the initialization of force-directed layout
algorithms, such as t-SNE, UMAP~\citep{2018arXivUMAP}, and ForceAtlas2~\citep{jacomy2014forceatlas2}, largely dictates the global
consistency of resulting embeddings. When initialized randomly, clusters are
often arbitrarily scattered in the embedding space. However, when using
initalization schemes based on PCA or spectral embeddings, the clusters
identified in the resulting embedding are typically grouped in a globally
coherent manner.

In standard t-SNE, distances between data points are converted to similarities
through the use of Gaussian kernels. The perplexity of these kernels governs the
sizes of the local neighborhoods which are to be preserved. One easy way to
improve the global consistency of resulting embeddings is to increase the sizes
of these neighborhoods. This can be achieved by increasing the perplexity
parameter. However, this often comes at the price of local structure.
\citet{kobak2019art} propose to instead use mixtures of Gaussians to better
preserve short-range, as well as long-range distances and, thus, achieve a
better trade-off between local and global structure.

Embeddings produced by standard t-SNE often makes use all available space and
separates clusters by only thin boundaries. When working with large data sets,
this often obscures the global relationships between clusters as all neighboring
groups appear equidistant from one another.  Other dimensionality reduction
methods, such as UMAP or
ForceAtlas2, produce embeddings where clusters
appear more compact and the white-space separating the clusters may be
interpreted as a loose measure of distance. Recently, \citet{bohm2020unifying}
showed that the exaggeration factor $\rho$ could be used to produce layouts more
similar to UMAP and ForceAtlas2. By incorporating exaggeration into later phases
of the optimization, t-SNE introduces more white-space between clusters which better
reflects the global relations between clusters.

Standard t-SNE reveals the clustering structure at a single level of resolution.
While the perplexity parameter can be used to control the trade-off between
local and global structure, this can be time-consuming, and small, well-defined
clusters can be missed. Alternatively, \citet{kobak2019heavy} suggest that
varying the degrees of freedom in the $t$-distribution can be used to explore
the clustering structure at different levels of resolution. Modifying
Eqn.~(\ref{eq:cauchy_kernel}) to $q_{ij} \propto \left ( 1 + || \mathbf{y}_i -
\mathbf{y}_j ||^2 / \alpha \right )^{-\alpha}$ allows us to use heavier-tailed
distributions to model distances in the embedding space, which acts to highlight
small subgoups in the resulting visualization.

\section{Implementation} \label{sec:implementation}

We introduce \opentsne, a comprehensive Python library that implements t-SNE and
its numerous recently proposed extensions, including the implementation of
efficient approximation schemes allowing the embedding of millions of data
points~\citep{van2014accelerating,linderman2019fast}, the addition of new
data points into existing embeddings~\citep{policar2019embedding}, various
initialization improvements~\citep{kobak2019umap},
extensions~\citep{kobak2019heavy,kobak2019art}, and improved parameter
defaults~\citep{belkina2019automated}. Combined, these improvements result in
the faster generation of more informative visualizations.

\opentsne\ is compatible with the Python data science ecosystem (e.g.
\pkg{numpy}, \pkg{scikit-learn}, \pkg{scanpy}), providing a familiar
and intuitive API to users. Its modular design encourages extensibility and
experimentation with various parameter settings and changes to the analysis
pipeline. \opentsne\ aims to make the latest theoretical advances in t-SNE
accessible to the wider data-science community. These features are easily
accessible through an intuitive API, which closely follows the style
\pkg{scikit-learn}~\citep{sklearn_api}, a popular and widely used toolkit for
machine learning. The following snippet exemplifies this interface and
demonstrates how \opentsne\ can be used to quickly generate several
visualizations, each highlighting different aspects of the underlying data
structure.
\begin{CodeChunk}
\begin{CodeInput}
from openTSNE import TSNE

# Vanilla t-SNE
tsne_standard = TSNE().fit(x)

# Adapt parameters to better highlight small clusters in the data
tsne_extensions = TSNE(perplexity=50, exaggeration=2, dof=0.8).fit(x)

# Use multiscale kernel for better local/global structure trade-off
from openTSNE.affinity import Multiscale

affinities = Multiscale(x, perplexities=[50, 500])
tsne_multiscale = TSNE().fit(x, affinities=affinites)

# Use the generated embeddings to add new data points
new_embedding = tsne_extensions.transform(x_new)
\end{CodeInput}
\end{CodeChunk}
The code snippet considers a \pkg{numpy} array or \pkg{scipy} sparse
matrix \texttt{x} and creates three \texttt{TSNEEmbedding} objects. These
represent fully-optimized t-SNE embeddings and are internally stored as \pkg{numpy}
arrays. These can be manipulated and visualized using standard Python tools. We
then select one of these embeddings and use it to embed new data points from
\texttt{x\_new} into the embedding space, while keeping the reference embedding
fixed. Additionally, \opentsne\ provides a more advanced interface that enables
the user to change every aspect of the t-SNE optimization procedure. This
interface is described in detail in the accompanying documentation.

% \section{Existing t-SNE Implementations}

% \opentsne\ is not the only available open-source implementation of the t-SNE 
% algorithm, however, it provides the most user-friendly experience and is the most
% feature-complete. A detailed comparison of several t-SNE implementations is
% shown in Table~\ref{tab:features}.

% \pkg{scikit-learn} is one of the most widely-used machine learning toolkits
% in the Python data science ecosystem, and provides an implementation of the
% t-SNE algorithm~\citep{pedregosa2011scikit}. However, the implementation from
% \pkg{scikit-learn} utilizes an outdated approximation scheme and is
% prohibitively slow for even moderately-sized data sets containing tens of
% thousands of samples. \opentsne\ implements all available approximation schemes,
% and defaults to the FIt-SNE approximation algorithm, which allows it to scale to
% data sets containing up to millions of data points on standard, consumer-grade,
% laptop computers. Performance benchmarks are show in Fig.~\ref{fig:benchmarks}.
% \begin{figure*}[ht!]
%   \includegraphics[width=\textwidth]{benchmarks}
%   \caption{\label{fig:benchmarks}
%   We benchmark \opentsne\ (v0.6.0) against three popular open-source
%   implementations from \pkg{scikit-learn}~\citep{pedregosa2011scikit}
%   (v0.23.1), \pkg{MulticoreTSNE}~\citep{Ulyanov2016} (v0.1), and
%   \pkg{FIt-SNE}~\citep{linderman2019fast} (v1.1.0). Experiments were run on a
%   consumer-grade Intel Core i7-7700HQ processor found in laptop computers, and
%   on a server-grade Intel Xeon E5-2650. To generate benchmark data sets of
%   different sizes, we subsampled data from the 10X Genomics 1.3 million mouse
%   brain data set five times, resulting in five different data sets for each
%   size. In total, we run each implementation on 30 different data sets. Notice
%   that \opentsne\ scales similarly to \pkg{FIt-SNE}, as they both use the
%   same interpolation-based approximation scheme, while \pkg{scikit-learn} and
%   \pkg{MulticoreTSNE} utilize the Barnes-Hut approximation.
% }
% \end{figure*}

% Pure C++ implementations, such as \pkg{BHTSNE}~\citep{van2014accelerating},
% \pkg{MulticoreTSNE}~\citep{Ulyanov2016} and
% \pkg{FIt-SNE}~\citep{linderman2019fast} are fast, but require users to
% compile and install the packages themselves. This is especially problematic for
% users of the Windows operating system, where the C/C++ compiler is not bundled
% with the operating system, making their correct configuration and compilation
% cumbersome. Additionally, these packages were not built with the Python ecosystem in
% mind, and cannot be directly used as dependencies in other Python packages.
% \opentsne\ provides several, well-documented installation options and can
% readily be used as a dependency in larger projects. We provide precompiled
% binaries through the Python Package Index (\pkg{PyPI}) or \pkg{conda} from
% the \pkg{conda-forge} channel, allowing \opentsne\ to work out-of-the-box all
% major platforms.

% Another concern is the adoption of recent theoretical improvements to the t-SNE
% algorithm. Some implementations such as \pkg{BHTSNE} and \pkg{MulticoreTSNE}
% appear to have been discontinued, as there has been little to no activity on
% their respective GitHub repositories in recent years. \pkg{scikit-learn} has a
% very active community, however, adoption of recent developments such as improved
% initialization schemes, better default parameter settings, and more flexible
% affinity kernels has been slow. Among the existing implementations,
% \pkg{FIt-SNE} includes most of the latest features but lacks crucial ones e.g.
% the ability to add new data points into existing embeddings, or flexible
% optimization strategies.

% Finally, with the exception of \pkg{scikit-learn}, the existing packages offer
% little to no user documentation, usage examples, or API reference. Additionally,
% these packages do not follow standard software development best practices, such
% as utilizing continuous integration checks or following standard style guides,
% making their maintenance difficult and unreliable.

%% -- Illustrations ------------------------------------------------------------

%% - Virtually all JSS manuscripts list source code along with the generated
%%   output. The style files provide dedicated environments for this.
%% - In R, the environments {Sinput} and {Soutput} - as produced by Sweave() or
%%   or knitr using the render_sweave() hook - are used (without the need to
%%   load Sweave.sty).
%% - Equivalently, {CodeInput} and {CodeOutput} can be used.
%% - The code input should use "the usual" command prompt in the respective
%%   software system.
%% - For R code, the prompt "R> " should be used with "+  " as the
%%   continuation prompt.
%% - Comments within the code chunks should be avoided - these should be made
%%   within the regular LaTeX text.

%\section{Illustrations} \label{sec:illustrations}

%% -- Summary/conclusions/discussion -------------------------------------------

\section{Summary and discussion} \label{sec:summary}

\subsection{Uncovering Structure in High-Dimensional Data}

Dimensionality reduction techniques implicitly assume that high-dimensional data
lies on a lower-dimensional manifold, which can accurately be captured by a
small number of dimensions. However, there is no evidence that every data set
can accurately be described using only two dimensions, and any such embedding
will inevitably lead to a loss of information. Thus, it is beneficial to examine
multiple embeddings, each of which provides a different perspective on topology
and other data characteristics.

\begin{figure*}[htbp]
  \center
  \includegraphics[width=0.75\textwidth]{tasic2018}
  \caption{\label{fig:tasic}
  We use \opentsne\ to create four different
  visualizations of the \citet{tasic2018shared} data,
  each providing a different perspective into the topology of the data.
  The data set contains 21,874 single-cells originating from the mouse
  neocortex. Cluster annotations and colors are taken from the original
  publication. Warm colors correspond to excitatory neurons, cool colors
  correspond to inhibitory neurons, and gray/brown colors correspond to
  non-neuronal cells. Standard t-SNE \textbf{(a)} emphasizes local
  structure while increasing perplexity \textbf{(b)} results in a more
  meaningful layout of the clusters. We can also combine the two
  perplexities by using a multiscale kernel affinity model \textbf{(c)}
  and obtain a trade-off between global and local structure.
  Alternatively, we can inspect more fine-grained structure and reveal
  smaller clusters by using a more heavy-tailed kernel \textbf{(d)}.}
\end{figure*}

We illustrate this point by generating four different embeddings of the data on
single-cell gene expression in mouse brain~\citep{tasic2018shared}.
Fig.~\ref{fig:tasic}.a shows an embedding using default t-SNE parameters. While
different clusters of excitatory and inhibitory neurons appear close to one
another, all clusters appear equidistant from their neighbors, and the overall
relations between groups are not obvious. The embedding in Fig.~\ref{fig:tasic}.b
focuses on preserving larger neighborhoods of points, resulting in a more
globally consistent layout where relations between clusters become more
apparent. Here, it is evident from the increased white space between groups that
there is one large class of excitatory neurons and two related classes of
inhibitory neurons. Unfortunately, focusing on preserving large neighborhoods
leads to the absorption of smaller clusters into larger ones. Alternatively,
Fig.~\ref{fig:tasic}.c uses multi-scale similarity kernels that aim to preserve
both the global organization of clusters and prevent smaller cluster absorption. We
constructed the embedding from Fig.~\ref{fig:tasic}.d with the settings used
for Fig.~\ref{fig:tasic}.a, but at a finer level of resolution.
The figure demonstrates that some clusters are composed of numerous, smaller
subgroups representing different cell subpopulations which are not visible under
standard parameter settings.

When dealing with data containing millions of data points, standard t-SNE
embeddings often become unwieldy -- cluster boundaries are blurred, large
clusters absorb smaller ones, and relationships between clusters become
increasingly difficult to interpret. We constructed Fig.~\ref{fig:cao}.a from
the data containing expression profiles of over two million single cells
captured at different time points in mouse development. The embedding indicates
numerous clusters with transitions between time points, as marked by the
color-coding, that are difficult to interpret. Kobak \& Berens observed that
increasing attractive forces between similar data points controlled via the
\textit{exaggeration} parameter leads to more compact clusters, and
subsequently, more informative visualizations~\citep{kobak2019art}. For instance,
\ref{fig:cao}.b doubles the default exaggeration, which uncovers some of the
data's overall structure. Further doubling the exaggeration in \ref{fig:cao}.c
allows us to observe that the data is comprised of two main groups of cells and
eight somewhat smaller clusters. The visualization also reveals several tiny
clusters, possibly corresponding to rare cell types.

\begin{figure*}[htbp]
  \includegraphics[width=\textwidth]{cao2019}
  \caption{\label{fig:cao}
  Increasing the exaggeration parameter leads to compact clusters, highlighting
  the data's global organization and emphasizing the continuous nature of
  cell state transitions. The data set from \citet{cao2019single}
  contains expression profiles from 2,058,652 single cells. The data were
  collected from mice embryos at different developmental stages at daily
  intervals after 9.5 to 13.5 days. \textbf{(c)} reveals that the data is
  comprised of two main components -- the neural tube and mesenchymal
  cells -- as well as several other smaller clusters. The
  colors indicate developmental progression with red indicating
  least-developed cells and blue indicating most developed cells. The
  overall developmental trajectory is most apparent with higher
  exaggeration levels, showing red cells slowly transitioning into blue
  cells. Progressively easing the exaggeration factor uncovers finer
  clusters within the larger groups, as shown in \textbf{(b)} with
  exaggeration of two and subsequently in \textbf{(a)}, where we show the
  standard t-SNE with no exaggeration. 32,011 putative doublets are
  excluded from the visualizations.
}
\end{figure*}

Exaggeration can highlight transitions between cell states in developmental
studies. Standard t-SNE often produces embeddings with clearly defined, discrete
clusters. We can adjust the level of granularity and resolution of the clusters
with several parameters in Fig.~\ref{fig:tasic}. However, discrete clusters are
often undesired in developmental studies where cells' state is assumed to follow
a continuous transition path. To this end, other embedding techniques such as
UMAP and ForceAtlas2 are used to better capture the continuity between cell
states. Recently, \citet{bohm2020unifying} showed that
embeddings produced by t-SNE with exaggeration values of 4 and $\sim30$
construct embeddings which are markedly similar to UMAP and ForceAtlas2,
respectively. For example, in Fig.~\ref{fig:cao}.a, the developmental trajectory
between different time points is difficult to observe due to many sprawled out
clusters. On the other hand, it is easier to trace the development when we
increase the exaggeration factor from $1$ to $2$ to $4$ in
Figs.~\ref{fig:cao}.b-c.

\subsection{Embedding New Samples}

Unlike other popular dimensionality reduction techniques such as principal
component analysis or autoencoders, t-SNE is a non-parametric method and does
not define an explicit mapping to the embedding space. Therefore, embeddings of
new data points need to be found through
optimization~\citep{policar2019embedding}. \opentsne\ is currently the only
publicly available library allowing users to add new samples to existing
embeddings in a principled manner.

Figs.~\ref{fig:macosko}.b and \ref{fig:macosko}.c demonstrate how we can use a
previously labeled single-cell data set and embed cells from a separate
experiment into the reference landscape. The reference data from
\citet{macosko2015highly} contains gene expression
profiles from mouse retinal cells. By embedding the samples from a similar
experiment on bipolar retinal cells by \citet{shekhar2016comprehensive}, we can correctly map the bipolar cell
clusters onto the reference embedding.

Embedding single cells into existing reference atlases can also be useful for
cell-type classification in cases of unknown cell identities. For instance, in
Fig.~\ref{fig:transform}, we construct a reference embedding using labeled data
from \citet{hochgerner2018conserved} containing
gene-expression profiles of cells from the mouse brain. The authors assign a
type to each cell. We can verify their classification accuracy by visualizing
the expression of well-established gene markers for the major cell types. We
then embed cells from \citet{harris2018classes} into the
constructed cell atlas. In \citet{harris2018classes}, labels are provided only for
neuronal cells. In the resulting mapping, we can quickly identify other
non-neuronal cell types, including oligodendrocytes and astrocytes. We can
further use marker genes to validate that the mapping in the reference landscape
is correct.

\begin{figure*}[htbp]
  \includegraphics[width=\textwidth]{transform_hochgerner}
  \caption{\label{fig:transform}
  \opentsne\ supports embedding new samples into an existing reference t-SNE
  landscape. For the series of visualizations shown in this figure, we
  first construct a t-SNE embedding for the data from 
  \citet{hochgerner2018conserved} containing 24,185
  developing, single cells from the mouse hippocampus. The data contains
  gene expression in different neurons, supporting glia, and other
  vascular cells (upper left). Data points representing cells are colored
  according to cell-types assigned in the original publication; see the
  legend from Fig.~\ref{fig:tasic} to map colors to cell-type. We then
  embed new, hippocampal cells collected in a study by
  \citet{harris2018classes} using the embedding of \citet{hochgerner2018conserved}
  data as a reference. In their study, \citet{harris2018classes}
  collected 6,971 single-cells and focused on identifying different
  types of inhibitory neurons. However, almost half of the collected cells
  are not neurons and were left uncharacterized. Inspecting the embeddings
  of these cells in the reference embedding (bottom left) reveals that in
  addition to inhibitory neurons, the data contains several supporting
  glial cells as well as a small population of endothelial cells. We can
  verify our approach's accuracy by inspecting marker genes for the major
  cell types in the reference (top row) and embedded samples (bottom row).
}
\end{figure*}

The examples presented above demonstrate how to use \opentsne\ to quickly gain
insight into newly-sequenced, single-cell data sets by utilizing existing cell
atlases. The approach is general and not limited to single-cell gene expression,
and one can, in principle, apply it to any tabular data set regardless of field.

\subsection{Versatility}

Versatility, the ability to use and combine different optimization approaches to
construct different embedding spaces, is another of \opentsne's core design
principles. \citet{kobak2019art} recently provided several recommendations and
tricks to obtain better and more meaningful t-SNE visualizations. These include
multi-scale similarity kernels, perplexity annealing, and increasing
exaggeration when working with massive data sets. \opentsne\ provides a flexible
API to incorporate these improvements in just a few lines of code. Furthermore,
\opentsne\ supports custom affinity models, enabling users to construct t-SNE
embeddings on non-tabular relational data: the only requirement imposed by the
affinity-model is some notion of similarity between data points. Finally,
\opentsne's comprehensive callback system can be utilized to monitor and adapt
different stages of the optimization phase and has been used to construct
visually appealing animations of the t-SNE optimization process.

\subsection{Speed}

One of the t-SNE's common criticisms is limited scalability to large data sets
containing, for instance, millions of data
points~\citep{becht2019dimensionality}. The culprit for slow response time stems
from a particular optimization procedure and its specific implementation in
popular Python libraries. Until quite recently, most popular implementations of
t-SNE were based on the Barnes-Hut approximation scheme developed by
\citet{van2014accelerating} with asymptotic time complexity $\mathcal{O}(N \log
N)$, where $N$ is the number of data items ({\em e.g.} cells). The most
widely-used implementation of t-SNE came from \pkg{scikit-learn}, which exhibits
long runtimes when compared to its C++ counterpart -- \pkg{MulticoreTSNE}. The
multi-threaded \pkg{MulticoreTSNE} implementation~\citep{Ulyanov2016} can
construct t-SNE embeddings of millions of data points in a matter of hours on
widely-accessible, consumer-grade processors (Fig.~\ref{fig:benchmarks}).
However, recently, \citet{linderman2019fast} developed a new approximation
scheme -- FIt-SNE -- which further reduces the asymptotic time complexity to
$\mathcal{O}(N)$. We include this approach in \opentsne, enabling the embedding
of large data sets in a matter of minutes.

Fig.~\ref{fig:benchmarks} benchmarks four popular Python t-SNE implementations,
including those from \pkg{scikit-learn} (v0.23.1), \pkg{MulticoreTSNE} (v0.1),
\pkg{FIt-SNE} (v1.1.0), and our \opentsne\ (v0.6.0). We perform benchmarks on
two computational platforms, one representing the usage on a personal computer
and the other the utility of these libraries on a high-performance computing
platform. The Intel(R) Core i7 is commonly found in consumer-grade laptop
computers, while Intel(R) Xeon(R) processors appear in high-performance
computing machines. Benchmarks were run for $1,000$ iterations with the original
t-SNE parameters, as some implementations do not allow their modification.

\begin{figure*}[ht]
  \includegraphics[width=\textwidth]{benchmarks}
  \caption{\label{fig:benchmarks}
  We benchmark \opentsne\ (v0.6.0) against three popular open-source
  implementations from \pkg{scikit-learn}~\citep{pedregosa2011scikit}
  (v0.23.1), \pkg{MulticoreTSNE}~\citep{Ulyanov2016} (v0.1), and
  \pkg{FIt-SNE}~\citep{linderman2019fast} (v1.1.0). Experiments were run on a
  consumer-grade Intel Core i7-7700HQ processor found in laptop computers, and
  on a server-grade Intel Xeon E5-2650. To generate benchmark data sets of
  different sizes, we subsampled data from the 10X Genomics 1.3 million mouse
  brain data set five times, resulting in five different data sets for each
  size. In total, we run each implementation on 30 different data sets. Notice
  that \opentsne\ scales similarly to \pkg{FIt-SNE}, as they both use the
  same interpolation-based approximation scheme, while \pkg{scikit-learn} and
  \pkg{MulticoreTSNE} utilize the Barnes-Hut approximation.
}
\end{figure*}

Benchmark results (Fig.~\ref{fig:benchmarks}) confirm that both \pkg{FIt-SNE}
and \opentsne\ scale better than their Barnes-Hut counterparts --
\pkg{scikit-learn} and \pkg{MulticoreTSNE}. While \opentsne's implementation
uses Python which incurs some runtime overhead compared to its C++ counterpart
\pkg{FIt-SNE}, the two libraries' speeds are surprisingly comparable. Modern
computer processors contain multiple cores, allowing us to use multi-threading,
which further reduces the gap between \opentsne\ and \pkg{FIt-SNE}. On the
server-grade processor, \opentsne\ is slightly faster than its pure C++
counterpart when utilizing multiple cores. \opentsne\ uses \pkg{numpy} for most
linear algebra operations, which may make better use of the Intel Math Kernel
Library (MKL), which is aggressively optimized on Intel(R) Xeon(R) processors.

\opentsne\ provides a flexible API, allowing splitting up the
embedding-construction process into several parts, caching slow operations.
Running t-SNE optimization in stages enables users to quickly experiment with
different parameter settings and iterate on their final visualizations.

\subsection{Ease of Use}

Intuitive access and simple installation procedures almost universally correlate
with the widespread adoption of novel computational techniques. While the t-SNE
implementation from \pkg{scikit-learn} fits this requirement, the implementation
proves prohibitively slow for even moderately-sized data sets that span tens of
thousands of data records. Other C++ implementations such as \pkg{MulticoreTSNE}
and \pkg{FIt-SNE} exhibit better scaling in more massive data sets, but do not
provide precompiled binaries and require users to compile the software
themselves. This problem is critical, for instance, for users of the Windows
operating system, where the C++ compiler does not come with the system, making
the correct configuration of current t-SNE implementations cumbersome.

We designed \opentsne\ to be accessible to a broader audience. We provide
precompiled binaries for all major Python versions on all major platforms,
making the installation process as seamless as possible. One can install
\opentsne\ through the Python Package Index (\pkg{PyPI}) or \pkg{conda} from the
\pkg{conda-forge} channel, the two most widely adopted Python package managers.
\opentsne's interface is inspired by \pkg{scikit-learn}, which is well
established in the Python data science ecosystem. \opentsne\ implements
multi-threaded versions of both the Barnes-Hut and FIt-SNE approximation
schemes, enabling it to be applied to data sets containing millions of data
points. While the Python virtual machine inevitably introduces some performance
overhead, the runtime is comparable to its C++ counterpart -- \pkg{FIt-SNE}.
Finally, \opentsne\ is extensible. Its modular design enables researchers to
quickly experiment with different parameter settings and easily incorporate
custom components into the software. We provide a full feature list and
comparison to other popular t-SNE implementations in Table~\ref{tab:features}.

\begin{table}
\begin{center}\small
\newcommand*\rot{\rotatebox{45}}
\renewcommand{\arraystretch}{1.25}

\begin{tabular}{l c c c c c c c}
\toprule
\setlength\tabcolsep{6pt}
& \rot{\pkg{scikit-learn}} & \rot{\pkg{BHTSNE}} & \rot{\pkg{MulticoreTSNE}} & \rot{\pkg{FIt-SNE}} & \rot{\pkg{openTSNE}} \\

\toprule
\textbf{Ease of Installation}\\
Native Python installation & \checkmark & & & & \checkmark \\
\pkg{PyPI} package & \checkmark & & \checkmark & & \checkmark \\
\pkg{conda} package & \checkmark & & & & \checkmark \\

\hline
\textbf{Approximation Schemes}\\
Barnes-Hut ($\mathcal{O}(N \log N)$) & \checkmark & \checkmark & \checkmark & & \checkmark \\
FIt-SNE ($\mathcal{O}(N)$) & & & & \checkmark & \checkmark \\

\hline
\textbf{Advanced Features and Extensions}\\
Multiscale Gaussian kernels & & & & \checkmark & \checkmark \\
Fully-custom affinity kernels & & & & & \checkmark \\
Variable degrees of freedom & & & & \checkmark & \checkmark \\
Variable exaggeration & & & & \checkmark & \checkmark \\
Globally-consistent initialization & & & & \checkmark & \checkmark \\
Automatic learning rate & & & & \checkmark & \checkmark \\
Adding samples to embeddings & & & & & \checkmark \\
Callback system & & & & & \checkmark \\
Interactive optimization  & & & & & \checkmark \\

\hline
\textbf{Project Quality}\\
User documentation & \checkmark & & & & \checkmark \\
End-to-end usage examples & \checkmark & & & \checkmark & \checkmark \\
API reference & \checkmark & & & & \checkmark \\
Continuous integration & \checkmark & & \checkmark & & \checkmark \\
\bottomrule
\end{tabular}
\end{center}

\caption{\label{tab:features}
  We compare the features of \opentsne~(v0.6.0) to four popular open-source
  t-SNE implementations from \pkg{scikit-learn}~(v0.23.1),
  \pkg{BHTSNE}~(master), \pkg{MulticoreTSNE}~(v0.1), and
  \pkg{FIt-SNE}~(v1.1.0).
}
\end{table}

\section{Conclusion}
We introduce \opentsne, the most feature-complete, open-source, Python
implementation of the widely popular t-SNE visualization algorithm. The library
incorporates the latest developments to the t-SNE algorithm and makes them
easily accessible through a familiar API. \opentsne\ implements efficient
approximation schemes, making it suitable for the visualization of large data
sets, containing up to millions of data points. Finally, \opentsne\ is easy to
install and comes with precompiled binaries available for all major platforms.

%% -- Optional special unnumbered sections -------------------------------------

\section*{Availability and Documentation}

\opentsne\ is distributed under the BSD-3-Clause License and its source code is
publicly available at \url{https://github.com/pavlin-policar/openTSNE}.
\opentsne\ can also be installed from \pkg{PyPI} and \pkg{conda-forge}.
End-to-end usage examples, documentation, and API reference are all available at
\url{https://opentsne.readthedocs.io}.

\section*{Computational details}

The results in this paper were obtained using \proglang{Python}~3.7.6 with the
\pkg{openTSNE}~XXX package. \opentsne\ has three direct dependencies:
\pkg{numpy}, \pkg{scipy}, and \pkg{scikit-learn}. In our experiments, we used
\pkg{numpy}~v1.18.1, \pkg{scipy}~v1.4.1. \pkg{scikit-learn}~v0.23.1. As these
packages include their own downstream dependencies, the full environment used to
produce the figures is included in the accompanying GitHub repository
\url{https://github.com/pavlin-policar/opentsne-paper}.

\section*{Acknowledgments}

We want to thank Dmitry Kobak for helpful discussions regarding t-SNE and
his contributions to the source code. We would also like to thank George
Linderman for his help with the FIt-SNE algorithm. This work was supported by
Slovenian Research Agency (P2-0209).

%% -- Bibliography -------------------------------------------------------------
%% - References need to be provided in a .bib BibTeX database.
%% - All references should be made with \cite, \citet, \citep, \citealp etc.
%%   (and never hard-coded). See the FAQ for details.
%% - JSS-specific markup (\proglang, \pkg, \code) should be used in the .bib.
%% - Titles in the .bib should be in title case.
%% - DOIs should be included where available.

\bibliography{references}

\end{document}

### 2021-04-26 paper/main.tex
Remove pdf from repo, minor edits

@@
 for efficient dimensionality reduction techniques and their implementations,
-which can produce informative data visualizations. Popular approaches include
-principal component analysis, multidimensional scaling, t-distributed stochastic
+capable of producing informative data visualizations. Popular approaches include
+principal component analysis, multidimensional scaling, $t$-distributed stochastic
 neighbor embedding (t-SNE)~\citep{maaten2008visualizing}, and uniform manifold
@@
 of these shortcomings have recently been addressed. \citet{linderman2019fast}
-developed FIt-SNE, an efficient approximation scheme which massively improves
+developed FIt-SNE, an efficient approximation scheme that massively improves
 the scalability of t-SNE, achieving linear time complexity in the number of
@@
 
-\subsection{t-distributed stochastic neighbor embedding} \label{sec:meth.tsne}
+\subsection{$t$-distributed stochastic neighbor embedding} \label{sec:meth.tsne}
 
-t-distributed stochastic neighbor embedding (t-SNE) is a non-linear
+$t$-distributed stochastic neighbor embedding (t-SNE) is a non-linear
 dimensionality reduction method that to aims to find a low-dimensional embedding
@@
 This enables t-SNE to adapt to the varying density of the data in the
-multi-dimensional space. The perplexity can be interpreted as as the continuous
+multi-dimensional space. The perplexity can be interpreted as the continuous
 analogue to the number of nearest neighbors to which the distances will be
@@
 Optimization is performed using batch gradient descent using the delta-bar-delta
-update rule~\citep{jacobs1988increased}. The t-SNE optimization proceedure
+update rule~\citep{jacobs1988increased}. The t-SNE optimization procedure
 consists of two phases: in the first \textit{early exaggeration} phase, the
@@
 reduce the time complexity of t-SNE to $\mathcal{O}(N)$. This allows modern
-t-SNE implementations to be leveraged for the visulization of data sets
+t-SNE implementations to be leveraged for the visualization of data sets
 containing up to millions of data points.
@@
 \subsubsection*{Repulsive Forces}
-The repulsive term can similarily be approximated, motivated by methods from
+Similarly, the repulsive term can also be approximated, motivated by methods from
 particle simulations. \citet{van2014accelerating} proposed an approach based on
@@
 $\mathcal{O}(M \log N)$ and $\mathcal{O}(M)$, respectively. Special care must be
-taken to tune the learning rate during optimization as the existing values
-$\eta=200$ and $\eta=N/12$ result in unstable optimization.
+taken to tune the learning rate during optimization as the default values
+$\eta=200$ and $\eta=N/12$ result in highly unstable optimization.
 
@@
 often arbitrarily scattered in the embedding space. However, when using
-initalization schemes based on PCA or spectral embeddings, the clusters
+initialization schemes based on PCA or spectral embeddings, the clusters
 identified in the resulting embedding are typically grouped in a globally
@@
 
-Embeddings produced by standard t-SNE often makes use all available space and
+Embeddings produced by standard t-SNE often make use of all available space and
 separates clusters by only thin boundaries. When working with large data sets,
@@
 groups appear equidistant from one another.  Other dimensionality reduction
-methods, such as UMAP or
-ForceAtlas2, produce embeddings where clusters
-appear more compact and the white-space separating the clusters may be
-interpreted as a loose measure of distance. Recently, \citet{bohm2020unifying}
-showed that the exaggeration factor $\rho$ could be used to produce layouts more
-similar to UMAP and ForceAtlas2. By incorporating exaggeration into later phases
-of the optimization, t-SNE introduces more white-space between clusters which better
+methods, such as UMAP and ForceAtlas2, produce embeddings where clusters appear
+more compact and the white space separating the clusters may be interpreted as a
+loose measure of distance. Recently, \citet{bohm2020unifying} showed that the
+exaggeration factor $\rho$ could be used to produce layouts more similar to UMAP
+and ForceAtlas2. By incorporating exaggeration into later phases of the
+optimization, t-SNE introduces more white space between clusters which better
 reflects the global relations between clusters.
@@
 pipeline. \opentsne\ aims to make the latest theoretical advances in t-SNE
-accessible to the wider data-science community. These features are easily
+accessible to the wider data science community. These features are easily
 accessible through an intuitive API, which closely follows the style
@@
 The figure demonstrates that some clusters are composed of numerous, smaller
-subgroups representing different cell subpopulations which are not visible under
+subgroups representing different cell subpopulations that are not visible under
 standard parameter settings.
@@
 The results in this paper were obtained using \proglang{Python}~3.7.6 with the
-\pkg{openTSNE}~XXX package. \opentsne\ has three direct dependencies:
+\pkg{openTSNE}~v0.6.0 package. \opentsne\ has three direct dependencies:
 \pkg{numpy}, \pkg{scipy}, and \pkg{scikit-learn}. In our experiments, we used

### 2021-05-13 paper/main.tex
minor changes

@@
 \Abstract{
-One of the most popular techniques for constructing visualizations of large,
-high-dimensional data sets is $t$-distributed stochastic neighbor embedding
-(t-SNE). Recently, several extensions have been proposed to address
-issues of scalability and the quality of the resulting visualizations. We
-introduce \opentsne, a modular Python library that implements the core t-SNE
-algorithm and its many extensions. The library is fast and scales up to data
-sets containing millions of data points. \opentsne\ is distributed under the
-BSD-3-Clause License and its source code is publicly available at
-\url{https://github.com/pavlin-policar/openTSNE}.
+One of the most popular techniques for constructing visualizations of large, high-dimensional data sets is $t$-distributed stochastic neighbor embedding (t-SNE).
+Recently, several extensions have been proposed to address scalability issues and the quality of the resulting visualizations.
+We introduce \opentsne, a modular Python library that implements the core t-SNE algorithm and its many extensions.
+The library is fast and can treat data sets containing millions of data points.
+We distribute \opentsne\ under the BSD-3-Clause License.
+Its source code is available at \url{https://github.com/pavlin-policar/openTSNE}.
 }
@@
 
-The ever-growing volumes of high-dimensional data sets in machine learning calls
-for efficient dimensionality reduction techniques and their implementations,
-capable of producing informative data visualizations. Popular approaches include
-principal component analysis, multidimensional scaling, $t$-distributed stochastic
-neighbor embedding (t-SNE)~\citep{maaten2008visualizing}, and uniform manifold
-approximation and projections (UMAP)~\citep{2018arXivUMAP}. Among these, t-SNE
-has received much attention as it can address high volumes of data and reveal
-the underlying data structure. For instance, t-SNE is widely used in the
-bioinformatics community in areas such as single-cell
-transcriptomics~\citep{macosko2015highly,cao2019single,tasic2018shared}, human
-genetics~\citep{hirata2019genetic}, metagenomic
-assembly~\citep{beaulaurier2018metagenomic}, the spatial organization of
-microbial communities~\citep{sheth2019spatial}, and
-metabolomics~\citep{tkachev2019differences}. Reports on single-cell gene
-expression data, our running example, often start with an overview of the cell
-landscape, where t-SNE is used to embed high-dimensional expression profiles
-into a two-dimensional space. Figs.~\ref{fig:macosko}.a and \ref{fig:macosko}.b
-show two such embeddings.
+The ever-growing volumes of high-dimensional data sets in machine learning call for efficient dimensionality reduction techniques and their implementations, capable of producing informative data visualizations.
+Popular approaches include principal component analysis, multidimensional scaling, $t$-distributed stochastic neighbor embedding (t-SNE)~\citep{maaten2008visualizing}, and uniform manifold approximation and projections (UMAP)~\citep{2018arXivUMAP}.
+Among these, t-SNE has received much attention as it can address high volumes of data and reveal the underlying data structure.
+For instance, t-SNE is widely used in the bioinformatics community in areas such as single-cell transcriptomics~\citep{macosko2015highly,cao2019single,tasic2018shared}, human genetics~\citep{hirata2019genetic}, metagenomic assembly~\citep{beaulaurier2018metagenomic}, the spatial organization of microbial communities~\citep{sheth2019spatial}, and metabolomics~\citep{tkachev2019differences}.
+To visually explore the data structure, the objects of interest, such as diseased and healthy tissues or single cells, are profiled through thousands of features that include, for example, expression of genes or concentration of metabolites.
+The task of t-SNE is to embed the objects of interest within a low, preferably two-dimensional embedding.
+Reports on single-cell gene expression data, our running example, often start with an overview of the cell landscape and use t-SNE to embed high-dimensional expression profiles into a two-dimensional space.
+Figs.~\ref{fig:macosko}.a and \ref{fig:macosko}.b show two such embeddings.
 
@@
   \caption{\label{fig:macosko}
-  We use \opentsne\ to generate three t-SNE embeddings and demonstrate recent
-  theoretical advances. The data in \textbf{(a)} and \textbf{(b)} represent
-  44,808 single-cell gene-expression profiles of mouse retinal cells from 
-  \citet{macosko2015highly}. The data in \textbf{(c)} additionally contains
-  27,499 expression profiles from mouse retinal cells from
-  \citet{shekhar2016comprehensive}. \textbf{(a)} We construct a t-SNE embedding
-  following the parameter choices from the original publication by
-  \citet{maaten2008visualizing}. The visualization shows no preservation of the
-  global organization of clusters, resulting from random initialization and an
-  affinity model focused on preserving local neighborhoods. \textbf{(b)} A
-  modern t-SNE embedding, utilizing the latest theoretical advances and
-  practical recommendations constructed using a multi-scale affinity model,
-  preserving both short-range and long-range interactions between data points
-  and initialized so that the global layout is as meaningful as possible. Unlike
-  in \textbf{(a)}, the green and blue clusters representing different sub-types
-  of amacrine and bipolar cells are now localized to the same regions of the
-  space, indicating a higher level of similarity than to other cell types. The
-  embedding in \textbf{(c)} shows how existing t-SNE reference atlases can be
-  used to place new samples into existing embeddings. The positions of new data
-  points correspond to cell types from the reference atlas.
+We use \opentsne\ to generate three t-SNE embeddings and demonstrate recent theoretical advances.
+The data in \textbf{(a)} and \textbf{(b)} includes 44,808 mouse retinal cells (dots in the plots) that are described with high-dimensional gene-expression profiles from \citet{macosko2015highly}.
+The data in \textbf{(c)} additionally contains 27,499 expression profiles from mouse retinal cells from \citet{shekhar2016comprehensive}.
+\textbf{(a)} We construct t-SNE embedding following the parameter choices from the original publication by \citet{maaten2008visualizing}.
+The visualization shows no preservation of the global organization of clusters, resulting from random initialization and an affinity model focused on preserving local neighborhoods.
+\textbf{(b)} A modern t-SNE embedding, utilizing the latest theoretical advances and practical recommendations constructed using a multi-scale affinity model, preserving both short-range and long-range interactions between data points and initialized so that the global layout is as meaningful as possible.
+Unlike in \textbf{(a)}, the green and blue clusters representing different sub-types of amacrine and bipolar cells are now localized to the same regions of the space, indicating a higher level of similarity than to other cell types.
+The embedding in \textbf{(c)} shows how existing t-SNE reference atlases can be used to place new samples into existing embeddings.
+The positions of new data points correspond to cell types from the reference atlas.
 }
@@
 
-Despite its utility, t-SNE has often been criticized for its limited
-scalability, lack of global organization -- t-SNE identifies well-defined
-clusters that may be arbitrarily scattered throughout the low-dimensional space
--- and the absence of theoretically-founded methods to map new data into
-existing embeddings~\citep{ding2018interpretable,becht2019dimensionality}. Most
-of these shortcomings have recently been addressed. \citet{linderman2019fast}
-developed FIt-SNE, an efficient approximation scheme that massively improves
-the scalability of t-SNE, achieving linear time complexity in the number of
-samples. \citet{kobak2019art} proposed several techniques to improve global
-cluster coherence, including estimating similarities with a mixture of Gaussian
-kernels. In our previous work, we introduced a principled approach for embedding
-new samples into existing visualizations~\citep{policar2019embedding}.
+Despite its utility, t-SNE has often been criticized for its limited scalability, lack of global organization -- t-SNE identifies well-defined clusters that may be arbitrarily scattered throughout the embedding, and the absence of theoretically-founded methods to map new data into existing embeddings~\citep{ding2018interpretable,becht2019dimensionality}.
+Most of these shortcomings have recently been addressed.
+\citet{linderman2019fast} developed FIt-SNE, an efficient approximation scheme that massively improves the scalability of t-SNE, achieving linear time complexity in the number of samples.
+\citet{kobak2019art} proposed several techniques to improve global cluster coherence, including estimating similarities with a mixture of Gaussian kernels.
+In our previous work, we introduced a principled approach for embedding new samples into existing visualizations~\citep{policar2019embedding}.
 
-However, easy access to efficient implementations almost universally correlates
-with the widespread adoption of novel computational techniques. Despite the many
-theoretical advancements, most popular t-SNE libraries have been slow to
-incorporate them into their implementations. In this paper, we present \opentsne,
-our open-source Python implementation of t-SNE and its many recently proposed
-extensions. \opentsne\ is easy to install and includes precompiled binaries
-available through popular Python package managers. It provides a familiar API,
-making it suitable as a drop-in replacement for existing t-SNE implementations.
+However, easy access to efficient implementations almost universally correlates with the widespread adoption of novel computational techniques.
+Despite the many theoretical advancements, the most popular t-SNE libraries have been slow to incorporate them into their implementations.
+In this paper, we present \opentsne, our open-source Python implementation of t-SNE, and the recently proposed extensions to t-SNE that \opentsne\ implements.
+\opentsne\ is easy to install and includes precompiled binaries available through popular Python package managers.
+It provides a familiar API, making it suitable as a drop-in replacement for existing t-SNE implementations.
 
@@
 
-Optimization is performed using batch gradient descent using the delta-bar-delta
-update rule~\citep{jacobs1988increased}. The t-SNE optimization procedure
-consists of two phases: in the first \textit{early exaggeration} phase, the
-attractive forces between data points are increased by some factor $\rho$,
-typically set to $12$, so that points in the embedding can more easily move
-throughout the space and settle near their respective neighbors. In the second
-phase of the optimization, the attractive forces are then reverted to their
-original values with $\rho=1$.
+Optimization is performed using batch gradient descent using the delta-bar-delta update rule~\citep{jacobs1988increased}.
+The t-SNE optimization procedure consists of two phases: in the first \textit{early exaggeration} phase, the attractive forces between data points are increased by some factor $\rho$, typically set to $12$, so that points in the embedding can more easily move throughout the space and settle near their respective neighbors.
+In the second phase of the optimization, the attractive forces are then reverted to their original values with $\rho=1$.
 
@@
 
-One of t-SNE's criticisms is that it fails to preserve long-range distances but
-instead focuses on capturing the local structure of the
-data manifold~\citep{becht2019dimensionality}. Recently, several approaches have been
-proposed to improve the global organization of the resulting embeddings.
+One of t-SNE's criticisms is that it fails to preserve long-range distances but instead focuses on capturing the local structure of the data manifold~\citep{becht2019dimensionality}.
+Recently, several approaches have been proposed to improve the global organization of the resulting embeddings.
 
-\citet{kobak2019umap} showed that the initialization of force-directed layout
-algorithms, such as t-SNE, UMAP~\citep{2018arXivUMAP}, and ForceAtlas2~\citep{jacomy2014forceatlas2}, largely dictates the global
-consistency of resulting embeddings. When initialized randomly, clusters are
-often arbitrarily scattered in the embedding space. However, when using
-initialization schemes based on PCA or spectral embeddings, the clusters
-identified in the resulting embedding are typically grouped in a globally
-coherent manner.
+\citet{kobak2019umap} showed that the initialization of force-directed layout algorithms, such as t-SNE, UMAP~\citep{2018arXivUMAP}, and ForceAtlas2~\citep{jacomy2014forceatlas2}, largely dictates the global consistency of resulting embeddings.
+When initialized randomly, clusters are often arbitrarily scattered in the embedding space.
+However, when using initialization schemes based on PCA or spectral embeddings, the clusters identified in the resulting embedding are typically grouped in a globally coherent manner.
 
-In standard t-SNE, distances between data points are converted to similarities
-through the use of Gaussian kernels. The perplexity of these kernels governs the
-sizes of the local neighborhoods which are to be preserved. One easy way to
-improve the global consistency of resulting embeddings is to increase the sizes
-of these neighborhoods. This can be achieved by increasing the perplexity
-parameter. However, this often comes at the price of local structure.
-\citet{kobak2019art} propose to instead use mixtures of Gaussians to better
-preserve short-range, as well as long-range distances and, thus, achieve a
-better trade-off between local and global structure.
+In standard t-SNE, distances between data points are converted to similarities through the use of Gaussian kernels.
+The perplexity of these kernels governs the sizes of the local neighborhoods which are to be preserved.
+One easy way to improve the global consistency of resulting embeddings is to increase the sizes of these neighborhoods.
+This can be achieved by increasing the perplexity parameter.
+However, this often comes at the price of local structure.
+\citet{kobak2019art} propose to use mixtures of Gaussians instead to preserve better short-range and long-range distances and, thus, achieve a better trade-off between local and global structure.
 
-Embeddings produced by standard t-SNE often make use of all available space and
-separates clusters by only thin boundaries. When working with large data sets,
-this often obscures the global relationships between clusters as all neighboring
-groups appear equidistant from one another.  Other dimensionality reduction
-methods, such as UMAP and ForceAtlas2, produce embeddings where clusters appear
-more compact and the white space separating the clusters may be interpreted as a
-loose measure of distance. Recently, \citet{bohm2020unifying} showed that the
-exaggeration factor $\rho$ could be used to produce layouts more similar to UMAP
-and ForceAtlas2. By incorporating exaggeration into later phases of the
-optimization, t-SNE introduces more white space between clusters which better
-reflects the global relations between clusters.
+Embeddings produced by standard t-SNE often use all available space and separate clusters by only thin boundaries.
+When working with large data sets, this often obscures the global relationships between clusters as all neighboring groups appear equidistant from one another.
+Other dimensionality reduction methods, such as UMAP and ForceAtlas2, produce embeddings where clusters appear more compact, and the white space separating the clusters may be interpreted as a loose measure of distance.
+Recently, \citet{bohm2020unifying} showed that the exaggeration factor $\rho$ could be used to produce layouts more similar to UMAP and ForceAtlas2.
+By incorporating exaggeration into later phases of the optimization, t-SNE introduces white space between clusters, which better reflects the global relations between clusters.
 
 Standard t-SNE reveals the clustering structure at a single level of resolution.
-While the perplexity parameter can be used to control the trade-off between
-local and global structure, this can be time-consuming, and small, well-defined
-clusters can be missed. Alternatively, \citet{kobak2019heavy} suggest that
-varying the degrees of freedom in the $t$-distribution can be used to explore
-the clustering structure at different levels of resolution. Modifying
-Eqn.~(\ref{eq:cauchy_kernel}) to $q_{ij} \propto \left ( 1 + || \mathbf{y}_i -
-\mathbf{y}_j ||^2 / \alpha \right )^{-\alpha}$ allows us to use heavier-tailed
-distributions to model distances in the embedding space, which acts to highlight
-small subgoups in the resulting visualization.
+While the perplexity parameter can be used to control the trade-off between local and global structure, this can be time-consuming, and small, well-defined clusters can be missed.
+Alternatively, \citet{kobak2019heavy} suggest that varying the degrees of freedom in the $t$-distribution can be used to explore the clustering structure at different levels of resolution.
+Modifying Eqn.~(\ref{eq:cauchy_kernel}) to $q_{ij} \propto \left ( 1 + || \mathbf{y}_i - \mathbf{y}_j ||^2 / \alpha \right )^{-\alpha}$ allows us to use heavier-tailed distributions to model distances in the embedding space, which acts to highlight small subgroups in the resulting visualization.
 
@@
 
-We introduce \opentsne, a comprehensive Python library that implements t-SNE and
-its numerous recently proposed extensions, including the implementation of
-efficient approximation schemes allowing the embedding of millions of data
-points~\citep{van2014accelerating,linderman2019fast}, the addition of new
-data points into existing embeddings~\citep{policar2019embedding}, various
-initialization improvements~\citep{kobak2019umap},
-extensions~\citep{kobak2019heavy,kobak2019art}, and improved parameter
-defaults~\citep{belkina2019automated}. Combined, these improvements result in
-the faster generation of more informative visualizations.
+We introduce \opentsne, a comprehensive Python library that implements t-SNE and its numerous recently proposed extensions, including the implementation of efficient approximation schemes allowing the embedding of millions of data points~\citep{van2014accelerating,linderman2019fast}, the addition of new data points into existing embeddings~\citep{policar2019embedding}, various initialization improvements~\citep{kobak2019umap}, extensions~\citep{kobak2019heavy,kobak2019art}, and improved parameter defaults~\citep{belkina2019automated}.
+When combined, these improvements result in the faster generation of more informative visualizations.
 
 \opentsne\ is compatible with the Python data science ecosystem (e.g.
-\pkg{numpy}, \pkg{scikit-learn}, \pkg{scanpy}), providing a familiar
-and intuitive API to users. Its modular design encourages extensibility and
-experimentation with various parameter settings and changes to the analysis
-pipeline. \opentsne\ aims to make the latest theoretical advances in t-SNE
-accessible to the wider data science community. These features are easily
-accessible through an intuitive API, which closely follows the style
-\pkg{scikit-learn}~\citep{sklearn_api}, a popular and widely used toolkit for
-machine learning. The following snippet exemplifies this interface and
-demonstrates how \opentsne\ can be used to quickly generate several
-visualizations, each highlighting different aspects of the underlying data
-structure.
+\pkg{numpy}, \pkg{scikit-learn}, \pkg{scanpy}), providing a familiar and intuitive API to users.
+Its modular design encourages extensibility and experimentation with various parameter settings and changes to the analysis pipeline.
+\opentsne\ aims to make the latest theoretical advances in t-SNE accessible to the wider data science community.
+These features are easily accessible through an intuitive API, which closely follows the style \pkg{scikit-learn}~\citep{sklearn_api}, a popular and widely used toolkit for machine learning.
+The following snippet exemplifies this interface and demonstrates how \opentsne\ can be used to quickly generate several visualizations, each highlighting different aspects of the underlying data structure.
 \begin{CodeChunk}
@@
 \end{CodeChunk}
-The code snippet considers a \pkg{numpy} array or \pkg{scipy} sparse
-matrix \texttt{x} and creates three \texttt{TSNEEmbedding} objects. These
-represent fully-optimized t-SNE embeddings and are internally stored as \pkg{numpy}
-arrays. These can be manipulated and visualized using standard Python tools. We
-then select one of these embeddings and use it to embed new data points from
-\texttt{x\_new} into the embedding space, while keeping the reference embedding
-fixed. Additionally, \opentsne\ provides a more advanced interface that enables
-the user to change every aspect of the t-SNE optimization procedure. This
-interface is described in detail in the accompanying documentation.
-
-% \section{Existing t-SNE Implementations}
-
-% \opentsne\ is not the only available open-source implementation of the t-SNE 
-% algorithm, however, it provides the most user-friendly experience and is the most
-% feature-complete. A detailed comparison of several t-SNE implementations is
-% shown in Table~\ref{tab:features}.
-
-% \pkg{scikit-learn} is one of the most widely-used machine learning toolkits
-% in the Python data science ecosystem, and provides an implementation of the
-% t-SNE algorithm~\citep{pedregosa2011scikit}. However, the implementation from
-% \pkg{scikit-learn} utilizes an outdated approximation scheme and is
-% prohibitively slow for even moderately-sized data sets containing tens of
-% thousands of samples. \opentsne\ implements all available approximation schemes,
-% and defaults to the FIt-SNE approximation algorithm, which allows it to scale to
-% data sets containing up to millions of data points on standard, consumer-grade,
-% laptop computers. Performance benchmarks are show in Fig.~\ref{fig:benchmarks}.
-% \begin{figure*}[ht!]
-%   \includegraphics[width=\textwidth]{benchmarks}
-%   \caption{\label{fig:benchmarks}
-%   We benchmark \opentsne\ (v0.6.0) against three popular open-source
-%   implementations from \pkg{scikit-learn}~\citep{pedregosa2011scikit}
-%   (v0.23.1), \pkg{MulticoreTSNE}~\citep{Ulyanov2016} (v0.1), and
-%   \pkg{FIt-SNE}~\citep{linderman2019fast} (v1.1.0). Experiments were run on a
-%   consumer-grade Intel Core i7-7700HQ processor found in laptop computers, and
-%   on a server-grade Intel Xeon E5-2650. To generate benchmark data sets of
-%   different sizes, we subsampled data from the 10X Genomics 1.3 million mouse
-%   brain data set five times, resulting in five different data sets for each
-%   size. In total, we run each implementation on 30 different data sets. Notice
-%   that \opentsne\ scales similarly to \pkg{FIt-SNE}, as they both use the
-%   same interpolation-based approximation scheme, while \pkg{scikit-learn} and
-%   \pkg{MulticoreTSNE} utilize the Barnes-Hut approximation.
-% }
-% \end{figure*}
-
-% Pure C++ implementations, such as \pkg{BHTSNE}~\citep{van2014accelerating},
-% \pkg{MulticoreTSNE}~\citep{Ulyanov2016} and
-% \pkg{FIt-SNE}~\citep{linderman2019fast} are fast, but require users to
-% compile and install the packages themselves. This is especially problematic for
-% users of the Windows operating system, where the C/C++ compiler is not bundled
-% with the operating system, making their correct configuration and compilation
-% cumbersome. Additionally, these packages were not built with the Python ecosystem in
-% mind, and cannot be directly used as dependencies in other Python packages.
-% \opentsne\ provides several, well-documented installation options and can
-% readily be used as a dependency in larger projects. We provide precompiled
-% binaries through the Python Package Index (\pkg{PyPI}) or \pkg{conda} from
-% the \pkg{conda-forge} channel, allowing \opentsne\ to work out-of-the-box all
-% major platforms.
-
-% Another concern is the adoption of recent theoretical improvements to the t-SNE
-% algorithm. Some implementations such as \pkg{BHTSNE} and \pkg{MulticoreTSNE}
-% appear to have been discontinued, as there has been little to no activity on
-% their respective GitHub repositories in recent years. \pkg{scikit-learn} has a
-% very active community, however, adoption of recent developments such as improved
-% initialization schemes, better default parameter settings, and more flexible
-% affinity kernels has been slow. Among the existing implementations,
-% \pkg{FIt-SNE} includes most of the latest features but lacks crucial ones e.g.
-% the ability to add new data points into existing embeddings, or flexible
-% optimization strategies.
-
-% Finally, with the exception of \pkg{scikit-learn}, the existing packages offer
-% little to no user documentation, usage examples, or API reference. Additionally,
-% these packages do not follow standard software development best practices, such
-% as utilizing continuous integration checks or following standard style guides,
-% making their maintenance difficult and unreliable.
-
-%% -- Illustrations ------------------------------------------------------------
-
-%% - Virtually all JSS manuscripts list source code along with the generated
-%%   output. The style files provide dedicated environments for this.
-%% - In R, the environments {Sinput} and {Soutput} - as produced by Sweave() or
-%%   or knitr using the render_sweave() hook - are used (without the need to
-%%   load Sweave.sty).
-%% - Equivalently, {CodeInput} and {CodeOutput} can be used.
-%% - The code input should use "the usual" command prompt in the respective
-%%   software system.
-%% - For R code, the prompt "R> " should be used with "+  " as the
-%%   continuation prompt.
-%% - Comments within the code chunks should be avoided - these should be made
-%%   within the regular LaTeX text.
-
-%\section{Illustrations} \label{sec:illustrations}
-
-%% -- Summary/conclusions/discussion -------------------------------------------
+The code snippet considers a \pkg{numpy} array or \pkg{scipy} sparse matrix \texttt{x} and creates three \texttt{TSNEEmbedding} objects.
+These represent fully-optimized t-SNE embeddings, and \opentsne\ internally stores them as \pkg{numpy} arrays.
+These can be manipulated and visualized using standard Python tools.
+We then select one of these embeddings and use it to embed new data points from \texttt{x\_new} into the embedding space while keeping the reference embedding fixed.
+Additionally, \opentsne\ provides a more advanced interface that enables the user to change every aspect of the t-SNE optimization procedure.
+We describe this interface in detail in the \opentsne's on-line documentation.
 
@@
 
-Dimensionality reduction techniques implicitly assume that high-dimensional data
-lies on a lower-dimensional manifold, which can accurately be captured by a
-small number of dimensions. However, there is no evidence that every data set
-can accurately be described using only two dimensions, and any such embedding
-will inevitably lead to a loss of information. Thus, it is beneficial to examine
-multiple embeddings, each of which provides a different perspective on topology
-and other data characteristics.
+Dimensionality reduction techniques implicitly assume that high-dimensional data lies on a lower-dimensional manifold, which a small number of dimensions can accurately capture.
+However, there is no evidence that every data set can accurately be described using only two dimensions, and any such embedding will inevitably lead to a loss of information.
+Thus, it is beneficial to examine multiple embeddings, each of which provides a different perspective on topology and other data characteristics.
 
@@
   meaningful layout of the clusters. We can also combine the two
-  perplexities by using a multiscale kernel affinity model \textbf{(c)}
+  perplexities by using a multi-scale kernel affinity model \textbf{(c)}
   and obtain a trade-off between global and local structure.
@@
 
-We illustrate this point by generating four different embeddings of the data on
-single-cell gene expression in mouse brain~\citep{tasic2018shared}.
-Fig.~\ref{fig:tasic}.a shows an embedding using default t-SNE parameters. While
-different clusters of excitatory and inhibitory neurons appear close to one
-another, all clusters appear equidistant from their neighbors, and the overall
-relations between groups are not obvious. The embedding in Fig.~\ref{fig:tasic}.b
-focuses on preserving larger neighborhoods of points, resulting in a more
-globally consistent layout where relations between clusters become more
-apparent. Here, it is evident from the increased white space between groups that
-there is one large class of excitatory neurons and two related classes of
-inhibitory neurons. Unfortunately, focusing on preserving large neighborhoods
-leads to the absorption of smaller clusters into larger ones. Alternatively,
-Fig.~\ref{fig:tasic}.c uses multi-scale similarity kernels that aim to preserve
-both the global organization of clusters and prevent smaller cluster absorption. We
-constructed the embedding from Fig.~\ref{fig:tasic}.d with the settings used
-for Fig.~\ref{fig:tasic}.a, but at a finer level of resolution.
-The figure demonstrates that some clusters are composed of numerous, smaller
-subgroups representing different cell subpopulations that are not visible under
-standard parameter settings.
+We illustrate this point by generating four different embeddings of the data on single-cell gene expression in mouse brain~\citep{tasic2018shared}.
+Fig.~\ref{fig:tasic}.a shows an embedding using default t-SNE parameters.
+While different clusters of excitatory and inhibitory neurons appear close to one another, all clusters appear equidistant from their neighbors, and the overall relations between groups are not obvious.
+The embedding in Fig.~\ref{fig:tasic}.b focuses on preserving larger neighborhoods of points, resulting in a more globally consistent layout where relations between clusters become more apparent.
+Here, it is evident from the increased white space between groups that there is one large class of excitatory neurons and two related classes of inhibitory neurons.
+Unfortunately, focusing on preserving large neighborhoods leads to the absorption of smaller clusters into larger ones.
+Alternatively, Fig.~\ref{fig:tasic}.c uses multi-scale similarity kernels that aim to preserve both the global organization of clusters and prevent smaller cluster absorption.
+We constructed the embedding from Fig.~\ref{fig:tasic}.d with the settings used for Fig.~\ref{fig:tasic}.a, but at a finer level of resolution.
+The figure demonstrates that some clusters are composed of numerous, smaller subgroups representing different cell subpopulations that are not visible under standard parameter settings.
 
-When dealing with data containing millions of data points, standard t-SNE
-embeddings often become unwieldy -- cluster boundaries are blurred, large
-clusters absorb smaller ones, and relationships between clusters become
-increasingly difficult to interpret. We constructed Fig.~\ref{fig:cao}.a from
-the data containing expression profiles of over two million single cells
-captured at different time points in mouse development. The embedding indicates
-numerous clusters with transitions between time points, as marked by the
-color-coding, that are difficult to interpret. Kobak \& Berens observed that
-increasing attractive forces between similar data points controlled via the
-\textit{exaggeration} parameter leads to more compact clusters, and
-subsequently, more informative visualizations~\citep{kobak2019art}. For instance,
-\ref{fig:cao}.b doubles the default exaggeration, which uncovers some of the
-data's overall structure. Further doubling the exaggeration in \ref{fig:cao}.c
-allows us to observe that the data is comprised of two main groups of cells and
-eight somewhat smaller clusters. The visualization also reveals several tiny
-clusters, possibly corresponding to rare cell types.
+When dealing with data containing millions of data points, standard t-SNE embeddings often become unwieldy -- cluster boundaries are blurred, large clusters absorb smaller ones, and relationships between clusters become increasingly difficult to interpret.
+We constructed Fig.~\ref{fig:cao}.a from the data containing expression profiles of over two million single cells captured at different time points in mouse development.
+The embedding indicates numerous clusters with transitions between time points, as marked by the color-coding, that are difficult to interpret.
+Kobak \& Berens observed that increasing attractive forces between similar data points controlled via the \textit{exaggeration} parameter leads to more compact clusters, and subsequently, more informative visualizations~\citep{kobak2019art}.
+For instance, \ref{fig:cao}.b doubles the default exaggeration, which uncovers some of the data's overall structure.
+Further doubling the exaggeration in \ref{fig:cao}.c allows us to observe that the data is comprised of two main groups of cells and eight somewhat smaller clusters.
+The visualization also reveals several tiny clusters, possibly corresponding to rare cell types.
 
@@
 \section{Conclusion}
-We introduce \opentsne, the most feature-complete, open-source, Python
-implementation of the widely popular t-SNE visualization algorithm. The library
-incorporates the latest developments to the t-SNE algorithm and makes them
-easily accessible through a familiar API. \opentsne\ implements efficient
-approximation schemes, making it suitable for the visualization of large data
-sets, containing up to millions of data points. Finally, \opentsne\ is easy to
-install and comes with precompiled binaries available for all major platforms.
+We introduce \opentsne, the most feature-complete, open-source, Python implementation of the widely popular t-SNE visualization algorithm.
+The library incorporates the latest developments to the t-SNE algorithm and makes them easily accessible through a familiar API.
+\opentsne\ implements efficient approximation schemes, making it suitable for the visualization of large data sets containing up to millions of data points.
+Finally, \opentsne\ is easy to install and comes with precompiled binaries available for all major platforms.
 

### 2022-03-09 paper/main.tex
Update citation to our ML manuscript

@@
 \citet{kobak2019art} proposed several techniques to improve global cluster coherence, including estimating similarities with a mixture of Gaussian kernels.
-In our previous work, we introduced a principled approach for embedding new samples into existing visualizations~\citep{policar2019embedding}.
+In our previous work, we introduced a principled approach for embedding new samples into existing visualizations~\citep{policar2021embedding}.
 
@@
 points need to be found through the use of optimization
-techniques~\citep{policar2019embedding}. When adding new data points to an
+techniques~\citep{policar2021embedding}. When adding new data points to an
 existing, reference embedding, the reference data points are fixed in place
@@
 new data points need to be found through
-optimization~\citep{policar2019embedding}. \opentsne\ is currently the only
+optimization~\citep{policar2021embedding}. \opentsne\ is currently the only
 publicly available library allowing users to add new samples to existing

### 2022-03-09 paper/main.tex
Update package versions

@@
 Fig.~\ref{fig:benchmarks} benchmarks four popular Python t-SNE implementations,
-including those from \pkg{scikit-learn} (v0.23.1), \pkg{MulticoreTSNE} (v0.1),
-\pkg{FIt-SNE} (v1.1.0), and our \opentsne\ (v0.6.0). We perform benchmarks on
+including those from \pkg{scikit-learn} (v1.0.1), \pkg{MulticoreTSNE} (v0.1),
+\pkg{FIt-SNE} (v1.1.0), and our \opentsne\ (v0.6.1). We perform benchmarks on
 two computational platforms, one representing the usage on a personal computer
@@
   \caption{\label{fig:benchmarks}
-  We benchmark \opentsne\ (v0.6.0) against three popular open-source
+  We benchmark \opentsne\ (v0.6.1) against three popular open-source
   implementations from \pkg{scikit-learn}~\citep{pedregosa2011scikit}
-  (v0.23.1), \pkg{MulticoreTSNE}~\citep{Ulyanov2016} (v0.1), and
+  (v1.0.1), \pkg{MulticoreTSNE}~\citep{Ulyanov2016} (v0.1), and
   \pkg{FIt-SNE}~\citep{linderman2019fast} (v1.1.0). Experiments were run on a
@@
 \caption{\label{tab:features}
-  We compare the features of \opentsne~(v0.6.0) to four popular open-source
-  t-SNE implementations from \pkg{scikit-learn}~(v0.23.1),
+  We compare the features of \opentsne~(v0.6.1) to four popular open-source
+  t-SNE implementations from \pkg{scikit-learn}~(v1.0.1),
   \pkg{BHTSNE}~(master), \pkg{MulticoreTSNE}~(v0.1), and
@@
 
-The results in this paper were obtained using \proglang{Python}~3.7.6 with the
-\pkg{openTSNE}~v0.6.0 package. \opentsne\ has three direct dependencies:
+The results in this paper were obtained using \proglang{Python}~3.9.7 with the
+\pkg{openTSNE}~v0.6.1 package. \opentsne\ has three direct dependencies:
 \pkg{numpy}, \pkg{scipy}, and \pkg{scikit-learn}. In our experiments, we used
-\pkg{numpy}~v1.18.1, \pkg{scipy}~v1.4.1. \pkg{scikit-learn}~v0.23.1. As these
+\pkg{numpy}~v1.21.3, \pkg{scipy}~v1.7.1. \pkg{scikit-learn}~v1.0.1. As these
 packages include their own downstream dependencies, the full environment used to

### 2022-03-14 benchmarks/run_python.sh
Benchmark: Fix numpy version after MCTSNE install

@@ -10,6 +10,9 @@ conda install -y python numpy scikit-learn
 pip install opentsne
 pip install -r requirements-benchmarks.txt
 
+# MulticoreTSNE forces lower version of numpy
+pip install --upgrade numpy
+
 # Setup FIt-SNE
 git clone git@github.com:KlugerLab/FIt-SNE.git || :
 conda install -c conda-forge fftw -y

### 2022-03-14 benchmarks/run_julia.sh
Benchmark: Fixup run scripts

@@ -22,7 +22,7 @@ wget -nc -P data http://file.biolab.si/opentsne/10x_mouse_zheng.pkl.gz
 mkdir -p logs
 
 # RUN BENCHMARKS
-SAMPLE_SIZES=(1000 5000 10000);
+SAMPLE_SIZES=(1000 5000 10000 20000 50000);
 REPETITIONS=6;
 
 # Single-threaded benchmarks

### 2022-03-14 benchmarks/run_python.sh

@@ -36,8 +36,7 @@ SAMPLE_SIZES=(1000 100000 250000 500000 750000 1000000);
 REPETITIONS=6;
 
 # Single-threaded benchmarks
-#METHODS=(openTSNEBH openTSNEFFT MulticoreTSNE FItSNE sklearn UMAP);
-METHODS=(MulticoreTSNE FItSNE UMAP);
+METHODS=(openTSNEBH openTSNEFFT MulticoreTSNE FItSNE sklearn UMAP);
 
 for method in ${METHODS[@]}; do
     for size in ${SAMPLE_SIZES[@]}; do
@@ -52,10 +51,7 @@ for method in ${METHODS[@]}; do
 done;
 
 # Multi-threaded benchmarks
-#METHODS=(openTSNEBH_8core openTSNEFFT_8core MulticoreTSNE_8core FItSNE_8core sklearn_8core UMAP_8core);
-#METHODS=(MulticoreTSNE_8core FItSNE_8core sklearn_8core);
-METHODS=(MulticoreTSNE_8core FItSNE_8core sklearn_8core UMAP_8core);
-
+METHODS=(openTSNEBH_8core openTSNEFFT_8core MulticoreTSNE_8core FItSNE_8core sklearn_8core UMAP_8core);
 for method in ${METHODS[@]}; do
     for size in ${SAMPLE_SIZES[@]}; do
         cmd="OMP_NUM_THREADS=8 NUMBA_NUM_THREADS=8 \

### 2022-03-14 benchmarks/benchmark.r
Benchmark: Force R to use seconds

@@ -40,5 +40,5 @@ for (i in 1:opt$repetitions) {
     start_time = Sys.time()
     embedding = Rtsne(sample_data, perplexity=30, theta=0.5, pca=FALSE, verbose=TRUE,
                       eta=200, num_threads=opt$`n-threads`)
-    cat("Rtsne benchmark time:", Sys.time() - start_time, "\n")
+    cat("Rtsne benchmark time:", difftime(Sys.time(), start_time, units="secs"), "\n")
 }

### 2022-03-14 benchmarks/run_r.sh

@@ -7,7 +7,7 @@ eval "$(conda shell.bash hook)"  # activate conda env inside script
 conda activate tsne_benchmarks_r
 
 # Setup R environment
-conda install -c r r -y
+mamba install -c r r -y
 Rscript -e 'install.packages("Rtsne", repos="https://cloud.r-project.org")'
 Rscript -e 'install.packages("optparse", repos="https://cloud.r-project.org")'
 
@@ -15,20 +15,22 @@ Rscript -e 'install.packages("optparse", repos="https://cloud.r-project.org")'
 mkdir -p data
 wget -nc -P data http://file.biolab.si/opentsne/10x_mouse_zheng.pkl.gz
 
-conda install -y python numpy
+mamba install -y python numpy
 python convert_pickle_to_csv.py
 
 # Prepare logs directory
 mkdir -p logs
 
+conda list > 00--run_r--conda_list.txt
+
 # RUN BENCHMARKS
-SAMPLE_SIZES=(1000 100000 250000 500000 750000 1000000);
 REPETITIONS=6;
 
 # Single-threaded benchmarks
+SAMPLE_SIZES=(1000 100000 250000 500000 750000);
 for size in ${SAMPLE_SIZES[@]}; do
     cmd="OMP_NUM_THREADS=1 \
-        Rscript benchmark.r \
+       Rscript benchmark.r \
         --repetitions $REPETITIONS \
         --n-threads 1 \
         --n-samples $size 2>&1 \
@@ -38,6 +40,7 @@ for size in ${SAMPLE_SIZES[@]}; do
 done;
 
 # Multi-threaded benchmarks
+SAMPLE_SIZES=(1000 100000 250000 500000 750000 1000000);
 for size in ${SAMPLE_SIZES[@]}; do
     cmd="OMP_NUM_THREADS=8 \
         Rscript benchmark.r \

### 2022-03-15 benchmarks/benchmark.r
Benchmarks: Keep better track of package versions

@@ -3,6 +3,8 @@ library(optparse)
 library(Rtsne)
 
 
+sessionInfo()
+
 option_list = list(
     make_option(c("--repetitions"), type="integer", default=1),
     make_option(c("--n-samples"), type="integer", default=FALSE),

### 2022-03-15 benchmarks/run_julia.sh

@@ -21,6 +21,8 @@ wget -nc -P data http://file.biolab.si/opentsne/10x_mouse_zheng.pkl.gz
 # Prepare logs directory
 mkdir -p logs
 
+conda env export > logs/00--run_julia--conda_env.txt
+
 # RUN BENCHMARKS
 SAMPLE_SIZES=(1000 5000 10000 20000 50000);
 REPETITIONS=6;

### 2022-03-15 benchmarks/run_python.sh

@@ -2,10 +2,10 @@
 
 set -e
 
-conda create --name tsne_benchmarks -y
+mamba create --name tsne_benchmarks -y
 eval "$(conda shell.bash hook)"  # activate conda env inside script
 conda activate tsne_benchmarks
-conda install -y python numpy scikit-learn
+mamba install -y python numpy scikit-learn
 
 pip install opentsne
 pip install -r requirements-benchmarks.txt
@@ -15,7 +15,7 @@ pip install --upgrade numpy
 
 # Setup FIt-SNE
 git clone git@github.com:KlugerLab/FIt-SNE.git || :
-conda install -c conda-forge fftw -y
+mamba install -c conda-forge fftw -y
 cd FIt-SNE
 CPLUS_INCLUDE_PATH="${CONDA_PREFIX}/include/" LIBRARY_PATH="${CONDA_PREFIX}/lib/" \
     g++ -std=c++11 -O3  src/sptree.cpp src/tsne.cpp src/nbodyfft.cpp  \
@@ -30,6 +30,8 @@ wget -nc -P data http://file.biolab.si/opentsne/10x_mouse_zheng.pkl.gz
 
 # Prepare logs directory
 mkdir -p logs
+
+conda env export > logs/00--run_python--conda_env.txt
 
 # RUN BENCHMARKS
 SAMPLE_SIZES=(1000 100000 250000 500000 750000 1000000);

### 2022-03-15 benchmarks/run_r.sh

@@ -21,7 +21,7 @@ python convert_pickle_to_csv.py
 # Prepare logs directory
 mkdir -p logs
 
-conda list > 00--run_r--conda_list.txt
+conda env export > logs/00--run_r--conda_env.txt
 
 # RUN BENCHMARKS
 REPETITIONS=6;

### 2022-03-15 benchmarks/run_python.sh
Benchmarks: Switch from mamba to conda in run scripts

@@ -2,10 +2,10 @@
 
 set -e
 
-mamba create --name tsne_benchmarks -y
+conda create --name tsne_benchmarks -y
 eval "$(conda shell.bash hook)"  # activate conda env inside script
 conda activate tsne_benchmarks
-mamba install -y python numpy scikit-learn
+conda install -y python numpy scikit-learn
 
 pip install opentsne
 pip install -r requirements-benchmarks.txt
@@ -15,7 +15,7 @@ pip install --upgrade numpy
 
 # Setup FIt-SNE
 git clone git@github.com:KlugerLab/FIt-SNE.git || :
-mamba install -c conda-forge fftw -y
+conda install -c conda-forge fftw -y
 cd FIt-SNE
 CPLUS_INCLUDE_PATH="${CONDA_PREFIX}/include/" LIBRARY_PATH="${CONDA_PREFIX}/lib/" \
     g++ -std=c++11 -O3  src/sptree.cpp src/tsne.cpp src/nbodyfft.cpp  \

### 2022-03-15 benchmarks/run_r.sh

@@ -7,7 +7,7 @@ eval "$(conda shell.bash hook)"  # activate conda env inside script
 conda activate tsne_benchmarks_r
 
 # Setup R environment
-mamba install -c r r -y
+conda install -c r r -y
 Rscript -e 'install.packages("Rtsne", repos="https://cloud.r-project.org")'
 Rscript -e 'install.packages("optparse", repos="https://cloud.r-project.org")'
 
@@ -15,7 +15,7 @@ Rscript -e 'install.packages("optparse", repos="https://cloud.r-project.org")'
 mkdir -p data
 wget -nc -P data http://file.biolab.si/opentsne/10x_mouse_zheng.pkl.gz
 
-mamba install -y python numpy
+conda install -y python numpy
 python convert_pickle_to_csv.py
 
 # Prepare logs directory
@@ -30,7 +30,7 @@ REPETITIONS=6;
 SAMPLE_SIZES=(1000 100000 250000 500000 750000);
 for size in ${SAMPLE_SIZES[@]}; do
     cmd="OMP_NUM_THREADS=1 \
-       Rscript benchmark.r \
+        Rscript benchmark.r \
         --repetitions $REPETITIONS \
         --n-threads 1 \
         --n-samples $size 2>&1 \

### 2022-03-15 paper/main.tex
Update abstract
Update benchmark figures and add generating script

@@
 The library is fast and can treat data sets containing millions of data points.
-We distribute \opentsne\ under the BSD-3-Clause License.
+\opentsne\ is open-source and is distributed under the BSD-3-Clause License.
 Its source code is available at \url{https://github.com/pavlin-policar/openTSNE}.
@@
 
-We introduce \opentsne, a comprehensive Python library that implements t-SNE and its numerous recently proposed extensions, including the implementation of efficient approximation schemes allowing the embedding of millions of data points~\citep{van2014accelerating,linderman2019fast}, the addition of new data points into existing embeddings~\citep{policar2019embedding}, various initialization improvements~\citep{kobak2019umap}, extensions~\citep{kobak2019heavy,kobak2019art}, and improved parameter defaults~\citep{belkina2019automated}.
+\subsection{Structure of openTSNE}
+
+\opentsne\ is designed to accomodate all users, including both a simple API that allows users to quickly obtain standard t-SNE visualization, as well as an advanced API that allows practitioners and researchers to modify different aspects of the t-SNE algorithm.
+
+The t-SNE algorithm is dependent on three key steps: the initial positions of points in the embedding, an affinity model describing the similarities between data points in the ambient space, and the optimization which optimizes the embedding layout.
+\opentsne\ mirrors these different components in its design.
+\opentsne\ contains three corresponding core modules -- \code{initialization}, \code{affinity}, and \code{tsne} -- each of which will be decribed in detail in later sections.
+These modules are designed to be easily extensible, allowing researchers to experiment with different configurations and produce the best possible t-SNE visualizations.
+For less demanding users, \opentsne\ also provides a familiar, scikit-learn-like interface through the \code{openTSNE.TSNE} class.
+
+\subsubsection{Initialization}
+
+just a numpy array
+
+\subsubsection{Affinities}
+
+\paragraph{Affinity}
+
+\paragraph{KNNIndex}
+
+\subsubsection{Embedding objects}
+
+\paragraph{TSNEEmbedding}
+
+\paragraph{PartialTSNEEmbedding}
+
+\subsection{Old text}
+
+We introduce \opentsne, a comprehensive Python library that implements t-SNE and its numerous recently proposed extensions, including the implementation of efficient approximation schemes allowing the embedding of millions of data points~\citep{van2014accelerating,linderman2019fast}, the addition of new data points into existing embeddings~\citep{policar2021embedding}, various initialization improvements~\citep{kobak2019umap}, extensions~\citep{kobak2019heavy,kobak2019art}, and improved parameter defaults~\citep{belkina2019automated}.
 When combined, these improvements result in the faster generation of more informative visualizations.
@@
 
+x: np.ndarray = ...
+
 # Vanilla t-SNE
@@
 
-# Use multiscale kernel for better local/global structure trade-off
-from openTSNE.affinity import Multiscale
-
-affinities = Multiscale(x, perplexities=[50, 500])
-tsne_multiscale = TSNE().fit(x, affinities=affinites)
+# Use multiple perplexity values for better local/global structure trade-off
+tsne_multiscale = TSNE(perplexity=[50, 500]).fit(x)
 
@@
 Fig.~\ref{fig:tasic}.a shows an embedding using default t-SNE parameters.
+We generate the four different embeddings using the following code snippet (which is shortened for clarity).
+\begin{CodeChunk}
+\begin{CodeInput}
+embedding_a = TSNE(perplexity=30).fit(x)
+embedding_b = TSNE(perplexity=500).fit(x)
+embedding_c = TSNE(perplexity=[30, 500]).fit(x)
+embedding_d = TSNE(perplexity=30, dof=0.6).fit(x)
+\end{CodeInput}
+\end{CodeChunk}
 While different clusters of excitatory and inhibitory neurons appear close to one another, all clusters appear equidistant from their neighbors, and the overall relations between groups are not obvious.
@@
 The visualization also reveals several tiny clusters, possibly corresponding to rare cell types.
+\begin{CodeChunk}
+\begin{CodeInput}
+# To improve the global structure of the final layout, we initialize the
+# embedding by creating a t-SNE plot on a sample of points.
+
+# Take a random sample of the data points
+indices = np.random.permutation(list(range(x.shape[0])))
+reverse = np.argsort(indices)
+
+x_sample, x_rest = x[indices[:25000]], x[indices[25000:]]
+
+# Create a t-SNE embedding on the sample, using a high perplexity value to
+# emphasize the high level structure
+init_sample = TSNE(perplexity=500).fit(x_sample)
+
+# We can next use the `transform` functionality to determine starting points for
+# the remaining data points. Since we don't actually want to optimize the
+# remaining points separately from the sample points, we don't need to call the
+# `transform` function. Instead, we'll call the `prepare_partial` function,
+# which `transform` calls anyways under the hood.
+init_rest = init_sample.prepare_partial(x_rest)
+
+# Lastly, we merge `init_sample` and `init_rest`
+init_full = np.vstack((init_sample, init_rest))[reverse]
+\end{CodeInput}
+\end{CodeChunk}
+We generate the four different embeddings using the following code snippet (which is shortened for clarity).
+\begin{CodeChunk}
+\begin{CodeInput}
+# We separately precompute the affinities as this can take quite a long time.
+# This way, we can reuse the precomputed affinities in various different
+# embeddings, without having to repeat the same, costly computation every time.
+affinities = openTSNE.affinity.PerplexityBasedNN(x, perplexity=30)
+
+# Create the TSNEEmbedding object using our previously calculated initialization
+# and precomputed affinities
+embedding = openTSNE.TSNEEmbedding(init_full, affinities)
+
+# Run the standard t-SNE early exaggeration phase
+embedding_ee =    embedding.optimize(n_iter=500, exaggeration=12, momentum=0.5)
+embedding_c  = embedding_ee.optimize(n_iter=500, exaggeration=4,  momentum=0.8)
+embedding_b  =  embedding_c.optimize(n_iter=500, exaggeration=2,  momentum=0.8)
+embedding_a  =  embedding_b.optimize(n_iter=500, exaggeration=1,  momentum=0.8)
+\end{CodeInput}
+\end{CodeChunk}
 
@@
 experiment on bipolar retinal cells by \citet{shekhar2016comprehensive}, we can correctly map the bipolar cell
-clusters onto the reference embedding.
+clusters onto the reference embedding. The following snippet shows the code necessary to generate figure~\ref{fig:macosko}.
+\begin{CodeChunk}
+\begin{CodeInput}
+# We specify the old default parameter values which are often used to show
+# why t-SNE misbehaves or fails to preserve global structure
+tsne_old = TSNE(initialization="random", learning_rate=200, n_iter=750).fit(x)
+
+# By default, openTSNE uses PCA-based initialization and automatically
+# determines the best learning rate. This also allows us to lower the default
+# number of iterations to 500, lowering the overall runtime
+tsne_standard = TSNE(perplexity=[50, 500]).fit(x)
+
+# Use the generated embeddings to add new data points
+new_embedding = tsne_standard.transform(x_new)
+\end{CodeInput}
+\end{CodeChunk}
 
@@
 
+% General problem statement
 One of the t-SNE's common criticisms is limited scalability to large data sets
@@
 points~\citep{becht2019dimensionality}. The culprit for slow response time stems
-from a particular optimization procedure and its specific implementation in
-popular Python libraries. Until quite recently, most popular implementations of
-t-SNE were based on the Barnes-Hut approximation scheme developed by
+from particular optimization procedures and their specific implementation in
+popular open-source libraries. Until quite recently, most popular implementations of
+t-SNE were based on either a direct implementation of the t-SNE algorithm with asymptotic time complexity $\mathcal{O}(N^2)$, or the Barnes-Hut approximation scheme developed by
 \citet{van2014accelerating} with asymptotic time complexity $\mathcal{O}(N \log
-N)$, where $N$ is the number of data items ({\em e.g.} cells). The most
+N)$, where $N$ is the number of data items ({\em e.g.} cells).
+
+% Problem with Python
+In Python, the most
 widely-used implementation of t-SNE came from \pkg{scikit-learn}, which exhibits
@@
 construct t-SNE embeddings of millions of data points in a matter of hours on
-widely-accessible, consumer-grade processors (Fig.~\ref{fig:benchmarks}).
+widely-accessible, consumer-grade processors (Fig.~\ref{fig:benchmarks_py}).
 However, recently, \citet{linderman2019fast} developed a new approximation
-scheme -- FIt-SNE -- which further reduces the asymptotic time complexity to
-$\mathcal{O}(N)$. We include this approach in \opentsne, enabling the embedding
-of large data sets in a matter of minutes.
+scheme -- FIt-SNE -- which reduces the asymptotic time complexity to
+$\mathcal{O}(N)$. This approximation scheme is included in \pkg{FIt-SNE} and \opentsne, enabling the embedding of large data sets in a matter of minutes.
 
-Fig.~\ref{fig:benchmarks} benchmarks four popular Python t-SNE implementations,
+% We benchmark Python
+Fig.~\ref{fig:benchmarks_py} benchmarks four popular Python t-SNE implementations,
 including those from \pkg{scikit-learn} (v1.0.1), \pkg{MulticoreTSNE} (v0.1),
 \pkg{FIt-SNE} (v1.1.0), and our \opentsne\ (v0.6.1). We perform benchmarks on
-two computational platforms, one representing the usage on a personal computer
-and the other the utility of these libraries on a high-performance computing
-platform. The Intel(R) Core i7 is commonly found in consumer-grade laptop
-computers, while Intel(R) Xeon(R) processors appear in high-performance
-computing machines. Benchmarks were run for $1,000$ iterations with the original
+a server-grade Intel(R) Xeon(R) CPU E5-1650 equipped with 128 GB of memory. Benchmarks were run for $1,000$ iterations with the original
 t-SNE parameters, as some implementations do not allow their modification.
+To generate benchmark data sets of different sizes, we subsampled data from the 10X Genomics 1.3 million mouse brain data set five times, resulting in five different data sets for each size.
+In total, we run each implementation on 30 different data sets.
 
 \begin{figure*}[ht]
-  \includegraphics[width=\textwidth]{benchmarks}
-  \caption{\label{fig:benchmarks}
+  \includegraphics[width=\textwidth]{benchmarks_python}
+  \caption{\label{fig:benchmarks_py}
   We benchmark \opentsne\ (v0.6.1) against three popular open-source
@@
   \pkg{FIt-SNE}~\citep{linderman2019fast} (v1.1.0). Experiments were run on a
-  consumer-grade Intel Core i7-7700HQ processor found in laptop computers, and
-  on a server-grade Intel Xeon E5-2650. To generate benchmark data sets of
-  different sizes, we subsampled data from the 10X Genomics 1.3 million mouse
-  brain data set five times, resulting in five different data sets for each
-  size. In total, we run each implementation on 30 different data sets. Notice
+  on a server-grade Intel(R) Xeon(R) CPU E5-1650 equipped with 128 GB of memory.Notice
   that \opentsne\ scales similarly to \pkg{FIt-SNE}, as they both use the
@@
 
-Benchmark results (Fig.~\ref{fig:benchmarks}) confirm that both \pkg{FIt-SNE}
+% Problem with other programming languages
+In R, the only widely used implementation of t-SNE -- \pkg{Rtsne}~\citep{krijthe2015rtsne} -- is based on the Barnes-Hut approximation, and is closely comparable to \pkg{MulticoreTSNE}.
+In Julia, the only available implementation -- \pkg{TSne.jl} -- is based on a direct $\mathcal{O}(N^2)$ implementation of the algorithm, which allows it to be used only on smaller data sets containing thousands of data points.
+Matlab is proprietary software, but also implements a naive version of the t-SNE algorithm, therefore we would expect its performance to mirror that of \pkg{TSne.jl} from Julia.
+
+% We benchmark other programming languages
+Using the same benchmarking setup as before, we ...
+The results are shown in Fig.~\ref{fig:benchmarks_lang}.
+
+\begin{figure*}[ht]
+  \includegraphics[width=\textwidth]{benchmarks_langs}
+  \caption{\label{fig:benchmarks_lang}
+  We choose the fastest available, open-source implementations of the t-SNE algorithm from the Python, R, and Julia programming languages.
+We choose \opentsne\ for Python, \pkg{Rtsne} for R, and \pkg{TSne.jl} for Julia.
+The benchmarks reflect the time complexity of the implemented algorithms.
+For instance, \pkg{TSne.jl} implements a naive $\mathcal{O}(N^2)$ algorithm, making it prohibitively expensive for all but the smallest data sets.
+\pkg{Rtsne} scales similarily to \pkg{MulticoreTSNE} in Python, as it implements the $\mathcal{O}(N \log N)$ Barnes-Hut approximation scheme.
+Interestingly, the multi-threaded implementation version of \pkg{Rtsne} crashes on benchmark data sets containing over 500,000 data points due to lack of memory.
+This would seriously hamper its utility on consumer-grade platforms, as our benchmark platform contained 128 GB of memory.
+\opentsne\ is the only library implementing the more recent FIt-SNE approximation scheme, which makes it suitable for even the largest data sets.
+}
+\end{figure*}
+
+Benchmark results (Fig.~\ref{fig:benchmarks_py}) confirm that both \pkg{FIt-SNE}
 and \opentsne\ scale better than their Barnes-Hut counterparts --
@@
 quickly experiment with different parameter settings and easily incorporate
-custom components into the software. We provide a full feature list and
+custom components into the software. We provide an extended list of features and
 comparison to other popular t-SNE implementations in Table~\ref{tab:features}.
@@
 \begin{center}\small
-\newcommand*\rot{\rotatebox{45}}
+\newcommand*\rot{\rotatebox{90}}
 \renewcommand{\arraystretch}{1.25}
 
-\begin{tabular}{l c c c c c c c}
+\begin{tabular}{l c c c c c|c c}
 \toprule
 \setlength\tabcolsep{6pt}
-& \rot{\pkg{scikit-learn}} & \rot{\pkg{BHTSNE}} & \rot{\pkg{MulticoreTSNE}} & \rot{\pkg{FIt-SNE}} & \rot{\pkg{openTSNE}} \\
+& \rot{\pkg{scikit-learn}} & \rot{\pkg{BHTSNE}} & \rot{\pkg{MulticoreTSNE}} & \rot{\pkg{FIt-SNE}} & \rot{\pkg{openTSNE}} & \rot{\pkg{Rtsne}} & \rot{\pkg{Tsne.jl}} \\
 
 \toprule
-\textbf{Ease of Installation}\\
-Native Python installation & \checkmark & & & & \checkmark \\
-\pkg{PyPI} package & \checkmark & & \checkmark & & \checkmark \\
-\pkg{conda} package & \checkmark & & & & \checkmark \\
+\textbf{Ease of Installation} \\
+Native Python installation & \checkmark & & & & \checkmark & & \\
+\pkg{PyPI} package & \checkmark & & \checkmark & & \checkmark & & \\
+\pkg{conda} package & \checkmark & & & & \checkmark & & \\
 
 \hline
-\textbf{Approximation Schemes}\\
-Barnes-Hut ($\mathcal{O}(N \log N)$) & \checkmark & \checkmark & \checkmark & & \checkmark \\
-FIt-SNE ($\mathcal{O}(N)$) & & & & \checkmark & \checkmark \\
+\textbf{Approximation Schemes} \\
+None ($\mathcal{O}(N^2)$) & & & & & & & \checkmark \\
+Barnes-Hut ($\mathcal{O}(N \log N)$) & \checkmark & \checkmark & \checkmark & & \checkmark & \checkmark & \\
+FIt-SNE ($\mathcal{O}(N)$) & & & & \checkmark & \checkmark & & \\
 
 \hline
-\textbf{Advanced Features and Extensions}\\
-Multiscale Gaussian kernels & & & & \checkmark & \checkmark \\
-Fully-custom affinity kernels & & & & & \checkmark \\
-Variable degrees of freedom & & & & \checkmark & \checkmark \\
-Variable exaggeration & & & & \checkmark & \checkmark \\
-Globally-consistent initialization & & & & \checkmark & \checkmark \\
-Automatic learning rate & & & & \checkmark & \checkmark \\
-Adding samples to embeddings & & & & & \checkmark \\
-Callback system & & & & & \checkmark \\
-Interactive optimization  & & & & & \checkmark \\
+\textbf{Advanced Features and Extensions} \\
+%Multiscale Gaussian kernels & & & & \checkmark & \checkmark & & \\
+Extensible affinity models & & & & & \checkmark & & \\
+Variable degrees of freedom & & & & \checkmark & \checkmark & & \\
+Variable exaggeration & & & & \checkmark & \checkmark & & \\
+Globally-consistent initialization & & & & \checkmark & \checkmark & & \\
+Automatic learning rate & & & & \checkmark & \checkmark & & \\
+Adding samples to embeddings & & & & & \checkmark & & \\
+Callback system & & & & & \checkmark & & \\
+Interactive optimization  & & & & & \checkmark & & \\
 
@@
 \textbf{Project Quality}\\
-User documentation & \checkmark & & & & \checkmark \\
-End-to-end usage examples & \checkmark & & & \checkmark & \checkmark \\
-API reference & \checkmark & & & & \checkmark \\
-Continuous integration & \checkmark & & \checkmark & & \checkmark \\
+Actively maintained & \checkmark & & & & \checkmark & \checkmark & \checkmark \\
+Continuous integration & \checkmark & & \checkmark & & \checkmark & \checkmark & \checkmark \\
+User documentation & \checkmark & & & & \checkmark & \checkmark  & \\
+End-to-end usage examples & \checkmark & & & \checkmark & \checkmark & \checkmark & \checkmark \\
+API reference & \checkmark & & & & \checkmark & \checkmark &  \\
 \bottomrule
@@
 \caption{\label{tab:features}
-  We compare the features of \opentsne~(v0.6.1) to four popular open-source
+  We compare the features of \opentsne~(v0.6.1) to four popular open-source Python
   t-SNE implementations from \pkg{scikit-learn}~(v1.0.1),
   \pkg{BHTSNE}~(master), \pkg{MulticoreTSNE}~(v0.1), and
-  \pkg{FIt-SNE}~(v1.1.0).
+  \pkg{FIt-SNE}~(v1.1.0), as well as the major implementations in R and Julia,
+  \pkg{Rtsne}~(v0.15), \pkg{Tsne.jl}~(v1.3.0). Active maintenance indicates whether the project has had any meaningful development in the past year from the time of writing.
 }
@@
 
-The results in this paper were obtained using \proglang{Python}~3.9.7 with the
+The results in this paper were obtained using \proglang{Python}~3.10.2 with the
 \pkg{openTSNE}~v0.6.1 package. \opentsne\ has three direct dependencies:

### 2022-03-15 benchmarks/generate_figures.py

@@ -1,0 +1,238 @@
+import os
+import re
+from dataclasses import dataclass
+from functools import wraps, cmp_to_key
+from typing import Callable, Dict, List
+
+import numpy as np
+import pandas as pd
+import matplotlib
+import matplotlib.pyplot as plt
+import seaborn as sns
+
+
+BENCHMARK_DIR = "../benchmarks/logs"
+
+
+def make_regex_func(pattern):
+    def _wrapper(fname):
+        matches = [re.findall(pattern, line) for line in open(fname)]
+        matches = list(filter(len, matches))
+        matches = list(map(lambda x: float(x[0]), matches))
+
+        return np.array(matches)[1:]  # skip first -- warmup
+
+    return _wrapper
+
+
+def from_minutes(f):
+    @wraps(f)
+    def _wrapper(*args, **kwargs):
+        return f(*args, **kwargs) * 60
+    return _wrapper
+
+
+@dataclass
+class Spec:
+    name: str
+    log_name: str
+    cores: int
+    lang: str
+    parser: Callable
+
+    @property
+    def display_name(self):
+        return f"{self.name} ({self.cores} {'cores' if self.cores > 1 else 'core'})"
+
+    def __hash__(self):
+        return hash(self.display_name)
+
+    def __eq__(self, other):
+        return self.display_name == other.display_name
+
+    def __gt__(self, other):
+        return self.display_name > other.display_name
+
+
+def get_benchmarks(spec: Spec, directory: str) -> List[Dict]:
+    # Create regex pattern to match file names
+    pattern = spec.log_name
+    if spec.cores > 1:
+        pattern += f"_{spec.cores}core"
+    pattern += r"_(\d+).log"
+
+    data = []
+    for f in os.listdir(directory):
+        if match := re.findall(pattern, f):
+            n_samples = int(match[0])
+
+            times = spec.parser(os.path.join(directory, f))
+            for t in times:
+                data.append({
+                    "Spec": spec,
+                    "Name": spec.display_name,
+                    "Num. Samples": n_samples,
+                    "Time (sec)": t,
+                    "Time (min)": t / 60,
[…]

### 2022-03-19 figures/download_h5ad.sh
Create scripts to generate all necessary data files

@@ -1,0 +1,31 @@
+#!/bin/bash
+
+
+set -e
+
+cat << EOF
+================================================================================
+Downloading datasets from file.biolab.si...
+================================================================================
+EOF
+
+# Download data, if not exists
+H5AD_DIR="./data/h5ad"
+mkdir -p "$H5AD_DIR"
+
+wget -nc -P "$H5AD_DIR" http://file.biolab.si/opentsne/h5ad/cao_2019.h5ad.tar.gz
+wget -nc -P "$H5AD_DIR" http://file.biolab.si/opentsne/h5ad/harris_2018.h5ad.tar.gz
+wget -nc -P "$H5AD_DIR" http://file.biolab.si/opentsne/h5ad/hochgerner_2018.h5ad.tar.gz
+wget -nc -P "$H5AD_DIR" http://file.biolab.si/opentsne/h5ad/macosko_2015.h5ad.tar.gz
+wget -nc -P "$H5AD_DIR" http://file.biolab.si/opentsne/h5ad/shekhar_2016.h5ad.tar.gz
+wget -nc -P "$H5AD_DIR" http://file.biolab.si/opentsne/h5ad/tasic_2018.h5ad.tar.gz
+
+cat << EOF
+================================================================================
+Extracting files...
+================================================================================
+EOF
+
+for f in "$H5AD_DIR/*.h5ad.tar.gz"; do
+    tar xvfa "$f" --remove-files
+done

### 2022-03-19 figures/generate_h5ad.sh

@@ -1,0 +1,79 @@
+#!/bin/bash
+
+# This script downloads the raw files from NCBI and converts them to h5ad files.
+# This can take a very long time, and we recommend downloading the files from
+# file.biolab.si/opentsne/ instead using the `download_h5ad.sh` script.
+
+set -e
+
+RAW_DIR="./data/raw"
+H5AD_DIR="./data/h5ad"
+
+cat << EOF
+================================================================================
+Downloading datasets...
+================================================================================
+EOF
+
+# Download data, if not exists
+mkdir -p "$RAW_DIR"
+
+# Cao 2019
+wget -nc -P "$RAW_DIR/cao_2019" ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE119nnn/GSE119945/suppl/GSE119945_gene_count.txt.gz
+wget -nc -P "$RAW_DIR/cao_2019" ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE119nnn/GSE119945/suppl/GSE119945_cell_annotate.csv.gz
+wget -nc -P "$RAW_DIR/cao_2019" ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE119nnn/GSE119945/suppl/GSE119945_gene_annotate.csv.gz
+
+# Macosko 2015
+wget -nc -P "$RAW_DIR/macosko_2015" https://ftp.ncbi.nlm.nih.gov/geo/series/GSE63nnn/GSE63472/suppl/GSE63472%5FP14Retina%5Fmerged%5Fdigital%5Fexpression%2Etxt%2Egz
+wget -nc -P "$RAW_DIR/macosko_2015" http://mccarrolllab.org/wp-content/uploads/2015/05/retina_clusteridentities.txt
+
+# Shekhar 2016
+wget -nc -P "$RAW_DIR/shekhar_2016" https://ftp.ncbi.nlm.nih.gov/geo/series/GSE81nnn/GSE81904/suppl/GSE81904_BipolarUMICounts_Cell2016.txt.gz
+wget -nc -P "$RAW_DIR/shekhar_2016" https://scrnaseq-public-datasets.s3.amazonaws.com/manual-data/shekhar/clust_retinal_bipolar.txt
+
+# Harris 2018
+wget -nc -P "$RAW_DIR/harris_2018" ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE99nnn/GSE99888/suppl/GSE99888_gene_expression.tab.gz
+
+# Tasic 2018
[…]

### 2022-03-19 figures/generate_h5ad/cao_2019.py

@@ -1,0 +1,60 @@
+import argparse
+import sys
+from os import path, listdir
+
+import numpy as np
+import pandas as pd
+import scipy.sparse as sp
+import anndata
+import scanpy as sc
+import utils
+
+
+parser = argparse.ArgumentParser()
+parser.add_argument("-i", "--in-dir", required=True)
+parser.add_argument("-o", "--out-dir", required=True)
+parser.add_argument("-f", "--force", action="store_true")
+args = parser.parse_args()
+
+fname = path.join(args.out_dir, "cao_2019.h5ad")
+
+if path.exists(fname) and not args.force:
+    print(f"`{fname}` exists. Skipping...")
+    sys.exit(0)
+
+adata = anndata.read_mtx(path.join(args.in_dir, "GSE119945_gene_count.txt.gz"))
+adata = adata.T
+
+cell_annotations = pd.read_csv(path.join(args.in_dir, "GSE119945_cell_annotate.csv.gz"))
+adata.obs = cell_annotations.set_index("sample")
+
+gene_annotations = pd.read_csv(path.join(args.in_dir, "GSE119945_gene_annotate.csv.gz"))
+adata.var = gene_annotations.set_index("gene_short_name")
+
+# Select main cluster
+adata.obs = adata.obs.rename(columns={"Main_Cluster": "cluster_id"})
+
+adata.obs["potential_doublet_cluster"] = adata.obs["potential_doublet_cluster"].astype(str)
+adata.obs["detected_doublet"] = adata.obs["detected_doublet"].astype(str)
+
+# Removing putative doublets
+print("Removing %d putative doublets" % np.sum(adata.obs["detected_doublet"] == "nan"))
+adata = adata[adata.obs["detected_doublet"] != "nan"]
+
+adata.write_h5ad(fname)
+
+print("Running single-cell pipeline...")
+sc.pp.filter_genes(adata, min_counts=10)
+gene_mask = utils.select_genes(adata.X, n=3000, threshold=0)
+
+sc.pp.normalize_per_cell(adata)
+sc.pp.log1p(adata)
+sc.pp.scale(adata, zero_center=False)
+
+X_pca = utils.pca(adata.X[:, gene_mask])
+
+# Reload the original, unprocessesed data set, and add the PCA coordinates
+adata = anndata.read_h5ad(fname)
+adata.obsm["X_pca"] = X_pca
+
+adata.write_h5ad(fname)

### 2022-03-19 figures/generate_h5ad/harris_2018.py

@@ -1,0 +1,36 @@
+import argparse
+import sys
+from os import path, listdir
+
+import pandas as pd
+import scipy.sparse as sp
+import anndata
+
+
+parser = argparse.ArgumentParser()
+parser.add_argument("-i", "--in-dir", required=True)
+parser.add_argument("-o", "--out-dir", required=True)
+parser.add_argument("-f", "--force", action="store_true")
+args = parser.parse_args()
+
+fname = path.join(args.out_dir, "harris_2018.h5ad")
+
+if path.exists(fname) and not args.force:
+    print(f"`{fname}` exists. Skipping...")
+    sys.exit(0)
+
+data = pd.read_table(path.join(args.in_dir, "GSE99888_gene_expression.tab.gz"), index_col=0)
+
+# Remove duplicate cells
+data = data[~data.index.duplicated()]
+
+
+x = sp.csr_matrix(data.values)
+
+adata = anndata.AnnData(
+    x.T,
+    var={"var_names": data.index.values},
+    obs={"obs_names": data.columns.values},
+)
+
+adata.write_h5ad(fname)

### 2022-03-19 figures/generate_h5ad/hochgerner_2018.py

@@ -1,0 +1,92 @@
+import argparse
+import sys
+from os import path, listdir
+
+import pandas as pd
+import scipy.sparse as sp
+import anndata
+import scanpy as sc
+import utils
+
+
+parser = argparse.ArgumentParser()
+parser.add_argument("-i", "--in-dir", required=True)
+parser.add_argument("-o", "--out-dir", required=True)
+parser.add_argument("-f", "--force", action="store_true")
+args = parser.parse_args()
+
+fname = path.join(args.out_dir, "hochgerner_2018.h5ad")
+
+if path.exists(fname) and not args.force:
+    print(f"`{fname}` exists. Skipping...")
+    sys.exit(0)
+
+data = pd.read_table(path.join(args.in_dir, "GSE104323_10X_expression_data_V2.tab.gz"), index_col=0)
+
+metadata = pd.read_table(path.join(args.in_dir, "GSE104323_metadata_barcodes_24185cells.txt.gz"), index_col=0)
+
+# Remove the cells for which we don't have the appropriate metadata
+nan_mask = metadata.isna().all(axis=1)
+metadata = metadata.loc[~nan_mask]
+data = data.loc[:, metadata.index]
+
+
+cell_types = metadata["characteristics: cell cluster"].copy()
+
+cell_types = cell_types.replace({
+    "Immature-Pyr": "immature pyramidal neuron",
+    "GC-juv": "juevnile granule cell",
+    "GC-adult": "granule cell",
+    "Immature-GC": "immature granule cell",
+    "Neuroblast": "neuroblast",
+    "Astro-adult": "astrocyte",
+    "Immature-GABA": "immature GABAergic neuron",
+    "Astro-juv": "juvenile astrocyte",
+    "OPC": "oligodendrocyte precursor cell",
+    "MOL": "oligodendrocyte",
+    "RGL_young": "radial glial cell",
+    "Immature-Astro": "immature astrocyte",
+    "Endothelial": "endothelial cell",
+    "Cajal-Retzius": "Cajal-Retzius cell",
+    "CA3-Pyr": "pyramidal neuron",
+    "nIPC-perin": "neuronal intermediate progenitor cells",
+    "nIPC": "neuronal intermediate progenitor cells",
+    "MiCajal-Retziusoglia": "glial cell",
+    "NFOL": "newly formed oligodendrocyte",
+    "GABA": "GABAergic neuron",
[…]

### 2022-03-19 figures/generate_h5ad/macosko_2015.py

@@ -1,0 +1,76 @@
+import argparse
+import sys
+from os import path, listdir
+
+import pandas as pd
+import scipy.sparse as sp
+import anndata
+import scanpy as sc
+import utils
+
+
+parser = argparse.ArgumentParser()
+parser.add_argument("-i", "--in-dir", required=True)
+parser.add_argument("-o", "--out-dir", required=True)
+parser.add_argument("-f", "--force", action="store_true")
+args = parser.parse_args()
+
+fname = path.join(args.out_dir, "macosko_2015.h5ad")
+
+if path.exists(fname) and not args.force:
+    print(f"`{fname}` exists. Skipping...")
+    sys.exit(0)
+
+data = pd.read_table(path.join(args.in_dir, "GSE63472_P14Retina_merged_digital_expression.txt.gz"), index_col=0)
+
+cluster_ids = pd.read_table(path.join(args.in_dir, "retina_clusteridentities.txt"), header=None, index_col=0, squeeze=True)
+
+# Drop unlabelled cells
+data = data.loc[:, cluster_ids.index]
+
+# Convert cluster ids to cell type names
+cell_types = cluster_ids.astype(object)
+
+cell_types.loc[cell_types == 1] = "Horizontal cells"
+cell_types.loc[cell_types == 2] = "Retinal ganglion cells"
+cell_types.loc[cell_types.isin(range(3, 24))] = "Amacrine cells"
+cell_types.loc[cell_types == 24] = "Rods"
+cell_types.loc[cell_types == 25] = "Cones"
+cell_types.loc[cell_types.isin(range(26, 34))] = "Bipolar cells"
+cell_types.loc[cell_types == 34] = "Muller glia"
+cell_types.loc[cell_types == 35] = "Astrocytes"
+cell_types.loc[cell_types == 36] = "Fibroblasts"
+cell_types.loc[cell_types == 37] = "Vascular endothelium"
+cell_types.loc[cell_types == 38] = "Pericytes"
+cell_types.loc[cell_types == 39] = "Microglia"
+
+# Create anndata object
+x = sp.csr_matrix(data.values)
+
+adata = anndata.AnnData(
+    x.T,
+    var={"var_names": data.index.values},
+    obs={
+        "obs_names": data.columns.values,
+        "cluster_ids": cluster_ids.values,
+        "labels": cell_types.values,
+    },
+)
+
[…]

### 2022-03-19 figures/generate_h5ad/shekhar_2016.py

@@ -1,0 +1,69 @@
+import argparse
+import sys
+from os import path, listdir
+
+import pandas as pd
+import scipy.sparse as sp
+import anndata
+
+
+parser = argparse.ArgumentParser()
+parser.add_argument("-i", "--in-dir", required=True)
+parser.add_argument("-o", "--out-dir", required=True)
+parser.add_argument("-f", "--force", action="store_true")
+args = parser.parse_args()
+
+fname = path.join(args.out_dir, "shekhar_2016.h5ad")
+
+if path.exists(fname) and not args.force:
+    print(f"`{fname}` exists. Skipping...")
+    sys.exit(0)
+
+data = pd.read_table(path.join(args.in_dir, "GSE81904_BipolarUMICounts_Cell2016.txt.gz"), index_col=0)
+
+metadata = pd.read_table(path.join(args.in_dir, "clust_retinal_bipolar.txt"), index_col=0)
+
+data = data.loc[:, metadata.index.values]
+metadata = metadata.loc[data.columns.values]
+
+
+cell_types = metadata["SUB-CLUSTER"].astype(object)
+
+cell_types[cell_types.str.contains("BC1")] = "type 1 cone bipolar cell"
+cell_types[cell_types.str.contains("BC2")] = "type 2 cone bipolar cell"
+cell_types[cell_types.str.contains("BC3")] = "type 3 cone bipolar cell"
+cell_types[cell_types.str.contains("BC4")] = "type 4 cone bipolar cell"
+cell_types[cell_types.str.contains("BC5")] = "type 5 cone bipolar cell"
+cell_types[cell_types.str.contains("BC6")] = "type 6 cone bipolar cell"
+cell_types[cell_types.str.contains("BC7")] = "type 7 cone bipolar cell"
+
+# This is a mixture, so go up one level in the ontology
+cell_types[cell_types == "BC8/9 (mixture of BC8 and BC9)"] = "type 8/9 cone bipolar cell"
+
+# Non-bipolar cells
+cell_types[cell_types == "RBC (Rod Bipolar cell)"] = "rod bipolar cell"
+cell_types[cell_types == "MG (Mueller Glia)"] = "Mueller cell"
+cell_types[cell_types == "AC (Amacrine cell)"] = "amacrine cell"
+cell_types[cell_types == "Rod Photoreceptors"] = "retinal rod cell"
+cell_types[cell_types == "Cone Photoreceptors"] = "retinal cone cell"
+
+
[…]

### 2022-03-19 figures/generate_h5ad/tasic_2018.py

@@ -1,0 +1,64 @@
+import argparse
+import sys
+from os import path, listdir
+
+import pandas as pd
+import scipy.sparse as sp
+import anndata
+import scanpy as sc
+import utils
+
+
+parser = argparse.ArgumentParser()
+parser.add_argument("-i", "--in-dir", required=True)
+parser.add_argument("-o", "--out-dir", required=True)
+parser.add_argument("-f", "--force", action="store_true")
+args = parser.parse_args()
+
+fname = path.join(args.out_dir, "tasic_2018.h5ad")
+
+if path.exists(fname) and not args.force:
+    print(f"`{fname}` exists. Skipping...")
+    sys.exit(0)
+
+data = pd.read_csv(path.join(args.in_dir, "GSE115746_cells_exon_counts.csv.gz"), index_col=0)
+
+metadata = pd.read_csv(path.join(args.in_dir, "GSE115746_complete_metadata_28706-cells.csv.gz"), index_col=0)
+
+x = sp.csr_matrix(data.values)
+
+adata = anndata.AnnData(
+    x.T,
+    var={"var_names": data.index.values},
+    obs={
+        "obs_names": data.columns.values,
+        "labels": metadata["cell_class"].astype(str),
+        "batch": metadata["source_name"].astype(str),
+    },
+)
+
+# Load colors
+color_metadata = pd.read_csv(path.join(args.in_dir, "tasic-sample_heatmap_plot_data.csv"), index_col=0)
+
+adata_mask = adata.obs_names.isin(color_metadata.index)
+adata = adata[adata_mask]
+
+adata.obs["cluster_color"] = color_metadata.loc[adata.obs_names.values, "cluster_color"]
+
+adata.write_h5ad(fname)
+
+print("Running single-cell pipeline...")
+sc.pp.filter_genes(adata, min_counts=10)
+gene_mask = utils.select_genes(adata.X, n=3000, threshold=0)
+
+sc.pp.normalize_per_cell(adata)
+sc.pp.log1p(adata)
+sc.pp.scale(adata, zero_center=False)
+
+X_pca = utils.pca(adata.X[:, gene_mask])
+
+# Reload the original, unprocessesed data set, and add the PCA coordinates
+adata = anndata.read_h5ad(fname)
+adata.obsm["X_pca"] = X_pca
+
+adata.write_h5ad(fname)

### 2022-03-19 figures/utils.py

@@ -1,0 +1,457 @@
+from os.path import abspath, dirname, join
+
+import numpy as np
+import scipy.sparse as sp
+
+FILE_DIR = dirname(abspath(__file__))
+DATA_DIR = join(FILE_DIR, "data")
+
+
+def calculate_cpm(x, axis=1):
+    """Calculate counts-per-million on data where the rows are genes.
+
+    Parameters
+    ----------
+    x : array_like
+    axis : int
+        Axis accross which to compute CPM. 0 for genes being in rows and 1 for
+        genes in columns.
+
+    """
+    normalization = np.sum(x, axis=axis)
+    # On sparse matrices, the sum will be 2d. We want a 1d array
+    normalization = np.squeeze(np.asarray(normalization))
+    # Straight up division is not an option since this will form a full dense
+    # matrix if `x` is sparse. Divison can be expressed as the dot product with
+    # a reciprocal diagonal matrix
+    normalization = sp.diags(1 / normalization, offsets=0)
+    if axis == 0:
+        cpm_counts = np.dot(x, normalization)
+    elif axis == 1:
+        cpm_counts = np.dot(normalization, x)
+    return cpm_counts * 1e6
+
+
+def log_normalize(data):
+    """Perform log transform log(x + 1).
+
+    Parameters
+    ----------
+    data : array_like
+
+    """
+    if sp.issparse(data):
+        data = data.copy()
+        data.data = np.log2(data.data + 1)
+        return data
+
+    return np.log2(data.astype(np.float64) + 1)
+
+
+def pca(x, n_components=50):
+    if sp.issparse(x):
+        x = x.toarray()
+    U, S, V = np.linalg.svd(x, full_matrices=False)
+    U[:, np.sum(V, axis=1) < 0] *= -1
+    x_reduced = np.dot(U, np.diag(S))
+    x_reduced = x_reduced[:, np.argsort(S)[::-1]][:, :n_components]
+    return x_reduced
+
+
+def select_genes(
+    data,
+    threshold=0,
+    atleast=10,
+    yoffset=0.02,
+    xoffset=5,
+    decay=1,
+    n=None,
+    plot=True,
+    markers=None,
+    genes=None,
+    figsize=(6, 3.5),
+    markeroffsets=None,
+    labelsize=10,
+    alpha=1,
+    verbose=False,
+):
+    if sp.issparse(data):
+        zeroRate = 1 - np.squeeze(np.array((data > threshold).mean(axis=0)))
[…]

### 2022-03-19 figures/download_h5ad.sh
Minor fixes to data gen scripts

@@ -26,6 +26,6 @@ Extracting files...
 ================================================================================
 EOF
 
-for f in "$H5AD_DIR/*.h5ad.tar.gz"; do
+for f in "$H5AD_DIR"/*.h5ad.tar.gz; do
     tar xvfa "$f" --remove-files
 done

### 2022-03-19 figures/generate_h5ad/cao_2019.py

@@ -1,14 +1,12 @@
+import anndata
 import argparse
-import sys
-from os import path, listdir
-
 import numpy as np
 import pandas as pd
-import scipy.sparse as sp
-import anndata
 import scanpy as sc
+import scipy.sparse as sp
+import sys
 import utils
-
+from os import path, listdir
 
 parser = argparse.ArgumentParser()
 parser.add_argument("-i", "--in-dir", required=True)

### 2022-03-19 figures/generate_h5ad/harris_2018.py

@@ -1,11 +1,9 @@
+import anndata
 import argparse
-import sys
-from os import path, listdir
-
 import pandas as pd
 import scipy.sparse as sp
-import anndata
-
+import sys
+from os import path, listdir
 
 parser = argparse.ArgumentParser()
 parser.add_argument("-i", "--in-dir", required=True)

### 2022-03-19 figures/generate_h5ad/hochgerner_2018.py

@@ -1,13 +1,11 @@
-import argparse
-import sys
-from os import path, listdir
-
-import pandas as pd
-import scipy.sparse as sp
 import anndata
+import argparse
+import pandas as pd
 import scanpy as sc
+import scipy.sparse as sp
+import sys
 import utils
-
+from os import path, listdir
 
 parser = argparse.ArgumentParser()
 parser.add_argument("-i", "--in-dir", required=True)

### 2022-03-19 figures/utils.py

@@ -3,9 +3,6 @@ from os.path import abspath, dirname, join
 import numpy as np
 import scipy.sparse as sp
 
-FILE_DIR = dirname(abspath(__file__))
-DATA_DIR = join(FILE_DIR, "data")
-
 
 def calculate_cpm(x, axis=1):
     """Calculate counts-per-million on data where the rows are genes.

### 2022-03-21 figures/fig1-macosko.py
Figs: Add (unfinished) scripts to generate figures

@@ -1,0 +1,269 @@
+import argparse
+import os
+
+import numpy as np
+import scanpy as sc
+import utils
+import openTSNE
+import matplotlib.pyplot as plt
+
+
+parser = argparse.ArgumentParser()
+parser.add_argument("-i", "--in-dir", required=True)
+parser.add_argument("-o", "--out-dir", required=True)
+parser.add_argument("-f", "--force", action="store_true")
+args = parser.parse_args()
+
+if not os.path.exists(args.out_dir):
+    os.mkdir(args.out_dir)
+
+print("Reading data...")
+adata = sc.read_h5ad(os.path.join(args.in_dir, "macosko_2015.h5ad"))
+new = sc.read_h5ad(os.path.join(args.in_dir, "shekhar_2016.h5ad"))
+
+### t-SNE Embeddings
+if (
+    "X_tsne_standard" not in adata.obsm or
+    "X_tsne_modern" not in adata.obsm or
+    "X_tsne_transform" not in new.obsm or
+    args.force
+):
+    # Standard embedding
+    print("Creating standard embedding...")
+    embedding_standard = openTSNE.TSNE(
+        perplexity=30,
+        initialization="random",
+        metric="cosine",
+        learning_rate=200,
+        n_iter=750,
+        n_jobs=8,
+        random_state=0,
+        verbose=True,
+    ).fit(adata.obsm["X_pca"])
+    
+    # Multiscale embedding
+    print("Creating multiscale embedding...")
+    embedding_modern = openTSNE.TSNE(
+        perplexity=[50, 500],
+        initialization="pca",
+        metric="cosine",
+        learning_rate="auto",
+        n_iter=500,
+        n_jobs=8,
+        random_state=0,
+        verbose=True,
+    ).fit(adata.obsm["X_pca"])
+    
+    ### Prepare new data to transform
+    print("Reloading data...")
+    adata = sc.read_h5ad(os.path.join(args.in_dir, "macosko_2015.h5ad"))
+    new = sc.read_h5ad(os.path.join(args.in_dir, "shekhar_2016.h5ad"))
+    
+    # Synchronize labels
+    new.obs["labels"] = new.obs["labels"].astype(str)
+    new.obs["labels"][new.obs["labels"].str.contains("bipolar cell")] = "retinal bipolar neuron"
+    new.obs["labels"] = new.obs["labels"].replace({
[…]

### 2022-03-21 figures/fig2-tasic.py

@@ -1,0 +1,112 @@
+import argparse
+import os
+
+import numpy as np
+import scanpy as sc
+import utils
+import openTSNE
+import matplotlib.pyplot as plt
+
+
+parser = argparse.ArgumentParser()
+parser.add_argument("-i", "--in-dir", required=True)
+parser.add_argument("-o", "--out-dir", required=True)
+parser.add_argument("-f", "--force", action="store_true")
+args = parser.parse_args()
+
+if not os.path.exists(args.out_dir):
+    os.mkdir(args.out_dir)
+
+print("Reading data...")
+adata = sc.read_h5ad(os.path.join(args.in_dir, "tasic_2018.h5ad"))
+
+### t-SNE Embeddings
+if (
+    "X_tsne_perp30" not in adata.obsm or
+    "X_tsne_perp500" not in adata.obsm or
+    "X_tsne_multiscale" not in adata.obsm or
+    "X_tsne_dof06" not in adata.obsm or
+    args.force
+):
+    init = openTSNE.initialization.rescale(adata.obsm["X_pca"][:, :2])
+    
+    print("Creating embedding perplexity=30...")
+    embedding_perp30 = openTSNE.TSNE(
+        perplexity=30,
+        initialization=init,
+        metric="cosine",
+        n_jobs=8,
+        random_state=0,
+        verbose=True,
+    ).fit(adata.obsm["X_pca"])
+    
+    print("Creating embedding perplexity=30...")
+    embedding_perp500 = openTSNE.TSNE(
+        perplexity=500,
+        initialization=init,
+        metric="cosine",
+        n_jobs=8,
+        random_state=0,
+        verbose=True,
+    ).fit(adata.obsm["X_pca"])
+    
+    print("Creating embedding perplexity=[30, 500]...")
+    embedding_multiscale = openTSNE.TSNE(
+        perplexity=[30, 500],
+        initialization=init,
+        metric="cosine",
+        n_jobs=8,
+        random_state=0,
+        verbose=True,
+    ).fit(adata.obsm["X_pca"])
+    
+    print("Creating embedding dof=0.6...")
+    embedding_dof06 = openTSNE.TSNE(
+        perplexity=30,
+        initialization=init,
+        metric="cosine",
+        dof=0.6,
+        n_jobs=8,
[…]

### 2022-03-21 figures/fig3-cao.py

@@ -1,0 +1,133 @@
+import argparse
+import os
+import gzip
+import pickle
+
+import numpy as np
+import scanpy as sc
+import utils
+import openTSNE
+import matplotlib.pyplot as plt
+
+
+parser = argparse.ArgumentParser()
+parser.add_argument("-i", "--in-dir", required=True)
+parser.add_argument("-o", "--out-dir", required=True)
+parser.add_argument("-f", "--force", action="store_true")
+args = parser.parse_args()
+
+if not os.path.exists(args.out_dir):
+    os.mkdir(args.out_dir)
+
+print("Reading data...")
+adata = sc.read_h5ad(os.path.join(args.in_dir, "cao_2019.h5ad"))
+
+# Generate indices for the random sample
+np.random.seed(0)
+
+x, y = adata.obsm["X_pca"], adata.obs["day"]
+
+indices = np.random.permutation(list(range(x.shape[0])))
+reverse = np.argsort(indices)
+    
+x_sample, x_rest = x[indices[:25000]], x[indices[25000:]]
+y_sample, y_rest = y[indices[:25000]], y[indices[25000:]]
+
+
+# Create sample embedding
+if not os.path.exists("_cache/cao_sample_embedding.pkl.gz"):
+    print("Creating sample embedding...")
+    
+    sample_affinities = openTSNE.affinity.PerplexityBasedNN(
+        x_sample,
+        perplexity=500,
+        metric="cosine",
+        n_jobs=24,
+        random_state=0,
+        verbose=True,
+    )
+    sample_init = openTSNE.initialization.spectral(sample_affinities.P)
+    
+    sample_embedding = openTSNE.TSNEEmbedding(
+        sample_init, sample_affinities, n_jobs=24, verbose=True,
+    )
+    sample_embedding.optimize(n_iter=250, exaggeration=12, momentum=0.5, inplace=True)
+    sample_embedding.optimize(n_iter=500, exaggeration=1, momentum=0.8, inplace=True)
+    
+    if not os.path.exists("_cache"):
+        os.mkdir("_cache")
+        
+    with gzip.open(os.path.join("_cache", "cao_sample_embedding.pkl.gz"), "wb") as f:
+        pickle.dump({"embedding": sample_embedding, "y": y_sample}, f)
+
+else:
+    print("Using precomputed sample embedding...")
[…]

### 2022-03-21 figures/fig4-hochgerner.py

@@ -1,0 +1,75 @@
+import argparse
+import os
+
+import numpy as np
+import scanpy as sc
+import utils
+import openTSNE
+import matplotlib.pyplot as plt
+
+
+parser = argparse.ArgumentParser()
+parser.add_argument("-i", "--in-dir", required=True)
+parser.add_argument("-o", "--out-dir", required=True)
+parser.add_argument("-f", "--force", action="store_true")
+args = parser.parse_args()
+
+if not os.path.exists(args.out_dir):
+    os.mkdir(args.out_dir)
+
+print("Reading data...")
+adata = sc.read_h5ad(os.path.join(args.in_dir, "hochgerner_2018.h5ad"))
+
+### t-SNE Embeddings
+if (
+    "X_tsne" not in adata.obsm or
+    args.force
+):
+    print("Creating reference embedding...")
+    init = openTSNE.initialization.rescale(adata.obsm["X_pca"][:, :2])
+    
+    embedding = openTSNE.TSNE(
+        perplexity=[50, 500],
+        initialization=init,
+        metric="cosine",
+        n_jobs=8,
+        random_state=0,
+        verbose=True,
+    ).fit(adata.obsm["X_pca"], initialization=init)
+    
+    adata.obsm["X_tsne"] = embedding.view(np.ndarray)
+
+    adata.write_h5ad(os.path.join(args.in_dir, "hochgerner_2018.h5ad"))
+
+### Make figure
+print("Generating figure...")
+
+colors = {
+    # Excitatory neurons
+    "immature pyramidal neuron": "#DE6093",
+    "pyramidal neuron": "#D5221C",
+    "immature granule cell": "#FEE00D",
+    "juevnile granule cell": "#FAB508",
+    "granule cell": "#FFA02C",
+    # Inhibitory neurons
+    "immature GABAergic neuron": "#74C9FF",
+    "GABAergic neuron": "#68A8E6",
+    "Cajal-Retzius cell": "#1C6865",
+    # Neural progenitors
+    "neuroblast": "#FEA388",
+    "neuronal intermediate progenitor cells": "#FF7390",
+    # Glia
+    "ependymal cell": "#967D78",
+    "oligodendrocyte precursor cell": "#9F9F9F",
+    "newly formed oligodendrocyte": "#787878",
+    "oligodendrocyte": "#505050",
+    "immature astrocyte": "#9B7E7B",
+    "juvenile astrocyte": "#846F57",
[…]

### 2022-03-21 figures/generate_figures.sh
Figs: Add bash script to generate all figures

@@ -1,0 +1,27 @@
+#!/bin/bash
+
+set -e
+
+H5AD_DIR="./data/h5ad"
+FIGURES_DIR="./figs"
+
+
+cat << EOF
+================================================================================
+Installing Python dependencies...
+================================================================================
+EOF
+
+pip install -r requirements-figures.txt
+
+
+cat << EOF
+================================================================================
+Generating figures...
+================================================================================
+EOF
+
+mkdir -p "$FIGURES_DIR"
+
+echo "Generating Figure 1..."
+python fig1-macosko.py -i "$H5AD_DIR" -o "$FIGURES_DIR"

### 2022-03-21 benchmarks/run_julia.sh
Change conda requirement files to yml

@@ -21,7 +21,7 @@ wget -nc -P data http://file.biolab.si/opentsne/10x_mouse_zheng.pkl.gz
 # Prepare logs directory
 mkdir -p logs
 
-conda env export > logs/00--run_julia--conda_env.txt
+conda env export > logs/00--run_julia--conda_env.yml
 
 # RUN BENCHMARKS
 SAMPLE_SIZES=(1000 5000 10000 20000 50000);

### 2022-03-21 benchmarks/run_python.sh

@@ -31,7 +31,7 @@ wget -nc -P data http://file.biolab.si/opentsne/10x_mouse_zheng.pkl.gz
 # Prepare logs directory
 mkdir -p logs
 
-conda env export > logs/00--run_python--conda_env.txt
+conda env export > logs/00--run_python--conda_env.yml
 
 # RUN BENCHMARKS
 SAMPLE_SIZES=(1000 100000 250000 500000 750000 1000000);

### 2022-03-21 benchmarks/run_r.sh

@@ -21,7 +21,7 @@ python convert_pickle_to_csv.py
 # Prepare logs directory
 mkdir -p logs
 
-conda env export > logs/00--run_r--conda_env.txt
+conda env export > logs/00--run_r--conda_env.yml
 
 # RUN BENCHMARKS
 REPETITIONS=6;

### 2022-03-22 paper/main.tex
Update benchmarking figures in manuscript

@@
 However, easy access to efficient implementations almost universally correlates with the widespread adoption of novel computational techniques.
-Despite the many theoretical advancements, the most popular t-SNE libraries have been slow to incorporate them into their implementations.
-In this paper, we present \opentsne, our open-source Python implementation of t-SNE, and the recently proposed extensions to t-SNE that \opentsne\ implements.
+Despite the many theoretical advancements, most popular t-SNE libraries have been slow to incorporate them into their implementations.
+In this paper, we present \opentsne, our open-source Python implementation of t-SNE, complete with its recently proposed extensions.
 \opentsne\ is easy to install and includes precompiled binaries available through popular Python package managers.
@@
 side-effect, embeddings converge faster, and the number of iterations can be
-lowered from the typical 1000 to 750, decreasing the overall runtime.  Modern
-t-SNE implementations, including \pkg{FIt-SNE} and \opentsne, have adopted this
-convention.
+lowered from the typical 1000 to 750, decreasing the overall runtime.
 
@@
 \citet{linderman2019fast} proposed an alternative approach, FIt-SNE, based on
-non-uniform convolutions which further reduces the time complexity to
+non-uniform convolutions, which further reduces the time complexity to
 $\mathcal{O}(N)$.
@@
 
-t-SNE is non-parametric and does not define an explicit mapping from the
+The t-SNE algorithm is non-parametric and does not define an explicit mapping from the
 high-dimensional space to the embedding space. Therefore embeddings of new data
-points need to be found through the use of optimization
+points need to be found through optimization
 techniques~\citep{policar2021embedding}. When adding new data points to an
@@
 
-One of t-SNE's criticisms is that it fails to preserve long-range distances but instead focuses on capturing the local structure of the data manifold~\citep{becht2019dimensionality}.
+One of t-SNE's long-standing criticisms is that it fails to preserve long-range distances but instead focuses on capturing the local structure of the data manifold~\citep{becht2019dimensionality}.
 Recently, several approaches have been proposed to improve the global organization of the resulting embeddings.
@@
 This can be achieved by increasing the perplexity parameter.
-However, this often comes at the price of local structure.
-\citet{kobak2019art} propose to use mixtures of Gaussians instead to preserve better short-range and long-range distances and, thus, achieve a better trade-off between local and global structure.
+However, increasing perplexity often leads to loss in local sturcture and tends to obscure small clusters.
+\citet{kobak2019art} instead propose to use mixtures of Gaussians to preserve better short-range and long-range distances and, thus, achieve a better trade-off between local and global structure.
 
@@
 Recently, \citet{bohm2020unifying} showed that the exaggeration factor $\rho$ could be used to produce layouts more similar to UMAP and ForceAtlas2.
-By incorporating exaggeration into later phases of the optimization, t-SNE introduces white space between clusters, which better reflects the global relations between clusters.
+By incorporating exaggeration into later phases of the optimization, t-SNE introduces white space between clusters, which may better reflect the global relations between clusters.
 
@@
 
-\subsection{Structure of openTSNE}
+We introduce \opentsne, a comprehensive Python library that implements the
+t-SNE algorithm and its numerous recentl extensions. \opentsne\ is
+designed to accomodate all types of users and includes both a simple API that
+allows users to quickly obtain informative t-SNE visualizations, as well as a
+more flexible API that allows seasoned practitioners to modify all the
+different aspects of the t-SNE algorithm.
 
-\opentsne\ is designed to accomodate all users, including both a simple API that allows users to quickly obtain standard t-SNE visualization, as well as an advanced API that allows practitioners and researchers to modify different aspects of the t-SNE algorithm.
+The t-SNE algorithm is comprised of three main steps, mirrored in the structure
+and flow of the \opentsne\ library. These are: the initial positions
+of points in the embedding, an affinity model describing the similarities
+between data points in the embedding, and the optimization phase which
+optimizes the embedding layout. Using the advanced API, each of these key
+elements can be modified or tuned according to produce the best possible
+visualization.
 
-The t-SNE algorithm is dependent on three key steps: the initial positions of points in the embedding, an affinity model describing the similarities between data points in the ambient space, and the optimization which optimizes the embedding layout.
-\opentsne\ mirrors these different components in its design.
-\opentsne\ contains three corresponding core modules -- \code{initialization}, \code{affinity}, and \code{tsne} -- each of which will be decribed in detail in later sections.
-These modules are designed to be easily extensible, allowing researchers to experiment with different configurations and produce the best possible t-SNE visualizations.
-For less demanding users, \opentsne\ also provides a familiar, scikit-learn-like interface through the \code{openTSNE.TSNE} class.
-
-\subsubsection{Initialization}
-
-just a numpy array
-
-\subsubsection{Affinities}
-
-\paragraph{Affinity}
-
-\paragraph{KNNIndex}
-
-\subsubsection{Embedding objects}
-
-\paragraph{TSNEEmbedding}
-
-\paragraph{PartialTSNEEmbedding}
-
-\subsection{Old text}
-
-We introduce \opentsne, a comprehensive Python library that implements t-SNE and its numerous recently proposed extensions, including the implementation of efficient approximation schemes allowing the embedding of millions of data points~\citep{van2014accelerating,linderman2019fast}, the addition of new data points into existing embeddings~\citep{policar2021embedding}, various initialization improvements~\citep{kobak2019umap}, extensions~\citep{kobak2019heavy,kobak2019art}, and improved parameter defaults~\citep{belkina2019automated}.
-When combined, these improvements result in the faster generation of more informative visualizations.
+\begin{figure*}[ht]
+  \centering
+  \includegraphics[width=0.75\textwidth]{opentsne-diagram}
+  \caption{\label{fig:diagram}
+  Bla
+  }
+\end{figure*}
 
 \opentsne\ is compatible with the Python data science ecosystem (e.g.
-\pkg{numpy}, \pkg{scikit-learn}, \pkg{scanpy}), providing a familiar and intuitive API to users.
+\pkg{numpy}, \pkg{scikit-learn}, \pkg{scanpy}), providing a familiar and intuitive API to end-users.
 Its modular design encourages extensibility and experimentation with various parameter settings and changes to the analysis pipeline.
 \opentsne\ aims to make the latest theoretical advances in t-SNE accessible to the wider data science community.
-These features are easily accessible through an intuitive API, which closely follows the style \pkg{scikit-learn}~\citep{sklearn_api}, a popular and widely used toolkit for machine learning.
+These features are easily accessible through an intuitive API, which closely follows the style \pkg{scikit-learn}~\citep{sklearn_api}, a popular and widely used toolkit in the machine learning community.
 The following snippet exemplifies this interface and demonstrates how \opentsne\ can be used to quickly generate several visualizations, each highlighting different aspects of the underlying data structure.
+
+\subsection{Example}
+\subsubsection{Installation}
+
+\opentsne\ is designed to be as user-friendly as possible and can be installed through either \pkg{PyPI} using \pkg{pip}
+\begin{CodeChunk}
+\begin{CodeInput}
+pip install opentsne
+\end{CodeInput}
+\end{CodeChunk}
+or using \pkg{conda} from the \pkg{conda-forge} channel
+\begin{CodeChunk}
+\begin{CodeInput}
+conda install -c conda-forge opentsne
+\end{CodeInput}
+\end{CodeChunk}
+\opentsne\ provides precompiled binaries for all major platforms (Windows, MacOS, Linux) for all currently supported Python versions (3.6---3.10).
+This is especially convenient for Windows users, as this obviates the need for a C/C++ compiler, which do not come bundled with the operating system.
+
+To verify that that \opentsne\ was installed correctly, try to import the module:
+\begin{CodeChunk}
+\begin{CodeInput}
+import openTSNE
+print(openTSNE.__version__)
+\end{CodeInput}
+\end{CodeChunk}
+
+\subsubsection{A simple workflow}
+
 \begin{CodeChunk}
@@
 
-\section{Summary and discussion} \label{sec:summary}
+\section{Discussion and Examples} \label{sec:summary}
 
@@
 
+\subsection{Embedding Massive Data Sets}
+
 When dealing with data containing millions of data points, standard t-SNE embeddings often become unwieldy -- cluster boundaries are blurred, large clusters absorb smaller ones, and relationships between clusters become increasingly difficult to interpret.
@@
 # To improve the global structure of the final layout, we initialize the
-# embedding by creating a t-SNE plot on a sample of points.
+# embedding by creating a t-SNE plot on a sample of points
 
@@
 
-# We can next use the `transform` functionality to determine starting points for
-# the remaining data points. Since we don't actually want to optimize the
+# We can next use the `transform` functionality to determine starting positions
+# for the remaining data points. Since we don't actually want to optimize the
 # remaining points separately from the sample points, we don't need to call the
@@
 
-# Lastly, we merge `init_sample` and `init_rest`
+# Lastly, we merge `init_sample` and `init_rest` and restore the original order
 init_full = np.vstack((init_sample, init_rest))[reverse]
@@
 \end{CodeChunk}
-We generate the four different embeddings using the following code snippet (which is shortened for clarity).
+We generate the three different embeddings using the following code snippet (which is shortened for clarity).
 \begin{CodeChunk}
@@
 One of the t-SNE's common criticisms is limited scalability to large data sets
-containing, for instance, millions of data
-points~\citep{becht2019dimensionality}. The culprit for slow response time stems
-from particular optimization procedures and their specific implementation in
-popular open-source libraries. Until quite recently, most popular implementations of
-t-SNE were based on either a direct implementation of the t-SNE algorithm with asymptotic time complexity $\mathcal{O}(N^2)$, or the Barnes-Hut approximation scheme developed by
-\citet{van2014accelerating} with asymptotic time complexity $\mathcal{O}(N \log
-N)$, where $N$ is the number of data items ({\em e.g.} cells).
+containing, for instance, millions of data points~\citep{becht2019dimensionality}.
+The culprit for slow response time stems from
+particular optimization procedures and their specific implementation in popular
+open-source libraries. Until quite recently, most popular implementations of
+t-SNE were based on either a direct implementation of the t-SNE algorithm with
+asymptotic time complexity $\mathcal{O}(N^2)$, or the Barnes-Hut approximation
+scheme developed by
+\citet{van2014accelerating} with asymptotic time complexity $\mathcal{O}
+ (N \log N)$, where $N$ is the number of data items ({\em e.g.} cells).
 
 % Problem with Python
-In Python, the most
-widely-used implementation of t-SNE came from \pkg{scikit-learn}, which exhibits
-long runtimes when compared to its C++ counterpart -- \pkg{MulticoreTSNE}. The
-multi-threaded \pkg{MulticoreTSNE} implementation~\citep{Ulyanov2016} can
-construct t-SNE embeddings of millions of data points in a matter of hours on
-widely-accessible, consumer-grade processors (Fig.~\ref{fig:benchmarks_py}).
-However, recently, \citet{linderman2019fast} developed a new approximation
-scheme -- FIt-SNE -- which reduces the asymptotic time complexity to
-$\mathcal{O}(N)$. This approximation scheme is included in \pkg{FIt-SNE} and \opentsne, enabling the embedding of large data sets in a matter of minutes.
+In Python, the most widely-used implementation of t-SNE came from \pkg{scikit-learn},
+which exhibits long runtimes when compared to its C++ counterpart --
+\pkg{MulticoreTSNE}. The multi-threaded \pkg{MulticoreTSNE}
+implementation~\citep{Ulyanov2016} can construct t-SNE embeddings of millions of
+data points in a matter of hours (Fig.~\ref{fig:benchmarks_py}).
+However, recently, \citet{linderman2019fast} developed a
+new approximation scheme -- FIt-SNE -- which futher reduces the asymptotic time
+complexity to $\mathcal{O}(N)$. This approximation scheme is included in \pkg{FIt-SNE} and \opentsne, enabling the embedding of large data sets in a matter
+of minutes.
 
@@
 including those from \pkg{scikit-learn} (v1.0.1), \pkg{MulticoreTSNE} (v0.1),
-\pkg{FIt-SNE} (v1.1.0), and our \opentsne\ (v0.6.1). We perform benchmarks on
-a server-grade Intel(R) Xeon(R) CPU E5-1650 equipped with 128 GB of memory. Benchmarks were run for $1,000$ iterations with the original
-t-SNE parameters, as some implementations do not allow their modification.
-To generate benchmark data sets of different sizes, we subsampled data from the 10X Genomics 1.3 million mouse brain data set five times, resulting in five different data sets for each size.
-In total, we run each implementation on 30 different data sets.
+\pkg{FIt-SNE} (v1.1.0), and our \opentsne\ (v0.6.1). Benchmarks were performed on
+an Intel(R) Xeon(R) CPU E5-1650 equipped with 128 GB of memory. Benchmarks were run for $1,000$ iterations with the original
+t-SNE parameters, as some implementations do not allow for their modification.
+
+To ensure a realistic benchmarking scenario, we utilize the 10X
+Genomics 1.3 million mouse brain data set, containing 1.3 million single-cell 
+gene-expression profiles. We preprocess the data using the standard single-cell
+analysis pipeline and extract the top 50 principal components. To generate
+benchmark data sets of different sizes, we subsample the data set five times,
+resulting in five different data sets for each sample size. In total, we run
+each implementation on 30 different data sets.
 
 \begin{figure*}[ht]
-  \includegraphics[width=\textwidth]{benchmarks_python}
+  \includegraphics[width=\textwidth]{benchmarks_python-final}
   \caption{\label{fig:benchmarks_py}
@@
   \pkg{FIt-SNE}~\citep{linderman2019fast} (v1.1.0). Experiments were run on a
-  on a server-grade Intel(R) Xeon(R) CPU E5-1650 equipped with 128 GB of memory.Notice
+  on a server-grade Intel(R) Xeon(R) CPU E5-1650 equipped with 128 GB of memory. Notice
   that \opentsne\ scales similarly to \pkg{FIt-SNE}, as they both use the
@@
 The results are shown in Fig.~\ref{fig:benchmarks_lang}. 
+Using 8 cores, Rtsne runs out of stack memory at 500k
 
 \begin{figure*}[ht]
-  \includegraphics[width=\textwidth]{benchmarks_langs}
+  \includegraphics[width=\textwidth]{benchmarks_langs-final}
+  %\includegraphics[width=\textwidth]{benchmarks-combined}
   \caption{\label{fig:benchmarks_lang}
@@
 \pkg{conda-forge} channel, the two most widely adopted Python package managers.
-\opentsne's interface is inspired by \pkg{scikit-learn}, which is well
+\opentsne's high-level interface is inspired by \pkg{scikit-learn}, which is well
 established in the Python data science ecosystem. \opentsne\ implements

### 2022-03-28 paper/main.tex
Add iris tutorial

@@
 \subsection{Example}
+
 \subsubsection{Installation}
@@
 \begin{CodeInput}
-import openTSNE
-print(openTSNE.__version__)
+>>> import openTSNE
+>>> print(openTSNE.__version__)
+0.6.2
 \end{CodeInput}
@@
 
+We demonstrate how to use \opentsne\ on the Iris data set from \citet{anderson1936species}.
+This small data set describes 150 iris flowers containing four variables (sepal
+length, sepal width, petal length, petal width) and their species (Iris Setosa,
+Iris Versicolor, Iris Virginica), which is the target variable.
+This data set is included in \pkg{scikit-learn}, making it easy to load:
 \begin{CodeChunk}
 \begin{CodeInput}
-from openTSNE import TSNE
-
-x: np.ndarray = ...
-
-# Vanilla t-SNE
-tsne_standard = TSNE().fit(x)
-
-# Adapt parameters to better highlight small clusters in the data
-tsne_extensions = TSNE(perplexity=50, exaggeration=2, dof=0.8).fit(x)
-
-# Use multiple perplexity values for better local/global structure trade-off
-tsne_multiscale = TSNE(perplexity=[50, 500]).fit(x)
-
-# Use the generated embeddings to add new data points
-new_embedding = tsne_extensions.transform(x_new)
+>>> from sklearn import datasets
+>>>
+>>> iris = datasets.load_iris()
+>>> iris.data[:3]
+array([[5.1, 3.5, 1.4, 0.2],
+       [4.9, 3. , 1.4, 0.2],
+       [4.7, 3.2, 1.3, 0.2]])
 \end{CodeInput}
 \end{CodeChunk}
-The code snippet considers a \pkg{numpy} array or \pkg{scipy} sparse matrix \texttt{x} and creates three \texttt{TSNEEmbedding} objects.
-These represent fully-optimized t-SNE embeddings, and \opentsne\ internally stores them as \pkg{numpy} arrays.
-These can be manipulated and visualized using standard Python tools.
-We then select one of these embeddings and use it to embed new data points from \texttt{x\_new} into the embedding space while keeping the reference embedding fixed.
-Additionally, \opentsne\ provides a more advanced interface that enables the user to change every aspect of the t-SNE optimization procedure.
-We describe this interface in detail in the \opentsne's on-line documentation.
+\pkg{scikit-learn} loads the data set into a \pkg{numpy} array, which can directly
+be used by \opentsne:
+\begin{CodeChunk}
+\begin{CodeInput}
+>>> import openTSNE
+>>>
+>>> embedding = openTSNE.TSNE().fit(iris.data)
+>>> embedding[:3]
+TSNEEmbedding([[-0.55581917, 17.56162344],
+               [-1.76742536, 20.07763453],
+               [-0.58444608, 20.00052495]])
+\end{CodeInput}
+\end{CodeChunk}
+\opentsne\ returns an \code{TSNEEmbedding} object, which is a subclass of
+\code{np.ndarray}. This allows users to easily manipulate the resulting embedding.
+The additional functionality provided by the \code{TSNEEmbedding} object allows
+the embedding to be further optimized using different parameter values, or be
+used as a reference when obtaining new data points.
+For users who do not need this additional functionality, this object can be
+converted to a standard \code{np.ndarray} with:
+\begin{CodeChunk}
+\begin{CodeInput}
+>>> import numpy as np
+>>>
+>>> X_tsne = embedding.view(np.ndarray)
+>>> X_tsne[:3]
+array([[-0.55581917, 17.56162344],
+       [-1.76742536, 20.07763453],
+       [-0.58444608, 20.00052495]])
+\end{CodeInput}
+\end{CodeChunk}
+Fig.~\ref{fig:iris} shows a visualization of the resulting t-SNE visualization.
+As expected, t-SNE is correctly able to group together samples from the same
+species. Additionally, the t-SNE embedding reveals that Iris Setosa is markedly
+different from the other two species, while flowers from Iris Versicolor and
+Iris Virginica may appear more similar to one another.
+\begin{figure*}[htbp]
+  \centering
+  \includegraphics[width=0.5\textwidth]{iris}
+  \caption{\label{fig:iris}
+  Visualization of the obtained t-SNE embedding on the Iris data set. The colors
+  correspond to different species of the Iris flower.
+}
+\end{figure*}
 

### 2022-03-28 paper/main.tex
Update benchmarks section

@@
 containing, for instance, millions of data points~\citep{becht2019dimensionality}.
-The culprit for slow response time stems from
-particular optimization procedures and their specific implementation in popular
+The culprit for slow response time stems from particular optimization procedures
+and their specific implementation in popular
 open-source libraries. Until quite recently, most popular implementations of
 t-SNE were based on either a direct implementation of the t-SNE algorithm with
-asymptotic time complexity $\mathcal{O}(N^2)$, or the Barnes-Hut approximation
-scheme developed by
-\citet{van2014accelerating} with asymptotic time complexity $\mathcal{O}
- (N \log N)$, where $N$ is the number of data items ({\em e.g.} cells).
-
-% Problem with Python
-In Python, the most widely-used implementation of t-SNE came from \pkg{scikit-learn},
-which exhibits long runtimes when compared to its C++ counterpart --
-\pkg{MulticoreTSNE}. The multi-threaded \pkg{MulticoreTSNE}
-implementation~\citep{Ulyanov2016} can construct t-SNE embeddings of millions of
-data points in a matter of hours (Fig.~\ref{fig:benchmarks_py}).
-However, recently, \citet{linderman2019fast} developed a
-new approximation scheme -- FIt-SNE -- which futher reduces the asymptotic time
-complexity to $\mathcal{O}(N)$. This approximation scheme is included in \pkg{FIt-SNE} and \opentsne, enabling the embedding of large data sets in a matter
-of minutes.
+asymptotic time complexity $\mathcal{O}(N^2)$~\citep{maaten2008visualizing}, or the Barnes-Hut approximation
+scheme developed by \citet{van2014accelerating} with asymptotic time complexity
+$\mathcal{O}(N \log N)$. Recently, however, \citet{linderman2019fast} developed
+a novel approximation scheme -- FIt-SNE -- which reduces the asymptotic time
+complexity to $\mathcal{O}(N)$, overcoming the difficulties of scaling t-SNE
+to large data sets.
 
 % We benchmark Python
-Fig.~\ref{fig:benchmarks_py} benchmarks four popular Python t-SNE implementations,
-including those from \pkg{scikit-learn} (v1.0.1), \pkg{MulticoreTSNE} (v0.1),
-\pkg{FIt-SNE} (v1.1.0), and our \opentsne\ (v0.6.1). Benchmarks were performed on
-an Intel(R) Xeon(R) CPU E5-1650 equipped with 128 GB of memory. Benchmarks were run for $1,000$ iterations with the original
-t-SNE parameters, as some implementations do not allow for their modification.
-
-To ensure a realistic benchmarking scenario, we utilize the 10X
-Genomics 1.3 million mouse brain data set, containing 1.3 million single-cell 
-gene-expression profiles. We preprocess the data using the standard single-cell
-analysis pipeline and extract the top 50 principal components. To generate
-benchmark data sets of different sizes, we subsample the data set five times,
-resulting in five different data sets for each sample size. In total, we run
-each implementation on 30 different data sets.
+Fig.~\ref{fig:benchmarks_py} benchmarks four popular Python t-SNE
+implementations, including \pkg{scikit-learn} (v1.0.1), \pkg{MulticoreTSNE}
+(v0.1), \pkg{FIt-SNE} (v1.1.0), and \opentsne\ (v0.6.2). Benchmarks were
+performed on an Intel(R) Xeon(R) CPU E5-1650 equipped with 128 GB of memory.
+Benchmarks were run for $1,000$ iterations with the original t-SNE parameters,
+as some implementations do not allow for their modification. To ensure a
+realistic benchmarking scenario, we utilize the 10X Genomics 1.3 million mouse
+brain data set~\citep{xxx}, containing 1.3 million single-cell gene-expression
+profiles. We preprocess the data using the standard single-cell analysis
+pipeline~\citep{XXX} and extract the top 50 principal components. To generate
+benchmark data sets of different sample sizes, we subsample the data set five
+times at each sample size. In total, we run each implementation on 30 different
+data sets.
 
 \begin{figure*}[ht]
-  \includegraphics[width=\textwidth]{benchmarks_python-final}
+  \centering
+  \includegraphics[width=0.75\textwidth]{benchmarks_python-final}
   \caption{\label{fig:benchmarks_py}
-  We benchmark \opentsne\ (v0.6.1) against three popular open-source
+  We benchmark \opentsne\ (v0.6.2) against three popular open-source
   implementations from \pkg{scikit-learn}~\citep{pedregosa2011scikit}
@@
   \pkg{FIt-SNE}~\citep{linderman2019fast} (v1.1.0). Experiments were run on a
-  on a server-grade Intel(R) Xeon(R) CPU E5-1650 equipped with 128 GB of memory. Notice
+  on an Intel(R) Xeon(R) CPU E5-1650 equipped with 128 GB of memory. Notice
   that \opentsne\ scales similarly to \pkg{FIt-SNE}, as they both use the
@@
 
+% Problem with Python
+In Python, the most widely-used implementation of t-SNE comes from \pkg{scikit-learn},
+which exhibits long runtimes when compared to other Python implementations.
+\pkg{scikit-learn}, like its C++ counterpart -- \pkg{MulticoreTSNE}~\citep{Ulyanov2016} ,
+is based on the Barnes-Hut approximation scheme. However, as
+shown on Fig.~\ref{fig:benchmarks_py}, \pkg{scikit-learn} exhibits much longer
+runtimes than \pkg{MulticoreTSNE}.
+Newer implementations, such as \pkg{FIt-SNE} and \opentsne, both implement the
+FIt-SNE approximation scheme, making them suitable for the visualization of
+millions of data points. While both libraries implement the same
+algorithm, \opentsne\ emphasizes ease of use and extensibility, and is primarily
+written in Python, which inevitably introduces some runtime overhead. This
+makes \opentsne\ marginally slower than its pure C++ counterpart -- \pkg{FIt-SNE}.
+However, these differences disappear almost entirely when
+utilizing multi-threading, as shown on Fig.~\ref{fig:benchmarks_py}.
+Additionally, \opentsne\ provides a flexible API, allowing the user to split up
+the embedding-construction process into several parts. This allows us to cache
+and reuse intermediate results, enabling users to quickly experiment with
+different parameter settings and iterate on their final visualizations. This
+kind of functionality is not supported in other implementations, which require
+the user to start from scratch for every embedding, resulting in large amounts
+of redundant computation.
+
 % Problem with other programming languages
-In R, the only widely used implementation of t-SNE -- \pkg{Rtsne}~\citep{krijthe2015rtsne} -- is based on the Barnes-Hut approximation, and is closely comparable to \pkg{MulticoreTSNE}.
-In Julia, the only available implementation -- \pkg{TSne.jl} -- is based on a direct $\mathcal{O}(N^2)$ implementation of the algorithm, which allows it to be used only on smaller data sets containing thousands of data points.
-Matlab is proprietary software, but also implements a naive version of the t-SNE algorithm, therefore we would expect its performance to mirror that of \pkg{TSne.jl} from Julia.
-
-% We benchmark other programming languages
-Using the same benchmarking setup as before, we ...
-The results are shown in Fig.~\ref{fig:benchmarks_lang}. 
-Using 8 cores, Rtsne runs out of stack memory at 500k
+In other programming languages, the selection of fast, open-source
+implementations of t-SNE is more limited. In R, the most widely used
+implementation of t-SNE -- \pkg{Rtsne}~\citep{krijthe2015rtsne} -- is based on
+the Barnes-Hut approximation, and is closely comparable to 
+\pkg{MulticoreTSNE}. In Julia, the most popular implementation -- 
+\pkg{TSne.jl}~\citep{julia_tsne} -- is based on a naive $\mathcal{O}(N^2)$ implementation of the
+algorithm, which allows it to be used only on smaller data sets containing up
+to thousands of data points. We benchmark \pkg{Rtsne} (v0.15) and 
+\pkg{TSne.jl} (v1.3.0) and compare it to \opentsne\ (v0.6.2).
+Fig.~\ref{fig:benchmarks_lang} illustrates the scaling of each approximation 
+scheme. Of these implementations, only \opentsne\ is able to scale to data sets
+containing millions of data points in a reasonable amount of time.
 
 \begin{figure*}[ht]
-  \includegraphics[width=\textwidth]{benchmarks_langs-final}
+  \centering
+  \includegraphics[width=0.75\textwidth]{benchmarks_langs-final}
   %\includegraphics[width=\textwidth]{benchmarks-combined}
   \caption{\label{fig:benchmarks_lang}
-  We choose the fastest available, open-source implementations of the t-SNE algorithm from the Python, R, and Julia programming languages.
-We choose \opentsne\ for Python, \pkg{Rtsne} for R, and \pkg{TSne.jl} for Julia.
-The benchmarks reflect the time complexity of the implemented algorithms.
-For instance, \pkg{TSne.jl} implements a naive $\mathcal{O}(N^2)$ algorithm, making it prohibitively expensive for all but the smallest data sets.
-\pkg{Rtsne} scales similarily to \pkg{MulticoreTSNE} in Python, as it implements the $\mathcal{O}(N \log N)$ Barnes-Hut approximation scheme.
-Interestingly, the multi-threaded implementation version of \pkg{Rtsne} crashes on benchmark data sets containing over 500,000 data points due to lack of memory.
-This would seriously hamper its utility on consumer-grade platforms, as our benchmark platform contained 128 GB of memory.
-\opentsne\ is the only library implementing the more recent FIt-SNE approximation scheme, which makes it suitable for even the largest data sets.
+  We select the fastest available, open-source implementations of the t-SNE
+  algorithm from the Python, R, and Julia programming languages. We choose
+  \opentsne\ (v0.6.2) for Python, \pkg{Rtsne}~\citep{krijthe2015rtsne} (v0.15)
+  for R, and \pkg{TSne.jl}~\citep{julia_tsne} (v1.3.0) for Julia. The benchmarks
+  reflect the time complexity of the
+  implemented approximation schemes. For instance, \pkg{TSne.jl} implements a
+  naive $\mathcal{O}(N^2)$ algorithm, making it prohibitively expensive for all
+  but the smallest of data sets. \pkg{Rtsne} scales similarily to
+  \pkg{MulticoreTSNE} in Python, as it implements the $\mathcal{O}(N \log N)$
+  Barnes-Hut approximation scheme. Interestingly, the multi-threaded
+  implementation version of \pkg{Rtsne} crashes on benchmark data sets
+  containing over 500,000 data points due to lack of stack memory. We found
+  these crashes to be inconsistent and somwewhat random. \opentsne\ is the only
+  library implementing the more recent FIt-SNE approximation scheme, which makes
+  it suitable for larger data sets.
 }
 \end{figure*}
-
-Benchmark results (Fig.~\ref{fig:benchmarks_py}) confirm that both \pkg{FIt-SNE}
-and \opentsne\ scale better than their Barnes-Hut counterparts --
-\pkg{scikit-learn} and \pkg{MulticoreTSNE}. While \opentsne's implementation
-uses Python which incurs some runtime overhead compared to its C++ counterpart
-\pkg{FIt-SNE}, the two libraries' speeds are surprisingly comparable. Modern
-computer processors contain multiple cores, allowing us to use multi-threading,
-which further reduces the gap between \opentsne\ and \pkg{FIt-SNE}. On the
-server-grade processor, \opentsne\ is slightly faster than its pure C++
-counterpart when utilizing multiple cores. \opentsne\ uses \pkg{numpy} for most
-linear algebra operations, which may make better use of the Intel Math Kernel
-Library (MKL), which is aggressively optimized on Intel(R) Xeon(R) processors.
-
-\opentsne\ provides a flexible API, allowing splitting up the
-embedding-construction process into several parts, caching slow operations.
-Running t-SNE optimization in stages enables users to quickly experiment with
-different parameter settings and iterate on their final visualizations.
 
@@
 
-\begin{tabular}{l c c c c c|c c}
+\begin{tabular}{l c c c c|c c}
 \toprule
 \setlength\tabcolsep{6pt}
-& \rot{\pkg{scikit-learn}} & \rot{\pkg{BHTSNE}} & \rot{\pkg{MulticoreTSNE}} & \rot{\pkg{FIt-SNE}} & \rot{\pkg{openTSNE}} & \rot{\pkg{Rtsne}} & \rot{\pkg{Tsne.jl}} \\
+& \rot{\pkg{scikit-learn}} & \rot{\pkg{MulticoreTSNE}} & \rot{\pkg{FIt-SNE}} & \rot{\pkg{openTSNE}} & \rot{\pkg{Rtsne}} & \rot{\pkg{Tsne.jl}} \\
 
@@
 \textbf{Ease of Installation} \\
-Native Python installation & \checkmark & & & & \checkmark & & \\
-\pkg{PyPI} package & \checkmark & & \checkmark & & \checkmark & & \\
-\pkg{conda} package & \checkmark & & & & \checkmark & & \\
+Native Python installation & \checkmark & & & \checkmark & & \\
+\pkg{PyPI} package & \checkmark & \checkmark & & \checkmark & & \\
+\pkg{conda} package & \checkmark & & & \checkmark & & \\
 
@@
 \textbf{Approximation Schemes} \\
-None ($\mathcal{O}(N^2)$) & & & & & & & \checkmark \\
-Barnes-Hut ($\mathcal{O}(N \log N)$) & \checkmark & \checkmark & \checkmark & & \checkmark & \checkmark & \\
-FIt-SNE ($\mathcal{O}(N)$) & & & & \checkmark & \checkmark & & \\
+None ($\mathcal{O}(N^2)$) & & & & & & \checkmark \\
+Barnes-Hut ($\mathcal{O}(N \log N)$) & \checkmark & \checkmark & & \checkmark & \checkmark & \\
+FIt-SNE ($\mathcal{O}(N)$) & & & \checkmark & \checkmark & & \\
 
@@
 %Multiscale Gaussian kernels & & & & \checkmark & \checkmark & & \\
-Extensible affinity models & & & & & \checkmark & & \\
-Variable degrees of freedom & & & & \checkmark & \checkmark & & \\
-Variable exaggeration & & & & \checkmark & \checkmark & & \\
-Globally-consistent initialization & & & & \checkmark & \checkmark & & \\
-Automatic learning rate & & & & \checkmark & \checkmark & & \\
-Adding samples to embeddings & & & & & \checkmark & & \\
-Callback system & & & & & \checkmark & & \\
-Interactive optimization  & & & & & \checkmark & & \\
+Extensible affinity models & & & & \checkmark & & \\
+Variable degrees of freedom & & & \checkmark & \checkmark & & \\
+Variable exaggeration & & & \checkmark & \checkmark & & \\
+Globally-consistent initialization & & & \checkmark & \checkmark & & \\
+Automatic learning rate & & & \checkmark & \checkmark & & \\
+Adding samples to embeddings & & & & \checkmark & & \\
+Callback system & & & & \checkmark & & \\
+Interactive optimization  & & & & \checkmark & & \\
 
@@
 \textbf{Project Quality}\\
-Actively maintained & \checkmark & & & & \checkmark & \checkmark & \checkmark \\
-Continuous integration & \checkmark & & \checkmark & & \checkmark & \checkmark & \checkmark \\
-User documentation & \checkmark & & & & \checkmark & \checkmark  & \\
-End-to-end usage examples & \checkmark & & & \checkmark & \checkmark & \checkmark & \checkmark \\
-API reference & \checkmark & & & & \checkmark & \checkmark &  \\
+Actively maintained & \checkmark & & & \checkmark & \checkmark & \checkmark \\
+Continuous integration & \checkmark & \checkmark & & \checkmark & \checkmark & \checkmark \\
+User documentation & \checkmark & & & \checkmark & \checkmark  & \\
+End-to-end usage examples & \checkmark & & \checkmark & \checkmark & \checkmark & \checkmark \\
+API reference & \checkmark & & & \checkmark & \checkmark &  \\
 \bottomrule
@@
 \caption{\label{tab:features}
-  We compare the features of \opentsne~(v0.6.1) to four popular open-source Python
-  t-SNE implementations from \pkg{scikit-learn}~(v1.0.1),
+  We compare the features of \opentsne~(v0.6.2) to \pkg{scikit-learn}~(v1.0.1),
   \pkg{BHTSNE}~(master), \pkg{MulticoreTSNE}~(v0.1), and
@@
 
-\section*{Computational details}
+\section*{Reproducibility}
 
 The results in this paper were obtained using \proglang{Python}~3.10.2 with the
-\pkg{openTSNE}~v0.6.1 package. \opentsne\ has three direct dependencies:
-\pkg{numpy}, \pkg{scipy}, and \pkg{scikit-learn}. In our experiments, we used
-\pkg{numpy}~v1.21.3, \pkg{scipy}~v1.7.1. \pkg{scikit-learn}~v1.0.1. As these
-packages include their own downstream dependencies, the full environment used to
-produce the figures is included in the accompanying GitHub repository
+\pkg{openTSNE}~v0.6.2 package. All results and figures can be reproduced using
+the scripts found in the accompanying GitHub repository
 \url{https://github.com/pavlin-policar/opentsne-paper}.

### 2022-03-28 benchmarks/generate_figures.py
Benchmark: Update figure script

@@ -9,6 +9,9 @@ import pandas as pd
 import matplotlib
 import matplotlib.pyplot as plt
 import seaborn as sns
+from matplotlib.lines import Line2D
+
+matplotlib.rcParams["pdf.fonttype"] = 42  # Make fonts editable in AI
 
 
 BENCHMARK_DIR = "../benchmarks/logs"
@@ -120,8 +123,8 @@ implementations = [
     Spec("UMAP", log_name="UMAP", cores=8, lang="Python", parser=parse_umap),
 
     # Rtsne
-    Spec("Rtsne", log_name="Rtsne", cores=1, lang="R", parser=from_minutes(parse_rtsne)),
-    Spec("Rtsne", log_name="Rtsne", cores=8, lang="R", parser=from_minutes(parse_rtsne)),
+    Spec("Rtsne", log_name="Rtsne", cores=1, lang="R", parser=parse_rtsne),
+    Spec("Rtsne", log_name="Rtsne", cores=8, lang="R", parser=parse_rtsne),
 
     # TSne.jl
     Spec("TSne.jl", log_name="TSne-jl", cores=1, lang="Julia", parser=parse_jl),
@@ -136,7 +139,7 @@ data = pd.DataFrame(data)
 
 
 def generate_benchmark_plot(colors, title, fname):
-    fig, ax = plt.subplots(figsize=(8, 4.5))
+    fig, ax = plt.subplots(figsize=(8, 6))
     sns.despine(offset=20)
 
     ax.set_title(
@@ -145,8 +148,8 @@ def generate_benchmark_plot(colors, title, fname):
         fontdict={"fontsize": "13"},
         pad=15,
     )
-    ax.set_xlabel("$N$ samples")
-    ax.set_ylabel("Time (min)")
+    ax.set_xlabel("Dataset size [samples]")
+    ax.set_ylabel("Time [min]")
 
     ax.grid(color="0.9", linestyle="--", linewidth=1)
 
@@ -190,13 +193,34 @@ def generate_benchmark_plot(colors, title, fname):
         lambda x, p: format(int(x), ",").replace(",", "."))
     )
 
-    legend_handle = ax.legend(
+    # The legend will only indicate solid/dashed lines for core count
+    # The line labels for the different implementations will be added later in AI
+    legend_patches = [
+        Line2D([0], [0], color="#000", lw=2, linestyle="solid"),
+        Line2D([0], [0], color="#000", lw=2, linestyle="dashed"),
+    ]
[…]

### 2022-04-08 paper/main.tex
Finish up implementation section

@@
 p_{j \mid i} = \frac{\exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{x}_j ) / \sigma_i^2 \right )}
-{\sum_{k \neq i } \exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{x}_k ) / \sigma_i^2 \right )}, \quad p_{i \mid i} = 0 \notag
-\label{eq:gaussian_kernel}
+{\sum_{k \neq i } \exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{x}_k ) / \sigma_i^2 \right )}, \quad p_{i \mid i} = 0
+\label{eq:p_ij}
 \end{equation}
@@
 
-One of t-SNE's long-standing criticisms is that it fails to preserve long-range distances but instead focuses on capturing the local structure of the data manifold~\citep{becht2019dimensionality}.
-Recently, several approaches have been proposed to improve the global organization of the resulting embeddings.
+One of t-SNE's long-standing criticisms is that it fails to preserve long-range
+distances but instead focuses on capturing the local structure of the data
+manifold~\citep{becht2019dimensionality}. Recently, several approaches have been
+proposed to improve the global organization of the resulting embeddings.
 
-\citet{kobak2019umap} showed that the initialization of force-directed layout algorithms, such as t-SNE, UMAP~\citep{2018arXivUMAP}, and ForceAtlas2~\citep{jacomy2014forceatlas2}, largely dictates the global consistency of resulting embeddings.
-When initialized randomly, clusters are often arbitrarily scattered in the embedding space.
-However, when using initialization schemes based on PCA or spectral embeddings, the clusters identified in the resulting embedding are typically grouped in a globally coherent manner.
+\citet{kobak2019umap} showed that the initialization of force-directed layout
+algorithms, such as t-SNE, UMAP~\citep{2018arXivUMAP}, and
+ForceAtlas2~\citep{jacomy2014forceatlas2}, largely dictates the global
+consistency of resulting embeddings. When initialized randomly, clusters are
+often arbitrarily scattered in the embedding space. However, when using
+initialization schemes based on PCA or spectral embeddings, the clusters
+identified in the resulting embedding are typically grouped in a globally
+coherent manner.
 
-In standard t-SNE, distances between data points are converted to similarities through the use of Gaussian kernels.
-The perplexity of these kernels governs the sizes of the local neighborhoods which are to be preserved.
-One easy way to improve the global consistency of resulting embeddings is to increase the sizes of these neighborhoods.
-This can be achieved by increasing the perplexity parameter.
-However, increasing perplexity often leads to loss in local sturcture and tends to obscure small clusters.
-\citet{kobak2019art} instead propose to use mixtures of Gaussians to preserve better short-range and long-range distances and, thus, achieve a better trade-off between local and global structure.
+In standard t-SNE, distances between data points are converted to similarities
+through the use of Gaussian kernels. The perplexity of these kernels governs the
+sizes of the local neighborhoods which are to be preserved. One easy way to
+improve the global consistency of resulting embeddings is to increase the sizes
+of these neighborhoods. This can be achieved by increasing the perplexity
+parameter. However, increasing perplexity often leads to loss in local sturcture
+and tends to obscure small clusters. \citet{kobak2019art} instead propose to use
+mixtures of Gaussians to preserve better short-range and long-range distances
+and, thus, achieve a better trade-off between local and global structure.
 
-Embeddings produced by standard t-SNE often use all available space and separate clusters by only thin boundaries.
-When working with large data sets, this often obscures the global relationships between clusters as all neighboring groups appear equidistant from one another.
-Other dimensionality reduction methods, such as UMAP and ForceAtlas2, produce embeddings where clusters appear more compact, and the white space separating the clusters may be interpreted as a loose measure of distance.
-Recently, \citet{bohm2020unifying} showed that the exaggeration factor $\rho$ could be used to produce layouts more similar to UMAP and ForceAtlas2.
-By incorporating exaggeration into later phases of the optimization, t-SNE introduces white space between clusters, which may better reflect the global relations between clusters.
+Embeddings produced by standard t-SNE often use all available space and separate
+clusters by only thin boundaries. When working with large data sets, this often
+obscures the global relationships between clusters as all neighboring groups
+appear equidistant from one another.  Other dimensionality reduction methods,
+such as UMAP and ForceAtlas2, produce embeddings where clusters appear more
+compact, and the white space separating the clusters may be interpreted as a
+loose measure of distance. Recently, \citet{bohm2020unifying} showed that the
+exaggeration factor $\rho$ could be used to produce layouts more similar to UMAP
+and ForceAtlas2. By incorporating exaggeration into later phases of the
+optimization, t-SNE introduces white space between clusters, which may better
+reflect the global relations between clusters.
 
 Standard t-SNE reveals the clustering structure at a single level of resolution.
-While the perplexity parameter can be used to control the trade-off between local and global structure, this can be time-consuming, and small, well-defined clusters can be missed.
-Alternatively, \citet{kobak2019heavy} suggest that varying the degrees of freedom in the $t$-distribution can be used to explore the clustering structure at different levels of resolution.
-Modifying Eqn.~(\ref{eq:cauchy_kernel}) to $q_{ij} \propto \left ( 1 + || \mathbf{y}_i - \mathbf{y}_j ||^2 / \alpha \right )^{-\alpha}$ allows us to use heavier-tailed distributions to model distances in the embedding space, which acts to highlight small subgroups in the resulting visualization.
+While the perplexity parameter can be used to control the trade-off between
+local and global structure, this can be time-consuming, and small, well-defined
+clusters can be missed. Alternatively, \citet{kobak2019heavy} suggest that
+varying the degrees of freedom in the $t$-distribution can be used to explore
+the clustering structure at different levels of resolution. Modifying
+Eqn.~(\ref{eq:cauchy_kernel}) to $q_{ij} \propto \left ( 1 + || \mathbf{y}_i -
+\mathbf{y}_j ||^2 / \alpha \right )^{-\alpha}$ allows us to use heavier-tailed
+distributions to model distances in the embedding space, which acts to highlight
+small subgroups in the resulting visualization.
 
@@
 
-We introduce \opentsne, a comprehensive Python library that implements the
-t-SNE algorithm and its numerous recentl extensions. \opentsne\ is
-designed to accomodate all types of users and includes both a simple API that
-allows users to quickly obtain informative t-SNE visualizations, as well as a
-more flexible API that allows seasoned practitioners to modify all the
+We introduce \opentsne, a comprehensive Python library that implements the t-SNE algorithm and its numerous recently proposed extensions.
+\opentsne\ is designed to accommodate all types of users and includes both a simple API that allows users to quickly obtain informative t-SNE visualizations, as well as a more flexible API that allows seasoned practitioners to modify all the
 different aspects of the t-SNE algorithm.
 
-The t-SNE algorithm is comprised of three main steps, mirrored in the structure
-and flow of the \opentsne\ library. These are: the initial positions
-of points in the embedding, an affinity model describing the similarities
-between data points in the embedding, and the optimization phase which
-optimizes the embedding layout. Using the advanced API, each of these key
-elements can be modified or tuned according to produce the best possible
-visualization.
+The core t-SNE algorithm is comprised of three main steps, mirrored in the
+structure of the \opentsne\ library:
 
-\begin{figure*}[ht]
+\begin{enumerate}
+\item Firstly, an affinity model describing the similarities between data points must be specified.
+Standard t-SNE predominantly deals with tabular data, and similarities are computed using Eqn.~\ref{eq:p_ij}.
+However, \citet{kobak2019art} showed that using mixtures of Gaussians instead of a single Gaussian often reveals better global consistency of the resulting visualization (albeit at some computational cost).
+Alternatively, one can speed up the overall computation by choosing a uniform kernel, which yields quantitatively similar results to standard t-SNE.
+\code{openTSNE.affinity} provides several affinity models to choose from but defaults to the standard perplexity-based model proposed in the original publication.
+
+Furthermore, it is straightforward to define and use a custom affinity model.
+For instance, one can easily extend the \code{openTSNE.affinity.Affinities} class and provide an affinity model for relational data using a normalized adjacency matrix.
+This enables rapid experimentation and opens the doors to the visualization of non-tabular data.
+
+\item Secondly, initial positions must be chosen for each data point in the embedding space.
+These can either be chosen randomly or can be used to impose some initial global structure on the embedding, e.g., using a PCA-based initialization.
+Accordingly, the \code{openTSNE.initialization} module provides several possible initialization schemes, or the user can also specify their own initialization by providing a \code{numpy.ndarray} object.
+By default, \opentsne\ opts for PCA-based initialization.
+
+\item Finally, equipped with an affinity model and initial point positions, the only remaining step is to optimize the embedding such that the positions of the points in the embedding space reflect their corresponding similarities in the affinity model.
+In \opentsne, the initialization and affinities are used to construct an \code{openTSNE.TSNEEmbedding} object, which can then be optimized using different parameter settings.
+Once an embedding object is constructed, it can subsequently be optimized via the \code{.optimize} method.
+
+Optimization can be performed using either the Barnes-Hut or FIt-SNE approximation scheme.
+While the FIt-SNE approximation scheme scales linearly with the number of samples, it introduces additional computational overhead, which is often unnecessary for smaller data sets.
+For smaller data sets, this sometimes results in longer runtimes than when using the Barnes-hut approximation scheme.
+By default, \opentsne\ uses a heuristic to ensure the fastest algorithm is chosen for the given data set, but this setting can also be specified by the user.
+\end{enumerate}
+
+Once the embedding is constructed and optimized, it is ready to use for any downstream tasks.
+\code{openTSNE.TSNEEmbedding} objects inherit from \code{numpy.ndarray}, which enables easy manipulation and provides interoperability with the broader \proglang{Python} ecosystem.
+
+Embedding new data points to an existing \code{openTSNE.TSNEEmbedding}
+follows the same three general steps as constructing a standard embedding:
+\begin{enumerate}
+\item Firstly, an affinity model must be provided describing the similarities between each new data point and the data points in the reference embedding.
+As before, \opentsne\ provides several possible affinity models in the \code{openTSNE.affinity} module.
+Most often, however, we want to utilize the same affinity model that was used to construct the reference embedding, as this leads to more spatially consistent embeddings.
+This is the default behavior of \opentsne.
+
+\item Secondly, the initial positions of the new data points in the embedding space must be specified.
+These can, again, be initialized randomly or be informed by their similarity to the data points in the reference embedding.
+As before, \opentsne\ provides several initialization options in the \code{openTSNE.initialization} module, but the user is free to provide their own initialization.
+
+\item Finally, what remains is to construct an embedding object, which can then be subject to optimization.
+Because this new embedding relies on an existing reference embedding, we construct a \code{openTSNE.PartialTSNEEmbedding} embedding object, which indicates its dependence on a reference \code{openTSNE.TSNEEmbedding} object.
+As before, \code{openTSNE.PartialTSNEEmbedding} inherits from \code{numpy.ndarray} and can therefore be easily manipulated.
+Optimization is, again, performed via the \code{.optimize} method.
+\end{enumerate}
+
+\begin{figure*}[t]
   \centering
@@
   \caption{\label{fig:diagram}
-  Bla
+  The high-level structure and flow of the \opentsne\ library.
   }
@@
 
-\opentsne\ is compatible with the Python data science ecosystem (e.g.
-\pkg{numpy}, \pkg{scikit-learn}, \pkg{scanpy}), providing a familiar and intuitive API to end-users.
-Its modular design encourages extensibility and experimentation with various parameter settings and changes to the analysis pipeline.
-\opentsne\ aims to make the latest theoretical advances in t-SNE accessible to the wider data science community.
-These features are easily accessible through an intuitive API, which closely follows the style \pkg{scikit-learn}~\citep{sklearn_api}, a popular and widely used toolkit in the machine learning community.
-The following snippet exemplifies this interface and demonstrates how \opentsne\ can be used to quickly generate several visualizations, each highlighting different aspects of the underlying data structure.
+Fig.~\ref{fig:diagram} shows the high-level structure and flow of the \opentsne\ library.
+
+Through different parameters, the user is given full control over every aspect of the t-SNE visualization algorithm, from different ways to define affinities to the delicate tuning of optimization parameters.
+This allows for rapid experimentation and enables users to quickly and efficiently inspect their data from numerous angles and at varying levels of resolution.
+
+While \opentsne\ allows for an unprecedented level of control at every step, most end-users often only need a quick and easy way to create informative visualizations of their data sets.
+\opentsne\ provides a high-level API that closely follows the style of \pkg{scikit-learn}~\citep{sklearn_api}, a popular and widely used toolkit in the machine learning community.
+This allows users of all levels to quickly pick up and make use of the \opentsne\ library.
 
 \subsection{Example}
+\label{sec:iris}
 
@@
 
-\opentsne\ is designed to be as user-friendly as possible and can be installed through either \pkg{PyPI} using \pkg{pip}
+\opentsne\ is designed to be easy to use and can be installed through either \pkg{PyPI} using \pkg{pip}
 \begin{CodeChunk}
@@
 \end{CodeChunk}
-\opentsne\ provides precompiled binaries for all major platforms (Windows, MacOS, Linux) for all currently supported Python versions (3.6---3.10).
-This is especially convenient for Windows users, as this obviates the need for a C/C++ compiler, which do not come bundled with the operating system.
+\opentsne\ provides precompiled binaries for all major platforms (Windows, MacOS, Linux) for all currently supported \proglang{Python} versions (3.7---3.10).
+This is especially convenient for Windows users, as this obviates the need for a \proglang{C}/\proglang{C++} compiler, which is typically not bundled with the operating system.
 
@@
 
-We demonstrate how to use \opentsne\ on the Iris data set from \citet{anderson1936species}.
-This small data set describes 150 iris flowers containing four variables (sepal
-length, sepal width, petal length, petal width) and their species (Iris Setosa,
-Iris Versicolor, Iris Virginica), which is the target variable.
-This data set is included in \pkg{scikit-learn}, making it easy to load:
+We demonstrate the simplest usage of \opentsne\ on the Iris data set from \citet{anderson1936species}.
+This small data set describes 150 iris flowers containing four variables (sepal length, sepal width, petal length, petal width) and their species (Iris Setosa, Iris Versicolor, Iris Virginica), which is the target variable.
+This data set is included in \pkg{scikit-learn}:
 \begin{CodeChunk}
@@
 \end{CodeChunk}
-\opentsne\ returns an \code{TSNEEmbedding} object, which is a subclass of
-\code{np.ndarray}. This allows users to easily manipulate the resulting embedding.
-The additional functionality provided by the \code{TSNEEmbedding} object allows
-the embedding to be further optimized using different parameter values, or be
-used as a reference when obtaining new data points.
-For users who do not need this additional functionality, this object can be
-converted to a standard \code{np.ndarray} with:
+Fig.~\ref{fig:iris} shows a visualization of the resulting t-SNE visualization.
+As expected, t-SNE is correctly able to group together samples from the same species.
+Additionally, the t-SNE embedding reveals that Iris Setosa is markedly different from the other two species, while flowers from Iris Versicolor and Iris Virginica may appear more similar to one another.
+\begin{figure*}[htbp]
+  \centering
+  \includegraphics[width=0.5\textwidth]{iris}
+  \caption{\label{fig:iris}
+  A visualization of the obtained t-SNE embedding on the Iris data set. The colors
+  correspond to different species of the Iris flower.
+}
+\end{figure*}
+We can now use this embedding as a reference for new data points. In this simple example, we will duplicate some of the original samples and add them into the embedding.
 \begin{CodeChunk}
 \begin{CodeInput}
->>> import numpy as np
+>>> new_data = iris.data[::3]
+>>> new_embedding = embedding.transform(new_data)
+>>> new_embedding[:3]
+PartialTSNEEmbedding([[-0.02637324, 15.69956674],
+                      [-2.84200792, 16.15478451],
+                      [-2.4594136 , 15.37910876]])
+\end{CodeInput}
+\end{CodeChunk}
+Fig.~\ref{fig:iris-transform} shows the resulting positions of the new data points, overlaid the original, reference embedding.
+As expected, t-SNE is correctly able to place the new data points to their corresponding groups.
+\begin{figure*}[htbp]
+  \centering
+  \includegraphics[width=0.5\textwidth]{iris-transform}
+  \caption{\label{fig:iris-transform}
+  A visualization of the original, reference data set (transparent points) overlaid with the new data points (opaque points).
+}
+\end{figure*}
+
+\subsubsection{Peeking under the hood}
+
+While the above example offers users a quick and easy way to generate informative visualizations, \opentsne\ allows us to control the t-SNE algorithm at every step.
+To demonstrate the flexibility of \opentsne, we will repeat the same analysis of the Iris data set as before, this time leveraging the more advanced features of the library.
+More practical examples of advanced usage will be shown in Section~\ref{sec:discussion}.
+
+We, again, begin by loading the Iris data set using \pkg{scikit-learn}:
+\begin{CodeChunk}
+\begin{CodeInput}
+>>> from sklearn import datasets
 >>>
->>> X_tsne = embedding.view(np.ndarray)
->>> X_tsne[:3]
-array([[-0.55581917, 17.56162344],
+>>> iris = datasets.load_iris()
+>>> iris.data[:3]
+array([[5.1, 3.5, 1.4, 0.2],
+       [4.9, 3. , 1.4, 0.2],
+       [4.7, 3.2, 1.3, 0.2]])
+\end{CodeInput}
+\end{CodeChunk}
+Next, we import the \opentsne\ library:
+\begin{CodeChunk}
+\begin{CodeInput}
+>>> import openTSNE
+\end{CodeInput}
+\end{CodeChunk}
+We now mirror the three steps listed above. First, we select an affinity model which describes the similarities between data points:
+\begin{CodeChunk}
+\begin{CodeInput}
+>>> affinities = openTSNE.affinity.PerplexityBasedNN(iris.data, perplexity=30)
+\end{CodeInput}
+\end{CodeChunk}
+Next, we determine the initial point positions in the embedding:
+\begin{CodeChunk}
+\begin{CodeInput}
+>>> initialization = openTSNE.initialization.pca(iris.data)
+>>> initialization[:3]
+array([[-1.30971087e-04,  1.55848907e-05],
+       [-1.32435711e-04, -8.63672049e-06],
+       [-1.40967409e-04, -7.07276279e-06]])
+\end{CodeInput}
+\end{CodeChunk}
+Next, using the affinity model and initialization, we can create an embedding object:
+\begin{CodeChunk}
+\begin{CodeInput}
+>>> embedding = openTSNE.TSNEEmbedding(initialization, affinities)
+>>> embedding[:3]
+TSNEEmbedding([[-1.30971087e-04,  1.55848907e-05],
+               [-1.32435711e-04, -8.63672049e-06],
+               [-1.40967409e-04, -7.07276279e-06]])
+\end{CodeInput}
+\end{CodeChunk}
+Notice that the embedding object contains the same point coordinates as the initialization.
+This is because although we have created an \code{openTSNE.TSNEEmbedding} object, we have not yet run any optimization.
+To optimize the embedding, we merely have to call the \code{embedding.optimize} method using our desired parameter settings.
+Here, we use the default parameter settings, including a shorter early-exaggeration phase, followed by a longer phase with no exaggeration.
+\begin{CodeChunk}
+\begin{CodeInput}
+>>> embedding.optimize(250, exaggeration=12, momentum=0.5, inplace=True)
+>>> embedding.optimize(500, momentum=0.8, inplace=True)
+>>> embedding[:3]
+TSNEEmbedding([[-0.55581917, 17.56162344],
                [-1.76742536, 20.07763453],
@@
 \end{CodeChunk}
-Fig.~\ref{fig:iris} shows a visualization of the resulting t-SNE visualization.
-As expected, t-SNE is correctly able to group together samples from the same
-species. Additionally, the t-SNE embedding reveals that Iris Setosa is markedly
-different from the other two species, while flowers from Iris Versicolor and
-Iris Virginica may appear more similar to one another.
-\begin{figure*}[htbp]
-  \centering
-  \includegraphics[width=0.5\textwidth]{iris}
-  \caption{\label{fig:iris}
-  Visualization of the obtained t-SNE embedding on the Iris data set. The colors
-  correspond to different species of the Iris flower.
-}
-\end{figure*}
+We can validate that method produces the exact same embedding as before:
+\begin{CodeChunk}
+\begin{CodeInput}
+>>> np.linalg.norm(openTSNE.TSNE().fit(iris.data) - embedding)
+0.0
+\end{CodeInput}
+\end{CodeChunk}
+Next, we can add new samples into the embedding:
+\begin{CodeChunk}
+\begin{CodeInput}
+>>> new_embedding = embedding.prepare_partial(iris.data[::3], perplexity=5)
+>>> new_embedding.optimize(
+...     exaggeration=1.5, n_iter=250, learning_rate=0.1, momentum=0.8,
+...     max_grad_norm=0.25, inplace=True
+... )
+>>> new_embedding[::3]
+PartialTSNEEmbedding([[-0.02637324, 15.69956674],
+                      [-2.84200792, 16.15478451],
+                      [-2.4594136 , 15.37910876]])
+\end{CodeInput}
+\end{CodeChunk}
+And validate that this also produces the exact same embedding as before:
+\begin{CodeChunk}
+\begin{CodeInput}
+>>> np.linalg.norm(new_embedding - embedding.transform(iris.data[::3]))
+0.0
+\end{CodeInput}
+\end{CodeChunk}
 
-\section{Discussion and Examples} \label{sec:summary}
+\section{Discussion and Examples} \label{sec:discussion}
 
@@
 
-Dimensionality reduction techniques implicitly assume that high-dimensional data lies on a lower-dimensional manifold, which a small number of dimensions can accurately capture.
-However, there is no evidence that every data set can accurately be described using only two dimensions, and any such embedding will inevitably lead to a loss of information.
-Thus, it is beneficial to examine multiple embeddings, each of which provides a different perspective on topology and other data characteristics.
+Dimensionality reduction techniques implicitly assume that high-dimensional data
+lies on a lower-dimensional manifold, which a small number of dimensions can
+accurately capture. However, there is no evidence that every data set can
+accurately be described using only two dimensions, and any such embedding will
+inevitably lead to a loss of information. Thus, it is beneficial to examine
+multiple embeddings, each of which provides a different perspective on topology
+and other data characteristics.
 
@@
 
-We illustrate this point by generating four different embeddings of the data on single-cell gene expression in mouse brain~\citep{tasic2018shared}.
-Fig.~\ref{fig:tasic}.a shows an embedding using default t-SNE parameters.
-We generate the four different embeddings using the following code snippet (which is shortened for clarity).
+We illustrate this point by generating four different embeddings of the data on
+single-cell gene expression in mouse brain~\citep{tasic2018shared}.
+Fig.~\ref{fig:tasic}.a shows an embedding using default t-SNE parameters. We generate
+the four different embeddings using the following code snippet (shortened for clarity).
 \begin{CodeChunk}
 \begin{CodeInput}
-embedding_a = TSNE(perplexity=30).fit(x)
-embedding_b = TSNE(perplexity=500).fit(x)
-embedding_c = TSNE(perplexity=[30, 500]).fit(x)
-embedding_d = TSNE(perplexity=30, dof=0.6).fit(x)
+embedding_a = openTSNE.TSNE(perplexity=30).fit(x)
+embedding_b = openTSNE.TSNE(perplexity=500).fit(x)
+embedding_c = openTSNE.TSNE(perplexity=[30, 500]).fit(x)
+embedding_d = openTSNE.TSNE(perplexity=30, dof=0.6).fit(x)
 \end{CodeInput}
@@
 
-When dealing with data containing millions of data points, standard t-SNE embeddings often become unwieldy -- cluster boundaries are blurred, large clusters absorb smaller ones, and relationships between clusters become increasingly difficult to interpret.
-We constructed Fig.~\ref{fig:cao}.a from the data containing expression profiles of over two million single cells captured at different time points in mouse development.
-The embedding indicates numerous clusters with transitions between time points, as marked by the color-coding, that are difficult to interpret.
-Kobak \& Berens observed that increasing attractive forces between similar data points controlled via the \textit{exaggeration} parameter leads to more compact clusters, and subsequently, more informative visualizations~\citep{kobak2019art}.
-For instance, \ref{fig:cao}.b doubles the default exaggeration, which uncovers some of the data's overall structure.
-Further doubling the exaggeration in \ref{fig:cao}.c allows us to observe that the data is comprised of two main groups of cells and eight somewhat smaller clusters.
-The visualization also reveals several tiny clusters, possibly corresponding to rare cell types.
+When dealing with data containing millions of data points, standard t-SNE
+embeddings often become unwieldy -- cluster boundaries are blurred, large
+clusters absorb smaller ones, and relationships between clusters become
+increasingly difficult to interpret. We constructed Fig.~\ref{fig:cao}.a from
+the data containing expression profiles of over two million single cells
+captured at different time points in mouse development. The embedding indicates
+numerous clusters with transitions between time points, as marked by the
+color-coding, that are difficult to interpret. \citet{kobak2019art} that
+increasing attractive forces between similar data points controlled via
+the \textit{exaggeration} parameter leads to more compact clusters, and
+subsequently, more informative visualizations. For instance, \ref{fig:cao}.b
+doubles the default exaggeration, which uncovers some
+of the data's overall structure. Further doubling the exaggeration in
+\ref{fig:cao}.c allows us to observe that the data is comprised of two main groups
+of cells and eight somewhat smaller clusters. The visualization also reveals
+several tiny clusters, possibly corresponding to rare cell types.
 \begin{CodeChunk}
@@
 # emphasize the high level structure
-init_sample = TSNE(perplexity=500).fit(x_sample)
+init_sample = openTSNE.TSNE(perplexity=500).fit(x_sample)
 
@@
 \end{CodeChunk}
-We generate the three different embeddings using the following code snippet (which is shortened for clarity).
+We generate the three different embeddings using the following code snippet
+(shortened for clarity).
 \begin{CodeChunk}
@@
 embeddings produced by t-SNE with exaggeration values of 4 and $\sim30$
-construct embeddings which are markedly similar to UMAP and ForceAtlas2,
+construct embeddings that are markedly similar to UMAP and ForceAtlas2,
 respectively. For example, in Fig.~\ref{fig:cao}.a, the developmental trajectory
@@
 experiment into the reference landscape. The reference data from
-\citet{macosko2015highly} contains gene expression
-profiles from mouse retinal cells. By embedding the samples from a similar
-experiment on bipolar retinal cells by \citet{shekhar2016comprehensive}, we can correctly map the bipolar cell
-clusters onto the reference embedding. The following snippet shows the code necessary to generate figure~\ref{fig:macosko}.
+\citet{macosko2015highly} contains gene expression profiles from mouse retinal
+cells. By embedding the samples from a similar experiment on bipolar retinal
+cells by \citet{shekhar2016comprehensive}, we can correctly map the bipolar
+cell clusters onto the reference embedding. The following snippet shows the
+code necessary to generate figure~\ref{fig:macosko}.
 \begin{CodeChunk}
@@
 # why t-SNE misbehaves or fails to preserve global structure
-tsne_old = TSNE(initialization="random", learning_rate=200, n_iter=750).fit(x)
+tsne_old = openTSNE.TSNE(
+    initialization="random", learning_rate=200, n_iter=750
+).fit(x)
 
@@
 # number of iterations to 500, lowering the overall runtime
-tsne_standard = TSNE(perplexity=[50, 500]).fit(x)
+tsne_standard = openTSNE.TSNE(perplexity=[50, 500]).fit(x)
 
@@
 Fig.~\ref{fig:transform}, we construct a reference embedding using labeled data
-from \citet{hochgerner2018conserved} containing
-gene-expression profiles of cells from the mouse brain. The authors assign a
-type to each cell. We can verify their classification accuracy by visualizing
-the expression of well-established gene markers for the major cell types. We
-then embed cells from \citet{harris2018classes} into the
-constructed cell atlas. In \citet{harris2018classes}, labels are provided only for
-neuronal cells. In the resulting mapping, we can quickly identify other
-non-neuronal cell types, including oligodendrocytes and astrocytes. We can
-further use marker genes to validate that the mapping in the reference landscape
-is correct.
+from \citet{hochgerner2018conserved} containing gene-expression profiles of
+cells from the mouse brain. The authors assign a type to each cell. We can
+verify their classification accuracy by visualizing the expression of
+well-established gene markers for the major cell types. We then embed cells
+from \citet{harris2018classes} into the constructed cell atlas. In
+\citet{harris2018classes}, labels are provided only for neuronal cells. In the
+resulting mapping, we can quickly identify other non-neuronal cell types,
+including oligodendrocytes and astrocytes. We can further use marker genes to
+validate that the mapping in the reference landscape is correct.
 
@@
 embeddings on non-tabular relational data: the only requirement imposed by the
-affinity-model is some notion of similarity between data points. Finally,
+affinity model is some notion of similarity between data points. Finally,
 \opentsne's comprehensive callback system can be utilized to monitor and adapt
@@
 % General problem statement
-One of the t-SNE's common criticisms is limited scalability to large data sets
-containing, for instance, millions of data points~\citep{becht2019dimensionality}.
-The culprit for slow response time stems from particular optimization procedures
-and their specific implementation in popular
-open-source libraries. Until quite recently, most popular implementations of
-t-SNE were based on either a direct implementation of the t-SNE algorithm with
-asymptotic time complexity $\mathcal{O}(N^2)$~\citep{maaten2008visualizing}, or the Barnes-Hut approximation
-scheme developed by \citet{van2014accelerating} with asymptotic time complexity
-$\mathcal{O}(N \log N)$. Recently, however, \citet{linderman2019fast} developed
-a novel approximation scheme -- FIt-SNE -- which reduces the asymptotic time
-complexity to $\mathcal{O}(N)$, overcoming the difficulties of scaling t-SNE
-to large data sets.
+One of the t-SNE's common criticisms is limited scalability to large data sets containing, for instance, millions of data points~\citep {becht2019dimensionality}.
+The culprit for slow response time stems from particular optimization procedures and their specific implementation in popular open-source libraries.
+Until quite recently, most popular implementations of t-SNE were based on either a direct implementation of the t-SNE algorithm with asymptotic time complexity $\mathcal{O}(N^2)$~\citep{maaten2008visualizing}, or the Barnes-Hut approximation scheme developed by \citet{van2014accelerating} with asymptotic time complexity $\mathcal{O}(N \log N)$.
+Recently, however, \citet{linderman2019fast} developed a novel approximation scheme -- FIt-SNE -- which reduces the asymptotic time complexity to $\mathcal {O}(N)$, overcoming the difficulties of scaling t-SNE to large data sets.
 
 % We benchmark Python
-Fig.~\ref{fig:benchmarks_py} benchmarks four popular Python t-SNE
+Fig.~\ref{fig:benchmarks_py} benchmarks four popular Python~\citep{vanrossum1995python} t-SNE
 implementations, including \pkg{scikit-learn} (v1.0.1), \pkg{MulticoreTSNE}
@@
 profiles. We preprocess the data using the standard single-cell analysis
-pipeline~\citep{XXX} and extract the top 50 principal components. To generate
+pipeline~\citep{xxx} and extract the top 50 principal components. To generate
 benchmark data sets of different sample sizes, we subsample the data set five
@@
 % Problem with Python
-In Python, the most widely-used implementation of t-SNE comes from \pkg{scikit-learn},
-which exhibits long runtimes when compared to other Python implementations.
-\pkg{scikit-learn}, like its C++ counterpart -- \pkg{MulticoreTSNE}~\citep{Ulyanov2016} ,
-is based on the Barnes-Hut approximation scheme. However, as
-shown on Fig.~\ref{fig:benchmarks_py}, \pkg{scikit-learn} exhibits much longer
-runtimes than \pkg{MulticoreTSNE}.
-Newer implementations, such as \pkg{FIt-SNE} and \opentsne, both implement the
-FIt-SNE approximation scheme, making them suitable for the visualization of
-millions of data points. While both libraries implement the same
-algorithm, \opentsne\ emphasizes ease of use and extensibility, and is primarily
-written in Python, which inevitably introduces some runtime overhead. This
-makes \opentsne\ marginally slower than its pure C++ counterpart -- \pkg{FIt-SNE}.
-However, these differences disappear almost entirely when
-utilizing multi-threading, as shown on Fig.~\ref{fig:benchmarks_py}.
-Additionally, \opentsne\ provides a flexible API, allowing the user to split up
-the embedding-construction process into several parts. This allows us to cache
-and reuse intermediate results, enabling users to quickly experiment with
-different parameter settings and iterate on their final visualizations. This
-kind of functionality is not supported in other implementations, which require
-the user to start from scratch for every embedding, resulting in large amounts
-of redundant computation.
+In Python, the most widely-used implementation of t-SNE comes from \pkg{scikit-learn}, which exhibits long runtimes when compared to other Python implementations.
+\pkg{scikit-learn}, like its \proglang{C++} counterpart -- \pkg{MulticoreTSNE}~\citep{Ulyanov2016} , is based on the Barnes-Hut approximation scheme.
+However, as shown on Fig.~\ref{fig:benchmarks_py}, \pkg{scikit-learn} exhibits much longer runtimes than \pkg{MulticoreTSNE}.
+Newer implementations, such as \pkg{FIt-SNE} and \opentsne, both implement the FIt-SNE approximation scheme, making them suitable for the visualization of millions of data points.
+While both libraries implement the same algorithm, \opentsne\ emphasizes ease of use and extensibility and is primarily written in Python, which inevitably introduces some runtime overhead.
+This makes \opentsne\ marginally slower than its pure \proglang{C++} counterpart -- \pkg{FIt-SNE}.
+However, these differences disappear almost entirely when utilizing multi-threading, as shown in Fig.~\ref{fig:benchmarks_py}.
+Additionally, \opentsne\ provides a flexible API, allowing the user to split up the embedding-construction process into several parts.
+This allows us to cache and reuse intermediate results, enabling users to quickly experiment with different parameter settings and iterate on their final visualizations.
+This kind of functionality is not supported in other implementations, which require the user to start from scratch for every embedding, resulting in large amounts of redundant computation.
 
 % Problem with other programming languages
-In other programming languages, the selection of fast, open-source
-implementations of t-SNE is more limited. In R, the most widely used
-implementation of t-SNE -- \pkg{Rtsne}~\citep{krijthe2015rtsne} -- is based on
-the Barnes-Hut approximation, and is closely comparable to 
-\pkg{MulticoreTSNE}. In Julia, the most popular implementation -- 
-\pkg{TSne.jl}~\citep{julia_tsne} -- is based on a naive $\mathcal{O}(N^2)$ implementation of the
-algorithm, which allows it to be used only on smaller data sets containing up
-to thousands of data points. We benchmark \pkg{Rtsne} (v0.15) and 
-\pkg{TSne.jl} (v1.3.0) and compare it to \opentsne\ (v0.6.2).
-Fig.~\ref{fig:benchmarks_lang} illustrates the scaling of each approximation 
-scheme. Of these implementations, only \opentsne\ is able to scale to data sets
-containing millions of data points in a reasonable amount of time.
+In other programming languages, the selection of fast, open-source implementations of t-SNE is more limited.
+In \proglang{R}~\citep{team2013r}, the most widely used implementation of t-SNE -- \pkg{Rtsne}~\citep{krijthe2015rtsne} -- is based on the Barnes-Hut approximation, and is closely comparable to \pkg{MulticoreTSNE}.
+In \proglang{Julia}~\citep{bezanson2017julia}, the most popular implementation -- \pkg{TSne.jl}~\citep{julia_tsne} -- is based on a naive $\mathcal{O}(N^2)$ implementation of the algorithm, which allows it to be used only on smaller data sets containing up to thousands of data points.
+We benchmark \pkg{Rtsne} (v0.15) and  \pkg{TSne.jl} (v1.3.0) and compare it to \opentsne\ (v0.6.2).
+Fig.~\ref{fig:benchmarks_lang} illustrates the scaling of each approximation scheme.
+Of these implementations, only \opentsne\ is able to scale to data sets containing millions of data points in a reasonable amount of time.
 
@@
   We select the fastest available, open-source implementations of the t-SNE
-  algorithm from the Python, R, and Julia programming languages. We choose
+  algorithm from the \proglang{Python}, \proglang{R}, and \proglang{Julia}
+  programming languages. We choose
   \opentsne\ (v0.6.2) for Python, \pkg{Rtsne}~\citep{krijthe2015rtsne} (v0.15)
-  for R, and \pkg{TSne.jl}~\citep{julia_tsne} (v1.3.0) for Julia. The benchmarks
-  reflect the time complexity of the
+  for \proglang{R}, and \pkg{TSne.jl}~\citep{julia_tsne} (v1.3.0) for
+  \proglang{Julia}. The benchmarks reflect the time complexity of the
   implemented approximation schemes. For instance, \pkg{TSne.jl} implements a
@@
   containing over 500,000 data points due to lack of stack memory. We found
-  these crashes to be inconsistent and somwewhat random. \opentsne\ is the only
+  these crashes to be inconsistent and somewhat random. \opentsne\ is the only
   library implementing the more recent FIt-SNE approximation scheme, which makes
@@
 proves prohibitively slow for even moderately-sized data sets that span tens of
-thousands of data records. Other C++ implementations such as \pkg{MulticoreTSNE}
+thousands of data records. Other \proglang{C++} implementations such as \pkg{MulticoreTSNE}
 and \pkg{FIt-SNE} exhibit better scaling in more massive data sets, but do not
@@
 themselves. This problem is critical, for instance, for users of the Windows
-operating system, where the C++ compiler does not come with the system, making
+operating system, where the \proglang{C++} compiler does not come with the system, making
 the correct configuration of current t-SNE implementations cumbersome.
@@
 points. While the Python virtual machine inevitably introduces some performance
-overhead, the runtime is comparable to its C++ counterpart -- \pkg{FIt-SNE}.
+overhead, the runtime is comparable to its \proglang{C++} counterpart -- \pkg{FIt-SNE}.
 Finally, \opentsne\ is extensible. Its modular design enables researchers to
@@
 custom components into the software. We provide an extended list of features and
-comparison to other popular t-SNE implementations in Table~\ref{tab:features}.
+comparison to other popular t-SNE implementations in Tab.~\ref{tab:features}.
 
@@
   \pkg{BHTSNE}~(master), \pkg{MulticoreTSNE}~(v0.1), and
-  \pkg{FIt-SNE}~(v1.1.0), as well as the major implementations in R and Julia,
-  \pkg{Rtsne}~(v0.15), \pkg{Tsne.jl}~(v1.3.0). Active maintenance indicates whether the project has had any meaningful development in the past year from the time of writing.
+  \pkg{FIt-SNE}~(v1.1.0), as well as the major implementations in \proglang{R}
+  and \proglang{Julia}, \pkg{Rtsne}~(v0.15), \pkg{Tsne.jl}~(v1.3.0). Active
+  maintenance indicates whether the project has had any meaningful development
+  in the past year from the time of writing.
 }
@@
 \section{Conclusion}
-We introduce \opentsne, the most feature-complete, open-source, Python implementation of the widely popular t-SNE visualization algorithm.
-The library incorporates the latest developments to the t-SNE algorithm and makes them easily accessible through a familiar API.
-\opentsne\ implements efficient approximation schemes, making it suitable for the visualization of large data sets containing up to millions of data points.
-Finally, \opentsne\ is easy to install and comes with precompiled binaries available for all major platforms.
+
+We introduce \opentsne, the most feature-complete, open-source, Python
+implementation of the widely popular t-SNE visualization algorithm. The library
+incorporates the latest developments to the t-SNE algorithm and makes them
+easily accessible through a familiar API. \opentsne\ implements efficient
+approximation schemes, making it suitable for the visualization of large data
+sets containing up to millions of data points. Finally, \opentsne\ is easy to
+install and comes with precompiled binaries available for all major platforms.
 
@@
 
-We want to thank Dmitry Kobak for helpful discussions regarding t-SNE and
-his contributions to the source code. We would also like to thank George
-Linderman for his help with the FIt-SNE algorithm. This work was supported by
-Slovenian Research Agency (P2-0209).
+We want to thank Dmitry Kobak for helpful discussions regarding t-SNE and his
+contributions to the source code. We would also like to thank George Linderman
+for his help with the FIt-SNE algorithm. This work was supported by Slovenian
+Research Agency (P2-0209).
 

### 2022-04-08 figures/fig3-iris.py
Add iris-transform script

@@ -20,6 +20,7 @@ iris = datasets.load_iris()
 
 print("Creating embedding...")
 embedding = openTSNE.TSNE().fit(iris.data)
+new_embedding = embedding.transform(iris.data[::3])
 
 print("Generating figure...")
 target = []
@@ -28,6 +29,7 @@ for t in iris.target_names[iris.target]:
 
 target = np.array(target)
 
+# Original embedding
 fig, ax = plt.subplots(figsize=(5, 5))
 utils.plot(embedding, target, s=64, alpha=0.75, ax=ax)
 ax.set(xlabel="t-SNE 1", ylabel="t-SNE 2")
@@ -44,3 +46,24 @@ plt.savefig(
     bbox_inches="tight",
     transparent=True,
 )
+plt.close()
+
+# Transform embedding
+fig, ax = plt.subplots(figsize=(5, 5))
+utils.plot(embedding, target, s=32, alpha=0.25, ax=ax)
+utils.plot(new_embedding, target[::3], s=64, alpha=1, ax=ax)
+ax.set(xlabel="t-SNE 1", ylabel="t-SNE 2")
+ax.axis("on")
+plt.savefig(
+    os.path.join(args.out_dir, "iris-transform.pdf"),
+    dpi=72,
+    bbox_inches="tight",
+    transparent=True,
+)
+plt.savefig(
+    os.path.join(args.out_dir, "iris-transform.png"),
+    dpi=72,
+    bbox_inches="tight",
+    transparent=True,
+)
+plt.close()

### 2022-04-08 figures/generate_figures.sh

@@ -29,11 +29,11 @@ python fig1-macosko.py -i "$H5AD_DIR" -o "$FIGURES_DIR"
 echo "Generating Figure 3..."
 python fig3-iris.py -o "$FIGURES_DIR"
 
-echo "Generating Figure 4..."
-python fig4-tasic.py -i "$H5AD_DIR" -o "$FIGURES_DIR"
-
 echo "Generating Figure 5..."
-python fig5-cao.py -i "$H5AD_DIR" -o "$FIGURES_DIR"
+python fig5-tasic.py -i "$H5AD_DIR" -o "$FIGURES_DIR"
 
 echo "Generating Figure 6..."
-python fig6-hochgerner.py -i "$H5AD_DIR" -o "$FIGURES_DIR"
+python fig6-cao.py -i "$H5AD_DIR" -o "$FIGURES_DIR"
+
+echo "Generating Figure 7..."
+python fig7-hochgerner.py -i "$H5AD_DIR" -o "$FIGURES_DIR"

### 2022-04-08 paper/main.tex
Better integrate the code snippets into the text

@@
 Recently, several extensions have been proposed to address scalability issues and the quality of the resulting visualizations.
-We introduce \opentsne, a modular Python library that implements the core t-SNE algorithm and its many extensions.
+We introduce \opentsne, a modular \proglang{Python} library that implements the core t-SNE algorithm and its many extensions.
 The library is fast and can treat data sets containing millions of data points.
@@
 Despite the many theoretical advancements, most popular t-SNE libraries have been slow to incorporate them into their implementations.
-In this paper, we present \opentsne, our open-source Python implementation of t-SNE, complete with its recently proposed extensions.
-\opentsne\ is easy to install and includes precompiled binaries available through popular Python package managers.
+In this paper, we present \opentsne, our open-source \proglang{Python} implementation of t-SNE, complete with its recently proposed extensions.
+\opentsne\ is easy to install and includes precompiled binaries available through popular \proglang{Python} package managers.
 It provides a familiar API, making it suitable as a drop-in replacement for existing t-SNE implementations.
@@
 
-We introduce \opentsne, a comprehensive Python library that implements the t-SNE algorithm and its numerous recently proposed extensions.
+We introduce \opentsne, a comprehensive \proglang{Python} library that implements the t-SNE algorithm and its numerous recently proposed extensions.
 \opentsne\ is designed to accommodate all types of users and includes both a simple API that allows users to quickly obtain informative t-SNE visualizations, as well as a more flexible API that allows seasoned practitioners to modify all the
@@
 
-\section{Discussion and Examples} \label{sec:discussion}
+\section{Case Studies} \label{sec:discussion}
 
@@
 Dimensionality reduction techniques implicitly assume that high-dimensional data
-lies on a lower-dimensional manifold, which a small number of dimensions can
-accurately capture. However, there is no evidence that every data set can
+lies on a lower-dimensional manifold, which can accurately be captured by a small number of dimensions. However, there is no evidence that every data set can
 accurately be described using only two dimensions, and any such embedding will
@@
 We illustrate this point by generating four different embeddings of the data on
-single-cell gene expression in mouse brain~\citep{tasic2018shared}.
-Fig.~\ref{fig:tasic}.a shows an embedding using default t-SNE parameters. We generate
-the four different embeddings using the following code snippet (shortened for clarity).
+single-cell gene expression in mouse brain~\citep{tasic2018shared}. We generate
+the embeddings using the following code snippet (shortened for clarity):
 \begin{CodeChunk}
@@
 \end{CodeChunk}
+Fig.~\ref{fig:tasic}.a shows an embedding using default t-SNE parameters.
 While different clusters of excitatory and inhibitory neurons appear close to one another, all clusters appear equidistant from their neighbors, and the overall relations between groups are not obvious.
@@
 several tiny clusters, possibly corresponding to rare cell types.
-\begin{CodeChunk}
-\begin{CodeInput}
-# To improve the global structure of the final layout, we initialize the
-# embedding by creating a t-SNE plot on a sample of points
-
-# Take a random sample of the data points
-indices = np.random.permutation(list(range(x.shape[0])))
-reverse = np.argsort(indices)
-
-x_sample, x_rest = x[indices[:25000]], x[indices[25000:]]
-
-# Create a t-SNE embedding on the sample, using a high perplexity value to
-# emphasize the high level structure
-init_sample = openTSNE.TSNE(perplexity=500).fit(x_sample)
-
-# We can next use the `transform` functionality to determine starting positions
-# for the remaining data points. Since we don't actually want to optimize the
-# remaining points separately from the sample points, we don't need to call the
-# `transform` function. Instead, we'll call the `prepare_partial` function,
-# which `transform` calls anyways under the hood.
-init_rest = init_sample.prepare_partial(x_rest)
-
-# Lastly, we merge `init_sample` and `init_rest` and restore the original order
-init_full = np.vstack((init_sample, init_rest))[reverse]
-\end{CodeInput}
-\end{CodeChunk}
-We generate the three different embeddings using the following code snippet
-(shortened for clarity).
-\begin{CodeChunk}
-\begin{CodeInput}
-# We separately precompute the affinities as this can take quite a long time.
-# This way, we can reuse the precomputed affinities in various different
-# embeddings, without having to repeat the same, costly computation every time.
-affinities = openTSNE.affinity.PerplexityBasedNN(x, perplexity=30)
-
-# Create the TSNEEmbedding object using our previously calculated initialization
-# and precomputed affinities
-embedding = openTSNE.TSNEEmbedding(init_full, affinities)
-
-# Run the standard t-SNE early exaggeration phase
-embedding_ee =    embedding.optimize(n_iter=500, exaggeration=12, momentum=0.5)
-embedding_c  = embedding_ee.optimize(n_iter=500, exaggeration=4,  momentum=0.8)
-embedding_b  =  embedding_c.optimize(n_iter=500, exaggeration=2,  momentum=0.8)
-embedding_a  =  embedding_b.optimize(n_iter=500, exaggeration=1,  momentum=0.8)
-\end{CodeInput}
-\end{CodeChunk}
 
@@
 
+Figs.~\ref{fig:cao}.a-c were generated using the following code snippet, which demonstrates how we can take advantage of the more advanced features of \opentsne\ to more quickly create meaningful visualizations of massive data sets.
+We first take a sample of the data and use t-SNE to construct a globally consistent initialization, which will then be used to initialize the remaning data points:
+\begin{CodeChunk}
+\begin{CodeInput}
+# To improve the global structure of the final layout, we initialize the
+# embedding by creating a t-SNE plot on a sample of points
+
+# Take a random sample of the data points
+indices = np.random.permutation(list(range(x.shape[0])))
+reverse = np.argsort(indices)
+
+x_sample, x_rest = x[indices[:25000]], x[indices[25000:]]
+
+# Create a t-SNE embedding on the sample, using a high perplexity value to
+# emphasize the high level structure
+init_sample = openTSNE.TSNE(perplexity=500).fit(x_sample)
+
+# We can next use the `transform` functionality to determine starting positions
+# for the remaining data points. Since we don't actually want to optimize the
+# remaining points separately from the sample points, we don't need to call the
+# `transform` function. Instead, we'll call the `prepare_partial` function,
+# which `transform` calls anyways under the hood.
+init_rest = init_sample.prepare_partial(x_rest)
+
+# Lastly, we merge `init_sample` and `init_rest` and restore the original order
+init_full = np.vstack((init_sample, init_rest))[reverse]
+\end{CodeInput}
+\end{CodeChunk}
+We then generate the three different embeddings using the following code snippet
+(shortened for clarity):
+\begin{CodeChunk}
+\begin{CodeInput}
+# We separately precompute the affinities as this can take quite a long time.
+# This way, we can reuse the precomputed affinities in various different
+# embeddings, without having to repeat the same, costly computation every time.
+affinities = openTSNE.affinity.PerplexityBasedNN(x, perplexity=30)
+
+# Create the TSNEEmbedding object using our previously calculated initialization
+# and precomputed affinities
+embedding = openTSNE.TSNEEmbedding(init_full, affinities)
+
+# Run the standard t-SNE early exaggeration phase
+embedding_ee =    embedding.optimize(n_iter=500, exaggeration=12, momentum=0.5)
+embedding_c  = embedding_ee.optimize(n_iter=500, exaggeration=4,  momentum=0.8)
+embedding_b  =  embedding_c.optimize(n_iter=500, exaggeration=2,  momentum=0.8)
+embedding_a  =  embedding_b.optimize(n_iter=500, exaggeration=1,  momentum=0.8)
+\end{CodeInput}
+\end{CodeChunk}
+
 \subsection{Embedding New Samples}
@@
 cell clusters onto the reference embedding. The following snippet shows the
-code necessary to generate figure~\ref{fig:macosko}.
+code necessary to generate Fig.~\ref{fig:macosko}:
 \begin{CodeChunk}
@@
 cell-type classification in cases of unknown cell identities. For instance, in
-Fig.~\ref{fig:transform}, we construct a reference embedding using labeled data
+Fig.~\ref{fig:transform}, we construct a reference embedding (using the same code snippet as above) using labeled data
 from \citet{hochgerner2018conserved} containing gene-expression profiles of
@@
 
+\section{Discussion} \label{sec:discussion}
+
 \subsection{Versatility}
@@
 % We benchmark Python
-Fig.~\ref{fig:benchmarks_py} benchmarks four popular Python~\citep{vanrossum1995python} t-SNE
+Fig.~\ref{fig:benchmarks_py} benchmarks four popular \proglang{Python}~\citep{vanrossum1995python} t-SNE
 implementations, including \pkg{scikit-learn} (v1.0.1), \pkg{MulticoreTSNE}
@@
 % Problem with Python
-In Python, the most widely-used implementation of t-SNE comes from \pkg{scikit-learn}, which exhibits long runtimes when compared to other Python implementations.
+In \proglang{Python}, the most widely-used implementation of t-SNE comes from \pkg{scikit-learn}, which exhibits long runtimes when compared to other \proglang{Python} implementations.
 \pkg{scikit-learn}, like its \proglang{C++} counterpart -- \pkg{MulticoreTSNE}~\citep{Ulyanov2016} , is based on the Barnes-Hut approximation scheme.
@@
 Newer implementations, such as \pkg{FIt-SNE} and \opentsne, both implement the FIt-SNE approximation scheme, making them suitable for the visualization of millions of data points.
-While both libraries implement the same algorithm, \opentsne\ emphasizes ease of use and extensibility and is primarily written in Python, which inevitably introduces some runtime overhead.
+While both libraries implement the same algorithm, \opentsne\ emphasizes ease of use and extensibility and is primarily written in \proglang{Python}, which inevitably introduces some runtime overhead.
 This makes \opentsne\ marginally slower than its pure \proglang{C++} counterpart -- \pkg{FIt-SNE}.
@@
   programming languages. We choose
-  \opentsne\ (v0.6.2) for Python, \pkg{Rtsne}~\citep{krijthe2015rtsne} (v0.15)
+  \opentsne\ (v0.6.2) for \proglang{Python}, \pkg{Rtsne}~\citep{krijthe2015rtsne} (v0.15)
   for \proglang{R}, and \pkg{TSne.jl}~\citep{julia_tsne} (v1.3.0) for
@@
   but the smallest of data sets. \pkg{Rtsne} scales similarily to
-  \pkg{MulticoreTSNE} in Python, as it implements the $\mathcal{O}(N \log N)$
+  \pkg{MulticoreTSNE} in \proglang{Python}, as it implements the $\mathcal{O}(N \log N)$
   Barnes-Hut approximation scheme. Interestingly, the multi-threaded
@@
 We designed \opentsne\ to be accessible to a broader audience. We provide
-precompiled binaries for all major Python versions on all major platforms,
+precompiled binaries for all major \proglang{Python} versions on all major platforms,
 making the installation process as seamless as possible. One can install
-\opentsne\ through the Python Package Index (\pkg{PyPI}) or \pkg{conda} from the
+\opentsne\ through the \proglang{Python} Package Index (\pkg{PyPI}) or \pkg{conda} from the
 \pkg{conda-forge} channel, the two most widely adopted Python package managers.
 \opentsne's high-level interface is inspired by \pkg{scikit-learn}, which is well
-established in the Python data science ecosystem. \opentsne\ implements
+established in the \proglang{Python} data science ecosystem. \opentsne\ implements
 multi-threaded versions of both the Barnes-Hut and FIt-SNE approximation
 schemes, enabling it to be applied to data sets containing millions of data
-points. While the Python virtual machine inevitably introduces some performance
+points. While the \proglang{Python} virtual machine inevitably introduces some performance
 overhead, the runtime is comparable to its \proglang{C++} counterpart -- \pkg{FIt-SNE}.
@@
 \textbf{Ease of Installation} \\
-Native Python installation & \checkmark & & & \checkmark & & \\
+Native \proglang{Python} installation & \checkmark & & & \checkmark & & \\
 \pkg{PyPI} package & \checkmark & \checkmark & & \checkmark & & \\
@@
 
-We introduce \opentsne, the most feature-complete, open-source, Python
+We introduce \opentsne, the most feature-complete, open-source, \proglang{Python}
 implementation of the widely popular t-SNE visualization algorithm. The library

### 2022-05-09 paper/main.tex
Update diagram
revision of implementation, mostly grammar

@@
 
-\newcommand{\opentsne}{\pkg{openTSNE}}
+\usepackage{xspace}
+\newcommand{\opentsne}{\pkg{openTSNE}\xspace}
 
@@
   Pavlin G. Poli\v{c}ar\\
-  Faculty of Computer and Information Science\\
-  University of Ljubljana\\
+  Faculty of Computer and Information Science, University of Ljubljana\\
   Ljubljana, Slovenia
@@
   Bla\v{z} Zupan\\
-  Faculty of Computer and Information Science\\
-  University of Ljubljana\\
+  Faculty of Computer and Information Science, University of Ljubljana\\
   Ljubljana, Slovenia
@@
 \Abstract{
-One of the most popular techniques for constructing visualizations of large, high-dimensional data sets is $t$-distributed stochastic neighbor embedding (t-SNE).
+One of the most popular techniques for visualizing large, high-dimensional data sets is $t$-distributed stochastic neighbor embedding (t-SNE).
 Recently, several extensions have been proposed to address scalability issues and the quality of the resulting visualizations.
@@
 The library is fast and can treat data sets containing millions of data points.
-\opentsne\ is open-source and is distributed under the BSD-3-Clause License.
+We release \opentsne in open source and distribute it under the BSD-3-Clause License.
 Its source code is available at \url{https://github.com/pavlin-policar/openTSNE}.
@@
 
-The ever-growing volumes of high-dimensional data sets in machine learning call for efficient dimensionality reduction techniques and their implementations, capable of producing informative data visualizations.
+The ever-growing volumes of high-dimensional data sets in machine learning call for efficient dimensionality reduction techniques and their implementations to produce informative data visualizations.
 Popular approaches include principal component analysis, multidimensional scaling, $t$-distributed stochastic neighbor embedding (t-SNE)~\citep{maaten2008visualizing}, and uniform manifold approximation and projections (UMAP)~\citep{2018arXivUMAP}.
@@
 For instance, t-SNE is widely used in the bioinformatics community in areas such as single-cell transcriptomics~\citep{macosko2015highly,cao2019single,tasic2018shared}, human genetics~\citep{hirata2019genetic}, metagenomic assembly~\citep{beaulaurier2018metagenomic}, the spatial organization of microbial communities~\citep{sheth2019spatial}, and metabolomics~\citep{tkachev2019differences}.
-To visually explore the data structure, the objects of interest, such as diseased and healthy tissues or single cells, are profiled through thousands of features that include, for example, expression of genes or concentration of metabolites.
+To visually explore the data structure, the objects of interest, such as diseased and healthy tissues or single cells, are profiled through thousands of features that include, for example, the expression of genes or the concentration of metabolites.
 The task of t-SNE is to embed the objects of interest within a low, preferably two-dimensional embedding.
@@
   \caption{\label{fig:macosko}
-We use \opentsne\ to generate three t-SNE embeddings and demonstrate recent theoretical advances.
+We use \opentsne to generate three t-SNE embeddings and demonstrate recent theoretical advances.
 The data in \textbf{(a)} and \textbf{(b)} includes 44,808 mouse retinal cells (dots in the plots) that are described with high-dimensional gene-expression profiles from \citet{macosko2015highly}.
@@
 In this paper, we present \opentsne, our open-source \proglang{Python} implementation of t-SNE, complete with its recently proposed extensions.
-\opentsne\ is easy to install and includes precompiled binaries available through popular \proglang{Python} package managers.
+\opentsne is easy to install and includes precompiled binaries available through popular \proglang{Python} package managers.
 It provides a familiar API, making it suitable as a drop-in replacement for existing t-SNE implementations.
@@
 We introduce \opentsne, a comprehensive \proglang{Python} library that implements the t-SNE algorithm and its numerous recently proposed extensions.
-\opentsne\ is designed to accommodate all types of users and includes both a simple API that allows users to quickly obtain informative t-SNE visualizations, as well as a more flexible API that allows seasoned practitioners to modify all the
-different aspects of the t-SNE algorithm.
-
-The core t-SNE algorithm is comprised of three main steps, mirrored in the
-structure of the \opentsne\ library:
+\opentsne includes a simple API that allows users to quickly obtain informative t-SNE visualizations and a more flexible API that allows seasoned practitioners to modify all the different aspects of the t-SNE algorithm.
+
+The core t-SNE algorithm comprises three main steps, mirrored in the structure of the \opentsne library:
 
 \begin{enumerate}
-\item Firstly, an affinity model describing the similarities between data points must be specified.
-Standard t-SNE predominantly deals with tabular data, and similarities are computed using Eqn.~\ref{eq:p_ij}.
+\item The t-SNE algorithm requires a specification of an affinity model to describe the similarities between data points.
+Standard t-SNE predominantly deals with tabular data and computes similarities using Eqn.~\ref{eq:p_ij}.
 However, \citet{kobak2019art} showed that using mixtures of Gaussians instead of a single Gaussian often reveals better global consistency of the resulting visualization (albeit at some computational cost).
@@
 
-Furthermore, it is straightforward to define and use a custom affinity model.
-For instance, one can easily extend the \code{openTSNE.affinity.Affinities} class and provide an affinity model for relational data using a normalized adjacency matrix.
-This enables rapid experimentation and opens the doors to the visualization of non-tabular data.
-
-\item Secondly, initial positions must be chosen for each data point in the embedding space.
-These can either be chosen randomly or can be used to impose some initial global structure on the embedding, e.g., using a PCA-based initialization.
-Accordingly, the \code{openTSNE.initialization} module provides several possible initialization schemes, or the user can also specify their own initialization by providing a \code{numpy.ndarray} object.
-By default, \opentsne\ opts for PCA-based initialization.
-
-\item Finally, equipped with an affinity model and initial point positions, the only remaining step is to optimize the embedding such that the positions of the points in the embedding space reflect their corresponding similarities in the affinity model.
-In \opentsne, the initialization and affinities are used to construct an \code{openTSNE.TSNEEmbedding} object, which can then be optimized using different parameter settings.
-Once an embedding object is constructed, it can subsequently be optimized via the \code{.optimize} method.
-
-Optimization can be performed using either the Barnes-Hut or FIt-SNE approximation scheme.
-While the FIt-SNE approximation scheme scales linearly with the number of samples, it introduces additional computational overhead, which is often unnecessary for smaller data sets.
-For smaller data sets, this sometimes results in longer runtimes than when using the Barnes-hut approximation scheme.
-By default, \opentsne\ uses a heuristic to ensure the fastest algorithm is chosen for the given data set, but this setting can also be specified by the user.
+In \opentsne, it is straightforward to define and use a custom affinity model.
+For instance, one can extend the \code{openTSNE.affinity.Affinities} class and provide an affinity model for relational data using a normalized adjacency matrix.
+Such extension enables rapid experimentation and opens the doors to the visualization of non-tabular data.
+
+\item The optimization in t-SNE starts with initial positions for each data point in the embedding space.
+We can set the initial positions randomly or impose some initial global structure on the embedding with, for example, a PCA-based initialization.
+Accordingly, the \code{openTSNE.initialization} module provides several possible initialization schemes, The user can also specify their initialization by providing a \code{numpy.ndarray} object.
+By default, \opentsne opts for PCA-based initialization.
+
+\item In optimizing the embedding, the t-SNE algorithm finds the positions of the points in the embedding space that reflect their corresponding similarities in the affinity model.
+\opentsne stores the initial positions and affinities in a \code{openTSNE.TSNEEmbedding} object that includes the \code{.optimize} method for embedding optimization that uses either the Barnes-Hut or FIt-SNE approximation scheme.
+While the FIt-SNE approximation scheme scales linearly with the number of samples, it introduces additional computational overhead, often unnecessary for smaller data sets, resulting in longer runtimes than the Barnes-hut approximation scheme.
+By default, \opentsne uses a heuristic to choose the fastest algorithm for a given data set.
 \end{enumerate}
@@
 \begin{enumerate}
-\item Firstly, an affinity model must be provided describing the similarities between each new data point and the data points in the reference embedding.
-As before, \opentsne\ provides several possible affinity models in the \code{openTSNE.affinity} module.
-Most often, however, we want to utilize the same affinity model that was used to construct the reference embedding, as this leads to more spatially consistent embeddings.
-This is the default behavior of \opentsne.
-
-\item Secondly, the initial positions of the new data points in the embedding space must be specified.
-These can, again, be initialized randomly or be informed by their similarity to the data points in the reference embedding.
-As before, \opentsne\ provides several initialization options in the \code{openTSNE.initialization} module, but the user is free to provide their own initialization.
-
-\item Finally, what remains is to construct an embedding object, which can then be subject to optimization.
+\item Firstly, we must provide an affinity model describing the similarities between each new data point and the data points in the reference embedding.
+As for the construction of the initial embedding, \opentsne provides several possible affinity models in the \code{openTSNE.affinity} module.
+Most often, however, we want to utilize the affinity model from the construction of the reference embedding, as this leads to more spatially consistent embeddings.
+Using an affinity model from reference embedding for the embedding of the new data is the default behavior of \opentsne.
+
+\item Secondly, we must specify the initial positions of the new data points in the embedding space.
+We can, again, initialize these positions randomly or according to their similarity to the data points in the reference embedding.
+As before, \opentsne provides several initialization options in the \code{openTSNE.initialization} module, but the user is free to provide their specific initialization.
+
+\item Finally, what remains is a construction of an embedding object, subject to optimization.
 Because this new embedding relies on an existing reference embedding, we construct a \code{openTSNE.PartialTSNEEmbedding} embedding object, which indicates its dependence on a reference \code{openTSNE.TSNEEmbedding} object.
@@
 
-\begin{figure*}[t]
+\begin{figure*}[htbp]
   \centering
-  \includegraphics[width=0.75\textwidth]{opentsne-diagram}
+  \includegraphics[width=0.9\textwidth]{opentsne-diagram}
   \caption{\label{fig:diagram}
-  The high-level structure and flow of the \opentsne\ library.
+  High-level structure and information flow of the \opentsne library.
+Solid black lines show the standard parameter flow, while red dashed lines indicate the utility of optional parameters, allowing more refined control over the t-SNE algorithm.
   }
@@
 
-Fig.~\ref{fig:diagram} shows the high-level structure and flow of the \opentsne\ library.
-
-Through different parameters, the user is given full control over every aspect of the t-SNE visualization algorithm, from different ways to define affinities to the delicate tuning of optimization parameters.
-This allows for rapid experimentation and enables users to quickly and efficiently inspect their data from numerous angles and at varying levels of resolution.
-
-While \opentsne\ allows for an unprecedented level of control at every step, most end-users often only need a quick and easy way to create informative visualizations of their data sets.
-\opentsne\ provides a high-level API that closely follows the style of \pkg{scikit-learn}~\citep{sklearn_api}, a popular and widely used toolkit in the machine learning community.
-This allows users of all levels to quickly pick up and make use of the \opentsne\ library.
+Fig.~\ref{fig:diagram} shows the high-level structure and information flow of the \opentsne library.
+Through different parameters, the user is given complete control over every aspect of the t-SNE data visualization algorithm, from different ways to define affinities to the delicate tuning of optimization parameters.
+The parametrization of every part of t-SNE allows for rapid experimentation.
+It enables users to quickly and efficiently inspect their data from numerous angles and at varying levels of resolution.
+
+While \opentsne allows for control at every step, most end-users often only need a quick and easy way to create informative visualizations of their data sets.
+\opentsne provides a high-level API that closely follows the style of \pkg{scikit-learn}~\citep{sklearn_api}, a popular and widely used machine learning toolkit.
 
@@
 
-\opentsne\ is designed to be easy to use and can be installed through either \pkg{PyPI} using \pkg{pip}
+\opentsne is designed to be easy to use and can be installed through either Python Package Index \pkg{PyPI} using package installer \pkg{pip},
 \begin{CodeChunk}
@@
 \end{CodeChunk}
-or using \pkg{conda} from the \pkg{conda-forge} channel
+or using package manager \pkg{conda} from the \pkg{conda-forge} repository,
 \begin{CodeChunk}
@@
 \end{CodeChunk}
-\opentsne\ provides precompiled binaries for all major platforms (Windows, MacOS, Linux) for all currently supported \proglang{Python} versions (3.7---3.10).
-This is especially convenient for Windows users, as this obviates the need for a \proglang{C}/\proglang{C++} compiler, which is typically not bundled with the operating system.
-
-To verify that that \opentsne\ was installed correctly, try to import the module:
+\opentsne provides precompiled binaries for all major platforms (Windows, macOS, Linux) for all currently supported \proglang{Python} versions (3.7---3.10).
+Precompiled binaries are especially convenient for Windows users, as they obviate the need for a \proglang{C}/\proglang{C++} compiler, which is typically not bundled with Windows operating system.
+
+To verify that that \opentsne was installed correctly, users can run Python and try to import the module by:
 \begin{CodeChunk}
@@
 
-We demonstrate the simplest usage of \opentsne\ on the Iris data set from \citet{anderson1936species}.
-This small data set describes 150 iris flowers containing four variables (sepal length, sepal width, petal length, petal width) and their species (Iris Setosa, Iris Versicolor, Iris Virginica), which is the target variable.
-This data set is included in \pkg{scikit-learn}:
+We here outline the most straightforward usage of \opentsne on the Iris data set from \citet{anderson1936species}.
+This small data set profiles 150 iris flowers with four variables (sepal length, sepal width, petal length, petal width) and provides information about their species ({\em Iris Setosa}, {\em Iris Versicolor}, {\em Iris Virginica}).
+Iris data set comes with \pkg{scikit-learn}:
 \begin{CodeChunk}
@@
 \end{CodeChunk}
-\pkg{scikit-learn} loads the data set into a \pkg{numpy} array, which can directly
-be used by \opentsne:
+\pkg{scikit-learn} loads the data set into a \pkg{numpy} array which can directly be used by \opentsne:
 \begin{CodeChunk}
@@
 \end{CodeChunk}
-Fig.~\ref{fig:iris} shows a visualization of the resulting t-SNE visualization.
-As expected, t-SNE is correctly able to group together samples from the same species.
-Additionally, the t-SNE embedding reveals that Iris Setosa is markedly different from the other two species, while flowers from Iris Versicolor and Iris Virginica may appear more similar to one another.
+Fig.~\ref{fig:iris} displays the resulting t-SNE visualization.
+As expected, t-SNE can correctly group samples from the same species.
+The t-SNE embedding reveals that flowers from {\em Iris Setosa} are markedly different from the other two species.
+At the same time, flowers from {\em Iris Versicolor} and {\em Virginica} are more similar to one another.
 \begin{figure*}[htbp]
@@
   \caption{\label{fig:iris}
-  A visualization of the obtained t-SNE embedding on the Iris data set. The colors
-  correspond to different species of the Iris flower.
+  A visualization of the constructed t-SNE embedding on the Iris data set. The colors correspond to different species of Iris.
 }
 \end{figure*}
-We can now use this embedding as a reference for new data points. In this simple example, we will duplicate some of the original samples and add them into the embedding.
+We can now use the resulting embedding as a reference for new data points. In this simple example, we will duplicate some of the original samples and add them to the embedding.
 \begin{CodeChunk}
@@
 
-While the above example offers users a quick and easy way to generate informative visualizations, \opentsne\ allows us to control the t-SNE algorithm at every step.
-To demonstrate the flexibility of \opentsne, we will repeat the same analysis of the Iris data set as before, this time leveraging the more advanced features of the library.
-More practical examples of advanced usage will be shown in Section~\ref{sec:discussion}.
+While the above example offers users a quick and easy way to generate informative visualizations, \opentsne allows us to control every step of the t-SNE embedding.
+To demonstrate this flexibility, we will repeat the exact analysis of the Iris data set as before, this time leveraging the library's more advanced features.
+We list other practical examples with more complex data sets in Section~\ref{sec:discussion}.
 
@@
 \end{CodeChunk}
-Next, we import the \opentsne\ library:
+Next, we import the \opentsne library:
 \begin{CodeChunk}
@@
 \end{CodeChunk}
-We now mirror the three steps listed above. First, we select an affinity model which describes the similarities between data points:
+We now mirror the three t-SNE steps from the previous section. First, we select an affinity model which describes the similarities between data points:
 \begin{CodeChunk}
@@
 \end{CodeChunk}
-Next, using the affinity model and initialization, we can create an embedding object:
+Finally, using the affinity model and initialization, we can create an embedding object:
 \begin{CodeChunk}
@@
 \end{CodeChunk}
-Notice that the embedding object contains the same point coordinates as the initialization.
-This is because although we have created an \code{openTSNE.TSNEEmbedding} object, we have not yet run any optimization.
-To optimize the embedding, we merely have to call the \code{embedding.optimize} method using our desired parameter settings.
-Here, we use the default parameter settings, including a shorter early-exaggeration phase, followed by a longer phase with no exaggeration.
+Notice that the embedding object contains the same point coordinates as the initialization: although we have created an \code{openTSNE.TSNEEmbedding} object, we have not yet run any optimization.
+To optimize the embedding, we have to call the \code{embedding.optimize} method using our desired parameter settings.
+We use the default parameter settings here, including a shorter early-exaggeration phase, followed by a longer phase with no exaggeration.
 \begin{CodeChunk}
@@
 \end{CodeChunk}
-Next, we can add new samples into the embedding:
+Next, we can add new samples into the embedding,
 \begin{CodeChunk}
@@
 \end{CodeChunk}
-And validate that this also produces the exact same embedding as before:
+and validate that this also produces the exact same embedding as before:
 \begin{CodeChunk}
@@
 
-Dimensionality reduction techniques implicitly assume that high-dimensional data
-lies on a lower-dimensional manifold, which can accurately be captured by a small number of dimensions. However, there is no evidence that every data set can
-accurately be described using only two dimensions, and any such embedding will
-inevitably lead to a loss of information. Thus, it is beneficial to examine
-multiple embeddings, each of which provides a different perspective on topology
-and other data characteristics.
+Dimensionality reduction techniques implicitly assume that high-dimensional data lies on a lower-dimensional manifold, which can accurately be captured by a small number of dimensions.
+However, there is no evidence that every data set can accurately be described using only two dimensions, and any such embedding will inevitably lead to a loss of information.
+Thus, it is beneficial to examine multiple embeddings, each providing a different perspective on topology and other data characteristics.
 
@@
   \caption{\label{fig:tasic}
-  We use \opentsne\ to create four different
+  We use \opentsne to create four different
   visualizations of the \citet{tasic2018shared} data,
@@
   perplexities by using a multi-scale kernel affinity model \textbf{(c)}
-  and obtain a trade-off between global and local structure.
-  Alternatively, we can inspect more fine-grained structure and reveal
+  and obtain a trade-off between global and local structures.
+  Alternatively, we can inspect more fine-grained structures and reveal
   smaller clusters by using a more heavy-tailed kernel \textbf{(d)}.}
@@
 
-We illustrate this point by generating four different embeddings of the data on
-single-cell gene expression in mouse brain~\citep{tasic2018shared}. We generate
-the embeddings using the following code snippet (shortened for clarity):
+We illustrate the benefits of creating different t-SNE plots of the same data set by generating four different embeddings of the data on single-cell gene expression in mouse brain~\citep{tasic2018shared}.
+We generate the embeddings using the following code snippet (shortened for clarity):
 \begin{CodeChunk}
@@
 
-When dealing with data containing millions of data points, standard t-SNE
-embeddings often become unwieldy -- cluster boundaries are blurred, large
-clusters absorb smaller ones, and relationships between clusters become
-increasingly difficult to interpret. We constructed Fig.~\ref{fig:cao}.a from
-the data containing expression profiles of over two million single cells
-captured at different time points in mouse development. The embedding indicates
-numerous clusters with transitions between time points, as marked by the
-color-coding, that are difficult to interpret. \citet{kobak2019art} that
-increasing attractive forces between similar data points controlled via
-the \textit{exaggeration} parameter leads to more compact clusters, and
-subsequently, more informative visualizations. For instance, \ref{fig:cao}.b
-doubles the default exaggeration, which uncovers some
-of the data's overall structure. Further doubling the exaggeration in
-\ref{fig:cao}.c allows us to observe that the data is comprised of two main groups
-of cells and eight somewhat smaller clusters. The visualization also reveals
-several tiny clusters, possibly corresponding to rare cell types.
+When dealing with data containing millions of data points, standard t-SNE embeddings often become unwieldy -- cluster boundaries are blurred, large clusters absorb smaller ones, and relationships between groups become increasingly difficult to interpret.
+We constructed Fig.~\ref{fig:cao}.a from the data containing expression profiles of over two million single cells captured at different time points in mouse development.
+The embedding reveals numerous clusters with transitions between time points marked by the color coding, which are difficult to interpret.
+\citet{kobak2019art} suggest that increasing attractive forces between similar data points controlled via the \textit{exaggeration} parameter leads to more compact clusters and, subsequently, more informative visualizations.
+For instance, \ref{fig:cao}.b doubles the default exaggeration, which uncovers some of the data's overall structure.
+Further doubling the exaggeration in \ref{fig:cao}.c allows us to observe that the data is comprised of two main groups of cells and eight somewhat smaller clusters.
+The visualization also reveals several tiny clusters, possibly corresponding to rare cell types.
 
@@
   \caption{\label{fig:cao}
-  Increasing the exaggeration parameter leads to compact clusters, highlighting
-  the data's global organization and emphasizing the continuous nature of
-  cell state transitions. The data set from \citet{cao2019single}
-  contains expression profiles from 2,058,652 single cells. The data were
-  collected from mice embryos at different developmental stages at daily
-  intervals after 9.5 to 13.5 days. \textbf{(c)} reveals that the data is
-  comprised of two main components -- the neural tube and mesenchymal
-  cells -- as well as several other smaller clusters. The
-  colors indicate developmental progression with red indicating
-  least-developed cells and blue indicating most developed cells. The
-  overall developmental trajectory is most apparent with higher
-  exaggeration levels, showing red cells slowly transitioning into blue
-  cells. Progressively easing the exaggeration factor uncovers finer
-  clusters within the larger groups, as shown in \textbf{(b)} with
-  exaggeration of two and subsequently in \textbf{(a)}, where we show the
-  standard t-SNE with no exaggeration. 32,011 putative doublets are
-  excluded from the visualizations.
+  Increasing the exaggeration parameter leads to compact clusters, highlighting the data's global organization and emphasizing the continuous nature of cell state transitions.
+The data set from \citet{cao2019single} contains expression profiles from 2,058,652 single cells.
+The data were collected from mice embryos at different developmental stages daily after 9.5 to 13.5 days.
+\textbf{(c)} reveals that the data is comprised of two main components -- the neural tube and mesenchymal cells -- as well as several other smaller clusters.
+The colors indicate developmental progression, where red indicates the least-developed cells and blue indicates the most developed cells.
+The overall developmental trajectory is most apparent with higher exaggeration levels, showing red cells slowly transitioning into blue cells.
+Progressively easing the exaggeration factor uncovers finer clusters within the larger groups, as shown in \textbf{(b)} with exaggeration of two and subsequently in \textbf{(a)}, where we show the standard t-SNE with no exaggeration.
+32,011 putative doublets are excluded from the visualizations.
 }
@@
 
-Exaggeration can highlight transitions between cell states in developmental
-studies. Standard t-SNE often produces embeddings with clearly defined, discrete
-clusters. We can adjust the level of granularity and resolution of the clusters
-with several parameters in Fig.~\ref{fig:tasic}. However, discrete clusters are
-often undesired in developmental studies where cells' state is assumed to follow
-a continuous transition path. To this end, other embedding techniques such as
-UMAP and ForceAtlas2 are used to better capture the continuity between cell
-states. Recently, \citet{bohm2020unifying} showed that
-embeddings produced by t-SNE with exaggeration values of 4 and $\sim30$
-construct embeddings that are markedly similar to UMAP and ForceAtlas2,
-respectively. For example, in Fig.~\ref{fig:cao}.a, the developmental trajectory
-between different time points is difficult to observe due to many sprawled out
-clusters. On the other hand, it is easier to trace the development when we
-increase the exaggeration factor from $1$ to $2$ to $4$ in
-Figs.~\ref{fig:cao}.b-c.
-
-Figs.~\ref{fig:cao}.a-c were generated using the following code snippet, which demonstrates how we can take advantage of the more advanced features of \opentsne\ to more quickly create meaningful visualizations of massive data sets.
-We first take a sample of the data and use t-SNE to construct a globally consistent initialization, which will then be used to initialize the remaning data points:
+Exaggeration can highlight transitions between cell states in developmental studies.
+Standard t-SNE often produces embeddings with clearly defined, discrete clusters.
+We can adjust the level of granularity and resolution of the clusters with several parameters, as shown in Fig.~\ref{fig:tasic}.
+However, discrete clusters are often undesired in developmental studies where cells' state is assumed to follow a continuous transition path.
+To this end, researchers have used other embedding techniques such as UMAP and ForceAtlas2 to better capture the continuity between cell states.
+Recently, \citet{bohm2020unifying} showed that embeddings produced by t-SNE with exaggeration values of 4 and $\sim30$ construct embeddings that are markedly similar to UMAP and ForceAtlas2, respectively.
+For example, in Fig.~\ref{fig:cao}.a, the developmental trajectory between different time points is difficult to observe due to many sprawled out clusters.
+On the other hand, it is easier to trace the development when we increase the exaggeration factor from $1$ to $2$ to $4$ in Figs.~\ref{fig:cao}.b-c.
+
+We have used the following snippet to create Figs.~\ref{fig:cao}.a-c and to demonstrate how we can take advantage of the more advanced features of \opentsne to more quickly create meaningful visualizations of massive data sets.
+Here, we first take a sample of the data and use t-SNE to construct a globally consistent embedding, which we will then use for initialization of the positions for the remaining data points:
 \begin{CodeChunk}
@@
 \end{CodeChunk}
-We then generate the three different embeddings using the following code snippet
-(shortened for clarity):
+We then generate the three different embeddings using the following code snippet (shortened for clarity):
 \begin{CodeChunk}
@@
 
-Unlike other popular dimensionality reduction techniques such as principal
-component analysis or autoencoders, t-SNE is a non-parametric method and does
-not define an explicit mapping to the embedding space. Therefore, embeddings of
-new data points need to be found through
-optimization~\citep{policar2021embedding}. \opentsne\ is currently the only
-publicly available library allowing users to add new samples to existing
-embeddings in a principled manner.
-
-Figs.~\ref{fig:macosko}.b and \ref{fig:macosko}.c demonstrate how we can use a
-previously labeled single-cell data set and embed cells from a separate
-experiment into the reference landscape. The reference data from
-\citet{macosko2015highly} contains gene expression profiles from mouse retinal
-cells. By embedding the samples from a similar experiment on bipolar retinal
-cells by \citet{shekhar2016comprehensive}, we can correctly map the bipolar
-cell clusters onto the reference embedding. The following snippet shows the
-code necessary to generate Fig.~\ref{fig:macosko}:
+Unlike other popular dimensionality reduction techniques such as principal component analysis or autoencoders, t-SNE is a non-parametric method and does not define an explicit mapping to the embedding space.
+Therefore, embeddings of new data points need to be found through optimization~\citep{policar2021embedding}.
+\opentsne is currently the only publicly available library allowing users to add new samples to existing embeddings in a principled manner.
+
+Figs.~\ref{fig:macosko}.b and \ref{fig:macosko}.c demonstrate how we can use a previously labeled single-cell data set and embed cells from a separate experiment into the reference landscape.
+The reference data from \citet{macosko2015highly} contains gene expression profiles from mouse retinal cells.
+By embedding the samples from a similar experiment on bipolar retinal cells by \citet{shekhar2016comprehensive}, we can correctly map the bipolar cell clusters onto the reference embedding.
+The following snippet shows the code necessary to generate Fig.~\ref{fig:macosko}:
 \begin{CodeChunk}
@@
 
-Embedding single cells into existing reference atlases can also be useful for
-cell-type classification in cases of unknown cell identities. For instance, in
-Fig.~\ref{fig:transform}, we construct a reference embedding (using the same code snippet as above) using labeled data
-from \citet{hochgerner2018conserved} containing gene-expression profiles of
-cells from the mouse brain. The authors assign a type to each cell. We can
-verify their classification accuracy by visualizing the expression of
-well-established gene markers for the major cell types. We then embed cells
-from \citet{harris2018classes} into the constructed cell atlas. In
-\citet{harris2018classes}, labels are provided only for neuronal cells. In the
-resulting mapping, we can quickly identify other non-neuronal cell types,
-including oligodendrocytes and astrocytes. We can further use marker genes to
-validate that the mapping in the reference landscape is correct.
+Embedding single cells into existing reference atlases can also be useful for cell-type classification in cases of unknown cell identities.
+For instance, in Fig.~\ref{fig:transform}, we construct a reference embedding (using the same code snippet as above) using labeled data from \citet{hochgerner2018conserved} containing gene-expression profiles of cells from the mouse brain.
+The authors assign a type to each cell.
+We can verify their classification accuracy by visualizing the expression of well-established gene markers for the major cell types.
+We then embed cells from \citet{harris2018classes} into the constructed cell atlas.
+In \citet{harris2018classes}, labels are provided only for neuronal cells.
+We can quickly identify other non-neuronal cell types in the resulting mapping, including oligodendrocytes and astrocytes.
+We can further use marker genes to validate that the mapping in the reference landscape is correct.
 
@@
   \caption{\label{fig:transform}
-  \opentsne\ supports embedding new samples into an existing reference t-SNE
-  landscape. For the series of visualizations shown in this figure, we
-  first construct a t-SNE embedding for the data from 
-  \citet{hochgerner2018conserved} containing 24,185
-  developing, single cells from the mouse hippocampus. The data contains
-  gene expression in different neurons, supporting glia, and other
-  vascular cells (upper left). Data points representing cells are colored
-  according to cell-types assigned in the original publication; see the
-  legend from Fig.~\ref{fig:tasic} to map colors to cell-type. We then
-  embed new, hippocampal cells collected in a study by
-  \citet{harris2018classes} using the embedding of \citet{hochgerner2018conserved}
-  data as a reference. In their study, \citet{harris2018classes}
-  collected 6,971 single-cells and focused on identifying different
-  types of inhibitory neurons. However, almost half of the collected cells
-  are not neurons and were left uncharacterized. Inspecting the embeddings
-  of these cells in the reference embedding (bottom left) reveals that in
-  addition to inhibitory neurons, the data contains several supporting
-  glial cells as well as a small population of endothelial cells. We can
-  verify our approach's accuracy by inspecting marker genes for the major
-  cell types in the reference (top row) and embedded samples (bottom row).
+  \opentsne supports embedding new samples into an existing reference t-SNE landscape.
+For the series of visualizations shown in this figure, we first construct a t-SNE embedding for the data from \citet{hochgerner2018conserved} containing 24,185 developing, single cells from the mouse hippocampus.
+The data contains gene expression in different neurons, supporting glia, and other vascular cells (upper left).
+Data points representing cells are colored according to cell types assigned in the original publication; see the legend from Fig.~\ref{fig:tasic} to map colors to cell type.
+We then embed new hippocampal cells collected in a study by \citet{harris2018classes} using the embedding of \citet{hochgerner2018conserved} data as a reference.
+In their research, \citet{harris2018classes} collected 6,971 single-cells and focused on identifying different types of inhibitory neurons.
+However, almost half of the collected cells are not neurons and were left uncharacterized.
+Inspecting the embeddings of these cells in the reference embedding (bottom left) reveals that in addition to inhibitory neurons, the data contains several supporting glial cells and a small population of endothelial cells.
+We can verify our approach's accuracy by inspecting marker genes for the major cell types in the reference (top row) and embedded samples (bottom row).
 }
@@
 
-The examples presented above demonstrate how to use \opentsne\ to quickly gain
-insight into newly-sequenced, single-cell data sets by utilizing existing cell
-atlases. The approach is general and not limited to single-cell gene expression,
-and one can, in principle, apply it to any tabular data set regardless of field.
+The examples presented above demonstrate how to use \opentsne to quickly gain insight into newly-sequenced, single-cell data sets by utilizing existing cell atlases.
+The approach is general and not limited to single-cell gene expression.
+One can, in principle, apply it to any tabular data set regardless of the research field or origin of the data.
 
@@
 
-Versatility, the ability to use and combine different optimization approaches to
-construct different embedding spaces, is another of \opentsne's core design
-principles. \citet{kobak2019art} recently provided several recommendations and
-tricks to obtain better and more meaningful t-SNE visualizations. These include
-multi-scale similarity kernels, perplexity annealing, and increasing
-exaggeration when working with massive data sets. \opentsne\ provides a flexible
-API to incorporate these improvements in just a few lines of code. Furthermore,
-\opentsne\ supports custom affinity models, enabling users to construct t-SNE
-embeddings on non-tabular relational data: the only requirement imposed by the
-affinity model is some notion of similarity between data points. Finally,
-\opentsne's comprehensive callback system can be utilized to monitor and adapt
-different stages of the optimization phase and has been used to construct
-visually appealing animations of the t-SNE optimization process.
+Versatility, the ability to use and combine different optimization approaches to construct different embedding spaces, is another of \opentsne's core design principles.
+\citet{kobak2019art} recently provided several recommendations and tricks to obtain better and more meaningful t-SNE visualizations.
+These include multi-scale similarity kernels, perplexity annealing, and increasing exaggeration when working with massive data sets.
+\opentsne provides a flexible API to incorporate these improvements in just a few lines of code.
+Furthermore, \opentsne supports custom affinity models, enabling users to construct t-SNE embeddings on non-tabular relational data: the only requirement imposed by the affinity model is some notion of similarity between data points.
+Finally, \opentsne's comprehensive callback system can be utilized to monitor and adapt different stages of the optimization phase and has been used to construct visually appealing animations of the t-SNE optimization process.
 
@@
 One of the t-SNE's common criticisms is limited scalability to large data sets containing, for instance, millions of data points~\citep {becht2019dimensionality}.
-The culprit for slow response time stems from particular optimization procedures and their specific implementation in popular open-source libraries.
+The culprit for slow response time stems from the optimization procedures and their specific implementation in popular open-source libraries.
 Until quite recently, most popular implementations of t-SNE were based on either a direct implementation of the t-SNE algorithm with asymptotic time complexity $\mathcal{O}(N^2)$~\citep{maaten2008visualizing}, or the Barnes-Hut approximation scheme developed by \citet{van2014accelerating} with asymptotic time complexity $\mathcal{O}(N \log N)$.
@@
 % We benchmark Python
-Fig.~\ref{fig:benchmarks_py} benchmarks four popular \proglang{Python}~\citep{vanrossum1995python} t-SNE
-implementations, including \pkg{scikit-learn} (v1.0.1), \pkg{MulticoreTSNE}
-(v0.1), \pkg{FIt-SNE} (v1.1.0), and \opentsne\ (v0.6.2). Benchmarks were
-performed on an Intel(R) Xeon(R) CPU E5-1650 equipped with 128 GB of memory.
-Benchmarks were run for $1,000$ iterations with the original t-SNE parameters,
-as some implementations do not allow for their modification. To ensure a
-realistic benchmarking scenario, we utilize the 10X Genomics 1.3 million mouse
-brain data set~\citep{xxx}, containing 1.3 million single-cell gene-expression
-profiles. We preprocess the data using the standard single-cell analysis
-pipeline~\citep{xxx} and extract the top 50 principal components. To generate
-benchmark data sets of different sample sizes, we subsample the data set five
-times at each sample size. In total, we run each implementation on 30 different
-data sets.
+Fig.~\ref{fig:benchmarks_py} benchmarks four popular \proglang{Python}~\citep{vanrossum1995python} t-SNE implementations, including \pkg{scikit-learn} (v1.0.1), \pkg{MulticoreTSNE} (v0.1), \pkg{FIt-SNE} (v1.1.0), and \opentsne (v0.6.2).
+Benchmarks were performed on an Intel(R) Xeon(R) CPU E5-1650 equipped with 128 GB of memory.
+Benchmarks were run for $1,000$ iterations with the original t-SNE parameters, as some implementations do not allow for their modification.
+To ensure a realistic benchmarking scenario, we utilize the 10X Genomics 1.3 million mouse brain data set~\citep{xxx}, containing 1.3 million single-cell gene-expression profiles.
+We preprocess the data using the standard single-cell analysis pipeline~\citep{xxx} and extract the top 50 principal components.
+To generate benchmark data sets of different sample sizes, we subsample the data set five times at each sample size.
+In total, we run each implementation on 30 different data sets.
 
@@
   \caption{\label{fig:benchmarks_py}
-  We benchmark \opentsne\ (v0.6.2) against three popular open-source
+  We benchmark \opentsne (v0.6.2) against three popular open-source
   implementations from \pkg{scikit-learn}~\citep{pedregosa2011scikit}
@@
   on an Intel(R) Xeon(R) CPU E5-1650 equipped with 128 GB of memory. Notice
-  that \opentsne\ scales similarly to \pkg{FIt-SNE}, as they both use the
+  that \opentsne scales similarly to \pkg{FIt-SNE}, as they both use the
   same interpolation-based approximation scheme, while \pkg{scikit-learn} and
@@
 Newer implementations, such as \pkg{FIt-SNE} and \opentsne, both implement the FIt-SNE approximation scheme, making them suitable for the visualization of millions of data points.
-While both libraries implement the same algorithm, \opentsne\ emphasizes ease of use and extensibility and is primarily written in \proglang{Python}, which inevitably introduces some runtime overhead.
-This makes \opentsne\ marginally slower than its pure \proglang{C++} counterpart -- \pkg{FIt-SNE}.
+While both libraries implement the same algorithm, \opentsne emphasizes ease of use and extensibility and is primarily written in \proglang{Python}, which inevitably introduces some runtime overhead.
+This makes \opentsne marginally slower than its pure \proglang{C++} counterpart -- \pkg{FIt-SNE}.
 However, these differences disappear almost entirely when utilizing multi-threading, as shown in Fig.~\ref{fig:benchmarks_py}.
-Additionally, \opentsne\ provides a flexible API, allowing the user to split up the embedding-construction process into several parts.
-This allows us to cache and reuse intermediate results, enabling users to quickly experiment with different parameter settings and iterate on their final visualizations.
-This kind of functionality is not supported in other implementations, which require the user to start from scratch for every embedding, resulting in large amounts of redundant computation.
+
+Additionally, \opentsne provides a flexible API, allowing the user to split the embedding-construction process into several parts.
+This modularity enables us to cache and reuse intermediate results, allowing users to experiment with different parameter settings and iterate on their final visualizations.
+Such functionality is not supported in other t-SNE implementations, which require the user to start from scratch for every embedding, resulting in large amounts of redundant computation.
 
@@
 In \proglang{Julia}~\citep{bezanson2017julia}, the most popular implementation -- \pkg{TSne.jl}~\citep{julia_tsne} -- is based on a naive $\mathcal{O}(N^2)$ implementation of the algorithm, which allows it to be used only on smaller data sets containing up to thousands of data points.
-We benchmark \pkg{Rtsne} (v0.15) and  \pkg{TSne.jl} (v1.3.0) and compare it to \opentsne\ (v0.6.2).
+We benchmark \pkg{Rtsne} (v0.15) and  \pkg{TSne.jl} (v1.3.0) and compare it to \opentsne (v0.6.2).
 Fig.~\ref{fig:benchmarks_lang} illustrates the scaling of each approximation scheme.
-Of these implementations, only \opentsne\ is able to scale to data sets containing millions of data points in a reasonable amount of time.
+Of these implementations, only \opentsne can scale to data sets containing millions of data points in a reasonable amount of time.
 
@@
   \caption{\label{fig:benchmarks_lang}
-  We select the fastest available, open-source implementations of the t-SNE
-  algorithm from the \proglang{Python}, \proglang{R}, and \proglang{Julia}
-  programming languages. We choose
-  \opentsne\ (v0.6.2) for \proglang{Python}, \pkg{Rtsne}~\citep{krijthe2015rtsne} (v0.15)
-  for \proglang{R}, and \pkg{TSne.jl}~\citep{julia_tsne} (v1.3.0) for
-  \proglang{Julia}. The benchmarks reflect the time complexity of the
-  implemented approximation schemes. For instance, \pkg{TSne.jl} implements a
-  naive $\mathcal{O}(N^2)$ algorithm, making it prohibitively expensive for all
-  but the smallest of data sets. \pkg{Rtsne} scales similarily to
-  \pkg{MulticoreTSNE} in \proglang{Python}, as it implements the $\mathcal{O}(N \log N)$
-  Barnes-Hut approximation scheme. Interestingly, the multi-threaded
-  implementation version of \pkg{Rtsne} crashes on benchmark data sets
-  containing over 500,000 data points due to lack of stack memory. We found
-  these crashes to be inconsistent and somewhat random. \opentsne\ is the only
-  library implementing the more recent FIt-SNE approximation scheme, which makes
-  it suitable for larger data sets.
+  We select the fastest available, open-source implementations of the t-SNE algorithm from the \proglang{Python}, \proglang{R}, and \proglang{Julia} programming languages.
+We choose \opentsne (v0.6.2) for \proglang{Python}, \pkg{Rtsne}~\citep{krijthe2015rtsne} (v0.15) for \proglang{R}, and \pkg{TSne.jl}~\citep{julia_tsne} (v1.3.0) for \proglang{Julia}.
+The benchmarks reflect the time complexity of the implemented approximation schemes.
+For instance, \pkg{TSne.jl} implements a naive $\mathcal{O}(N^2)$ algorithm, making it prohibitively expensive for all but the smallest of data sets.
+\pkg{Rtsne} scales similarly to \pkg{MulticoreTSNE} in \proglang{Python}, as it implements the $\mathcal{O}(N \log N)$ Barnes-Hut approximation scheme.
+Interestingly, the multi-threaded implementation version of \pkg{Rtsne} crashes on benchmark data sets containing over 500,000 data points due to a lack of stack memory.
+We found these crashes to be inconsistent and somewhat random.
+\opentsne is the only library implementing the more recent FIt-SNE approximation scheme, which makes it suitable for larger data sets.
 }
@@
 
-Intuitive access and simple installation procedures almost universally correlate
-with the widespread adoption of novel computational techniques. While the t-SNE
-implementation from \pkg{scikit-learn} fits this requirement, the implementation
-proves prohibitively slow for even moderately-sized data sets that span tens of
-thousands of data records. Other \proglang{C++} implementations such as \pkg{MulticoreTSNE}
-and \pkg{FIt-SNE} exhibit better scaling in more massive data sets, but do not
-provide precompiled binaries and require users to compile the software
-themselves. This problem is critical, for instance, for users of the Windows
-operating system, where the \proglang{C++} compiler does not come with the system, making
-the correct configuration of current t-SNE implementations cumbersome.
-
-We designed \opentsne\ to be accessible to a broader audience. We provide
-precompiled binaries for all major \proglang{Python} versions on all major platforms,
-making the installation process as seamless as possible. One can install
-\opentsne\ through the \proglang{Python} Package Index (\pkg{PyPI}) or \pkg{conda} from the
-\pkg{conda-forge} channel, the two most widely adopted Python package managers.
-\opentsne's high-level interface is inspired by \pkg{scikit-learn}, which is well
-established in the \proglang{Python} data science ecosystem. \opentsne\ implements
-multi-threaded versions of both the Barnes-Hut and FIt-SNE approximation
-schemes, enabling it to be applied to data sets containing millions of data
-points. While the \proglang{Python} virtual machine inevitably introduces some performance
-overhead, the runtime is comparable to its \proglang{C++} counterpart -- \pkg{FIt-SNE}.
-Finally, \opentsne\ is extensible. Its modular design enables researchers to
-quickly experiment with different parameter settings and easily incorporate
-custom components into the software. We provide an extended list of features and
-comparison to other popular t-SNE implementations in Tab.~\ref{tab:features}.
+Intuitive access and simple installation procedures typically correlate with the widespread adoption of novel computational techniques.
+While the t-SNE implementation from \pkg{scikit-learn} fits this requirement, its implementation is prohibitively slow for even moderately-sized data sets that span tens of thousands of data records.
+Other \proglang{C++} implementations such as \pkg{MulticoreTSNE} and \pkg{FIt-SNE} exhibit better scaling in more massive data sets, but the packages do not provide precompiled binaries and require users to compile the software themselves.
+This problem is critical, for instance, for users of the Windows operating system, where the \proglang{C++} compiler does not come with the system, making the correct configuration of current t-SNE implementations cumbersome.
+
+We designed \opentsne to be accessible to a broader audience.
+We provide precompiled binaries for all major \proglang{Python} versions on all major platforms, making the installation process as seamless as possible.
+One can install \opentsne through the \proglang{Python} Package Index (\pkg{PyPI}) or \pkg{conda} from the \pkg{conda-forge} channel, the two most widely adopted Python package managers.
+\opentsne's high-level interface is inspired by \pkg{scikit-learn}, which is well established in the \proglang{Python} data science ecosystem.
+\opentsne implements multi-threaded versions of the Barnes-Hut and FIt-SNE approximation schemes, enabling it to be applied to data sets containing millions of data points.
+While the \proglang{Python} virtual machine inevitably introduces some performance overhead, the runtime is comparable to its \proglang{C++} counterpart -- \pkg{FIt-SNE}.
+Finally, \opentsne is extensible.
+Its modular design enables researchers to experiment quickly with different parameter settings and easily incorporate custom components into the software.
+We provide an extended list of features and comparison to other popular t-SNE implementations in Tab.~\ref{tab:features}.
 
@@
 
-We introduce \opentsne, the most feature-complete, open-source, \proglang{Python}
-implementation of the widely popular t-SNE visualization algorithm. The library
-incorporates the latest developments to the t-SNE algorithm and makes them
-easily accessible through a familiar API. \opentsne\ implements efficient
-approximation schemes, making it suitable for the visualization of large data
-sets containing up to millions of data points. Finally, \opentsne\ is easy to
-install and comes with precompiled binaries available for all major platforms.
+We introduce \opentsne, the most feature-complete, open-source, \proglang{Python} implementation of the widely popular t-SNE visualization algorithm.
+The library incorporates the latest developments to the t-SNE algorithm and makes them easily accessible through a familiar API.
+\opentsne implements efficient approximation schemes, making it suitable for visualizing large data sets containing up to millions of data points.
+Finally, \opentsne is easy to install and comes with precompiled binaries available for all major platforms.
 
@@
 
-\opentsne\ is distributed under the BSD-3-Clause License and its source code is
+\opentsne is distributed under the BSD-3-Clause License and its source code is
 publicly available at \url{https://github.com/pavlin-policar/openTSNE}.
-\opentsne\ can also be installed from \pkg{PyPI} and \pkg{conda-forge}.
+\opentsne can also be installed from \pkg{PyPI} and \pkg{conda-forge}.
 End-to-end usage examples, documentation, and API reference are all available at
@@
 for his help with the FIt-SNE algorithm. This work was supported by Slovenian
-Research Agency (P2-0209).
+Research Agency (P2-0209, L2-3170).
 

### 2022-05-22 paper/main.tex
Text edits.

@@
 We introduce \opentsne, a modular \proglang{Python} library that implements the core t-SNE algorithm and its many extensions.
-The library is fast and can treat data sets containing millions of data points.
+The library is faster than existing implementations and can compute projections of data sets containing millions of data points in minutes.
 We release \opentsne in open source and distribute it under the BSD-3-Clause License.
@@
 Popular approaches include principal component analysis, multidimensional scaling, $t$-distributed stochastic neighbor embedding (t-SNE)~\citep{maaten2008visualizing}, and uniform manifold approximation and projections (UMAP)~\citep{2018arXivUMAP}.
-Among these, t-SNE has received much attention as it can address high volumes of data and reveal the underlying data structure.
-For instance, t-SNE is widely used in the bioinformatics community in areas such as single-cell transcriptomics~\citep{macosko2015highly,cao2019single,tasic2018shared}, human genetics~\citep{hirata2019genetic}, metagenomic assembly~\citep{beaulaurier2018metagenomic}, the spatial organization of microbial communities~\citep{sheth2019spatial}, and metabolomics~\citep{tkachev2019differences}.
-To visually explore the data structure, the objects of interest, such as diseased and healthy tissues or single cells, are profiled through thousands of features that include, for example, the expression of genes or the concentration of metabolites.
-The task of t-SNE is to embed the objects of interest within a low, preferably two-dimensional embedding.
-Reports on single-cell gene expression data, our running example, often start with an overview of the cell landscape and use t-SNE to embed high-dimensional expression profiles into a two-dimensional space.
-Figs.~\ref{fig:macosko}.a and \ref{fig:macosko}.b show two such embeddings.
+Among these, t-SNE has achieved widespread adoption as it can address high volumes of data and reveal the underlying data manifold.
+For instance, t-SNE is widely used in the bioinformatics community in areas such as single-cell transcriptomics~\citep{macosko2015highly,cao2019single,tasic2018shared}, human genetics~\citep{hirata2019genetic}, metagenomic assembly~\citep{beaulaurier2018metagenomic}, spatial organization of microbial communities~\citep{sheth2019spatial}, and metabolomics~\citep{tkachev2019differences}.
+To visually explore the data manifold, the objects of interest, such as diseased and healthy tissues or single cells, are profiled through thousands of features that include, for example, the expression of genes or the concentration of metabolites.
+The objective of t-SNE is to embed data points within a low-dimensional space, where the preservation of distances is rewarded in a non-linear fashion, thus resulting in a compact grouping of highly similar data points.
+Reports on single-cell gene expression data, where increasingly specific cell types are distinguished by a decreasing number of genes,
+often rely on t-SNE to embed high-dimensional expression profiles into a two-dimensional space and discover constituent cell types (Fig.~\ref{fig:macosko}.a).
 
@@
 We use \opentsne to generate three t-SNE embeddings and demonstrate recent theoretical advances.
-The data in \textbf{(a)} and \textbf{(b)} includes 44,808 mouse retinal cells (dots in the plots) that are described with high-dimensional gene-expression profiles from \citet{macosko2015highly}.
+The data in \textbf{(a)} and \textbf{(b)} include 44,808 mouse retinal cells that are described with high-dimensional gene-expression profiles from \citet{macosko2015highly}.
 The data in \textbf{(c)} additionally contains 27,499 expression profiles from mouse retinal cells from \citet{shekhar2016comprehensive}.
@@
 
-Despite its utility, t-SNE has often been criticized for its limited scalability, lack of global organization -- t-SNE identifies well-defined clusters that may be arbitrarily scattered throughout the embedding, and the absence of theoretically-founded methods to map new data into existing embeddings~\citep{ding2018interpretable,becht2019dimensionality}.
+Despite its utility, t-SNE has often been criticized for its limited scalability, lack of global organization -- t-SNE identifies well-defined clusters that may be arbitrarily scattered throughout the embedding -- and the absence of theoretically-founded methods to map new data into existing embeddings~\citep{ding2018interpretable,becht2019dimensionality}.
 Most of these shortcomings have recently been addressed.
-\citet{linderman2019fast} developed FIt-SNE, an efficient approximation scheme that massively improves the scalability of t-SNE, achieving linear time complexity in the number of samples.
-\citet{kobak2019art} proposed several techniques to improve global cluster coherence, including estimating similarities with a mixture of Gaussian kernels.
-In our previous work, we introduced a principled approach for embedding new samples into existing visualizations~\citep{policar2021embedding}.
+\citet{linderman2019fast} developed FIt-SNE, an efficient approximation that massively improves the scalability of t-SNE, achieving linear time complexity in the number of samples.
+\citet{kobak2019art} proposed several techniques to improve global cluster coherence (Fig.
+\ref{fig:macosko}.b), including estimating similarities with a mixture of Gaussian kernels.
+In our previous work, we introduced a principled approach for embedding new samples into existing visualizations (Fig.~\ref{fig:macosko}.c and~\citet{policar2021embedding}).
 
-However, easy access to efficient implementations almost universally correlates with the widespread adoption of novel computational techniques.
+Easy access to efficient implementations almost universally correlates with the widespread adoption of novel computational techniques.
 Despite the many theoretical advancements, most popular t-SNE libraries have been slow to incorporate them into their implementations.
@@
 dimensionality reduction method that to aims to find a low-dimensional embedding
-where neighborhoods are preserved. More formally, given a multi-dimensional data set
+where neighborhoods are preserved. Given a multi-dimensional data set with $N$ data points
 $\mathbf{X} = \left \{ \mathbf{x}_1, \mathbf{x}_2, \dots, \mathbf{x}_N \right \}
-\in \mathbb{R}^D$ where $N$ is the number of data points in the data set, t-SNE
+\in \mathbb{R}^D$, t-SNE
 aims to find a low dimensional embedding $\mathbf{Y} = \left \{ \mathbf{y}_1,
@@
 
-A direct evaluation of t-SNE gradients requires $\mathcal{O}(N^2)$ operations,
-which makes its application impractical to any reasonably-sized data set and
-highlights the need for the development of efficient approximation schemes.
+Evaluation of t-SNE gradients requires $\mathcal{O}(N^2)$ operations,
+which makes it impractical with large data sets and
+calls for efficient approximation schemes.
 \citet{van2014accelerating} observed that the t-SNE gradient can be cast as an
@@
 \citet{van2014accelerating} observed that evaluating the attractive forces
-between all pairs of data points is excessive, and that considering only a
-handful of nearest neighbors at each point are sufficient to obtain a good
-approximation. Therefore, instead of calculating all pairwise interactions, it
-is sufficient to only find and evaluate the attraction to a certain number of
+between all pairs of data points is excessive. In practice, it 
+suffices to approximate the attraction forces against a small number of
 $k$ nearest neighbors. Using tree-based nearest-neighbor search methods,
 \citet{van2014accelerating} reduced the time complexity to $\mathcal{O}(N \log
-N)$. \citet{linderman2019fast} further realized that, qualitatively, embeddings
-are visually indistinguishable when using only \textit{approximate} nearest
+N)$. \citet{linderman2019fast} noted that embeddings
+are visually indistinguishable when using merely \textit{approximate} nearest
 neighbors, further reducing time complexity to $\mathcal{O}(N)$.
@@
 The t-SNE algorithm is non-parametric and does not define an explicit mapping from the
-high-dimensional space to the embedding space. Therefore embeddings of new data
+high-dimensional space to the embedding space. Embeddings of new data
 points need to be found through optimization
-techniques~\citep{policar2021embedding}. When adding new data points to an
+schemes~\citep{policar2021embedding}. When adding new data points to an
 existing, reference embedding, the reference data points are fixed in place
@@
 \item The optimization in t-SNE starts with initial positions for each data point in the embedding space.
-We can set the initial positions randomly or impose some initial global structure on the embedding with, for example, a PCA-based initialization.
+We can set the initial positions randomly or impose an initial global structure on the embedding with, for example, a PCA-based initialization.
 Accordingly, the \code{openTSNE.initialization} module provides several possible initialization schemes, The user can also specify their initialization by providing a \code{numpy.ndarray} object.
@@
 
-While \opentsne allows for control at every step, most end-users often only need a quick and easy way to create informative visualizations of their data sets.
+Most end-users often only need a quick and easy way to create informative visualizations of their data sets.
 \opentsne provides a high-level API that closely follows the style of \pkg{scikit-learn}~\citep{sklearn_api}, a popular and widely used machine learning toolkit.
@@
 
-Exaggeration can highlight transitions between cell states in developmental studies.
+Exaggeration can highlight transitions between cell stages in developmental studies.
 Standard t-SNE often produces embeddings with clearly defined, discrete clusters.
 We can adjust the level of granularity and resolution of the clusters with several parameters, as shown in Fig.~\ref{fig:tasic}.
-However, discrete clusters are often undesired in developmental studies where cells' state is assumed to follow a continuous transition path.
-To this end, researchers have used other embedding techniques such as UMAP and ForceAtlas2 to better capture the continuity between cell states.
+However, discrete clusters are often undesired in developmental studies where cells' stage is assumed to follow a continuous transition path.
+To this end, researchers have used other embedding techniques such as UMAP and ForceAtlas2 to better capture the continuity between cell stages.
 Recently, \citet{bohm2020unifying} showed that embeddings produced by t-SNE with exaggeration values of 4 and $\sim30$ construct embeddings that are markedly similar to UMAP and ForceAtlas2, respectively.
@@
 
-Versatility, the ability to use and combine different optimization approaches to construct different embedding spaces, is another of \opentsne's core design principles.
+The ability to use and combine different optimization approaches to construct different embedding spaces is another of \opentsne's core design principles.
 \citet{kobak2019art} recently provided several recommendations and tricks to obtain better and more meaningful t-SNE visualizations.
@@
 \opentsne provides a flexible API to incorporate these improvements in just a few lines of code.
-Furthermore, \opentsne supports custom affinity models, enabling users to construct t-SNE embeddings on non-tabular relational data: the only requirement imposed by the affinity model is some notion of similarity between data points.
+Furthermore, \opentsne supports custom affinity models, enabling users to construct t-SNE embeddings on non-tabular relational data: the only requirement imposed by the affinity model is a notion of similarity between data points.
 Finally, \opentsne's comprehensive callback system can be utilized to monitor and adapt different stages of the optimization phase and has been used to construct visually appealing animations of the t-SNE optimization process.
@@
 One of the t-SNE's common criticisms is limited scalability to large data sets containing, for instance, millions of data points~\citep {becht2019dimensionality}.
-The culprit for slow response time stems from the optimization procedures and their specific implementation in popular open-source libraries.
-Until quite recently, most popular implementations of t-SNE were based on either a direct implementation of the t-SNE algorithm with asymptotic time complexity $\mathcal{O}(N^2)$~\citep{maaten2008visualizing}, or the Barnes-Hut approximation scheme developed by \citet{van2014accelerating} with asymptotic time complexity $\mathcal{O}(N \log N)$.
-Recently, however, \citet{linderman2019fast} developed a novel approximation scheme -- FIt-SNE -- which reduces the asymptotic time complexity to $\mathcal {O}(N)$, overcoming the difficulties of scaling t-SNE to large data sets.
+Slow response times stem from the optimization procedures and their specific implementation in popular open-source libraries.
+Most implementations of t-SNE have been based on either a direct implementation of the t-SNE algorithm with asymptotic time complexity $\mathcal{O}(N^2)$~\citep{maaten2008visualizing}, or the Barnes-Hut approximation with asymptotic time complexity $\mathcal{O}(N \log N)$~\citep{van2014accelerating}.
+Recently, however, \citet{linderman2019fast} developed a novel approximation -- FIt-SNE -- which reduces the asymptotic time complexity to $\mathcal {O}(N)$, enabling scaling to large data sets.
 
@@
 To generate benchmark data sets of different sample sizes, we subsample the data set five times at each sample size.
-In total, we run each implementation on 30 different data sets.
+In total, we test each implementation on 30 different data sets.
 

### 2022-05-23 paper/main.tex
Add missing references

@@
 Benchmarks were run for $1,000$ iterations with the original t-SNE parameters, as some implementations do not allow for their modification.
-To ensure a realistic benchmarking scenario, we utilize the 10X Genomics 1.3 million mouse brain data set~\citep{xxx}, containing 1.3 million single-cell gene-expression profiles.
-We preprocess the data using the standard single-cell analysis pipeline~\citep{xxx} and extract the top 50 principal components.
+To ensure a realistic benchmarking scenario, we utilize the 10X Genomics 1.3 million mouse brain data set~\citep{cao2019single}, containing 1.3 million single-cell gene-expression profiles.
+We preprocess the data using the standard single-cell analysis pipeline~\citep{kiselev2019challenges} and extract the top 50 principal components.
 To generate benchmark data sets of different sample sizes, we subsample the data set five times at each sample size.

### 2022-05-23 benchmarks/run_python.sh
benchmarks: clone fitsne using https instead of git

@@ -14,7 +14,7 @@ pip install -r requirements-benchmarks.txt
 pip install --upgrade numpy
 
 # Setup FIt-SNE
-git clone git@github.com:KlugerLab/FIt-SNE.git || :
+git clone https://github.com/KlugerLab/FIt-SNE.git || :
 conda install -c conda-forge fftw -y
 cd FIt-SNE
 CPLUS_INCLUDE_PATH="${CONDA_PREFIX}/include/" LIBRARY_PATH="${CONDA_PREFIX}/lib/" \

### 2022-07-07 paper/main.tex
Improve cosmetics in the manuscript

@@
 
-\subsection{Efficient Approximation Schemes} \label{sec:meth.approx}
+\subsection{Efficient approximation schemes} \label{sec:meth.approx}
 
@@
 N-body problem where data points represent particles that attract and repel each
-other. The gradient from Eqn.~(\ref{eq:tsne_gradient}) can be rewritten as
+other. The gradient from Equation~(\ref{eq:tsne_gradient}) can be rewritten as
 \begin{equation}
@@
 
-\subsubsection*{Attractive Forces}
+\subsubsection*{Attractive forces}
 \citet{van2014accelerating} observed that evaluating the attractive forces
@@
 
-\subsubsection*{Repulsive Forces}
+\subsubsection*{Repulsive forces}
 Similarly, the repulsive term can also be approximated, motivated by methods from
@@
 
-\subsection{Embedding New Samples} \label{sec:meth.transform}
+\subsection{Embedding new samples} \label{sec:meth.transform}
 
@@
 \mathbf{w}_M \right \} \in \mathbb{R}^d$. Additionally, we omit the
-symmetrization step in Eqn.~(\ref{eq:symmetrize}). Plugging these terms into
-Eqn.~(\ref{eq:kl_divergence}), we obtain the following gradient
+symmetrization step in Equation~(\ref{eq:symmetrize}). Plugging these terms into
+Equation~(\ref{eq:kl_divergence}), we obtain the following gradient
 \begin{equation}
@@
 
-\subsection{Improving the Global Structure of Embeddings} \label{sec:meth.global}
+\subsection{Improving the global structure of embeddings} \label{sec:meth.global}
 
@@
 the clustering structure at different levels of resolution. Modifying
-Eqn.~(\ref{eq:cauchy_kernel}) to $q_{ij} \propto \left ( 1 + || \mathbf{y}_i -
+Equation~(\ref{eq:cauchy_kernel}) to $q_{ij} \propto \left ( 1 + || \mathbf{y}_i -
 \mathbf{y}_j ||^2 / \alpha \right )^{-\alpha}$ allows us to use heavier-tailed
@@
 \item The t-SNE algorithm requires a specification of an affinity model to describe the similarities between data points.
-Standard t-SNE predominantly deals with tabular data and computes similarities using Eqn.~\ref{eq:p_ij}.
+Standard t-SNE predominantly deals with tabular data and computes similarities using Equation~\ref{eq:p_ij}.
 However, \citet{kobak2019art} showed that using mixtures of Gaussians instead of a single Gaussian often reveals better global consistency of the resulting visualization (albeit at some computational cost).
@@
 Because this new embedding relies on an existing reference embedding, we construct a \code{openTSNE.PartialTSNEEmbedding} embedding object, which indicates its dependence on a reference \code{openTSNE.TSNEEmbedding} object.
-As before, \code{openTSNE.PartialTSNEEmbedding} inherits from \code{numpy.ndarray} and can therefore be easily manipulated.
-Optimization is, again, performed via the \code{.optimize} method.
+As before, \\
+\code{openTSNE.PartialTSNEEmbedding} inherits from \code{numpy.ndarray} and can therefore be easily manipulated. Optimization is, again, performed via the \code{.optimize} method.
 \end{enumerate}
@@
 
-\opentsne is designed to be easy to use and can be installed through either Python Package Index \pkg{PyPI} using package installer \pkg{pip},
+\opentsne is designed to be easy to use and can be installed through either \proglang{Python} Package Index \pkg{PyPI} using package installer \pkg{pip},
 \begin{CodeChunk}
@@
 
-To verify that that \opentsne was installed correctly, users can run Python and try to import the module by:
+To verify that that \opentsne was installed correctly, users can run \proglang{Python} and try to import the module by:
 \begin{CodeChunk}
@@
 \begin{CodeInput}
->>> affinities = openTSNE.affinity.PerplexityBasedNN(iris.data, perplexity=30)
+>>> affinities = openTSNE.affinity.PerplexityBasedNN(
+...     iris.data, perplexity=30
+... )
 \end{CodeInput}
@@
 
-\section{Case Studies} \label{sec:discussion}
+\section{Case studies} \label{sec:discussion}
 
-\subsection{Uncovering Structure in High-Dimensional Data}
+\subsection{Uncovering structure in high-dimensional data}
 
@@
 
-\subsection{Embedding Massive Data Sets}
+\subsection{Embedding massive data sets}
 
@@
 # To improve the global structure of the final layout, we initialize the
-# embedding by creating a t-SNE plot on a sample of points
+# embedding by creating a t-SNE plot on a sample of points.
 
-# Take a random sample of the data points
+# Take a random sample of the data points.
 indices = np.random.permutation(list(range(x.shape[0])))
@@
 # Create a t-SNE embedding on the sample, using a high perplexity value to
-# emphasize the high level structure
+# emphasize the high level structure.
 init_sample = openTSNE.TSNE(perplexity=500).fit(x_sample)
 
-# We can next use the `transform` functionality to determine starting positions
-# for the remaining data points. Since we don't actually want to optimize the
-# remaining points separately from the sample points, we don't need to call the
-# `transform` function. Instead, we'll call the `prepare_partial` function,
-# which `transform` calls anyways under the hood.
+# We can next use the `transform` functionality to determine starting
+# positions for the remaining data points. Since we don't actually want to
+# optimize the remaining points separately from the sample points, we don't
+# need to call the `transform` function. Instead, we'll call the
+# `prepare_partial` function, which `transform` calls anyways under the hood.
 init_rest = init_sample.prepare_partial(x_rest)
 
-# Lastly, we merge `init_sample` and `init_rest` and restore the original order
+# Lastly, we merge `init_sample` and `init_rest` and restore the original
+# ordering.
 init_full = np.vstack((init_sample, init_rest))[reverse]
@@
 # This way, we can reuse the precomputed affinities in various different
-# embeddings, without having to repeat the same, costly computation every time.
+# embeddings, without having to repeat the same, costly computation
+# every time.
 affinities = openTSNE.affinity.PerplexityBasedNN(x, perplexity=30)
 
-# Create the TSNEEmbedding object using our previously calculated initialization
-# and precomputed affinities
+# Create the TSNEEmbedding object using our previously calculated
+# initialization and precomputed affinities.
 embedding = openTSNE.TSNEEmbedding(init_full, affinities)
@@
 
-\subsection{Embedding New Samples}
+\subsection{Embedding new samples}
 
@@
 # We specify the old default parameter values which are often used to show
-# why t-SNE misbehaves or fails to preserve global structure
+# why t-SNE misbehaves or fails to preserve global structure.
 tsne_old = openTSNE.TSNE(
@@
 # By default, openTSNE uses PCA-based initialization and automatically
-# determines the best learning rate. This also allows us to lower the default
-# number of iterations to 500, lowering the overall runtime
+# determines the best learning rate. This also allows us to lower the
+# default number of iterations to 500, lowering the overall runtime
 tsne_standard = openTSNE.TSNE(perplexity=[50, 500]).fit(x)
@@
 
-\subsection{Ease of Use}
+\subsection{Ease of use}
 
@@
 We provide precompiled binaries for all major \proglang{Python} versions on all major platforms, making the installation process as seamless as possible.
-One can install \opentsne through the \proglang{Python} Package Index (\pkg{PyPI}) or \pkg{conda} from the \pkg{conda-forge} channel, the two most widely adopted Python package managers.
+One can install \opentsne through the \proglang{Python} Package Index (\pkg{PyPI}) or \pkg{conda} from the \pkg{conda-forge} channel, the two most widely adopted \proglang{Python} package managers.
 \opentsne's high-level interface is inspired by \pkg{scikit-learn}, which is well established in the \proglang{Python} data science ecosystem.
@@
 \toprule
-\textbf{Ease of Installation} \\
+\textbf{Ease of installation} \\
 Native \proglang{Python} installation & \checkmark & & & \checkmark & & \\
@@
 \hline
-\textbf{Approximation Schemes} \\
+\textbf{Approximation schemes} \\
 None ($\mathcal{O}(N^2)$) & & & & & & \checkmark \\
@@
 \hline
-\textbf{Advanced Features and Extensions} \\
+\textbf{Advanced features and extensions} \\
 %Multiscale Gaussian kernels & & & & \checkmark & \checkmark & & \\
@@
 \hline
-\textbf{Project Quality}\\
+\textbf{Project quality}\\
 Actively maintained & \checkmark & & & \checkmark & \checkmark & \checkmark \\
@@
 
-\section*{Availability and Documentation}
+\section*{Availability and documentation}
 

### 2022-07-07 figures/download_h5ad.sh
download_h5ad: fix untar into correct directory

@@ -27,5 +27,5 @@ Extracting files...
 EOF
 
 for f in "$H5AD_DIR"/*.h5ad.tar.gz; do
-    tar xvfa "$f" --remove-files
+    tar xvfa "$f" -C "$H5AD_DIR" --remove-files
 done

### 2022-09-26 benchmarks/run_julia.sh
benchmarks: fixup double quotes in bash scripts

@@ -28,7 +28,7 @@ SAMPLE_SIZES=(1000 5000 10000 20000 50000);
 REPETITIONS=6;
 
 # Single-threaded benchmarks
-for size in ${SAMPLE_SIZES[@]}; do
+for size in "${SAMPLE_SIZES[@]}"; do
     cmd="OMP_NUM_THREADS=1 \
         julia benchmark.jl \
         --repetitions $REPETITIONS \

### 2022-09-26 benchmarks/run_python.sh

@@ -40,8 +40,8 @@ REPETITIONS=6;
 # Single-threaded benchmarks
 METHODS=(openTSNEBH openTSNEFFT MulticoreTSNE FItSNE sklearn UMAP);
 
-for method in ${METHODS[@]}; do
-    for size in ${SAMPLE_SIZES[@]}; do
+for method in "${METHODS[@]}"; do
+    for size in "${SAMPLE_SIZES[@]}"; do
         cmd="OMP_NUM_THREADS=1 NUMBA_NUM_THREADS=1 \
             python benchmark.py $method run_multiple \
             --n $REPETITIONS \
@@ -54,8 +54,8 @@ done;
 
 # Multi-threaded benchmarks
 METHODS=(openTSNEBH_8core openTSNEFFT_8core MulticoreTSNE_8core FItSNE_8core sklearn_8core UMAP_8core);
-for method in ${METHODS[@]}; do
-    for size in ${SAMPLE_SIZES[@]}; do
+for method in "${METHODS[@]}"; do
+    for size in "${SAMPLE_SIZES[@]}"; do
         cmd="OMP_NUM_THREADS=8 NUMBA_NUM_THREADS=8 \
             python benchmark.py $method run_multiple \
             --n $REPETITIONS \

### 2022-09-26 benchmarks/run_r.sh

@@ -28,7 +28,7 @@ REPETITIONS=6;
 
 # Single-threaded benchmarks
 SAMPLE_SIZES=(1000 100000 250000 500000 750000);
-for size in ${SAMPLE_SIZES[@]}; do
+for size in "${SAMPLE_SIZES[@]}"; do
     cmd="OMP_NUM_THREADS=1 \
         Rscript benchmark.r \
         --repetitions $REPETITIONS \
@@ -41,7 +41,7 @@ done;
 
 # Multi-threaded benchmarks
 SAMPLE_SIZES=(1000 100000 250000 500000 750000 1000000);
-for size in ${SAMPLE_SIZES[@]}; do
+for size in "${SAMPLE_SIZES[@]}"; do
     cmd="OMP_NUM_THREADS=8 \
         Rscript benchmark.r \
         --repetitions $REPETITIONS \

### 2022-09-26 benchmarks/convert_pickle_to_csv.py
benchmarks: convert_pickle_to_csv now uses command line arguments

@@ -1,17 +1,32 @@
 """
 To run benchmarks in non-Python programming languages, we need to convert the
 pickle file to something these languages can read. We choose CSV.
+
+The pickle file must adhere to the format:
+{
+    "pca_50": np.ndarray  # the data matrix,
+    ...
+}
+
 """
+import argparse
 import gzip
-import pickle
-from os import path
-
 import numpy as np
+import os
+import pickle
+import sys
 
-CSV_FNAME = path.join("data", "10x_mouse_zheng.csv")
+parser = argparse.ArgumentParser(description="Convert a .pkl.gz file to csv")
+parser.add_argument("-i", "--input", help="Input file", required=True)
+parser.add_argument("-o", "--output", help="Output file", required=True)
+parser.add_argument("-f", "--force", help="Force overwrite output file", action="store_true")
+args = parser.parse_args()
 
-if not path.exists(CSV_FNAME):
-    with gzip.open(path.join("data", "10x_mouse_zheng.pkl.gz"), "rb") as f:
-        data = pickle.load(f)
+if os.path.exists(args.output) and not args.force:
+    print(f"File `{args.output}` exists. Doing nothing.")
+    sys.exit(0)
 
-    np.savetxt(CSV_FNAME, data["pca_50"])
+with gzip.open(args.input, "rb") as f:
+    data = pickle.load(f)
+
+np.savetxt(args.output, data["pca_50"])

### 2022-09-26 benchmarks/run_julia.sh

@@ -18,6 +18,9 @@ julia -e 'using Pkg; Pkg.add("TSne")'
 mkdir -p data
 wget -nc -P data http://file.biolab.si/opentsne/benchmark/10x_mouse_zheng.pkl.gz
 
+conda install -y python numpy
+python convert_pickle_to_csv.py -i data/10x_mouse_zheng.pkl.gz -o data/10x_mouse_zheng.csv
+
 # Prepare logs directory
 mkdir -p logs
 

### 2022-09-26 benchmarks/run_r.sh

@@ -16,7 +16,7 @@ mkdir -p data
 wget -nc -P data http://file.biolab.si/opentsne/benchmark/10x_mouse_zheng.pkl.gz
 
 conda install -y python numpy
-python convert_pickle_to_csv.py
+python convert_pickle_to_csv.py -i data/10x_mouse_zheng.pkl.gz -o data/10x_mouse_zheng.csv
 
 # Prepare logs directory
 mkdir -p logs

### 2022-09-26 benchmarks/run_julia.sh
benchmarks: remove prefix from generated environment files

@@ -24,7 +24,7 @@ python convert_pickle_to_csv.py -i data/10x_mouse_zheng.pkl.gz -o data/10x_mouse
 # Prepare logs directory
 mkdir -p logs
 
-conda env export > logs/00--run_julia--conda_env.yml
+conda env export | sed '/prefix:/d' > logs/00--run_julia--conda_env.yml
 
 # RUN BENCHMARKS
 SAMPLE_SIZES=(1000 5000 10000 20000 50000);

### 2022-09-26 benchmarks/run_python.sh

@@ -31,7 +31,7 @@ wget -nc -P data http://file.biolab.si/opentsne/benchmark/10x_mouse_zheng.pkl.gz
 # Prepare logs directory
 mkdir -p logs
 
-conda env export > logs/00--run_python--conda_env.yml
+conda env export | sed '/prefix:/d' > logs/00--run_python--conda_env.yml
 
 # RUN BENCHMARKS
 SAMPLE_SIZES=(1000 100000 250000 500000 750000 1000000);

### 2022-09-26 benchmarks/run_r.sh

@@ -21,7 +21,7 @@ python convert_pickle_to_csv.py -i data/10x_mouse_zheng.pkl.gz -o data/10x_mouse
 # Prepare logs directory
 mkdir -p logs
 
-conda env export > logs/00--run_r--conda_env.yml
+conda env export | sed '/prefix:/d' > logs/00--run_r--conda_env.yml
 
 # RUN BENCHMARKS
 REPETITIONS=6;

### 2022-09-26 benchmarks/convert_pickle_to_csv.py
benchmarks: add info message to generate csv

@@ -26,6 +26,7 @@ if os.path.exists(args.output) and not args.force:
     print(f"File `{args.output}` exists. Doing nothing.")
     sys.exit(0)
 
+print(f"Generating `{args.output}`...")
 with gzip.open(args.input, "rb") as f:
     data = pickle.load(f)
 

### 2022-09-26 benchmarks/convert_csv_to_rds.r
benchmarks: add conversion script to rds

@@ -1,0 +1,20 @@
+#!/usr/bin/env Rscript
+library(optparse)
+
+option_list = list(
+    make_option(c("-i", "--input"), type="character"),
+    make_option(c("-o", "--output"), type="character"),
+    make_option(c("-f", "--force"), default=FALSE, action="store_true")
+)
+
+opt_parser = OptionParser(option_list=option_list)
+opt = parse_args(opt_parser)
+
+if (file.exists(opt$output) && !opt$force) {
+    cat(sprintf("File `%s` exists. Doing nothing.", opt$output))
+    quit()
+}
+
+cat(sprintf("Generating `%s`...", opt$output))
+data = read.csv(opt$input, sep=" ", header=F)
+saveRDS(data, opt$output)

### 2022-09-27 benchmarks/benchmark.r
benchmarks: fixup r script

@@ -6,22 +6,15 @@ library(Rtsne)
 sessionInfo()
 
 option_list = list(
+    make_option(c("--fname"), type="character"),
     make_option(c("--repetitions"), type="integer", default=1),
     make_option(c("--n-samples"), type="integer", default=FALSE),
     make_option(c("--n-threads"), type="integer", default=1)
 )
-
 opt_parser = OptionParser(option_list=option_list)
 opt = parse_args(opt_parser)
 
-
-# Load data from RDS. If not exists, generate RDS
-RDS_FNAME = file.path("data", "10x_mouse_zheng.rds")
-if (!file.exists(RDS_FNAME)) {
-    data = read.csv(file.path("data", "10x_mouse_zheng.csv"), sep=" ", header=F)
-    saveRDS(data, RDS_FNAME)
-}
-data = readRDS(RDS_FNAME)
+data = readRDS(opt$fname)
 
 for (i in 1:opt$repetitions) {
     set.seed(i)

### 2022-09-27 benchmarks/convert_csv_to_rds.r

@@ -11,7 +11,7 @@ opt_parser = OptionParser(option_list=option_list)
 opt = parse_args(opt_parser)
 
 if (file.exists(opt$output) && !opt$force) {
-    cat(sprintf("File `%s` exists. Doing nothing.", opt$output))
+    cat(sprintf("File `%s` exists. Doing nothing.\n", opt$output))
     quit()
 }
 

### 2022-09-27 benchmarks/benchmark.py
benchmarks: bundle benchmarks.py in package

@@ -1,0 +1,285 @@
+import gzip
+import pickle
+import signal
+import time
+from contextlib import contextmanager
+
+import fire
+import numpy as np
+from MulticoreTSNE import MulticoreTSNE as MulticoreTSNE_
+from sklearn.manifold import TSNE as SKLTSNE
+from sklearn.utils import check_random_state
+
+import openTSNE
+import openTSNE.callbacks
+
+
+def raise_timeout(signum, frame):
+    raise TimeoutError
+
+
+@contextmanager
+def _timeout(time):
+    """Limit runtime of a particular chunk of code.
+
+    Parameters
+    ----------
+    time: int
+        Time in minutes
+
+    """
+    # Register a function to raise a TimeoutError on the signal.
+    signal.signal(signal.SIGALRM, raise_timeout)
+    # Schedule the signal to be sent after ``time``.
+    signal.alarm(int(time * 60))
+
+    try:
+        yield
+    except TimeoutError:
+        raise TimeoutError(f"Function call timed out after {time} minutes!")
+    finally:
+        # Unregister the signal, so it won't be triggered
+        # if the timeout is not reached.
+        signal.signal(signal.SIGALRM, signal.SIG_IGN)
+
+
+class TSNEBenchmark:
+    perplexity = 30
+    learning_rate = 200
+
+    def run(self, fname: str, n_samples=1000, repetitions=1, n_jobs=1, timeout=120):
+        x, y = self.load_data(fname=fname, n_samples=n_samples)
+
+        for idx in range(repetitions):
+            with _timeout(timeout):
+                self._run(x, random_state=idx, n_jobs=n_jobs)
+
+    def _run(self, data, random_state=None, n_jobs=1):
+        raise NotImplementedError()
+
+    def load_data(self, fname: str, n_samples: int = None):
+        with gzip.open(fname, "rb") as f:
+            data = pickle.load(f)
+
+        x, y = data["pca_50"], data["CellType1"]
+        print(f"Full data set dimensions: {x.shape}")
+
+        if n_samples is not None:
+            indices = np.random.choice(
+                list(range(x.shape[0])), n_samples, replace=False
+            )
+            x, y = x[indices], y[indices]
+
+        print(f"Benchmark data set dimensions: {x.shape}")
+        return x, y
+
+
+class openTSNENNDescent(TSNEBenchmark):
+    def _run(self, x, random_state=None, n_jobs=1):
+        print("-" * 80)
+        print("Random state", random_state)
[…]

### 2022-09-27 benchmarks/benchmark.jl
benchmarks: add fname option to julia script

@@ -8,6 +8,8 @@ using Pkg
 
 s = ArgParseSettings()
 @add_arg_table s begin
+    "--fname"
+        arg_type = String
     "--repetitions"
         arg_type = Int
         default = 1
@@ -19,7 +21,7 @@ end
 args = parse_args(s)
 
 
-dataset = CSV.File("data/10x_mouse_zheng.csv", delim=' ', header=false) |> DataFrame
+dataset = CSV.File(args["fname"], delim=' ', header=false) |> DataFrame
 data = Matrix{Float64}(dataset)
 
 

### 2022-09-27 benchmarks/run.sh
benchmarks: replace python/r/jl scripts with single run script

@@ -1,0 +1,214 @@
+#!/bin/bash
+
+set -e
+
+CONDA_ENVIRONMENT_NAME="tsne_benchmarks"
+
+is_installed() {
+  [ -x "$(command -v "$1")" ]
+}
+
+activate_environment() {
+  eval "$(conda shell.bash hook)"
+  conda activate "$1"
+}
+
+install_fitsne() {
+  # Setup FIt-SNE
+  git clone https://github.com/KlugerLab/FIt-SNE.git || :
+  conda install -c conda-forge fftw -y
+  cd FIt-SNE
+  CPLUS_INCLUDE_PATH="${CONDA_PREFIX}/include/" LIBRARY_PATH="${CONDA_PREFIX}/lib/" \
+      g++ -std=c++11 -O3  src/sptree.cpp src/tsne.cpp src/nbodyfft.cpp  \
+      -o bin/fast_tsne -pthread -lfftw3 -lm -Wno-address-of-packed-member
+  cd ..
+}
+
+install_python_dependencies() {
+  conda install -y python numpy scikit-learn
+
+  pip install opentsne --no-binary :all:
+  pip install -r requirements-benchmarks.txt
+
+  install_fitsne
+}
+
+install_r_dependencies() {
+  conda install -c r r -y
+  Rscript -e 'install.packages("Rtsne", repos="https://cloud.r-project.org")'
+  Rscript -e 'install.packages("optparse", repos="https://cloud.r-project.org")'
+}
+
+install_julia_dependencies() {
+  conda install -c conda-forge julia -y
+  julia -e 'using Pkg; Pkg.add("ArgParse")'
+  julia -e 'using Pkg; Pkg.add("CSV")'
+  julia -e 'using Pkg; Pkg.add("DataFrames")'
+  julia -e 'using Pkg; Pkg.add("StatsBase")'
+  julia -e 'using Pkg; Pkg.add("TSne")'
+}
+
+# Download benchmark datasets from remote
+# $1: dataset_name -- e.g. macosko_2015, which matches remote file name
+download_datasets() {
+  mkdir -p data
+
+  wget -nc -P data "http://file.biolab.si/opentsne/benchmark/$1.pkl.gz"
+  is_installed python && python convert_pickle_to_csv.py -i "data/$1.pkl.gz" -o "data/$1.csv"
+  is_installed Rscript && Rscript convert_csv_to_rds.r -i "data/$1.csv" -o "data/$1.rds"
+}
+
[…]

### 2022-09-27 benchmarks/run_julia.sh

@@ -1,42 +1,0 @@
-#!/bin/bash
-
-set -e
-
-conda create --name tsne_benchmarks_julia -y
-eval "$(conda shell.bash hook)"  # activate conda env inside script
-conda activate tsne_benchmarks_julia
-
-# Setup Julia environment
-conda install -c conda-forge julia -y
-julia -e 'using Pkg; Pkg.add("ArgParse")'
-julia -e 'using Pkg; Pkg.add("CSV")'
-julia -e 'using Pkg; Pkg.add("DataFrames")'
-julia -e 'using Pkg; Pkg.add("StatsBase")'
-julia -e 'using Pkg; Pkg.add("TSne")'
-
-# Download benchmark dataset
-mkdir -p data
-wget -nc -P data http://file.biolab.si/opentsne/benchmark/10x_mouse_zheng.pkl.gz
-
-conda install -y python numpy
-python convert_pickle_to_csv.py -i data/10x_mouse_zheng.pkl.gz -o data/10x_mouse_zheng.csv
-
-# Prepare logs directory
-mkdir -p logs
-
-conda env export | sed '/prefix:/d' > logs/00--run_julia--conda_env.yml
-
-# RUN BENCHMARKS
-SAMPLE_SIZES=(1000 5000 10000 20000 50000);
-REPETITIONS=6;
-
-# Single-threaded benchmarks
-for size in "${SAMPLE_SIZES[@]}"; do
-    cmd="OMP_NUM_THREADS=1 \
-        julia benchmark.jl \
-        --repetitions $REPETITIONS \
-        --n-samples $size 2>&1 \
-        | tee -a logs/TSne-jl_${size}.log";
-    echo "$cmd";
-    eval "$cmd";
-done;

### 2022-09-27 benchmarks/run_python.sh

@@ -1,67 +1,0 @@
-#!/bin/bash
-
-set -e
-
-conda create --name tsne_benchmarks -y
-eval "$(conda shell.bash hook)"  # activate conda env inside script
-conda activate tsne_benchmarks
-conda install -y python numpy scikit-learn
-
-pip install opentsne
-pip install -r requirements-benchmarks.txt
-
-# MulticoreTSNE forces lower version of numpy
-pip install --upgrade numpy
-
-# Setup FIt-SNE
-git clone https://github.com/KlugerLab/FIt-SNE.git || :
-conda install -c conda-forge fftw -y
-cd FIt-SNE
-CPLUS_INCLUDE_PATH="${CONDA_PREFIX}/include/" LIBRARY_PATH="${CONDA_PREFIX}/lib/" \
-    g++ -std=c++11 -O3  src/sptree.cpp src/tsne.cpp src/nbodyfft.cpp  \
-    -o bin/fast_tsne -pthread -lfftw3 -lm -Wno-address-of-packed-member
-cd ..
-
-# Fetch benchmarking module from openTSNE
-wget -nc https://raw.githubusercontent.com/pavlin-policar/openTSNE/master/benchmarks/benchmark.py
-# Download benchmark dataset
-mkdir -p data
-wget -nc -P data http://file.biolab.si/opentsne/benchmark/10x_mouse_zheng.pkl.gz
-
-# Prepare logs directory
-mkdir -p logs
-
-conda env export | sed '/prefix:/d' > logs/00--run_python--conda_env.yml
-
-# RUN BENCHMARKS
-SAMPLE_SIZES=(1000 100000 250000 500000 750000 1000000);
-REPETITIONS=6;
-
-# Single-threaded benchmarks
-METHODS=(openTSNEBH openTSNEFFT MulticoreTSNE FItSNE sklearn UMAP);
-
-for method in "${METHODS[@]}"; do
-    for size in "${SAMPLE_SIZES[@]}"; do
-        cmd="OMP_NUM_THREADS=1 NUMBA_NUM_THREADS=1 \
-            python benchmark.py $method run_multiple \
-            --n $REPETITIONS \
-            --n-samples $size 2>&1 \
-            | tee -a logs/${method}_${size}.log";
-        echo "$cmd";
-        eval "$cmd";
-    done;
-done;
-
-# Multi-threaded benchmarks
[…]

### 2022-09-27 benchmarks/run_r.sh

@@ -1,53 +1,0 @@
-#!/bin/bash
-
-set -e
-
-conda create --name tsne_benchmarks_r -y
-eval "$(conda shell.bash hook)"  # activate conda env inside script
-conda activate tsne_benchmarks_r
-
-# Setup R environment
-conda install -c r r -y
-Rscript -e 'install.packages("Rtsne", repos="https://cloud.r-project.org")'
-Rscript -e 'install.packages("optparse", repos="https://cloud.r-project.org")'
-
-# Download benchmark dataset
-mkdir -p data
-wget -nc -P data http://file.biolab.si/opentsne/benchmark/10x_mouse_zheng.pkl.gz
-
-conda install -y python numpy
-python convert_pickle_to_csv.py -i data/10x_mouse_zheng.pkl.gz -o data/10x_mouse_zheng.csv
-
-# Prepare logs directory
-mkdir -p logs
-
-conda env export | sed '/prefix:/d' > logs/00--run_r--conda_env.yml
-
-# RUN BENCHMARKS
-REPETITIONS=6;
-
-# Single-threaded benchmarks
-SAMPLE_SIZES=(1000 100000 250000 500000 750000);
-for size in "${SAMPLE_SIZES[@]}"; do
-    cmd="OMP_NUM_THREADS=1 \
-        Rscript benchmark.r \
-        --repetitions $REPETITIONS \
-        --n-threads 1 \
-        --n-samples $size 2>&1 \
-        | tee -a logs/Rtsne_${size}.log";
-    echo "$cmd";
-    eval "$cmd";
-done;
-
-# Multi-threaded benchmarks
-SAMPLE_SIZES=(1000 100000 250000 500000 750000 1000000);
-for size in "${SAMPLE_SIZES[@]}"; do
-    cmd="OMP_NUM_THREADS=8 \
-        Rscript benchmark.r \
-        --repetitions $REPETITIONS \
-        --n-threads 8 \
-        --n-samples $size 2>&1 \
-        | tee -a logs/Rtsne_8core_${size}.log";
-    echo "$cmd";
-    eval "$cmd";
-done;

### 2022-09-27 benchmarks/benchmark.py
benchmarks: fixup run script

@@ -6,7 +6,6 @@ from contextlib import contextmanager
 
 import fire
 import numpy as np
-from MulticoreTSNE import MulticoreTSNE as MulticoreTSNE_
 from sklearn.manifold import TSNE as SKLTSNE
 from sklearn.utils import check_random_state
 
@@ -77,6 +76,7 @@ class TSNEBenchmark:
 class openTSNENNDescent(TSNEBenchmark):
     def _run(self, x, random_state=None, n_jobs=1):
         print("-" * 80)
+        print(f"openTSNE v{openTSNE.__version__}")
         print("Random state", random_state)
         print("-" * 80, flush=True)
 
@@ -117,6 +117,7 @@ class openTSNENNDescent(TSNEBenchmark):
 class openTSNEBH(TSNEBenchmark):
     def _run(self, x, random_state=None, n_jobs=1):
         print("-" * 80)
+        print(f"openTSNE v{openTSNE.__version__}")
         print("Random state", random_state)
         print("-" * 80, flush=True)
 
@@ -157,6 +158,7 @@ class openTSNEBH(TSNEBenchmark):
 class openTSNEFFT(TSNEBenchmark):
     def _run(self, x, random_state=None, n_jobs=1):
         print("-" * 80)
+        print(f"openTSNE v{openTSNE.__version__}")
         print("Random state", random_state)
         print("-" * 80, flush=True)
 
@@ -196,6 +198,8 @@ class openTSNEFFT(TSNEBenchmark):
 
 class MulticoreTSNE(TSNEBenchmark):
     def _run(self, x, random_state=None, n_jobs=1):
+        from MulticoreTSNE import MulticoreTSNE as MulticoreTSNE_
+
         print("-" * 80)
         print("Random state", random_state)
         print("-" * 80, flush=True)
@@ -216,7 +220,8 @@ class MulticoreTSNE(TSNEBenchmark):
 
 class FItSNE(TSNEBenchmark):
     def _run(self, x, random_state=None, n_jobs=1):
-        import sys; sys.path.append("FIt-SNE")
+        import sys;
+        sys.path.append("FIt-SNE")
         from fast_tsne import fast_tsne
 
         print("-" * 80)

### 2022-09-27 benchmarks/run.sh

@@ -25,16 +25,16 @@ install_fitsne() {
 }
 
 install_python_dependencies() {
-  conda install -y python numpy scikit-learn
+  conda install python numpy scikit-learn -y
 
-  pip install opentsne --no-binary :all:
+  pip install opentsne --no-binary opentsne
   pip install -r requirements-benchmarks.txt
 
   install_fitsne
 }
 
 install_r_dependencies() {
-  conda install -c r r -y
+  conda install -c r -c conda-forge r -y
   Rscript -e 'install.packages("Rtsne", repos="https://cloud.r-project.org")'
   Rscript -e 'install.packages("optparse", repos="https://cloud.r-project.org")'
 }

### 2022-09-27 benchmarks/run.sh
benchmarks: allow julia install to fail

@@ -120,7 +120,7 @@ if ! conda env list | grep "\s*$CONDA_ENVIRONMENT_NAME\s*" >/dev/null 2>&1; then
   activate_environment "$CONDA_ENVIRONMENT_NAME"
   install_python_dependencies
   install_r_dependencies
-  install_julia_dependencies
+  install_julia_dependencies || true  # allowed to fail on mac
 else
   activate_environment "$CONDA_ENVIRONMENT_NAME"
 fi

### 2022-09-27 figures/manuscript_iris.py
figures: add iris example from manuscript, line by line

@@ -1,0 +1,56 @@
+"""
+This code is used in the manuscript and demonstrates the usage of openTSNE
+on a simple example. The recommended way to use this is to run ipython and
+paste in the code line-by-line. This way, the different types of objects and
+flow will be most apparent.
+"""
+import openTSNE
+openTSNE.__version__
+
+# Simple usage
+from sklearn import datasets
+
+iris = datasets.load_iris()
+iris.data[:3]
+
+import openTSNE
+embedding = openTSNE.TSNE().fit(iris.data)
+embedding[:3]
+
+new_data = iris.data[::3]
+new_embedding = embedding.transform(new_data)
+new_embedding[:3]
+
+# Advanced usage
+from sklearn import datasets
+
+iris = datasets.load_iris()
+iris.data[:3]
+
+import openTSNE
+
+affinities = openTSNE.affinity.PerplexityBasedNN(
+    iris.data, perplexity=30
+)
+
+initialization = openTSNE.initialization.pca(iris.data)
+initialization[:3]
+
+embedding = openTSNE.TSNEEmbedding(initialization, affinities)
+embedding[:3]
+
+embedding.optimize(250, exaggeration=12, momentum=0.5, inplace=True)
+embedding.optimize(500, momentum=0.8, inplace=True)
+embedding[:3]
+
+import numpy as np
+np.linalg.norm(openTSNE.TSNE().fit(iris.data) - embedding)
+
+new_embedding = embedding.prepare_partial(iris.data[::3], perplexity=5)
+new_embedding.optimize(
+    exaggeration=1.5, n_iter=250, learning_rate=0.1, momentum=0.8,
+    max_grad_norm=0.25, inplace=True
+)
+new_embedding[::3]
+
+np.linalg.norm(new_embedding - embedding.transform(iris.data[::3]))

### 2022-09-28 benchmarks/run.sh
benchmarks: remove timeout used for testing in py benchmarks

@@ -147,8 +147,7 @@ run_python_benchmarks() {
               --fname data/$dataset_name.pkl.gz \
               --repetitions $repetitions \
               --n-samples $size \
-              --n-jobs $cores \
-              --timeout 0.5 2>&1 \
+              --n-jobs $cores 2>&1 \
               | tee -a logs/${dataset_name}--${method}--${cores}_core--${size}_samples.log";
           echo "$cmd" | tr -s " ";
           eval "$cmd";

### 2022-09-28 benchmarks/benchmark.py
benchmarks: remove timetout from benchmark.py

@@ -46,12 +46,11 @@ class TSNEBenchmark:
     perplexity = 30
     learning_rate = 200
 
-    def run(self, fname: str, n_samples=1000, repetitions=1, n_jobs=1, timeout=120):
+    def run(self, fname: str, n_samples=1000, repetitions=1, n_jobs=1):
         x, y = self.load_data(fname=fname, n_samples=n_samples)
 
         for idx in range(repetitions):
-            with _timeout(timeout):
-                self._run(x, random_state=idx, n_jobs=n_jobs)
+            self._run(x, random_state=idx, n_jobs=n_jobs)
 
     def _run(self, data, random_state=None, n_jobs=1):
         raise NotImplementedError()

### 2022-09-28 benchmarks/benchmark.py
benchmark: set FIt-SNE to 1000 iterations

@@ -241,6 +241,7 @@ class FItSNE(TSNEBenchmark):
             initialization=init,
             perplexity=self.perplexity,
             stop_early_exag_iter=250,
+            max_iter=1000,
             early_exag_coeff=12,
             nthreads=n_jobs,
             seed=random_state,

### 2022-09-28 benchmarks/benchmark.py
benchmarks: fix learning rate in fitsne benchmark

@@ -240,6 +240,7 @@ class FItSNE(TSNEBenchmark):
             map_dims=2,
             initialization=init,
             perplexity=self.perplexity,
+            learning_rate=self.learning_rate,
             stop_early_exag_iter=250,
             max_iter=1000,
             early_exag_coeff=12,

### 2022-09-29 benchmarks/generate_figures.py
benchmarks: update figure script for new benchmark format

@@ -59,15 +59,13 @@ class Spec:
 
 def get_benchmarks(spec: Spec, directory: str) -> List[Dict]:
     # Create regex pattern to match file names
-    pattern = spec.log_name
-    if spec.cores > 1:
-        pattern += f"_{spec.cores}core"
-    pattern += r"_(\d+).log"
+    pattern = r"(.+)" + f"--{spec.log_name}--{spec.cores}_core--" + r"(\d+)_samples.log"
 
     data = []
     for f in os.listdir(directory):
         if match := re.findall(pattern, f):
-            n_samples = int(match[0])
+            dataset = match[0][0]
+            n_samples = int(match[0][1])
 
             times = spec.parser(os.path.join(directory, f))
             for t in times:
@@ -77,6 +75,7 @@ def get_benchmarks(spec: Spec, directory: str) -> List[Dict]:
                     "Num. Samples": n_samples,
                     "Time (sec)": t,
                     "Time (min)": t / 60,
+                    "Benchmark": dataset,
                 })
 
     return data
@@ -86,11 +85,11 @@ def get_benchmarks(spec: Spec, directory: str) -> List[Dict]:
 # openTSNE
 parse_opentsne = make_regex_func(r"openTSNE: Full (\d+\.\d+)")
 # FIt-SNE
-parse_fitsne = make_regex_func(r"FIt-SNE: (\d+\.\d+)")
+parse_fitsne = make_regex_func(r"FItSNE: (\d+\.\d+)")
 # MulticoreTSNE
-parse_multicore = make_regex_func(r"Multicore t-SNE: (\d+\.\d+)")
+parse_multicore = make_regex_func(r"MulticoreTSNE: (\d+\.\d+)")
 # scikit-learn
-parse_sklearn = make_regex_func(r"scikit-learn t-SNE: (\d+\.\d+)")
+parse_sklearn = make_regex_func(r"sklearn: (\d+\.\d+)")
 # UMAP
 parse_umap = make_regex_func(r"UMAP: (\d+\.\d+)")
 # Rtsne
@@ -155,7 +154,7 @@ def generate_benchmark_plot(colors, title, fname):
 
     agg = data.groupby(["Spec", "Name", "Num. Samples"])["Time (min)"].aggregate(["mean", "std"]).reset_index()
 
-    # Sort the labels so they appear in the correct order in the legend
+    # Sort the labels, so they appear in the correct order in the legend
[…]

### 2022-09-29 paper/main.tex
paper: update manuscript

@@
 The objective of t-SNE is to embed data points within a low-dimensional space, where the preservation of distances is rewarded in a non-linear fashion, thus resulting in a compact grouping of highly similar data points.
-Reports on single-cell gene expression data, where increasingly specific cell types are distinguished by a decreasing number of genes,
-often rely on t-SNE to embed high-dimensional expression profiles into a two-dimensional space and discover constituent cell types (Fig.~\ref{fig:macosko}.a).
+Reports on single-cell gene expression data, where increasingly specific cell types are distinguished by a decreasing number of genes, often rely on t-SNE to embed high-dimensional expression profiles into a two-dimensional space and discover constituent cell types (Fig.~\ref{fig:macosko}.a).
 
@@
 Despite the many theoretical advancements, most popular t-SNE libraries have been slow to incorporate them into their implementations.
-In this paper, we present \opentsne, our open-source \proglang{Python} implementation of t-SNE, complete with its recently proposed extensions.
+In \proglang{Python}, the most commonly used implementation of the t-SNE algorithm comes from \pkg{scikit-learn}~\citep{pedregosa2011scikit}, which suffers from poor scalability.
+Other implementations from the \proglang{R}~\citep{team2013r,krijthe2015rtsne} and \proglang{Julia}~\citep{bezanson2017julia,julia_tsne} programming languages suffer from similar drawbacks.
+\pkg{MulticoreTSNE}~\citep{Ulyanov2016} enjoys better scalability but is likewise limited when applied to data sets containing millions of data points.
+Additionally, none of these implementations include recently proposed extensions that produce more globally consistent embeddings.
+\pkg{FIt-SNE}~\citep{linderman2019fast} alleviates these problems and provides bindings to both the \proglang{Python} and \proglang{R} programming languages.
+However, \pkg{FIt-SNE} is difficult to install and distribute and provides an API largely inconsistent with the \proglang{Python} data science ecosystem.
+None of these existing implementations provide solutions to adding new samples into existing embeddings.
+To alleviate these problems, we present \opentsne, our open-source \proglang{Python} implementation of the t-SNE algorithm, complete with its recently proposed extensions.
 \opentsne is easy to install and includes precompiled binaries available through popular \proglang{Python} package managers.
@@
 >>> import openTSNE
->>> print(openTSNE.__version__)
+>>> openTSNE.__version__
 0.6.2
@@
 \section{Case studies} \label{sec:discussion}
+
+We provide three case studies demonstrating the usage of \opentsne in different settings using different combinations of hyperparameters to highlight different aspects of the given data sets.
+The first case study highlights how varying the perplexity and resolution parameters can reveal different aspects of the underlying topology of the studied data set.
+The second case study demonstrates how \opentsne may be used to create embeddings of massive data sets containing millions of data points and how its flexible API allows the quick and efficient construction of embeddings.
+The final case study illustrates how we can use existing t-SNE embeddings to add new samples into the embedding space.
+This enables us to reuse carefully annotated visualizations for the rapid characterization of new, unseen data points.
 
@@
 \url{https://github.com/pavlin-policar/opentsne-paper}.
+Detailed instructions are provided showing how to reproduce all materials and results shown in this manuscript.
+
+All case studies can easily be reproduced on a consumer-grade laptop computer in minutes using the accompanying scripts.
+Note, however, that the second case study features a large data set and may require up to several hours to complete.
+
+Scripts are included to reproduce the benchmarks of the different implementations of the t-SNE algorithm.
+However, please be aware that the full benchmark suite can take several days to complete due to the poor performance of other libraries.
+A smaller, illustratory benchmark suite is provided, more suitable for consumer-grade laptop computers, demonstrating the runtimes of the different t-SNE implementations.
+Please note, however, that the scaling advantages of \opentsne are diminished when running on smaller data sets and will not appear as significant as they would otherwise.
 

### 2022-10-19 paper/main.tex
minor changes

@@
 Easy access to efficient implementations almost universally correlates with the widespread adoption of novel computational techniques.
-Despite the many theoretical advancements, most popular t-SNE libraries have been slow to incorporate them into their implementations.
-In \proglang{Python}, the most commonly used implementation of the t-SNE algorithm comes from \pkg{scikit-learn}~\citep{pedregosa2011scikit}, which suffers from poor scalability.
-Other implementations from the \proglang{R}~\citep{team2013r,krijthe2015rtsne} and \proglang{Julia}~\citep{bezanson2017julia,julia_tsne} programming languages suffer from similar drawbacks.
-\pkg{MulticoreTSNE}~\citep{Ulyanov2016} enjoys better scalability but is likewise limited when applied to data sets containing millions of data points.
-Additionally, none of these implementations include recently proposed extensions that produce more globally consistent embeddings.
+Despite the many recent theoretical advancements, popular t-SNE libraries have slowly incorporated them into their implementations.
+In \proglang{Python}, the most commonly used implementation of the t-SNE algorithm comes from \pkg{scikit-learn}~\citep{pedregosa2011scikit}, which is nicely integrated with \proglang{Python} data science environment but suffers from poor scalability.
+Other implementations from the \proglang{R}~\citep{team2013r,krijthe2015rtsne} and \proglang{Julia}~\citep{bezanson2017julia,julia_tsne} programming languages also scale poorly (see Sec.~\ref{s:speed}).
+\pkg{MulticoreTSNE}~\citep{Ulyanov2016} enjoys better scalability but is likewise limited when applied to larger data sets.
+Additionally, none of these implementations include recently proposed extensions that can result in more globally consistent embeddings.
 \pkg{FIt-SNE}~\citep{linderman2019fast} alleviates these problems and provides bindings to both the \proglang{Python} and \proglang{R} programming languages.
 However, \pkg{FIt-SNE} is difficult to install and distribute and provides an API largely inconsistent with the \proglang{Python} data science ecosystem.
-None of these existing implementations provide solutions to adding new samples into existing embeddings.
-To alleviate these problems, we present \opentsne, our open-source \proglang{Python} implementation of the t-SNE algorithm, complete with its recently proposed extensions.
+None of the existing t-SNE implementations support adding new samples into existing embeddings.
+To alleviate all these problems, and to support newly proposed additions to t-SNE algorithms, we here present \opentsne, an open-source \proglang{Python} implementation of the t-SNE algorithm.
 \opentsne is easy to install and includes precompiled binaries available through popular \proglang{Python} package managers.
@@
 
-\subsection{Speed}
+\subsection{Speed\label{s:speed}}
 

### 2022-11-02 paper/main.tex
paper: reflect new benchmarks

--- 2022-09-29 paper/main.tex
@@
 % We benchmark Python
-Fig.~\ref{fig:benchmarks_py} benchmarks four popular \proglang{Python}~\citep{vanrossum1995python} t-SNE implementations, including \pkg{scikit-learn} (v1.0.1), \pkg{MulticoreTSNE} (v0.1), \pkg{FIt-SNE} (v1.1.0), and \opentsne (v0.6.2).
+Fig.~\ref{fig:benchmarks_py} benchmarks four popular \proglang{Python}~\citep{vanrossum1995python} t-SNE implementations, including \pkg{scikit-learn} (v1.1.2), \pkg{MulticoreTSNE} (v0.1), \pkg{FIt-SNE} (v1.1.0), and \opentsne (v0.6.2).
 Benchmarks were performed on an Intel(R) Xeon(R) CPU E5-1650 equipped with 128 GB of memory.
@@
   implementations from \pkg{scikit-learn}~\citep{pedregosa2011scikit}
-  (v1.0.1), \pkg{MulticoreTSNE}~\citep{Ulyanov2016} (v0.1), and
+  (v1.1.2), \pkg{MulticoreTSNE}~\citep{Ulyanov2016} (v0.1), and
   \pkg{FIt-SNE}~\citep{linderman2019fast} (v1.1.0). Experiments were run on a
@@
 \pkg{scikit-learn}, like its \proglang{C++} counterpart -- \pkg{MulticoreTSNE}~\citep{Ulyanov2016} , is based on the Barnes-Hut approximation scheme.
-However, as shown on Fig.~\ref{fig:benchmarks_py}, \pkg{scikit-learn} exhibits much longer runtimes than \pkg{MulticoreTSNE}.
-Newer implementations, such as \pkg{FIt-SNE} and \opentsne, both implement the FIt-SNE approximation scheme, making them suitable for the visualization of millions of data points.
-While both libraries implement the same algorithm, \opentsne emphasizes ease of use and extensibility and is primarily written in \proglang{Python}, which inevitably introduces some runtime overhead.
-This makes \opentsne marginally slower than its pure \proglang{C++} counterpart -- \pkg{FIt-SNE}.
-However, these differences disappear almost entirely when utilizing multi-threading, as shown in Fig.~\ref{fig:benchmarks_py}.
+However, as shown in Fig.~\ref{fig:benchmarks_py}, \pkg{scikit-learn} exhibits somewhat longer runtimes than \pkg{MulticoreTSNE}.
+Newer implementations, such as \pkg{FIt-SNE} and \opentsne, implement the FIt-SNE approximation scheme, making them suitable for the visualization of millions of data points.
+While both libraries implement the same algorithm, \opentsne emphasizes ease of use and extensibility and is primarily written in \proglang{Python}.
+Despite this, both libraries exhibit similar runtimes, with \opentsne even being marginally faster when utilizing multi-threading, as shown in Fig.~\ref{fig:benchmarks_py}.
 
@@
 
+We note here that the benchmarks in Figs.~\ref{fig:benchmarks_py}~and~\ref{fig:benchmarks_lang} were run for 1,000 iterations to ensure a fair comparison between different implementations.
+However, it has recently been shown that, using an appropriate learning rate, we can safely reduce the number of iterations to 750~\citep{belkina2019automated}.
+This means that the actual runtime will be faster than reported in these benchmarks in everyday usage of \opentsne.
+
 \subsection{Ease of use}
@@
 \opentsne implements multi-threaded versions of the Barnes-Hut and FIt-SNE approximation schemes, enabling it to be applied to data sets containing millions of data points.
-While the \proglang{Python} virtual machine inevitably introduces some performance overhead, the runtime is comparable to its \proglang{C++} counterpart -- \pkg{FIt-SNE}.
 Finally, \opentsne is extensible.

### 2023-05-22 2023-05-revision-1/main.tex
Setup revision-1 folder

--- 2022-11-02 paper/main.tex
@@
 N-body problem where data points represent particles that attract and repel each
-other. The gradient from Equation~(\ref{eq:tsne_gradient}) can be rewritten as
+other. The gradient from Equation~\ref{eq:tsne_gradient} can be rewritten as
 \begin{equation}
@@
 \mathbf{w}_M \right \} \in \mathbb{R}^d$. Additionally, we omit the
-symmetrization step in Equation~(\ref{eq:symmetrize}). Plugging these terms into
-Equation~(\ref{eq:kl_divergence}), we obtain the following gradient
+symmetrization step in Equation~\ref{eq:symmetrize}. Plugging these terms into
+Equation~\ref{eq:kl_divergence}, we obtain the following gradient
 \begin{equation}
@@
 the clustering structure at different levels of resolution. Modifying
-Equation~(\ref{eq:cauchy_kernel}) to $q_{ij} \propto \left ( 1 + || \mathbf{y}_i -
+Equation~\ref{eq:cauchy_kernel} to $q_{ij} \propto \left ( 1 + || \mathbf{y}_i -
 \mathbf{y}_j ||^2 / \alpha \right )^{-\alpha}$ allows us to use heavier-tailed
@@
 
-We have used the following snippet to create Figs.~\ref{fig:cao}.a-c and to demonstrate how we can take advantage of the more advanced features of \opentsne to more quickly create meaningful visualizations of massive data sets.
-Here, we first take a sample of the data and use t-SNE to construct a globally consistent embedding, which we will then use for initialization of the positions for the remaining data points:
-\begin{CodeChunk}
-\begin{CodeInput}
-# To improve the global structure of the final layout, we initialize the
-# embedding by creating a t-SNE plot on a sample of points.
-
-# Take a random sample of the data points.
+We have used the following code to create Figs.~\ref{fig:cao}.a-c and to demonstrate how we can take advantage of the more advanced features of \opentsne to more quickly create meaningful visualizations of massive data sets.
+
+First, we take small subset of data points and create a t-SNE embedding using a high perplexity value to emphasize the high level structure:
+\begin{CodeChunk}
+\begin{CodeInput}
 indices = np.random.permutation(list(range(x.shape[0])))
-reverse = np.argsort(indices)
-
 x_sample, x_rest = x[indices[:25000]], x[indices[25000:]]
 
-# Create a t-SNE embedding on the sample, using a high perplexity value to
-# emphasize the high level structure.
 init_sample = openTSNE.TSNE(perplexity=500).fit(x_sample)
-
-# We can next use the `transform` functionality to determine starting
-# positions for the remaining data points. Since we don't actually want to
-# optimize the remaining points separately from the sample points, we don't
-# need to call the `transform` function. Instead, we'll call the
-# `prepare_partial` function, which `transform` calls anyways under the hood.
+\end{CodeInput}
+\end{CodeChunk}
+Next, we will use this embedding to determine the starting positions for the remaining data points:
+\begin{CodeChunk}
+\begin{CodeInput}
 init_rest = init_sample.prepare_partial(x_rest)
-
-# Lastly, we merge `init_sample` and `init_rest` and restore the original
-# ordering.
-init_full = np.vstack((init_sample, init_rest))[reverse]
-\end{CodeInput}
-\end{CodeChunk}
-We then generate the three different embeddings using the following code snippet (shortened for clarity):
-\begin{CodeChunk}
-\begin{CodeInput}
-# We separately precompute the affinities as this can take quite a long time.
-# This way, we can reuse the precomputed affinities in various different
-# embeddings, without having to repeat the same, costly computation
-# every time.
+\end{CodeInput}
+\end{CodeChunk}
+Lastly, we merge \code{init_sample} and \code{init_rest} and restore the original ordering to obtain the full initialization:
+\begin{CodeChunk}
+\begin{CodeInput}
+init_full = np.vstack((init_sample, init_rest))[np.argsort(indices)]
+\end{CodeInput}
+\end{CodeChunk}
+
+Next, we precompute the affinity model.
+This can take quite a long time and, this way, we can reuse the same affinity model in various different embeddings, without having to repeat the same, costly computation every time:
+\begin{CodeChunk}
+\begin{CodeInput}
 affinities = openTSNE.affinity.PerplexityBasedNN(x, perplexity=30)
-
-# Create the TSNEEmbedding object using our previously calculated
-# initialization and precomputed affinities.
+\end{CodeInput}
+\end{CodeChunk}
+
+Using our precomputed initialization and affinity model, we are now ready to create our \code{TSNEEmbedding} object and optimize the embedding to our liking: 
+\begin{CodeChunk}
+\begin{CodeInput}
 embedding = openTSNE.TSNEEmbedding(init_full, affinities)
 
-# Run the standard t-SNE early exaggeration phase
 embedding_ee = embedding.optimize(n_iter=500, exaggeration=12, momentum=0.5)
@@
 By embedding the samples from a similar experiment on bipolar retinal cells by \citet{shekhar2016comprehensive}, we can correctly map the bipolar cell clusters onto the reference embedding.
-The following snippet shows the code necessary to generate Fig.~\ref{fig:macosko}:
-\begin{CodeChunk}
-\begin{CodeInput}
-# We specify the old default parameter values which are often used to show
-# why t-SNE misbehaves or fails to preserve global structure.
+
+The following snippet shows the code necessary to generate Fig.~\ref{fig:macosko}.
+
+We first construct Fig.~\ref{fig:macosko}.a with the most commonly used parameter values which aims to show that these can lead to poor cluster separation and an overall lack of global consistency:
+\begin{CodeChunk}
+\begin{CodeInput}
 tsne_old = openTSNE.TSNE(
@@
 ).fit(x)
-
-# By default, openTSNE uses PCA-based initialization and automatically
-# determines the best learning rate. This also allows us to lower the
-# default number of iterations to 500, lowering the overall runtime
-tsne_standard = openTSNE.TSNE(perplexity=[50, 500]).fit(x)
-
-# Use the generated embeddings to add new data points
+\end{CodeInput}
+\end{CodeChunk}
+We generate the embedding from Fig.~\ref{fig:macosko}.b using parameter values emphasizing the global coherence of the resulting visualization.
+By default, \opentsne uses PCA-based initialization and automatically determines the optimal learning rate.
+This allows \opentsne to lower the default number of iterations to 500.
+To emphasize the global consistency of the resulting embedding, we can also specify multiple perplexity values, resulting in a multiscale affinity model:
+\begin{CodeChunk}
+\begin{CodeInput}
+tsne_multiscale = openTSNE.TSNE(perplexity=[50, 500]).fit(x)
+\end{CodeInput}
+\end{CodeChunk}
+
+Finally, we can embed new samples into our obtained embedding space:
+\begin{CodeChunk}
+\begin{CodeInput}
 new_embedding = tsne_standard.transform(x_new)
@@
 
-\subsection{Speed\label{s:speed}}
+\subsection{Speed}\label{s:speed}
 

### 2023-05-22 2023-05-revision-1/main.tex
Mention Laplacian eigenmaps in initialization section

@@
 often arbitrarily scattered in the embedding space. However, when using
-initialization schemes based on PCA or spectral embeddings, the clusters
-identified in the resulting embedding are typically grouped in a globally
+initialization schemes based on PCA or Laplacian eigenmaps~\citep{belkin2001laplacian}, the clusters identified in the resulting embedding are typically grouped in a globally
 coherent manner.
@@
 \item The optimization in t-SNE starts with initial positions for each data point in the embedding space.
-We can set the initial positions randomly or impose an initial global structure on the embedding with, for example, a PCA-based initialization.
-Accordingly, the \code{openTSNE.initialization} module provides several possible initialization schemes, The user can also specify their initialization by providing a \code{numpy.ndarray} object.
+We can set the initial positions randomly or impose an initial global structure on the embedding.
+% with, for example, a PCA-based initialization.
+The \code{openTSNE.initialization} module provides several possible initialization schemes including random, PCA-based, and Laplacian Eigenmap-based intialization schemes.
+The user can also specify their own initialization by providing a \code{numpy.ndarray} object.
 By default, \opentsne opts for PCA-based initialization.

### 2023-05-23 figures/fig1-macosko.py
figs: set matplotlib font for pdf

@@ -9,6 +9,10 @@ import scanpy as sc
 
 import utils
 
+import matplotlib
+matplotlib.rcParams["pdf.fonttype"] = 42
+matplotlib.rcParams["ps.fonttype"] = 42
+
 parser = argparse.ArgumentParser()
 parser.add_argument("-i", "--in-dir", required=True)
 parser.add_argument("-o", "--out-dir", required=True)

### 2023-05-23 figures/fig3-iris.py

@@ -8,6 +8,10 @@ from sklearn import datasets
 
 import utils
 
+import matplotlib
+matplotlib.rcParams["pdf.fonttype"] = 42
+matplotlib.rcParams["ps.fonttype"] = 42
+
 parser = argparse.ArgumentParser()
 parser.add_argument("-o", "--out-dir", required=True)
 args = parser.parse_args()

### 2023-05-23 figures/fig6-cao.py

@@ -7,6 +7,10 @@ import numpy as np
 import openTSNE
 import scanpy as sc
 
+import matplotlib
+matplotlib.rcParams["pdf.fonttype"] = 42
+matplotlib.rcParams["ps.fonttype"] = 42
+
 parser = argparse.ArgumentParser()
 parser.add_argument("-i", "--in-dir", required=True)
 parser.add_argument("-o", "--out-dir", required=True)

### 2023-05-23 figures/fig7-hochgerner.py

@@ -7,6 +7,10 @@ import utils
 import openTSNE
 import matplotlib.pyplot as plt
 
+import matplotlib
+matplotlib.rcParams["pdf.fonttype"] = 42
+matplotlib.rcParams["ps.fonttype"] = 42
+
 
 parser = argparse.ArgumentParser()
 parser.add_argument("-i", "--in-dir", required=True)

### 2023-05-23 figures/fig6-cao.py
figs: remove old momentum parameters

@@ -54,8 +54,8 @@ if (
     sample_embedding = openTSNE.TSNEEmbedding(
         sample_init, sample_affinities, n_jobs=24, verbose=True,
     )
-    sample_embedding.optimize(n_iter=250, exaggeration=12, momentum=0.5, inplace=True)
-    sample_embedding.optimize(n_iter=500, exaggeration=1, momentum=0.8, inplace=True)
+    sample_embedding.optimize(n_iter=250, exaggeration=12, inplace=True)
+    sample_embedding.optimize(n_iter=500, exaggeration=1, inplace=True)
 
     # Calculate full affinities
     print("Calculating full affinities...")
@@ -80,10 +80,10 @@ if (
         init_full, affinities, n_jobs=8, random_state=42, verbose=True
     )
     
-    embedding_ee = embedding.optimize(n_iter=500, exaggeration=12, momentum=0.5)
-    embedding_exag4 = embedding_ee.optimize(n_iter=500, exaggeration=4, momentum=0.8)
-    embedding_exag2 = embedding_exag4.optimize(n_iter=500, exaggeration=2, momentum=0.8)
-    embedding_exag1 = embedding_exag2.optimize(n_iter=500, exaggeration=1, momentum=0.8)
+    embedding_ee = embedding.optimize(n_iter=500, exaggeration=12)
+    embedding_exag4 = embedding_ee.optimize(n_iter=500, exaggeration=4)
+    embedding_exag2 = embedding_exag4.optimize(n_iter=500, exaggeration=2)
+    embedding_exag1 = embedding_exag2.optimize(n_iter=500, exaggeration=1)
     
     # Add computed embeddings to h5ad files and save them to disk
     adata.obsm["X_tsne_ee"] = embedding_ee.view(np.ndarray)

### 2023-05-23 figures/fig1-macosko.py
figs: fixup text location in fig1

@@ -247,11 +247,11 @@ for ax_ in ax.ravel():
     
 ax[2].set_xlim(ax[1].get_xlim()), ax[2].set_ylim(ax[1].get_ylim())
 
-plt.text(38, -12.5, "Amacrine\ncells", color="k", fontsize=10, horizontalalignment="center", transform=ax[0].transData)
+plt.text(38, -14, "Amacrine\ncells", color="k", fontsize=10, horizontalalignment="center", transform=ax[0].transData)
 plt.text(27, 30, "Bipolar\ncells", color="k", fontsize=10, horizontalalignment="center", transform=ax[0].transData)
 
-plt.text(32, 24, "Amacrine\ncells", color="k", fontsize=10, horizontalalignment="center", transform=ax[1].transData)
-plt.text(30, -21.5, "Bipolar\ncells", color="k", fontsize=10, horizontalalignment="center", transform=ax[1].transData)
+plt.text(45, 30, "Amacrine\ncells", color="k", fontsize=10, horizontalalignment="center", transform=ax[1].transData)
+plt.text(32, -31.5, "Bipolar\ncells", color="k", fontsize=10, horizontalalignment="center", transform=ax[1].transData)
 
 for ax_, letter in zip(ax.ravel(), string.ascii_lowercase):
     plt.text(0, 1.02, letter, transform=ax_.transAxes, fontsize=16, va="baseline", fontweight="bold")

### 2023-05-23 figures/fig1-macosko.py
figs: add old momentum to fig1, which shows old standard

@@ -41,6 +41,8 @@ if (
         metric="cosine",
         learning_rate=200,
         n_iter=750,
+        initial_momentum=0.5,
+        final_momentum=0.8,
         n_jobs=8,
         random_state=0,
         verbose=True,

### 2023-05-23 2023-05-revision-1/main.tex
paper: ensure that uniform kernel doesn't come out of nowhere

@@
 \begin{enumerate}
-\item The t-SNE algorithm requires a specification of an affinity model to describe the similarities between data points.
-Standard t-SNE predominantly deals with tabular data and computes similarities using Equation~\ref{eq:p_ij}.
-However, \citet{kobak2019art} showed that using mixtures of Gaussians instead of a single Gaussian often reveals better global consistency of the resulting visualization (albeit at some computational cost).
-Alternatively, one can speed up the overall computation by choosing a uniform kernel, which yields quantitatively similar results to standard t-SNE.
-\code{openTSNE.affinity} provides several affinity models to choose from but defaults to the standard perplexity-based model proposed in the original publication.
-
-In \opentsne, it is straightforward to define and use a custom affinity model.
-For instance, one can extend the \code{openTSNE.affinity.Affinities} class and provide an affinity model for relational data using a normalized adjacency matrix.
-Such extension enables rapid experimentation and opens the doors to the visualization of non-tabular data.
+\item The t-SNE algorithm requires the specification of an affinity model to describe the similarities between data points $p_{ij}$.
+Standard t-SNE uses perplexity-based Gaussian kernels with varying bandwidths as specified in Equation~\ref{eq:p_ij}.
+However, other affinity models can also be used to highlight different aspects of the data or visualize other data modalities.
+For instance, \citet{kobak2019art} showed that using mixtures of Gaussians often better reveals global cluster organization in the resulting visualization (albeit at some computational cost).
+Alternatively, one can use a uniform affinity model that assigns equal similarities $p_{ij}$ to a pre-specified number of nearest neighbors.
+This yields quantitatively similar results to standard t-SNE at a lower computational cost, making it particularly suitable for larger data sets.
+Visualizing different data modalities, such as graphs, can also be achieved by providing a precomputed distance matrix that might consist of, for instance, the shortest paths between nodes.
+The \code{openTSNE.affinity} module provides a selection of commonly used affinity models but defaults to the standard perplexity-based Gaussian model, as proposed in the original publication.
+Users can also define custom affinity models by extending the \code{openTSNE.affinity.Affinities} class.
+The resulting class can then be used as a drop-in replacement for the built-in affinity models.
 

### 2023-05-23 2023-05-revision-1/main.tex
paper: update momentum parameters to reflect recent changes in the library

@@
 \begin{CodeInput}
->>> embedding.optimize(250, exaggeration=12, momentum=0.5, inplace=True)
->>> embedding.optimize(500, momentum=0.8, inplace=True)
+>>> embedding.optimize(250, exaggeration=12, inplace=True)
+>>> embedding.optimize(500, inplace=True)
 >>> embedding[:3]
@@
 >>> new_embedding.optimize(
-...     exaggeration=1.5, n_iter=250, learning_rate=0.1, momentum=0.8,
+...     exaggeration=1.5, n_iter=250, learning_rate=0.1,
 ...     max_grad_norm=0.25, inplace=True
@@
 
-embedding_ee = embedding.optimize(n_iter=500, exaggeration=12, momentum=0.5)
-embedding_c = embedding_ee.optimize(n_iter=500, exaggeration=4, momentum=0.8)
-embedding_b = embedding_c.optimize(n_iter=500, exaggeration=2, momentum=0.8)
-embedding_a = embedding_b.optimize(n_iter=500, exaggeration=1, momentum=0.8)
+embedding_ee = embedding.optimize(n_iter=500, exaggeration=12)
+embedding_c = embedding_ee.optimize(n_iter=500, exaggeration=4)
+embedding_b = embedding_c.optimize(n_iter=500, exaggeration=2)
+embedding_a = embedding_b.optimize(n_iter=500, exaggeration=1)
 \end{CodeInput}

### 2023-05-23 2023-05-revision-1/main.tex
paper: update text and table 1 with FIt-SNE pypi/conda checkmarks

@@
 \pkg{FIt-SNE}~\citep{linderman2019fast} alleviates these problems and provides bindings to both the \proglang{Python} and \proglang{R} programming languages.
-However, \pkg{FIt-SNE} is difficult to install and distribute and provides an API largely inconsistent with the \proglang{Python} data science ecosystem.
-None of the existing t-SNE implementations support adding new samples into existing embeddings.
+However, none of the existing t-SNE implementations support adding new samples into existing embeddings.
 To alleviate all these problems, and to support newly proposed additions to t-SNE algorithms, we here present \opentsne, an open-source \proglang{Python} implementation of the t-SNE algorithm.
@@
 \textbf{Ease of installation} \\
-Native \proglang{Python} installation & \checkmark & & & \checkmark & & \\
-\pkg{PyPI} package & \checkmark & \checkmark & & \checkmark & & \\
-\pkg{conda} package & \checkmark & & & \checkmark & & \\
+\pkg{PyPI} package & \checkmark & \checkmark & \checkmark & \checkmark & & \\
+\pkg{conda} package & \checkmark & & \checkmark & \checkmark & & \\
 
@@
 \textbf{Advanced features and extensions} \\
-%Multiscale Gaussian kernels & & & & \checkmark & \checkmark & & \\
 Extensible affinity models & & & & \checkmark & & \\

### 2023-05-23 2023-05-revision-1/main.tex
paper: add scRNA-seq pipeline to appendix

@@
 
+\newpage
+
+\begin{appendix}
+
+\section{Single-cell RNA-seq Data Preprocessing}
+
+We here describe the preprocessing steps applied to the single-cell RNA-seq data sets used in Fig.~\ref{fig:macosko}~\citep{macosko2015highly}, Fig.~\ref{fig:tasic}~\citep{tasic2018shared}, Fig.~\ref{fig:cao}~\citep{cao2019single}, and Fig.~\ref{fig:transform}~\citep{hochgerner2018conserved}.
+The standard single-cell preprocessing pipeline consists of gene filtering, gene selection, data normalization, and dimensionality reduction.
+
+First, we discard genes detected fewer than ten times, then select the 3,000 most highly-variable genes following the gene-selection procedure from \citet{kobak2019art}.
+We then perform counts-per-median library-size normalization followed by log normalization and standardization without zero-centering.
+Finally, we extract the top 50 principal components, which are then used to construct subsequent t-SNE embeddings.
+This pipeline is applied to all the aforementioned data sets with the following exceptions, which serve to better illustrate our points:
+\begin{itemize}
+  \item In \citet{macosko2015highly}, we perform library-size normalization by computing counts-per-million and apply centering during standardization.
+  \item In \citet{hochgerner2018conserved}, we select the 1,000 most variable genes.
+\end{itemize}
+
+When embedding additional data points into the existing embedding in Figs.~\ref{fig:macosko}.c and \ref{fig:transform}, we follow the procedure from \citet{policar2021embedding}.
+Therefore, no preprocessing is applied to the data sets from \citet{shekhar2016comprehensive} and \citet{harris2018classes}, and the similarities between the reference data set and the new data set are computed on the raw counts.
+
+\end{appendix}
+
 \end{document}

### 2023-05-23 figures/manuscript_iris.py
figs: remove momentum from iris script

@@ -39,8 +39,8 @@ initialization[:3]
 embedding = openTSNE.TSNEEmbedding(initialization, affinities)
 embedding[:3]
 
-embedding.optimize(250, exaggeration=12, momentum=0.5, inplace=True)
-embedding.optimize(500, momentum=0.8, inplace=True)
+embedding.optimize(250, exaggeration=12, inplace=True)
+embedding.optimize(500, inplace=True)
 embedding[:3]
 
 import numpy as np
@@ -48,7 +48,7 @@ np.linalg.norm(openTSNE.TSNE().fit(iris.data) - embedding)
 
 new_embedding = embedding.prepare_partial(iris.data[::3], perplexity=5)
 new_embedding.optimize(
-    exaggeration=1.5, n_iter=250, learning_rate=0.1, momentum=0.8,
+    exaggeration=1.5, n_iter=250, learning_rate=0.1,
     max_grad_norm=0.25, inplace=True
 )
 new_embedding[::3]

### 2023-05-23 2023-05-revision-1/main.tex
paper: fixup figure font sizes
paper: update openTSNE version to 0.7.1

@@
 \end{CodeChunk}
-\opentsne provides precompiled binaries for all major platforms (Windows, macOS, Linux) for all currently supported \proglang{Python} versions (3.7---3.10).
+\opentsne provides precompiled binaries for all major platforms (Windows, macOS, Linux) for all currently supported \proglang{Python} versions (3.7---3.11).
 Precompiled binaries are especially convenient for Windows users, as they obviate the need for a \proglang{C}/\proglang{C++} compiler, which is typically not bundled with Windows operating system.
@@
 >>> openTSNE.__version__
-0.6.2
+0.7.1
 \end{CodeInput}
@@
   \center
-  \includegraphics[width=0.75\textwidth]{tasic2018}
+  \includegraphics[width=\textwidth]{tasic2018}
   \caption{\label{fig:tasic}
@@
 % We benchmark Python
-Fig.~\ref{fig:benchmarks_py} benchmarks four popular \proglang{Python}~\citep{vanrossum1995python} t-SNE implementations, including \pkg{scikit-learn} (v1.1.2), \pkg{MulticoreTSNE} (v0.1), \pkg{FIt-SNE} (v1.1.0), and \opentsne (v0.6.2).
+Fig.~\ref{fig:benchmarks_py} benchmarks four popular \proglang{Python}~\citep{vanrossum1995python} t-SNE implementations, including \pkg{scikit-learn} (v1.1.2), \pkg{MulticoreTSNE} (v0.1), \pkg{FIt-SNE} (v1.1.0), and \opentsne (v0.7.1).
 Benchmarks were performed on an Intel(R) Xeon(R) CPU E5-1650 equipped with 128 GB of memory.
@@
   \caption{\label{fig:benchmarks_py}
-  We benchmark \opentsne (v0.6.2) against three popular open-source
+  We benchmark \opentsne (v0.7.1) against three popular open-source
   implementations from \pkg{scikit-learn}~\citep{pedregosa2011scikit}
@@
 In \proglang{Julia}~\citep{bezanson2017julia}, the most popular implementation -- \pkg{TSne.jl}~\citep{julia_tsne} -- is based on a naive $\mathcal{O}(N^2)$ implementation of the algorithm, which allows it to be used only on smaller data sets containing up to thousands of data points.
-We benchmark \pkg{Rtsne} (v0.15) and  \pkg{TSne.jl} (v1.3.0) and compare it to \opentsne (v0.6.2).
+We benchmark \pkg{Rtsne} (v0.15) and  \pkg{TSne.jl} (v1.3.0) and compare it to \opentsne (v0.7.1).
 Fig.~\ref{fig:benchmarks_lang} illustrates the scaling of each approximation scheme.
@@
   We select the fastest available, open-source implementations of the t-SNE algorithm from the \proglang{Python}, \proglang{R}, and \proglang{Julia} programming languages.
-We choose \opentsne (v0.6.2) for \proglang{Python}, \pkg{Rtsne}~\citep{krijthe2015rtsne} (v0.15) for \proglang{R}, and \pkg{TSne.jl}~\citep{julia_tsne} (v1.3.0) for \proglang{Julia}.
+We choose \opentsne (v0.7.1) for \proglang{Python}, \pkg{Rtsne}~\citep{krijthe2015rtsne} (v0.15) for \proglang{R}, and \pkg{TSne.jl}~\citep{julia_tsne} (v1.3.0) for \proglang{Julia}.
 The benchmarks reflect the time complexity of the implemented approximation schemes.
@@
 \caption{\label{tab:features}
-  We compare the features of \opentsne~(v0.6.2) to \pkg{scikit-learn}~(v1.0.1),
+  We compare the features of \opentsne~(v0.7.1) to \pkg{scikit-learn}~(v1.0.1),
   \pkg{BHTSNE}~(master), \pkg{MulticoreTSNE}~(v0.1), and
@@
 The results in this paper were obtained using \proglang{Python}~3.10.2 with the
-\pkg{openTSNE}~v0.6.2 package. All results and figures can be reproduced using
+\pkg{openTSNE}~v0.7.1 package. All results and figures can be reproduced using
 the scripts found in the accompanying GitHub repository

### 2023-05-23 2023-05-revision-1/main.tex
paper: elaborate on unstable transform optimization

@@
 
-Similarly to standard t-SNE, a direct calculation of gradients takes
-$\mathcal{O}(N \cdot M)$ time, but it is straightforward to adapt the Barnes-Hut
-and FIt-SNE approximation schemes, reducing the time complexity to
-$\mathcal{O}(M \log N)$ and $\mathcal{O}(M)$, respectively. Special care must be
-taken to tune the learning rate during optimization as the default values
-$\eta=200$ and $\eta=N/12$ result in highly unstable optimization.
+Similarly to standard t-SNE, a direct calculation of gradients takes $\mathcal{O}(N \cdot M)$ time, but it is straightforward to adapt the Barnes-Hut and FIt-SNE approximation schemes, reducing the time complexity to $\mathcal{O}(M \log N)$ and $\mathcal{O}(M)$, respectively.
+Special care must be taken to adjust the learning rate during optimization as the parameter values used during standard optimization lead to highly unstable behaviour.
+Empirically, we have found that reducing the learning rate to $\eta \sim 0.1$ results in more stable optimization, as discussed in \citet{policar2021embedding}.
+Moreover, during standard optimization, data points are positioned in the embedding such that an equilibrium between the attractive and repulsive forces is achieved.
+However, when embedding new data points, the reference embedding remains fixed.
+Consequently, if a new data point needed to be positioned between existing reference data points, it may not be possible to reach such an equilibrium, as the reference data points could not adjust to accomodate the new data point.
+To compensate for the increased repulsive forces, we have found it beneficial to run this optimization using the exaggeration factor $\rho=1.5$.
+The training dynamics of embedding new data points into existing embeddings are yet fully understood and are left to future exploration.
 

### 2023-05-23 2023-05-revision-1/main.tex
paper: describe BH/FFT heuristic

@@
 \item In optimizing the embedding, the t-SNE algorithm finds the positions of the points in the embedding space that reflect their corresponding similarities in the affinity model.
-\opentsne stores the initial positions and affinities in a \code{openTSNE.TSNEEmbedding} object that includes the \code{.optimize} method for embedding optimization that uses either the Barnes-Hut or FIt-SNE approximation scheme.
-While the FIt-SNE approximation scheme scales linearly with the number of samples, it introduces additional computational overhead, often unnecessary for smaller data sets, resulting in longer runtimes than the Barnes-hut approximation scheme.
-By default, \opentsne uses a heuristic to choose the fastest algorithm for a given data set.
+\opentsne stores the initial positions and affinities in an \code{openTSNE.TSNEEmbedding} object that includes the \code{.optimize} method for embedding optimization that uses either the Barnes-Hut or FIt-SNE approximation scheme.
+While the FIt-SNE approximation scheme scales linearly with the number of samples, it introduces additional computational overhead, often unnecessary for smaller data sets, resulting in longer overall runtimes than the Barnes-Hut approximation scheme.
+By default, \opentsne uses an empirically-determined heuristic to choose the fastest algorithm for a given data set based on its number of samples.
+If the data set contains fewer than 10,000 samples, the Barnes-Hut approximation scheme is selected, while FIt-SNE is used for larger data sets.
+
 \end{enumerate}

### 2023-05-24 2023-05-revision-1/main.tex
paper: reword confusing cluster colors caption in Fig. 7

@@
 The data contains gene expression in different neurons, supporting glia, and other vascular cells (upper left).
-Data points representing cells are colored according to cell types assigned in the original publication; see the legend from Fig.~\ref{fig:tasic} to map colors to cell type.
+Each data point corresponds to a single cell colored according to its inferred cell type as determined in the original publication.
+We use the same color scheme as in Fig.~\ref{fig:tasic} where warm colors correspond to excitatory neurons, cool colors correspond to inhibitory neurons, and gray/brown colors correspond to non-neuronal cells.
 We then embed new hippocampal cells collected in a study by \citet{harris2018classes} using the embedding of \citet{hochgerner2018conserved} data as a reference.
@@
 Inspecting the embeddings of these cells in the reference embedding (bottom left) reveals that in addition to inhibitory neurons, the data contains several supporting glial cells and a small population of endothelial cells.
-We can verify our approach's accuracy by inspecting marker genes for the major cell types in the reference (top row) and embedded samples (bottom row).
-}
+We can verify our approach's accuracy by inspecting marker genes for the major cell types in the reference (top row) and embedded samples (bottom row).}
 \end{figure*}

### 2023-05-24 2023-05-revision-1/main.tex
paper: clarify what prepare_partial does

@@
 \end{CodeChunk}
-and validate that this also produces the exact same embedding as before:
+We here use the \code{embedding.prepare_partial} method to determine initial positions for the new data points, then call the \code{new_embedding.optimize} method to run the optimization process.
+Again, we can validate that this also produces the exact same embedding as before:
 \begin{CodeChunk}

### 2023-05-24 2023-05-revision-1/main.tex
paper: update all openTSNE references to v1.0.0

@@
 >>> openTSNE.__version__
-0.7.1
+1.0.0
 \end{CodeInput}
@@
 % We benchmark Python
-Fig.~\ref{fig:benchmarks_py} benchmarks four popular \proglang{Python}~\citep{vanrossum1995python} t-SNE implementations, including \pkg{scikit-learn} (v1.1.2), \pkg{MulticoreTSNE} (v0.1), \pkg{FIt-SNE} (v1.1.0), and \opentsne (v0.7.1).
+Fig.~\ref{fig:benchmarks_py} benchmarks four popular \proglang{Python}~\citep{vanrossum1995python} t-SNE implementations, including \pkg{scikit-learn} (v1.1.2), \pkg{MulticoreTSNE} (v0.1), \pkg{FIt-SNE} (v1.1.0), and \opentsne (v1.0.0).
 Benchmarks were performed on an Intel(R) Xeon(R) CPU E5-1650 equipped with 128 GB of memory.
@@
   \caption{\label{fig:benchmarks_py}
-  We benchmark \opentsne (v0.7.1) against three popular open-source
+  We benchmark \opentsne (v1.0.0) against three popular open-source
   implementations from \pkg{scikit-learn}~\citep{pedregosa2011scikit}
@@
 In \proglang{Julia}~\citep{bezanson2017julia}, the most popular implementation -- \pkg{TSne.jl}~\citep{julia_tsne} -- is based on a naive $\mathcal{O}(N^2)$ implementation of the algorithm, which allows it to be used only on smaller data sets containing up to thousands of data points.
-We benchmark \pkg{Rtsne} (v0.15) and  \pkg{TSne.jl} (v1.3.0) and compare it to \opentsne (v0.7.1).
+We benchmark \pkg{Rtsne} (v0.15) and  \pkg{TSne.jl} (v1.3.0) and compare it to \opentsne (v1.0.0).
 Fig.~\ref{fig:benchmarks_lang} illustrates the scaling of each approximation scheme.
@@
   We select the fastest available, open-source implementations of the t-SNE algorithm from the \proglang{Python}, \proglang{R}, and \proglang{Julia} programming languages.
-We choose \opentsne (v0.7.1) for \proglang{Python}, \pkg{Rtsne}~\citep{krijthe2015rtsne} (v0.15) for \proglang{R}, and \pkg{TSne.jl}~\citep{julia_tsne} (v1.3.0) for \proglang{Julia}.
+We choose \opentsne (v1.0.0) for \proglang{Python}, \pkg{Rtsne}~\citep{krijthe2015rtsne} (v0.15) for \proglang{R}, and \pkg{TSne.jl}~\citep{julia_tsne} (v1.3.0) for \proglang{Julia}.
 The benchmarks reflect the time complexity of the implemented approximation schemes.
@@
 \caption{\label{tab:features}
-  We compare the features of \opentsne~(v0.7.1) to \pkg{scikit-learn}~(v1.0.1),
+  We compare the features of \opentsne~(v1.0.0) to \pkg{scikit-learn}~(v1.0.1),
   \pkg{BHTSNE}~(master), \pkg{MulticoreTSNE}~(v0.1), and
@@
 The results in this paper were obtained using \proglang{Python}~3.10.2 with the
-\pkg{openTSNE}~v0.7.1 package. All results and figures can be reproduced using
+\pkg{openTSNE}~v1.0.0 package. All results and figures can be reproduced using
 the scripts found in the accompanying GitHub repository

### 2023-05-24 2023-05-revision-1/main.tex
paper: ensure we define attractive forces before referring to them

@@
 and represents the unnormalized $q_{ij}$.
+Alternatively, this gradient can be rewritten as
+\begin{equation}
+\frac{\partial C}{\partial \mathbf{y}_i} = 4 \Bigg [
+\underbrace{\sum_{j \neq i} p_{ij} q_{ij} Z \left ( \mathbf{y}_i - \mathbf{y}_j \right )}_{\text{attractive forces}}  -
+\underbrace{\sum_{j \neq i} q_{ij}^2 Z \left ( \mathbf{y}_i - \mathbf{y}_j \right )}_{\text{repulsive forces}}
+\Bigg ], \label{eq:grad_attr_rep}
+\end{equation}
+where $Z = \sum_{k \neq l}\left ( 1 + || \mathbf{y}_k - \mathbf{y}_l ||^2 \right )^{-1}$.
+This formulation implies that the t-SNE algorithm can be interpreted as a force-directed layout algorithm, where individual data points act as particles exerting both attractive and repulsive on each other until a state of equilibrium is reached.
 
@@
 
-Evaluation of t-SNE gradients requires $\mathcal{O}(N^2)$ operations,
-which makes it impractical with large data sets and
-calls for efficient approximation schemes.
-\citet{van2014accelerating} observed that the t-SNE gradient can be cast as an
-N-body problem where data points represent particles that attract and repel each
-other. The gradient from Equation~\ref{eq:tsne_gradient} can be rewritten as
-\begin{equation}
-\frac{\partial C}{\partial \mathbf{y}_i} = 4 \Bigg [
-\underbrace{\sum_{j \neq i} p_{ij} q_{ij} Z \left ( \mathbf{y}_i - \mathbf{y}_j \right )}_{\text{attractive forces}}  -
-\underbrace{\sum_{j \neq i} q_{ij}^2 Z \left ( \mathbf{y}_i - \mathbf{y}_j \right )}_{\text{repulsive forces}}
-\Bigg ], \label{eq:grad_attr_rep} \notag
-\end{equation}
-where $Z = \sum_{k \neq l}\left ( 1 + || \mathbf{y}_k - \mathbf{y}_l ||^2 \right
-)^{-1}$. We can view this equation as a particle simulation, where the
-two terms represent the attractive and repulsive forces between individual
-particles. Each term lends itself to efficient approximations, enabling us to
-reduce the time complexity of t-SNE to $\mathcal{O}(N)$. This allows modern
-t-SNE implementations to be leveraged for the visualization of data sets
-containing up to millions of data points.
+A direct evaluation of the t-SNE gradients requires $\mathcal{O}(N^2)$ operations, which makes it impractical with large data sets and calls for efficient approximation schemes.
+The formulation in Equation~\ref{eq:grad_attr_rep} casts the t-SNE gradient as an N-body simulation where data points act as particles exerting attractive and repulsive forces onto one another.
+Both terms lend themselves to efficient approximations, enabling us to reduce the time complexity of the t-SNE algorithm to $\mathcal{O}(N)$.
+This allows modern t-SNE implementations to be leveraged for the visualization of data sets containing up to millions of data points.
 

### 2023-05-25 benchmarks/benchmark.py
benchmarks: add uniform tsne to benchmarks

@@ -46,13 +46,19 @@ class TSNEBenchmark:
     perplexity = 30
     learning_rate = 200
 
-    def run(self, fname: str, n_samples=1000, repetitions=1, n_jobs=1):
+    def run(self, fname: str, n_samples=1000, repetitions=1, exaggeration=1, n_jobs=1, old_tsne=True):
         x, y = self.load_data(fname=fname, n_samples=n_samples)
 
         for idx in range(repetitions):
-            self._run(x, random_state=idx, n_jobs=n_jobs)
+            self._run(
+                x,
+                random_state=idx,
+                exaggeration=exaggeration,
+                n_jobs=n_jobs,
+                old_tsne=old_tsne,
+            )
 
-    def _run(self, data, random_state=None, n_jobs=1):
+    def _run(self, data, random_state=None, exaggeration=1, n_jobs=1, old_tsne=True):
         raise NotImplementedError()
 
     def load_data(self, fname: str, n_samples: int = None):
@@ -73,7 +79,7 @@ class TSNEBenchmark:
 
 
 class openTSNENNDescent(TSNEBenchmark):
-    def _run(self, x, random_state=None, n_jobs=1):
+    def _run(self, x, random_state=None, exaggeration=1, n_jobs=1, old_tsne=True):
         print("-" * 80)
         print(f"openTSNE v{openTSNE.__version__}")
         print("Random state", random_state)
@@ -107,14 +113,23 @@ class openTSNENNDescent(TSNEBenchmark):
             random_state=random_state,
             verbose=True,
         )
-        embedding.optimize(250, exaggeration=12, momentum=0.8, inplace=True)
-        embedding.optimize(750, momentum=0.5, inplace=True)
+
+        if old_tsne:
+            initial_momentum = 0.8
+            final_momentum = 0.5
+            ee_iter, rest_iter = 250, 750
+        else:
+            initial_momentum = final_momentum = 0.8
+            ee_iter, rest_iter = 250, 500
+
+        embedding.optimize(ee_iter, exaggeration=12, momentum=initial_momentum, inplace=True)
+        embedding.optimize(rest_iter, exaggeration=exaggeration, momentum=final_momentum, inplace=True)
         print("openTSNE: Optimization", time.time() - start_optim)
         print("openTSNE: Full", time.time() - start, flush=True)
 
 
 class openTSNEBH(TSNEBenchmark):
-    def _run(self, x, random_state=None, n_jobs=1):
[…]

### 2023-05-25 benchmarks/run.sh

@@ -98,7 +98,7 @@ if $small; then
 
   dataset_name="macosko_2015"
   sample_sizes=(1000 2000)
-  repetitions=6;
+  repetitions=3;
 
 else
   echo "Running full benchmark suite."
@@ -155,6 +155,73 @@ run_python_benchmarks() {
   done;
 }
 
+run_python_umap_opentsne_benchmarks() {
+  # Regular UMAP
+  for size in "${sample_sizes[@]}"; do
+    cmd="OMP_NUM_THREADS=$cores NUMBA_NUM_THREADS=$cores \
+        LD_LIBRARY_PATH="${CONDA_PREFIX}/lib/" \
+        DYLD_LIBRARY_PATH="${CONDA_PREFIX}/lib/" \
+        python benchmark.py UMAP run \
+        --fname data/$dataset_name.pkl.gz \
+        --repetitions $repetitions \
+        --n-samples $size \
+        --n-jobs $cores 2>&1 \
+        | tee -a logs/${dataset_name}--UMAP_standard--${cores}_core--${size}_samples.log";
+    echo "$cmd" | tr -s " ";
+    eval "$cmd";
+  done;
+  
+  # Regular t-SNE, but with newer default parameters
+  for size in "${sample_sizes[@]}"; do
+    cmd="OMP_NUM_THREADS=$cores NUMBA_NUM_THREADS=$cores \
+        LD_LIBRARY_PATH="${CONDA_PREFIX}/lib/" \
+        DYLD_LIBRARY_PATH="${CONDA_PREFIX}/lib/" \
+        python benchmark.py openTSNEFFT run \
+        --fname data/$dataset_name.pkl.gz \
+        --repetitions $repetitions \
+        --n-samples $size \
+        --n-jobs $cores 2>&1 \
+        --old_tsne False \
+        | tee -a logs/${dataset_name}--openTSNE_standard--${cores}_core--${size}_samples.log";
+    echo "$cmd" | tr -s " ";
+    eval "$cmd";
+  done;
+  
+  # t-SNE with uniform affinity kernel with k-neighbors 30
+  for size in "${sample_sizes[@]}"; do
+    cmd="OMP_NUM_THREADS=$cores NUMBA_NUM_THREADS=$cores \
+        LD_LIBRARY_PATH="${CONDA_PREFIX}/lib/" \
+        DYLD_LIBRARY_PATH="${CONDA_PREFIX}/lib/" \
[…]

### 2023-05-25 2023-05-revision-1/response-to-reviewers.tex
paper: minor revisions, add response to reviewers tex file

\documentclass{article}

\usepackage[T1]{fontenc}
\usepackage{natbib}
\usepackage{todonotes}
\usepackage{gensymb}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage[top=1in,bottom=1in,left=1in,right=1in]{geometry}
\usepackage{alltt}
\usepackage{framed}
\usepackage{xcolor}
\usepackage{comment}
\usepackage{graphicx}

\usepackage{xr}[]
\externaldocument{main}
\usepackage{url}

\renewcommand{\thetable}{\Alph{table}}
\renewcommand{\thefigure}{\Alph{figure}}

\newcommand{\beq}{\begin{equation}}
\newcommand{\eeq}{\end{equation}}
\newcommand{\beqs}{\begin{equation}}
\newcommand{\eeqs}{\end{equation}}
\newcommand{\mx}[1]{\ensuremath{\boldsymbol{\mathrm{{#1}}}}}   % Simple matrix format.

% \newcommand{\review}[1]{\noindent{{\textbf{\textit{#1}}}}\vspace{3mm}}
% \definecolor{colframe}{rgb}{0.01,0.199,0.1}

\newcounter{rtaskno}
\newcommand{\rtask}[1]{\refstepcounter{rtaskno}\label{#1}}

\newcommand{\reviewc}[2]{\begin{leftbar}\noindent\rtask{#1}\ref{#1}. #2\end{leftbar}}
\newcommand{\review}[1]{\begin{leftbar}\noindent #1\end{leftbar}}
\newcommand{\answer}[1]{\noindent #1\\[0mm]}
\newcommand{\copyquote}[1]{\vspace{2mm} {\color{blue} #1}\vspace{2mm}}

\newlength{\leftbarwidth}
\setlength{\leftbarwidth}{3pt}
\newlength{\leftbarsep}
\setlength{\leftbarsep}{10pt}
\renewenvironment{leftbar}{%
    \def\FrameCommand{{\color{black}{\vrule width \leftbarwidth\relax\hspace {\leftbarsep}}}}%
    \MakeFramed {\advance \hsize -\width \FrameRestore }%
}{\endMakeFramed}

% \usepackage[colorlinks=True,citecolor=black]{hyperref}

\begin{document}

\noindent{\bf Poli\v{c}ar, Stra\v{z}ar, and Zupan: 
\\
openTSNE: a modular Python library for t-SNE dimensionality reduction and embedding}

\vspace{5mm}
\noindent {\bf RESPONSE TO THE REVIEWERS}

\vspace{5mm}

\noindent Dear Editor and the Reviewers, \\

\vspace{3mm}

\noindent We thank you for the comprehensive feedback that helped us improve the manuscript and the permission to submit a revised version. Below, we provide detailed, point-by-point responses to the reviewers comments and concerns. To facilitate the review process, we also include a version of the manuscript with the highlighted changes from the initial submission.

\vspace{3mm}

\noindent Kind regards,
\vspace{3mm}

\noindent Pavlin Poli\v{c}ar, 
Martin Stra\v{z}ar, 
Bla\v{z} Zupan \\ 
\noindent Ljubljana, May 26, 2023

\vspace{10mm}

\subsection*{Reviewer B}

\reviewc{r1q1}{I just have a few minor comments regarding the text. Firstly, the authors claim that FIt-SNE is not a PyPI/conda package in table 1, and that is just not the case. There is a similar comment in the introduction at the bottom of page 2 that needs to be remedied in this regard. "However, FIt-SNE is difficult to install and distribute, and provides and API largely inconsistent with the Python data science ecosystem".  I also find it fairly the comment fairly vague in the context of being inconsistent with Python data-science ecosystem, as the functional approach with most of the parameters being set by default is also a commonly used API for software.}

\answer{
We apologize for this oversight and would like to thank the reviewer for pointing out the availability of PyPI/conda packages for FIt-SNE. We have updated the text and Table 1 to reflect this.
}

\reviewc{r1q2}{Finally, this is one of the only works that switches between the Barnes-Hut approach to the FIt-SNE algorithm and the authors mention that they have a heuristic for doing so. This is an extremely useful feature as well, as it chooses the best of both worlds, and it would be wonderful if the authors could provide a brief description for the heuristic they use in their code. }

\answer{We have now included a sentence in Section 3 on page 8 that describes our heuristic.}

\subsection*{Reviewer C}

\reviewc{r2q1}{The paper describes openTSNE version 0.6, but the current version is 0.7, and it seems there were recent changes affecting some of the text and the code snippets in the paper (learning rate / momentum). This needs to be updated throughout. More generally, if the paper goes on to be published in JOSS, would it make sense to release version 1.0 and describe it as such in the paper? If the library is ready for a peer-reviewed publication, then is it ready to be designated 1.0?}

\answer{
We agree with the reviewer and have released version 1.0 of the software. We have updated all the code snippets in the manuscript and the accompanying code files to reflect the recent changes in the openTSNE package. We have also re-generated the figures accordingly.
}

\reviewc{r2q2}{There are no details given about preprocessing transcriptomic data (normalization, log-transformation, gene selection, PCA, etc.) As the paper uses RNAseq datasets for all examples, these details should be provided somewhere.}

\answer{We have added the details of the single-cell RNA-seq preprocessing pipeline to the Appendix.
}

\reviewc{r2q3}{page 7: "one can speed up the overall computation by choosing a uniform kernel" -- this sounds cryptic as the uniform kernel was not explained anywhere. It could make sense to elaborate on it somewhere: what exactly is it? how to use it?}

\answer{
We have reworded the paragraph so that the uniform affinity kernel is now briefly described.
}

\reviewc{r2q4}{page 7: initialization paragraph only mentions PCA, would it make sense to mention that openTSNE also supports Laplacian eigenmaps as initialization?}

\answer{
We now mention this option in the corresponding paragraph on page 8 and have also added a reference to Laplacian eigenmaps in Section 2.4 in the paragraph on global initialization.
}

\reviewc{r2q5}{It would be interesting to add UMAP to the runtime benchmark. UMAP is a very popular alternative to t-SNE, and is often thought to be much faster, which is not actually the case for openTNSE. Perhaps a separate benchmark figure could focus on openTSNE vs UMAP (depending on the sample size)? This may depend on the used affinity model, as uniform affinity with 15 neighbors should be much faster than the standard affinity model (with 90 neighbors) for large sample sizes. If so, it may be of interest to show both and discuss.}

\answer{
TODO bla
}

\reviewc{r2q6}{top of page 5: you talk about attractive forces but they have not been introduced yet. It may make sense to move the equation from secton 2.2 into 2.1 after Eq (5).}

\answer{We would like to thank the reviewer for the suggestion, and we have now moved the equation from section 2.2 to the end of section 2.1, and updated the text accordingly.}

\reviewc{r2q7}{page 6: "Special care must be taken to tune the learning rate during optimization" -- the details are missing here. So what values are used in openTSNE?}

\answer{We have expanded the paragraph to elaborate on the instability and what parameter values we use in openTSNE.}

\reviewc{r2q8}{page 13: it is unclear what embedding.prepare\_partial() does here -- how are the new samples initialized, before you are running optimize()? Same for the prepare\_partial call on page 17.}

\answer{We agree that this is somewhat crpytic and have added a sentence below the code snippet describing what each of the function calls does.}

\reviewc{r2q9}{Figure 7 -- "Data points representing cells are colored according to cell types assigned in the original publication; see the legend from Fig. 5 to map colors to cell type." The last part ("see the legend from Fig. 5 to map colors to cell type") is unclear as Figure 5 shows different dataset. Does this study use the same colormap? This is confusing.}

\answer{We agree that this sentence is confusing and have reworded it to be more clear.}

\reviewc{r2q10}{References: Boehm et al is published.}

\answer{We have updated the citation.}

\end{document}

### 2023-05-25 2023-05-revision-1/main.tex
paper: minor revisions, add response to reviewers tex file

@@
 However, when embedding new data points, the reference embedding remains fixed.
-Consequently, if a new data point needed to be positioned between existing reference data points, it may not be possible to reach such an equilibrium, as the reference data points could not adjust to accomodate the new data point.
-To compensate for the increased repulsive forces, we have found it beneficial to run this optimization using the exaggeration factor $\rho=1.5$.
+Consequently, if a new data point needs to be positioned between existing reference data points, it may not be possible to reach such an equilibrium, as the reference data points cannot adjust to accomodate the new data point.
+To compensate for the increased repulsive forces, we have found it beneficial to increase the exaggeration factor to $\rho=1.5$ during optimization.
 The training dynamics of embedding new data points into existing embeddings are yet fully understood and are left to future exploration.
@@
 For instance, \citet{kobak2019art} showed that using mixtures of Gaussians often better reveals global cluster organization in the resulting visualization (albeit at some computational cost).
-Alternatively, one can use a uniform affinity model that assigns equal similarities $p_{ij}$ to a pre-specified number of nearest neighbors.
-This yields quantitatively similar results to standard t-SNE at a lower computational cost, making it particularly suitable for larger data sets.
-Visualizing different data modalities, such as graphs, can also be achieved by providing a precomputed distance matrix that might consist of, for instance, the shortest paths between nodes.
+Alternatively, a uniform affinity model can produce quantitively similar embeddings to standard t-SNE at a third of the computational cost, making it particularly useful for larger data sets.
+Unlike the standard affinity model, which uses Gaussian kernels to determine the similarities $p_{ij}$, the uniform affinity model uses a uniform kernel to assign equal similarities to a pre-specified number of nearest neighbors.
+Visualizing different data modalities, such as graphs, can also be achieved by providing precomputed distance matrices that might consist of, for instance, the shortest paths between nodes.
 The \code{openTSNE.affinity} module provides a selection of commonly used affinity models but defaults to the standard perplexity-based Gaussian model, as proposed in the original publication.
@@
 By default, \opentsne uses an empirically-determined heuristic to choose the fastest algorithm for a given data set based on its number of samples.
-If the data set contains fewer than 10,000 samples, the Barnes-Hut approximation scheme is selected, while FIt-SNE is used for larger data sets.
+If the data set contains fewer than 10,000 samples, the Barnes-Hut approximation scheme is selected, otherwise, the FIt-SNE is used.
 

### 2023-05-26 2023-05-revision-1/response-to-reviewers.tex
paper: add comparison to UMAP

@@
 \answer{
-TODO bla
+Initially, we decided to omit a comparison to UMAP, as it is a different algorithm and a one-to-one comparison is difficult and not the focal point of the manuscript.
+However, given the popularity of both algorithms, we agree it makes sense to include a benchmark comparison between their implementations.
+However, given the often misunderstood differences between the two algorithms, we feel it is necessary to provide a throughout discussion on their benchmark results.
+We have added a subsection in Section 5.2, comparing the performance of the \texttt{openTSNE} and \texttt{umap-learn} libraries.
+The uniform affinity model is also an interesting and useful addition in the openTSNE framework, and we have added a comparison to that as well, as it could also be used in real-world scenarios when embedding massive data sets.
 }

### 2023-05-29 benchmarks/generate_figures.py
benchmarks: add comparison to UMAP in generate_figures.py

@@ -113,6 +113,12 @@ implementations = [
     Spec("openTSNE BH", log_name="openTSNEBH", cores=1, lang="Python", parser=parse_opentsne),
     Spec("openTSNE BH", log_name="openTSNEBH", cores=8, lang="Python", parser=parse_opentsne),
 
+    # Modern openTSNE
+    Spec("openTSNE FFT (modern)", log_name="openTSNE_standard", cores=1, lang="Python", parser=parse_opentsne),
+    Spec("openTSNE FFT (modern)", log_name="openTSNE_standard", cores=8, lang="Python", parser=parse_opentsne),
+    Spec("openTSNE FFT (uniform)", log_name="openTSNE_uniform30", cores=1, lang="Python", parser=parse_opentsne),
+    Spec("openTSNE FFT (uniform)", log_name="openTSNE_uniform30", cores=8, lang="Python", parser=parse_opentsne),
+
     # FIt-SNE
     Spec("FIt-SNE", log_name="FItSNE", cores=1, lang="Python", parser=parse_fitsne),
     Spec("FIt-SNE", log_name="FItSNE", cores=8, lang="Python", parser=parse_fitsne),
@@ -126,8 +132,8 @@ implementations = [
     Spec("scikit-learn", log_name="sklearn", cores=8, lang="Python", parser=parse_sklearn),
 
     # UMAP
-    Spec("UMAP", log_name="UMAP", cores=1, lang="Python", parser=parse_umap),
-    Spec("UMAP", log_name="UMAP", cores=8, lang="Python", parser=parse_umap),
+    Spec("UMAP", log_name="UMAP_standard", cores=1, lang="Python", parser=parse_umap),
+    Spec("UMAP", log_name="UMAP_standard", cores=8, lang="Python", parser=parse_umap),
 
     # Rtsne
     Spec("Rtsne", log_name="Rtsne", cores=1, lang="R", parser=parse_rtsne),
@@ -145,7 +151,7 @@ for impl in implementations:
 data = pd.DataFrame(data)
 
 
-def generate_benchmark_plot(colors, title, fname):
+def generate_benchmark_plot(colors, title, fname, ylim=130):
     fig, ax = plt.subplots(figsize=(8, 6))
     sns.despine(offset=20)
 
@@ -197,8 +203,8 @@ def generate_benchmark_plot(colors, title, fname):
[…]

### 2023-05-29 2023-05-revision-1/main.tex
paper: update benchmarks figures and finalize umap figure

@@
   \centering
-  \includegraphics[width=0.75\textwidth]{benchmarks_python-final}
+  \includegraphics{benchmarks_python-final}
   \caption{\label{fig:benchmarks_py}
@@
   \centering
-  \includegraphics[width=0.75\textwidth]{benchmarks_langs-final}
+  \includegraphics{benchmarks_langs-final}
   %\includegraphics[width=\textwidth]{benchmarks-combined}
@@
 
+\subsubsection{Comparison to UMAP}
+
+In this section, we compare the performance of the \opentsne (v1.0.0) and \pkg{umap-learn} (0.5.3) libraries.
+\pkg{umap-learn} is the official \textsf{Python} implementation of the UMAP algorithm~\citep{2018arXivUMAP}, another dimensionality reduction technique used for the visualization of high-dimensional data.
+A throughout comparison of the two algorithms is outside the scope of this manuscript, and we here instead highlight only their similarities and differences as they relate to the runtime of the implementations.
+
+Both t-SNE and UMAP are instances of force-directed layout algorithms in which each data point acts as a particle, attracting and repelling other particles.
+In the first step, both algorithms use the high-dimensional data set to construct a $k$-nearest neighbors graph (KNNG).
+Then, the low-dimensional embedding is optimized such that data points connected in the KNNG attract one another, and all data points exert some small, repulsive force.
+The two methods differ in the KNNG construction process, the specification of the attractive and repulsive forces, and their optimization approaches.
+
+The primary factor dictating the runtime of both methods is the number of nearest neighbors used when constructing the KNNG.
+Apart from the runtime required to find the $k$-nearest neighbors, the number of neighbors directly influences the runtime of each optimization step.
+In each step, every data point exerts an attractive force on each of its neighbors in the low-dimensional embedding.
+Therefore, the larger the number of neighbors considered, the longer the runtime.
+Modern implementations of the t-SNE algorithm construct a KNNG with $3 \cdot u$ nearest neighbors where $u$ is the desired perplexity.
+By default, \opentsne (v1.0.0) uses a perplexity value of 30, meaning that 90 nearest neighbors need to be found and considered for each data point.
+In contrast, the \pkg{umap-learn} package (v0.5.3) constructs a KNNG using only 15 neighbors by default, greatly reducing the time spent on nearest neighbor search and point interactions compared to t-SNE.
+
+Alternatively, empirical evidence indicates that replacing the standard perplexity-based Gaussian affinity model in t-SNE with a uniform affinity model produces markedly similar embeddings to that of standard t-SNE.
+For instance, running standard t-SNE with perplexity $u=30$ constructs a KNNG with $3 \cdot u = 90$ neighbors.
+However, when using the equivalent uniform affinity model, only $u$ neighbors are needed.
+This makes the uniform affinity model an attractive option for the visualization of large data sets.
+
+In Fig.~\ref{fig:benchmarks_umap}, we include benchmarks of the standard perplexity-based t-SNE algorithm, its uniform affinity model variant, and UMAP.
+Differently from the previous benchmarks which were run using older parameter settings to accomodate different t-SNE implementations, we now run both libraries using their default parameter settings to measure the runtime of typical, real-world use-cases.
+\begin{figure*}[ht]
+  \centering
+  \includegraphics{benchmarks_umap-final}
+  \caption{\label{fig:benchmarks_umap} We benchmark \opentsne (v1.0.0) against \pkg{umap-learn} (v0.5.3), the official \textsf{Python} implementation of the UMAP algorithm.
+Notice that the standard single-threaded perplexity-based variant of t-SNE runs slower than UMAP but outperforms UMAP in the multi-threaded case.
+We also benchmark the uniform affinity model variant of t-SNE, which outperforms both standard t-SNE and UMAP.}
+\end{figure*}
+
+Given that, by default, UMAP considers six times fewer neighbors than standard t-SNE, it is unsurprising that the single-threaded variant of UMAP outperforms the single-threaded implementation of standard t-SNE.
+Suprisingly, however, the single-threaded uniform affinity variant of t-SNE outperforms not only than the single-threaded variant of UMAP, but also its multi-threaded variant, despite dealing with half as many neighbors.
+This result likely stems from the fact that UMAP is written in pure \textsf{Python}, but uses \pkg{numba}~\citep{lam2015numba} to accelerate computation.
+On the other hand, critical elements of the \opentsne library are written in \pkg{cython}~\citep{behnel2011cython}, which is subsequently transpiled into \textsf{C++}, which may be able to be more aggresively optimized by the \textsf{C++} compiler.
+Another contributing factor may be the efficiency of the approximation schemes.
+While both the UMAP and FIt-SNE algorithms are asymptotically linear in the number of samples, their real-world runtimes may differ significantly.
+
+The multi-threaded benchmarks show that \opentsne outperforms UMAP in both its standard and uniform variants.
+The reason for this likely lies in the different optimization procedures used by the two algorithms.
+At each step of the optimization, \opentsne computes all pairwise interactions between data points in the embedding.
+While a direct calculation of these interactions would result in quadratic time complexity, \opentsne implements efficient approximation schemes that reduce this to linear time complexity (see Sec.~\ref{sec:meth.approx}).
+Importantly, these approximation schemes are very amenable to parallelization.
+This is clearly demonstrated by the runtime differences between the single-core and multi-core performance of \opentsne.
+On the other hand, UMAP bypasses the need to compute all pairwise interactions between data points using negative sampling.
+This allows UMAP to scale linearly in the number of samples without the need for additional approximation schemes.
+However, negative sampling is less amenable to paralellization and does not benefit from the availability of additional cores as much as t-SNE.
+
+Lastly, we turn our attention to an often overlooked runtime metric -- the runtime per optimization step.
+Both t-SNE and UMAP comprise two steps.
+First, both methods construct a KNNG, which runtime is dominated by $k$-nearest neighbor search.
+In the second step, a low-dimensional embedding is found via optimization.
+By default, \opentsne runs optimization for 750 iterations.
+UMAP, on the other hand, uses a heuristic based on the data set size to determine the number of iterations.
+For data sets containing fewer than 10,000 data points, UMAP opts for 500 iterations.
+For larger data sets, however, this is reduced to 200 iterations.
+Fig.~\ref{fig:benchmarks_umap} indicates that even in the single-threaded scenario, the perplexity-based variant of \opentsne requires significantly less time per iteration compared to \pkg{umap-learn}, as each \opentsne benchmark runs $3.75$ times more iterations than \pkg{umap-learn} despite considering more neighbors at each optimization step.
+
 \subsection{Ease of use}

### 2023-05-31 benchmarks/generate_figures.py
benchmarks: update openTSNE run to v1.0.0

@@ -290,6 +290,6 @@ generate_benchmark_plot(
     colors,
     title="Comparison with UMAP",
     fname=os.path.join("..", "paper", "benchmarks_umap.pdf"),
-    ylim=65,
+    ylim=50,
 )
 plt.show()

### 2023-06-02 2023-05-revision-1/main.tex
minor changes

--- 2023-05-25 2023-05-revision-1/main.tex
@@
 and represents the unnormalized $q_{ij}$.
-Alternatively, this gradient can be rewritten as
+We can rewrite this gradient to
 \begin{equation}
@@
 where $Z = \sum_{k \neq l}\left ( 1 + || \mathbf{y}_k - \mathbf{y}_l ||^2 \right )^{-1}$.
-This formulation implies that the t-SNE algorithm can be interpreted as a force-directed layout algorithm, where individual data points act as particles exerting both attractive and repulsive on each other until a state of equilibrium is reached.
+With this formulation, we can interpret t-SNE algorithm as a force-directed layout algorithm, where individual data points act as particles exerting both attractive and repulsive on each other until a state of equilibrium is reached.
 
@@
 Similarly to standard t-SNE, a direct calculation of gradients takes $\mathcal{O}(N \cdot M)$ time, but it is straightforward to adapt the Barnes-Hut and FIt-SNE approximation schemes, reducing the time complexity to $\mathcal{O}(M \log N)$ and $\mathcal{O}(M)$, respectively.
-Special care must be taken to adjust the learning rate during optimization as the parameter values used during standard optimization lead to highly unstable behaviour.
-Empirically, we have found that reducing the learning rate to $\eta \sim 0.1$ results in more stable optimization, as discussed in \citet{policar2021embedding}.
-Moreover, during standard optimization, data points are positioned in the embedding such that an equilibrium between the attractive and repulsive forces is achieved.
+Special care must be taken to adjust the learning rate during optimization as the parameter values used during standard optimization may lead to  unstable behaviour.
+We have observed that reducing the learning rate to $\eta \sim 0.1$ results in more stable optimization, as discussed in \citet{policar2021embedding}.
+Standard optimization positions data points in the embedding to achieve an equilibrium between the attractive and repulsive forces.
 However, when embedding new data points, the reference embedding remains fixed.
-Consequently, if a new data point needs to be positioned between existing reference data points, it may not be possible to reach such an equilibrium, as the reference data points cannot adjust to accomodate the new data point.
-To compensate for the increased repulsive forces, we have found it beneficial to increase the exaggeration factor to $\rho=1.5$ during optimization.
-The training dynamics of embedding new data points into existing embeddings are yet fully understood and are left to future exploration.
+Consequently, if a new data point needs to be positioned among existing reference data points, it may not be possible to reach such an equilibrium, as the reference data points cannot adjust to accomodate the new data point.
+To compensate for the increased repulsive forces, it is beneficial to increase the exaggeration factor to $\rho=1.5$ during optimization.
 
@@
 
+\subsubsection{Comparison to UMAP}
+
+We compare the performance of the \opentsne (v1.0.0) and \pkg{umap-learn} (0.5.3) libraries.
+\pkg{umap-learn} is the official \textsf{Python} implementation of the UMAP algorithm~\citep{2018arXivUMAP}, an alternative dimensionality reduction technique used for the visualization of high-dimensional data.
+A throughout comparison of the two algorithms is outside the scope of this manuscript, and we here instead highlight only their similarities and differences as they relate to the runtime of the implementations.
+
+Both t-SNE and UMAP are force-directed layout algorithms in which each data point acts as a particle, attracting and repelling other particles.
+In the first step, both algorithms use the high-dimensional data set to construct a $k$-nearest neighbors graph (KNNG).
+Then, the low-dimensional embedding is optimized such that data points connected in the KNNG attract one another, and all data points exert some small, repulsive force.
+The two methods differ in the KNNG construction process, the attractive and repulsive forces specification, and their optimization approaches.
+
+The primary factor dictating the runtime of both methods is the number of nearest neighbors used when constructing the KNNG.
+Apart from the runtime required to find the $k$-nearest neighbors, the number of neighbors directly influences the runtime of each optimization step.
+In each step, every data point exerts an attractive force on each neighbor in the low-dimensional embedding.
+Therefore, the larger the number of neighbors considered, the longer the runtime.
+Modern implementations of the t-SNE algorithm construct a KNNG with $3 \cdot u$ nearest neighbors where $u$ is the desired perplexity.
+By default, \opentsne (v1.0.0) uses a perplexity value of 30, meaning that 90 nearest neighbors must be found and considered for each data point.
+In contrast, the \pkg{umap-learn} package (v0.5.3) constructs a KNNG using only 15 neighbors by default, substantially reducing the time spent on nearest neighbor search and point interactions compared to t-SNE.
+
+Alternatively, empirical evidence indicates that replacing the standard perplexity-based Gaussian affinity model in t-SNE with a uniform affinity model produces markedly similar embeddings to that of standard t-SNE.
+For instance, running standard t-SNE with the perplexity of $u=30$ constructs a KNNG with $3 \cdot u = 90$ neighbors.
+However, when using the equivalent uniform affinity model, only $u$ neighbors are needed, making the uniform affinity model attractive for visualizing large data sets.
+
+In Fig.~\ref{fig:benchmarks_umap}, we include benchmarks of the standard perplexity-based t-SNE algorithm, its uniform affinity model variant, and UMAP.
+Differently from the previous benchmarks, which were run using older parameter settings to accommodate different t-SNE implementations, we now run both libraries using their default parameter settings to measure the runtime of typical, real-world use cases.
+\begin{figure*}[ht]
+  \centering
+  \includegraphics{benchmarks_umap-final}
+  \caption{\label{fig:benchmarks_umap} We benchmark \opentsne (v1.0.0) against \pkg{umap-learn} (v0.5.3), the official \textsf{Python} implementation of the UMAP algorithm.
+Notice that the standard single-threaded perplexity-based variant of t-SNE runs slower than UMAP but outperforms UMAP in the multi-threaded case.
+We also benchmark the uniform affinity model variant of t-SNE, which outperforms both standard t-SNE and UMAP.}
+\end{figure*}
+
+Given that, by default, UMAP considers six times fewer neighbors than standard t-SNE, it is unsurprising that the single-threaded variant of UMAP outperforms the single-threaded implementation of standard t-SNE.
+Surprisingly, however, the single-threaded uniform affinity variant of t-SNE outperforms the single-threaded variant of UMAP and its multi-threaded variant, despite dealing with half as many neighbors.
+This result likely stems from the fact that UMAP is written in pure \textsf{Python} but uses \pkg{numba}~\citep{lam2015numba} to accelerate computation.
+On the other hand, we wrote critical elements of the \opentsne library in \pkg{cython}~\citep{behnel2011cython}, which is subsequently transpiled into \textsf{C++}, which may be able to be more aggressively optimized by the \textsf{C++} compiler.
+Another contributing factor may be the efficiency of the approximation schemes.
+While the UMAP and FIt-SNE algorithms are asymptotically linear in the number of samples, their real-world runtimes may differ significantly.
+
+The multi-threaded benchmarks show that \opentsne outperforms UMAP in standard and uniform variants.
+We can attribute this to different optimization procedures used by the two algorithms.
+At each optimization step, \opentsne computes all pairwise interactions between data points in the embedding.
+While directly calculating these interactions would result in quadratic time complexity, \opentsne implements efficient approximation schemes that reduce this to linear time complexity (see Sec.~\ref{sec:meth.approx}).
+Importantly, these approximation schemes are very amenable to parallelization.
+The runtime differences between the single-core and multi-core performance of \opentsne clearly demonstrate this.
+On the other hand, UMAP bypasses the need to compute all pairwise interactions between data points using negative sampling, allowing it to scale linearly in the number of samples without additional approximation schemes.
+However, negative sampling is less amenable to parallelization and does not benefit from the availability of additional cores as much as t-SNE.
+
+Lastly, we focus on an often overlooked runtime metric -- the runtime per optimization step.
+Both t-SNE and UMAP comprise two steps.
+First, both methods construct a KNNG, which runtime is dominated by $k$-nearest neighbor search.
+In the second step, a low-dimensional embedding is found via optimization.
+By default, \opentsne runs optimization for 750 iterations.
+UMAP, on the other hand, uses a heuristic based on the data set size to determine the number of iterations.
+For data sets containing fewer than 10,000 data points, UMAP opts for 500 iterations.
+For larger data sets, however, this is reduced to 200 iterations.
+Fig.~\ref{fig:benchmarks_umap} indicates that even in the single-threaded scenario, the perplexity-based variant of \opentsne requires significantly less time per iteration compared to \pkg{umap-learn}, as each \opentsne benchmark runs $3.75$ times more iterations than \pkg{umap-learn} despite considering more neighbors at each optimization step.
+
 \subsection{Ease of use}

### 2023-06-02 2023-05-revision-1/response-to-reviewers.tex
minor changes

--- 2023-05-25 2023-05-revision-1/response-to-reviewers.tex
@@
 \answer{
-We apologize for this oversight and would like to thank the reviewer for pointing out the availability of PyPI/conda packages for FIt-SNE. We have updated the text and Table 1 to reflect this.
+We apologize for this oversight and would like to thank the reviewer for pointing out the availability of PyPI/conda packages for FIt-SNE.
+We have updated the text and the respective content of Table 1.
 }
@@
 
-\answer{We have now included a sentence in Section 3 on page 8 that describes our heuristic.}
+\answer{We have now included a sentence in Section 3 on page 8 that describes our heuristics.}
 
@@
 \answer{
-We have reworded the paragraph so that the uniform affinity kernel is now briefly described.
+We have reworded the paragraph to describe the uniform affinity kernel.
 }
@@
 \answer{
-TODO bla
+Initially, we decided to omit a comparison to UMAP, as it is a different algorithm, and a one-to-one comparison is difficult and not the focal point of the manuscript.
+However, given the popularity of both algorithms, it makes sense to include a benchmark comparison between their implementations.
+Given the often misunderstood differences between the two algorithms, it is necessary to provide a throughout discussion of their benchmark results.
+We have added a subsection in Section 5.2, comparing the performance of the \texttt{openTSNE} and \texttt{umap-learn} libraries.
+The uniform affinity model is also an interesting and valuable addition to the openTSNE framework, and we have added a comparison to that as well, as it could also be used in real-world scenarios when embedding massive data sets.
 }
@@
 
-\answer{We would like to thank the reviewer for the suggestion, and we have now moved the equation from section 2.2 to the end of section 2.1, and updated the text accordingly.}
+\answer{We would like to thank the reviewer for the suggestion. We have now moved the equation from section 2.2 to the end of section 2.1, and updated the text accordingly.}
 
@@
 
-\answer{We agree that this is somewhat crpytic and have added a sentence below the code snippet describing what each of the function calls does.}
+\answer{We agree and have added a sentence below the code snippet describing what each of the function calls does.}
 
@@
 
-\answer{We agree that this sentence is confusing and have reworded it to be more clear.}
+\answer{We have reworded this section of text accordingly.}
 

### 2023-06-03 2023-05-revision-1/main.tex
minor changes in appendix
paper: minor typos

@@
 and represents the unnormalized $q_{ij}$.
-We can rewrite this gradient to
+We can rewrite this gradient as
 \begin{equation}
@@
 Alternatively, empirical evidence indicates that replacing the standard perplexity-based Gaussian affinity model in t-SNE with a uniform affinity model produces markedly similar embeddings to that of standard t-SNE.
-For instance, running standard t-SNE with the perplexity of $u=30$ constructs a KNNG with $3 \cdot u = 90$ neighbors.
+For instance, running standard t-SNE with a perplexity of $u=30$ constructs a KNNG with $3 \cdot u = 90$ neighbors.
 However, when using the equivalent uniform affinity model, only $u$ neighbors are needed, making the uniform affinity model attractive for visualizing large data sets.
@@
 We here describe the preprocessing steps applied to the single-cell RNA-seq data sets used in Fig.~\ref{fig:macosko}~\citep{macosko2015highly}, Fig.~\ref{fig:tasic}~\citep{tasic2018shared}, Fig.~\ref{fig:cao}~\citep{cao2019single}, and Fig.~\ref{fig:transform}~\citep{hochgerner2018conserved}.
-The standard single-cell preprocessing pipeline consists of gene filtering, gene selection, data normalization, and dimensionality reduction.
+The standard single-cell preprocessing pipeline that have used consists of gene filtering, gene selection, data normalization, and dimensionality reduction.
 

### 2023-08-21 benchmarks/run.sh
benchmarks: fixup run.sh

@@ -25,22 +25,26 @@ install_fitsne() {
 }
 
 install_python_dependencies() {
-  conda install python numpy scikit-learn -y
+  conda install python pip=22.2.2 numpy scikit-learn -y
 
   pip install opentsne --no-binary opentsne
   pip install -r requirements-benchmarks.txt
+}
 
-  install_fitsne
+install_r () {
+  conda install -c r -c conda-forge r -y
 }
 
 install_r_dependencies() {
-  conda install -c r -c conda-forge r -y
-  Rscript -e 'install.packages("Rtsne", repos="https://cloud.r-project.org")'
-  Rscript -e 'install.packages("optparse", repos="https://cloud.r-project.org")'
+  Rscript -e 'if (!require("Rtsne")) install.packages("Rtsne", repos="https://cloud.r-project.org")'
+  Rscript -e 'if (!require("optparse")) install.packages("optparse", repos="https://cloud.r-project.org")'
+}
+
+install_julia () {
+  conda install -c conda-forge julia -y
 }
 
 install_julia_dependencies() {
-  conda install -c conda-forge julia -y
   julia -e 'using Pkg; Pkg.add("ArgParse")'
   julia -e 'using Pkg; Pkg.add("CSV")'
   julia -e 'using Pkg; Pkg.add("DataFrames")'
@@ -97,7 +101,7 @@ if $small; then
   echo "Running small benchmark suite."
 
   dataset_name="macosko_2015"
-  sample_sizes=(1000 2000)
+  sample_sizes=(1000 2000 5000)
   repetitions=3;
 
 else
@@ -119,11 +123,19 @@ if ! conda env list | grep "\s*$CONDA_ENVIRONMENT_NAME\s*" >/dev/null 2>&1; then
   conda create --name "$CONDA_ENVIRONMENT_NAME" -y
   activate_environment "$CONDA_ENVIRONMENT_NAME"
   install_python_dependencies
-  install_r_dependencies
-  install_julia_dependencies || true  # allowed to fail on mac
+  install_r
+  install_julia || true # allowed to fail on mac
 else
   activate_environment "$CONDA_ENVIRONMENT_NAME"
 fi
+
+# Install FIt-SNE, R, and Julia dependencies separately, since replicating
+# the conda environment doesn't install these packages
+if [ ! -d "FIt-SNE" ]; then
+  install_fitsne
+fi
+install_r_dependencies
[…]

### 2023-08-22 2023-08-final-revision/response-to-reviewers.tex
paper: add final revision files

\documentclass{article}

\usepackage[T1]{fontenc}
\usepackage{natbib}
\usepackage{todonotes}
\usepackage{gensymb}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage[top=1in,bottom=1in,left=1in,right=1in]{geometry}
\usepackage{alltt}
\usepackage{framed}
\usepackage{xcolor}
\usepackage{comment}
\usepackage{graphicx}

\usepackage{xr}[]
\externaldocument{main}
\usepackage{url}

\renewcommand{\thetable}{\Alph{table}}
\renewcommand{\thefigure}{\Alph{figure}}

\newcommand{\beq}{\begin{equation}}
\newcommand{\eeq}{\end{equation}}
\newcommand{\beqs}{\begin{equation}}
\newcommand{\eeqs}{\end{equation}}
\newcommand{\mx}[1]{\ensuremath{\boldsymbol{\mathrm{{#1}}}}}   % Simple matrix format.

% \newcommand{\review}[1]{\noindent{{\textbf{\textit{#1}}}}\vspace{3mm}}
% \definecolor{colframe}{rgb}{0.01,0.199,0.1}

\newcounter{rtaskno}
\newcommand{\rtask}[1]{\refstepcounter{rtaskno}\label{#1}}

\newcommand{\reviewc}[2]{\begin{leftbar}\noindent\rtask{#1}\ref{#1}. #2\end{leftbar}}
\newcommand{\review}[1]{\begin{leftbar}\noindent #1\end{leftbar}}
\newcommand{\answer}[1]{\noindent #1\\[0mm]}
\newcommand{\copyquote}[1]{\vspace{2mm} {\color{blue} #1}\vspace{2mm}}

\newlength{\leftbarwidth}
\setlength{\leftbarwidth}{3pt}
\newlength{\leftbarsep}
\setlength{\leftbarsep}{10pt}
\renewenvironment{leftbar}{%
    \def\FrameCommand{{\color{black}{\vrule width \leftbarwidth\relax\hspace {\leftbarsep}}}}%
    \MakeFramed {\advance \hsize -\width \FrameRestore }%
}{\endMakeFramed}

% \usepackage[colorlinks=True,citecolor=black]{hyperref}

\begin{document}

\noindent{\bf Poli\v{c}ar, Stra\v{z}ar, and Zupan: 
\\
openTSNE: A Modular Python Library for t-SNE Dimensionality Reduction and Embedding}

\vspace{5mm}
\noindent {\bf RESPONSE TO THE REVIEWERS}

\vspace{5mm}

\noindent Dear Editor and the Reviewers, \\

\vspace{3mm}

\noindent Thank you for your valuable feedback through which we have been able to greatly improve our manuscript. Below, we provide detailed, point-by-point responses to the Editor's and Reviewer's comments and concerns. To facilitate the review process, we also include a version of the manuscript with the highlighted changes from the previous revision.

\vspace{3mm}

\noindent Kind regards,
\vspace{3mm}

\noindent Pavlin Poli\v{c}ar, 
Martin Stra\v{z}ar, 
Bla\v{z} Zupan \\ 
\noindent Ljubljana, August 21, 2023

\vspace{10mm}

% Editor
\subsection*{Editor}

\noindent\textbf{Manuscript style comments}
\reviewc{r1q1}{Code should have enough spaces to facilitate reading.  Please include spaces before and after operators and after commas (unless spaces have syntactical meaning).}

\answer{As previously stated, when formatting the code used in the manuscript, we followed the standard Python code-style guidelines specified in PEP 8 (https://peps.python.org/pep-0008/). The style guide specifies that spaces should sometimes be omitted around operators, particularily when specifying parameter values. This is different from some other popular programming languages, such as R, but is considered standard practice in the Python community.}

\reviewc{r1q2}{As a reminder, please make sure that \textbackslash proglang,  \textbackslash pkg and \textbackslash code have been used for highlighting throughout the paper (including titles and references), except where explicitly escaped.}

\answer{We have reviewed the manuscript and have ensured that all package names, programming language references, and code snippets are surrounded by the appropriate Latex tags.}

\noindent\textbf{References}
\reviewc{r1q3}{Please make sure that all software packages are \textbackslash cite\{\}'d properly.}

\answer{We have ensured that all software packages are cited properly. Where the software package has been published in a journal or at a conference, we have provided the paper citation. Where such a publication is not available, we have followed the guidelines from the online JSS style guide (https://www.jstatsoft.org/style\#how-to-cite-software).}

\reviewc{r1q4}{All references should be in title style.}

\answer{We have reviewed the manuscript and ensured that the capitalization of the title and reference titles matches the title style. We have also ensured that the section titles and figure annotations match the sentence style.}

\noindent\textbf{Code}
\reviewc{r1q5}{As a reminder, please make sure that the files needed to replicate all code/examples within the manuscript are included in a standalone replication script.}

\answer{
We have carefully prepared the replication material so that they may be run as easily as possible and include them in the submission. The replication materials include several scripts that can be run to reproduce both the figures and the benchmark results presented in this manuscript. The benchmarking script can take days or weeks to complete, so we include a smaller, illustratory benchmark run. We also include exact conda environment specifications so that the results can be reproduced exactly. Detailed instructions on each script and reproduction steps are available in the reproducibility materials attached, as well as the accompanying Github repository (https://github.com/pavlin-policar/openTSNE-paper).
}

% Reviewer A
\subsection*{Reviewer A}

\reviewc{r2q1}{Was unable to run the benchmarks on either a linux box or an intel Mac with python 3.7/3.9, but was able to reproduce all the figures. However, I also couldn't resolve the conda environment in the figures directory either.}

\answer{
We would like to thank the reviewer for alerting us that our environment-replication files were Linux-specific. We have now remedied this and have tested the environment-replication script on an OSX system.
}

% Reviewer B
\subsection*{Reviewer B}

\reviewc{r3q1}{Page 5 mentions that Belkina et al suggest learning rate N/12. However, I checked the defaults in openTSNE 1.0 and the default learning rate seems to be N. Would be good to comment on that.}

\answer{
We have added a clarifying sentence to Page 5, and describe the default behaviour of openTSNE in the Implementation section in point 3 on page 8.
}

\reviewc{r3q2}{Page 21 says you used scikit-learn 1.1 for the benchmark, page 27 gives version 1.0. Also, given that there is version 1.2 out and there were some differences to t-SNE implementation in it, I wonder if it makes sense to use 1.2 for benchmarking (not sure if anything is going to change or not).}

\answer{
Thank you for pointing out this oversight, we have corrected the scikit-learn version to 1.1.2 in the caption of Table 1. We have also run a short benchmark comparing the performance of scikit-learn v1.1 vs v1.2 and the newest v1.3. We found no discernible differences in runtime, therefore, we feel that updating the manuscript in any large way is unnecessary. We have, however, added a corresponding footnote on page 20.
}

\reviewc{r3q3}{Page 23 explains that tSNE results with perplexity 30 are similar to the results with uniform affinity with 30 neighbors. It says "empirical evidence indicates" but gives no evidence (there is no figure) and no citations. One possible citation would be Boehm et al. JMLR 2022: https://jmlr.org/papers/v23/21-0055.html, see Figure 2 in there, but they use uniform affinity with 15 neighbors, not 30. By the way, this is the same values as in UMAP. Anyway, I think the authors should either give a reference (and then use the same values as in the reference) or alternatively provide some evidence themselves.
}

\answer{We have added the citation to the text.}

\reviewc{r3q4}{What is not mentioned in the paper, is the library useed for KNNG construction. openTSNE allows several libraries, so it may be good to mention it somewhere, give relevant citations (to Annoy, HNSW, etc.), and also explain that different libraries support different metrics and/or sparsity. And this can affect the runtime. Maybe this can be mentioned in the benchmarking section, or in the comparison to UMAP section, or elsewhere.}

\answer{
We have added a paragraph to the Implementation section in point 1 describing the different avialable nearest neighbor search methods included in openTSNE.
}

\end{document}

### 2023-08-22 2023-08-final-revision/main.tex
paper: add final revision files

--- 2023-06-03 2023-05-revision-1/main.tex
@@
 %% - \Shorttitle{} with LaTeX markup (if any), used as running title
-\title{openTSNE: a modular \proglang{Python} library for t-SNE dimensionality reduction and embedding}
-\Plaintitle{openTSNE: a modular Python library for t-SNE dimensionality reduction and embedding}
-\Shorttitle{openTSNE: Extensible, parallel implementations of t-SNE}
+\title{\pkg{openTSNE}: A Modular \proglang{Python} Library for t-SNE Dimensionality Reduction and Embedding}
+\Plaintitle{openTSNE: A Modular Python Library for t-SNE Dimensionality Reduction and Embedding}
+\Shorttitle{\pkg{openTSNE}: Extensible, parallel implementations of t-SNE}
 
@@
 %% - Should be comma-separated and in sentence case.
-\Keywords{t-SNE, embedding, visualization, dimensionality reduction, \proglang{Python}}
-\Plainkeywords{t-SNE, embedding, visualization, dimensionality reduction}
+\Keywords{t-SNE, Embedding, Visualization, Dimensionality reduction, \proglang{Python}}
+\Plainkeywords{t-SNE, Embedding, Visualization, Dimensionality reduction, Python}
 
@@
 The ever-growing volumes of high-dimensional data sets in machine learning call for efficient dimensionality reduction techniques and their implementations to produce informative data visualizations.
-Popular approaches include principal component analysis, multidimensional scaling, $t$-distributed stochastic neighbor embedding (t-SNE)~\citep{maaten2008visualizing}, and uniform manifold approximation and projections (UMAP)~\citep{2018arXivUMAP}.
+Popular approaches include principal component analysis (PCA), multidimensional scaling, $t$-distributed stochastic neighbor embedding (t-SNE)~\citep{maaten2008visualizing}, and uniform manifold approximation and projections (UMAP)~\citep{2018arXivUMAP}.
 Among these, t-SNE has achieved widespread adoption as it can address high volumes of data and reveal the underlying data manifold.
@@
 We use \opentsne to generate three t-SNE embeddings and demonstrate recent theoretical advances.
-The data in \textbf{(a)} and \textbf{(b)} include 44,808 mouse retinal cells that are described with high-dimensional gene-expression profiles from \citet{macosko2015highly}.
-The data in \textbf{(c)} additionally contains 27,499 expression profiles from mouse retinal cells from \citet{shekhar2016comprehensive}.
-\textbf{(a)} We construct t-SNE embedding following the parameter choices from the original publication by \citet{maaten2008visualizing}.
+The data in (a) and (b) include 44,808 mouse retinal cells that are described with high-dimensional gene-expression profiles from \citet{macosko2015highly}.
+The data in (c) additionally contains 27,499 expression profiles from mouse retinal cells from \citet{shekhar2016comprehensive}.
+(a) We construct t-SNE embedding following the parameter choices from the original publication by \citet{maaten2008visualizing}.
 The visualization shows no preservation of the global organization of clusters, resulting from random initialization and an affinity model focused on preserving local neighborhoods.
-\textbf{(b)} A modern t-SNE embedding, utilizing the latest theoretical advances and practical recommendations constructed using a multi-scale affinity model, preserving both short-range and long-range interactions between data points and initialized so that the global layout is as meaningful as possible.
-Unlike in \textbf{(a)}, the green and blue clusters representing different sub-types of amacrine and bipolar cells are now localized to the same regions of the space, indicating a higher level of similarity than to other cell types.
-The embedding in \textbf{(c)} shows how existing t-SNE reference atlases can be used to place new samples into existing embeddings.
+(b) A modern t-SNE embedding, utilizing the latest theoretical advances and practical recommendations constructed using a multi-scale affinity model, preserving both short-range and long-range interactions between data points and initialized so that the global layout is as meaningful as possible.
+Unlike in (a), the green and blue clusters representing different sub-types of amacrine and bipolar cells are now localized to the same regions of the space, indicating a higher level of similarity than to other cell types.
+The embedding in (c) shows how existing t-SNE reference atlases can be used to place new samples into existing embeddings.
 The positions of new data points correspond to cell types from the reference atlas.
@@
 Easy access to efficient implementations almost universally correlates with the widespread adoption of novel computational techniques.
-Despite the many recent theoretical advancements, popular t-SNE libraries have slowly incorporated them into their implementations.
-In \proglang{Python}, the most commonly used implementation of the t-SNE algorithm comes from \pkg{scikit-learn}~\citep{pedregosa2011scikit}, which is nicely integrated with \proglang{Python} data science environment but suffers from poor scalability.
+Despite the many recent theoretical advancements, popular t-SNE libraries have been slow to incorporate them into their implementations.
+In \proglang{Python}, the most commonly used implementation of the t-SNE algorithm comes from \pkg{scikit-learn}~\citep{pedregosa2011scikit}, which is tightly integrated with the \proglang{Python} data science environment but suffers from poor scalability.
 Other implementations from the \proglang{R}~\citep{team2013r,krijthe2015rtsne} and \proglang{Julia}~\citep{bezanson2017julia,julia_tsne} programming languages also scale poorly (see Sec.~\ref{s:speed}).
-\pkg{MulticoreTSNE}~\citep{Ulyanov2016} enjoys better scalability but is likewise limited when applied to larger data sets.
+Another \proglang{Python} implementation, \pkg{MulticoreTSNE}~\citep{Ulyanov2016}, enjoys better scalability but is likewise limited when applied to larger data sets.
 Additionally, none of these implementations include recently proposed extensions that can result in more globally consistent embeddings.
@@
 However, none of the existing t-SNE implementations support adding new samples into existing embeddings.
-To alleviate all these problems, and to support newly proposed additions to t-SNE algorithms, we here present \opentsne, an open-source \proglang{Python} implementation of the t-SNE algorithm.
+To alleviate all these problems, and to support newly proposed additions to the t-SNE algorithm, we here present \opentsne, an open-source \proglang{Python} implementation of the t-SNE algorithm.
 \opentsne is easy to install and includes precompiled binaries available through popular \proglang{Python} package managers.
@@
 where $Z = \sum_{k \neq l}\left ( 1 + || \mathbf{y}_k - \mathbf{y}_l ||^2 \right )^{-1}$.
-With this formulation, we can interpret t-SNE algorithm as a force-directed layout algorithm, where individual data points act as particles exerting both attractive and repulsive on each other until a state of equilibrium is reached.
+Under this formulation, we can cast t-SNE algorithm as a force-directed layout algorithm, where individual data points act as particles exerting both attractive and repulsive on each other until a state of equilibrium is reached.
 
 Optimization is performed using batch gradient descent using the delta-bar-delta update rule~\citep{jacobs1988increased}.
-The t-SNE optimization procedure consists of two phases: in the first \textit{early exaggeration} phase, the attractive forces between data points are increased by some factor $\rho$, typically set to $12$, so that points in the embedding can more easily move throughout the space and settle near their respective neighbors.
+The t-SNE optimization procedure consists of two phases: in the first \textit{early exaggeration} phase, the attractive forces between data points are increased by some \textit{exaggeration} factor $\rho$, typically set to $12$, so that points in the embedding can more easily move throughout the space and settle near their respective neighbors.
 In the second phase of the optimization, the attractive forces are then reverted to their original values with $\rho=1$.
@@
 \citet{belkina2019automated} later found that convergence can be sped up by
-increasing the learning rate from the standard $\eta=200$ to $\eta=N/12$. As a
-side-effect, embeddings converge faster, and the number of iterations can be
+increasing the learning rate $\eta$ from the standard $\eta=200$ to $\eta = N/\rho$, where $\rho$ here refers to the exaggeration factor used during the early exaggeration phase.
+As a side-effect, embeddings converge faster, and the number of iterations can be
 lowered from the typical 1000 to 750, decreasing the overall runtime.
@@
 \citet{linderman2019fast} proposed an alternative approach, FIt-SNE, based on
-non-uniform convolutions, which further reduces the time complexity to
+non-uniform convolutions and interpolation, which further reduces the time complexity to
 $\mathcal{O}(N)$.
@@
 often arbitrarily scattered in the embedding space. However, when using
-initialization schemes based on PCA or Laplacian eigenmaps~\citep{belkin2001laplacian}, the clusters identified in the resulting embedding are typically grouped in a globally
+initialization schemes based on PCA or Laplacian Eigenmaps~\citep{belkin2001laplacian}, the clusters identified in the resulting embedding are typically grouped in a globally
 coherent manner.
@@
 
+A key component of the affinity model construction process is finding a specified number of nearest neighbors for each data point.
+\opentsne provides several backends for finding the nearest neighbors, each with its own advantages and drawbacks.
+For exact nearest neighbor search, we use the implementation from \pkg{scikit-learn}~\citep{pedregosa2011scikit}.
+While asymptotically slower than approximate nearest-neighbor search methods, exact approaches introduce minimal computation overhead and tend to run faster for smaller data sets.
+The \pkg{annoy}~\citep{annoylib} library offers a fast and efficient implementation of random-projection trees and supports several commonly used metrics but lacks support for sparse matrices.
+The \pkg{pynndescent}~\citep{pynndescent} library supports sparse matrices and allows users to define and use custom metrics.
+Finally, we include the \pkg{hnswlib}~\citep{hnswlib} library as it provides a third alternative algorithm for approximate nearest neighbor search, which may perform better on certain data sets.
+By default, \opentsne automatically selects the most appropriate algorithm for a given data set based on data set size, sparsity, and metric support.
+These can be overridden by the user and easily extended to support custom nearest-neighbor search solutions.
+
 \item The optimization in t-SNE starts with initial positions for each data point in the embedding space.
 We can set the initial positions randomly or impose an initial global structure on the embedding.
-% with, for example, a PCA-based initialization.
 The \code{openTSNE.initialization} module provides several possible initialization schemes including random, PCA-based, and Laplacian Eigenmap-based intialization schemes.
@@
 \item In optimizing the embedding, the t-SNE algorithm finds the positions of the points in the embedding space that reflect their corresponding similarities in the affinity model.
-\opentsne stores the initial positions and affinities in an \code{openTSNE.TSNEEmbedding} object that includes the \code{.optimize} method for embedding optimization that uses either the Barnes-Hut or FIt-SNE approximation scheme.
+\opentsne stores the initial positions and affinities in an \code{openTSNE.TSNEEmbedding} object that includes the \code{.optimize} method for embedding optimization.
+
+We adopt the learning rate $\eta=N/\rho$ proposed by \citet{belkina2019automated}, who suggest using a global learning rate computed from the early exaggeration factor $\rho$.
+Empirically, we have found that adapting the learning rate to the exaggeration factor $\rho$ used in the second phase of the optimization as well further speeds up convergence.
+
+Optimization can be performed using either the Barnes-Hut or FIt-SNE approximation scheme.
 While the FIt-SNE approximation scheme scales linearly with the number of samples, it introduces additional computational overhead, often unnecessary for smaller data sets, resulting in longer overall runtimes than the Barnes-Hut approximation scheme.
@@
 Using an affinity model from reference embedding for the embedding of the new data is the default behavior of \opentsne.
+The same rationale applies to the nearest neighbor search algorithm.
 
@@
   \caption{\label{fig:diagram}
-  High-level structure and information flow of the \opentsne library.
+  The high-level structure and information flow of the \opentsne library.
 Solid black lines show the standard parameter flow, while red dashed lines indicate the utility of optional parameters, allowing more refined control over the t-SNE algorithm.
@@
 Fig.~\ref{fig:diagram} shows the high-level structure and information flow of the \opentsne library.
-Through different parameters, the user is given complete control over every aspect of the t-SNE data visualization algorithm, from different ways to define affinities to the delicate tuning of optimization parameters.
+Through different parameters, the user is given complete control over every aspect of the t-SNE algorithm, from different ways to define affinities to the delicate tuning of optimization parameters.
 The parametrization of every part of t-SNE allows for rapid experimentation.
@@
 
-\opentsne is designed to be easy to use and can be installed through either \proglang{Python} Package Index \pkg{PyPI} using package installer \pkg{pip},
+\opentsne is designed to be easy to use and can be installed through either \proglang{Python} Package Index \pkg{PyPI} using the \pkg{pip} package manager,
 \begin{CodeChunk}
@@
 \end{CodeChunk}
-or using package manager \pkg{conda} from the \pkg{conda-forge} repository,
+or using the \pkg{conda} package manager from the \pkg{conda-forge} repository,
 \begin{CodeChunk}
@@
 This small data set profiles 150 iris flowers with four variables (sepal length, sepal width, petal length, petal width) and provides information about their species ({\em Iris Setosa}, {\em Iris Versicolor}, {\em Iris Virginica}).
-Iris data set comes with \pkg{scikit-learn}:
+The Iris data set comes bundled with \pkg{scikit-learn}:
 \begin{CodeChunk}
@@
 While the above example offers users a quick and easy way to generate informative visualizations, \opentsne allows us to control every step of the t-SNE embedding.
-To demonstrate this flexibility, we will repeat the exact analysis of the Iris data set as before, this time leveraging the library's more advanced features.
+To demonstrate this flexibility, we will repeat the same analysis of the Iris data set as before, this time leveraging the library's more advanced features.
 We list other practical examples with more complex data sets in Section~\ref{sec:discussion}.
@@
 Notice that the embedding object contains the same point coordinates as the initialization: although we have created an \code{openTSNE.TSNEEmbedding} object, we have not yet run any optimization.
-To optimize the embedding, we have to call the \code{embedding.optimize} method using our desired parameter settings.
+To optimize the embedding, we call the \code{embedding.optimize} method using our desired parameter settings.
 We use the default parameter settings here, including a shorter early-exaggeration phase, followed by a longer phase with no exaggeration.
@@
 However, there is no evidence that every data set can accurately be described using only two dimensions, and any such embedding will inevitably lead to a loss of information.
-Thus, it is beneficial to examine multiple embeddings, each providing a different perspective on topology and other data characteristics.
+Thus, it is beneficial to examine multiple embeddings, each providing a different perspective on the topology and other data characteristics.
 
@@
   We use \opentsne to create four different
-  visualizations of the \citet{tasic2018shared} data,
+  visualizations of the \citet{tasic2018shared} data set,
   each providing a different perspective into the topology of the data.
@@
   correspond to inhibitory neurons, and gray/brown colors correspond to
-  non-neuronal cells. Standard t-SNE \textbf{(a)} emphasizes local
-  structure while increasing perplexity \textbf{(b)} results in a more
+  non-neuronal cells. Standard t-SNE (a) emphasizes local
+  structure while increasing perplexity (b) results in a more
   meaningful layout of the clusters. We can also combine the two
-  perplexities by using a multi-scale kernel affinity model \textbf{(c)}
+  perplexities by using a multi-scale kernel affinity model (c)
   and obtain a trade-off between global and local structures.
   Alternatively, we can inspect more fine-grained structures and reveal
-  smaller clusters by using a more heavy-tailed kernel \textbf{(d)}.}
+  smaller clusters by using a more heavy-tailed kernel (d).}
 \end{figure*}
@@
 
-When dealing with data containing millions of data points, standard t-SNE embeddings often become unwieldy -- cluster boundaries are blurred, large clusters absorb smaller ones, and relationships between groups become increasingly difficult to interpret.
+When dealing with data sets containing millions of data points, standard t-SNE embeddings often become unwieldy -- cluster boundaries are blurred, large clusters absorb smaller ones, and relationships between groups become increasingly difficult to interpret.
 We constructed Fig.~\ref{fig:cao}.a from the data containing expression profiles of over two million single cells captured at different time points in mouse development.
@@
   \caption{\label{fig:cao}
-  Increasing the exaggeration parameter leads to compact clusters, highlighting the data's global organization and emphasizing the continuous nature of cell state transitions.
+  Increasing the exaggeration factor leads to compact clusters, highlighting the data's global organization and emphasizing the continuous nature of cell state transitions.
 The data set from \citet{cao2019single} contains expression profiles from 2,058,652 single cells.
 The data were collected from mice embryos at different developmental stages daily after 9.5 to 13.5 days.
-\textbf{(c)} reveals that the data is comprised of two main components -- the neural tube and mesenchymal cells -- as well as several other smaller clusters.
+(c) reveals that the data is comprised of two main components -- the neural tube and mesenchymal cells -- as well as several other smaller clusters.
 The colors indicate developmental progression, where red indicates the least-developed cells and blue indicates the most developed cells.
 The overall developmental trajectory is most apparent with higher exaggeration levels, showing red cells slowly transitioning into blue cells.
-Progressively easing the exaggeration factor uncovers finer clusters within the larger groups, as shown in \textbf{(b)} with exaggeration of two and subsequently in \textbf{(a)}, where we show the standard t-SNE with no exaggeration.
+Progressively easing the exaggeration factor uncovers finer clusters within the larger groups, as shown in (b) with exaggeration of two and subsequently in (a), where we show the standard t-SNE with no exaggeration.
 32,011 putative doublets are excluded from the visualizations.
@@
 
-Unlike other popular dimensionality reduction techniques such as principal component analysis or autoencoders, t-SNE is a non-parametric method and does not define an explicit mapping to the embedding space.
+Unlike other popular dimensionality reduction techniques such as PCA or autoencoders, t-SNE is a non-parametric method and does not define an explicit mapping to the embedding space.
 Therefore, embeddings of new data points need to be found through optimization~\citep{policar2021embedding}.
@@
 One of the t-SNE's common criticisms is limited scalability to large data sets containing, for instance, millions of data points~\citep {becht2019dimensionality}.
-Slow response times stem from the optimization procedures and their specific implementation in popular open-source libraries.
+Slow response times stem from the optimization procedures and their specific implementations in popular open-source libraries.
 Most implementations of t-SNE have been based on either a direct implementation of the t-SNE algorithm with asymptotic time complexity $\mathcal{O}(N^2)$~\citep{maaten2008visualizing}, or the Barnes-Hut approximation with asymptotic time complexity $\mathcal{O}(N \log N)$~\citep{van2014accelerating}.
@@
 % We benchmark Python
-Fig.~\ref{fig:benchmarks_py} benchmarks four popular \proglang{Python}~\citep{vanrossum1995python} t-SNE implementations, including \pkg{scikit-learn} (v1.1.2), \pkg{MulticoreTSNE} (v0.1), \pkg{FIt-SNE} (v1.1.0), and \opentsne (v1.0.0).
+Fig.~\ref{fig:benchmarks_py} benchmarks four popular \proglang{Python}~\citep{vanrossum1995python} t-SNE implementations, including \pkg{scikit-learn} (v1.1.2)\footnote{We also compared the differences in performance between \pkg{scikit-learn} v1.1.2 and more the more recent releases v1.2.2 and v1.3.0, but found no discernible differences in the overall runtime.}, \pkg{MulticoreTSNE} (v0.1), \pkg{FIt-SNE} (v1.1.0), and \opentsne (v1.0.0).
 Benchmarks were performed on an Intel(R) Xeon(R) CPU E5-1650 equipped with 128 GB of memory.
@@
   (v1.1.2), \pkg{MulticoreTSNE}~\citep{Ulyanov2016} (v0.1), and
-  \pkg{FIt-SNE}~\citep{linderman2019fast} (v1.1.0). Experiments were run on a
-  on an Intel(R) Xeon(R) CPU E5-1650 equipped with 128 GB of memory. Notice
+  \pkg{FIt-SNE}~\citep{linderman2019fast} (v1.1.0). Experiments were run on an Intel(R) Xeon(R) CPU E5-1650 equipped with 128 GB of memory. Notice
   that \opentsne scales similarly to \pkg{FIt-SNE}, as they both use the
@@
 
-Alternatively, empirical evidence indicates that replacing the standard perplexity-based Gaussian affinity model in t-SNE with a uniform affinity model produces markedly similar embeddings to that of standard t-SNE.
+Alternatively, empirical evidence indicates that replacing the standard perplexity-based Gaussian affinity model in t-SNE with a uniform affinity model produces markedly similar embeddings to that of standard t-SNE~\citep{bohm2020unifying}.
 For instance, running standard t-SNE with a perplexity of $u=30$ constructs a KNNG with $3 \cdot u = 90$ neighbors.
@@
 Given that, by default, UMAP considers six times fewer neighbors than standard t-SNE, it is unsurprising that the single-threaded variant of UMAP outperforms the single-threaded implementation of standard t-SNE.
-Surprisingly, however, the single-threaded uniform affinity variant of t-SNE outperforms the single-threaded variant of UMAP and its multi-threaded variant, despite dealing with half as many neighbors.
-This result likely stems from the fact that UMAP is written in pure \textsf{Python} but uses \pkg{numba}~\citep{lam2015numba} to accelerate computation.
-On the other hand, we wrote critical elements of the \opentsne library in \pkg{cython}~\citep{behnel2011cython}, which is subsequently transpiled into \textsf{C++}, which may be able to be more aggressively optimized by the \textsf{C++} compiler.
+Surprisingly, however, the single-threaded uniform affinity variant of t-SNE outperforms both the single-threaded variant of UMAP and its multi-threaded variant, despite dealing with twice as many neighbors.
+This result likely stems from the fact that UMAP is written in pure \textsf{Python} and uses \pkg{numba}~\citep{lam2015numba} to accelerate computation.
+On the other hand, we implemented critical elements of the \opentsne library in \pkg{Cython}~\citep{behnel2011cython}, which is subsequently transpiled into \textsf{C++}, which may be able to be more aggressively optimized by the \textsf{C++} compiler.
 Another contributing factor may be the efficiency of the approximation schemes.
@@
 
-The multi-threaded benchmarks show that \opentsne outperforms UMAP in standard and uniform variants.
+The multi-threaded benchmarks show that \opentsne outperforms \pkg{umap-learn} in standard and uniform variants.
 We can attribute this to different optimization procedures used by the two algorithms.
@@
 For larger data sets, however, this is reduced to 200 iterations.
-Fig.~\ref{fig:benchmarks_umap} indicates that even in the single-threaded scenario, the perplexity-based variant of \opentsne requires significantly less time per iteration compared to \pkg{umap-learn}, as each \opentsne benchmark runs $3.75$ times more iterations than \pkg{umap-learn} despite considering more neighbors at each optimization step.
+Fig.~\ref{fig:benchmarks_umap} indicates that even in the single-threaded scenario, the perplexity-based variant of \opentsne requires significantly less time per iteration compared to \pkg{umap-learn}, as each \opentsne benchmark runs $3.75$ times more iterations than \pkg{umap-learn} despite considering six times more neighbors at each optimization step.
 
@@
 Other \proglang{C++} implementations such as \pkg{MulticoreTSNE} and \pkg{FIt-SNE} exhibit better scaling in more massive data sets, but the packages do not provide precompiled binaries and require users to compile the software themselves.
-This problem is critical, for instance, for users of the Windows operating system, where the \proglang{C++} compiler does not come with the system, making the correct configuration of current t-SNE implementations cumbersome.
+This problem is critical, for instance, for users of the Windows operating system, where the \proglang{C++} compiler does not come bundled with the system, making the correct configuration of current t-SNE implementations cumbersome.
 
@@
 \caption{\label{tab:features}
-  We compare the features of \opentsne~(v1.0.0) to \pkg{scikit-learn}~(v1.0.1),
+  We compare the features of \opentsne~(v1.0.0) to \pkg{scikit-learn}~(v1.1.2),
   \pkg{BHTSNE}~(master), \pkg{MulticoreTSNE}~(v0.1), and
@@
 
-\section{Single-cell RNA-seq Data Preprocessing}
+\section{Single-cell RNA-seq data preprocessing}
 

### 2023-08-24 2023-08-final-revision/response-to-reviewers.tex
minor changes

@@
 
-\noindent Thank you for your valuable feedback through which we have been able to greatly improve our manuscript.
-Below, we provide detailed, point-by-point responses to the Editor's and Reviewer's comments and concerns.
+\noindent Thank you for your valuable feedback for the final version of the manuscript.
+Below, we provide detailed, point-by-point responses to minor comments and change requests.
 To facilitate the review process, we also include a version of the manuscript with the highlighted changes from the previous revision.
@@
 
-\answer{We have reviewed the manuscript and have ensured that all package names, programming language references, and code snippets are surrounded by the appropriate Latex tags.}
+\answer{We have inspected the manuscript to make sure that all package names, programming language references, and code snippets are surrounded by the appropriate Latex tags.}
 
@@
 Thank you for pointing out this oversight, we have corrected the scikit-learn version to 1.1.2 in the caption of Table 1.
-We have also run a short benchmark comparing the performance of scikit-learn v1.1 vs v1.2 and the newest v1.3.
+We have also run several benchmarks comparing the performance of scikit-learn v1.1 vs v1.2 and the newest v1.3.
 We found no discernible differences in runtime, therefore, we feel that updating the manuscript in any large way is unnecessary.

### 2023-08-24 2023-08-final-revision/main.tex
minor changes

@@
 \opentsne provides several backends for finding the nearest neighbors, each with its own advantages and drawbacks.
-For exact nearest neighbor search, we use the implementation from \pkg{scikit-learn}~\citep{pedregosa2011scikit}.
+We use the implementation from \pkg{scikit-learn}~\citep{pedregosa2011scikit} for the exact nearest neighbor search.
 While asymptotically slower than approximate nearest-neighbor search methods, exact approaches introduce minimal computation overhead and tend to run faster for smaller data sets.
@@
 The \pkg{pynndescent}~\citep{pynndescent} library supports sparse matrices and allows users to define and use custom metrics.
-Finally, we include the \pkg{hnswlib}~\citep{hnswlib} library as it provides a third alternative algorithm for approximate nearest neighbor search, which may perform better on certain data sets.
+Finally, we include the \pkg{hnswlib}~\citep{hnswlib} library as it provides a third alternative algorithm for approximate nearest neighbor search, which may perform better on specific data sets.
 By default, \opentsne automatically selects the most appropriate algorithm for a given data set based on data set size, sparsity, and metric support.
@@
 We adopt the learning rate $\eta=N/\rho$ proposed by \citet{belkina2019automated}, who suggest using a global learning rate computed from the early exaggeration factor $\rho$.
-Empirically, we have found that adapting the learning rate to the exaggeration factor $\rho$ used in the second phase of the optimization as well further speeds up convergence.
+Empirically, we have found that adapting the learning rate to the exaggeration factor $\rho$ used in the second phase of the optimization also further speeds up convergence.
 

### 2023-08-24 2023-08-final-revision/main.tex
paper: keywords should be lower-case

@@
 %% - Should be comma-separated and in sentence case.
-\Keywords{t-SNE, Embedding, Visualization, Dimensionality reduction, \proglang{Python}}
-\Plainkeywords{t-SNE, Embedding, Visualization, Dimensionality reduction, Python}
+\Keywords{t-SNE, embedding, visualization, dimensionality reduction, \proglang{Python}}
+\Plainkeywords{t-SNE, embedding, visualization, dimensionality reduction, Python}
 
